### 2008-08-25 docs/oboe-guide/main.tex

\documentclass[article,oneside]{memoir}

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[colorlinks=true, linkcolor=blue]{hyperref}
\usepackage{memhfixc}
\usepackage{listings}

\oddsidemargin 0.0in
\textwidth 6.5in 
\topmargin 0.0in 
\textheight 8.5in

\newcommand{\obs}{\textsc{oboe}}
\newcommand{\owl}{\textsf{OWL}}
\newcommand{\seek}{\href{http://seek.ecoinformatics.org/}{SEEK}}
\lstset{language=XML, basicstyle=\tiny\ttfamily}

\title{The \obs{} Tutorial}
\author{Joshua S. Madin, Shawn Bowers, Mark P. Schildhauer, and Matt Jones}

\begin{document}

\maketitle

\tableofcontents

%%%
\pagebreak
\chapter{Introduction}
The \textbf{E}xtensible \textbf{OB}servation \textbf{O}ntology
(reversed to give \obs{}) was developed to describe scientific
observations and measurements with particular emphasis on capturing
observational context.  This emphasis is important because automated
reasoning tasks, such as searching for data or merging data sets,
require ``knowledge'' of contextual structure in order to determine if
data from different sources are compatible (e.g., equivalent or
subsumable).  For example, if a human or machine process can determine
that two data points have the same temporal and spatial context, then
the points are potentially comparable, and might be merged to increase
the power of a scientific analysis. This tutorial will focus on the
use of \obs{} for annotating ecological and environmental data sets.
A more detailed description of the of \obs{}'s ontological structure
can be found elsewhere (Madin \textit{et al.}, 2007).

\section{Observation and Measurement}

Observations and Measurements form the core of \obs{}; represented as
black ellipses in Figure~\ref{fig:oboe}.  An Observation is defined as
an assertion that a \textit{thing} (physical, abstract, or conceptual)
was observed to belong to a specified class of Entity in a specified
ontology.  For example, a field researcher might assert that an entity
they observed belongs in the class Tree in the ontology portrayed in
Figure~\ref{fig:entity} (i.e., the observed thing is an
\textit{instance} of class Tree which is a subclass of Entity).

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering  
 \includegraphics[width=0.9\textwidth]{./figures/tut_fig_2.png}
 \caption{A simplified representation of \obs{}'s core concepts.
   Ellipses signify ontology classes and arrows signify properties.
   Maximum and minimum cardinalities are given for each property and
   thus indicating the necessary number of connections required for a
   given data instance.}
  \label{fig:oboe}
  \end{minipage}
  }
\end{figure}

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.45\textwidth]{./figures/tut_fig_4.png}
  \caption{A simple ontology of observable entities illustrating a
    continuum from coarser to finer level classification.  The
    property "is-a" applies to class instances (not the classes
    themselves).  As an example, this ontology states that an instance
    of the class Tree is also an instance of the class Plant, but not
    an instance of the class Rock.}
  \label{fig:entity}
  \end{minipage}
  }
\end{figure}

A Measurement is the subsequent documentation of a characteristic of
the observed entity as data (or sometimes metadata).  For example, a
data set may contain a set of tree heights (Table~\ref{tab:lobata}),
where the characteristic Height was measured for a number of instances
of the class \textit{Quercus lobata}.  A Measurement is composed from
four parts (Figure~\ref{fig:oboe}).  The first two parts are the
Characteristic that was measured (e.g., weight, color, or name) and
the Measurement Standard used to make the measurement (e.g., physical
unit or classification scheme), both of which are selected from their
respective ontology extensions of \obs{}; similar to observed Entity.
The other two parts are Value (the actual datum in the data set) and
an estimate of precision, both of which are literals (i.e., not
represented in an ontology).

\begin{table}[htbp]
  \centering
  \begin{tabular}{ @{} c @{} }
    \toprule
    Height (m) \\
    \midrule
    1.2 \\
    9.5 \\
    4.8 \\
    1.5 \\
    2.1 \\
    ... \\
    \bottomrule
  \end{tabular}
  \caption{A data set containing tree heights for the oak \textit{Quercus lobata}.}
 \label{tab:lobata}
\end{table}

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.4\textwidth]{./figures/tut_eg_10_tab1.png}
  \caption{The \obs{} representation of the first data point in
    Table~\ref{tab:lobata}.  A. Entities, Characteristics, and
    Standards are represented as instances of ontology classes.
    B. Measurement Precisions and Values are represented using codes
    or literals.  The other data points in Table~\ref{tab:lobata} are
    represented identically except for Value, and therefore the
    \textsf{hasValue} property is usually omitted in \obs{}
    representations.}
  \label{fig:lobata}
  \end{minipage}
  }
\end{figure}

Generally, the observed Entity is chosen to be the lowest common
denominator for the focal set of data; that is, in
Table~\ref{tab:lobata} all the height measurement data apply to
independent instances of \textit{Quercus lobata} and nothing else
(e.g., not Rock, Bush, nor \textit{Quercus durata}).  However, in
Table~\ref{tab:oaks} the observed Entity is coarser than for
Table~\ref{tab:lobata}, and the lowest common denominator is now Tree.
In this second example, the classification of instances of Tree into
``Species" is now a Measurement, and the Height of each instance is
also measured within the same instance of Tree that was observed
(Figure~\ref{fig:tut_eg_10_tab2}).

\begin{table}[htbp]
  \centering
  \caption{A data set containing heights for multiple oak species.}
  \begin{tabular}{ @{} cc @{} }
    \toprule
    Species & Height (m) \\
    \midrule
    \textit{Q. lobata} & 15.5 \\
    \textit{Q. berberidifolia} & 9.0 \\
    \textit{Q. berberidifolia} & 8.8 \\
    \textit{Q. lobata} & 16.5 \\
    \textit{Q. durata} & 5.1 \\
    ... & ... \\
    \bottomrule
  \end{tabular}
 \label{tab:oaks}
\end{table}

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.45\textwidth]{./figures/tut_eg_10_tab2.png}
  \caption{The \obs{} representation of the first data point in Table~\ref{tab:oaks}.}
  \label{fig:tut_eg_10_tab2}
  \end{minipage}
  }
\end{figure}

These examples illustrate several key aspects of \obs{}:

\begin{itemize}

\item An Observation implies a \textit{definite} assertion according
  to a given ontology (or world view).  For example, the Heights
  measured in Table~\ref{tab:lobata} apply to instances of a category
  of things called \textit{Quercus lobata} according to the world view
  shown in Figure~\ref{fig:entity}.  But world views differ between
  research groups and change through time (e.g., one group might clump
  instances of the class \textit{Quercus lobata} with \textit{Quercus
    durata}).  \obs{} does not attempt to capture or track mappings
  among different world views.

\item A Measurement is, in essence, the further classification of an
  entity based its characteristics (e.g., an instance of the class
  ``7.45 Meter High Tree'') but implies an inherent degree of
  \textit{uncertainty}.  An observed entity can be measured (or
  classified) more than once (e.g., height, width, and name).
  However, each measurement must apply to the \textit{whole} entity.
  For example, it is not correct to observe a Lion and measure the
  length of its tail within the same observation.  In this case, the
  observation of Lion provides \textit{context} for a second
  observation of Tail (discussed in the next section).  Measurements
  can be made at each level of observation; e.g., the length of the
  Lion and the length of the Tail (which is associated to the
  particular instance of Lion via context).

\end{itemize}

\pagebreak
\section{Context: The Backbone of Observation}

Context forms the critical backbone of observational data.  That is,
Observations and their corresponding Measurements can only be
understood within the context they are made.  For example, there are
few interesting questions that can be asked about a set of tree
heights without any contextual information (e.g., the data in
Table~\ref{tab:lobata}).  Whereas, if there is also observational
information about location where the trees were observed, then these
data are potentially comparable with measurements taken at different
locations.  Correctly representing the contextual structure of a data
set use \obs{} takes practice, but if done correctly provides a
powerful mechanism for automatically determining the compatibility and
mergability of data.

A data set (including metadata) will typically contain information
about a object or phenomena observed, as well as the time and place it
was observed.  However, several contextual structures can potentially
describe the relationships among the three entities---e.g., Time,
Place, and Object (Figure~\ref{fig:tut_eg_11C}).  At this point, it is
important to understand that observing what we might be considered the
``same" Entity at a different time or place typically results in a
different instance of Entity.  For example, trees change through time
(i.e., they are perdurants).  In this sense, observing perdurant
Entities at different times can be treated similarly to observing them
at different places: they do not represent the same instance of
Entity.

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.8\textwidth]{./figures/tut_eg_11C.png}
  \caption{Examples of Entity instance dependence in space and time.
    Panel \textbf{A} illustrates the case when place and time provide
    independent context for Tree instance observations; that is, each
    combination of place and time contains a different (unique)
    instance of Tree.  Panel \textbf{B} shows the case when space is
    dependent of time.  This might happen if space changes through
    time, such as when space is sampled randomly at different time
    steps.  Finally, panel \textbf{C} shows the case when time is
    dependent of place.  This might happen when time-series are taken
    at different places or sampling times are selected randomly within
    a space.}
  \label{fig:tut_eg_11C}
  \end{minipage}
  }
\end{figure}

However, this is not always the case, and Space and Time are good
examples of this.  Figure~\ref{fig:tut_eg_11C} shows three possible
contextual structures for Time, Place, and Object.
Figure~\ref{fig:tut_eg_11C}A represents the contextual structure when
Time and Place provide independent context for Object.  This means
that a given instance of Place can be observed within different
instances of Time (and visa versa), however a given instance of Object
cannot be observed within different instances of Time nor Place.  The
dependency matrix for Figure~\ref{fig:tut_eg_11C}A illustrates that a
different instance of Time does not necessarily imply a different
instance of Place (denoted ``N" for No).  However, a different Time
does imply a different Object (denoted ``Y" for Yes).  Furthermore, a
different instance of Place does not imply a different instance of
Time, but it does imply a different Object.  Finally, a different
instance of Object does not necessarily imply a different Time nor
Place, because in this example it is possible to observed more than
one object at a specific Time and Place combination.  Following the
grey arrows determines the correct \obs{} context representation in
\obs{} (to the right of the matrix).  However, Time and Place are not
always independent.  Figure~\ref{fig:tut_eg_11C}B shows an example for
when Places are different (e.g., randomly chosen) at each for the
times observations were made; i.e., Place is dependent on Time.
Figure~\ref{fig:tut_eg_11C}C shows an example where Time is now
dependent on Place, which might happen when time-series are conducted
at different places.

\begin{table}[htbp]
  \centering
 \caption{A data set containing heights for multiple oak species within plots at Summer Hill on two occassions.}
 \begin{tabular}{ @{} ccccc @{} }
   \toprule
   Date & Location & Plot & Species & Height (m) \\
   \midrule
   12/04/2007 & Summer Hill & A & \textit{Q. lobata} & 2.5 \\
   12/04/2007 & Summer Hill & A & \textit{Q. lobata} & 3.1 \\
   12/04/2007 & Summer Hill & B & \textit{Q. berberidifolia} & 4.1 \\
   12/04/2007 & Summer Hill & B & \textit{Q. lobata} & 8.5 \\
   12/04/2007 & Summer Hill & C & \textit{Q. durata} & 2.2 \\
   12/06/2008 & Summer Hill & A & \textit{Q. lobata} & 7.8 \\
   12/06/2008 & Summer Hill & A & \textit{Q. lobata} & 7.7 \\
   12/06/2008 & Summer Hill & A & \textit{Q. lobata} & 2.1 \\
   12/06/2008 & Summer Hill & A & \textit{Q. lobata} & 2.6 \\
   12/06/2008 & Summer Hill & A & \textit{Q. lobata} & 3.2 \\
   12/06/2008 & Summer Hill & B & \textit{Q. berberidifolia} & 9.1 \\
   12/06/2008 & Summer Hill & B & \textit{Q. lobata} & 9.8 \\
   12/06/2008 & Summer Hill & B & \textit{Q. berberidifolia} & 2.2 \\
   12/06/2008 & Summer Hill & C & \textit{Q. durata} & 7.5 \\
   \bottomrule
 \end{tabular}
\label{tab:context}
\end{table}

Table~\ref{tab:context} contains another hypothetical data set that
was collected to monitor the growth of oak trees at a specified
location through time.  Data collection was accomplished by measuring
the names and heights of individual trees within 10 x 10 meter square
plots at two different times.  In this table, each row (or tuple) is
comprised from four Observations: one each of Temporal Point, Spatial
Location, Replicate Plot (a second spacial treatment), and Tree
(selected from the \obs{} Entity extension shown in
Figure~\ref{fig:tut_ont_1}).  Figure~\ref{fig:tut_eg_11D} illustrates
two possible ways in which the contextual structure might be
represented depending on the research protocol.  The key difference
between the two representations is whether or not the plots are (A)
permanent (i.e., the same instance of Replicate Plot can exist at
different instances of Temporal Point) or (B) randomly placed during
each census (i.e., an instance of Plot is dependent on a specific
instance of Temporal Point).

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.85\textwidth]{./figures/tut_eg_11D.png}
  \caption{Venn diagrams illustrating two possible contextual
    structures for the Entities observed in Table~\ref{tab:context}.
    Panel \textbf{A} shows the contextual structure for when Plots are
    randomly placed at each consecutive time step; so an instance of
    Plot is dependent of an instance of Time.  Alternatively, panel
    \textbf{B} shows the structure for when Plots are permanently
    placed and thus ``endure" through time; so an instance of Plot is
    not wholly dependent on an instance of Time.}
  \label{fig:tut_eg_11D}
  \end{minipage}
  }
\end{figure}

\pagebreak

\begin{figure}[hbtp]
  \fbox{
    \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=1\textwidth]{./figures/tut_ont_1.png}
  \caption{Ontology extensions of core \obs{} classes used in this tutorial.}
  \label{fig:tut_ont_1}
  \end{minipage}
  }
\end{figure}

%%%
\pagebreak
\section{Semantic Annotation}
%%%

Semantic annotation is the mapping between data and \obs{}.  Outlined
below are the steps necessary to annotate data sets using
Table~\ref{tab:context} as the working example.  Computer software is
currently being develop to assist the annotation process.

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.75\textwidth]{./figures/tut_eg_10_att.png}
  \caption{A...}
  \label{fig: tut_eg_10_att}
  \end{minipage}
  }
\end{figure}

\begin{enumerate}

\item Identify the data attributes (in this case, columns) to be
  capture by the semantic annotation (top row in Figure~\ref{fig:
    tut_eg_10_att}).  Meta data that applies to whole-attributes
  should also be identified.  In this example, all replicate plots
  were 10 meters square, and so there was no need to include a column
  expressing this in the raw data set (and thus it was relegated to
  meta data).  However, such information is critical for determining
  important measures like areal density, especially when comparing and
  potentially merging data sets.

\item The appropriate Entities for each data attribute are then
  selected from a specified extension ontology (second row in
  Figure~\ref{fig: tut_eg_10_att}; selected from the Entity extension
  in Figure~\ref{fig:tut_ont_1}), remembering that sometimes more than
  one attribute pertains to the same Entity.  In this example, Species
  and Height both pertain to a instance of Tree.

\item Choose the Characteristics from the extension ontology that best
  describes that the attribute measured, as well as the Measurement
  Standard used to represent the Characteristic (third and fourth rows
  of Figure~\ref{fig: tut_eg_10_att}; both selected from the
  respective extensions in Figure~\ref{fig:tut_ont_1}).

\item Value points to the set of data for which the annotation
  applies.  In this case, the annotation is applied to each data point
  in the column vectors (bottom row of Figure~\ref{fig:
    tut_eg_10_att}).  Precision only applies to continuous variables,
  and is by default the number of significant decimal places (fifth
  row of Figure~\ref{fig: tut_eg_10_att}).

\item Determine the contextual structure.  In this data set, the plots
  are permanent and therefore the representation in
  Figure~\ref{fig:tut_eg_11D}B is correct.

\item Complete the \obs{} annotation graphically using the information
  in Figure~\ref{fig: tut_eg_10_att} and the context structure in
  Figure~\ref{fig:tut_eg_11D}B to give Figure~\ref{fig:
    tut_eg_10_ann}).

\item Represent the annotation using the Semantic Annotation Language
  (SAL) that can then be saved in the data sets metadata.  The SAL is
  described elsewhere and will be done automatically when the
  annotation software is ready.

\end{enumerate}

\begin{figure}[hbtp]
  \fbox{
  \begin{minipage} {16 cm}
  \centering
  \includegraphics[width=0.55\textwidth]{./figures/tut_eg_10_ann.png}
  \caption{The \obs{} representation of the first data point in Table~\ref{tab:context}.}
  \label{fig: tut_eg_10_ann}
  \end{minipage}
  }
\end{figure}

\newpage
\textbf{ANNOTATION 1---}
\begin{lstlisting}[frame=single]
<annotation emlPackage="..." dataTable="..." xmlns:oboe="..." xmlns:ext="...">

  <observation label="o1" attributes="Date">
    <entity class="ext:TemporalPoint"/>
    <measurement>
      <characteristic class="oboe:Date"/>
      <standard class="ext:DateTime"/>
      <value attribute="Date"/>
    </measurement>
  </observation>

  <observation label="o2" attributes="Location">
    <entity class="ext:SpatialLocation"/>
    <measurement>
      <characteristic class="oboe:EntityName"/>
      <standard class="ext:SiteNameCatalog"/>
      <value attribute="Location"/>
    </measurement>
  </observation>

  <observation label="o3" attributes="Date Location ReplicatePlot">
    <entity class="ext:ReplicatePlot"/>
    <measurement>
      <characteristic class="oboe:EntityName"/>
      <standard class="oboe:NominalStandard"/>
      <value attribute="Plot"/>
    </measurement>
    <measurement precision="1">
      <characteristic class="ext:Area"/>
      <standard class="ext:MeterSquare"/>
      <value constant="10"/>
    </measurement>
    <context observation="o1" property="oboe:hasContainmentContext"/>
    <context observation="o2" property="oboe:hasContainmentContext"/>
  </observation>

  <observation label="o4" attributes="*">
    <entity class="ext:Tree"/>
    <measurement>
      <characteristic class="oboe:EntityName"/>
      <standard class="ext:TaxonomicStandard"/>
      <value attribute="Species"/>
    </measurement>
    <measurement precision="0.1">
      <characteristic class="ext:Height"/>
      <standard class="ext:Meter"/>
      <value attribute="Height"/>
    </measurement>
    <context observation="o3" property="oboe:hasContainmentContext"/>
  </observation>

</annotation>
\end{lstlisting}

\clearpage

%%%
\chapter{Examples of Semantic Annotation}

In this section, a number of real data sets are annotated using \obs{} starting with basic examples and getting progressively harder.

\section{Height Measurement}
\section{Tree Measurements I}
\section{Tree Measurements II}
\section{Leaf Traits}
\section{GCE Fertilisation Plots}
\section{Specimen Record}
\section{Summary of Organism Heights from Example 3}
\section{Predator-Prey Study}
\section{Parasitoid--Parasite--Host Study}

\end{document}

### 2008-09-08 docs/presentations/research-coordination-mtg-abq-2008/sonet-handout.tex

\documentclass[10pt,twocolumn]{article}

\usepackage{fullpage}
\usepackage{graphicx}

\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\mypara}{\vspace{6pt}\noindent}
\newcommand{\sonet}{{SONet}}

\begin{document}

%\title{SONet (Scientific Observations Network): Facilitating Data
%  Interoperability within the Environmental and Ecological Sciences}

\title{\textbf{\sonet: A Scientific Observations Network for
    Environmental and Ecological Data Interoperability}}

\author{Mark Schildhauer$^1$, Shawn Bowers$^2$, Josh Madin$^4$,
  Corinna Gries$^3$, Deborah McGuinness$^5$, \\
  Matt Jones$^1$, Phillip Dibner$^6$, Luis
  Bermudez$^7$, John Graybeal$^7$ \\[12pt]
  $^1$NCEAS, UC Santa Barbara, $^2$UC Davis Genome Center,
  $^3$CAP/LTER and \\ Arizona State University, $^4$Macquarie
  University, Australia, $^5$McGuinness \\ Associates, $^6$Open
  Geospatial Consortium Interoperability Institute, $^7$Monterey \\
  Bay Aquarium Research Institute}

\date{September, 8, 2008}

\maketitle

\thispagestyle{empty}

\mypara Advances in ecological, biodiversity, and other environmental
and life sciences increasingly depend on the use of information from
multiple disciplines to tackle broader and more complex questions
about the natural world.  Such advances, however, are hindered by data
heterogeneity, which impedes the ability of researchers to discover,
interpret, and integrate relevant data that have been collected by
others.

\mypara The \emph{\textbf{Scientific Observations Network}}, or
  \emph{\textbf{\sonet}}, is a recently funded NSF project intended to
facilitate the development of a community-sanctioned, unified data
model for representing observational data.  \sonet\ brings together
researchers in the ecological, biodiversity, and environmental
sciences, working in close conjunction with computer scientists and
information managers, to define and develop specifications and
open-source technologies to facilitate the interpretation and
integration of scientific data.

\begin{figure}[!b]
  \begin{footnotesize}
  \centering
  \includegraphics[scale=.4]{sonet-subgroups}
  \vspace{-8pt}
  \caption{{\small Primary \sonet\ working groups}}
  \label{fig:groups}
  \end{footnotesize}
\end{figure}

\mypara A key aspect of \sonet's approach has grown out of earlier
workshops, in which domain and information-management experts agreed
that the ``scientific observation'' provides a promising conceptual
linchpin for improved software interoperability and data integration
across disciplines. In the sense used here, a scientific observation
consists of the measurement of a characteristic of some entity, at a
given place and time.  The specifics of the entities and the
characteristics measured can be captured through emerging approaches
in semantic technologies, most promisingly via formal ontologies.

\begin{figure}[!h]
  \begin{footnotesize}
  \centering
  \includegraphics[scale=.35]{sonet-timeline}
  \vspace{-8pt}
  \caption{{\small Overview of \sonet\ workshops and activies}}
  \label{fig:timeline}
  \end{footnotesize}
\end{figure}

\mypara The \sonet\ effort provides resources to build a
cross-disciplinary community to refine and promote a unified approach
to modeling scientific data, through working groups
(\figref{fig:groups}) and a series of workshops
(\figref{fig:timeline}) and to begin the process of building
interoperable ontologies that capture the semantic nuances needed for
effectively communicating the meaning and interpretation of data to
specialists as well as scientists pursuing synthetic analyses.
Working groups together with network experts will also develop a
series of demonstration projects to illustrate the capabilities for
data interoperability.

\end{document}

### 2008-11-05 docs/pubs/ISEI6/main.tex

\documentclass[preprint,number]{elsarticle} 
% other options: preprint, review, 1p, 3p, 5p, 
% bib options: authoryear, number, longtitle,
% other options: times

\title{Owlifier: Creating OWL-DL Ontologies from Simple
  Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}}
\tnotetext[t1]{This work supported in part by NSF grants ...}

\author[smb]{Shawn Bowers\corref{cor1}}
\ead{sbowers@ucdavis.edu}

\author[jsm]{Joshua S. Madin}
\ead{jmadin@bio.mq.edu.au}

\author[mps]{Mark P. Schildhauer}
\ead[url]{schild@nceas.ucsb.edu}

\cortext[cor1]{Corresponding author}

\address[smb]{UC Davis Genome Center}

\address[jsm]{Dept. of Biological Sciences, Macquarie University, Australia}

\address[mps]{National Center for Ecological Analysis and Synthesis,
  UC Santa Barbara}

\begin{document}

\begin{abstract}
Abstract goes here ... 
\end{abstract}

\maketitle

\section{Introduction}

\bibliographystyle{elsarticle-num}
\bibliography{main}

\end{document}

### 2008-11-05 docs/pubs/ISEI6/owlifier-2008.tex

\documentclass[11pt,onecolumn]{article}

\usepackage{fullpage}

\newcommand{\owlifier}{\textsf{owlifier}}

\title{\textsf{Owlifier}: An Application for Creating Simple OWL
  Ontologies from Spreadsheet-Based Knowledge Descriptions}

\author{Shawn Bowers \and Mark Schildhauer \and Joshua Madin \and
  Mathew Jones}

\date{}

\begin{document}

\maketitle

\begin{abstract}
\end{abstract}

\section{Introduction}

\section{\textsf{Owlifier} Syntax}

An \owlifier\ table defines an OWL-DL \cite{owl} ontology through a
set of \emph{blocks}.  Each non-empty row in an \owlifier\ table
corresponds to a block. The type of the block is given in the first
column of the row. The types of blocks supported by \owlifier\ are as
follows.  It is assumed below that if any of the properties or
concepts used in a block are defined, i.e., are not imported from
another ontology, then they are added to the current ontology.

\paragraph{Import Block.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, concepts and
properties of imported ontologies can be used within other blocks of
the table.  An import block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \texttt{import} & $\textit{NS}$ & $\textit{URI}$ \\ \hline
  \end{tabular}
\end{itemize}
where $\textit{NS}$ is a namespace label and $\textit{URI}$ is an OWL
ontology URI. Concepts and properties from imported ontologies are
referenced by prefixing the namespace label $\textit{NS}$ to the
corresponding concept or property name in the normal way.

\paragraph{Concept Block.} Concept blocks specify concept and
subconcept relationships. A concept block has the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{concept} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each concept $C_i$ is asserted in the current ontology to
subsume a concept $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a
concept block induces a DL axiom \[C_i \sqsubseteq C_{i+1}\] If both
$C_i$ and $C_{i+1}$ are imported concepts, we say that the block
defines an ``articulation'' (i.e., mapping) between them.

\paragraph{Synonym Block.} Synonym blocks define an equivalence
relationship between concepts.  A synonym block has the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{synonym} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each concept $C_i$ is asserted as being equivalent to concept
$C_{i+1}$ in the current ontology, for $1 \le i < n$.

\paragraph{Overlap Block.} Except in certain situations (described
further below), defined concepts are assumed to be disjoint.  Overlap
blocks explicitly relax this assumption for a given set of
concepts. An overlap block has the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{overlap} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each concept $C_i$ is allowed to share instances with each
concept $C_j$, for $1 \le i,j \le n$. In particular, $C_i$ and $C_j$
are not defined to be disjoint concepts in the current ontology.

\paragraph{Property Block.} Property blocks define the required
\emph{object} properties of concepts. A property block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{property} & $P$ & $C_1$
& $C_2$ & \dots & $C_n$ \\ \hline
\end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $P$ is an object property and each $C_i$ is a concept, for $1
\le i \le n$.  For every concept $C_i$, the property block induces the
DL axiom \[C_i \sqsubseteq \exists P . C_{i+1} \] stating that each
instance of $C_i$ has, amongst possibly other things, a relationship
through $P$ to some instance of $C_{i+1}$.  For example, the block
\begin{quote}\texttt{property} \texttt{hasPart} \texttt{Body}
  \texttt{Head} \texttt{Eye} \texttt{Retina}
\end{quote} states that a body has at least one head, a head has at
least one eye, and an eye has at least one retina.

\paragraph{Attribute Block.} Attribute blocks are used to define the
required \emph{datatype} properties of concepts. An attribute block
has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $P$ & $D$ & $C_1$
    & $C_2$ & \dots & $C_n$ \\ \hline
\end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where $P$ is a datatype property, each $C_i$ is a concept for $1 \le i
\le n$, and $D$ is a datatype (\texttt{anyValueType}, \texttt{string},
\texttt{int}, etc.). For every concept $C_i$, the property block
induces the DL axiom \[C_i \sqsubseteq (\exists P . D) \sqcap (\le 1 ~
P)\] stating that each instance of $C_i$ has, amongst possibly other
things, a relationship through $P$ to exactly one data value of type
$D$.

\paragraph{Value Block.} Value blocks define required datatype
property \emph{constant values} for concepts. A value block has the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{value} & $P$ & $V$ & $C_1$
    & $C_2$ & \dots & $C_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
n$, and $V$ is a datatype value. For each concept $C_i$, the value
block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
each instance of $C_i$ has a value $V$ for property $P$.  The value
restrictions stated by value blocks are often used for defining
so-called \emph{value partitions} \cite{co-ode}.

\paragraph{Inverse Block.} Inverse blocks state that two object
properties are inverses of each other. That is, for inverse properties
$P_1$ and $P_2$ and concept instances $O_1$ and $O_2$, if $P_1(O_1) =
O_2$, then $P_2(O_2) = O_1$.  An inverse block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \texttt{inverse} & $P_1$ & $P_2$
\\ \hline
  \end{tabular}
\end{itemize}
where $P_1$ and $P_2$ are object properties.

\paragraph{Transitive Block.} Transitive blocks state that a property
is transitive. That is, if $P$ is transitive and a concept instance
$O_1$ is related to an instance $O_2$ by $P$, and $O_2$ is related to
an instance $O_3$ by $P$, then $O_1$ is also by definition related to
$O_3$ by $P$. A transitive block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|}\hline \texttt{transitive} & $P$ 
\\ \hline
  \end{tabular}
\end{itemize}
where $P$ is an object property.

\paragraph{Cardinality Block.} Cardinality blocks state the maximum
number of properties $P$ an instance of a concept may have.
Cardinality blocks have the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|}\hline \texttt{cardinality} & $N$ & $P$ & 
    $C_1$ & $C_2$ 
    \\ \hline
  \end{tabular}
\end{itemize}
where $N$ is the maximum number of properties $P$ that instances of
concept $C_1$ may have to instances of concept $C_2$. A cardinality
block induces the DL axiom \[C_1 \sqsubseteq (\le N P.C_2)\] stating
that each instance of $C_1$ can be related to at most $N$ unique
instances of $C_2$ via $P$. For example, the blocks
\begin{quote}
\texttt{cardinality} 1 \texttt{hasPart} \texttt{Body} \texttt{Head} \\
\texttt{cardinality} 2 \texttt{hasPart} \texttt{Head} \texttt{Eye}
\end{quote}
states that a body has at most one head and a head has at most two
eyes.

\paragraph{Sufficient Block.} Sufficient blocks state that any
instance having a property $P$ to an instance of a concept $C_2$ is a
sufficient condition for being an instance of a concept $C_1$. A
sufficient block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|}\hline \texttt{sufficient} & $P$ & $C_1$ & $C_2$
    \\ \hline
  \end{tabular}
\end{itemize}
where $C_1$ is the target concept (i.e., denoting the concept
definition), $P$ is the sufficient property, and $C_2$ is the
sufficient concept. A sufficient block induces the DL axiom \[C_1
\equiv \exists P.C_2\] Sufficient blocks provide a mechansism to
construct simple class definitions (i.e., classes defined precisely by
other classes), primarily for use with value partitions.

\paragraph{Description Block.} Description blocks assign plain-text
definitions to concepts and properties. A description block has the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \texttt{description} & $T$ & $S$
\\ \hline
  \end{tabular}
\end{itemize}
where $T$ is either a property or a concept and $S$ is a description
string.

\paragraph{Note Block.} Note blocks add comments to the current
ontology, and are ignored by \owlifier.  A note block has the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \texttt{note} & $S$
\\ \hline
  \end{tabular}
\end{itemize}
where $S$ is a comment string.

*** Say something about relaxing block syntax ... to make it easier to
specify ontologies. Also, allow blocks to be given in any order.

\section{\textsf{Owlifier} Reasoning}

e.g., Disjoint Concept Inference. Need to describe here when we say
two concepts are disjoint. Other inferences are now possible as well.

Errors:
\begin{itemize}
\item Blocks with syntactic errors
\item Inverse properties can be between at most two properties. For
  instance, inverse(P1,P2) and inverse(P1,P3) is not allowed.
\item At most one description is allowed per property or concept.
\item Property and concept names must be disjoint
\end{itemize}

Warnings: 
\begin{itemize}
\item Cyclic concept hierarchies
\item Re-definition of imported concepts (have to define what this means)
\item Introduction of an inconsistency (can we show that this will
  never happen in a fully defined ontology, i.e., one without imports)? 
\item ...
\end{itemize}

\section{\textsf{Owlifier} Examples}

\begin{itemize}
\item Simple example, no imports, no warnings
\item Extension example
\item Articulation example
\end{itemize}

\subsection{\textsf{Owlifier} for OBOE}

How it works with OBOE.

\section{\textsf{Owlifier} API}

Flags:
\begin{itemize}
\item Turn on/off consistency checking/validation
\item Output format (OWL/RDF, 
\item Output inferred axioms
\item Ontology URI to use
\end{itemize}

\section{Discussion}

Implementation, etc.

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2008-11-21 docs/pubs/ISEI6/owlifier-2008.tex

@@
 \item[]
-  \begin{tabular}{|l|l|}\hline \texttt{transitive} & $P$ 
+  \begin{tabular}{|l|l|l|l|}\hline \texttt{transitive} & $P_1$ &
+    $\dots$ & $P_n$
 \\ \hline
@@
 
-\paragraph{Cardinality Block.} Cardinality blocks state the maximum
+[others?, e.g., associatiave
+
+\paragraph{Minimum Block.} Minimum blocks state the minimum
 number of properties $P$ an instance of a concept may have.
-Cardinality blocks have the form
+Minimum blocks have the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|}\hline \texttt{cardinality} & $N$ & $P$ & 
-    $C_1$ & $C_2$ 
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{minimum} & $P$ & $N$ & 
+    $C_1$ & $C_2$ & $\dots$ & $C_m$
+    \\ \hline
+  \end{tabular}
+\end{itemize}
+where $N$ is the minimum number of properties $P$ that instances of
+concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
+and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
+(\le N P.C_{i+1})\] stating that each instance of $C_i$ must be related to
+at least $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
+\begin{quote}
+\texttt{minimum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
+\texttt{minimum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
+\end{quote}
+states that a body has at least one head and at least two eyes.
+
+\paragraph{Maximum Block.} Maximum blocks state the maximum
+number of properties $P$ an instance of a concept may have.
+Maximum blocks have the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{maximum} & $P$ & $N$ & 
+    $C_1$ & $C_2$ & $\dots$ & $C_m$
     \\ \hline
@@
 where $N$ is the maximum number of properties $P$ that instances of
-concept $C_1$ may have to instances of concept $C_2$. A cardinality
-block induces the DL axiom \[C_1 \sqsubseteq (\le N P.C_2)\] stating
-that each instance of $C_1$ can be related to at most $N$ unique
-instances of $C_2$ via $P$. For example, the blocks
+concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
+and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
+(\ge N P.C_{i+1})\] stating that each instance of $C_i$ may be related to
+at most $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
 \begin{quote}
-\texttt{cardinality} 1 \texttt{hasPart} \texttt{Body} \texttt{Head} \\
-\texttt{cardinality} 2 \texttt{hasPart} \texttt{Head} \texttt{Eye}
+\texttt{maximum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
+\texttt{maximum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
 \end{quote}
-states that a body has at most one head and a head has at most two
-eyes.
+states that a body has at least one head and at least two eyes.
 
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|}\hline \texttt{sufficient} & $P$ & $C_1$ & $C_2$
+  \begin{tabular}{|l|l|l|l|}\hline \texttt{sufficient} & $C_1$ & $P$ & $C_2$
     \\ \hline
@@
 construct simple class definitions (i.e., classes defined precisely by
-other classes), primarily for use with value partitions.
+other classes), primarily for use with value partitions. [NOTE: these
+should be anded together?]
 

### 2008-11-25 docs/pubs/ISEI6/main.tex

@@
 
+\newcommand{\owlifier}{Owlifier}
+\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}
+
 \title{Owlifier: Creating OWL-DL Ontologies from Simple
   Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}}
-\tnotetext[t1]{This work supported in part by NSF grants ...}
+%\tnotetext[t1]{This work supported in part by NSF grants ...}
 
@@
 \begin{abstract}
-Abstract goes here ... 
+abstract here...
 \end{abstract}
@@
 
-\bibliographystyle{elsarticle-num}
+The use of formal ontologies can significantly enhance metadata
+descriptions of ecological data by providing richer and more
+consistent terminologies. Annotating data with ontology terms can both
+help users interpret data as well as enable advanced capabilities for
+data discovery and integration, e.g., by exploiting subsumption and
+part-of hierarchies as well as more formal constraints such as
+cardinality restrictions on properties and term equivalence. Many
+efforts are underway to engage scientists in the development of
+ontologies to represent common ecological and biodiversity
+concepts. Most of these projects leverage OWL (the W3C Web Ontology
+Language) as a standard XML syntax for representing and sharing
+ontologies. A key advantage of OWL is that it is supported by a wide
+range of generic tools, including editors, reasoning systems, query
+languages, and storage technologies. These tools, however, are
+primarily targeted at experts in knowledge engineering and software
+development. This is especially true with ontology editors (such as
+Prot?g?, OWLed, SWOOP, etc.), which allow for very detailed ontology
+specifications, but at the same time require a considerable amount of
+understanding of the underlying ontology formalisms and syntax. Thus,
+we see the lack of suitable ontology editing tools for scientists as
+one of the major barriers for more wide-scale adoption of ontologies
+in ecology.
+
+Here, we present a novel approach for ontology creation that aims at
+being more intuitive for ecologists and that can be used to rapidly
+construct large ontologies for describing scientific data. Our
+approach is to allow scientists to use common spreadsheet tools to
+describe, in an intuitive way, different aspects of an ontology, and
+then to take these descriptions and convert them into full-fledged OWL
+ontologies using a software application called owlifier. An owlifier
+spreadsheet consists of a set of "blocks" that have a predefined
+template structure for users to fill in. Each non-empty row in an
+owlifier table constitutes a block, where blocks can be used to define
+ontology concepts (including sub-concepts), concept synonyms,
+overlapping concepts (i.e., that by definition are not disjoint),
+properties (including their domain and range), and inverse
+properties. We also provide blocks for plain-text descriptions of
+concepts and properties, and for referencing one or more existing
+ontologies (e.g., to extend an ontology or to define ontology
+articulations). Blocks can be sparse (inheriting from previous
+blocks), which simplifies the creation of large ontologies. Although
+not as expressive as full-fledged OWL ontologies, our approach can
+produce ontology structures essential for enhanced data discovery and
+integration, while at the same time providing a more accessible user
+interface for ecologists. Further, compared to many existing editors,
+our approach can be used to rapidly construct concept hierarchies,
+e.g., from long lists of keywords, within familiar spreadsheet
+software. We have field-tested the owlifier approach with ecologists
+and evolutionary biologists working with trait data, and found this
+application enabled them to quickly and easily comprehend and
+construct ontologies.
+
+\section{\owlifier\ Syntax and Semantics}
+
+An \owlifier\ table defines an OWL-DL
+\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
+\emph{blocks} representing one or more ontology definitions.  Each
+non-empty row in an \owlifier\ table corresponds to a block. The type
+of the block is given in the first column of the row. We describe each
+type of block supported by \owlifier\ below. Here we assume that if
+any of the properties or concepts used in a block are defined, i.e.,
+are not imported from another ontology, then they are added to the
+ontology being defined by the \owlifier\ table.
+
+\mypara{Import Block}
+
+\mypara{}Import blocks assign namespace labels to external
+ontologies. Each external ontology is imported into the current
+ontology. We refer to the ontologies of import blocks as
+\emph{imported ontologies}.  Using import blocks, concepts and
+properties of imported ontologies can be used within other blocks of
+the table.  Import blocks take the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|}\hline
+    \texttt{import} & $\textit{N}$ & $\textit{U}$ \\ \hline 
+  \end{tabular} 
+\end{itemize}
+where $\textit{N}$ is a namespace label and $\textit{U}$ is an OWL
+ontology URI. Classes and properties from imported ontologies are
+referenced by prefixing the namespace label $\textit{N}$ to the
+corresponding class or property name in the normal way. The following
+block imports the ``earth realm'' ontology from the set of SWEET
+ontologies \cite{raskin:_seman_web_for_earth_and}.
+\begin{tabbing}
+  ~~\texttt{import sweet
+    http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
+\end{tabbing}
+where the ontology class denoting marine ecosystems can be refered to
+from within the corresponding \owlifier\ table using the syntax
+\texttt{sweet:MarineEcosystem}. This class is referred to an an
+``imported'' class.
+
+\mypara{Class Block}
+
+\mypara{}Class blocks specify class and subclass relationships. Class
+blocks take the form
+\begin{itemize}
+\item[] 
+  \begin{tabular}{|l|l|l|l|l|}\hline
+    \texttt{class} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+  \end{tabular} \hfill ($n \ge 1$)
+\end{itemize}
+where each class $C_i$ is asserted in the current ontology to subsume
+a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in an entity block
+induces a DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
+$C_{i+1}$ are imported classes, we say that the block defines an
+``articulation'' (i.e., mapping) between them. The following class
+block defines a simple subclass hierarchy
+\begin{tabbing}
+~~\texttt{class PhysicalFeature AquaticPhysicalFeature River}
+\end{tabbing}
+stating that a river is a subclass of aquatic physical features and
+that aquatic physical features are a type of physical feature.
+
+\mypara{Synonym Block}
+
+\mypara{}Synonym blocks define an equivalence relationship between
+classes.  A synonym block has the form
+\begin{itemize}
+\item[] 
+  \begin{tabular}{|l|l|l|l|l|}\hline
+    \texttt{synonym} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+  \end{tabular} \hfill ($n \ge 2$)
+\end{itemize}
+where each class $C_i$ is asserted as being equivalent to class
+$C_{i+1}$ in the current ontology, for $1 \le i < n$. For example, the
+block
+\begin{tabbing}
+~~\texttt{synonym Maize Corn}
+\end{tabbing}
+defines maize and corn as synonym (i.e., equivalent) classes.
+
+\mypara{Overlap Block}
+
+\mypara{}Except in certain situations (described further below),
+classes are assumed to be disjoint.  Overlap blocks explicitly relax
+this assumption for a given set of classes by stating that a set of
+classes may have overlapping instances. An overlap block has the form
+\begin{itemize}
+\item[] 
+  \begin{tabular}{|l|l|l|l|l|}\hline
+    \texttt{overlap} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+  \end{tabular} \hfill ($n \ge 2$)
+\end{itemize}
+where each class $C_i$ is allowed to share instances with each concept
+$C_j$, for $1 \le i,j \le n$. In particular, $C_i$ and $C_j$ are not
+defined to be disjoint classes in the current ontology.
+
+--- an example here ---
+
+\mypara{Relationship Block}
+
+\mypara{}Relationship blocks define required \emph{object} properties
+of classes. A property block has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $R$ & $C_1$
+    & $C_2$ & \dots & $C_n$ \\ \hline
+  \end{tabular} \hfill ($n \ge 2$)
+\end{itemize}
+where $R$ is an object property and each $C_i$ is a class, for $1 \le
+i \le n$.  For every class $C_i$, the property block induces the DL
+axiom \[C_i \sqsubseteq \exists R . C_{i+1} \] stating that each
+instance of $C_i$ has, amongst possibly other things, a relationship
+through $R$ to some instance of $C_{i+1}$.  The following example
+\begin{tabbing}
+   ~~need example here
+\end{tabbing}
+
+\mypara{Transitive Block}
+
+\mypara{}Transitive blocks state that a property is transitive. That
+is, if $P$ is transitive and a concept instance $O_1$ is related to an
+instance $O_2$ by $P$, and $O_2$ is related to an instance $O_3$ by
+$P$, then $O_1$ is also by definition related to $O_3$ by $P$. A
+transitive block has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $R$ & $C_1$
+    & $C_2$ & \dots & $C_n$ \\ \hline
+  \end{tabular} \hfill ($n \ge 2$)
+\end{itemize}
+where $P$ is an object property. For example, the block
+\begin{quote}\texttt{transitive} \texttt{hasPart} \texttt{Body}
+  \texttt{Head} \texttt{Eye} \texttt{Retina}
+\end{quote} states that a body has at least one head, a head has at
+least one eye, and an eye has at least one retina.
+
+\paragraph{Attribute Block.} Attribute blocks are used to define the
+required \emph{datatype} properties of concepts. An attribute block
+has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $P$ & $D$ & $C_1$
+    & $C_2$ & \dots & $C_n$ \\ \hline
+  \end{tabular} \hfill ($n \ge 1$)
+\end{itemize}
+where $P$ is a datatype property, each $C_i$ is a concept for $1 \le i
+\le n$, and $D$ is a datatype (\texttt{anyValueType}, \texttt{string},
+\texttt{int}, etc.). For every concept $C_i$, the property block
+induces the DL axiom \[C_i \sqsubseteq (\exists P . D)\] stating that
+each instance of $C_i$ has, amongst possibly other things, a
+relationship through $P$ to a data value of type $D$.
+
+\paragraph{Value Block.} Value blocks define required datatype
+property \emph{constant values} for concepts. A value block has the
+form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{value} & $P$ & $V$ & $C_1$
+    & $C_2$ & \dots & $C_n$ \\ \hline
+  \end{tabular} \hfill ($n \ge 1$)
+\end{itemize}
+where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
+n$, and $V$ is a datatype value. For each concept $C_i$, the value
+block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
+each instance of $C_i$ has a value $V$ for property $P$.  The value
+restrictions stated by value blocks are often used for defining
+so-called \emph{value partitions} \cite{co-ode}.
+
+\paragraph{Inverse Block.} Inverse blocks state that two object
+properties are inverses of each other. That is, for inverse properties
+$P_1$ and $P_2$ and concept instances $O_1$ and $O_2$, if $P_1(O_1) =
+O_2$, then $P_2(O_2) = O_1$.  An inverse block has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|}\hline \texttt{inverse} & $P_1$ & $P_2$
+\\ \hline
+  \end{tabular}
+\end{itemize}
+where $P_1$ and $P_2$ are object properties.
+
+\paragraph{Minimum Block.} Minimum blocks state the minimum
+number of properties $P$ an instance of a concept may have.
+Minimum blocks have the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{minimum} & $P$ & $N$ & 
+    $C_1$ & $C_2$ & $\dots$ & $C_m$
+    \\ \hline
+  \end{tabular}
+\end{itemize}
+where $N$ is the minimum number of properties $P$ that instances of
+concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
+and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
+(\le N P.C_{i+1})\] stating that each instance of $C_i$ must be related to
+at least $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
+\begin{quote}
+\texttt{minimum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
+\texttt{minimum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
+\end{quote}
+states that a body has at least one head and at least two eyes.
+
+\paragraph{Maximum Block.} Maximum blocks state the maximum
+number of properties $P$ an instance of a concept may have.
+Maximum blocks have the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{maximum} & $P$ & $N$ & 
+    $C_1$ & $C_2$ & $\dots$ & $C_m$
+    \\ \hline
+  \end{tabular}
+\end{itemize}
+where $N$ is the maximum number of properties $P$ that instances of
+concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
+and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
+(\ge N P.C_{i+1})\] stating that each instance of $C_i$ may be related to
+at most $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
+\begin{quote}
+\texttt{maximum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
+\texttt{maximum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
+\end{quote}
+states that a body has at least one head and at least two eyes.
+
+\paragraph{Sufficient Block.} Sufficient blocks state that any
+instance having a property $P$ to an instance of a concept $C_2$ is a
+sufficient condition for being an instance of a concept $C_1$. A
+sufficient block has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|}\hline \texttt{sufficient} & $C_1$ & $P$ & $C_2$
+    \\ \hline
+  \end{tabular}
+\end{itemize}
+where $C_1$ is the target concept (i.e., denoting the concept
+definition), $P$ is the sufficient property, and $C_2$ is the
+sufficient concept. A sufficient block induces the DL axiom \[C_1
+\equiv \exists P.C_2\] Sufficient blocks provide a mechansism to
+construct simple class definitions (i.e., classes defined precisely by
+other classes), primarily for use with value partitions. [NOTE: these
+should be anded together?]
+
+\paragraph{Description Block.} Description blocks assign plain-text
+definitions to concepts and properties. A description block has the
+form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|}\hline \texttt{description} & $T$ & $S$
+\\ \hline
+  \end{tabular}
+\end{itemize}
+where $T$ is either a property or a concept and $S$ is a description
+string.
+
+\paragraph{Note Block.} Note blocks add comments to the current
+ontology, and are ignored by \owlifier.  A note block has the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|}\hline \texttt{note} & $S$
+\\ \hline
+  \end{tabular}
+\end{itemize}
+where $S$ is a comment string.
+
+*** Say something about relaxing block syntax ... to make it easier to
+specify ontologies. Also, allow blocks to be given in any order.
+
+%\bibliographystyle{elsarticle-num}
+\bibliographystyle{abbrv}
 \bibliography{main}

### 2008-11-25 docs/pubs/ISEI6/main.tex

@@
 construct ontologies.
+
+Notes: 
+\begin{itemize}
+\item we generally follow the guidelines in ontology creation of
+  Rector et al. in \cite{rector04:_owl_pizzas}.
+\item we also try to simplify certain aspects, e.g., disjoint classes
+  and open-world reasoning, which can be confusing for non-experts
+\item we try to use more meaningful terms, when applicable (e.g., instead of datatype property, attribute, etc.)
+\item our goal is to not support all constructs in OWL-DL, but to
+  provide the most commonly used ones for defining typical ontologies,
+  but still allow experts to refine the ontology in ontology editing
+  tools if necessary
+\end{itemize}
 
@@
 
+\section{\owlifier\ Properties and Reasoning}
+\label{sec:properties}
+
+Some desirable properties: 
+\begin{itemize}
+\item non-ambiguous (no block leads to ambiguous DL axioms)
+\item ``reasonable'' (not everything has to be explicitly stated)
+\item ???
+\end{itemize}
+
+\section{\owlifier\ Implementation}
+\label{sec:implementation}
+
+Flags: 
+\begin{itemize}
+\item delimeter characters 
+\item perform classification
+\item warnings
+\item owlifier to owl and owl to owlifier
+\item ??? 
+\end{itemize}
+
+\section{Conclusion}
+\label{sec:conclusion}
+
 %\bibliographystyle{elsarticle-num}

### 2008-11-26 docs/pubs/ISEI6/main.tex

@@
 
-\newcommand{\owlifier}{Owlifier}
+\newcommand{\owlifier}{\textsf{Owlifier}}
 \newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}
@@
 \begin{abstract}
-abstract here...
+  Formal ontlogies promise to provide many added advantages to
+  ecological data management ... Most ontology editing frameworks and
+  tools, however, are design for knowledge-representation experts and
+  those with considerable understanding of the complexities of
+  ontology languages (i.e., description logics). This paper proposes
+  an approach for ontology creation based on ...
 \end{abstract}
@@
 descriptions of ecological data by providing richer and more
-consistent terminologies. Annotating data with ontology terms can both
-help users interpret data as well as enable advanced capabilities for
-data discovery and integration, e.g., by exploiting subsumption and
-part-of hierarchies as well as more formal constraints such as
-cardinality restrictions on properties and term equivalence. Many
-efforts are underway to engage scientists in the development of
-ontologies to represent common ecological and biodiversity
-concepts. Most of these projects leverage OWL (the W3C Web Ontology
-Language) as a standard XML syntax for representing and sharing
-ontologies. A key advantage of OWL is that it is supported by a wide
-range of generic tools, including editors, reasoning systems, query
-languages, and storage technologies. These tools, however, are
-primarily targeted at experts in knowledge engineering and software
-development. This is especially true with ontology editors (such as
-Prot?g?, OWLed, SWOOP, etc.), which allow for very detailed ontology
-specifications, but at the same time require a considerable amount of
-understanding of the underlying ontology formalisms and syntax. Thus,
-we see the lack of suitable ontology editing tools for scientists as
-one of the major barriers for more wide-scale adoption of ontologies
-in ecology.
+consistent terminologies \cite{tree,arees,parr}. Annotating data with
+ontology terms can both help users interpret data as well as enable
+advanced capabilities for data discovery and integration, e.g., by
+exploiting subsumption and part-of hierarchies as well as more formal
+constraints such as cardinality restrictions on properties and term
+equivalence \cite{ecoinf,other?}. Many efforts are underway to engage
+scientists in the development of ontologies to represent common
+ecological and biodiversity concepts. Most of these projects leverage
+the W3C Web Ontology Language (OWL) as a standard XML syntax for
+representing and sharing ontologies. A key advantage of OWL is that it
+is supported by a wide range of generic tools, including editors
+\cite{protege,swoop}, reasoning systems \cite{pellet,fact++}, query
+languages \cite{???}, and storage technologies \cite{jena,???}. Most
+of these tools, however, are primarily targeted at experts in
+knowledge engineering and software development. This is especially
+true with ontology editors (such as Protege, OWLed, SWOOP, etc.),
+which allow for very detailed ontology specifications, but at the same
+time require a considerable amount of understanding of the underlying
+ontology formalisms and syntax. Thus, we see the lack of suitable
+ontology editing tools for scientists as one of the major barriers for
+more wide-scale adoption of ontologies in ecology.
 
@@
 construct large ontologies for describing scientific data. Our
-approach is to allow scientists to use common spreadsheet tools to
-describe, in an intuitive way, different aspects of an ontology, and
-then to take these descriptions and convert them into full-fledged OWL
-ontologies using a software application called owlifier. An owlifier
-spreadsheet consists of a set of "blocks" that have a predefined
-template structure for users to fill in. Each non-empty row in an
-owlifier table constitutes a block, where blocks can be used to define
-ontology concepts (including sub-concepts), concept synonyms,
-overlapping concepts (i.e., that by definition are not disjoint),
-properties (including their domain and range), and inverse
-properties. We also provide blocks for plain-text descriptions of
-concepts and properties, and for referencing one or more existing
-ontologies (e.g., to extend an ontology or to define ontology
-articulations). Blocks can be sparse (inheriting from previous
-blocks), which simplifies the creation of large ontologies. Although
-not as expressive as full-fledged OWL ontologies, our approach can
-produce ontology structures essential for enhanced data discovery and
-integration, while at the same time providing a more accessible user
-interface for ecologists. Further, compared to many existing editors,
-our approach can be used to rapidly construct concept hierarchies,
-e.g., from long lists of keywords, within familiar spreadsheet
-software. We have field-tested the owlifier approach with ecologists
-and evolutionary biologists working with trait data, and found this
-application enabled them to quickly and easily comprehend and
-construct ontologies.
+approach is to allow scientists to use common spreadsheet-based tools
+to describe, in an intuitive way, different aspects of an ontology,
+and then to take these descriptions and convert them into full-fledged
+OWL ontologies using a software application called \owlifier.
+
+An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
+a predefined template structure for users to fill in. Each non-empty
+row in an owlifier table constitutes a block, where blocks can be used
+to define ontology classes (including subclasses), class synonyms,
+data- and object-type properties (and constraints), etc.  We also
+provide blocks for plain-text descriptions of concepts and properties,
+and for referencing one or more existing ontologies (e.g., to extend
+an existing ontology or to define ontology articulations). Blocks can
+be sparse (inheriting from previous blocks), which simplifies the
+creation of large ontologies.
+
+Although not as expressive as full-fledged OWL ontologies, our
+approach can produce ontology structures essential for enhanced data
+discovery and integration, while at the same time providing a more
+accessible user interface for ecologists. Further, compared to many
+existing editors, our approach can be used to rapidly construct
+concept hierarchies, e.g., from long lists of keywords, within
+familiar spreadsheet software. We have field-tested the owlifier
+approach with ecologists and evolutionary biologists working with
+trait data, and found this application enabled them to quickly and
+easily comprehend and construct ontologies.
+
+In this paper, we describe the syntax and semantics of \owlifier. We
+consider a set of blocks that generally follow the ...
 
@@
   and open-world reasoning, which can be confusing for non-experts
-\item we try to use more meaningful terms, when applicable (e.g., instead of datatype property, attribute, etc.)
+\item we try to use more meaningful terms, when applicable (e.g.,
+  instead of datatype property, attribute, etc.)
 \item our goal is to not support all constructs in OWL-DL, but to
@@
 
-\mypara{Import Block}
-
-\mypara{}Import blocks assign namespace labels to external
-ontologies. Each external ontology is imported into the current
-ontology. We refer to the ontologies of import blocks as
-\emph{imported ontologies}.  Using import blocks, concepts and
+\mypara{Import Block.} Import blocks assign namespace labels to
+external ontologies. Each external ontology is imported into the
+current ontology. We refer to the ontologies of import blocks as
+\emph{imported ontologies}.  Using import blocks, classes and
 properties of imported ontologies can be used within other blocks of
-the table.  Import blocks take the form
+an \owlifier\ table.  Import blocks take the form
 \begin{itemize}
@@
 block imports the ``earth realm'' ontology from the set of SWEET
-ontologies \cite{raskin:_seman_web_for_earth_and}.
+ontologies \cite{raskin:_seman_web_for_earth_and}
 \begin{tabbing}
@@
 where the ontology class denoting marine ecosystems can be refered to
-from within the corresponding \owlifier\ table using the syntax
-\texttt{sweet:MarineEcosystem}. This class is referred to an an
-``imported'' class.
+from within the corresponding \owlifier\ table using the syntax, e.g.,
+\sloppy\texttt{sweet:MarineEcosystem}. Because this class refers to a
+class in another ontology, we call it an ``imported'' class.
 
-\mypara{Class Block}
-
-\mypara{}Class blocks specify class and subclass relationships. Class
-blocks take the form
+\mypara{Concept Block.} Concept blocks introduce new OWL classes and
+specify subclass relationships. Imported classes may also be used
+within a concept block by prefixing the class name with a namespace
+label, e.g., when an ontology is being extended.  We use the term
+`concept' to avoid confusion with the use of the term 'class' in
+biological taxonomies and because `concept' is often more intuitive
+for users. Concept blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{class} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+    \texttt{concept} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 1$)
@@
 where each class $C_i$ is asserted in the current ontology to subsume
-a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in an entity block
-induces a DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
+a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a concept block
+induces the DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
 $C_{i+1}$ are imported classes, we say that the block defines an
-``articulation'' (i.e., mapping) between them. The following class
-block defines a simple subclass hierarchy
+``articulation'' (i.e., mapping) between the two classes. The
+following concept block defines a simple subclass hierarchy
 \begin{tabbing}
-~~\texttt{class PhysicalFeature AquaticPhysicalFeature River}
+  ~~\texttt{concept PhysicalFeature AquaticPhysicalFeature River}
 \end{tabbing}
-stating that a river is a subclass of aquatic physical features and
-that aquatic physical features are a type of physical feature.
+which states that river is a subclass of aquatic physical feature and
+that aquatic physical feature is a subclass of physical feature.
 
-\mypara{Synonym Block}
-
-\mypara{}Synonym blocks define an equivalence relationship between
-classes.  A synonym block has the form
+\mypara{Synonym Block.} Synonym blocks define an equivalence
+relationship between classes.  A synonym block has the form
 \begin{itemize}

### 2008-12-01 docs/pubs/ISEI6/main.tex

@@
 \begin{abstract}
-  Formal ontlogies promise to provide many added advantages to
-  ecological data management ... Most ontology editing frameworks and
-  tools, however, are design for knowledge-representation experts and
-  those with considerable understanding of the complexities of
-  ontology languages (i.e., description logics). This paper proposes
-  an approach for ontology creation based on ...
+  Data discovery and integration is important in many ecological
+  studies, especially those that concern broad-scale ecological
+  questions. While important, the discovery and integration of
+  ecological data is often a complex and time-consuming task for
+  researchers. This situation is often due to the use of informal,
+  ambiguous, and sometimes inconsistent terms for describing
+  ecological data.  Ontologies can help address this problem by
+  providing a mechanism to formally define ecological concepts that
+  can be used to more consistently annotate, relate, and search for
+  data sets.
+
+  Unlike in molecular biology or biomedicine, however, few ontology
+  development efforts exist within ecology. These efforts often
+  require considerable expertise in ontology languages and development
+  tools that limits the creation of ontologies by ecologists.  This
+  paper addresses this problem by allowing ecologists to use common
+  spreadsheet tools to describe different aspects of an ontology. We
+  introduce novel conventions for creating, relating, and constraining
+  concepts through spreadsheets, and provide software tools for
+  converting ontologies created in this way into equivalent OWL-DL
+  representations. We also consider inverse translations, i.e., that
+  convert ontologies represented using OWL-DL into our spreadsheet
+  format. Our approach allows large lists of terms to be easily
+  related and organized into concept hierarchies, and generally
+  provides a more intuitive and natural interface for ontology
+  creation by ecologists.
 \end{abstract}
@@
 
+Within the fields of molecular biology and biomedicine considerable
+effort has gone into developing ontologies for improving data
+discovery and integration
+\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
+benefits can be obtained for ecological data, far fewer efforts exist
+to develop richer and more consistent terminologies of ecology
+concepts
+\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
 The use of formal ontologies can significantly enhance metadata
-descriptions of ecological data by providing richer and more
-consistent terminologies \cite{tree,arees,parr}. Annotating data with
-ontology terms can both help users interpret data as well as enable
-advanced capabilities for data discovery and integration, e.g., by
-exploiting subsumption and part-of hierarchies as well as more formal
-constraints such as cardinality restrictions on properties and term
-equivalence \cite{ecoinf,other?}. Many efforts are underway to engage
-scientists in the development of ontologies to represent common
-ecological and biodiversity concepts. Most of these projects leverage
-the W3C Web Ontology Language (OWL) as a standard XML syntax for
-representing and sharing ontologies. A key advantage of OWL is that it
-is supported by a wide range of generic tools, including editors
-\cite{protege,swoop}, reasoning systems \cite{pellet,fact++}, query
-languages \cite{???}, and storage technologies \cite{jena,???}. Most
-of these tools, however, are primarily targeted at experts in
-knowledge engineering and software development. This is especially
-true with ontology editors (such as Protege, OWLed, SWOOP, etc.),
-which allow for very detailed ontology specifications, but at the same
-time require a considerable amount of understanding of the underlying
-ontology formalisms and syntax. Thus, we see the lack of suitable
-ontology editing tools for scientists as one of the major barriers for
-more wide-scale adoption of ontologies in ecology.
+descriptions of ecological data. Annotating data with ontology terms
+can both help users interpret data as well as enable advanced
+capabilities for data discovery and integration, e.g., by exploiting
+subsumption and part-of hierarchies as well as more formal constraints
+such as cardinality restrictions on properties and term
+equivalence. Many efforts are underway to engage scientists in the
+development of ontologies to represent common ecological and
+biodiversity concepts. Most of these projects leverage the W3C Web
+Ontology Language (OWL) \cite{smith04:_owl_web_ontol_languag_guide} as
+a standard XML syntax for representing and sharing ontologies. A key
+advantage of OWL is that it is supported by a wide range of generic
+tools, including editors
+\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
+reasoning systems
+\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
+languages
+\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
+and storage technologies
+\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
+however, are primarily targeted at experts in knowledge engineering
+and software development. This is especially true with ontology
+editors (such as Protege, SWOOP, etc.), which allow for very detailed
+ontology specifications, but at the same time require a considerable
+amount of understanding of the underlying ontology formalisms and
+syntax. Thus, we see the lack of suitable ontology editing tools for
+scientists as one of the major barriers for more wide-scale adoption
+of ontologies in ecology.
 

### 2008-12-01 docs/pubs/ISEI6/main.tex
abstract and introduction complete draft

@@
 
-\newcommand{\owlifier}{\textsf{Owlifier}}
+\newcommand{\Owlifier}{\textsf{Owlifier}}
+\newcommand{\owlifier}{\textsf{owlifier}}
 \newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}
+
+\newcommand{\secref}[1]{Sect.~\ref{#1}}
+\newcommand{\figref}[1]{Fig.~\ref{#1}}
 
@@
 \begin{abstract}
-  Data discovery and integration is important in many ecological
+  Discovery and integration of data is important in many ecological
   studies, especially those that concern broad-scale ecological
-  questions. While important, the discovery and integration of
-  ecological data is often a complex and time-consuming task for
-  researchers. This situation is often due to the use of informal,
-  ambiguous, and sometimes inconsistent terms for describing
-  ecological data.  Ontologies can help address this problem by
-  providing a mechanism to formally define ecological concepts that
-  can be used to more consistently annotate, relate, and search for
-  data sets.
-
-  Unlike in molecular biology or biomedicine, however, few ontology
-  development efforts exist within ecology. These efforts often
-  require considerable expertise in ontology languages and development
-  tools that limits the creation of ontologies by ecologists.  This
-  paper addresses this problem by allowing ecologists to use common
+  questions. Discovery and integration is often a difficult and
+  time-consuming task for researchers, in part because informal,
+  ambiguous, and sometimes inconsistent terms are used to describe
+  data semantics.  Ontologies can help address this problem by
+  providing a means to define ecological concepts to more consistently
+  annotate, relate, and search for data sets.
+  %
+  However, unlike in molecular biology or biomedicine, few ontology
+  development efforts exist within ecology. Ontology development often
+  requires considerable expertise in ontology languages and
+  development tools, which is often a barrier for ontology creation in
+  ecology. In this paper we address this problem by providing an
+  approach for ontology creation that allows ecologists to use common
   spreadsheet tools to describe different aspects of an ontology.  We
-  introduce novel conventions for creating, relating, and constraining
+  present conventions for creating, relating, and constraining
   concepts through spreadsheets, and provide software tools for
-  converting ontologies created in this way into equivalent OWL-DL
-  representations. We also consider inverse translations, i.e., that
+  converting these ontologies into equivalent OWL-DL
+  representations. We also consider inverse translations, i.e., to
   convert ontologies represented using OWL-DL into our spreadsheet
@@
    provides a more intuitive and natural interface for ontology
-  creation by ecologists.
+  development by ecologists.
 \end{abstract}
@@
 subsumption and part-of hierarchies as well as more formal constraints
-such as cardinality restrictions on properties and term
-equivalence. Many efforts are underway to engage scientists in the
-development of ontologies to represent common ecological and
-biodiversity concepts. Most of these projects leverage the W3C Web
-Ontology Language (OWL) \cite{smith04:_owl_web_ontol_languag_guide} as
-a standard XML syntax for representing and sharing ontologies. A key
-advantage of OWL is that it is supported by a wide range of generic
-tools, including editors
+such as cardinality restrictions on properties and term equivalence.
+
+Efforts to engage scientists in the development of ontologies
+typically leverage the W3C Web Ontology Language (OWL)
+\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
+for representing and sharing ontologies. A key advantage of OWL is
+that it is supported by a wide range of generic tools, including
+editors
 \cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
@@
 however, are primarily targeted at experts in knowledge engineering
-and software development. This is especially true with ontology
-editors (such as Protege, SWOOP, etc.), which allow for very detailed
-ontology specifications, but at the same time require a considerable
-amount of understanding of the underlying ontology formalisms and
-syntax. Thus, we see the lack of suitable ontology editing tools for
-scientists as one of the major barriers for more wide-scale adoption
-of ontologies in ecology.
+and software development familiar with the underlying description
+logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
+with ontology editors (such as Protege, SWOOP, etc.), which allow for
+very detailed ontology specifications, but at the same time require a
+considerable amount of understanding of the underlying ontology
+formalisms and syntax. Thus, we see the lack of suitable ontology
+editing tools for scientists as one of the major barriers for more
+wide-scale adoption of ontologies in ecology.
 
-Here, we present a novel approach for ontology creation that aims at
-being more intuitive for ecologists and that can be used to rapidly
+This paper presents a novel approach for ontology creation that aims
+at being more intuitive for ecologists and that can be used to rapidly
 construct large ontologies for describing scientific data. Our
@@
 and then to take these descriptions and convert them into full-fledged
-OWL ontologies using a software application called \owlifier.
+OWL ontologies using a software application called \owlifier. An
+\owlifier\ spreadsheet consists of a set of \emph{blocks} that have a
+predefined template structure for users to fill in. Each non-empty row
+in an \owlifier\ table constitutes a block. Each block defines
+different aspects of an ontology including ontology classes,
+subclasses, synonyms, and properties.  We also provide blocks for
+plain-text descriptions of classes and properties, and for referencing
+one or more existing ontologies (e.g., to extend an existing ontology
+or to define ontology articulations). Blocks can be sparse (inheriting
+from previous blocks), which can further simplify the creation of
+large ontologies.
 
-An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
-a predefined template structure for users to fill in. Each non-empty
-row in an owlifier table constitutes a block, where blocks can be used
-to define ontology classes (including subclasses), class synonyms,
-data- and object-type properties (and constraints), etc.  We also
-provide blocks for plain-text descriptions of concepts and properties,
-and for referencing one or more existing ontologies (e.g., to extend
-an existing ontology or to define ontology articulations). Blocks can
-be sparse (inheriting from previous blocks), which simplifies the
-creation of large ontologies.
-
-Although not as expressive as full-fledged OWL ontologies, our
-approach can produce ontology structures essential for enhanced data
-discovery and integration, while at the same time providing a more
-accessible user interface for ecologists. Further, compared to many
-existing editors, our approach can be used to rapidly construct
-concept hierarchies, e.g., from long lists of keywords, within
-familiar spreadsheet software. We have field-tested the owlifier
-approach with ecologists and evolutionary biologists working with
-trait data, and found this application enabled them to quickly and
+While not as expressive as OWL-DL, our approach can produce ontology
+structures essential for enhanced data discovery and integration,
+while at the same time provide a more accessible user interface for
+ecologists. Further, our approach can be used to rapidly construct
+class hierarchies from long lists of keywords using familiar
+spreadsheet software. For instance, an ecologist can easily list (or
+import) a set of terms, and then incrementally organize these into a
+class hierarchy, define properties and constraints, etc. We have used
+the \owlifier\ approach with ecologists and evolutionary biologists
+working with trait data, and found that it enabled them to quickly and
 easily comprehend and construct ontologies.
 
-In this paper, we describe the syntax and semantics of \owlifier. We
-consider a set of blocks that generally follow the ...
+The rest of this paper is organized as follows. In
+\secref{sec:owlifier} we describe the basic syntax and semantics of
+\owlifier. We define blocks that support a large subset of OWL-DL and
+that also generally follow the ontology creation guidelines defined in
+\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
+ontology creation using OWL-DL, e.g., by assuming classes are disjoint
+by default (unless specified otherwise) and by applying implicit
+property restriction closures \cite{rector04:_owl_pizzas}. In
+\secref{sec:properties} we describe additional characteristics of
+\owlifier\ and discuss issues with respect to classification and
+reasoning. In \secref{sec:implementation} we briefly describe the
+\owlifier\ implementation, and conlcude in \secref{sec:conclusion}
+with related and future work. In general, the goal of \owlifier\ is
+not to support all constructs in OWL-DL, but instead to provide a
+higher-level ontology syntax (via spreadsheet blocks) that is easy for
+ecologists to use and understand while also providing the necessary
+constructs for developing typical ecological ontologies. By compiling
+\owlifier\ to OWL-DL, we also allow for experts to refine and extend
+the ontology using more advanced ontology editing tools if necessary.
 
-Notes:
-\begin{itemize}
-\item we generally follow the guidelines in ontology creation of
-  Rector et al. in \cite{rector04:_owl_pizzas}.
-\item we also try to simplify certain aspects, e.g., disjoint classes
-  and open-world reasoning, which can be confusing for non-experts
-\item we try to use more meaningful terms, when applicable (e.g.,
-  instead of datatype property, attribute, etc.)
-\item our goal is to not support all constructs in OWL-DL, but to
-  provide the most commonly used ones for defining typical ontologies,
-  but still allow experts to refine the ontology in ontology editing
-  tools if necessary
-\end{itemize}
-
-\section{\owlifier\ Syntax and Semantics}
+\section{\Owlifier\ Syntax and Semantics}
+\label{sec:owlifier}
 
@@
 
-\mypara{Concept Block.} Concept blocks introduce new OWL classes and
+\mypara{Entity Block.} Entity blocks introduce new OWL classes and
 specify subclass relationships. Imported classes may also be used
@@
 label, e.g., when an ontology is being extended.  We use the term
-`concept' to avoid confusion with the use of the term 'class' in
-biological taxonomies and because `concept' is often more intuitive
-for users. Concept blocks take the form
+`entity to avoid confusion with the use of the term 'class' in
+biological taxonomies and because `entity' is used in \cite{oboe} to
+define ecological concepts. Entity blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{concept} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+    \texttt{entity} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 1$)
@@
 \begin{tabbing}
-  ~~\texttt{concept PhysicalFeature AquaticPhysicalFeature River}
+  ~~\texttt{entity PhysicalFeature AquaticPhysicalFeature River}
 \end{tabbing}

### 2008-12-01 docs/pubs/ISEI6/main.tex
up to relationship blocks

@@
 \newcommand{\owlifier}{\textsf{owlifier}}
-\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}
+
+\newcommand{\myblock}[1]{\vspace{12pt}\noindent\textbf{#1}}
 
@@
 type of block supported by \owlifier\ below. Here we assume that if
-any of the properties or concepts used in a block are defined, i.e.,
-are not imported from another ontology, then they are added to the
-ontology being defined by the \owlifier\ table.
+any properties or classes used in a block are undefined, i.e., are not
+imported from another ontology, then they are added to the ontology
+being specified via the \owlifier\ table. In general, we name blocks
+according to the more generic terms used in
+\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
+instead of those of OWL-DL. This allows \owlifier\ to generate
+ontologies that extend the observational model of
+\cite{madina07:_ontol_for_descr_and_synth} for data annotation, and in
+certain cases avoids confusion with established terms used within
+biology.
 
-\mypara{Import Block.} Import blocks assign namespace labels to
+\myblock{Import Blocks.} Import blocks assign namespace labels to
 external ontologies. Each external ontology is imported into the
@@
 properties of imported ontologies can be used within other blocks of
-an \owlifier\ table.  Import blocks take the form
+an \owlifier\ table.  Rows containing import blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|}\hline
-    \texttt{import} & $\textit{N}$ & $\textit{U}$ \\ \hline 
+    \texttt{import} & $n$ & $u$ \\ \hline 
   \end{tabular} 
 \end{itemize}
-where $\textit{N}$ is a namespace label and $\textit{U}$ is an OWL
-ontology URI. Classes and properties from imported ontologies are
-referenced by prefixing the namespace label $\textit{N}$ to the
-corresponding class or property name in the normal way. The following
-block imports the ``earth realm'' ontology from the set of SWEET
-ontologies \cite{raskin:_seman_web_for_earth_and}
+where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
+and properties from imported ontologies are referenced by prefixing
+the namespace label $n$ to the corresponding class or property name in
+the normal way. The following example block imports the ``earth
+realm'' ontology from the set of SWEET ontologies
+\cite{raskin:_seman_web_for_earth_and}
 \begin{tabbing}
@@
 \end{tabbing}
-where the ontology class denoting marine ecosystems can be refered to
-from within the corresponding \owlifier\ table using the syntax, e.g.,
-\sloppy\texttt{sweet:MarineEcosystem}. Because this class refers to a
-class in another ontology, we call it an ``imported'' class.
+With this import block the class denoting marine ecosystems can be
+referred to from within an \owlifier\ table using the expression
+\texttt{sweet:MarineEcosystem}. Because this class refers to a class
+in another ontology, we refer to it as an \emph{imported class}.
 
-\mypara{Entity Block.} Entity blocks introduce new OWL classes and
+\myblock{Entity Blocks.} Entity blocks introduce new OWL classes and
 specify subclass relationships. Imported classes may also be used
-within a concept block by prefixing the class name with a namespace
-label, e.g., when an ontology is being extended.  We use the term
-`entity to avoid confusion with the use of the term 'class' in
-biological taxonomies and because `entity' is used in \cite{oboe} to
-define ecological concepts. Entity blocks take the form
+within entity blocks by prefixing class names with namespace labels
+(see above).  We use the term `entity' to conform to the observational
+model of \cite{bowers08:_concep_model_framew_for_expres} and to avoid
+confusion with the term 'class' used within biological
+taxonomies. Rows containing entity blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{entity} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+    \texttt{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 1$)
 \end{itemize}
-where each class $C_i$ is asserted in the current ontology to subsume
-a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a concept block
-induces the DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
-$C_{i+1}$ are imported classes, we say that the block defines an
-``articulation'' (i.e., mapping) between the two classes. The
-following concept block defines a simple subclass hierarchy
+where each $c$ is a class, and each $c_i$ is asserted in the current
+ontology to subsume $c_{i+1}$, for $1 \le i < n$. That is, each $c_i$
+in a concept block induces the description-logic axiom $c_{i+1}
+\sqsubseteq c_i$.  If both $c_i$ and $c_{i+1}$ are imported classes,
+we say that the block defines an ``articulation'' (i.e., mapping)
+between the two classes. The following concept block defines a simple
+subclass hierarchy
 \begin{tabbing}
@@
 \end{tabbing}
-which states that river is a subclass of aquatic physical feature and
-that aquatic physical feature is a subclass of physical feature.
+stating that Physical Feature, Aquatic Physical Feature, and River are
+classes, and that River is a subclass of Aquatic Physical Feature, and
+Aquatic Physical Feature is a subclass of Physical Feature.
 
-\mypara{Synonym Block.} Synonym blocks define an equivalence
-relationship between classes.  A synonym block has the form
+\myblock{Synonym Blocks.} Synonym blocks define equivalence
+relationships between ontology classes.  Rows containing synonym
+blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{synonym} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+    \texttt{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-where each class $C_i$ is asserted as being equivalent to class
-$C_{i+1}$ in the current ontology, for $1 \le i < n$. For example, the
-block
+stating that each class $c_i$ is equivalent to class $c_{i+1}$ in the
+current ontology, for $1 \le i < n$. Each $c_i$ in a synonym blocks
+induces a description logic axiom of the form $c_i \equiv c_{i+1}$. As
+an example, the block
 \begin{tabbing}
@@
 \end{tabbing}
-defines maize and corn as synonym (i.e., equivalent) classes.
+defines Maize and Corn as synonym (i.e., equivalent) classes.
 
-\mypara{Overlap Block}
-
-\mypara{}Except in certain situations (described further below),
-classes are assumed to be disjoint.  Overlap blocks explicitly relax
-this assumption for a given set of classes by stating that a set of
-classes may have overlapping instances. An overlap block has the form
+\myblock{Overlap Blocks.} Except in certain situations (see
+\secref{sec:properties}), classes are assumed to be disjoint.  Overlap
+blocks explicitly relax this assumption by stating that a set of
+classes may have overlapping instances. Rows containing overlap blocks
+take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{overlap} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
+    \texttt{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-where each class $C_i$ is allowed to share instances with each concept
-$C_j$, for $1 \le i,j \le n$. In particular, $C_i$ and $C_j$ are not
-defined to be disjoint classes in the current ontology.
+stating that each class $c_i$ is allowed to share instances with each
+class $c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and
+$c_j$ in an overlap block are not defined to be disjoint classes in
+the current ontology. As an example, consider the following entity
+blocks that define the classes Estuary, Lagoon, and Marsh as
+subclasses of Ecological Habitats.
+\begin{tabbing}
+\begin{tabular}{lll}
+  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Estuary} \\ 
+  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Lagoon} \\ 
+  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Marsh} \\ 
+\end{tabular}
+\end{tabbing}
+Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
+as disjoint classes. To relax this assumption and allow Lagoons to
+also be kinds of Estuaries, we explicitly define the overlap block
+\begin{tabbing}
+\begin{tabular}{lll}
+  \texttt{overlap} & \texttt{Estuary} & \texttt{Lagoon}
+\end{tabular}
+\end{tabbing}
+In general, overlap blocks are rarely used but provide a mechanism to
+override the default behavior of \owlifier\ in asserting disjoint
+classes.
 
---- an example here ---
-
-\mypara{Relationship Block}
-
-\mypara{}Relationship blocks define required \emph{object} properties
-of classes. A property block has the form
+\myblock{Relationship Blocks.} Relationship blocks define required
+\emph{object} properties of classes. A property block has the form
 \begin{itemize}
@@
 
-\mypara{Transitive Block}
-
-\mypara{}Transitive blocks state that a property is transitive. That
-is, if $P$ is transitive and a concept instance $O_1$ is related to an
-instance $O_2$ by $P$, and $O_2$ is related to an instance $O_3$ by
-$P$, then $O_1$ is also by definition related to $O_3$ by $P$. A
-transitive block has the form
+\myblock{Transitive Blocks.} Transitive blocks state that a property
+is transitive. That is, if $P$ is transitive and a concept instance
+$O_1$ is related to an instance $O_2$ by $P$, and $O_2$ is related to
+an instance $O_3$ by $P$, then $O_1$ is also by definition related to
+$O_3$ by $P$. A transitive block has the form
 \begin{itemize}
@@
 
-\paragraph{Attribute Block.} Attribute blocks are used to define the
+\myblock{Attribute Blocks.} Attribute blocks are used to define the
 required \emph{datatype} properties of concepts. An attribute block
@@
 
-\paragraph{Value Block.} Value blocks define required datatype
-property \emph{constant values} for concepts. A value block has the
-form
+\myblock{Value Blocks.} Value blocks define required datatype property
+\emph{constant values} for concepts. A value block has the form
 \begin{itemize}
@@
 
-\paragraph{Inverse Block.} Inverse blocks state that two object
+\myblock{Inverse Blocks.} Inverse blocks state that two object
 properties are inverses of each other. That is, for inverse properties
@@
 
-\paragraph{Minimum Block.} Minimum blocks state the minimum
-number of properties $P$ an instance of a concept may have.
-Minimum blocks have the form
+\myblock{Minimum Blocks.} Minimum blocks state the minimum number of
+properties $P$ an instance of a concept may have.  Minimum blocks have
+the form
 \begin{itemize}
@@
 
-\paragraph{Maximum Block.} Maximum blocks state the maximum
+\myblock{Maximum Blocks.} Maximum blocks state the maximum
 number of properties $P$ an instance of a concept may have.
@@
 
-\paragraph{Sufficient Block.} Sufficient blocks state that any
-instance having a property $P$ to an instance of a concept $C_2$ is a
+\myblock{Sufficient Blocks.} Sufficient blocks state that any instance
+having a property $P$ to an instance of a concept $C_2$ is a
 sufficient condition for being an instance of a concept $C_1$. A
@@
 
-\paragraph{Description Block.} Description blocks assign plain-text
+\myblock{Description Blocks.} Description blocks assign plain-text
 definitions to concepts and properties. A description block has the
@@
 
-\paragraph{Note Block.} Note blocks add comments to the current
+\myblock{Note Blocks.} Note blocks add comments to the current
 ontology, and are ignored by \owlifier.  A note block has the form

### 2008-12-02 docs/pubs/ISEI6/main.tex
upto attribute blocks

@@
 imported from another ontology, then they are added to the ontology
-being specified via the \owlifier\ table. In general, we name blocks
+being specified by the \owlifier\ table. In general, we name blocks
 according to the more generic terms used in
@@
 ontologies that extend the observational model of
-\cite{madina07:_ontol_for_descr_and_synth} for data annotation, and in
-certain cases avoids confusion with established terms used within
-biology.
+\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
+for data annotation, and in certain cases avoids confusion with
+established terms used within biology.
 
@@
 the namespace label $n$ to the corresponding class or property name in
-the normal way. The following example block imports the ``earth
+the normal way. As an example, the following block imports the ``earth
 realm'' ontology from the set of SWEET ontologies
@@
 \begin{tabbing}
-  ~~\texttt{import sweet
-    http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
+  ~~\texttt{import}  \texttt{sweet} 
+  \texttt{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
 \end{tabbing}
-With this import block the class denoting marine ecosystems can be
-referred to from within an \owlifier\ table using the expression
+With this import block the class denoting Marine Ecosystems (a class
+defined in the above ontology) can be referred to from within an
+\owlifier\ table using the expression
 \texttt{sweet:MarineEcosystem}. Because this class refers to a class
@@
 
-\myblock{Entity Blocks.} Entity blocks introduce new OWL classes and
-specify subclass relationships. Imported classes may also be used
+\myblock{Entity Blocks.} Entity blocks are the primary blocks used to
+define ontologies. An entity block introduces new OWL classes and
+specifies subclass relationships. Imported classes may also be used
 within entity blocks by prefixing class names with namespace labels
-(see above).  We use the term `entity' to conform to the observational
-model of \cite{bowers08:_concep_model_framew_for_expres} and to avoid
-confusion with the term 'class' used within biological
+(as described above).  We use the term `entity' to conform to the
+observational model of \cite{bowers08:_concep_model_framew_for_expres}
+and to avoid confusion with the term `class' used within biological
 taxonomies. Rows containing entity blocks take the form
@@
 \begin{tabbing}
-  ~~\texttt{entity PhysicalFeature AquaticPhysicalFeature River}
+  ~~\texttt{entity} \texttt{PhysicalFeature} 
+    \texttt{AquaticPhysicalFeature} \texttt{River}
 \end{tabbing}
@@
 \begin{tabbing}
-~~\texttt{synonym Maize Corn}
+  ~~\texttt{synonym} \texttt{Maize} \texttt{Corn}
 \end{tabbing}
-defines Maize and Corn as synonym (i.e., equivalent) classes.
+defines Maize and Corn as synonyms, i.e., as equivalent classes.
 
 \myblock{Overlap Blocks.} Except in certain situations (see
-\secref{sec:properties}), classes are assumed to be disjoint.  Overlap
-blocks explicitly relax this assumption by stating that a set of
-classes may have overlapping instances. Rows containing overlap blocks
-take the form
+\secref{sec:properties}), classes are assumed to be disjoint in
+\owlifier.  Overlap blocks explicitly relax this assumption by stating
+that a set of classes may have overlapping instances. Rows containing
+overlap blocks take the form
 \begin{itemize}
@@
 subclasses of Ecological Habitats.
+begin
 \begin{tabbing}
-\begin{tabular}{lll}
-  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Estuary} \\ 
-  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Lagoon} \\ 
-  \texttt{entity} & \texttt{EcologicalHabitat} & \texttt{Marsh} \\ 
-\end{tabular}
+  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Estuary} \\ 
+  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Lagoon} \\ 
+  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Marsh} \\ 
 \end{tabbing}
 Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
-as disjoint classes. To relax this assumption and allow Lagoons to
-also be kinds of Estuaries, we explicitly define the overlap block
+as disjoint classes. To relax this assumption and allow, e.g., types
+of Lagoons to also be types of Estuaries, we explicitly add the
+following overlap block
 \begin{tabbing}
-\begin{tabular}{lll}
-  \texttt{overlap} & \texttt{Estuary} & \texttt{Lagoon}
-\end{tabular}
+  ~~\texttt{overlap} \texttt{Estuary} \texttt{Lagoon}
 \end{tabbing}
@@
 
-\myblock{Relationship Blocks.} Relationship blocks define required
-\emph{object} properties of classes. A property block has the form
+\myblock{Relationship Blocks.} Relationship blocks define
+\emph{required} object properties of classes. An object property
+within OWL is a property defined between two class instances. Rows
+containing relationship blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $R$ & $C_1$
-    & $C_2$ & \dots & $C_n$ \\ \hline
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $p$ & $c_1$
+    & $c_2$ & \dots & $c_n$ \\ \hline
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-where $R$ is an object property and each $C_i$ is a class, for $1 \le
-i \le n$.  For every class $C_i$, the property block induces the DL
-axiom \[C_i \sqsubseteq \exists R . C_{i+1} \] stating that each
-instance of $C_i$ has, amongst possibly other things, a relationship
-through $R$ to some instance of $C_{i+1}$.  The following example
+where $p$ is an object property and each $c_i$ is a class, for $1 \le
+i \le n$.  For every class $c_i$, the relationship block induces the
+DL axiom $c_i \sqsubseteq \exists p . c_{i+1}$ stating that each
+instance of $c_i$ has, amongst possibly other things, a relationship
+through $p$ to some instance of $c_{i+1}$.  For example, the following
+block states that the class California Voles live in Grassy Areas.
 \begin{tabbing}
-   ~~need example here
+  ~~\texttt{relationship} \texttt{livesIn} \texttt{CaliforniaVole} 
+    \texttt{GrassyArea}
 \end{tabbing}
+In some cases, a particular property can apply to a sequence of
+classes, and for convenience, each such class can be specified in a
+single block. For example, consider the following block.
+\begin{tabbing}
+  ~~\texttt{relationship} \texttt{directlyBelow} \texttt{Hypolimnion} 
+    \texttt{Thermocline} \texttt{Epilimnion} 
+\end{tabbing}
+This block states that, e.g., within a thermally stratified lake, the
+Hypolimnion layer is directly below the Thermocline layer, and the
+Thermocline layer is directly below the Epilimnion layer.
 
-\myblock{Transitive Blocks.} Transitive blocks state that a property
-is transitive. That is, if $P$ is transitive and a concept instance
-$O_1$ is related to an instance $O_2$ by $P$, and $O_2$ is related to
-an instance $O_3$ by $P$, then $O_1$ is also by definition related to
-$O_3$ by $P$. A transitive block has the form
+\myblock{Transitive Blocks.} Transitive blocks are special cases of
+relationship blocks in which the object property is asserted to be
+transitive. That is, if $p$ is a transitive property from individuals
+$o_1$ to $o_2$ and from $o_2$ to $o_3$, then $p$ is also assumed to be
+from $o_1$ to $o_3$. Rows containing transitive blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $R$ & $C_1$
-    & $C_2$ & \dots & $C_n$ \\ \hline
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $p$ & $c_1$
+    & $c_2$ & \dots & $c_n$ \\ \hline
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-where $P$ is an object property. For example, the block
-\begin{quote}\texttt{transitive} \texttt{hasPart} \texttt{Body}
+where $p$ is an object property and each $c_i$ is a class, for $1 \le
+i \le n$. As a simple example, the following transitive block
+\begin{tabbing}
+  ~~\texttt{transitive} \texttt{hasPart} \texttt{Body}
   \texttt{Head} \texttt{Eye} \texttt{Retina}
-\end{quote} states that a body has at least one head, a head has at
-least one eye, and an eye has at least one retina.
+\end{tabbing} 
+states that every Body has a Head as a part, every Head has an Eye as
+a part, and every Eye has a Retina as a part. Moreover, because the
+part-of property here is defined to be transitive, it is possible to
+infer that every Body also has an Eye and a Retina.
 
-\myblock{Attribute Blocks.} Attribute blocks are used to define the
-required \emph{datatype} properties of concepts. An attribute block
-has the form
+\myblock{Attribute Blocks.} Attribute blocks define required
+\emph{datatype} properties of concepts. Unlike object properties, a
+datatype property within OWL is a property defined between a class
+instance and a literal value (e.g., string, integer, etc.). Rows
+containing attribute blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $P$ & $D$ & $C_1$
-    & $C_2$ & \dots & $C_n$ \\ \hline
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $p$ & $c_1$
+    & $c_2$ & \dots & $c_n$ \\ \hline
   \end{tabular} \hfill ($n \ge 1$)
 \end{itemize}
-where $P$ is a datatype property, each $C_i$ is a concept for $1 \le i
-\le n$, and $D$ is a datatype (\texttt{anyValueType}, \texttt{string},
-\texttt{int}, etc.). For every concept $C_i$, the property block
-induces the DL axiom \[C_i \sqsubseteq (\exists P . D)\] stating that
-each instance of $C_i$ has, amongst possibly other things, a
-relationship through $P$ to a data value of type $D$.
+where $p$ is a datatype property and $c_i$ is a class, for $1 \le i
+\le n$. For each concept $c_i$, the attribute block induces the
+description logic axiom $c_i \sqsubseteq \exists p$ stating that each
+instance of $c_i$ has, amongst possibly other things, a property $p$
+with a literal value. As an example, the following ...
 

### 2008-12-02 docs/pubs/ISEI6/main.tex

@@
 
-\newcommand{\secref}[1]{Sect.~\ref{#1}}
-\newcommand{\figref}[1]{Fig.~\ref{#1}}
+\newcommand{\secref}[1]{Section~\ref{#1}}
+\newcommand{\figref}[1]{Figure~\ref{#1}}
 
@@
 with ontology editors (such as Protege, SWOOP, etc.), which allow for
-very detailed ontology specifications, but at the same time require a
-considerable amount of understanding of the underlying ontology
-formalisms and syntax. Thus, we see the lack of suitable ontology
-editing tools for scientists as one of the major barriers for more
-wide-scale adoption of ontologies in ecology.
+very detailed ontology specifications, but at the same time require
+considerable understanding of the underlying ontology formalisms and
+syntax. Thus, we see the lack of suitable ontology editing tools for
+scientists as one of the major barriers for more wide-scale adoption
+of ontologies in ecology.
 
@@
 While not as expressive as OWL-DL, our approach can produce ontology
-structures essential for enhanced data discovery and integration,
-while at the same time provide a more accessible user interface for
-ecologists. Further, our approach can be used to rapidly construct
-class hierarchies from long lists of keywords using familiar
-spreadsheet software. For instance, an ecologist can easily list (or
-import) a set of terms, and then incrementally organize these into a
-class hierarchy, define properties and constraints, etc. We have used
-the \owlifier\ approach with ecologists and evolutionary biologists
-working with trait data, and found that it enabled them to quickly and
-easily comprehend and construct ontologies.
+structures essential for enhanced data discovery and integration
+\cite{madin07:_ontol_for_descr_and_synth}, while at the same time
+provide a more accessible user interface for ecologists. Further, our
+approach can be used to rapidly construct class hierarchies from long
+lists of keywords using familiar spreadsheet software. For instance,
+an ecologist can easily list (or import) a set of terms, and then
+incrementally organize these into class hierarchies, and define
+relevant properties and their constraints. Based on our experiences
+using \owlifier\ with ecologists and evolutionary biologists studying
+trait data, we found that this approach can enable them to quickly and
+easily comprehend and construct useful ontologies.
 
@@
 property restriction closures \cite{rector04:_owl_pizzas}. In
-\secref{sec:properties} we describe additional characteristics of
+\secref{sec:characteristics} we describe additional characteristics of
 \owlifier\ and discuss issues with respect to classification and
@@
 
-\section{\Owlifier\ Syntax and Semantics}
+\section{The Syntax and Semantics of \Owlifier}
 \label{sec:owlifier}
@@
 type of block supported by \owlifier\ below. Here we assume that if
-any properties or classes used in a block are undefined, i.e., are not
-imported from another ontology, then they are added to the ontology
-being specified by the \owlifier\ table. In general, we name blocks
+any properties or classes used in a block are not imported from
+another ontology, then they are to be added to the ontology being
+specified by the \owlifier\ table. In general, we name blocks
 according to the more generic terms used in
-\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
+\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
 instead of those of OWL-DL. This allows \owlifier\ to generate
 ontologies that extend the observational model of
-\cite{bowers08:_concep_model_framew_for_expres,madina07:_ontol_for_descr_and_synth}
+\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
 for data annotation, and in certain cases avoids confusion with
-established terms used within biology.
+established terms commonly used within ecology.
 
@@
   \begin{tabular}{|l|l|l|}\hline
-    \texttt{import} & $n$ & $u$ \\ \hline 
+    \textsf{import} & $n$ & $u$ \\ \hline 
   \end{tabular} 
@@
 the namespace label $n$ to the corresponding class or property name in
-the normal way. As an example, the following block imports the ``earth
-realm'' ontology from the set of SWEET ontologies
-\cite{raskin:_seman_web_for_earth_and}
+the normal way. As an example, the following block imports the SWEET
+``Earth Realm'' ontology \cite{raskin:_seman_web_for_earth_and}
 \begin{tabbing}
-  ~~\texttt{import}  \texttt{sweet} 
-  \texttt{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
+  ~~\textsf{import}  \textsf{sweet} 
+  \textsf{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
 \end{tabbing}
 With this import block the class denoting Marine Ecosystems (a class
-defined in the above ontology) can be referred to from within an
+defined in the SWEET ontology) can be referred to from within an
 \owlifier\ table using the expression
-\texttt{sweet:MarineEcosystem}. Because this class refers to a class
+\textsf{sweet:MarineEcosystem}. Because this class refers to a class
 in another ontology, we refer to it as an \emph{imported class}.
@@
 within entity blocks by prefixing class names with namespace labels
-(as described above).  We use the term `entity' to conform to the
-observational model of \cite{bowers08:_concep_model_framew_for_expres}
-and to avoid confusion with the term `class' used within biological
-taxonomies. Rows containing entity blocks take the form
+(as described above).  Rows containing entity blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
+    \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 1$)
 \end{itemize}
-where each $c$ is a class, and each $c_i$ is asserted in the current
-ontology to subsume $c_{i+1}$, for $1 \le i < n$. That is, each $c_i$
-in a concept block induces the description-logic axiom $c_{i+1}
-\sqsubseteq c_i$.  If both $c_i$ and $c_{i+1}$ are imported classes,
-we say that the block defines an ``articulation'' (i.e., mapping)
-between the two classes. The following concept block defines a simple
-subclass hierarchy
+where each class $c_i$ is asserted in the current ontology to subsume
+$c_{i+1}$, for $1 \le i < n$. That is, each $c_i$ in an entity block
+induces the description-logic axiom $c_{i+1} \sqsubseteq c_i$.  If
+both $c_i$ and $c_{i+1}$ are imported classes, we say that the block
+defines an ``articulation'' (i.e., mapping) between the two
+classes. The following entity block defines a simple subclass
+hierarchy.
 \begin{tabbing}
-  ~~\texttt{entity} \texttt{PhysicalFeature} 
-    \texttt{AquaticPhysicalFeature} \texttt{River}
+  ~~\textsf{entity} \textsf{PhysicalFeature} 
+    \textsf{AquaticPhysicalFeature} \textsf{River}
 \end{tabbing}
-stating that Physical Feature, Aquatic Physical Feature, and River are
-classes, and that River is a subclass of Aquatic Physical Feature, and
-Aquatic Physical Feature is a subclass of Physical Feature.
+This block states that Physical Feature, Aquatic Physical Feature, and
+River are classes; River is a subclass of Aquatic Physical Feature;
+and Aquatic Physical Feature is a subclass of Physical Feature. The
+following entity block introduces a new class via an imported class.
+\begin{tabbing}
+  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
+    \textsf{IntertidalEcosystem}
+\end{tabbing}
+This block states that Intertidal Ecosystem is a subclass of the
+Marine Ecosystem class imported from the SWEET ontology. Similarly,
+assuming ``marine'' denotes an existing ontology of marine ecosystem
+concepts, the following entity block defines a simple class
+articulation.
+\begin{tabbing}
+  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
+    \textsf{marine:DeapSeaEcosystem}
+\end{tabbing}
+This block states that the Deap Sea Ecosystem class of the marine
+ontology is a subclass of the Marine Ecosystem class of the SWEET
+ontology.
 
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
+    \textsf{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-stating that each class $c_i$ is equivalent to class $c_{i+1}$ in the
-current ontology, for $1 \le i < n$. Each $c_i$ in a synonym blocks
-induces a description logic axiom of the form $c_i \equiv c_{i+1}$. As
-an example, the block
+where each class $c_i$ is equivalent to class $c_{i+1}$ in the current
+ontology, for $1 \le i < n$. That is, each $c_i$ in a synonym blocks
+induces a description logic axiom of the form $c_i \equiv
+c_{i+1}$. The following synonym block defines a simple equivalence
+relationship.
 \begin{tabbing}
-  ~~\texttt{synonym} \texttt{Maize} \texttt{Corn}
+  ~~\textsf{synonym} \textsf{Maize} \textsf{Corn}
 \end{tabbing}
-defines Maize and Corn as synonyms, i.e., as equivalent classes.
+This block states that the Maize and Corn classes are synonyms, i.e.,
+equivalent classes. Similar to entity blocks, synonym blocks often
+contain imported classes, e.g., to extend an existing ontology or to
+define ontology articulations.
 
-\myblock{Overlap Blocks.} Except in certain situations (see
-\secref{sec:properties}), classes are assumed to be disjoint in
-\owlifier.  Overlap blocks explicitly relax this assumption by stating
-that a set of classes may have overlapping instances. Rows containing
-overlap blocks take the form
+\myblock{Overlap Blocks.} Except in certain situations (described
+further in \secref{sec:characteristics}), classes are generally
+assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
+this assumption by stating that a give set of classes may have
+overlapping instances. Rows containing overlap blocks take the form
 \begin{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|}\hline
-    \texttt{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
+    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
   \end{tabular} \hfill ($n \ge 2$)
 \end{itemize}
-stating that each class $c_i$ is allowed to share instances with each
-class $c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and
-$c_j$ in an overlap block are not defined to be disjoint classes in
-the current ontology. As an example, consider the following entity
-blocks that define the classes Estuary, Lagoon, and Marsh as
-subclasses of Ecological Habitats.
-begin
+where each class $c_i$ is allowed to share instances with each class
+$c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and $c_j$
+in an overlap block are not defined to be disjoint classes in the
+current ontology. As an example, consider the following entity blocks
+that define the classes Estuary, Lagoon, and Marsh as subclasses of
+Ecological Habitats.
 \begin{tabbing}
-  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Estuary} \\ 
-  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Lagoon} \\ 
-  ~~\texttt{entity} \texttt{EcologicalHabitat} \texttt{Marsh} \\ 
+  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
+  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
+  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
 \end{tabbing}
@@
 \begin{tabbing}
-  ~~\texttt{overlap} \texttt{Estuary} \texttt{Lagoon}
+  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
 \end{tabbing}
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{relationship} & $p$ & $c_1$
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{relationship} & $p$ & $c_1$
     & $c_2$ & \dots & $c_n$ \\ \hline
@@
 \begin{tabbing}
-  ~~\texttt{relationship} \texttt{livesIn} \texttt{CaliforniaVole} 
-    \texttt{GrassyArea}
+  ~~\textsf{relationship} \textsf{livesIn} \textsf{CaliforniaVole} 
+    \textsf{GrassyArea}
 \end{tabbing}
@@
 \begin{tabbing}
-  ~~\texttt{relationship} \texttt{directlyBelow} \texttt{Hypolimnion} 
-    \texttt{Thermocline} \texttt{Epilimnion} 
+  ~~\textsf{relationship} \textsf{directlyBelow} \textsf{Hypolimnion} 
+    \textsf{Thermocline} \textsf{Epilimnion} 
 \end{tabbing}
@@
 \myblock{Transitive Blocks.} Transitive blocks are special cases of
-relationship blocks in which the object property is asserted to be
-transitive. That is, if $p$ is a transitive property from individuals
-$o_1$ to $o_2$ and from $o_2$ to $o_3$, then $p$ is also assumed to be
-from $o_1$ to $o_3$. Rows containing transitive blocks take the form
+relationship blocks where the object property is asserted to be
+transitive. If a property $p$ is declared to be transitive, whenever
+$p$ relates an individual $o_1$ to an individual $o_2$, and an
+individual $o_2$ to an individual $o_3$, then $p$ is also assumed to
+relate $o_1$ to $o_3$. Rows containing transitive blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|}\hline \texttt{transitive} & $p$ & $c_1$
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{transitive} & $p$ & $c_1$
     & $c_2$ & \dots & $c_n$ \\ \hline
@@
 where $p$ is an object property and each $c_i$ is a class, for $1 \le
-i \le n$. As a simple example, the following transitive block
+i \le n$. The following block is a simple example of a transitive
+relationship. 
 \begin{tabbing}
-  ~~\texttt{transitive} \texttt{hasPart} \texttt{Body}
-  \texttt{Head} \texttt{Eye} \texttt{Retina}
+  ~~\textsf{transitive} \textsf{hasPart} \textsf{Body}
+  \textsf{Head} \textsf{Eye} \textsf{Retina}
 \end{tabbing} 
-states that every Body has a Head as a part, every Head has an Eye as
-a part, and every Eye has a Retina as a part. Moreover, because the
-part-of property here is defined to be transitive, it is possible to
-infer that every Body also has an Eye and a Retina.
+This block states that every Body has a Head as a part, every Head has
+an Eye as a part, and every Eye has a Retina as a part. Moreover,
+because the part-of property here is defined to be transitive, it is
+possible to infer that every Body also has an Eye and a Retina through
+the corresponding relationship restrictiosn, i.e., $\textsf{Body}
+\sqsubseteq \exists \textsf{hasPart} . \textsf{Head}$, $\textsf{Head}
+\sqsubseteq \exists \textsf{hasPart} . \textsf{Eye}$, and
+$\textsf{Eye} \sqsubseteq \exists \textsf{hasPart} . \textsf{Retina}$.
+
+\myblock{Inverse Blocks.} Inverse blocks state that two object
+properties are inverses of each other. If $p_1$ and $p_2$ are defined
+to be inverse properties, whenever $p_1$ relates an individual $o_1$
+to an individual $o_2$ then $p_2$ is assumd to relate $o_2$ to $o_1$.
+Rows containing inverse blocks take the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|}\hline \textsf{inverse} & $p_1$ & $p_2$
+\\ \hline
+  \end{tabular}
+\end{itemize}
+where $p_1$ and $p_2$ are object properties. A common example of
+inverse properties are \texttt{partOf} and \texttt{hasPart}.
+
+\myblock{}*** HERE ***
 
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{attribute} & $p$ & $c_1$
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{attribute} & $p$ & $c_1$
     & $c_2$ & \dots & $c_n$ \\ \hline
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{value} & $P$ & $V$ & $C_1$
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{value} & $P$ & $V$ & $C_1$
     & $C_2$ & \dots & $C_n$ \\ \hline
@@
 
-\myblock{Inverse Blocks.} Inverse blocks state that two object
-properties are inverses of each other. That is, for inverse properties
-$P_1$ and $P_2$ and concept instances $O_1$ and $O_2$, if $P_1(O_1) =
-O_2$, then $P_2(O_2) = O_1$.  An inverse block has the form
-\begin{itemize}
-\item[]
-  \begin{tabular}{|l|l|l|}\hline \texttt{inverse} & $P_1$ & $P_2$
-\\ \hline
-  \end{tabular}
-\end{itemize}
-where $P_1$ and $P_2$ are object properties.
-
 \myblock{Minimum Blocks.} Minimum blocks state the minimum number of
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{minimum} & $P$ & $N$ & 
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{minimum} & $P$ & $N$ & 
     $C_1$ & $C_2$ & $\dots$ & $C_m$
@@
 \begin{quote}
-\texttt{minimum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
-\texttt{minimum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
+\textsf{minimum} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} \textsf{Nose}\\
+\textsf{minimum} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
 \end{quote}
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \texttt{maximum} & $P$ & $N$ & 
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{maximum} & $P$ & $N$ & 
     $C_1$ & $C_2$ & $\dots$ & $C_m$
@@
 \begin{quote}
-\texttt{maximum} \texttt{hasPart} 1 \texttt{Body} \texttt{Head} \\
-\texttt{maximum} \texttt{hasPart} 2 \texttt{Head} \texttt{Eye}
+\textsf{maximum} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} \\
+\textsf{maximum} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
 \end{quote}
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|}\hline \texttt{sufficient} & $C_1$ & $P$ & $C_2$
+  \begin{tabular}{|l|l|l|l|}\hline \textsf{sufficient} & $C_1$ & $P$ & $C_2$
     \\ \hline
@@
 \item[]
-  \begin{tabular}{|l|l|l|}\hline \texttt{description} & $T$ & $S$
+  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $T$ & $S$
     \\ \hline
@@
 \item[]
-  \begin{tabular}{|l|l|l|}\hline \texttt{note} & $S$
+  \begin{tabular}{|l|l|l|}\hline \textsf{note} & $S$
 \\ \hline
@@
 
-*** Say something about relaxing block syntax ... to make it easier to
-specify ontologies. Also, allow blocks to be given in any order.
-
-\section{\owlifier\ Properties and Reasoning}
-\label{sec:properties}
+\section{Additional Characteristics of \Owlifier}
+\label{sec:characteristics}
 
@@
 
-\section{\owlifier\ Implementation}
+*** Say something about relaxing block syntax ... to make it easier to
+specify ontologies. Also, allow blocks to be given in any order.
+
+\section{Implementation of \Owlifier}
 \label{sec:implementation}

### 2008-12-02 docs/pubs/ISEI6/main.tex

@@
   providing a means to define ecological concepts to more consistently
-  annotate, relate, and search for data sets.
-  %
-  However, unlike in molecular biology or biomedicine, few ontology
-  development efforts exist within ecology. Ontology development often
-  requires considerable expertise in ontology languages and
-  development tools, which is often a barrier for ontology creation in
-  ecology. In this paper we address this problem by providing an
-  approach for ontology creation that allows ecologists to use common
-  spreadsheet tools to describe different aspects of an ontology.  We
-  present conventions for creating, relating, and constraining
-  concepts through spreadsheets, and provide software tools for
-  converting these ontologies into equivalent OWL-DL
-  representations. We also consider inverse translations, i.e., to
-  convert ontologies represented using OWL-DL into our spreadsheet
-  format. Our approach allows large lists of terms to be easily
-  related and organized into concept hierarchies, and generally
-   provides a more intuitive and natural interface for ontology
-  development by ecologists.
+  annotate, relate, and search for data sets.  However, unlike in
+  molecular biology or biomedicine, few ontology development efforts
+  exist within ecology. Ontology development often requires
+  considerable expertise in ontology languages and development tools,
+  which is often a barrier for ontology creation in ecology. In this
+  paper we address this problem by providing an approach for ontology
+  creation that allows ecologists to use common spreadsheet tools to
+  describe different aspects of an ontology.  We present conventions
+  for creating, relating, and constraining concepts through
+  spreadsheets, and provide software tools for converting these
+  ontologies into equivalent OWL-DL representations. We also consider
+  inverse translations, i.e., to convert ontologies represented using
+  OWL-DL into our spreadsheet format. Our approach allows large lists
+  of terms to be easily related and organized into concept
+  hierarchies, and generally provides a more intuitive and natural
+  interface for ontology development by ecologists.
 \end{abstract}
@@
 $c_{i+1}$, for $1 \le i < n$. That is, each $c_i$ in an entity block
-induces the description-logic axiom $c_{i+1} \sqsubseteq c_i$.  If
+induces the description logic axiom $c_{i+1} \sqsubseteq c_i$.  If
 both $c_i$ and $c_{i+1}$ are imported classes, we say that the block
@@
 
+% \myblock{Property Blocks.} Property blocks define basic object property
+% domain and range constraints. , where the domain of an
+% object property denotes the set of individuals that can have the
+% property, and the range denotes the set of individuals that can be
+% related to domain individuals through the property. Rows containing
+% relationship blocks take the form 
+% \begin{itemize}
+% \item[]
+%   \begin{tabular}{|l|l|l|l|}\hline \textsf{property} & $p$ & $c_1$
+%     & $c_2$ \\ \hline
+%   \end{tabular} \hfill ($n \ge 2$)
+% \end{itemize}
+% where $p$ is the object property, class $c_1$ is the property domain,
+% and class $c_2$ is the property range.
+
 \myblock{Relationship Blocks.} Relationship blocks define
@@
 \end{itemize}
-where $p$ is an object property and each $c_i$ is a class, for $1 \le
-i \le n$.  For every class $c_i$, the relationship block induces the
-DL axiom $c_i \sqsubseteq \exists p . c_{i+1}$ stating that each
-instance of $c_i$ has, amongst possibly other things, a relationship
-through $p$ to some instance of $c_{i+1}$.  For example, the following
-block states that the class California Voles live in Grassy Areas.
+where $p$ is an object property and each $c$ is a class.  For every
+class $c_i$, the relationship block induces the DL axiom $c_i
+\sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
+is $p$-related to some instance of $c_{i+1}$, for $1 \le i < n$.  For
+example, the following block states that instances of the class
+California Voles live in Grassy Areas.
 \begin{tabbing}
@@
 \end{itemize}
-where $p$ is an object property and each $c_i$ is a class, for $1 \le
-i \le n$. The following block is a simple example of a transitive
-relationship. 
+where $p$ is an object property and each $c$ is a class.  The
+following block is a simple example of a transitive relationship.
 \begin{tabbing}
@@
 \end{tabbing} 
-This block states that every Body has a Head as a part, every Head has
-an Eye as a part, and every Eye has a Retina as a part. Moreover,
-because the part-of property here is defined to be transitive, it is
-possible to infer that every Body also has an Eye and a Retina through
-the corresponding relationship restrictiosn, i.e., $\textsf{Body}
-\sqsubseteq \exists \textsf{hasPart} . \textsf{Head}$, $\textsf{Head}
-\sqsubseteq \exists \textsf{hasPart} . \textsf{Eye}$, and
-$\textsf{Eye} \sqsubseteq \exists \textsf{hasPart} . \textsf{Retina}$.
+This block states that every instance of the class Body has a Head as
+a part, every instance of the class Head has an Eye as a part, and
+every instance of the class Eye has a Retina as a part. Moreover,
+because the \textsf{hasPart} property above is defined to be
+transitive, it is possible to infer that every instance of Body also
+has an Eye and a Retina as a part through the implied relationship
+restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
+. \textsf{Head}$, $\textsf{Head} \sqsubseteq \exists \textsf{hasPart}
+. \textsf{Eye}$, and $\textsf{Eye} \sqsubseteq \exists
+\textsf{hasPart} . \textsf{Retina}$.
+
+\myblock{Cardinality Blocks.} Cardinality blocks are also similar to
+relationship blocks.  We consider three types of cardinality blocks
+for defining minimum, maximum, and exact cardinality restrictions. A
+minimum block states the smallest number of properties $p$ to distinct
+individuals that an individual of a class may have.  Rows containing
+minimum blocks take the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{min} & $p$ & $m$ & 
+    $c_1$ & $c_2$ & $\dots$ & $c_n$
+    \\ \hline 
+  \end{tabular} \hfill $(n \ge 2)$
+\end{itemize}
+where $m$ is the minimum number of properties $p$ that instances of
+class $c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i
+< n$.  For each class $c_i$, a minimum cardinality block induces the
+description logic axiom $c_{i} \sqsubseteq \; (\le m) \; p.c_{i+1}$
+stating that each instance of $c_i$ must be $p$-related to at least
+$m$ unique instances of $c_{i+1}$. The following two blocks
+demonstrate simple minimum cardinality constraints.
+\begin{tabbing}
+  ~~\textsf{min} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
+  \textsf{Nose} \\
+  ~~\textsf{min} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
+\end{tabbing}
+The first block states that instances of the class Body have at least
+one Head as a part, which in turn have at least one Nose as a
+part.\footnote{Cardinality restrictions ensuring participation to at
+  least one property are typically not given through minimum
+  cardinality blocks since they are also implied by relationship
+  blocks.} The second block states that instances of the class Head
+have at least two Eyes as parts.
+
+A maximum block states the largest number of properties $p$ to
+distinct individuals that an individual of a class may have.  Rows
+containing maximum blocks take the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{max} & $p$ & $m$ & 
+    $c_1$ & $c_2$ & $\dots$ & $c_n$
+    \\ \hline
+  \end{tabular} \hfill $(n \ge 2)$
+\end{itemize}
+where $m$ is the maximum number of properties $p$ that instances of
+concept $c_i$ may have to instances of concept $c_{i+1}$, for $1 \le i
+< n$.  For each class $c_i$, a maximum cardinality block induces the
+description logic axiom $c_i \sqsubseteq \; (\ge m) \; p.c_{i+1}$
+stating that each instance of $c_i$ may be $p$-related to at most $m$
+unique instances of $c_{i+1}$. The following two blocks demonstrate
+simple maximum cardinality constraints.
+\begin{tabbing}
+  ~~\textsf{max} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
+  \textsf{Nose} \\
+  ~~\textsf{max} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
+\end{tabbing}
+The first block states that instances of the class Body have at most
+one Head as a part, which in turn has at most one Nose as a part. The
+second block states that instances of the class Head have at most two
+Eyes as parts.
+
+An exact block ensures both a minimum and maximum number $m$ of
+properties $p$ to distinct individuals that an individual of a class
+must have. Rows containing exact blocks take the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{exact} & $p$ & $m$ & 
+    $c_1$ & $c_2$ & $\dots$ & $c_n$
+    \\ \hline
+  \end{tabular} \hfill $(n \ge 2)$
+\end{itemize}
+where $m$ is the number of properties $p$ that instances of concept
+$c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
+For each class $c_i$, an exactly block induces the description logic
+axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
+instance of $c_i$ must be $p$-related to at $m$ unique instances of
+$c_{i+1}$.
 
@@
 to be inverse properties, whenever $p_1$ relates an individual $o_1$
-to an individual $o_2$ then $p_2$ is assumd to relate $o_2$ to $o_1$.
-Rows containing inverse blocks take the form
+to an individual $o_2$ then $p_2$ (as the inverse of $p_1$) is assumed
+to relate $o_2$ to $o_1$.  Rows containing inverse blocks take the
+form
 \begin{itemize}
@@
 where $p_1$ and $p_2$ are object properties. A common example of
-inverse properties are \texttt{partOf} and \texttt{hasPart}.
+inverse properties are \texttt{hasPart} and \texttt{partOf}, i.e., if
+an individual $o_1$ has an individual $o_2$ as a part, then $o_2$ is
+by definition a part of $o_1$.
 
-\myblock{}*** HERE ***
+% \myblock{Attribute Blocks.} Attribute blocks define required
+% \emph{datatype} properties of classes. Unlike object properties, a
+% datatype property within OWL is a property defined between a class
+% instance and a literal value (e.g., a string or integer). Rows
+% containing attribute blocks take the form
+% \begin{itemize}
+% \item[]
+%   \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{attribute} & $p$ & $c_1$
+%     & $c_2$ & \dots & $c_n$ \\ \hline
+%   \end{tabular} \hfill ($n \ge 1$)
+% \end{itemize}
+% where $p$ is a datatype property and $c_i$ is a class, for $1 \le i
+% \le n$. For each concept $c_i$, the attribute block induces the
+% description logic axiom $c_i \sqsubseteq \exists p$ stating that
+% each instance of $c_i$ has, amongst possibly other things, a
+% property $p$ with a literal value. Attributes are often useful for
+% As an example, the following ...
 
-\myblock{Attribute Blocks.} Attribute blocks define required
-\emph{datatype} properties of concepts. Unlike object properties, a
-datatype property within OWL is a property defined between a class
-instance and a literal value (e.g., string, integer, etc.). Rows
-containing attribute blocks take the form
+% \myblock{Value Blocks.} Value blocks define required datatype property
+% \emph{constant values} for concepts. A value block has the form
+% \begin{itemize}
+% \item[]
+%   \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{value} & $P$ & $V$ & $C_1$
+%     & $C_2$ & \dots & $C_n$ \\ \hline
+%   \end{tabular} \hfill ($n \ge 1$)
+% \end{itemize}
+% where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
+% n$, and $V$ is a datatype value. For each concept $C_i$, the value
+% block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
+% each instance of $C_i$ has a value $V$ for property $P$.  The value
+% restrictions stated by value blocks are often used for defining
+% so-called \emph{value partitions} \cite{co-ode}.
+
+\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
+blocks in that they state equivalences between classes. However,
+equivalence relationships are defined in sufficient blocks through
+property restrictions (i.e., via anonymous classes). We consider two
+main types of sufficient blocks. Rows containing sufficient blocks
+take one of the following forms.
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{attribute} & $p$ & $c_1$
-    & $c_2$ & \dots & $c_n$ \\ \hline
-  \end{tabular} \hfill ($n \ge 1$)
-\end{itemize}
-where $p$ is a datatype property and $c_i$ is a class, for $1 \le i
-\le n$. For each concept $c_i$, the attribute block induces the
-description logic axiom $c_i \sqsubseteq \exists p$ stating that each
-instance of $c_i$ has, amongst possibly other things, a property $p$
-with a literal value. As an example, the following ...
-
-\myblock{Value Blocks.} Value blocks define required datatype property
-\emph{constant values} for concepts. A value block has the form
-\begin{itemize}
+  \begin{tabular}{|l|l|l|l|l|l|}\hline 
+    \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
+    \textsf{exact} & $m$ & $c_1$ & $p$ & $c_2$\\ \hline
+  \end{tabular}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{value} & $P$ & $V$ & $C_1$
-    & $C_2$ & \dots & $C_n$ \\ \hline
-  \end{tabular} \hfill ($n \ge 1$)
-\end{itemize}
-where $P$ is a datatype property, $C_i$ is a concept for $1 \le i \le
-n$, and $V$ is a datatype value. For each concept $C_i$, the value
-block induces the DL axiom \[C_i \sqsubseteq (V \in P)\] stating that
-each instance of $C_i$ has a value $V$ for property $P$.  The value
-restrictions stated by value blocks are often used for defining
-so-called \emph{value partitions} \cite{co-ode}.
-
-\myblock{Minimum Blocks.} Minimum blocks state the minimum number of
-properties $P$ an instance of a concept may have.  Minimum blocks have
-the form
-\begin{itemize}
-\item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{minimum} & $P$ & $N$ & 
-    $C_1$ & $C_2$ & $\dots$ & $C_m$
-    \\ \hline
+  \begin{tabular}{|l|l|l|l|l|}\hline 
+    \textsf{sufficient} & \textsf{only} & $c_1$ & $p$ & \{\textsf{not}\} $c_2$\\ \hline
   \end{tabular}
 \end{itemize}
-where $N$ is the minimum number of properties $P$ that instances of
-concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
-and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
-(\le N P.C_{i+1})\] stating that each instance of $C_i$ must be related to
-at least $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
-\begin{quote}
-\textsf{minimum} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} \textsf{Nose}\\
-\textsf{minimum} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
-\end{quote}
-states that a body has at least one head and at least two eyes.
+The first form states that individuals that are $p$-related to at
+least, at most, or exactly $m$ individuals of class $c_2$ are by
+definition individuals of the class $c_1$. Sufficient blocks of this
+form induce either the description logic axiom $c_1 \equiv \; (\le m)
+\; p . c_2$, $c_1 \equiv \; (\ge m) \; p . c_2$, or $c_1 \equiv \; (=
+m) \; p . c_2$ depending on whether the qualifier \textsf{min},
+\textsf{max}, or \textsf{exact} is used, respectively. The second form
+states that individuals that are $p$-related to only individuals of
+class $c_2$ are by definition individuals of the class
+$c_1$. Sufficient blocks of this form induce description logic axioms
+$c_1 \equiv \forall p . c_2$. The complement of $c_2$ may also be used
+within these blocks via the expression \textsf{not $c_2$}. Sufficient
+blocks of this form induce description logic axioms $c_1 \equiv
+\forall p . \neg c_2$ stating that individuals that are $p$-related to
+only individual that are not of class $c_2$ (i.e., the complement of
+$c_2$) are by definition individuals of class $c_1$.  Multiple
+sufficient blocks for a class $c_1$ are treated as a single
+description logic axiom, e.g., if each of the above blocks are defined
+for $c_1$ we have $c_1 \equiv \; (\le m \; p .  c_2) \sqcap (\ge m \;
+p . c_2) \sqcap (\forall p.c_2)$. The following blocks are simple
+examples of sufficient conditions.
+\begin{tabbing}
+  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
+  \textsf{hasPart} \textsf{MammalProperEar}
+  ~~\textsf{sufficient} \textsf{maximum} 2 \textsf{MammalProper} 
+  \textsf{hasPart} \textsf{MammalProperEar}
+  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
+  \textsf{hasPart} \textsf{Hair}
 
-\myblock{Maximum Blocks.} Maximum blocks state the maximum
-number of properties $P$ an instance of a concept may have.
-Maximum blocks have the form
-\begin{itemize}
-\item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{maximum} & $P$ & $N$ & 
-    $C_1$ & $C_2$ & $\dots$ & $C_m$
-    \\ \hline
-  \end{tabular}
-\end{itemize}
-where $N$ is the maximum number of properties $P$ that instances of
-concept $C_1$ may have to instances of concept $C_2$, $C_2$ to $C_3$,
-and so on. A cardinality block induces the DL axiom \[C_i \sqsubseteq
-(\ge N P.C_{i+1})\] stating that each instance of $C_i$ may be related to
-at most $N$ unique instances of $C_{i+1}$ via $P$. For example, the blocks
-\begin{quote}
-\textsf{maximum} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} \\
-\textsf{maximum} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
-\end{quote}
-states that a body has at least one head and at least two eyes.
+  ~~\textsf{sufficient} \textsf{mimimum} 3 \textsf{MammalProper} 
+  \textsf{hasPart} \textsf{MiddleEarBone}
+\end{tabbing}
 
-\myblock{Sufficient Blocks.} Sufficient blocks state that any instance
-having a property $P$ to an instance of a concept $C_2$ is a
-sufficient condition for being an instance of a concept $C_1$. A
-sufficient block has the form
-\begin{itemize}
-\item[]
-  \begin{tabular}{|l|l|l|l|}\hline \textsf{sufficient} & $C_1$ & $P$ & $C_2$
-    \\ \hline
-  \end{tabular}
-\end{itemize}
-where $C_1$ is the target concept (i.e., denoting the concept
-definition), $P$ is the sufficient property, and $C_2$ is the
-sufficient concept. A sufficient block induces the DL axiom \[C_1
-\equiv \exists P.C_2\] Sufficient blocks provide a mechansism to
-construct simple class definitions (i.e., classes defined precisely by
-other classes), primarily for use with value partitions. [NOTE: these
-should be anded together?]
+\begin{tabbing}
+  ~~\textsf{sufficient} \textsf{relationship} \textsf{MammalProper} \textsf{hasPart} \textsf{SweatGland} \\
+  ~~\textsf{sufficient} \textsf{MammalProper} \textsf{hasPart} \textsf{Hair}
+\end{tabbing}
+These blocks state that
 
@@
 
+*** Using owlifier to represent values, value types, and untangling
+from owl-pizzas ***
+
 *** Say something about relaxing block syntax ... to make it easier to
@@
 
+related work: similar to dave's work on ``global taxonomic
+constraints'', which assume certain structures of hierarchies, e.g.,
+disjoint siblings.
+
 %\bibliographystyle{elsarticle-num}

### 2008-12-03 docs/pubs/ISEI6/main.tex
section 2 draft done

@@
   studies, especially those that concern broad-scale ecological
-  questions. Discovery and integration is often a difficult and
-  time-consuming task for researchers, in part because informal,
-  ambiguous, and sometimes inconsistent terms are used to describe
-  data semantics.  Ontologies can help address this problem by
-  providing a means to define ecological concepts to more consistently
-  annotate, relate, and search for data sets.  However, unlike in
-  molecular biology or biomedicine, few ontology development efforts
-  exist within ecology. Ontology development often requires
-  considerable expertise in ontology languages and development tools,
-  which is often a barrier for ontology creation in ecology. In this
-  paper we address this problem by providing an approach for ontology
-  creation that allows ecologists to use common spreadsheet tools to
-  describe different aspects of an ontology.  We present conventions
-  for creating, relating, and constraining concepts through
-  spreadsheets, and provide software tools for converting these
-  ontologies into equivalent OWL-DL representations. We also consider
-  inverse translations, i.e., to convert ontologies represented using
-  OWL-DL into our spreadsheet format. Our approach allows large lists
-  of terms to be easily related and organized into concept
-  hierarchies, and generally provides a more intuitive and natural
-  interface for ontology development by ecologists.
+  questions. Data discovery and integration is often a difficult and
+  time-consuming task for researchers, which is due in part to the use
+  of informal, ambiguous, and sometimes inconsistent terms for
+  describing data content.  Ontologies offer a solution to this
+  problem by providing consistent definitions of ecological concepts
+  that in turn can be used to annotate, relate, and search for data
+  sets.  However, unlike in molecular biology or biomedicine, few
+  ontology development efforts exist within ecology. Ontology
+  development often requires considerable expertise in ontology
+  languages and development tools, which is often a barrier for
+  ontology creation in ecology. In this paper we describe an approach
+  for ontology creation that allows ecologists to use common
+  spreadsheet tools to describe different aspects of an ontology.  We
+  present conventions for creating, relating, and constraining
+  concepts through spreadsheets, and provide software tools for
+  converting these ontologies into equivalent OWL-DL
+  representations. We also consider inverse translations, i.e., to
+  convert ontologies represented using OWL-DL into our spreadsheet
+  format. Our approach allows large lists of terms to be easily
+  related and organized into concept hierarchies, and generally
+  provides a more intuitive and natural interface for ontology
+  development by ecologists.
 \end{abstract}
@@
 discovery and integration
-\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
+\cite{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
 benefits can be obtained for ecological data, far fewer efforts exist
-to develop richer and more consistent terminologies of ecology
-concepts
+to develop broad and consistent terminologies within ecology
 \cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
 The use of formal ontologies can significantly enhance metadata
-descriptions of ecological data. Annotating data with ontology terms
-can both help users interpret data as well as enable advanced
-capabilities for data discovery and integration, e.g., by exploiting
-subsumption and part-of hierarchies as well as more formal constraints
-such as cardinality restrictions on properties and term equivalence.
+descriptions of ecological data. For instance, annotating data with
+ontology terms can both help users interpret data as well as enable
+advanced capabilities for data discovery and integration, e.g., by
+exploiting subsumption and part-of hierarchies as well as more formal
+constraints such as cardinality restrictions on properties and term
+equivalence \cite{madin08:_advan_ecolog_resear_with_ontol}.
 
@@
 considerable understanding of the underlying ontology formalisms and
-syntax. Thus, we see the lack of suitable ontology editing tools for
-scientists as one of the major barriers for more wide-scale adoption
-of ontologies in ecology.
+syntax. We see the lack of suitable ontology editing tools for
+scientists without expertise in knowledge representation as one of the
+major barriers for more wide-scale adoption of ontologies in ecology.
 
@@
 While not as expressive as OWL-DL, our approach can produce ontology
-structures essential for enhanced data discovery and integration
-\cite{madin07:_ontol_for_descr_and_synth}, while at the same time
-provide a more accessible user interface for ecologists. Further, our
-approach can be used to rapidly construct class hierarchies from long
-lists of keywords using familiar spreadsheet software. For instance,
-an ecologist can easily list (or import) a set of terms, and then
-incrementally organize these into class hierarchies, and define
-relevant properties and their constraints. Based on our experiences
-using \owlifier\ with ecologists and evolutionary biologists studying
-trait data, we found that this approach can enable them to quickly and
-easily comprehend and construct useful ontologies.
+structures that are essential for improved data discovery and
+integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
+important, because spreadsheet tools are frequently used by ecologists
+to store and analyze data, \owlifier\ can provide ecologists with a
+familiar and accessible user interface for ontology creation. This
+approach also leverages the ability of spreadsheet tools to organize
+and manipulate tabular data, e.g., allowing users to rapidly construct
+class hierarchies from long lists of keywords.  In this way, an
+ecologist can easily construct (or import) a set of terms, and then
+incrementally organize these into class hierarchies, properties, and
+constraints. In initial experiments with ecologists and evolutionary
+biologists studying trait data, we found that \owlifier\ enabled them
+to quickly and easily comprehend and construct useful ontologies.
 
@@
 
-An \owlifier\ table defines an OWL-DL
+As described above, an \owlifier\ table defines an OWL-DL
 \cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
@@
 non-empty row in an \owlifier\ table corresponds to a block. The type
-of the block is given in the first column of the row. We describe each
-type of block supported by \owlifier\ below. Here we assume that if
-any properties or classes used in a block are not imported from
-another ontology, then they are to be added to the ontology being
-specified by the \owlifier\ table. In general, we name blocks
-according to the more generic terms used in
+of the block is given in the first column of the row. 
+% This section describes in detail each type of block supported by
+% \owlifier.
+We assume that if any properties or classes used in a block are not
+imported from another ontology, then they are to be added to the
+ontology being specified by the \owlifier\ table (i.e., the
+``current'' ontology). In general, we name blocks according to the
+terms used in
 \cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
-instead of those of OWL-DL. This allows \owlifier\ to generate
-ontologies that extend the observational model of
-\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
-for data annotation, and in certain cases avoids confusion with
-established terms commonly used within ecology.
+as opposed to the names used for corresponding constructs in
+OWL-DL. This choice of block names helps to simplify terminology
+(e.g., we use ``relation'' below instead of ``object property''),
+allows \owlifier\ to easilty generate ontologies that extend the
+observational model of
+\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
+and avoids confusion with established terms commonly used within
+ecology (e.g., ``class'').
 
@@
 assuming ``marine'' denotes an existing ontology of marine ecosystem
-concepts, the following entity block defines a simple class
-articulation.
+concepts, the following block defines a simple class articulation.
 \begin{tabbing}
@@
 ontology is a subclass of the Marine Ecosystem class of the SWEET
-ontology.
+ontology (thus defining a mapping between these two ontologies).
 
@@
 where each class $c_i$ is equivalent to class $c_{i+1}$ in the current
-ontology, for $1 \le i < n$. That is, each $c_i$ in a synonym blocks
+ontology, for $1 \le i < n$. That is, each $c_i$ in a synonym block
 induces a description logic axiom of the form $c_i \equiv
-c_{i+1}$. The following synonym block defines a simple equivalence
+c_{i+1}$. The following synonym block creates a simple equivalence
 relationship.
@@
 \end{tabbing}
-This block states that the Maize and Corn classes are synonyms, i.e.,
-equivalent classes. Similar to entity blocks, synonym blocks often
-contain imported classes, e.g., to extend an existing ontology or to
-define ontology articulations.
+This block states that the Maize and Corn classes are synonyms
+(equivalent classes). Similar to entity blocks, synonym blocks often
+contain imported classes for extending existing ontologies or defining
+ontology mappings.
 
@@
 assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
-this assumption by stating that a give set of classes may have
+this assumption by stating that a given set of classes may have
 overlapping instances. Rows containing overlap blocks take the form
@@
 where each class $c_i$ is allowed to share instances with each class
-$c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and $c_j$
-in an overlap block are not defined to be disjoint classes in the
-current ontology. As an example, consider the following entity blocks
-that define the classes Estuary, Lagoon, and Marsh as subclasses of
+$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
+overlap block are not defined to be disjoint classes in the current
+ontology. As an example, consider the following entity blocks that
+define the classes Estuary, Lagoon, and Marsh as subclasses of
 Ecological Habitats.
@@
 \myblock{Relationship Blocks.} Relationship blocks define
-\emph{required} object properties of classes.  An object property
-within OWL is a property defined between two class instances. Rows
-containing relationship blocks take the form
+\emph{required} class object properties.  An object property within
+OWL is a property defined between two class instances. Rows containing
+relationship blocks take the form
 \begin{itemize}
@@
 where $p$ is an object property and each $c$ is a class.  For every
-class $c_i$, the relationship block induces the DL axiom $c_i
+class $c_i$, the relationship block induces the description logic axiom $c_i
 \sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
@@
 In some cases, a particular property can apply to a sequence of
-classes, and for convenience, each such class can be specified in a
-single block. For example, consider the following block.
+classes, and for convenience, each such class can be specified in
+\owlifier\ using a single block. For example, consider the following
+block.
 \begin{tabbing}
@@
 transitive. If a property $p$ is declared to be transitive, whenever
-$p$ relates an individual $o_1$ to an individual $o_2$, and an
-individual $o_2$ to an individual $o_3$, then $p$ is also assumed to
-relate $o_1$ to $o_3$. Rows containing transitive blocks take the form
+$p$ relates an individual $x$ to an individual $y$, and an
+individual $y$ to an individual $z$, then $p$ is also assumed to
+relate $x$ to $z$. Rows containing transitive blocks take the form
 \begin{itemize}
@@
 transitive, it is possible to infer that every instance of Body also
-has an Eye and a Retina as a part through the implied relationship
+has an Eye and a Retina as a part through the inherited relationship
 restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
@@
 $c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
-For each class $c_i$, an exactly block induces the description logic
+For each class $c_i$, an exact block induces the description logic
 axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
-instance of $c_i$ must be $p$-related to at $m$ unique instances of
+instance of $c_i$ must be $p$-related to $m$ unique instances of
 $c_{i+1}$.
@@
 properties are inverses of each other. If $p_1$ and $p_2$ are defined
-to be inverse properties, whenever $p_1$ relates an individual $o_1$
-to an individual $o_2$ then $p_2$ (as the inverse of $p_1$) is assumed
-to relate $o_2$ to $o_1$.  Rows containing inverse blocks take the
+to be inverse properties, whenever $p_1$ relates an individual $x$
+to an individual $y$ then $p_2$ (as the inverse of $p_1$) is assumed
+to relate $y$ to $x$.  Rows containing inverse blocks take the
 form
@@
 where $p_1$ and $p_2$ are object properties. A common example of
-inverse properties are \texttt{hasPart} and \texttt{partOf}, i.e., if
-an individual $o_1$ has an individual $o_2$ as a part, then $o_2$ is
-by definition a part of $o_1$.
+inverse properties are \textsf{hasPart} and \textsf{partOf}, i.e., if
+an individual $x$ has an individual $y$ as a part, then $y$ is
+by definition a part of $x$.
 
@@
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|}\hline 
+  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline 
     \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
-    \textsf{exact} & $m$ & $c_1$ & $p$ & $c_2$\\ \hline
-  \end{tabular}
+    \textsf{exact} & $m$ & $p$ & $c_1$ & $c_2$ & \dots & $c_n$\\ \hline
+  \end{tabular} \hfill $(n \ge 2)$
 \item[]
-  \begin{tabular}{|l|l|l|l|l|}\hline 
-    \textsf{sufficient} & \textsf{only} & $c_1$ & $p$ & \{\textsf{not}\} $c_2$\\ \hline
-  \end{tabular}
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline 
+    \textsf{sufficient} & \textsf{only} & $p$ & $c_1$ & $c_2$ & \dots & 
+    \{\textsf{not}\} $c_n$\\ \hline
+  \end{tabular} \hfill $(n \ge 2)$
 \end{itemize}
-The first form states that individuals that are $p$-related to at
-least, at most, or exactly $m$ individuals of class $c_2$ are by
-definition individuals of the class $c_1$. Sufficient blocks of this
-form induce either the description logic axiom $c_1 \equiv \; (\le m)
-\; p . c_2$, $c_1 \equiv \; (\ge m) \; p . c_2$, or $c_1 \equiv \; (=
-m) \; p . c_2$ depending on whether the qualifier \textsf{min},
-\textsf{max}, or \textsf{exact} is used, respectively. The second form
-states that individuals that are $p$-related to only individuals of
-class $c_2$ are by definition individuals of the class
-$c_1$. Sufficient blocks of this form induce description logic axioms
-$c_1 \equiv \forall p . c_2$. The complement of $c_2$ may also be used
-within these blocks via the expression \textsf{not $c_2$}. Sufficient
-blocks of this form induce description logic axioms $c_1 \equiv
-\forall p . \neg c_2$ stating that individuals that are $p$-related to
-only individual that are not of class $c_2$ (i.e., the complement of
-$c_2$) are by definition individuals of class $c_1$.  Multiple
-sufficient blocks for a class $c_1$ are treated as a single
-description logic axiom, e.g., if each of the above blocks are defined
-for $c_1$ we have $c_1 \equiv \; (\le m \; p .  c_2) \sqcap (\ge m \;
-p . c_2) \sqcap (\forall p.c_2)$. The following blocks are simple
-examples of sufficient conditions.
+The first form states that individuals $p$-related to at least, at
+most, or exactly $m$ individuals of class $c_{i+1}$ are by definition
+individuals of the class $c_i$, for $1 \le i < n$. Each class $c_i$ of
+a sufficient block of this form induces either the description logic
+axiom $c_i \equiv \; (\le m) \; p . c_{i+1}$, $c_i \equiv \; (\ge m)
+\; p . c_{i+1}$, or $c_i \equiv \; (= m) \; p . c_{i+1}$ depending on
+whether the qualifier \textsf{min}, \textsf{max}, or \textsf{exact} is
+used, respectively.
+
+The second form states that individuals $p$-related to only
+individuals of class $c_{i+1}$ are by definition individuals of the
+class $c_i$, for $1 \le i < n$. For each $c_i$ a sufficient block of
+this form induces the description logic axiom $c_i \equiv \forall p
+. c_{i+1}$. The complement of the last class in a sequence may also be
+used within these blocks via the expression \textsf{not
+  $c_n$}. Sufficient blocks of this form induce description logic
+axioms $c_{n-1} \equiv \forall p . \neg c_n$ stating that individuals
+that are $p$-related to only individuals that are not of class $c_n$
+(i.e., the complement of $c_n$) are by definition individuals of class
+$c_{n-1}$.  Multiple sufficient blocks for a class are treated as a
+single description logic axiom through intersection. For example, the
+blocks
 \begin{tabbing}
-  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
-  \textsf{hasPart} \textsf{MammalProperEar}
-  ~~\textsf{sufficient} \textsf{maximum} 2 \textsf{MammalProper} 
-  \textsf{hasPart} \textsf{MammalProperEar}
-  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
-  \textsf{hasPart} \textsf{Hair}
+  ~~\textsf{sufficient} \= \textsf{only} \= $m$ \= $c_1$ \= $c_2$ \= \kill
+  ~~\textsf{sufficient} \> \textsf{min} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
+  ~~\textsf{sufficient} \> \textsf{max} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
+  ~~\textsf{sufficient} \> \textsf{only} \> $p$ \> $c_1$ \> $c_2$
+\end{tabbing}
+result in the single axiom $c_1 \equiv \; (\le m \; p .  c_2) \sqcap
+(\ge m \; p . c_2) \sqcap (\forall p.c_2)$. The following blocks are
+simple examples of sufficient conditions.
+\begin{tabbing}
+  ~~\textsf{sufficient} \textsf{exact} \= \kill
+  ~~\textsf{sufficient} \textsf{min} \> 1 \textsf{hasPart} 
+  \textsf{Mammal} \textsf{Hair} \\
+  ~~\textsf{sufficient} \textsf{exact} \> 2  \textsf{hasPart} 
+  \textsf{Mammal} \textsf{MammalianEar} \\
+  ~~\textsf{sufficient} \textsf{only} \> \textsf{feedsYoungWith} 
+  \textsf{Mammal} \textsf{MammaryGland}\\
+  ~~\textsf{sufficient} \textsf{exact} \> 3 \textsf{hasPart} 
+  \textsf{MammalianEar} \textsf{MiddleEarBone}
+\end{tabbing}
+The first three blocks state that individuals that have Hair as a
+part, that have exactly two Mammalian Ears as parts, and that feed
+their young via Mammary Glands are Mammals. The last block states
+individuals that have exactly three Middle Ear Bones are Mammalian
+Ears.
 
-  ~~\textsf{sufficient} \textsf{mimimum} 3 \textsf{MammalProper} 
-  \textsf{hasPart} \textsf{MiddleEarBone}
-\end{tabbing}
-
-\begin{tabbing}
-  ~~\textsf{sufficient} \textsf{relationship} \textsf{MammalProper} \textsf{hasPart} \textsf{SweatGland} \\
-  ~~\textsf{sufficient} \textsf{MammalProper} \textsf{hasPart} \textsf{Hair}
-\end{tabbing}
-These blocks state that
-
-\myblock{Description Blocks.} Description blocks assign plain-text
-definitions to concepts and properties. A description block has the
-form
+\myblock{Comment Blocks.} Comment blocks provide a mechanisms to add
+plain-text comments to \owlifier\ tables. A description block attaches
+a plain-text comment to classes and properties. Rows containing
+description blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $T$ & $S$
+  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $c$ or $p$ & $s$
     \\ \hline
@@
 \end{itemize}
-where $T$ is either a property or a concept and $S$ is a description
-string.
-
-\myblock{Note Blocks.} Note blocks add comments to the current
-ontology, and are ignored by \owlifier.  A note block has the form
+where the string $s$ is associated as a comment to the class $c$ or
+property $p$. A note block allows attaches comments to the current
+ontology. Rows containing note blocks take the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|}\hline \textsf{note} & $S$
+  \begin{tabular}{|l|l|l|}\hline \textsf{note} & $s$
 \\ \hline
@@
 \end{itemize}
-where $S$ is a comment string.
+where $s$ is a comment string.
 
-\section{Additional Characteristics of \Owlifier}
+\section{Additional Features of \Owlifier}
 \label{sec:characteristics}
 
-Some desirable properties: 
-\begin{itemize}
-\item non-ambiguous (no block leads to ambiguous DL axioms)
-\item ``reasonable'' (not everything has to be explicitly stated)
-\item ???
-\end{itemize}
+1. non-ambiguous blocks (no block leads to an ambiguous DL axiom)
 
-*** Using owlifier to represent values, value types, and untangling
-from owl-pizzas ***
+2. disjoint classes are computed by default
 
-*** Say something about relaxing block syntax ... to make it easier to
-specify ontologies. Also, allow blocks to be given in any order.
+3. property closures
+
+4. is reasonable, e.g., it is possible to define values, value types,
+and untangling
+
+5. we allow sparse blocks, and blocks can be given in any order
 
@@
 
-Flags: 
-\begin{itemize}
-\item delimeter characters 
-\item perform classification
-\item warnings
-\item owlifier to owl and owl to owlifier
-\item ??? 
-\end{itemize}
+figure showing basics owlifier interaction, description.
+
+current status of implementation (subset of the syntax described here), perform classification via pellet/jena, performs disjoint check
+
+ongoing extensions: additional syntax; specify delimeter characters;
+perform inverse translations (easy to perform, e.g., ...); perform
+closure computation (again, easy).
 
@@
 constraints'', which assume certain structures of hierarchies, e.g.,
-disjoint siblings.
+disjoint siblings. protege approach/plugin
+
+future work: the ongoing extensions described above, support seamless
+conversion into and out of owlifier, e.g., if someone extends owlifier
+ontology, re-apply extensions after new owlifier changes. Also,
+support translation to oboe.
 

### 2008-12-03 docs/pubs/ISEI6/main.tex

@@
 \myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
-blocks in that they state equivalences between classes. However,
-equivalence relationships are defined in sufficient blocks through
-property restrictions (i.e., via anonymous classes). We consider two
-main types of sufficient blocks. Rows containing sufficient blocks
-take one of the following forms.
-\begin{itemize}
-\item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline 
-    \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
-    \textsf{exact} & $m$ & $p$ & $c_1$ & $c_2$ & \dots & $c_n$\\ \hline
-  \end{tabular} \hfill $(n \ge 2)$
-\item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|}\hline 
-    \textsf{sufficient} & \textsf{only} & $p$ & $c_1$ & $c_2$ & \dots & 
-    \{\textsf{not}\} $c_n$\\ \hline
-  \end{tabular} \hfill $(n \ge 2)$
-\end{itemize}
-The first form states that individuals $p$-related to at least, at
-most, or exactly $m$ individuals of class $c_{i+1}$ are by definition
-individuals of the class $c_i$, for $1 \le i < n$. Each class $c_i$ of
-a sufficient block of this form induces either the description logic
-axiom $c_i \equiv \; (\le m) \; p . c_{i+1}$, $c_i \equiv \; (\ge m)
-\; p . c_{i+1}$, or $c_i \equiv \; (= m) \; p . c_{i+1}$ depending on
-whether the qualifier \textsf{min}, \textsf{max}, or \textsf{exact} is
-used, respectively.
-
-The second form states that individuals $p$-related to only
-individuals of class $c_{i+1}$ are by definition individuals of the
-class $c_i$, for $1 \le i < n$. For each $c_i$ a sufficient block of
-this form induces the description logic axiom $c_i \equiv \forall p
-. c_{i+1}$. The complement of the last class in a sequence may also be
-used within these blocks via the expression \textsf{not
-  $c_n$}. Sufficient blocks of this form induce description logic
-axioms $c_{n-1} \equiv \forall p . \neg c_n$ stating that individuals
-that are $p$-related to only individuals that are not of class $c_n$
-(i.e., the complement of $c_n$) are by definition individuals of class
-$c_{n-1}$.  Multiple sufficient blocks for a class are treated as a
-single description logic axiom through intersection. For example, the
-blocks
+blocks in that they state equivalences between classes. A sufficient
+block serves as a modifier to entity blocks, relationship blocks
+(including transitive blocks), and cardinality blocks, modifying the
+associated description logic axioms by using equivalence ($\equiv$) in
+place of sublass ($\sqsubseteq$). For example the block
 \begin{tabbing}
-  ~~\textsf{sufficient} \= \textsf{only} \= $m$ \= $c_1$ \= $c_2$ \= \kill
-  ~~\textsf{sufficient} \> \textsf{min} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
-  ~~\textsf{sufficient} \> \textsf{max} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
-  ~~\textsf{sufficient} \> \textsf{only} \> $p$ \> $c_1$ \> $c_2$
+  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
 \end{tabbing}
-result in the single axiom $c_1 \equiv \; (\le m \; p .  c_2) \sqcap
-(\ge m \; p . c_2) \sqcap (\forall p.c_2)$. The following blocks are
-simple examples of sufficient conditions.
+induces the description logic axiom $\textsf{Mammal} \equiv \exists
+\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
+that has Hair as a part is a Mammal. Additionally, we allow
+relationship blocks within a sufficient block to be extended with
+\textsf{not} to state that absense of the property is a defining
+characteristic of the class. For example the following blocks
 \begin{tabbing}
-  ~~\textsf{sufficient} \textsf{exact} \= \kill
-  ~~\textsf{sufficient} \textsf{min} \> 1 \textsf{hasPart} 
-  \textsf{Mammal} \textsf{Hair} \\
-  ~~\textsf{sufficient} \textsf{exact} \> 2  \textsf{hasPart} 
-  \textsf{Mammal} \textsf{MammalianEar} \\
-  ~~\textsf{sufficient} \textsf{only} \> \textsf{feedsYoungWith} 
-  \textsf{Mammal} \textsf{MammaryGland}\\
-  ~~\textsf{sufficient} \textsf{exact} \> 3 \textsf{hasPart} 
-  \textsf{MammalianEar} \textsf{MiddleEarBone}
+  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
+  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
 \end{tabbing}
-The first three blocks state that individuals that have Hair as a
-part, that have exactly two Mammalian Ears as parts, and that feed
-their young via Mammary Glands are Mammals. The last block states
-individuals that have exactly three Middle Ear Bones are Mammalian
-Ears.
+induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
+\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
+any Mammal that does not have an Epipubic Bone as a part is a
+Eutheria. Note also that multiple sufficient blocks for a particular
+class result in a single axiom in which each constraint is combined
+via conjunction ($\sqcap$).
 
@@
 
-1. non-ambiguous blocks (no block leads to an ambiguous DL axiom)
+In this section, we briefly describe some of the additional features
+of \owlifier, specifically focusing on the use of disjoint classes,
+property closures, \owlifier\ reasoning, and additional block syntax.
 
-2. disjoint classes are computed by default
+\subsection{Disjoint Class Inference}
 
-3. property closures
+OWL is based on the open world assumption, which can lead to a number
+of ontology development ``pitfalls'' for those new to the language
+\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
+example is in the creation of disjoint classes (another concerns
+property closures, described below). In particular, unless explicitly
+stated, distinct classes within an OWL-DL ontology are never assumed
+to be disjoint. However, in many ontologies a large number of classes
+are typically defined as being disjoint (e.g., sibling classes), and
+stating these disjoint constraints is often time-consuming since each
+pair of classes must be given an explicit disjoint assertion. Editors
+such as Protege \cite{knublauch04:_editin_descr_logic_ontol_with}
+provide shortcuts via the user interface to create specific sets of
+disjoint assertions, e.g., by allowing a user to define all children
+of a particular class as disjoint. In general, however, many users
+expect such classes to be disjoint by default
+\cite{rector04:_owl_pizzas} and this expectation often leads to
+modeling errors.
+
+Alternatively, the default assumption in \owlifier\ is that distinct
+classes are disjoint. Specifically, as an \owlifier\ table is
+converted to an OWL-DL ontology, the system analyzes the class
+hierarchy structure and identifies pairs of classes that are: (1) not
+related via subclass relations (either direct or indirect subclasses);
+(2) not defined as synonyms; and (3) not explicitly defined to overlap
+via an overlap block. Each such pair of classes is then asserted by
+\owlifier\ in the resulting ontology as being disjoint. As described
+in \cite{rector04:_owl_pizzas}, undeclared disjoint classes are a
+common problem in ontology development using OWL-DL and often limit
+the utility of reasoning systems (by limiting the inferences that can
+be obtained). The approach employed in \owlifier\ for handling
+disjoint classes makes the common expectations of users the default
+case, which in general should lead to a more intuitive ontology
+editing environment and an overall fewer number of modeling mistakes.
+
+\subsection{Object Property Closure}
+
+Another common mistake for new users of OWL-DL concerns the user of
+universal rather than existential property restrictions
+\cite{rector04:_owl_pizzas}. In particular, an expression such as
+$\texttt{Body} \sqsubseteq \forall \textsf{hasPart}$
+
+\subsection{Reasoning in \owlifier}
 
@@
 and untangling
+
+\subsection{Sparse Blocks}
+
+1. non-ambiguous blocks (no block leads to an ambiguous DL axiom)
 
@@
 ontology, re-apply extensions after new owlifier changes. Also,
-support translation to oboe.
+support translation to oboe. Additional testing and evaulation.
 

### 2008-12-03 docs/pubs/ISEI6/main.tex

@@
 % other options: times
+\usepackage{graphicx}
 
@@
 ontology creation using OWL-DL, e.g., by assuming classes are disjoint
-by default (unless specified otherwise) and by applying implicit
-property restriction closures \cite{rector04:_owl_pizzas}. In
-\secref{sec:characteristics} we describe additional characteristics of
+by default (unless specified otherwise) and supporting sufficient
+restrictions \cite{rector04:_owl_pizzas}. In
+\secref{sec:features} we describe additional characteristics of
 \owlifier\ and discuss issues with respect to classification and
@@
 \myblock{Overlap Blocks.} Except in certain situations (described
-further in \secref{sec:characteristics}), classes are generally
+further in \secref{sec:features}), classes are generally
 assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
@@
 \section{Additional Features of \Owlifier}
-\label{sec:characteristics}
+\label{sec:features}
 
@@
 of \owlifier, specifically focusing on the use of disjoint classes,
-property closures, \owlifier\ reasoning, and additional block syntax.
+\owlifier\ reasoning, and additional block syntax.
 
@@
 \cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
-example is in the creation of disjoint classes (another concerns
-property closures, described below). In particular, unless explicitly
-stated, distinct classes within an OWL-DL ontology are never assumed
-to be disjoint. However, in many ontologies a large number of classes
-are typically defined as being disjoint (e.g., sibling classes), and
-stating these disjoint constraints is often time-consuming since each
-pair of classes must be given an explicit disjoint assertion. Editors
-such as Protege \cite{knublauch04:_editin_descr_logic_ontol_with}
-provide shortcuts via the user interface to create specific sets of
-disjoint assertions, e.g., by allowing a user to define all children
-of a particular class as disjoint. In general, however, many users
-expect such classes to be disjoint by default
-\cite{rector04:_owl_pizzas} and this expectation often leads to
-modeling errors.
+example is in the creation of disjoint classes.  In particular, unless
+explicitly stated, distinct classes within an OWL-DL ontology are
+never assumed to be disjoint. However, in many ontologies a large
+number of classes are typically defined as being disjoint (e.g.,
+sibling classes), and stating these disjoint constraints is often
+time-consuming since each pair of classes must be given an explicit
+disjoint assertion. Editors such as Protege
+\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
+via the user interface to create specific sets of disjoint assertions,
+e.g., by allowing a user to define all children of a particular class
+as disjoint. In general, however, many users expect such classes to be
+disjoint by default \cite{rector04:_owl_pizzas} and this expectation
+often leads to modeling errors.
 
@@
 
-\subsection{Object Property Closure}
+\subsection{Reasoning in \Owlifier}
 
-Another common mistake for new users of OWL-DL concerns the user of
-universal rather than existential property restrictions
-\cite{rector04:_owl_pizzas}. In particular, an expression such as
-$\texttt{Body} \sqsubseteq \forall \textsf{hasPart}$
+Blocks in \owlifier\ are unambiguous, i.e., for every \owlifier\ block
+(or set of blocks in the case of sufficient blocks) there is a
+well-defined set of corresponding OWL-DL axioms. This property is
+important because it implies that reasoning can be performed over
+ontologies defined in \owlifier\ using standard OWL-DL reasoners.  We
+use this capability in our current \owlifier\ implementation
+(described further in \secref{sec:implementation}) to verify
+ontologies defined using \owlifier\ and report possible errors to
+users.
 
-\subsection{Reasoning in \owlifier}
+In general, new axioms are inferred from an \owlifier\ ontology
+primarily from the use of synonym blocks, sufficient blocks, and
+transitive blocks (whose inferences are described in the previous
+section). For instance, let $A$, $B$, and $C$ be classes and $P$ be an
+object property. From an axiom $A \equiv B$ generated from a synonym
+block, and an axiom $B \sqsubseteq \exists P.C$ generated from a
+relationship block, the axiom $A \sqsubseteq \exists P.B$ would be
+inferred. Thus, the axioms of a particular class are ``inherited'' by
+all of the synonyms of the class. Similarly, from an axiom $A \equiv
+\exists P. C$ generated from a sufficent block, and an axiom $B
+\sqsubseteq \exists P.C$ generated from a relationship block, the axiom
+$B \sqsubseteq A$ would be inferred. For both the case of synonym and
+sufficient blocks, the use of equivalence permits a number of
+additional inferences to be made via an OWL-DL reasoner.
 
-4. is reasonable, e.g., it is possible to define values, value types,
-and untangling
+As described in \cite{rector04:_owl_pizzas} additional reasoning can
+occur within OWL-DL ontologies when domain and range axioms are
+providedq (as well as, e.g., property closure axioms).  We explicitly
+do not consider these constraints in the current version of \owlifier\
+because they also often result in modeling errors for inexperienced
+OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the more
+traditional description-logic approaches that do not have explicit
+domain and range axioms.  Although not currently supported in
+\owlifier, domain and range constraints could be inferred from given
+relationship blocks as well as property closure axioms.
 
@@
 
-1. non-ambiguous blocks (no block leads to an ambiguous DL axiom)
+To help simplify the creation of class hierarchies and property
+sequences (including transitive, cardinality, and sufficient blocks),
+we allow for ``sparse'' blocks that inherit missing information from
+their closest proceeding block. For instance the following entity
+blocks
+\begin{tabbing}
+  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
+  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
+  \textsf{Bay}\\
+  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
+  \textsf{Fjord} \\
+  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
+  \textsf{TidalMarsh} \\ 
+  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
+  \textsf{SaltMarsh}
+\end{tabbing}
+can be equivalently represented in \owlifier\ using the following
+sparse blocks.
+\begin{tabbing}
+  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
+  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
+  \textsf{Bay}\\
+  \> \> \> \textsf{Fjord} \\
+  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
+  \> \> \> \textsf{SaltMarsh}
+\end{tabbing}
+In general, the use of sparse blocks simplifies ontology creation by
+not requiring users to enter every (redundant, leading) field to be
+explicitly given, which in turn can simplify the overall layout of the
+ontology within a spreadsheet.  Additionaly, \owlifier\ does not place
+constraints on the order of blocks within a spreadsheet. Classes also
+do not need to explicitly be defined within an entity block, e.g.,
+classes without corresponding entity blocks can be introduced simply
+through synonym and relationship blocks. This typically occurs when a
+particular class does not participate as a subclass or superclass of
+another class in the current ontology. Overall, one of our goals is
+for \owlifier\ to allow minimal information to be entered by limitting
+redundancy as well as by supporting a range of inferences, while at
+the same time reducing the number of errors commonly made in ontology
+development by non-experts.
 
-5. we allow sparse blocks, and blocks can be given in any order
-
-\section{Implementation of \Owlifier}
+\section{The \Owlifier Application}
 \label{sec:implementation}
 
-figure showing basics owlifier interaction, description.
+\figref{fig:owlifier} shows the general architecture of the \owlifier\
+application. A scientist or researcher first creates a spreadsheet
+containing a set of ontology definitions using \owlifier\ conventions.
+The spreadsheet is then exported as a plain-text file containing the
+\owlifier\ table, e.g., using a CSV or tab-delimited format. The
+\owlifier\ table is then sent to the \owlifier\ application, which
+performs a number of steps that include: (i) parsing the file; (ii)
+generating the corresponding OWL-DL representation (e.g., in the
+current implementation, the Jena API is used \cite{carroll04:_jena});
+(iii) validating the ontology and ensuring it is consistent via an
+OWL-DL reasoner (e.g., the current implementation uses Pellet
+\cite{sirin07:_pellet}); and (iv) outputing the corresponding OWL-DL
+file. The resulting OWL file can then be refined and extended via
+existing OWL tools (e.g., Protege or SWOOP). It is possible that the
+refined ontology is converted back to a corresponding \owlifier\
+representation (shown using dashed lines in
+\figref{fig:owlifier}). While simple to perform, this last conversion
+is not information preserving since not all OWL-DL constructs are
+represented through \owlifier\ blocks.
 
-current status of implementation (subset of the syntax described here), perform classification via pellet/jena, performs disjoint check
+The current implementation of the \owlifier\ application is written in
+Java and supports a subset of the blocks defined in
+\secref{sec:owlifier}. In particular, we are currently extending
+\owlifier\ to support cardinality and sufficient blocks as well as to
+perform the above conversion from OWL-DL files to corresponding
+\owlifier\ blocks. As noted above, the current implementation,
+although it does not support all of the blocks defined here, is being
+used within a project focused on supporting trait data and many of the
+extensions outlined here are based on these experiences. In addition,
+\owlifier\ has been used to construct term hierarchies from large sets
+of keywords (harvested from existing metadata documents) as well as to
+define articulations into the observation ontology defined in
+\cite{madin07:_ontol_for_descr_and_synth}.
 
-ongoing extensions: additional syntax; specify delimeter characters;
-perform inverse translations (easy to perform, e.g., ...); perform
-closure computation (again, easy).
+\begin{figure}
+  \centering
+  \includegraphics[scale=.5]{architecture.pdf}
+  \caption{The basic \owlifier\ application and relation to other
+    technologies.}
+  \label{fig:owlifier}
+\end{figure}
 
@@
 
-related work: similar to dave's work on ``global taxonomic
-constraints'', which assume certain structures of hierarchies, e.g.,
-disjoint siblings. protege approach/plugin
+We describe a new To address barriers for ontology development and adoption within
+ecology, we have presented a new approach that allows the \owlifier,
+which consists of a set orelated work: similar to dave's work on
+``global taxonomic constraints'', which assume certain structures of
+hierarchies, e.g., disjoint siblings. protege approach/plugin
 

### 2008-12-03 docs/pubs/ISEI6/main.tex

@@
 by default (unless specified otherwise) and supporting sufficient
-restrictions \cite{rector04:_owl_pizzas}. In
-\secref{sec:features} we describe additional characteristics of
-\owlifier\ and discuss issues with respect to classification and
-reasoning. In \secref{sec:implementation} we briefly describe the
-\owlifier\ implementation, and conlcude in \secref{sec:conclusion}
-with related and future work. In general, the goal of \owlifier\ is
-not to support all constructs in OWL-DL, but instead to provide a
-higher-level ontology syntax (via spreadsheet blocks) that is easy for
-ecologists to use and understand while also providing the necessary
-constructs for developing typical ecological ontologies. By compiling
-\owlifier\ to OWL-DL, we also allow for experts to refine and extend
-the ontology using more advanced ontology editing tools if necessary.
+restrictions \cite{rector04:_owl_pizzas}. In \secref{sec:features} we
+describe additional characteristics of \owlifier\ and discuss issues
+with respect to classification and reasoning. In
+\secref{sec:implementation} we briefly describe the \owlifier\
+implementation, and conclude in \secref{sec:conclusion} with related
+and future work. In general, the goal of \owlifier\ is not to support
+all constructs in OWL-DL, but instead to provide a higher-level
+ontology syntax (via spreadsheet blocks) that is easy for ecologists
+to use and understand while also providing the necessary constructs
+for developing typical ecological ontologies. By compiling \owlifier\
+to OWL-DL, we also allow for experts to refine and extend the ontology
+using more advanced ontology editing tools if necessary
+(cf.~\figref{fig:owlifier}).
 
@@
 (e.g., we use ``relation'' below instead of ``object property''),
-allows \owlifier\ to easilty generate ontologies that extend the
+allows \owlifier\ to easily generate ontologies that extend the
 observational model of
@@
   ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
-    \textsf{marine:DeapSeaEcosystem}
+    \textsf{marine:DeepSeaEcosystem}
 \end{tabbing}
-This block states that the Deap Sea Ecosystem class of the marine
+This block states that the Deep Sea Ecosystem class of the marine
 ontology is a subclass of the Marine Ecosystem class of the SWEET
@@
 associated description logic axioms by using equivalence ($\equiv$) in
-place of sublass ($\sqsubseteq$). For example the block
+place of subclass ($\sqsubseteq$). For example the block
 \begin{tabbing}
@@
 relationship blocks within a sufficient block to be extended with
-\textsf{not} to state that absense of the property is a defining
+\textsf{not} to state that absence of the property is a defining
 characteristic of the class. For example the following blocks
@@
 all of the synonyms of the class. Similarly, from an axiom $A \equiv
-\exists P. C$ generated from a sufficent block, and an axiom $B
+\exists P. C$ generated from a sufficient block, and an axiom $B
 \sqsubseteq \exists P.C$ generated from a relationship block, the axiom
@@
 occur within OWL-DL ontologies when domain and range axioms are
-providedq (as well as, e.g., property closure axioms).  We explicitly
+provided (as well as, e.g., property closure axioms).  We explicitly
 do not consider these constraints in the current version of \owlifier\
@@
 explicitly given, which in turn can simplify the overall layout of the
-ontology within a spreadsheet.  Additionaly, \owlifier\ does not place
+ontology within a spreadsheet.  Additionally, \owlifier\ does not place
 constraints on the order of blocks within a spreadsheet. Classes also
@@
 another class in the current ontology. Overall, one of our goals is
-for \owlifier\ to allow minimal information to be entered by limitting
+for \owlifier\ to allow minimal information to be entered by limiting
 redundancy as well as by supporting a range of inferences, while at
@@
 OWL-DL reasoner (e.g., the current implementation uses Pellet
-\cite{sirin07:_pellet}); and (iv) outputing the corresponding OWL-DL
+\cite{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
 file. The resulting OWL file can then be refined and extended via
@@
 
-We describe a new To address barriers for ontology development and adoption within
-ecology, we have presented a new approach that allows the \owlifier,
-which consists of a set orelated work: similar to dave's work on
-``global taxonomic constraints'', which assume certain structures of
-hierarchies, e.g., disjoint siblings. protege approach/plugin
+In this paper, we have presented a new approach for developing
+ontologies to address barriers in ontology development and adoption by
+ecologists. Our approach allows scientists to use familiar spreadsheet
+software (frequently used, e.g., by ecologists for storing and
+analyzing data) to create ontologies by filling in a set of templates,
+or blocks, that generate OWL-DL class hierarchies, properties, and
+constraints via the \owlifier\ application.  This approach can provide
+a more intuitive and accessible ontology editing environment for
+ecologists, especially compared to existing OWL-based tools that
+require considerable expertise in the underlying logical formalisms.
 
-future work: the ongoing extensions described above, support seamless
-conversion into and out of owlifier, e.g., if someone extends owlifier
-ontology, re-apply extensions after new owlifier changes. Also,
-support translation to oboe. Additional testing and evaulation.
+Protege provides a variety of ontology editing plug-ins, including a
+simple interface for text-based editing of class hierarchies. In
+\cite{kola07:_impor_spread_data_into_proteg}, an approach is described
+for importing spreadsheet-based ontology descriptions into
+Protege. However, this approach aims at supporting \emph{ad-hoc}
+spreadsheet structures and then provides an intermediate step and
+interface for mapping these structures into ontology axioms. This
+approach is similar to others (e.g.,
+\cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
+for defining mappings from relational data to RDF and OWL ontology
+class and instance data. To the best of our knowledge, however, our
+approach is the first to consider a detailed set of templates that can
+support a large subset of existing OWL constructs. In addition, we
+define a number of shortcuts for creating \owlifier\ tables, including
+sparse blocks, as well as default semantics (e.g., disjoint classes,
+which is a form of global taxonomic constraint defined in
+\cite{thau07:_reason_about_taxon_in_first_order_logi}) that further
+simplify ontology creation for end users.
+
+As future work, we intend to extend our current \owlifier\
+implementation to support the full set of blocks defined above. We
+also intend to explore approaches to fully support conversions between
+\owlifier\ tables and OWL-DL ontologies. In particular, we want to
+extend \owlifier\ to re-apply changes made by knowledge representation
+experts that previously modified the OWL-DL ontology generated by
+\owlifier. For instance, if an OWL-DL ontology generated from an
+\owlifier\ table is refined and extended by a knowledge representation
+expert, then converted back into an \owlifier\ ontology that is
+further edited and extended by a scientist, and then converted again
+into an OWL-DL ontology, we want to maintain (i.e., re-apply) the
+original extensions and edits created by the knowledge representation
+expert that are still relevant. We are also developing extensions to
+\owlifier\ to support conversion into the observation ontology
+framework presented in \cite{bowers08:_concep_model_framew_for_expres}
+and intend to perform additional testing and evaluation of the
+\owlifier\ approach with ecologists.
 

### 2008-12-03 docs/pubs/ISEI6/main.tex

@@
 biologists studying trait data, we found that \owlifier\ enabled them
-to quickly and easily comprehend and construct useful ontologies.
+to quickly and easily comprehend and construct relatively complex and
+meaningful ontologies.
 
@@
 ontology creation using OWL-DL, e.g., by assuming classes are disjoint
-by default (unless specified otherwise) and supporting sufficient
-restrictions \cite{rector04:_owl_pizzas}. In \secref{sec:features} we
-describe additional characteristics of \owlifier\ and discuss issues
-with respect to classification and reasoning. In
+by default (unless specified otherwise)
+\cite{rector04:_owl_pizzas}. In \secref{sec:features} we describe
+additional characteristics of \owlifier\ and discuss issues with
+respect to classification and reasoning. In
 \secref{sec:implementation} we briefly describe the \owlifier\
@@
 relationship blocks where the object property is asserted to be
-transitive. If a property $p$ is declared to be transitive, whenever
-$p$ relates an individual $x$ to an individual $y$, and an
-individual $y$ to an individual $z$, then $p$ is also assumed to
-relate $x$ to $z$. Rows containing transitive blocks take the form
+transitive. That is, if a property $p$ is declared to be transitive,
+and $p$ relates an individual $x$ to an individual $y$ and $y$ to an
+individual $z$, then $p$ is assumed to also relate $x$ to $z$. Rows
+containing transitive blocks take the form
 \begin{itemize}
@@
 where $p$ is an object property and each $c$ is a class.  The
-following block is a simple example of a transitive relationship.
+following block is a simple example of a transitive relationship
+definition.
 \begin{tabbing}
@@
 relationship blocks.  We consider three types of cardinality blocks
-for defining minimum, maximum, and exact cardinality restrictions. A
-minimum block states the smallest number of properties $p$ to distinct
-individuals that an individual of a class may have.  Rows containing
-minimum blocks take the form
+for defining minimum, maximum, and exact cardinality restrictions.
+Rows containing minimum blocks take the form
 \begin{itemize}
@@
 
-A maximum block states the largest number of properties $p$ to
-distinct individuals that an individual of a class may have.  Rows
-containing maximum blocks take the form
+Rows containing maximum blocks take the form
 \begin{itemize}
@@
 
-An exact block ensures both a minimum and maximum number $m$ of
-properties $p$ to distinct individuals that an individual of a class
-must have. Rows containing exact blocks take the form
+Rows containing exact blocks take the form
 \begin{itemize}
@@
 relationship blocks within a sufficient block to be extended with
-\textsf{not} to state that absence of the property is a defining
+\textsf{not} to state that the absence of the property is a defining
 characteristic of the class. For example the following blocks
@@
 Eutheria. Note also that multiple sufficient blocks for a particular
-class result in a single axiom in which each constraint is combined
-via conjunction ($\sqcap$).
+class result in a single axiom in which each constraint (i.e., class
+constructor) is combined via intersection ($\sqcap$).
 
-\myblock{Comment Blocks.} Comment blocks provide a mechanisms to add
+\myblock{Comment Blocks.} Comment blocks provide a mechanism to add
 plain-text comments to \owlifier\ tables. A description block attaches
@@
 where the string $s$ is associated as a comment to the class $c$ or
-property $p$. A note block allows attaches comments to the current
+property $p$. A note block attaches comments to the current
 ontology. Rows containing note blocks take the form
@@
 
-In this section, we briefly describe some of the additional features
-of \owlifier, specifically focusing on the use of disjoint classes,
-\owlifier\ reasoning, and additional block syntax.
+Here we briefly describe some of the additional features of \owlifier,
+specifically focusing on the use of disjoint classes, \owlifier\
+reasoning, and additional block syntax.
 
@@
 example is in the creation of disjoint classes.  In particular, unless
-explicitly stated, distinct classes within an OWL-DL ontology are
-never assumed to be disjoint. However, in many ontologies a large
-number of classes are typically defined as being disjoint (e.g.,
-sibling classes), and stating these disjoint constraints is often
-time-consuming since each pair of classes must be given an explicit
+explicitly asserted, distinct classes within an OWL-DL ontology are
+not assumed to be disjoint. However, in many ontologies a large number
+of classes are typically defined as being disjoint (e.g., sibling
+classes), and stating these disjoint constraints is often
+time consuming since each pair of classes must be given an explicit
 disjoint assertion. Editors such as Protege
@@
 because they also often result in modeling errors for inexperienced
-OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the more
-traditional description-logic approaches that do not have explicit
-domain and range axioms.  Although not currently supported in
-\owlifier, domain and range constraints could be inferred from given
-relationship blocks as well as property closure axioms.
+OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the
+approach of more traditional description logics, which do not have
+explicit domain and range axioms.  Although not currently supported in
+\owlifier, domain and range constraints as well as property closure
+axioms can be inferred from given relationship blocks.
 
@@
 In general, the use of sparse blocks simplifies ontology creation by
-not requiring users to enter every (redundant, leading) field to be
-explicitly given, which in turn can simplify the overall layout of the
-ontology within a spreadsheet.  Additionally, \owlifier\ does not place
-constraints on the order of blocks within a spreadsheet. Classes also
-do not need to explicitly be defined within an entity block, e.g.,
-classes without corresponding entity blocks can be introduced simply
-through synonym and relationship blocks. This typically occurs when a
-particular class does not participate as a subclass or superclass of
-another class in the current ontology. Overall, one of our goals is
-for \owlifier\ to allow minimal information to be entered by limiting
-redundancy as well as by supporting a range of inferences, while at
-the same time reducing the number of errors commonly made in ontology
-development by non-experts.
+not requiring users to enter every redundant field explicitly, which
+in turn can simplify the overall layout of the ontology within a
+spreadsheet.  Additionally, \owlifier\ does not place constraints on
+the order of blocks within a spreadsheet. Classes also do not need to
+be explicitly defined within an entity block, e.g., classes without
+corresponding entity blocks can be introduced simply through synonym
+and relationship blocks. This typically occurs when a particular class
+does not participate as a subclass or superclass of another class in
+the current ontology. Taken together, these appraoches allow users to
+enter minimal information by limiting redundancy and by supporting a
+range of inferences, while at the same time reducing the causes of
+many errors commonly made in ontology development by non-experts.
 
-\section{The \Owlifier Application}
+\section{The \Owlifier\ Implementation}
 \label{sec:implementation}
@@
 representation (shown using dashed lines in
-\figref{fig:owlifier}). While simple to perform, this last conversion
-is not information preserving since not all OWL-DL constructs are
-represented through \owlifier\ blocks.
+\figref{fig:owlifier}). While straightforward to perform, this last
+conversion is not information preserving since not all OWL-DL
+constructs are represented through \owlifier\ blocks.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=.5]{architecture.pdf}
+  \caption{The basic \owlifier\ application and relation to other
+    technologies.}
+  \label{fig:owlifier}
+\end{figure}
 
@@
 of keywords (harvested from existing metadata documents) as well as to
-define articulations into the observation ontology defined in
-\cite{madin07:_ontol_for_descr_and_synth}.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.5]{architecture.pdf}
-  \caption{The basic \owlifier\ application and relation to other
-    technologies.}
-  \label{fig:owlifier}
-\end{figure}
+define articulations between existing ontologies and the observation
+ontology defined in \cite{madin07:_ontol_for_descr_and_synth}.
 
@@
 
-In this paper, we have presented a new approach for developing
-ontologies to address barriers in ontology development and adoption by
+This paper presents a new approach for developing ontologies to
+address barriers in ontology development and adoption by
 ecologists. Our approach allows scientists to use familiar spreadsheet
-software (frequently used, e.g., by ecologists for storing and
+software (e.g., frequently used by ecologists for storing and
 analyzing data) to create ontologies by filling in a set of templates,
@@
 ecologists, especially compared to existing OWL-based tools that
-require considerable expertise in the underlying logical formalisms.
+require considerable expertise in the underlying logic formalisms.
 
@@
 Protege. However, this approach aims at supporting \emph{ad-hoc}
-spreadsheet structures and then provides an intermediate step and
-interface for mapping these structures into ontology axioms. This
-approach is similar to others (e.g.,
+spreadsheet structures by providing an intermediate interface for
+mapping these structures into ontology axioms. This approach is
+similar to others (e.g.,
 \cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
@@
 define a number of shortcuts for creating \owlifier\ tables, including
-sparse blocks, as well as default semantics (e.g., disjoint classes,
-which is a form of global taxonomic constraint defined in
+sparse blocks and default semantics (e.g., disjoint classes, which is
+a form of global taxonomic constraint defined in
 \cite{thau07:_reason_about_taxon_in_first_order_logi}) that further

### 2009-08-27 docs/pubs/ISEI6/response.txt
start of responses and edits

Mark's Revisions:

On p. 7, the "sufficient blocks" are not given a template
representation, whereas all the other blocks are...

Also on p. 7, you describe how the relationship block can be negated
in a sufficient block.  Is negation also allowed in the relationship
block itself?  This would be nice, and if permissible, should be
described in the relationship block section.

==> add to possible future extensions

Section 3.2 on reasoning is where some additional explanation could be
provided about difficulties of adding domain/range constraints to an
"advanced owlifier" .  Also, does the relationship block
simultaneously imply both someValuesFrom AND allValuesFrom (in
Manchester syntax)?  If not, it would be nice to provide a closure
axiom as an option on the relationship block.  Is this low-hanging
fruit?

==> add to possible future extensions

In second paragraph of conclusion, last sentence, I am not sure why
you refer to "disjoint classes" as a form of global taxonomic
constraint defined in [22]-- I would think it is more the converse
that global taxonomic constraints often require disjoint classes.

==> add to possible future extensions

It would be nice to clearly say what currently DOES or DOES NOT YET
exist in owlifier, either in the Conclusion, or the preceding
paragraph.

==> add to implementation part, along with a url to the OWLifier webpage

----------------------------------------------------------------------
Reviewer #1: 
----------------------------------------------------------------------
1). "Not mentioned is the list of OWL features not supported."

- Probably add a list to the implementation?

[josh]: 
Features not supported:
  - Data-type properties
  - Property domains and ranges
  - Symmetric properties
  - Others?
  I would say at page 2, third full paragraph, after first sentence:
  Primarily, we omit much of the complexities associated with
  properties (e.g., domains and ranges, data-type properties) that are
  often redundant and should be implemented where necessary, by more
  experienced knowledge modellers.

[mark]: I'd amend Josh's sentence slightly, and place it as the last
sentence of the preceding paragraph (para. 2; before paragraph
starting with "While not as expressive") rather than second sentence
of following para:

"We omit several OWL features associated with properties (e.g.,
domains and ranges, data-type properties) as these can be confusing to
construct, and more suitable for experienced knowledge modelers."

----------------------------------------------------------------------
2). "Overriding the OWL open world assumption should be fine, so long as
that assumption is made clear to the users int eh documentation."

We primarily provide a different approach for specifying disjoint
classes, as opposed to overriding the OWA in general. However, we
agree that describing this feature will be particularly important to
users of OWLifier that are familiar with OWL-DL.

[josh]:
- This is currently made clear in section 3.1 Disjoint Class Inference
- Maybe there should be a brief mention of this in the introduction?

[mark]: I think it is fine as is. Hard to be brief about this without
being confusing.

----------------------------------------------------------------------
3). "There is no mention of the competing CMAP-COE tool that makes it
easy to create ontologies visually, with help from drop-down lists.
It would be helpful to include a line that compares your approach to
CMAP (which also supports only a subset of OWL)."

We added additional text in the introduction that further describes
our motivation for the spreadsheet-based approach and some of its
advantages for ecologists/sciencts over CMAP-COE.

[josh]:

- Even installing and using this tool is too complicated and a pain
for people. Introduction, third paragraph, second sentence, maybe
change to something like:

Our approach is to allow scientists to use common and familiar
spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
Office Spreadsheet) to describe, in an intuitive way, different
aspects of an ontology, and then to take these descriptions and
convert them into full-fledged OWL ontologies using a software
application called owlifier.  The use of a familiar editing
environment will improve the speed and understanding of ontology
construction and avoid locating, downloading, and interpreting
additional software packages and add-ons (e.g., CMAP-COE).  Many
spread-sheet applications also contain easy-to-use features that
support scientists with the structuring process (e.g., cut / copy /
paste, search / replace, sort, split-window, annotation,
track-changes, freeze panes, and so on).

[mark]: Above sounds good to me.

----------------------------------------------------------------------
4). "Something that might be pointed out - the tool could be used to
help enforce some community standards (e.g., a common use of a
"part-of" relation)."

[josh]:
- It would be nice to say something like (although I not sure where it
  would be placed):

Furthermore, blocks can be reused (e.g., relationship blocks) to
ensure that future ontology development adheres to any community
agreed-upon representations, and will result in the smoother ontology
integration within ecology.

[mark]: Nice

[shawn]: Because owlifier spreadsheets can be converted to owl
ontologies, these can be further imported in owlifier spreadsheets
(e.g., for extension).  This follows the basic OWL infrastructure.

----------------------------------------------------------------------
5). "In the references: Usually, the word "pages" is abbreviated to
pp. Ref 18: should be to a journal article, not a web site - perhaps
Raskin and Pan (2005) in Computers & Geosciences. Ref 19: to EKAW?
What is that?  Compare Refs 4 and 7 - different wordings are used to
describe the same conference."

FIXED

----------------------------------------------------------------------
Reviewer #2:
----------------------------------------------------------------------
6). "In the early stages of the paper there are some assertions
without reference about what is and is not difficult. See p.2 end of
2nd, 3rd and 4th paragraphs.  Can a little more detail be added to
support the statements? If not, then noting that these are
suppositions would be useful for the reader."

[josh]: Honestly, I don't find a problem here.  I think paragraph 2 is
fine and obvious to anyone who's used Protege (we could potential say
"based on our experience in an ecological informatics project
constructing ontologies [SEEK]").  I can't find anything not
referenced in paragraph 3.  And the example used by the reviewer in
paragraph 4 IS referenced and not so much about the level of logical
description, and basically says "you can capture the level of
description shown in this cited paper as necessary for improved
discovery and integration".

[mark]: I think the assertion about "lack of suitable ontology editing
tools is a major barrier for widescale adoption in ecology" is the
issue, and begs the question as to why this isn't the case in other
domains.  So, I suggest we delete the last sentence of paragraph 2 on
pg 2, and add a following paragraph that elaborates and clarifies that
point:

"A well-funded ontology development effort, Gene Ontology (GO) has
been underway for over a decade within the molecular biology
community, and has led to some major successes enriching genomic and
biomedical data resources with annotations to a community-based
knowledge representation. A major lesson learned from these efforts
was the importance of keeping the ontology focused, and indeed GO has
a highly delimited scope (Bada et al.)  Ecology, on the other hand,
encompasses an extremely broad range of scales and disciplines, such
that a coordinated ontology development effort would need to involve a
large number of experts to cover the diverse types of information
relevant to ecological investigations.  At this point, we believe that
ecological researchers need to begin constructing knowledge
representations for their own specialized subdisciplines.  Most
ontology editing tools, however, typically require substantial
expertise in formal knowled ge representation, and present a major
barrier for more wide-scale creation and adoption of ontologies in
ecology."

[Bada, Michael, Robert Stevens, Carol Goble, A. Gil, Michael
Ashburner, Judith A. Blake, J. Michael Cherry, Midori Harris, Suzanna
Lewis.  2004. A short study on the success of the gene ontology.
Journal of Web Semantics, vol 1: 235-240.]

----------------------------------------------------------------------
7). "Was there any quantifiable basis for choosing the predicates?  In
particular, the statement" "While not as expressive as OWL-DL, our
approach can produce ontology structures that are essential for
improved data discovery and integration [14]."  indicates some level
of decision in the reduced description but without any details,
especially in terms for DL-expressivity, it is SHIQ, SHOIN, or what?"

As far as I am aware, the arrangement of DL constructs used in
OWLifier does not directly map to any previously named DL sublanguage,
however, it does provides only OWL-DL based constructs, and thus can
be considered best as a sublanguage of SHOIN(D). Note that although
many of the OWL-DL constructs can be used, they can only be used in
certain contexts. Also, some of the more foundational constructs of
DLs are not considered, e.g., value restrictions and concept
disjunction. While additional blocks can easily be added to OWLifier
spreadsheet conventions, the proposed set of blocks were chosen to
provide a simplified syntax (for a novice DL user) while leading to a
"natural" or "intuitive" semantics of blocks for these users.

----------------------------------------------------------------------
8). "There are no mention of instances, that I could find, at all of
instances?  Are these part of the use case or not? If not, this really
needs to be stated.  An ontology without instances has limited
application."

[josh]: Maybe at the start of third paragraph of introduction: This
paper presents a novel approach for ontology creation that aims at
being more intuitive for ecologists that can be used to rapidly
construct large ontologies of ecological terms (classes) and the
subsequent annotation of scientific data (instances).

----------------------------------------------------------------------
9). "Fig.1 suggests a round-trip approach to knowledge engineering,
but in fact it is not. How does the present formulation represent to
scientiest what gets dropped in the back conversion from OWL to the
spreadsheet format?"

[josh]: I guess we need be a bit more transparent in the last sentence
of the bottom paragraph, page 10; something like:
   While straightforward to perform, this last conversion will only
preserve OWL-DL constructs that are represented by owlifier blocks.

[mark]: Substitute term "back conversion" instead of "last
conversion", and "that are" to "that can be"

Note misspelling of "appraoches" on p. 10.

----------------------------------------------------------------------
10). "Most of all, there is almost no (or none at all) description of
the implemenation or any evaluation, i.e. the how, why, etc.  What has
this application been applied to?"

[josh]: We could mention that current efforts are underway using
owlifier to structure list of keywords from the US LTER program.  And
that the spread-sheet approach works well for these kind of mid-level
term sorting exercises (e.g., 50-5000) terms.  Nonetheless, (from
experience) larger numbers of terms does get difficult to navigate and
structure, although the use of blocks means that you can focus on
tractable chunks of terms and properties.  Furthermore, the use of
spread-sheet software features greatly improve navigation, commenting,
and editing capabilities (e.g., sort, freeze panes, split window,
search, multiple worksheets / windows, comments, track-changes).

----------------------------------------------------------------------
11). "It seems that it would be easy to generate complex blocks. How
are moderate size ontologies developed?"

----------------------------------------------------------------------
12). "What is the rationale for the proposed extensions later in the
paper, i.e. what are the use cases or evaluation-driven itersations?"

----------------------------------------------------------------------
Reviewer #3:
----------------------------------------------------------------------

----------------------------------------------------------------------
13). "Possibly the authors could signal at the end of section 1
something like "Readers unfamiliar with logic notation may safely
ignore the small amount they may encounter in what follows.""

[josh]: I like the reviewer's idea about a sentence at the end of
section 1 saying that "Readers unfamiliar with logic notation may
safely ignore the small amount they may encounter in what follows" and
focus on the examples accompanying each block description.

[mark]: I agree

----------------------------------------------------------------------
14). "Although not enough to recommend required additions, I am left a
little puzzled as to why the authors seem to be silent on two issues:
OWL data type properties and owlifier instance entities. Both of these
would seem easy to represent in their owlifier syntax. I would also
imagine that both could arise in representing and reasoning about both
dataset metadata and record-level metadata on biodiversity
databases. So I am left wondering whether I am wrong about the ease of
representing them in spreadsheets and attaching DL axioms, or whether
instead, the authors believe that these two OWL constructs are better
treated in as data for reasoners or elsewhere."

[josh]: I think we might need to be clearer that we are talking about
class / term structuring and not instantiating with data (class
instantiating comes later, after the formal conversion to OWL and by
the knowledge engineer).  Maybe this could be added to the figure in
some way.

I admit data-type properties are an obvious exclusion and should be
justified.  This could be done by simply stating that owlifier is a
mechanism for structuring classes with no additional, complicating
information at this stage (although we do allow for cardinality...).

[mark]: Yes, it is hard to convey the reason for this focus on
Classes.  Well, I feel that our use case is driven by using these
ontologies to annotate data-- values of which become the
instances. Maybe this is not a proper understanding.  But also even
Rector et al. OWL Pizzas paper states: "this paper only concerns
issues in defining OWL classes, since this is the strength of OWL-DL,
and most existing classifiers deal with individuals completely or not
at all." So we are in good company by just advancing capability to
work with classes and relationships.

Would such support for data-type properties be low-hanging fruit for
extending owlifier?  Especially the ability to capture enumerated
lists might be very useful in the future.

----------------------------------------------------------------------
15). "On page 4 (Subsection "Overlap blocks", the second sentence has a
small English ambiguity: "[...]a given set of classes may have
overlapping instances." could reasonably be understood to mean that
some instances overlap one another, whereas I believe the clause means
"several classes may have common instances." [Grammatically, the
overlapping instances do not belong to the collective "set of classes"
but rather to each of the members of that collective]"

[josh]: I agree with the grammatical comment to help the lay
reader. overlapping --> common

[mark]: ow about "Overlap blocks explicitly relax this assumption by
stating that a given set of classes may share common instances" (2nd
sent in para on Overlap blocks)

----------------------------------------------------------------------
16). "Reference 22 is missing 'c' in the article title.  "logi" should
be "logic"."

FIXED

### 2009-08-27 docs/pubs/ISEI6/main.tex
start of reformatting
start of responses and edits

@@
-\documentclass[preprint,number]{elsarticle} 
+\documentclass[5p,authoryear]{elsarticle} 
 % other options: preprint, review, 1p, 3p, 5p, 
@@
   Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}}
-%\tnotetext[t1]{This work supported in part by NSF grants ...}
+\tnotetext[t1]{This work supported in part by NSF grants DBI-0743429
+  and DBI-0753144.}
 
@@
 
-\address[smb]{UC Davis Genome Center}
+\address[smb]{UC Davis Genome Center \& Dept. of Computer Science,
+  Gonzaga University}
 
@@
 construct large ontologies for describing scientific data.  Our
-approach is to allow scientists to use common spreadsheet-based tools
-to describe, in an intuitive way, different aspects of an ontology,
-and then to take these descriptions and convert them into full-fledged
-OWL ontologies using a software application called \owlifier. An
-\owlifier\ spreadsheet consists of a set of \emph{blocks} that have a
-predefined template structure for users to fill in. Each non-empty row
-in an \owlifier\ table constitutes a block. Each block defines
+approach is to allow scientists to use common and familiar
+spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
+Office Spreadsheet) to describe, in an intuitive way, different
+aspects of an ontology, and then to take these descriptions and
+convert them into full-fledged OWL ontologies using a software
+application called \owlifier. The use of a familiar editing
+environment can improve the speed and understanding of ontology
+construction and avoid locating, downloading, and interpreting
+additional software packages and add-ons (e.g., CMAP-COE
+\cite{hayes05:_collab_knowl_captur_in_ontol}).  Many spreadsheet
+applications also contain easy-to-use features that support scientists
+with the structuring process (e.g., cut/copy/paste, search/replace,
+sort, split-window, annotation, track-changes, freeze panes, and so
+on).
+
+An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
+a predefined template structure for users to fill in. Each non-empty
+row in an \owlifier\ table constitutes a block. Each block defines
 different aspects of an ontology including ontology classes,
@@
 from previous blocks), which can further simplify the creation of
-large ontologies.
+large ontologies. We omit several OWL features associated with
+properties (e.g., role inclusion axioms, domain and range property
+restrictions, and data-type properties among others) as these can be
+confusing to construct for non-experts and are more suitable for
+experienced knowledge modelers.
 
@@
 the normal way. As an example, the following block imports the SWEET
-``Earth Realm'' ontology \cite{raskin:_seman_web_for_earth_and}
+``Earth Realm'' ontology \cite{raskin05}
+\begin{small}
 \begin{tabbing}
@@
 \end{tabbing}
+\end{small}
 With this import block the class denoting Marine Ecosystems (a class
@@
 
-\myblock{Comment Blocks.} Comment blocks provide a mechanism to add
-plain-text comments to \owlifier\ tables. A description block attaches
-a plain-text comment to classes and properties. Rows containing
-description blocks take the form
+\myblock{Comment Blocks.} There are two types of comment blocks that
+each provide a mechanism to add plain-text comments to \owlifier\
+tables. A description block attaches a plain-text comment to classes
+and properties. Rows containing description blocks take the form
 \begin{itemize}
@@
 property $p$. A note block attaches comments to the current
-ontology. Rows containing note blocks take the form
+ontology. Rows containing note blocks can occur anywhere within an
+\owlifier\ table, and take the form
 \begin{itemize}
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[scale=.75]{architecture.pdf}
+  \caption{The basic \owlifier\ application and relation to other
+    technologies.}
+  \label{fig:owlifier}
+\end{figure*}
+
 \section{The \Owlifier\ Implementation}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[scale=.5]{architecture.pdf}
-  \caption{The basic \owlifier\ application and relation to other
-    technologies.}
-  \label{fig:owlifier}
-\end{figure}
-
 The current implementation of the \owlifier\ application is written in
@@
 \owlifier\ to support cardinality and sufficient blocks as well as to
-perform the above conversion from OWL-DL files to corresponding
-\owlifier\ blocks. As noted above, the current implementation,
-although it does not support all of the blocks defined here, is being
-used within a project focused on supporting trait data and many of the
-extensions outlined here are based on these experiences. In addition,
-\owlifier\ has been used to construct term hierarchies from large sets
+perform the (potentially lossy) back conversion from OWL-DL files to
+corresponding \owlifier\ blocks. The current implementation of
+\owlifier\ is being used within a project focused on integrating
+vegetation trait data.  As noted above, not all the blocks described
+here are yet supported, however, these priority needs are clearly
+emerging from our interactions with the vegetation scientists.
+Nevertheless, these scientists were able to rapidly prototype and
+revise their trait ontologies using \owlifier, with little to no
+instruction in formal logic or knowledge modeling.  In addition,
+\owlifier\ is being used to construct term hierarchies from large sets
 of keywords (harvested from existing metadata documents) as well as to
@@
 %\bibliographystyle{elsarticle-num}
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 \bibliography{main}

### 2009-08-28 docs/pubs/ISEI6/response.txt
increment

@@
 
+To address this issue we added additional text on the need for new
+tools in ontology development for ecologists (e.g., compared with ...)
+Also, much of the motivation for the work is based on our work with
+ecologists and information managers, which we refer to in the paper.
+
 [josh]: Honestly, I don't find a problem here.  I think paragraph 2 is
@@
 
+Our main goal is to provide a mechanism for describing ecological
+concepts, and then to allow additional tools to use these concepts for
+annotating scientific data (which represents the instances/values). We
+briefly mention this now in the introduction, and also elaborate on
+the scope of owlifier later in the paper.
+
 [josh]: Maybe at the start of third paragraph of introduction: This
@@
 
+We reworded the implementation section to make it clearer that the
+conversion is not information preserving. ... the owlifier tool will
+notify the user of this situation, but at this time no additional
+support is provided.
+
 [josh]: I guess we need be a bit more transparent in the last sentence
@@
 conversion", and "that are" to "that can be"
-
-Note misspelling of "appraoches" on p. 10.
 

### 2009-08-28 docs/pubs/ISEI6/main.tex
increment

@@
 discovery and integration
-\cite{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
+\citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
 benefits can be obtained for ecological data, far fewer efforts exist
 to develop broad and consistent terminologies within ecology
-\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
+\citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
 The use of formal ontologies can significantly enhance metadata
@@
 constraints such as cardinality restrictions on properties and term
-equivalence \cite{madin08:_advan_ecolog_resear_with_ontol}.
+equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.
 
@@
 typically leverage the W3C Web Ontology Language (OWL)
-\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
+\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
 for representing and sharing ontologies. A key advantage of OWL is
@@
 editors
-\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
+\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
 reasoning systems
-\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
+\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
 languages
-\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
+\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
 and storage technologies
-\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
+\citep{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
 however, are primarily targeted at experts in knowledge engineering
 and software development familiar with the underlying description
-logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
+logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true
 with ontology editors (such as Protege, SWOOP, etc.), which allow for
@@
 considerable understanding of the underlying ontology formalisms and
-syntax. We see the lack of suitable ontology editing tools for
-scientists without expertise in knowledge representation as one of the
-major barriers for more wide-scale adoption of ontologies in ecology.
+syntax.  A well-funded ontology development effort, the Gene Ontology
+(GO), has been underway for over a decade within the molecular biology
+community, and has led to some major successes enriching genomic and
+biomedical data resources with annotations to community-based
+ontologies. A major lesson learned from these efforts was the
+importance of keeping the ontology focused, and indeed GO has a highly
+delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}
+Ecology, on the other hand, encompasses an extremely broad range of
+scales and disciplines, such that a coordinated ontology development
+effort would need to involve a large number of experts to cover the
+diverse types of information relevant to ecological investigations.
+At this point, we believe that ecological researchers need to begin
+constructing ontologies for their own specialized subdisciplines.
+Most ontology editing tools, however, typically require substantial
+expertise in formal knowledge representation, and present a major
+barrier for more wide-scale creation and adoption of ontologies in
+ecology.
 
@@
 at being more intuitive for ecologists and that can be used to rapidly
-construct large ontologies for describing scientific data.  Our
-approach is to allow scientists to use common and familiar
-spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
-Office Spreadsheet) to describe, in an intuitive way, different
-aspects of an ontology, and then to take these descriptions and
-convert them into full-fledged OWL ontologies using a software
-application called \owlifier. The use of a familiar editing
-environment can improve the speed and understanding of ontology
-construction and avoid locating, downloading, and interpreting
-additional software packages and add-ons (e.g., CMAP-COE
-\cite{hayes05:_collab_knowl_captur_in_ontol}).  Many spreadsheet
+construct large ontologies of ecological terms (classes) and the
+subsequent annotation of scientific data (instances).  Our approach is
+to allow scientists to use common and familiar spreadsheet-based tools
+(e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to
+describe, in an intuitive way, different aspects of an ontology, and
+then to take these descriptions and convert them into full-fledged OWL
+ontologies using a software application called \owlifier. The use of a
+\emph{familiar} editing environment can improve the speed and
+understanding of ontology construction and can help users avoid the
+often time-consuming task of locating, downloading, installing, and
+learning fundamentally new software applications and interfaces (e.g.,
+CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
+\citep{knublauch04:_editin_descr_logic_ontol_with}).  Many spreadsheet
 applications also contain easy-to-use features that support scientists
-with the structuring process (e.g., cut/copy/paste, search/replace,
-sort, split-window, annotation, track-changes, freeze panes, and so
-on).
+in the process of editing and managing \owlifier\ ontology documents
+(e.g., via cut/copy/paste, search/replace, sort, split-window,
+track-changes, freeze panes, and so on).
 
@@
 structures that are essential for improved data discovery and
-integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
+integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
 important, because spreadsheet tools are frequently used by ecologists
@@
 that also generally follow the ontology creation guidelines defined in
-\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
+\citep{rector04:_owl_pizzas}. We also simplify certain aspects of
 ontology creation using OWL-DL, e.g., by assuming classes are disjoint
 by default (unless specified otherwise)
-\cite{rector04:_owl_pizzas}. In \secref{sec:features} we describe
+\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
 additional characteristics of \owlifier\ and discuss issues with
@@
 As described above, an \owlifier\ table defines an OWL-DL
-\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
+\citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of
 \emph{blocks} representing one or more ontology definitions.  Each
@@
 terms used in
-\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
+\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
 as opposed to the names used for corresponding constructs in
@@
 observational model of
-\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
+\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
 and avoids confusion with established terms commonly used within
@@
 the normal way. As an example, the following block imports the SWEET
-``Earth Realm'' ontology \cite{raskin05}
+``Earth Realm'' ontology \citep{raskin05}
 \begin{small}
@@
 % restrictions stated by value blocks are often used for defining
-% so-called \emph{value partitions} \cite{co-ode}.
+% so-called \emph{value partitions} \citep{co-ode}.
 
@@
 of ontology development ``pitfalls'' for those new to the language
-\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
+\citep{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
 example is in the creation of disjoint classes.  In particular, unless
@@
 disjoint assertion. Editors such as Protege
-\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
+\citep{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
 via the user interface to create specific sets of disjoint assertions,
@@
 as disjoint. In general, however, many users expect such classes to be
-disjoint by default \cite{rector04:_owl_pizzas} and this expectation
+disjoint by default \citep{rector04:_owl_pizzas} and this expectation
 often leads to modeling errors.
@@
 \owlifier\ in the resulting ontology as being disjoint. As described
-in \cite{rector04:_owl_pizzas}, undeclared disjoint classes are a
+in \citep{rector04:_owl_pizzas}, undeclared disjoint classes are a
 common problem in ontology development using OWL-DL and often limit
@@
 
-As described in \cite{rector04:_owl_pizzas} additional reasoning can
+As described in \citep{rector04:_owl_pizzas} additional reasoning can
 occur within OWL-DL ontologies when domain and range axioms are
@@
 because they also often result in modeling errors for inexperienced
-OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the
+OWL-DL users \citep{rector04:_owl_pizzas}. Instead, we adopt the
 approach of more traditional description logics, which do not have
@@
 does not participate as a subclass or superclass of another class in
-the current ontology. Taken together, these appraoches allow users to
+the current ontology. Taken together, these approaches allow users to
 enter minimal information by limiting redundancy and by supporting a
@@
 generating the corresponding OWL-DL representation (e.g., in the
-current implementation, the Jena API is used \cite{carroll04:_jena});
+current implementation, the Jena API is used \citep{carroll04:_jena});
 (iii) validating the ontology and ensuring it is consistent via an
 OWL-DL reasoner (e.g., the current implementation uses Pellet
-\cite{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
+\citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
 file. The resulting OWL file can then be refined and extended via
@@
 representation (shown using dashed lines in
-\figref{fig:owlifier}). While straightforward to perform, this last
+\figref{fig:owlifier}). While straightforward to perform, this back
 conversion is not information preserving since not all OWL-DL
-constructs are represented through \owlifier\ blocks.
+statements can be represented through \owlifier\ blocks.
 
@@
 define articulations between existing ontologies and the observation
-ontology defined in \cite{madin07:_ontol_for_descr_and_synth}.
+ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.
 
@@
 simple interface for text-based editing of class hierarchies. In
-\cite{kola07:_impor_spread_data_into_proteg}, an approach is described
+\citep{kola07:_impor_spread_data_into_proteg}, an approach is described
 for importing spreadsheet-based ontology descriptions into
@@
 similar to others (e.g.,
-\cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
+\citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
 for defining mappings from relational data to RDF and OWL ontology
@@
 a form of global taxonomic constraint defined in
-\cite{thau07:_reason_about_taxon_in_first_order_logi}) that further
+\citep{thau07:_reason_about_taxon_in_first_order_logi}) that further
 simplify ontology creation for end users.
@@
 \owlifier\ to support conversion into the observation ontology
-framework presented in \cite{bowers08:_concep_model_framew_for_expres}
+framework presented in \citep{bowers08:_concep_model_framew_for_expres}
 and intend to perform additional testing and evaluation of the

### 2009-08-29 docs/pubs/ISEI6/response.txt
almost done...

@@
-Mark's Revisions:
+We thank the reviewers for their excellent comments and suggestions for improvements to the paper.
+We revised the paper to address each of the suggestions.
+Below we describe the changes we made that were requested by the reviewers as well as answers to the questions of the reviewers.
 
-On p. 7, the "sufficient blocks" are not given a template
-representation, whereas all the other blocks are...
+---------------------------------------------------------------------- Reviewer #1: ---------------------------------------------------------------------- 1).
+"Not mentioned is the list of OWL features not supported."
 
-Also on p. 7, you describe how the relationship block can be negated
-in a sufficient block.  Is negation also allowed in the relationship
-block itself?  This would be nice, and if permissible, should be
-described in the relationship block section.
+We added a sentence to the implementation section clarifying this, and also a brief description of what is not supported to the introduction (along with some motivation for why we consider only a subset of constructs).
 
-==> add to possible future extensions
+---------------------------------------------------------------------- 2).
+"Overriding the OWL open world assumption should be fine, so long as that assumption is made clear to the users in the documentation."
 
-Section 3.2 on reasoning is where some additional explanation could be
-provided about difficulties of adding domain/range constraints to an
-"advanced owlifier" .  Also, does the relationship block
-simultaneously imply both someValuesFrom AND allValuesFrom (in
-Manchester syntax)?  If not, it would be nice to provide a closure
-axiom as an option on the relationship block.  Is this low-hanging
-fruit?
+We primarily provide a different approach (e.g., than Protege) for specifying disjoint classes, as opposed to a general override of the OWA.
+However, we agree that describing this feature will be particularly important to users of owlifier that are familiar with OWL-DL.
+We also describe this in Section 3.1.
 
-==> add to possible future extensions
+---------------------------------------------------------------------- 3).
+"There is no mention of the competing CMAP-COE tool that makes it easy to create ontologies visually, with help from drop-down lists.
+It would be helpful to include a line that compares your approach to CMAP (which also supports only a subset of OWL)."
 
-In second paragraph of conclusion, last sentence, I am not sure why
-you refer to "disjoint classes" as a form of global taxonomic
-constraint defined in [22]-- I would think it is more the converse
-that global taxonomic constraints often require disjoint classes.
+We added additional text in the introduction that further describes our motivation for the spreadsheet-based approach and some of its advantages for ecologists/scientists over CMAP-COE.
+This is also now briefly mentioned in the conclusion (which lists additional related work).
 
-==> add to possible future extensions
+---------------------------------------------------------------------- 4).
+"Something that might be pointed out - the tool could be used to help enforce some community standards (e.g., a common use of a "part-of" relation)."
 
-It would be nice to clearly say what currently DOES or DOES NOT YET
-exist in owlifier, either in the Conclusion, or the preceding
-paragraph.
+We added two sentences to the conclusion that highlights this feature of owlifier.
 
-==> add to implementation part, along with a url to the OWLifier webpage
+---------------------------------------------------------------------- 5).
+"In the references: Usually, the word "pages" is abbreviated to pp.
+Ref 18: should be to a journal article, not a web site - perhaps Raskin and Pan (2005) in Computers & Geosciences.
+Ref 19: to EKAW?
+What is that?
+Compare Refs 4 and 7 - different wordings are used to describe the same conference."
 
-----------------------------------------------------------------------
-Reviewer #1: 
-----------------------------------------------------------------------
-1). "Not mentioned is the list of OWL features not supported."
+We fixed each of these.
 
-- Probably add a list to the implementation?
+---------------------------------------------------------------------- Reviewer #2: ---------------------------------------------------------------------- 6).
+"In the early stages of the paper there are some assertions without reference about what is and is not difficult.
+See p.2 end of 2nd, 3rd and 4th paragraphs.
+Can a little more detail be added to support the statements?
+If not, then noting that these are suppositions would be useful for the reader."
 
-[josh]: 
-Features not supported:
-  - Data-type properties
-  - Property domains and ranges
-  - Symmetric properties
-  - Others?
-  I would say at page 2, third full paragraph, after first sentence:
-  Primarily, we omit much of the complexities associated with
-  properties (e.g., domains and ranges, data-type properties) that are
-  often redundant and should be implemented where necessary, by more
-  experienced knowledge modellers.
+To address this issue we added additional text in the introduction on why there is a need for new tools familiar to ecological scientists for ontology development, comparing our approach to CMAP-COE and Protege as well as efforts such as OBO.
+Also, much of the motivation for the work is based on our work with ecologists and information managers, which we refer to in the paper (in the implementation section).
 
-[mark]: I'd amend Josh's sentence slightly, and place it as the last
-sentence of the preceding paragraph (para. 2; before paragraph
-starting with "While not as expressive") rather than second sentence
-of following para:
+---------------------------------------------------------------------- 7).
+"Was there any quantifiable basis for choosing the predicates?
+In particular, the statement" "While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration [14]."  indicates some level of decision in the reduced description but without any details, especially in terms for DL-expressivity, it is SHIQ, SHOIN, or what?"
 
-"We omit several OWL features associated with properties (e.g.,
-domains and ranges, data-type properties) as these can be confusing to
-construct, and more suitable for experienced knowledge modelers."
+As far as we are aware, the arrangement of DL constructs used in owlifier does not directly map to any previously named DL sublanguage, however, it does provide only OWL-DL based constructs, and thus can be considered best as a sublanguage of SHOIN(D).
+Note that although many of the OWL-DL constructs can be used, they can only be used in certain contexts.
+Also, some of the more foundational constructs of DLs are not considered, e.g., value restrictions and concept disjunction.
+While additional blocks can easily be added to owlifier spreadsheet conventions, the proposed set of blocks were chosen to provide a simplified syntax (for a novice DL user) while leading to a "natural" or "intuitive" semantics of blocks for these users.
 
-----------------------------------------------------------------------
-2). "Overriding the OWL open world assumption should be fine, so long as
-that assumption is made clear to the users int eh documentation."
+---------------------------------------------------------------------- 8).
+"There are no mention of instances, that I could find, at all of instances?
+Are these part of the use case or not?
+If not, this really needs to be stated.
+An ontology without instances has limited application."
 
-We primarily provide a different approach for specifying disjoint
-classes, as opposed to overriding the OWA in general. However, we
-agree that describing this feature will be particularly important to
-users of OWLifier that are familiar with OWL-DL.
+Our main goal is to provide a mechanism for describing ecological concepts, and then to allow additional tools to use these concepts for annotating scientific data (which represents the instances/values).
+We briefly mention this now in the introduction, and also elaborate on the scope of owlifier later in the paper.
+Also see our response to (14) below.
 
-[josh]:
-- This is currently made clear in section 3.1 Disjoint Class Inference
-- Maybe there should be a brief mention of this in the introduction?
+---------------------------------------------------------------------- 9).
+"Fig.1 suggests a round-trip approach to knowledge engineering, but in fact it is not.
+How does the present formulation represent to scientist what gets dropped in the back conversion from OWL to the spreadsheet format?"
 
-[mark]: I think it is fine as is. Hard to be brief about this without
-being confusing.
+We reworded the implementation section to make it clearer that the conversion is not information preserving.
+In general, the owlifier tool will only notify the user of this situation, but at this time no additional support is provided.
+We also tried to better clarify and describe the extensions to owlifier needed to better support the round trip conversion in the conclusion (as part of our future work).
 
-----------------------------------------------------------------------
-3). "There is no mention of the competing CMAP-COE tool that makes it
-easy to create ontologies visually, with help from drop-down lists.
-It would be helpful to include a line that compares your approach to
-CMAP (which also supports only a subset of OWL)."
+---------------------------------------------------------------------- 10).
+"Most of all, there is almost no (or none at all) description of the implementation or any evaluation, i.e.
+the how, why, etc.
+What has this application been applied to?"
 
-We added additional text in the introduction that further describes
-our motivation for the spreadsheet-based approach and some of its
-advantages for ecologists/sciencts over CMAP-COE.
+The areas the owlifier application has been applied to is described in the implementation section, as well as in the introduction (briefly).
+In particular, we are applying owlifier to the Traitnet project (ecologists studying/using biological trait data) and to an LTER project that is attempting to impose structure over a large list of keywords extracted from existing metadata documents.
+We have not performed a formal evaluation of the owlifier application, which we state in the paper as future work, however, we have received positive (informal) feedback on the tool and we briefly describe this at the end of the implementation section.
 
-[josh]:
+---------------------------------------------------------------------- 11). "It seems that it would be easy to generate complex blocks. How are moderate size ontologies developed?"
 
-- Even installing and using this tool is too complicated and a pain
-for people. Introduction, third paragraph, second sentence, maybe
-change to something like:
+Our use of owlifier has included both small and moderate sized ontologies.
+Ecologists often create spreadsheets with large amounts of data, and similarly owlifier can be used in this way.
+In addition, sparse blocks significantly simplify the size of large hierarchies, since not all information needs to be repeated per row.
+In the LTER project mentioned above, there are thousands of terms (organized into a flat list) that are easily imported into Excel, and then using the Excel editing features, have been fairly easily organized into hiearchies (i.e., entity blocks and property blocks).
+In general, using a combination of tools (e.g., that could show a portion of the spreadsheet visually) would also probably be extremely useful, however, we have not developed this type of technology for owlifier.
 
-Our approach is to allow scientists to use common and familiar
-spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
-Office Spreadsheet) to describe, in an intuitive way, different
-aspects of an ontology, and then to take these descriptions and
-convert them into full-fledged OWL ontologies using a software
-application called owlifier.  The use of a familiar editing
-environment will improve the speed and understanding of ontology
-construction and avoid locating, downloading, and interpreting
-additional software packages and add-ons (e.g., CMAP-COE).  Many
-spread-sheet applications also contain easy-to-use features that
-support scientists with the structuring process (e.g., cut / copy /
-paste, search / replace, sort, split-window, annotation,
-track-changes, freeze panes, and so on).
+---------------------------------------------------------------------- 12).
+"What is the rationale for the proposed extensions later in the paper, i.e.
+what are the use cases or evaluation-driven itersations?"
 
-[mark]: Above sounds good to me.
-
-----------------------------------------------------------------------
-4). "Something that might be pointed out - the tool could be used to
-help enforce some community standards (e.g., a common use of a
-"part-of" relation)."
-
-[josh]:
-- It would be nice to say something like (although I not sure where it
-  would be placed):
-
-Furthermore, blocks can be reused (e.g., relationship blocks) to
-ensure that future ontology development adheres to any community
-agreed-upon representations, and will result in the smoother ontology
-integration within ecology.
-
-[mark]: Nice
-
-[shawn]: Because owlifier spreadsheets can be converted to owl
-ontologies, these can be further imported in owlifier spreadsheets
-(e.g., for extension).  This follows the basic OWL infrastructure.
-
-----------------------------------------------------------------------
-5). "In the references: Usually, the word "pages" is abbreviated to
-pp. Ref 18: should be to a journal article, not a web site - perhaps
-Raskin and Pan (2005) in Computers & Geosciences. Ref 19: to EKAW?
-What is that?  Compare Refs 4 and 7 - different wordings are used to
-describe the same conference."
-
-FIXED
-
-----------------------------------------------------------------------
-Reviewer #2:
-----------------------------------------------------------------------
-6). "In the early stages of the paper there are some assertions
-without reference about what is and is not difficult. See p.2 end of
-2nd, 3rd and 4th paragraphs.  Can a little more detail be added to
-support the statements? If not, then noting that these are
-suppositions would be useful for the reader."
-
-To address this issue we added additional text on the need for new
-tools in ontology development for ecologists (e.g., compared with ...)
-Also, much of the motivation for the work is based on our work with
-ecologists and information managers, which we refer to in the paper.
-
-[josh]: Honestly, I don't find a problem here.  I think paragraph 2 is
-fine and obvious to anyone who's used Protege (we could potential say
-"based on our experience in an ecological informatics project
-constructing ontologies [SEEK]").  I can't find anything not
-referenced in paragraph 3.  And the example used by the reviewer in
-paragraph 4 IS referenced and not so much about the level of logical
-description, and basically says "you can capture the level of
-description shown in this cited paper as necessary for improved
-discovery and integration".
-
-[mark]: I think the assertion about "lack of suitable ontology editing
-tools is a major barrier for widescale adoption in ecology" is the
-issue, and begs the question as to why this isn't the case in other
-domains.  So, I suggest we delete the last sentence of paragraph 2 on
-pg 2, and add a following paragraph that elaborates and clarifies that
-point:
-
-"A well-funded ontology development effort, Gene Ontology (GO) has
-been underway for over a decade within the molecular biology
-community, and has led to some major successes enriching genomic and
-biomedical data resources with annotations to a community-based
-knowledge representation. A major lesson learned from these efforts
-was the importance of keeping the ontology focused, and indeed GO has
-a highly delimited scope (Bada et al.)  Ecology, on the other hand,
-encompasses an extremely broad range of scales and disciplines, such
-that a coordinated ontology development effort would need to involve a
-large number of experts to cover the diverse types of information
-relevant to ecological investigations.  At this point, we believe that
-ecological researchers need to begin constructing knowledge
-representations for their own specialized subdisciplines.  Most
-ontology editing tools, however, typically require substantial
-expertise in formal knowled ge representation, and present a major
-barrier for more wide-scale creation and adoption of ontologies in
-ecology."
-
-[Bada, Michael, Robert Stevens, Carol Goble, A. Gil, Michael
-Ashburner, Judith A. Blake, J. Michael Cherry, Midori Harris, Suzanna
-Lewis.  2004. A short study on the success of the gene ontology.
-Journal of Web Semantics, vol 1: 235-240.]
-
-----------------------------------------------------------------------
-7). "Was there any quantifiable basis for choosing the predicates?  In
-particular, the statement" "While not as expressive as OWL-DL, our
-approach can produce ontology structures that are essential for
-improved data discovery and integration [14]."  indicates some level
-of decision in the reduced description but without any details,
-especially in terms for DL-expressivity, it is SHIQ, SHOIN, or what?"
-
-As far as I am aware, the arrangement of DL constructs used in
-OWLifier does not directly map to any previously named DL sublanguage,
-however, it does provides only OWL-DL based constructs, and thus can
-be considered best as a sublanguage of SHOIN(D). Note that although
-many of the OWL-DL constructs can be used, they can only be used in
-certain contexts. Also, some of the more foundational constructs of
-DLs are not considered, e.g., value restrictions and concept
-disjunction. While additional blocks can easily be added to OWLifier
-spreadsheet conventions, the proposed set of blocks were chosen to
-provide a simplified syntax (for a novice DL user) while leading to a
-"natural" or "intuitive" semantics of blocks for these users.
-
-----------------------------------------------------------------------
-8). "There are no mention of instances, that I could find, at all of
-instances?  Are these part of the use case or not? If not, this really
-needs to be stated.  An ontology without instances has limited
-application."
-
-Our main goal is to provide a mechanism for describing ecological
-concepts, and then to allow additional tools to use these concepts for
-annotating scientific data (which represents the instances/values). We
-briefly mention this now in the introduction, and also elaborate on
-the scope of owlifier later in the paper.
-
-[josh]: Maybe at the start of third paragraph of introduction: This
-paper presents a novel approach for ontology creation that aims at
-being more intuitive for ecologists that can be used to rapidly
-construct large ontologies of ecological terms (classes) and the
-subsequent annotation of scientific data (instances).
-
-----------------------------------------------------------------------
-9). "Fig.1 suggests a round-trip approach to knowledge engineering,
-but in fact it is not. How does the present formulation represent to
-scientiest what gets dropped in the back conversion from OWL to the
-spreadsheet format?"
-
-We reworded the implementation section to make it clearer that the
-conversion is not information preserving. ... the owlifier tool will
-notify the user of this situation, but at this time no additional
-support is provided.
-
-[josh]: I guess we need be a bit more transparent in the last sentence
-of the bottom paragraph, page 10; something like:
-   While straightforward to perform, this last conversion will only
-preserve OWL-DL constructs that are represented by owlifier blocks.
-
-[mark]: Substitute term "back conversion" instead of "last
-conversion", and "that are" to "that can be"
-
-----------------------------------------------------------------------
-10). "Most of all, there is almost no (or none at all) description of
-the implemenation or any evaluation, i.e. the how, why, etc.  What has
-this application been applied to?"
-
-[josh]: We could mention that current efforts are underway using
-owlifier to structure list of keywords from the US LTER program.  And
-that the spread-sheet approach works well for these kind of mid-level
-term sorting exercises (e.g., 50-5000) terms.  Nonetheless, (from
-experience) larger numbers of terms does get difficult to navigate and
-structure, although the use of blocks means that you can focus on
-tractable chunks of terms and properties.  Furthermore, the use of
-spread-sheet software features greatly improve navigation, commenting,
-and editing capabilities (e.g., sort, freeze panes, split window,
-search, multiple worksheets / windows, comments, track-changes).
-
-----------------------------------------------------------------------
-11). "It seems that it would be easy to generate complex blocks. How
-are moderate size ontologies developed?"
-
-----------------------------------------------------------------------
-12). "What is the rationale for the proposed extensions later in the
-paper, i.e. what are the use cases or evaluation-driven itersations?"
+The point of these extensions are primarily for round-trip support of owlifier to OWL-DL and back. We added and revised the last paragraph to better articulate the goals and motivation of this work.
 
@@
 
-----------------------------------------------------------------------
-13). "Possibly the authors could signal at the end of section 1
-something like "Readers unfamiliar with logic notation may safely
-ignore the small amount they may encounter in what follows.""
+---------------------------------------------------------------------- 13).
+"Possibly the authors could signal at the end of section 1 something like "Readers unfamiliar with logic notation may safely ignore the small amount they may encounter in what follows.""
 
-[josh]: I like the reviewer's idea about a sentence at the end of
-section 1 saying that "Readers unfamiliar with logic notation may
-safely ignore the small amount they may encounter in what follows" and
-focus on the examples accompanying each block description.
+We added a sentence stating this.
 
-[mark]: I agree
+---------------------------------------------------------------------- 14).
+"Although not enough to recommend required additions, I am left a little puzzled as to why the authors seem to be silent on two issues: OWL data type properties and owlifier instance entities.
+Both of these would seem easy to represent in their owlifier syntax.
+I would also imagine that both could arise in representing and reasoning about both dataset metadata and record-level metadata on biodiversity databases.
+So I am left wondering whether I am wrong about the ease of representing them in spreadsheets and attaching DL axioms, or whether instead, the authors believe that these two OWL constructs are better treated in as data for reasoners or elsewhere."
 
-----------------------------------------------------------------------
-14). "Although not enough to recommend required additions, I am left a
-little puzzled as to why the authors seem to be silent on two issues:
-OWL data type properties and owlifier instance entities. Both of these
-would seem easy to represent in their owlifier syntax. I would also
-imagine that both could arise in representing and reasoning about both
-dataset metadata and record-level metadata on biodiversity
-databases. So I am left wondering whether I am wrong about the ease of
-representing them in spreadsheets and attaching DL axioms, or whether
-instead, the authors believe that these two OWL constructs are better
-treated in as data for reasoners or elsewhere."
+Our use cases are largely driven by separating the specification of conceptual information (i.e., t-box assertions) from instance-level information.
+In particular, in our applications instance level information is provided through "semantic annotations" that map data sets to ontology instances.
+In this way, too, multiple data sets can be mapped to the same ontologies.
+Our decision also is inspired in part by the work of Rector et al.
+that state in the OWL Pizza's paper: "this paper only concerns issues in defining OWL classes, since this is the strength of OWL-DL, and most existing classifiers deal with individuals completely or not at all."  However, we may extend owlifier in the future to include datatype properties, which would be a natural and simple extension.
 
-[josh]: I think we might need to be clearer that we are talking about
-class / term structuring and not instantiating with data (class
-instantiating comes later, after the formal conversion to OWL and by
-the knowledge engineer).  Maybe this could be added to the figure in
-some way.
+---------------------------------------------------------------------- 15).
+"On page 4 (Subsection "Overlap blocks", the second sentence has a small English ambiguity: "[...]a given set of classes may have overlapping instances." could reasonably be understood to mean that some instances overlap one another, whereas I believe the clause means "several classes may have common instances." [Grammatically, the overlapping instances do not belong to the collective "set of classes" but rather to each of the members of that collective]"
 
-I admit data-type properties are an obvious exclusion and should be
-justified.  This could be done by simply stating that owlifier is a
-mechanism for structuring classes with no additional, complicating
-information at this stage (although we do allow for cardinality...).
+We fixed this, which now reads: "Overlap blocks explicitly relax this assumption by stating that a given set of classes may share common instances."
 
-[mark]: Yes, it is hard to convey the reason for this focus on
-Classes.  Well, I feel that our use case is driven by using these
-ontologies to annotate data-- values of which become the
-instances. Maybe this is not a proper understanding.  But also even
-Rector et al. OWL Pizzas paper states: "this paper only concerns
-issues in defining OWL classes, since this is the strength of OWL-DL,
-and most existing classifiers deal with individuals completely or not
-at all." So we are in good company by just advancing capability to
-work with classes and relationships.
+---------------------------------------------------------------------- 16). "Reference 22 is missing 'c' in the article title.  "logi" should be "logic"."
 
-Would such support for data-type properties be low-hanging fruit for
-extending owlifier?  Especially the ability to capture enumerated
-lists might be very useful in the future.
-
-----------------------------------------------------------------------
-15). "On page 4 (Subsection "Overlap blocks", the second sentence has a
-small English ambiguity: "[...]a given set of classes may have
-overlapping instances." could reasonably be understood to mean that
-some instances overlap one another, whereas I believe the clause means
-"several classes may have common instances." [Grammatically, the
-overlapping instances do not belong to the collective "set of classes"
-but rather to each of the members of that collective]"
-
-[josh]: I agree with the grammatical comment to help the lay
-reader. overlapping --> common
-
-[mark]: ow about "Overlap blocks explicitly relax this assumption by
-stating that a given set of classes may share common instances" (2nd
-sent in para on Overlap blocks)
-
-----------------------------------------------------------------------
-16). "Reference 22 is missing 'c' in the article title.  "logi" should
-be "logic"."
-
-FIXED
+We fixed this.

### 2009-08-29 docs/pubs/ISEI6/main.tex
almost done...

@@
 \usepackage{graphicx}
+\usepackage{url}
+\usepackage[colorlinks=true,hyperindex=false,bookmarks=false]{hyperref}
 
@@
 
-\address[smb]{UC Davis Genome Center \& Dept. of Computer Science,
-  Gonzaga University}
+\address[smb]{Dept. of Computer Science,
+ Gonzaga University \& UC Davis Genome Center}
+
+%\address[smb]{UC Davis Genome Center}
 
@@
 
-Within the fields of molecular biology and biomedicine considerable
-effort has gone into developing ontologies for improving data
-discovery and integration
-\citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
-benefits can be obtained for ecological data, far fewer efforts exist
-to develop broad and consistent terminologies within ecology
-\citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
-The use of formal ontologies can significantly enhance metadata
-descriptions of ecological data. For instance, annotating data with
-ontology terms can both help users interpret data as well as enable
-advanced capabilities for data discovery and integration, e.g., by
-exploiting subsumption and part-of hierarchies as well as more formal
-constraints such as cardinality restrictions on properties and term
-equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.
+Within the fields of molecular biology and biomedicine considerable effort has gone into developing ontologies for improving data discovery and integration \citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}.
+While similar benefits can be obtained for ecological data, far fewer efforts exist to develop broad and consistent terminologies within ecology \citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
+The use of formal ontologies can significantly enhance metadata descriptions of ecological data.
+For instance, annotating data with ontology terms can both help users interpret data as well as enable advanced capabilities for data discovery and integration, e.g., by exploiting subsumption and part-of hierarchies as well as more formal constraints such as cardinality restrictions on properties and term equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.
 
-Efforts to engage scientists in the development of ontologies
-typically leverage the W3C Web Ontology Language (OWL)
-\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
-for representing and sharing ontologies. A key advantage of OWL is
-that it is supported by a wide range of generic tools, including
-editors
-\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
-reasoning systems
-\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
-languages
-\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
-and storage technologies
-\citep{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
-however, are primarily targeted at experts in knowledge engineering
-and software development familiar with the underlying description
-logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true
-with ontology editors (such as Protege, SWOOP, etc.), which allow for
-very detailed ontology specifications, but at the same time require
-considerable understanding of the underlying ontology formalisms and
-syntax.  A well-funded ontology development effort, the Gene Ontology
-(GO), has been underway for over a decade within the molecular biology
-community, and has led to some major successes enriching genomic and
-biomedical data resources with annotations to community-based
-ontologies. A major lesson learned from these efforts was the
-importance of keeping the ontology focused, and indeed GO has a highly
-delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}
-Ecology, on the other hand, encompasses an extremely broad range of
-scales and disciplines, such that a coordinated ontology development
-effort would need to involve a large number of experts to cover the
-diverse types of information relevant to ecological investigations.
-At this point, we believe that ecological researchers need to begin
-constructing ontologies for their own specialized subdisciplines.
-Most ontology editing tools, however, typically require substantial
-expertise in formal knowledge representation, and present a major
-barrier for more wide-scale creation and adoption of ontologies in
-ecology.
+Efforts to engage scientists in the development of ontologies typically leverage the W3C Web Ontology Language (OWL) \citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax for representing and sharing ontologies.
+A key advantage of OWL is that it is supported by a wide range of generic tools, including editors \citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop}, reasoning systems \citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query languages \citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules}, and storage technologies \citep{carroll04:_jena,broekstra02:_sesam}.
+However, most of these tools are primarily targeted at experts in knowledge engineering and software development familiar with the underlying description logic semantics of OWL-DL \citep{grau08:_owl}.
+This is especially true with ontology editors (such as Protege, SWOOP, etc.), which allow for very detailed ontology specifications, but at the same time require considerable understanding of the underlying ontology formalisms and syntax.
+A well-funded ontology development effort, the Gene Ontology (GO), has been underway for over a decade within the molecular biology community, and has had success in enriching genomic and biomedical data resources with annotations to community-based ontologies.
+A major lesson learned from these efforts was the importance of keeping the ontology focused, and indeed GO has a highly delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}.
+Ecology, on the other hand, encompasses an extremely broad range of scales and disciplines, such that a coordinated ontology development effort would need to involve a large number of experts to cover the diverse types of information relevant to ecological investigations.
+At this point, we believe that ecological researchers need to begin constructing ontologies for their own specialized sub-disciplines, and due to the substantial expertise in formal knowledge representation required in most ontology editing tools, new tools are needed to enable wide-scale creation and adoption of ontologies in ecology.
 
-This paper presents a novel approach for ontology creation that aims
-at being more intuitive for ecologists and that can be used to rapidly
-construct large ontologies of ecological terms (classes) and the
-subsequent annotation of scientific data (instances).  Our approach is
-to allow scientists to use common and familiar spreadsheet-based tools
-(e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to
-describe, in an intuitive way, different aspects of an ontology, and
-then to take these descriptions and convert them into full-fledged OWL
-ontologies using a software application called \owlifier. The use of a
-\emph{familiar} editing environment can improve the speed and
-understanding of ontology construction and can help users avoid the
-often time-consuming task of locating, downloading, installing, and
-learning fundamentally new software applications and interfaces (e.g.,
-CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
-\citep{knublauch04:_editin_descr_logic_ontol_with}).  Many spreadsheet
-applications also contain easy-to-use features that support scientists
-in the process of editing and managing \owlifier\ ontology documents
-(e.g., via cut/copy/paste, search/replace, sort, split-window,
-track-changes, freeze panes, and so on).
+This paper presents a novel approach for ontology creation that aims at being more intuitive for ecologists and that can be used to rapidly construct large ontologies of ecological terms (classes and properties), which can then be used in a subsequent step for annotation of scientific data (instances) \citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.
+Our approach is to allow scientists to use common and familiar spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to describe, in an intuitive way, different aspects of an ontology, and then to take these descriptions and convert them into full-fledged OWL ontologies using a software application called \owlifier.
 
-An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
-a predefined template structure for users to fill in. Each non-empty
-row in an \owlifier\ table constitutes a block. Each block defines
-different aspects of an ontology including ontology classes,
-subclasses, synonyms, and properties.  We also provide blocks for
-plain-text descriptions of classes and properties, and for referencing
-one or more existing ontologies (e.g., to extend an existing ontology
-or to define ontology articulations). Blocks can be sparse (inheriting
-from previous blocks), which can further simplify the creation of
-large ontologies. We omit several OWL features associated with
-properties (e.g., role inclusion axioms, domain and range property
-restrictions, and data-type properties among others) as these can be
-confusing to construct for non-experts and are more suitable for
-experienced knowledge modelers.
+An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have a predefined template structure for users to fill in.
+Each non-empty row in an \owlifier\ table constitutes a block.
+Each block defines different aspects of an ontology including classes, subclasses, synonyms, and properties.
+We also provide blocks for plain-text descriptions of classes and properties, and for referencing one or more existing ontologies (e.g., to extend an existing ontology or to define ontology articulations).
+Blocks can be sparse (inheriting from previous blocks), which can further simplify the creation of large ontologies.
+We omit several OWL features associated with properties (e.g., role inclusion axioms, domain and range property restrictions, and data-type properties among others) as these can be confusing to construct for non-experts and are more suitable for experienced knowledge modelers.
 
-While not as expressive as OWL-DL, our approach can produce ontology
-structures that are essential for improved data discovery and
-integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
-important, because spreadsheet tools are frequently used by ecologists
-to store and analyze data, \owlifier\ can provide ecologists with a
-familiar and accessible user interface for ontology creation. This
-approach also leverages the ability of spreadsheet tools to organize
-and manipulate tabular data, e.g., allowing users to rapidly construct
-class hierarchies from long lists of keywords.  In this way, an
-ecologist can easily construct (or import) a set of terms, and then
-incrementally organize these into class hierarchies, properties, and
-constraints. In initial experiments with ecologists and evolutionary
-biologists studying trait data, we found that \owlifier\ enabled them
-to quickly and easily comprehend and construct relatively complex and
-meaningful ontologies.
+While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration \citep{madin07:_ontol_for_descr_and_synth}.
+Just as important, because spreadsheet tools are already frequently used by ecologists to store and analyze data, \owlifier\ can provide ecologists with a familiar and accessible user interface for ontology creation.
+This approach also leverages the easy-to-use interfaces provided by many spreadsheet tools for organizing and manipulating tabular data, e.g., via cut/copy/paste, search/replace, sort, split-window, track-changes, freeze panes, and so on.
+In this way, an ecologist can easily construct (or load) a set of terms, and then incrementally organize these into class hierarchies, properties, and constraints.
+The use of an editing environment that is familiar to scientists can significantly help improve the speed and understanding of ontology construction and avoid the often time-consuming task of locating, downloading, installing, and learning fundamentally new software applications and interfaces (e.g., CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege \citep{knublauch04:_editin_descr_logic_ontol_with}).
+In initial experiments with ecologists and evolutionary biologists studying trait data, we found that \owlifier\ enabled them to quickly and easily comprehend and construct relatively complex and meaningful ontologies.
 
-The rest of this paper is organized as follows. In
-\secref{sec:owlifier} we describe the basic syntax and semantics of
-\owlifier. We define blocks that support a large subset of OWL-DL and
-that also generally follow the ontology creation guidelines defined in
-\citep{rector04:_owl_pizzas}. We also simplify certain aspects of
-ontology creation using OWL-DL, e.g., by assuming classes are disjoint
-by default (unless specified otherwise)
-\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
-additional characteristics of \owlifier\ and discuss issues with
-respect to classification and reasoning. In
-\secref{sec:implementation} we briefly describe the \owlifier\
-implementation, and conclude in \secref{sec:conclusion} with related
-and future work. In general, the goal of \owlifier\ is not to support
-all constructs in OWL-DL, but instead to provide a higher-level
-ontology syntax (via spreadsheet blocks) that is easy for ecologists
-to use and understand while also providing the necessary constructs
-for developing typical ecological ontologies. By compiling \owlifier\
-to OWL-DL, we also allow for experts to refine and extend the ontology
-using more advanced ontology editing tools if necessary
-(cf.~\figref{fig:owlifier}).
+The rest of this paper is organized as follows.
+In \secref{sec:owlifier} we describe the basic syntax and semantics of \owlifier.
+The semantics of \owlifier\ blocks is given by mappings from \owlifier\ expressions to description-logic statements.
+Readers unfamiliar with description-logic notation may safely ignore these mappings, focusing instead on the descriptions and examples of blocks given in \secref{sec:owlifier}.
+We define blocks that support a large subset of OWL-DL and that also generally follow the ontology creation guidelines defined in \citep{rector04:_owl_pizzas}.
+We also simplify certain aspects of ontology creation using OWL, e.g., by assuming classes are disjoint by default (unless specified otherwise) \citep{rector04:_owl_pizzas}.
+In \secref{sec:features} we describe additional characteristics of \owlifier\ and discuss issues with respect to classification and reasoning.
+In \secref{sec:implementation} we briefly describe the \owlifier\ implementation, and conclude in \secref{sec:conclusion} with related and future work.
+In general, the goal of \owlifier\ is not to support all constructs in OWL-DL, but instead to provide a higher-level ontology syntax (via spreadsheet blocks) that is easy for ecologists to use and understand while also providing the necessary constructs for developing typical ecological ontologies.
+By compiling \owlifier\ to OWL-DL, we also allow for experts to refine and extend the ontology using more advanced ontology editing tools if necessary (cf.~\figref{fig:owlifier}).
 
@@
 
-As described above, an \owlifier\ table defines an OWL-DL
-\citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of
-\emph{blocks} representing one or more ontology definitions.  Each
-non-empty row in an \owlifier\ table corresponds to a block. The type
-of the block is given in the first column of the row. 
+As described above, an \owlifier\ table defines an OWL-DL \citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of \emph{blocks} representing one or more ontology definitions.
+Each non-empty row in an \owlifier\ table corresponds to a block.
+The type of the block is given in the first column of the row.
 % This section describes in detail each type of block supported by
 % \owlifier.
-We assume that if any properties or classes used in a block are not
-imported from another ontology, then they are to be added to the
-ontology being specified by the \owlifier\ table (i.e., the
-``current'' ontology). In general, we name blocks according to the
-terms used in
-\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
-as opposed to the names used for corresponding constructs in
-OWL-DL. This choice of block names helps to simplify terminology
-(e.g., we use ``relation'' below instead of ``object property''),
-allows \owlifier\ to easily generate ontologies that extend the
-observational model of
-\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
-and avoids confusion with established terms commonly used within
-ecology (e.g., ``class'').
+We assume that if any properties or classes used in a block are not imported from another ontology, then they are to be added to the ontology being specified by the \owlifier\ table (i.e., the ``current'' ontology).
+In general, we name blocks according to the terms used in \citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth} as opposed to the names used for corresponding constructs in OWL-DL.
+This choice of block names helps to simplify terminology (e.g., we use ``relationship'' below instead of ``object property''), allows \owlifier\ to easily generate ontologies that extend the observational model of \citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}, and avoids confusion with established terms commonly used within ecology (e.g., ``class'').
 
@@
 further in \secref{sec:features}), classes are generally
-assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
-this assumption by stating that a given set of classes may have
-overlapping instances. Rows containing overlap blocks take the form
+assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax this assumption by
+stating that a given set of classes may share common instances. Rows containing overlap blocks take the form
 \begin{itemize}
@@
 
-\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
-blocks in that they state equivalences between classes. A sufficient
-block serves as a modifier to entity blocks, relationship blocks
-(including transitive blocks), and cardinality blocks, modifying the
-associated description logic axioms by using equivalence ($\equiv$) in
-place of subclass ($\sqsubseteq$). For example the block
+\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym blocks in that they state equivalences between classes.
+We consider two types of sufficient blocks.
+A sufficient entity block takes the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$
+\\ \hline
+\end{tabular} \hfill ($n \ge 2$)
+\end{itemize}
+where each $c_i$ is a class.
+A sufficient entity block induces the description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$ stating that the class $c_1$ is equivalent to the intersection of the classes $c_2$ through $c_n$.
+Similarly, a sufficient relationship block takes the form
+\begin{itemize}
+\item[]
+  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ \hline \end{tabular} \end{itemize} where $p$ is an object property, $c_1$ and $c_2$ are classes, and \textsf{not} is an optional qualifier.
+A sufficient relationship block (without the \textsf{not} qualifier) induces the description logic axiom $c_1 \equiv \exists p .
+c_2$ stating that the class $c_1$ is equivalent to the set of individuals that are $p$-related to at least one individual of class $c_2$.
+The \textsf{not} qualifier states that the absence of the property is a defining characteristic of the class.
+Thus, a sufficient relationship block containing a \textsf{not} qualifier induce the description logic axiom $c_1 \equiv \neg \exists p .
+c_2$.
+The use of multiple sufficient blocks for a particular class results in a single axiom in which each constraint (i.e., class constructor) is combined via intersection ($\sqcap$).
+For example, the following blocks
+\begin{small}
 \begin{tabbing}
-  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
+~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \= \textsf{Mammal} \= \kill 
+~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \> \textsf{Hair} \\
+  ~~\textsf{sufficient} \> \textsf{entity} \> \textsf{Eutheria} \> \textsf{Mammal} \\
+  ~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \> \textsf{Eutheria}  \textsf{EpipubicBone} 
 \end{tabbing}
-induces the description logic axiom $\textsf{Mammal} \equiv \exists
-\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
-that has Hair as a part is a Mammal. Additionally, we allow
-relationship blocks within a sufficient block to be extended with
-\textsf{not} to state that the absence of the property is a defining
-characteristic of the class. For example the following blocks
+induce the following description logic axioms.
 \begin{tabbing}
-  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
-  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
+~~$\textsf{Mammal} \equiv \exists \textsf{hasPart}.\textsf{Hair}$\\ 
+~~$\textsf{Eutheria} \equiv \textsf{Mamma} \sqcap \neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$
 \end{tabbing}
-induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
-\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
-any Mammal that does not have an Epipubic Bone as a part is a
-Eutheria. Note also that multiple sufficient blocks for a particular
-class result in a single axiom in which each constraint (i.e., class
-constructor) is combined via intersection ($\sqcap$).
+\end{small}
+These axioms state that \emph{any} individual that has Hair as a part is a Mammal and that \emph{any} Mammal that does not have an Epipubic Bone as a part is a Eutheria.
 
@@
 editing environment and an overall fewer number of modeling mistakes.
+
+\begin{figure*}
+  \centering
+  \includegraphics[scale=.75]{architecture.pdf}
+  \caption{The basic \owlifier\ application and relation to other
+    technologies.}
+  \label{fig:owlifier}
+\end{figure*}
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[scale=.75]{architecture.pdf}
-  \caption{The basic \owlifier\ application and relation to other
-    technologies.}
-  \label{fig:owlifier}
-\end{figure*}
-
 \section{The \Owlifier\ Implementation}
@@
 
-\figref{fig:owlifier} shows the general architecture of the \owlifier\
-application. A scientist or researcher first creates a spreadsheet
-containing a set of ontology definitions using \owlifier\ conventions.
-The spreadsheet is then exported as a plain-text file containing the
-\owlifier\ table, e.g., using a CSV or tab-delimited format. The
-\owlifier\ table is then sent to the \owlifier\ application, which
-performs a number of steps that include: (i) parsing the file; (ii)
-generating the corresponding OWL-DL representation (e.g., in the
-current implementation, the Jena API is used \citep{carroll04:_jena});
-(iii) validating the ontology and ensuring it is consistent via an
-OWL-DL reasoner (e.g., the current implementation uses Pellet
-\citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
-file. The resulting OWL file can then be refined and extended via
-existing OWL tools (e.g., Protege or SWOOP). It is possible that the
-refined ontology is converted back to a corresponding \owlifier\
-representation (shown using dashed lines in
-\figref{fig:owlifier}). While straightforward to perform, this back
-conversion is not information preserving since not all OWL-DL
-statements can be represented through \owlifier\ blocks.
+\figref{fig:owlifier} shows the general architecture of the \owlifier\ application.
+A scientist first creates a spreadsheet containing a set of ontology definitions using \owlifier\ conventions.
+The spreadsheet is then exported as a plain-text file containing the \owlifier\ table, e.g., using a CSV or tab-delimited format.
+The \owlifier\ table is then sent to the \owlifier\ application, which performs a number of steps that include: (i) parsing the file; (ii) generating the corresponding OWL-DL representation (e.g., in the current implementation, the OWL-API is used \citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena} could be used as well); (iii) validating the ontology and ensuring it is consistent via an OWL-DL reasoner (e.g., the current implementation uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL file.
+The resulting OWL file can then be refined and extended via existing OWL tools (e.g., Protege or SWOOP).
+It is also possible for the refined ontology to be converted back to a corresponding \owlifier\ representation (shown using dashed lines in \figref{fig:owlifier}).
+While straightforward to perform, this ``back'' conversion is not information preserving since not all OWL-DL statements can be represented through \owlifier\ blocks.
+That is, the conversion from OWL-DL to \owlifier\ will only preserve OWL-DL constructs that are represented by \owlifier\ blocks.
+In particular, \owlifier\ does not support a number of OWL-DL language constructs, which include domain and range constraints, datatype properties, sub properties, property constraints other than transitive (i.e., functional, symmetric, inverse, inverse functional), value restrictions, individuals, etc., as well as various combinations of OWL-DL constructs.
 
-The current implementation of the \owlifier\ application is written in
-Java and supports a subset of the blocks defined in
-\secref{sec:owlifier}. In particular, we are currently extending
-\owlifier\ to support cardinality and sufficient blocks as well as to
-perform the (potentially lossy) back conversion from OWL-DL files to
-corresponding \owlifier\ blocks. The current implementation of
-\owlifier\ is being used within a project focused on integrating
-vegetation trait data.  As noted above, not all the blocks described
-here are yet supported, however, these priority needs are clearly
-emerging from our interactions with the vegetation scientists.
-Nevertheless, these scientists were able to rapidly prototype and
-revise their trait ontologies using \owlifier, with little to no
-instruction in formal logic or knowledge modeling.  In addition,
-\owlifier\ is being used to construct term hierarchies from large sets
-of keywords (harvested from existing metadata documents) as well as to
-define articulations between existing ontologies and the observation
-ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.
+The current implementation of \owlifier\ is written as an open-source Java application\footnote{See \url{https://semtools.ecoinformatics.org/owlifier}} and supports a subset of the blocks defined in \secref{sec:owlifier}.
+In particular, we are currently extending \owlifier\ to fully support sufficient blocks (the remaining blocks described in \secref{sec:owlifier} are implemented) as well as the potentially lossy, back conversion from OWL-DL files to corresponding \owlifier\ blocks.
+The current implementation of \owlifier\ is being used within a project focused on integrating vegetation trait data.
+As noted above, not all the blocks described here are yet supported, however, these priority needs are clearly emerging from our interactions with the vegetation scientists.
+Nevertheless, these scientists were able to rapidly prototype and revise their trait ontologies using \owlifier, with little to no instruction in formal logic or knowledge modeling.
+In addition, \owlifier\ is being used to construct term hierarchies from large sets of keywords (harvested from existing metadata documents) as well as to define articulations between existing ontologies and the observation ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.
+As we continue to use \owlifier\ within these projects, we intend to extend the application as needed to support additional blocks and services.
 
@@
 
-This paper presents a new approach for developing ontologies to
-address barriers in ontology development and adoption by
-ecologists. Our approach allows scientists to use familiar spreadsheet
-software (e.g., frequently used by ecologists for storing and
-analyzing data) to create ontologies by filling in a set of templates,
-or blocks, that generate OWL-DL class hierarchies, properties, and
-constraints via the \owlifier\ application.  This approach can provide
-a more intuitive and accessible ontology editing environment for
-ecologists, especially compared to existing OWL-based tools that
-require considerable expertise in the underlying logic formalisms.
+This paper presents a new approach for developing ontologies to address barriers in ontology development and adoption by ecologists.
+Our approach allows scientists to use familiar spreadsheet software (e.g., frequently used by ecologists for storing and analyzing data) to create ontologies by filling in a set of templates, or blocks, that generate OWL-DL class hierarchies, properties, and constraints via the \owlifier\ application.
+This approach can provide a more intuitive and accessible ontology editing environment for ecologists, especially compared to existing OWL-based tools that require considerable expertise in the underlying logic formalisms.
+Similar to standard OWL ontologies, \owlifier\ spreadsheets can import and extend existing OWL-DL ontologies (including those generated from \owlifier\ spreadsheets), which further supports the community development and interoperability of ontologies.
+Further, \owlifier\ blocks (or similarly, partially filled in blocks) can be reused during ontology creation to help ensure that future ontology development adheres to best practices and community standards (e.g., similar to ``part-of'' properties defined in the Gene Ontology \citep{bada04:_short_study_succes_of_gene_ontol}).
 
-Protege provides a variety of ontology editing plug-ins, including a
-simple interface for text-based editing of class hierarchies. In
-\citep{kola07:_impor_spread_data_into_proteg}, an approach is described
-for importing spreadsheet-based ontology descriptions into
-Protege. However, this approach aims at supporting \emph{ad-hoc}
-spreadsheet structures by providing an intermediate interface for
-mapping these structures into ontology axioms. This approach is
-similar to others (e.g.,
-\citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
-for defining mappings from relational data to RDF and OWL ontology
-class and instance data. To the best of our knowledge, however, our
-approach is the first to consider a detailed set of templates that can
-support a large subset of existing OWL constructs. In addition, we
-define a number of shortcuts for creating \owlifier\ tables, including
-sparse blocks and default semantics (e.g., disjoint classes, which is
-a form of global taxonomic constraint defined in
-\citep{thau07:_reason_about_taxon_in_first_order_logi}) that further
-simplify ontology creation for end users.
+Protege provides a variety of ontology editing plug-ins, including a simple interface for text-based editing of class hierarchies.
+In \citep{kola07:_impor_spread_data_into_proteg}, an approach is described for importing spreadsheet-based ontology descriptions into Protege.
+However, this approach aims at supporting \emph{ad-hoc} spreadsheet structures by providing an intermediate interface for mapping these structures into ontology axioms.
+This approach is similar to others (e.g., \citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin}) for defining mappings from relational data to RDF and OWL ontology class and instance data.
+In addition to these approaches, a number of visual editing environments have been developed to help novice users create OWL-based ontologies (e.g., \citep{hayes05:_collab_knowl_captur_in_ontol}).
+To the best of our knowledge, however, our approach is the first to consider an intuitive spreadsheet-based approach together with a detailed set of templates that can support a large subset of existing OWL constructs.
+In addition, we define a number of shortcuts for creating \owlifier\ tables, including sparse blocks and default semantics (e.g., disjoint classes) that further simplify ontology creation for end users.
 
-As future work, we intend to extend our current \owlifier\
-implementation to support the full set of blocks defined above. We
-also intend to explore approaches to fully support conversions between
-\owlifier\ tables and OWL-DL ontologies. In particular, we want to
-extend \owlifier\ to re-apply changes made by knowledge representation
-experts that previously modified the OWL-DL ontology generated by
-\owlifier. For instance, if an OWL-DL ontology generated from an
-\owlifier\ table is refined and extended by a knowledge representation
-expert, then converted back into an \owlifier\ ontology that is
-further edited and extended by a scientist, and then converted again
-into an OWL-DL ontology, we want to maintain (i.e., re-apply) the
-original extensions and edits created by the knowledge representation
-expert that are still relevant. We are also developing extensions to
-\owlifier\ to support conversion into the observation ontology
-framework presented in \citep{bowers08:_concep_model_framew_for_expres}
-and intend to perform additional testing and evaluation of the
-\owlifier\ approach with ecologists.
+As future work, we plan to extend our current \owlifier\ implementation to support the full set of blocks defined here as well as introduce additional blocks, e.g., for creating OWL datatype properties.
+We also would like to explore approaches for supporting round-trip conversions between \owlifier\ tables and OWL-DL ontologies that allow scientists and knowledge-representation experts to incrementally develop \owlifier-based ontologies.
+Specifically, we want to extend the \owlifier\ application so that it can re-apply changes made by knowledge-representation experts that previously modified the OWL-DL ontology generated by \owlifier.
+For instance, if an OWL-DL ontology generated from an \owlifier\ spreadsheet is refined and extended by a knowledge-representation expert, then converted back into an \owlifier\ ontology that is further edited and extended by a scientist, and then converted again into an OWL-DL ontology, we want to maintain (i.e., re-apply) the original extensions and edits created by the knowledge representation expert that are still relevant (thus supporting lossless conversions).
+We also plan to extend the \owlifier\ application to support translation into the observation ontology framework presented in \citep{bowers08:_concep_model_framew_for_expres} and perform additional testing and evaluation of the \owlifier\ approach with a wide range of ecologists.
 

### 2009-08-29 docs/pubs/ISEI6/main.tex
final version

@@
 
-Within the fields of molecular biology and biomedicine considerable effort has gone into developing ontologies for improving data discovery and integration \citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}.
-While similar benefits can be obtained for ecological data, far fewer efforts exist to develop broad and consistent terminologies within ecology \citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
-The use of formal ontologies can significantly enhance metadata descriptions of ecological data.
-For instance, annotating data with ontology terms can both help users interpret data as well as enable advanced capabilities for data discovery and integration, e.g., by exploiting subsumption and part-of hierarchies as well as more formal constraints such as cardinality restrictions on properties and term equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.
+Within the fields of molecular biology and biomedicine considerable
+effort has gone into developing ontologies for improving data
+discovery and integration
+\citep{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
+benefits can be obtained for ecological data, far fewer efforts exist
+to develop broad and consistent terminologies within ecology
+\citep{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
+The use of formal ontologies can significantly enhance metadata
+descriptions of ecological data. For instance, annotating data with
+ontology terms can both help users interpret data as well as enable
+advanced capabilities for data discovery and integration, e.g., by
+exploiting subsumption and part-of hierarchies as well as more formal
+constraints such as cardinality restrictions on properties and term
+equivalence \citep{madin08:_advan_ecolog_resear_with_ontol}.
 
-Efforts to engage scientists in the development of ontologies typically leverage the W3C Web Ontology Language (OWL) \citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax for representing and sharing ontologies.
-A key advantage of OWL is that it is supported by a wide range of generic tools, including editors \citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop}, reasoning systems \citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query languages \citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules}, and storage technologies \citep{carroll04:_jena,broekstra02:_sesam}.
-However, most of these tools are primarily targeted at experts in knowledge engineering and software development familiar with the underlying description logic semantics of OWL-DL \citep{grau08:_owl}.
-This is especially true with ontology editors (such as Protege, SWOOP, etc.), which allow for very detailed ontology specifications, but at the same time require considerable understanding of the underlying ontology formalisms and syntax.
-A well-funded ontology development effort, the Gene Ontology (GO), has been underway for over a decade within the molecular biology community, and has had success in enriching genomic and biomedical data resources with annotations to community-based ontologies.
-A major lesson learned from these efforts was the importance of keeping the ontology focused, and indeed GO has a highly delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}.
-Ecology, on the other hand, encompasses an extremely broad range of scales and disciplines, such that a coordinated ontology development effort would need to involve a large number of experts to cover the diverse types of information relevant to ecological investigations.
-At this point, we believe that ecological researchers need to begin constructing ontologies for their own specialized sub-disciplines, and due to the substantial expertise in formal knowledge representation required in most ontology editing tools, new tools are needed to enable wide-scale creation and adoption of ontologies in ecology.
+Efforts to engage scientists in the development of ontologies
+typically leverage the W3C Web Ontology Language (OWL)
+\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
+for representing and sharing ontologies. A key advantage of OWL is
+that it is supported by a wide range of generic tools, including
+editors
+\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
+reasoning systems
+\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
+languages
+\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
+and storage technologies
+\citep{carroll04:_jena,broekstra02:_sesam}. However, most of these
+tools are primarily targeted at experts in knowledge engineering and
+software development familiar with the underlying description logic
+semantics of OWL-DL \citep{grau08:_owl}. This is especially true with
+ontology editors (such as Protege, SWOOP, etc.), which allow for very
+detailed ontology specifications, but at the same time require
+considerable understanding of the underlying ontology formalisms and
+syntax.  The development of the Gene Ontology (GO) has been underway
+for over a decade within the molecular biology community, and has lead
+to significant improvements in data interoperability by enriching
+genomic and biomedical data resources with annotations to
+community-based ontologies. A major lesson learned from these efforts
+was the importance of keeping the ontology focused, and indeed GO has
+a highly delimited scope
+\citep{bada04:_short_study_succes_of_gene_ontol}. Ecology, on the
+other hand, encompasses an extremely broad range of scales and
+disciplines, such that a coordinated ontology development effort would
+need to involve a large number of experts to cover the diverse types
+of information relevant to ecological investigations.  We believe that
+ecological researchers must also begin constructing ontologies for
+their own specialized subdisciplines. However, due to the substantial
+expertise in formal knowledge representation that is required to use
+most existing ontology editing tools, new tools and approaches are
+needed to enable the wide-scale creation and adoption of ontologies in
+ecology.
 
-This paper presents a novel approach for ontology creation that aims at being more intuitive for ecologists and that can be used to rapidly construct large ontologies of ecological terms (classes and properties), which can then be used in a subsequent step for annotation of scientific data (instances) \citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.
-Our approach is to allow scientists to use common and familiar spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to describe, in an intuitive way, different aspects of an ontology, and then to take these descriptions and convert them into full-fledged OWL ontologies using a software application called \owlifier.
+This paper presents a novel approach for ontology creation that aims
+at being more intuitive for ecologists and that can be used to rapidly
+construct large ontologies of ecological terms (classes and
+properties). These terms can then be used in subsequent steps to
+annotate ecological data (instances), e.g., as in
+\citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.
+Our approach is to allow scientists to use common and familiar
+spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
+Office Spreadsheet) to describe, in an intuitive way, different
+aspects of an ontology, and then to take these descriptions
+(represented as spreadsheets) and convert them into full-fledged OWL
+ontologies using a software application called \owlifier.
 
-An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have a predefined template structure for users to fill in.
-Each non-empty row in an \owlifier\ table constitutes a block.
-Each block defines different aspects of an ontology including classes, subclasses, synonyms, and properties.
-We also provide blocks for plain-text descriptions of classes and properties, and for referencing one or more existing ontologies (e.g., to extend an existing ontology or to define ontology articulations).
-Blocks can be sparse (inheriting from previous blocks), which can further simplify the creation of large ontologies.
-We omit several OWL features associated with properties (e.g., role inclusion axioms, domain and range property restrictions, and data-type properties among others) as these can be confusing to construct for non-experts and are more suitable for experienced knowledge modelers.
+An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
+a predefined template structure for users to fill in. Each non-empty
+row in an \owlifier\ spreadsheet constitutes a block. Each block
+defines different aspects of an ontology including classes,
+subclasses, synonyms, and properties.  We also provide blocks for
+plain-text descriptions of classes and properties, and for referencing
+one or more existing ontologies (e.g., to extend an existing ontology
+or to define ontology articulations). Blocks can be sparse (inheriting
+from previous blocks), which further simplifies the creation of large
+ontologies by minimizing the amount of redundant information that must
+be provided in the \owlifier\ tabular representation. Several of the
+more advanced OWL features are omitted in \owlifier, primarily
+associated with properties (e.g., role inclusion axioms, domain and
+range property restrictions, and data-type properties among others),
+as these can be confusing for non-experts and are more suitable for
+experienced knowledge modelers.
 
-While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration \citep{madin07:_ontol_for_descr_and_synth}.
-Just as important, because spreadsheet tools are already frequently used by ecologists to store and analyze data, \owlifier\ can provide ecologists with a familiar and accessible user interface for ontology creation.
-This approach also leverages the easy-to-use interfaces provided by many spreadsheet tools for organizing and manipulating tabular data, e.g., via cut/copy/paste, search/replace, sort, split-window, track-changes, freeze panes, and so on.
-In this way, an ecologist can easily construct (or load) a set of terms, and then incrementally organize these into class hierarchies, properties, and constraints.
-The use of an editing environment that is familiar to scientists can significantly help improve the speed and understanding of ontology construction and avoid the often time-consuming task of locating, downloading, installing, and learning fundamentally new software applications and interfaces (e.g., CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege \citep{knublauch04:_editin_descr_logic_ontol_with}).
-In initial experiments with ecologists and evolutionary biologists studying trait data, we found that \owlifier\ enabled them to quickly and easily comprehend and construct relatively complex and meaningful ontologies.
+While not as expressive as OWL-DL, \owlifier\ can be used to produce
+ontology structures that are essential for improved data discovery and
+integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
+important, because spreadsheet tools are already frequently used by
+ecologists to store and analyze data, \owlifier\ can provide
+ecologists with a familiar and accessible user interface for ontology
+creation. This approach also leverages the easy-to-use interfaces
+provided by many spreadsheet tools for organizing and manipulating
+tabular data, e.g., via cut/copy/paste, search/replace, sort,
+split-window, track-changes, freeze panes, and so on.  In this way, an
+ecologist can easily construct (or load) a set of terms, and then
+incrementally organize these into class hierarchies, properties, and
+constraints.  The use of an editing environment that is familiar to
+scientists can significantly help improve the speed and understanding
+of ontology construction and avoid the often time-consuming task of
+locating, downloading, installing, and learning fundamentally new
+software applications and interfaces (e.g., CMAP-COE
+\citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
+\citep{knublauch04:_editin_descr_logic_ontol_with}). In initial
+experiments with ecologists and evolutionary biologists studying trait
+data, we found that \owlifier\ enabled them to quickly and easily
+comprehend and construct relatively complex and meaningful ontologies.
 
-The rest of this paper is organized as follows.
-In \secref{sec:owlifier} we describe the basic syntax and semantics of \owlifier.
-The semantics of \owlifier\ blocks is given by mappings from \owlifier\ expressions to description-logic statements.
-Readers unfamiliar with description-logic notation may safely ignore these mappings, focusing instead on the descriptions and examples of blocks given in \secref{sec:owlifier}.
-We define blocks that support a large subset of OWL-DL and that also generally follow the ontology creation guidelines defined in \citep{rector04:_owl_pizzas}.
-We also simplify certain aspects of ontology creation using OWL, e.g., by assuming classes are disjoint by default (unless specified otherwise) \citep{rector04:_owl_pizzas}.
-In \secref{sec:features} we describe additional characteristics of \owlifier\ and discuss issues with respect to classification and reasoning.
-In \secref{sec:implementation} we briefly describe the \owlifier\ implementation, and conclude in \secref{sec:conclusion} with related and future work.
-In general, the goal of \owlifier\ is not to support all constructs in OWL-DL, but instead to provide a higher-level ontology syntax (via spreadsheet blocks) that is easy for ecologists to use and understand while also providing the necessary constructs for developing typical ecological ontologies.
-By compiling \owlifier\ to OWL-DL, we also allow for experts to refine and extend the ontology using more advanced ontology editing tools if necessary (cf.~\figref{fig:owlifier}).
+The rest of this paper is organized as follows. In
+\secref{sec:owlifier} we describe the basic syntax and semantics of
+\owlifier. The semantics of \owlifier\ blocks is given by mapping
+\owlifier\ expressions to description-logic statements. Readers
+unfamiliar with description-logic notation may safely ignore these
+mappings, focusing instead on the descriptions and examples of blocks
+given in \secref{sec:owlifier}.  We define blocks that support a large
+subset of OWL-DL and that also generally follow the ontology creation
+guidelines defined in \citep{rector04:_owl_pizzas}. We also simplify
+certain aspects of ontology creation using OWL, e.g., by assuming
+classes are disjoint by default (unless specified otherwise)
+\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
+additional characteristics of \owlifier\ and discuss issues with
+respect to classification and reasoning. In
+\secref{sec:implementation} we briefly describe the
+\owlifier\ implementation, and conclude in \secref{sec:conclusion}
+with related and future work. In general, the goal of \owlifier\ is
+not to support all constructs in OWL-DL, but instead to provide a
+higher-level ontology syntax (via spreadsheet blocks) that is easy for
+ecologists to use and understand while also providing the necessary
+constructs for developing typical ecological ontologies. By compiling
+\owlifier\ to OWL-DL, we also allow for experts to refine and extend
+the ontology using more advanced ontology editing tools if necessary
+(cf.~\figref{fig:owlifier}).
 
@@
 the normal way. As an example, the following block imports the SWEET
-``Earth Realm'' ontology \citep{raskin05}
+``Earth Realm'' ontology \citep{raskin05}.
 \begin{small}
@@
 block.
+\begin{small}
 \begin{tabbing}
@@
 \end{tabbing}
+\end{small}
 This block states that, e.g., within a thermally stratified lake, the
@@
 
-\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym blocks in that they state equivalences between classes.
-We consider two types of sufficient blocks.
-A sufficient entity block takes the form
+\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
+blocks in that they state equivalences between classes. We consider
+two types of sufficient blocks. A sufficient entity block takes the
+form
 \begin{itemize}
@@
 \end{itemize}
-where each $c_i$ is a class.
-A sufficient entity block induces the description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$ stating that the class $c_1$ is equivalent to the intersection of the classes $c_2$ through $c_n$.
-Similarly, a sufficient relationship block takes the form
+where each $c_i$ is a class. A sufficient entity block induces the
+description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$
+stating that the class $c_1$ is equivalent to the intersection of the
+classes $c_2$ through $c_n$.  Similarly, a sufficient relationship
+block takes the form
 \begin{itemize}
 \item[]
-  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ \hline \end{tabular} \end{itemize} where $p$ is an object property, $c_1$ and $c_2$ are classes, and \textsf{not} is an optional qualifier.
-A sufficient relationship block (without the \textsf{not} qualifier) induces the description logic axiom $c_1 \equiv \exists p .
-c_2$ stating that the class $c_1$ is equivalent to the set of individuals that are $p$-related to at least one individual of class $c_2$.
-The \textsf{not} qualifier states that the absence of the property is a defining characteristic of the class.
-Thus, a sufficient relationship block containing a \textsf{not} qualifier induce the description logic axiom $c_1 \equiv \neg \exists p .
-c_2$.
-The use of multiple sufficient blocks for a particular class results in a single axiom in which each constraint (i.e., class constructor) is combined via intersection ($\sqcap$).
-For example, the following blocks
+  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ 
+\hline \end{tabular} 
+\end{itemize} 
+where $p$ is an object property, $c_1$ and $c_2$ are classes, and
+\textsf{not} is an optional qualifier.  A sufficient relationship
+block (without a \textsf{not} qualifier) induces the description logic
+axiom $c_1 \equiv \exists p . c_2$ stating that the class $c_1$ is
+equivalent to the set of individuals that are $p$-related to at least
+one individual of class $c_2$. The \textsf{not} qualifier states that
+the absence of the property is a defining characteristic of the
+class. Thus, a sufficient relationship block containing a \textsf{not}
+qualifier induces the description logic axiom $c_1 \equiv \neg \exists
+p . c_2$.  The use of multiple sufficient blocks for a particular
+class results in a single axiom in which each constraint (i.e., class
+constructor) is combined via intersection ($\sqcap$). For example, the
+following blocks
 \begin{small}
 \begin{tabbing}
-~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \= \textsf{Mammal} \= \kill 
-~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \> \textsf{Hair} \\
-  ~~\textsf{sufficient} \> \textsf{entity} \> \textsf{Eutheria} \> \textsf{Mammal} \\
-  ~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \> \textsf{Eutheria}  \textsf{EpipubicBone} 
+~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \=
+  \textsf{Mammal} \= \kill ~~\textsf{sufficient} \>
+  \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \>
+  \textsf{Hair} \\ ~~\textsf{sufficient} \> \textsf{entity} \>
+  \textsf{Eutheria} \> \textsf{Mammal} \\ ~~\textsf{sufficient} \>
+  \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \>
+  \textsf{Eutheria} \textsf{EpipubicBone}
 \end{tabbing}
@@
 \end{small}
-These axioms state that \emph{any} individual that has Hair as a part is a Mammal and that \emph{any} Mammal that does not have an Epipubic Bone as a part is a Eutheria.
+These axioms state that \emph{any} individual that has Hair as a part
+is a Mammal and that \emph{any} Mammal that does not have an Epipubic
+Bone as a part is a Eutheria.
 
 \myblock{Comment Blocks.} There are two types of comment blocks that
-each provide a mechanism to add plain-text comments to \owlifier\
-tables. A description block attaches a plain-text comment to classes
-and properties. Rows containing description blocks take the form
+each provide a mechanism to add plain-text comments to
+\owlifier\ tables. A description block attaches a plain-text comment
+to classes and properties. Rows containing description blocks take the
+form
 \begin{itemize}
@@
 Here we briefly describe some of the additional features of \owlifier,
-specifically focusing on the use of disjoint classes, \owlifier\
-reasoning, and additional block syntax.
+specifically focusing on the use of disjoint classes,
+\owlifier\ reasoning, and additional block syntax.
 
@@
 of classes are typically defined as being disjoint (e.g., sibling
-classes), and stating these disjoint constraints is often
-time consuming since each pair of classes must be given an explicit
+classes), and stating these disjoint constraints is often time
+consuming since each pair of classes must be given an explicit
 disjoint assertion. Editors such as Protege
@@
 \exists P. C$ generated from a sufficient block, and an axiom $B
-\sqsubseteq \exists P.C$ generated from a relationship block, the axiom
-$B \sqsubseteq A$ would be inferred. For both the case of synonym and
-sufficient blocks, the use of equivalence permits a number of
-additional inferences to be made via an OWL-DL reasoner.
+\sqsubseteq \exists P.C$ generated from a relationship block, the
+axiom $B \sqsubseteq A$ would be inferred. For both the case of
+synonym and sufficient blocks, the use of equivalence permits a number
+of additional inferences to be made via an OWL-DL reasoner.
 
@@
 provided (as well as, e.g., property closure axioms).  We explicitly
-do not consider these constraints in the current version of \owlifier\
-because they also often result in modeling errors for inexperienced
-OWL-DL users \citep{rector04:_owl_pizzas}. Instead, we adopt the
-approach of more traditional description logics, which do not have
-explicit domain and range axioms.  Although not currently supported in
-\owlifier, domain and range constraints as well as property closure
-axioms can be inferred from given relationship blocks.
+do not consider these constraints in the current version of
+\owlifier\ because they also often result in modeling errors for
+inexperienced OWL-DL users \citep{rector04:_owl_pizzas}. Instead, we
+adopt the approach of more traditional description logics, which do
+not have explicit domain and range axioms.  Although not currently
+supported in \owlifier, domain and range constraints as well as
+property closure axioms can be inferred from given relationship
+blocks.
 
@@
 
-\figref{fig:owlifier} shows the general architecture of the \owlifier\ application.
-A scientist first creates a spreadsheet containing a set of ontology definitions using \owlifier\ conventions.
-The spreadsheet is then exported as a plain-text file containing the \owlifier\ table, e.g., using a CSV or tab-delimited format.
-The \owlifier\ table is then sent to the \owlifier\ application, which performs a number of steps that include: (i) parsing the file; (ii) generating the corresponding OWL-DL representation (e.g., in the current implementation, the OWL-API is used \citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena} could be used as well); (iii) validating the ontology and ensuring it is consistent via an OWL-DL reasoner (e.g., the current implementation uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL file.
-The resulting OWL file can then be refined and extended via existing OWL tools (e.g., Protege or SWOOP).
-It is also possible for the refined ontology to be converted back to a corresponding \owlifier\ representation (shown using dashed lines in \figref{fig:owlifier}).
-While straightforward to perform, this ``back'' conversion is not information preserving since not all OWL-DL statements can be represented through \owlifier\ blocks.
-That is, the conversion from OWL-DL to \owlifier\ will only preserve OWL-DL constructs that are represented by \owlifier\ blocks.
-In particular, \owlifier\ does not support a number of OWL-DL language constructs, which include domain and range constraints, datatype properties, sub properties, property constraints other than transitive (i.e., functional, symmetric, inverse, inverse functional), value restrictions, individuals, etc., as well as various combinations of OWL-DL constructs.
+\figref{fig:owlifier} shows the general architecture of the
+\owlifier\ application. A scientist first creates a spreadsheet
+containing a set of ontology definitions using \owlifier\ conventions.
+The spreadsheet is then exported as a plain-text file containing the
+\owlifier\ table, e.g., using a CSV or tab-delimited format. The
+\owlifier\ table is then sent to the \owlifier\ application, which
+performs a number of steps that include: (i) parsing the file; (ii)
+generating the corresponding OWL-DL representation (e.g., in the
+current implementation, the OWL-API is used
+\citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena}
+could be used as well); (iii) validating the ontology and ensuring it
+is consistent via an OWL-DL reasoner (e.g., the current implementation
+uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the
+corresponding OWL-DL file. The resulting OWL file can then be refined
+and extended via existing OWL tools (e.g., Protege or SWOOP). It is
+also possible for the refined ontology to be converted back to a
+corresponding \owlifier\ representation (shown using dashed lines in
+\figref{fig:owlifier}). While straightforward to perform, this
+``back'' conversion is not information preserving since not all OWL-DL
+statements can be represented through \owlifier\ blocks.  That is, the
+conversion from OWL-DL to \owlifier\ will only preserve OWL-DL
+constructs that are represented by \owlifier\ blocks.  In particular,
+\owlifier\ does not support a number of OWL-DL language constructs,
+including domain and range constraints, datatype properties, sub
+properties, property constraints (i.e., functional, symmetric, and
+inverse functional), value restrictions, individuals, etc., as well as
+various combinations of OWL-DL constructs.
 
-The current implementation of \owlifier\ is written as an open-source Java application\footnote{See \url{https://semtools.ecoinformatics.org/owlifier}} and supports a subset of the blocks defined in \secref{sec:owlifier}.
-In particular, we are currently extending \owlifier\ to fully support sufficient blocks (the remaining blocks described in \secref{sec:owlifier} are implemented) as well as the potentially lossy, back conversion from OWL-DL files to corresponding \owlifier\ blocks.
-The current implementation of \owlifier\ is being used within a project focused on integrating vegetation trait data.
-As noted above, not all the blocks described here are yet supported, however, these priority needs are clearly emerging from our interactions with the vegetation scientists.
-Nevertheless, these scientists were able to rapidly prototype and revise their trait ontologies using \owlifier, with little to no instruction in formal logic or knowledge modeling.
-In addition, \owlifier\ is being used to construct term hierarchies from large sets of keywords (harvested from existing metadata documents) as well as to define articulations between existing ontologies and the observation ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.
-As we continue to use \owlifier\ within these projects, we intend to extend the application as needed to support additional blocks and services.
+The current implementation of \owlifier\ is written as an open-source
+Java application\footnote{See
+  \url{https://semtools.ecoinformatics.org/owlifier}} and supports a
+subset of the blocks defined in \secref{sec:owlifier}. In particular,
+we are currently extending \owlifier\ to fully support sufficient
+blocks (the remaining blocks described in \secref{sec:owlifier} are
+implemented) as well as the potentially lossy, back conversion from
+OWL-DL files to corresponding \owlifier\ blocks. The current
+implementation of \owlifier\ is being used within a project focused on
+integrating vegetation trait data.  As noted above, not all the blocks
+described here are yet supported, however, these priority needs are
+clearly emerging from our interactions with the vegetation scientists.
+Nevertheless, these scientists were able to rapidly prototype and
+revise their trait ontologies using \owlifier, with little to no
+instruction in formal logic or knowledge modeling.  In addition,
+\owlifier\ is being used to construct term hierarchies from large sets
+of keywords (harvested from existing metadata documents) as well as to
+define articulations between existing ontologies and the observation
+ontology defined in \citep{madin07:_ontol_for_descr_and_synth}. As we
+continue to use \owlifier\ within these projects, we intend to extend
+the application as needed to support additional blocks and services.
 
@@
 
-This paper presents a new approach for developing ontologies to address barriers in ontology development and adoption by ecologists.
-Our approach allows scientists to use familiar spreadsheet software (e.g., frequently used by ecologists for storing and analyzing data) to create ontologies by filling in a set of templates, or blocks, that generate OWL-DL class hierarchies, properties, and constraints via the \owlifier\ application.
-This approach can provide a more intuitive and accessible ontology editing environment for ecologists, especially compared to existing OWL-based tools that require considerable expertise in the underlying logic formalisms.
-Similar to standard OWL ontologies, \owlifier\ spreadsheets can import and extend existing OWL-DL ontologies (including those generated from \owlifier\ spreadsheets), which further supports the community development and interoperability of ontologies.
-Further, \owlifier\ blocks (or similarly, partially filled in blocks) can be reused during ontology creation to help ensure that future ontology development adheres to best practices and community standards (e.g., similar to ``part-of'' properties defined in the Gene Ontology \citep{bada04:_short_study_succes_of_gene_ontol}).
+This paper presents a new approach for developing ontologies to
+address barriers in ontology development and adoption by
+ecologists. Our approach allows scientists to use familiar spreadsheet
+software (e.g., frequently used by ecologists for storing and
+analyzing data) to create ontologies by filling in a set of templates,
+or blocks, that generate OWL-DL class hierarchies, properties, and
+constraints via the \owlifier\ application.  This approach can provide
+a more intuitive and accessible ontology editing environment for
+ecologists, especially compared to existing OWL-based tools that
+require considerable expertise in the underlying logic
+formalisms. Similar to standard OWL ontologies,
+\owlifier\ spreadsheets can import and extend existing OWL-DL
+ontologies (including those generated from \owlifier\ spreadsheets),
+which further supports the community development and interoperability
+of ontologies. Blocks in \owlifier\ (or similarly, partially filled in
+blocks) can also be reused during ontology creation to help ensure
+that future ontology development adheres to best practices and
+community standards (e.g., similar to ``part-of'' properties defined
+in the Gene Ontology
+\citep{bada04:_short_study_succes_of_gene_ontol}).
 
-Protege provides a variety of ontology editing plug-ins, including a simple interface for text-based editing of class hierarchies.
-In \citep{kola07:_impor_spread_data_into_proteg}, an approach is described for importing spreadsheet-based ontology descriptions into Protege.
-However, this approach aims at supporting \emph{ad-hoc} spreadsheet structures by providing an intermediate interface for mapping these structures into ontology axioms.
-This approach is similar to others (e.g., \citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin}) for defining mappings from relational data to RDF and OWL ontology class and instance data.
-In addition to these approaches, a number of visual editing environments have been developed to help novice users create OWL-based ontologies (e.g., \citep{hayes05:_collab_knowl_captur_in_ontol}).
-To the best of our knowledge, however, our approach is the first to consider an intuitive spreadsheet-based approach together with a detailed set of templates that can support a large subset of existing OWL constructs.
-In addition, we define a number of shortcuts for creating \owlifier\ tables, including sparse blocks and default semantics (e.g., disjoint classes) that further simplify ontology creation for end users.
+Protege provides a variety of ontology editing plug-ins, including a
+simple interface for text-based editing of class hierarchies. In
+\citep{kola07:_impor_spread_data_into_proteg}, an approach is
+described for importing spreadsheet-based ontology descriptions into
+Protege. However, this approach aims at supporting \emph{ad-hoc}
+spreadsheet structures by providing an intermediate interface for
+mapping these structures into ontology axioms. This approach is
+similar to others (e.g.,
+\citep{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
+for defining mappings from relational data to RDF and OWL ontology
+class and instance data. In addition to these approaches, a number of
+visual editing environments have been developed to help novice users
+create OWL-based ontologies (e.g.,
+\citep{hayes05:_collab_knowl_captur_in_ontol}). To the best of our
+knowledge, however, our approach is the first to consider an intuitive
+spreadsheet-based approach together with a detailed set of templates
+that can support a large subset of existing OWL constructs. In
+addition, we define a number of shortcuts for creating
+\owlifier\ tables, including sparse blocks and default semantics
+(e.g., disjoint classes) that further simplify ontology creation for
+end users.
 
-As future work, we plan to extend our current \owlifier\ implementation to support the full set of blocks defined here as well as introduce additional blocks, e.g., for creating OWL datatype properties.
-We also would like to explore approaches for supporting round-trip conversions between \owlifier\ tables and OWL-DL ontologies that allow scientists and knowledge-representation experts to incrementally develop \owlifier-based ontologies.
-Specifically, we want to extend the \owlifier\ application so that it can re-apply changes made by knowledge-representation experts that previously modified the OWL-DL ontology generated by \owlifier.
-For instance, if an OWL-DL ontology generated from an \owlifier\ spreadsheet is refined and extended by a knowledge-representation expert, then converted back into an \owlifier\ ontology that is further edited and extended by a scientist, and then converted again into an OWL-DL ontology, we want to maintain (i.e., re-apply) the original extensions and edits created by the knowledge representation expert that are still relevant (thus supporting lossless conversions).
-We also plan to extend the \owlifier\ application to support translation into the observation ontology framework presented in \citep{bowers08:_concep_model_framew_for_expres} and perform additional testing and evaluation of the \owlifier\ approach with a wide range of ecologists.
+As future work, we plan to extend our current
+\owlifier\ implementation to support the full set of blocks defined
+here as well as introduce additional blocks, e.g., for creating OWL
+datatype properties. We also would like to explore approaches for
+supporting round-trip conversions between \owlifier\ tables and OWL-DL
+ontologies that allow scientists and knowledge-representation experts
+to incrementally develop \owlifier-based ontologies.  Specifically, we
+want to extend the \owlifier\ application so that it can store and
+re-apply changes made by knowledge-representation experts that
+previously modified an OWL-DL ontology generated by \owlifier. For
+instance, if an OWL-DL ontology generated from an
+\owlifier\ spreadsheet is refined and extended by a
+knowledge-representation expert, then converted back into an
+\owlifier\ ontology that is further edited and extended by a
+scientist, and then converted again into an OWL-DL ontology, we want
+to maintain (i.e., re-apply) the original extensions and edits created
+by the knowledge-representation expert that are still relevant (thus
+supporting lossless conversions).  We also plan to extend the
+\owlifier\ application to support translation into the observation
+ontology framework presented in
+\citep{bowers08:_concep_model_framew_for_expres} and perform
+additional testing and evaluation of the \owlifier\ approach with a
+wide range of ecologists.
 

### 2010-01-11 docs/algdraft/Materialize_DB.tex
add meetingnotes and algorithm draft directory.

\documentclass[10pt]{article}
%\title{}
%\author{}
%\date{}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{a4}
\usepackage{color}
\renewcommand\floatpagefraction{0.99}
\renewcommand\topfraction{0.99}
\renewcommand\bottomfraction{0.99}
\renewcommand\textfraction{.05}
\setcounter{totalnumber}{5}
%\setlength{\parindent}{0pt}
%\pdfpagewidth 8.5in
%\pdfpageheight 11in

\begin{document}

Note that in our description, we use
$A.*$ to represent the annotation information.
More specifically,
\begin{itemize}
\item MeasType = $\{\langle \underline{MeasTypeId}, ObsTypeId, CharType, StdType, ProtType, Precision, Value, isKey \rangle\}$;
\item ObservationType = $\{\langle \underline{ObsTypeId}, EntTypeId, isUnique \rangle\}$ \\
	{\bf HP Question}: I did not use ``AnnotId'' in the algorithm, so I remove it here. How shall this be used?
\item ContextType = $\{\langle \underline{ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$\\
	{\bf HP Question}: Add the {\em isIdentify} for the purpose of checking whether we need to use one observation's context for key, is this ok?
\item Map = $\{\langle \underline{MeasTypeId, ObsTypeId, Cond}, Val\rangle\}$
\end{itemize}

$OBOE.*$, on the other hand, represent the {\em OUTPUT} data represented in the OBOE model. In detail,
\begin{itemize}
\item $Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances materialized from $Dataset$.\\
	{\bf HP Question}: I did not use ``AnnotId'' in the algorithm, so I remove it here. How shall this be used?
\item $Measurenebt=\{\langle \underline{MeasId}, ObsId, Characteristic, Val\rangle\}$\\     
	%{\bf HP Question}: I changed ``MeasType'' in the ppt to ``Characteristic'', is it right? It is consistent with the example.
         \textcolor{blue}{Changed to: }\\
	$Measurenebt=\{\langle \underline{MeasId}, \textcolor{blue}{MeasTypeId}, ObsId, Val\rangle\}$\\   
          {\bf HP note}: the other information about the measurement type, e.g., standard, characteristic, can be gotten using {\em MeasTypeId}.
         In  real application, we can think of "duplicating the measurement type" information. Too detail. Ignore here. 
\item $Entity=\{\langle \underline{EntId}, EntType\rangle\}$
\item $Context =\{\langle \underline{ObsId, ConextObsId, RelType}\rangle\}$\\
	{\bf HP Question}: I did not use ``Relationship" table, do we really need to instantiate this information?
\end{itemize}

Each row in the input dataset represents the information related to one observation and its contexts.

The algorithm tries to catch the {\em key}, {\em unique} and {\em identifying} constraints in the annotation during the materialization process.

%\newpage
\begin{algorithm} [htb]
\caption{{\bf MaterializeDB} ($Dataset, A$)}
{\small
\begin{algorithmic}
%\STATE{{\bf MaterializeDB} ($Dataset, A$)}
\STATE /* {\em Dataset}: [Input] in the form of flat file */\\
\STATE /* {\em A }:  [Input] Annotations*/
\STATE
\STATE $ObsIdx =\emptyset$;  /* for keeping index $\langle ObsTypeId, \textcolor{blue}{KeyVal}\rangle \rightarrow ObsId$*/
\STATE $EntIdx = \emptyset$;  /* for keeping index $\langle MeasTypeId, \textcolor{blue}{KeyVal}\rangle \rightarrow EntId$*/
\FOR { (each $Row\langle A1, A2, \cdots, An\rangle \in Dataset$)}	
	%\STATE MapRow($Row, MapList, ObsIdx, EntIdx$);
    \STATE $MeasSet = \emptyset$; /* Keep the set of new measurement instances*/

    \STATE
    \STATE/* 1. Create new orphan measurement instances*/
	\FOR {(each $m = \langle MeasTypeId, \textcolor{blue}{ResAttribute}, Cond, Val\rangle \in A.Map$)}

		\IF{\textcolor{blue}{($m.ResAttribute != Row.Ai.Attrname$) OR} ($Row.Ai$ does not satisfy $m.Cond$)}
		\STATE cotinue;
		\ENDIF
		\STATE $ObsTypeId$ = GetObsTypeId ($A.MeasType, m.MeasTypeId$);
		\STATE
		\STATE $MeasId$ = GetNewMeasId($OBOE.Measurement$);
		%\STATE $Cha$ = GetCharacteristic4Meas($A.MeasType, m.MeasTypeId$);
		%\STATE Create a measurement instance $mi = \langle MeasId, null, Cha, m.val\rangle$;		
		\STATE \textcolor{blue}{{\bf if}($m.Val!=NULL$) $MeasVal$ = $m.Val$; }
		\STATE \textcolor{blue}{{\bf else} $MeasVal$ = $Row.Ai.Val$;}
		\STATE Create a measurement instance $mi = \langle MeasId, MeasTypeId, null, MeasVal\rangle$;
		\STATE Add $mi$ to $MeasSet$;
	\ENDFOR

    \STATE
    \STATE /* 2. Get observation types and measurement types with new instances*/
    \STATE $KeyIdx=\emptyset$ /* Keep index for $ObsTypeId \rightarrow \{mi\}$ */
    \FOR{(each $mi \in MeasSet$)}
    \STATE $ObsTypeId$ = GetObsTypeId ($A.MeasType, mi.MeasTypeId$);
	\STATE Update $KeyIdx$ by changing the item $ObsTypeId \rightarrow \{mi\}$;
    \ENDFOR

    \STATE
    \FOR{(each $ObsTypeId \in KeyIdx.keys$)}
        \STATE /*Get the key value for this observation. \\
                   Case 1: generally, it is the value for the ``key'' measurement. \\
                   Case 2: several measurement types are marked with ``key yes'', the key value is the combined value of these several measurement. \\
                   Case 3: this object type is marked with ``identifying yes, the key value '' is the combined value with its context observation's key measurement values. */
        \STATE $KeyVal$ = GetObsTypeKeys ($ObsTypeId, KeyIdx$);

        \STATE
        \STATE /* 3. Get an existing or create a new entity instance*/
        \STATE $HasKey$ = false;
        \IF{($ObsTypeId$ has key measurements)}
            \STATE $HasKey$ = true;
        \ENDIF
    	\STATE $EntId$ = {\bf MaterializeEntity}($ObsTypeId, HasKey, KeyVal, EntIdx, A, OBOE$)

        \STATE
        \STATE/* 4. Get an existing or create a new observation instance*/
         \STATE $ContextIdx = \emptyset$;  /* keep index $ObsTypeId \rightarrow ObsId$ to materialize context*/
        \FOR{(each $ObsTypeId \in KeyIdx.keys)$}
        \STATE $IsObsUnique$ = checkIfObsUnique($A.ObservationType, ObsTypeId$);
        \STATE $ObsHasKey$ = $HasKey~\&\& IsObsUnique$ ;
        \STATE $ObsId$ = {\bf MaterializeObs}($ObsTypeId, ObsHasKey,KeyVal,EntId,ObsIdx, OBOE$);

        \STATE
        \STATE /*Maintain the measurement instances for this observation*/
        \STATE $miSet$ = GetMeasInst($KeyIdx, ObsTypeId$);
        \IF{($ObsId$ is a new one)}
            \STATE Set the $obsId$ to each $mi \in miSet$ so that $mi$-s are not orphans;
        \ELSE
            \STATE Discard all the $mi \in miSet$;
        \ENDIF
        \STATE Put all the $mi \in miSet$ to $OBOE.Measurement$;
        \STATE

    	\STATE $ContextIdx = ContextIdx \cup \{ObsTypeId \rightarrow ObsId\}$;
        \ENDFOR
    \ENDFOR
    \STATE
	\STATE /*5. Materialize context relationships among observation instances*/
	\STATE {\bf MaterializeContext}($ContextIdx, A, OBOE$)
\ENDFOR
\STATE Return OBOE;
\end{algorithmic}
}
\end{algorithm}

\begin{algorithm} [htb]
\caption{{\bf MaterializeEntity}($ObsTypeId, HasKey, KeyVal, EntIdx, A, OBOE)$}
\begin{algorithmic}
\STATE $EntType$ = GetObsEntityType ($A.ObservationType, ObsTypeId$)
\STATE $CrtNewEntInst$ = true;
\IF{ ($HasKey$==true) }
	\STATE $EntId$ = GetEntId($ObsTypeId, KeyVal, EntIdx$); %/*From the current entity key index get the id of the entity instance*/
	\IF{($EntId$ is not Null)}
		\STATE $CrtNewEntInst$ = false;
	\ENDIF
\ENDIF
\IF{($CrtNewEntInst==true$)}
	\STATE $EntId$ = CrtEntId($EntType$);
	\STATE Create an entity instance $ei = \langle EntId, EntType\rangle $;
	\STATE Put $ei$ to $OBOE.Entity$;
	\IF{$HasKey$==true)}
        \STATE /*Only when this is the key measurement, we need to maintain the index*/
        \STATE $EntIdx = EntIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow EntId$\};
    \ENDIF
\ENDIF
\STATE return $EntId$;
\end{algorithmic}
\end{algorithm}

\begin{algorithm} [htb]
\caption{{\bf MaterializeObs}($ObsTypeId, HasKey,KeyVal,EntId, ObsIdx, OBOE$)}
\begin{algorithmic}
		\STATE $CrtNewObsInst$ = true;
		\IF{ ($HasKey$==true) }
			\STATE $ObsId$ = GetObsId($ObsTypeId, KeyVal, ObsIdx$); %/*From the current observation key index get the id of the observation instance*/
			\IF{($ObsId$ is not Null)}
				\STATE $CrtNewObsInst$ = false;
			\ENDIF
		\ENDIF
		\IF{($CrtNewObsInst==true)$}
    		\STATE Create an observation instance $oi = \langle ObsId,  EntId\rangle $
    		\STATE Put $oi$ to $OBOE.Observation$;
    		\IF{($HasKey$==true)}
                \STATE /*Only when it has key measurement and it is identified as unique, we need to maintain the index*/
                \STATE $ObsIdx = ObsIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow ObsId$\};
            \ENDIF
		\ENDIF
		\STATE Return $ObsId$;
\end{algorithmic}
\end{algorithm}

\begin{algorithm} [htb]
\caption{{\bf MaterializeContext}($ContextIdx, A, OBOE$)}
\begin{algorithmic}
\FOR{ ($ObsTypeId \rightarrow ObsId \in ContextIdx$)}
\STATE $ContextObsTypeId, Rel$ = GetContextObsTypeRel($A.ContextType, ObsTypeId$);
\IF{($ContextObsTypeId$ is not Null)}
	\STATE $ContextObsId$ = GetContextObsId($ContextIdx, ContextObsTypeId$);
	\STATE Create a context $c=\langle ObsId, ContextObsId, Rel\rangle$;
	\STATE Put $c$ to $OBOE.Context$;
\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

We can run the example in page 6 in the powerpoint file to explain the algorithm.

For Row(2007, 1, piru, 35.8)
    \begin{itemize}
    \item Create a measurement $mi_1=\langle meas1, null, Year, 2007\rangle$
    \item Create a measurement $mi_2=\langle meas2, null, Year, 35.8\rangle$
    \item Create a measurement $mi_3=\langle meas3, null, TaxonomicTypeName, Picea~rubens\rangle$
    \item Create a measurement $mi_4=\langle meas4, null, EntityName, 1\rangle$
    \item Get $MeasSet = \{mi_1, mi_2, mi_3, mi_4\}$;
    \item Step 2: Get $KeyIdx = \{o_1\rightarrow \{mi_1\}, o_2 \rightarrow \{mi_2, mi_3, mi_4\}\}$
    \item Step 3-4: for each observation types $o_1$ and $o_2$
    \item for $o_1$
    \begin{itemize}
        \item Since $m_1$ is specified as key, get the $KeyVal=2007$;
        \item No entity with this key exists in $EntIdx$, create an entity $ei=\langle ent1, TemporalRange\rangle$; Now, $EntIdx=\{\langle o1, 2007\rightarrow ent1\rangle\}$.
        \item Since $o_1$ is specified as {\em distinct}, need to make sure we do not create redundant observations. No entry with this key exists in $ObsIdx$, so, create an observation
        $oi = \langle obs1, ent1\rangle$. Now, $ObsIdx=\{\langle o1, 2007 \rangle \rightarrow obs1\}$
    \end{itemize}
    \item When deal with $o_2$,
        \begin{itemize}
        \item $KeyVal=1$.
        \item Create an entity $ei=\langle ent2, Tree\rangle$; $EntIdx=\{\langle o1, 2007\rightarrow ent1\rangle, \langle o2, 1\rightarrow ent2\rangle\}$.
        \item Create an observation $\langle obs2, ent2 \rangle$. No need to update $ObsIdx$ because $o2$ is not identified as {\em unique}.
        \end{itemize}
    \item Connect $mi_1$ to $obs1$;
    \item Connect $mi_2$, $mi_3$ and $mi_4$ to $obs2$;
    \item Set the context relationship between $obs1$ and $obs2$;
    \end{itemize}

For Row (2008, 1, piru, 36.2)
    \begin{itemize}
    \item Create a measurement $mi_5=\langle meas5, null, Year, 2007\rangle$
    \item Create a measurement $mi_6=\langle meas6, null, Year, 35.8\rangle$
    \item Create a measurement $mi_7=\langle meas7, null, TaxonomicTypeName, Picea~rubens\rangle$
    \item Create a measurement $mi_8=\langle meas8, null, EntityName, 1\rangle$
    \item Get $MeasSet = \{mi_5, mi_6, mi_7, mi_8\}$;
    \item Step 2: Get $KeyIdx = \{o_1\rightarrow \{mi_5\}, o_2 \rightarrow \{mi_6, mi_7, mi_8\}\}$
    \item for type $o_1$
    \begin{itemize}
        \item  $KeyVal=2008$;
        \item Create an entity $\langle ent3, TemporalRange\rangle$; $EntIdx=\{\langle o1, 2007 \rangle \rightarrow ent1, \langle o2, 1 \rangle \rightarrow ent2, \langle o1, 2008 \rangle \rightarrow ent3\}$.
         \item Create an observation $\langle obs3, ent3 \rangle$, and $ObsIdx=\{\langle o1, 2007\rightarrow obs1\rangle, \langle o2, 1\rightarrow obs2\rangle, \langle o1, 2008\rightarrow obs3\rangle\}$
    \end{itemize}
    \item When deal with $o_2$,
        \begin{itemize}
         \item $KeyVal=1$.
        \item {\bf Some item $\langle o2, 1 \rangle \rightarrow ent2$ is already in $EntIdx$, so get the entity id $ent2$. No need to create an entity.}
        \item Since $o2$ is not specified with {\em unique yet}, we NEED to create an observation $\langle obs4, ent2 \rangle$. No need to update $ObsIdx$.
        \end{itemize}
    \end{itemize}
For ROW (2008, 2, abba, 33.2)
    \begin{itemize}
        \item For $o1$'s measurement $2008$, since $\langle o1, 2008 \rangle \rightarrow ent3$ already exists in $EntIdx$, {\bf No need to create a new ENTITY}.
        \item Since $o1$ is specified with {\em unique yes}, and since $\langle o1, 2008 \rangle \rightarrow obs3 $ already exists in $ObsIdx$, {\bf No need to create a new OBSERVATION} and no need to put the measurement for $2008$ into OBOE model.
    \end{itemize}

\end{document}

### 2010-01-15 docs/algdraft/Materialize_DB.tex
add mime type to these two files.

@@
 %\pdfpageheight 11in
-
+\newtheorem{example}{Example}[section]
 \begin{document}
@@
     \item Create a measurement $mi_1=\langle meas1, null, Year, 2007\rangle$
-    \item Create a measurement $mi_2=\langle meas2, null, Year, 35.8\rangle$
+    \item Create a measurement $mi_2=\langle meas2, null, DBH, 35.8\rangle$
     \item Create a measurement $mi_3=\langle meas3, null, TaxonomicTypeName, Picea~rubens\rangle$
@@
 
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{|l|l|l|}
+\hline
+plt & spp & dbh\\\hline
+A & piru & 35.8 \\\hline
+A & piru & 36.2 \\\hline
+B & piru &33.2 \\\hline
+\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset 2 for Example \ref{eg2}}
+\end{table}
+
+\begin{example} [Example with identifying]\label{eg2}
+Use the data in the following table as an example.
+
+For the {\bf first row}, after we run the first step, we got the three measurement instances $MeasSet=mi_1, mi_2, mi_3$. 
+After the second step, we get $KeyIdx = \{\langle o_1\rightarrow \{mi_1\} \rangle, \langle o_2 \rightarrow \{mi_2, mi_3\}\}$.
+
+Now we need to create observation and entity instances. 
+To create entity instance, we need to see whether there is/are key measurements for each observatioin type.  
+Here,  $o_1$ and $o_2$ have key measurements $m1$ and $m_2$ respectively. 
+The same value of the key measurement will denote the same entity. 
+For $o_1$, we get it's key value be $A$. Since there is no such a statement, we create an entity $ei_1$ of type $Plot$.
+Simiarly, for $o_2$, the key value is $(A, Picea~rubens)$ since it has context $o_1$ with {\em identifying yes}, we create an entity $ei_2$ of type $Tree$. \\
+$EntIdx = \{\langle o_1, A\rangle \rightarrow ei_1, \langle o_2, (A, Picea~rubens) \rangle \rightarrow ei_2\}$.
+
+To create observation instance, we need to see whethe the observation type is specified with distinct yes. 
+Here, $o_1$ is specified with {\em distinct yes} while $o_2$ is not. 
+We only check whether we need to merge observation instances for $o_1$ or not. 
+We create an observation instance $oi_1$ whose entity is $ei_1$. 
+We also create an observation instance $oi_2$ whose entity is $ei_2$. 
+$ObsIdx = \{\langle o_1, A\rangle \rightarrow oi_1\}$.
+
+The next step is to connect the measurement instances to observation instances,  $mi_1 \rightarrow oi_1$, $mi_2, mi_3 \rightarrow oi_2$.
+
+The last step for this row is to connect the observations using context relationship. For this instance, we connect $oi_1$ to $oi_2$ with context {\em Within}.
+
+For the {\bf second row}, after the first step, we also get three measurement instances $MeasSet=mi_4, mi_5, mi_6$. 
+After the second step, we get $KeyIdx = \{\langle o_1\rightarrow \{mi_4\} \rangle, \langle o_2 \rightarrow \{mi_5, mi_6\}\}$.
+
+When we create entities, 
+we need to check whether there is any entity corresponding to Key value $\langle o_1, A\rangle$ $\langle, o_2, (A, Picea~rubens)\rangle$ or not.
+We observe that they have corresponding $ei_1$ and $ei_2$. So, we don't need to create entity instances for them.
+
+Now, we create observatins, 
+Since $o_1$ is specified with {\em distinct yes}, we just need to see whether we need to create a new observation for it or get an existing one. 
+The key is $\langle o_1,A\rangle$, which corresponds to $oi_1$. So, we do not need to create a new observation for it. 
+For $o_2$, since no {\em distinct yes} is specified, we create an observation $oi_3$ for it.
+
+For the next step, when we try to connect the measurement instances to observation instances, we realize that we do not create a new observation for $o_1$,
+so it's related measurement instance $mi_4$ can be discarded. While for $o_2$, $m_5, m_6\rightarrow oi_3$.
+
+When we process the {\bf third row}, one thing to note is for $o_1$, now we have key value $\langle o_1, B\rangle$, we create a new entity for it.
+For $o_2$, we have key value $\langle o_2, (B,~Picea~rubens)\rangle$ and create a new entity for it.
+
+\end{example}
+
 \end{document}

### 2010-01-15 docs/algdraft/UseCaseKey.tex
add the usecase discussion draft

\documentclass[10pt]{article}
%\title{}
%\author{}
%\date{}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{a4}
\usepackage{color}
\renewcommand\floatpagefraction{0.99}
\renewcommand\topfraction{0.99}
\renewcommand\bottomfraction{0.99}
\renewcommand\textfraction{.05}
\setcounter{totalnumber}{5}
%\setlength{\parindent}{0pt}
%\pdfpagewidth 8.5in
%\pdfpageheight 11in
\newtheorem{example}{Example}[section]
\begin{document}

{\bf Use cases to show that it's needed to have {\em key yes}, {\em identifying yes} and {distinct yes}}

Q1: Why we need to distinguish the same entities using {\em key yes} and {\em identifying yes})?

Assume the following tabe is the measurement for some plant tree at different spots.

Consider this question that a user may ask. Give me the average dbh for every {\em piru} tree (i.e., tree entity). 
First, we have three observations here.
But how many tree entities here is a question.

There are several cases to consider:
\begin{itemize}
\item Case 1: The naive extreme way to interpret the data is that each observation is from different tree entity. Then, we have {\bf four} tree entities. 
	This may be too strict. People may say, well, I have some observations for the same entity. 
\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
That's obvious that {\em piru} is different from {\em abba}. 
With this constraint, we get {\bf two} tree entities. 
\item Case 3: the assumption of case two has some obvious problem. 
People want to further limit that the same {\em spp} in the same {\em plt} should represent the same tree entity set. 
To achieve this, we use {\em identifying yes}.
   Now, it should return
	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
\end{itemize}

\begin{table}[htb]
\begin{center}
\begin{tabular}{cc}
\begin{tabular}{|l|l|l|}
\hline
plt & spp & dbh\\\hline
A & piru & 35.8 \\\hline
A & piru & 36.2 \\\hline
B & piru &33.2 \\\hline
B&abba&34\\\hline
\end{tabular}
&
\begin{tabular}{|l|l|l|l|}
\hline
plt & area & spp & dbh\\\hline
A & 1.0 & piru & 35.8 \\\hline
A & 1.1 & piru & 36.2 \\\hline
B & &piru &33.2 \\\hline
B& & abba&34\\\hline
\end{tabular}\\
(a) & (b)
\end{tabular}
\end{center}
\vspace{-0.2in}
\caption{Dataset}
\end{table}

Q2: Why we need to distinguish the same entities using {\em key yes} and {\em distinct yes}) to identify the same observation? 
For the example, we have one observation for splot {\em A}.
What's the semantic purpose of this? What kind of query may need this? 
E.g., how many spots in this dataset?

{\bf Note 1}: if one observation type is marked with {\em distinct yes}, all its measurements should be marked with {\em key yes}. 
Otherwise, we may have the same observation with different measurement values. 
E.g., what will happen for the following:

\noindent
{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
\verb|    |{\bf  entity} ``Plot''\\
\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
\verb|        | {\bf characteristic} ``EntityName'' \\
\verb|        | {\bf standard} ``Nominal''\\
\verb|    | {\bf measurement} ``m2''\\
\verb|        | {\bf characteristic} ``area'' \\
\verb|        | {\bf standard} ``sqft''\\

Will $(A, 1.0)$ and $(A,1,1)$ be treated as the same observation? 
According to the semantic meaning, they are the same observation because they have the same value on the key measurement and this observation type is marked witht {\em distinct yes}. 
However, there is something wrong here.

\begin{table}[htb]
\begin{center}

\end{center}
\vspace{-0.2in}
\caption{Dataset}
\end{table}

Based on this note, it seems like it is not useful to denote {\em distinct yes}.
Basically, once all the measurements are marked with {\em key yes}, it automatically infers that it is distinct yes.

\end{document}

### 2010-01-22 docs/algdraft/Materialize_DB.tex
Updated materialization algorithm.

--- 2010-01-11 docs/algdraft/Materialize_DB.tex
@@
 \setcounter{totalnumber}{5}
+
 %\setlength{\parindent}{0pt}
-%\pdfpagewidth 8.5in
-%\pdfpageheight 11in
+\setlength\topmargin{0in}
+\setlength\headheight{0in}
+\setlength\headsep{0in}
+\setlength\textheight{8.5in}
+\setlength\textwidth{6.5in}
+\setlength\oddsidemargin{0in}
+\setlength\evensidemargin{0in}
 
+\newtheorem{example}{Example}[section]
 \begin{document}
 
-Note that in our description, we use
-$A.*$ to represent the annotation information.
+\section{Valid annotation constraints}
+Among the {\em key}, {\em distinct}, and {\em identifying} constraints,
+there are correlations between ``key yes'' and ``distinct yes''.
+\begin{itemize}
+\item ``Distinct yes'' on an observation type implies that its measurements collectively form a key of this observation type.
+\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
+In this case, when two rows (for the observation type) have the same values for these key measurements, they represent the different observation instances.
+It means that there may be other ``variables'' that can distinguish one observation instance from the other, but this dataset does not catch this variable ({\bf HP: is ``variable'' the right word to use?
+}).
+So, in real application, we do not encourage such situation to appear.
+\end{itemize}
+
+An example with invalid annotation: given the data in Table \ref{tb:pltarea}. The following annotation is {\em invalid}.
+According to the annotation, if two plots have the same value on EntityName, they represent the same plot observation. 
+Obviously, it will have probelem to interprete the data in Table \ref{tb:pltarea} with this annotatioin. 
+E.g., the first and second row catching plot with EntityName $A$ should be the same plot obervation.
+However, the data shows that this one plot has two different areas $1.0$ and $1.1$, so, there is confusion here.
+\\
+
+\noindent
+{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
+\verb|    |{\bf  entity} ``Plot''\\
+\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
+\verb|        | {\bf characteristic} ``EntityName'' \\
+\verb|        | {\bf standard} ``Nominal''\\
+\verb|    | {\bf measurement} ``m2''\\
+\verb|        | {\bf characteristic} ``area'' \\
+\verb|        | {\bf standard} ``sqft''\\
+\verb|       |$\cdots$ Not finished....
+
+\begin{table}[htb]
+\begin{center}
+%\begin{tabular}{cc}
+%\begin{tabular}{|l|l|l|}
+%\hline
+%plt & spp & dbh\\\hline
+%A & piru & 35.8 \\\hline
+%A & piru & 36.2 \\\hline
+%B & piru &33.2 \\\hline
+%B&abba&34\\\hline
+%\end{tabular}
+%&
+\begin{tabular}{|l|l|l|l|}
+\hline
+plt & area & spp & dbh\\\hline
+A & 1.0 & piru & 35.8 \\\hline
+A & 1.1 & piru & 36.2 \\\hline
+B & 2.0 &piru &33.2 \\\hline
+B& 2.0 & abba&34\\\hline
+\end{tabular}\\
+%(a) & (b)
+%\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset}
+\label{tb:pltarea}
+\end{table}
+
+In our algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the correlation rules among these constraints.
+
+{\bf From HP:} More to come on this part...
+
+\section{Data materialization algorithm}
+The {\bf MaterializeDB} algorithm catches the {\em key}, {\em distinct} and {\em identifying} constraints in the annotation during the materialization process. 
+The input of the algorithm is the $Dataset$ and the annotations $A.*$ on it. 
+Each row in the input dataset represents the information related to one or more observations and their contexts.
+The input $A.*$ represents the annotation information.
 More specifically,
 \begin{itemize}
-\item MeasType = $\{\langle \underline{MeasTypeId}, ObsTypeId, CharType, StdType, ProtType, Precision, Value, isKey \rangle\}$;
-\item ObservationType = $\{\langle \underline{ObsTypeId}, EntTypeId, isUnique \rangle\}$ \\
-	{\bf HP Question}: I did not use ``AnnotId'' in the algorithm, so I remove it here. How shall this be used?
-\item ContextType = $\{\langle \underline{ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$\\
-	{\bf HP Question}: Add the {\em isIdentify} for the purpose of checking whether we need to use one observation's context for key, is this ok?
-\item Map = $\{\langle \underline{MeasTypeId, ObsTypeId, Cond}, Val\rangle\}$
+\item A.MeasType = $\{\langle \underline{MeasTypeId}, ObsTypeId, CharType, StdType, ProtType, Precision, isKey \rangle\}$;
+\item A.ObservationType = $\{\langle \underline{ObsTypeId}, EntTypeId, isDistinct \rangle\}$ \\
+	Note that, we hide the ``AnnotId'' in this schema, which denotes the resource the annotation is on.
+    We do not include this in the algorithm because the algorithm focuses on dealing with annotations on one resource.
+    This way, we can simplify the description.
+\item A.ContextType = $\{\langle \underline{ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$
+\item A.Map = $\{\langle \underline{MeasTypeId, ResAttribute, Cond}, Val\rangle\}$
 \end{itemize}
 
-$OBOE.*$, on the other hand, represent the {\em OUTPUT} data represented in the OBOE model. In detail,
+The output of the algorithm is a set of materialized tables represented in the OBOE model and denoted by $OBOE.*$. In detail,
 \begin{itemize}
-\item $Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances materialized from $Dataset$.\\
-	{\bf HP Question}: I did not use ``AnnotId'' in the algorithm, so I remove it here. How shall this be used?
-\item $Measurenebt=\{\langle \underline{MeasId}, ObsId, Characteristic, Val\rangle\}$\\     
-	%{\bf HP Question}: I changed ``MeasType'' in the ppt to ``Characteristic'', is it right? It is consistent with the example.
-         \textcolor{blue}{Changed to: }\\
-	$Measurenebt=\{\langle \underline{MeasId}, \textcolor{blue}{MeasTypeId}, ObsId, Val\rangle\}$\\   
-          {\bf HP note}: the other information about the measurement type, e.g., standard, characteristic, can be gotten using {\em MeasTypeId}.
-         In  real application, we can think of "duplicating the measurement type" information. Too detail. Ignore here. 
-\item $Entity=\{\langle \underline{EntId}, EntType\rangle\}$
-\item $Context =\{\langle \underline{ObsId, ConextObsId, RelType}\rangle\}$\\
-	{\bf HP Question}: I did not use ``Relationship" table, do we really need to instantiate this information?
+\item $OBOE.Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances materialized from $Dataset$.
+\item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$
+\item $OBOE.Entity=\{\langle \underline{EntId}, EntType\rangle\}$
+\item $OBOE.Context =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$
 \end{itemize}
 
-Each row in the input dataset represents the information related to one observation and its contexts.
+Algorithm \ref{alg:matdb} shows the framework of our algorithm. 
+In this algorithm, $ObsIdx$ is the index structure maintained for the observation instances 
+whose types are specified with {\em distinct yes}. 
+This index maintains the mapping from observation type and key values to its corresponding observation instance id. 
+Obviously, only when observation type is specified with {\em distinct yes}, i.e., we want to keep track of the same observation instances, we need to maintain their instances in this index. 
+The key values can be calculated in three different cases.
+In the first case where just one measurement of an observation type is specified with ``key yes'' the key value is the value of this ``key'' measurement.
+In the second case, several measurement types are marked with ``key yes'', the key value is the combined value of these several measurement.
+In the third case, this object type is marked with ``identifying yes'', the key value is the combined value of its context observation's key measurement values.
 
-The algorithm tries to catch the {\em key}, {\em unique} and {\em identifying} constraints in the annotation during the materialization process.
+Similarly, in case that some measurement type(s) of an observation type is/are specified with ``key yes'', 
+if different observation instances have the same value on these key measurements, semantically, we interpret that they are of the same entity.
+$Entidx$ is the index structure for tracking the distinct entities.
+The key value is computed in the same way as that for an observation instance.
 
-%\newpage
+The procedure of this algorithm is very straightforward. 
+It processes the dataset in a row-wise manner. 
+In dealing with each row, five steps are involved. 
+The first step generates orphan measurement instances which are not connected any observation instances.
+The second step group these measurement instances according to their observation types.
+Then, for different observation types, we materialize entity instances and observation instances by either creating a new one or return an existing one.
+The last step assigns the context relationship among the different observation instances.
+
 \begin{algorithm} [htb]
-\caption{{\bf MaterializeDB} ($Dataset, A$)}
+\caption{{\bf MaterializeDB} ($Dataset, A.*$)}
+\label{alg:matdb}
 {\small
 \begin{algorithmic}
-%\STATE{{\bf MaterializeDB} ($Dataset, A$)}
-\STATE /* {\em Dataset}: [Input] in the form of flat file */\\
+\STATE /* {\em Dataset}: [Input] in the form of a flat file */\\
 \STATE /* {\em A }:  [Input] Annotations*/
 \STATE
-\STATE $ObsIdx =\emptyset$;  /* for keeping index $\langle ObsTypeId, \textcolor{blue}{KeyVal}\rangle \rightarrow ObsId$*/
-\STATE $EntIdx = \emptyset$;  /* for keeping index $\langle MeasTypeId, \textcolor{blue}{KeyVal}\rangle \rightarrow EntId$*/
+\STATE $ObsIdx =\emptyset$;  /* Keep index $\langle ObsTypeId, KeyVal\rangle \rightarrow ObsId$*/
+\STATE $EntIdx = \emptyset$;  /* Keep index $\langle ObsTypeId, KeyVal\rangle \rightarrow EntId$*/
+\STATE
 \FOR { (each $Row\langle A1, A2, \cdots, An\rangle \in Dataset$)}
-	%\STATE MapRow($Row, MapList, ObsIdx, EntIdx$);
-    \STATE $MeasSet = \emptyset$; /* Keep the set of new measurement instances*/
-
+    \STATE /* Step 1: Define measurement instances */\\
+            $MeasSet$ = CrtMeasurement($Row, A.*$);
     \STATE
-    \STATE/* 1. Create new orphan measurement instances*/
-	\FOR {(each $m = \langle MeasTypeId, \textcolor{blue}{ResAttribute}, Cond, Val\rangle \in A.Map$)}
-
-		\IF{\textcolor{blue}{($m.ResAttribute != Row.Ai.Attrname$) OR} ($Row.Ai$ does not satisfy $m.Cond$)}
-		\STATE cotinue;
-		\ENDIF
-		\STATE $ObsTypeId$ = GetObsTypeId ($A.MeasType, m.MeasTypeId$);
+    \STATE /* Step 2: Partition the measurement instances according to observation types*/\\
+        $ObsType2MeasIdx$ = PartMeas($MeasSet, A.*$);
     \STATE
-		\STATE $MeasId$ = GetNewMeasId($OBOE.Measurement$);
-		%\STATE $Cha$ = GetCharacteristic4Meas($A.MeasType, m.MeasTypeId$);
-		%\STATE Create a measurement instance $mi = \langle MeasId, null, Cha, m.val\rangle$;		
-		\STATE \textcolor{blue}{{\bf if}($m.Val!=NULL$) $MeasVal$ = $m.Val$; }
-		\STATE \textcolor{blue}{{\bf else} $MeasVal$ = $Row.Ai.Val$;}
-		\STATE Create a measurement instance $mi = \langle MeasId, MeasTypeId, null, MeasVal\rangle$;
-		\STATE Add $mi$ to $MeasSet$;
-	\ENDFOR
-
+    \STATE $ContextIdx = \emptyset$;  /* Keep index $ObsTypeId \rightarrow ObsId$ to materialize context*/
+    \FOR{(each $ObsTypeId \in ObsType2MeasIdx.keys$)}
+      \STATE /* Step 3: Find or create the entity instance for each observation type partition */
+        \STATE $EntId$ = MaterializeEntity($ObsTypeId, ObsType2MeasIdx, EntIdx, A.*, OBOE.*$);
         \STATE
-    \STATE /* 2. Get observation types and measurement types with new instances*/
-    \STATE $KeyIdx=\emptyset$ /* Keep index for $ObsTypeId \rightarrow \{mi\}$ */
-    \FOR{(each $mi \in MeasSet$)}
-    \STATE $ObsTypeId$ = GetObsTypeId ($A.MeasType, mi.MeasTypeId$);
-	\STATE Update $KeyIdx$ by changing the item $ObsTypeId \rightarrow \{mi\}$;
-    \ENDFOR
-
-    \STATE
-    \FOR{(each $ObsTypeId \in KeyIdx.keys$)}
-        \STATE /*Get the key value for this observation. \\
-                   Case 1: generally, it is the value for the ``key'' measurement. \\
-                   Case 2: several measurement types are marked with ``key yes'', the key value is the combined value of these several measurement. \\
-                   Case 3: this object type is marked with ``identifying yes, the key value '' is the combined value with its context observation's key measurement values. */
-        \STATE $KeyVal$ = GetObsTypeKeys ($ObsTypeId, KeyIdx$);
-
-        \STATE
-        \STATE /* 3. Get an existing or create a new entity instance*/
-        \STATE $HasKey$ = false;
-        \IF{($ObsTypeId$ has key measurements)}
-            \STATE $HasKey$ = true;
-        \ENDIF
-    	\STATE $EntId$ = {\bf MaterializeEntity}($ObsTypeId, HasKey, KeyVal, EntIdx, A, OBOE$)
-
-        \STATE
-        \STATE/* 4. Get an existing or create a new observation instance*/
-         \STATE $ContextIdx = \emptyset$;  /* keep index $ObsTypeId \rightarrow ObsId$ to materialize context*/
-        \FOR{(each $ObsTypeId \in KeyIdx.keys)$}
-        \STATE $IsObsUnique$ = checkIfObsUnique($A.ObservationType, ObsTypeId$);
-        \STATE $ObsHasKey$ = $HasKey~\&\& IsObsUnique$ ;
-        \STATE $ObsId$ = {\bf MaterializeObs}($ObsTypeId, ObsHasKey,KeyVal,EntId,ObsIdx, OBOE$);
-
-        \STATE
-        \STATE /*Maintain the measurement instances for this observation*/
-        \STATE $miSet$ = GetMeasInst($KeyIdx, ObsTypeId$);
-        \IF{($ObsId$ is a new one)}
-            \STATE Set the $obsId$ to each $mi \in miSet$ so that $mi$-s are not orphans;
-        \ELSE
-            \STATE Discard all the $mi \in miSet$;
-        \ENDIF
-        \STATE Put all the $mi \in miSet$ to $OBOE.Measurement$;
-        \STATE
-
-    	\STATE $ContextIdx = ContextIdx \cup \{ObsTypeId \rightarrow ObsId\}$;
-        \ENDFOR
+        \STATE /* Step 4 Find or create the observation instance for each observation type partition */
+        \STATE MaterializeObs($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, OBOE.*$);
     \ENDFOR
     \STATE
-	\STATE /*5. Materialize context relationships among observation instances*/
-	\STATE {\bf MaterializeContext}($ContextIdx, A, OBOE$)
+    \STATE /* Step 5: Assign the context observation instances */  \\
+            MaterializeContext($ContextIdx, A.*, OBOE.*$);
 \ENDFOR
-\STATE Return OBOE;
+\RETURN OBOE;
+\end{algorithmic}
+}
+\end{algorithm}
+
+{\bf Analysis of MaterializeDB}. \\
+{\bf Time}: As can be seen that  this algorihm scans the original data file in a row-by-row manner without revisiting the already seen rows. So, it is linear in the size of the dataset. 
+ We use $EntIdx$ and $ObsIdx$ to facilitate the checking of unique entity and observation instances. Let $m$ be the number of unique keys, each checking could take O(log(m)) time. 
+So, in total, the algorithm runs in O(n log(m)) where $m\ll n$ generally. \\
+{\bf Space}: $EntIdx$ and $ObsIdx$ are the intermdedite structures that we use in the algorithm. Since they keep the unique entity and observation key values. The space complexity is of $O(m)$.
+
+\begin{algorithm} [htb]
+\caption{{\bf CrtMeasurement} ($Row, A.*$)}
+{\small
+\begin{algorithmic}
+    \STATE/* Create new orphan measurement instances*/
+    \STATE $MeasSet = \emptyset$; /* Keep the set of new measurement instances*/
+	\FOR {(each $m = \langle MeasTypeId, ResAttribute, Cond, Val\rangle \in A.Map$)}
+
+		\STATE {\bf if}{(($m.ResAttribute != Row.Ai.Attrname$) OR ($Row.Ai$ does not satisfy $m.Cond$))}
+		  {\bf continue};
+		%\STATE $ObsTypeId$ = GetObsTypeId($A.MeasType, m.MeasTypeId$);
+		%\STATE
+		\STATE $mi_{id}$ = GetNewMeasId($OBOE.Measurement$);
+		\STATE {\bf if} ($m.Val!=NULL$) $MeasVal$ = $m.Val$;
+		\STATE {\bf else} $MeasVal$ = $Row.Ai.Val$;
+		\STATE Create a measurement instance $\langle mi_{id}, null, MeasType, MeasVal\rangle$ and add it to $MeasSet$;
+	\ENDFOR
+    \RETURN $MeasSet$
 \end{algorithmic}
@@
 \begin{algorithm} [htb]
-\caption{{\bf MaterializeEntity}($ObsTypeId, HasKey, KeyVal, EntIdx, A, OBOE)$}
+\caption{{\bf PartMeas} ($MeasSet, A.*$)}
+{\small
 \begin{algorithmic}
+    \STATE /* Partition measurement instances according to their observation types*/
+    \STATE $ObsType2MeasIdx=\emptyset$ /* Keep index for $ObsTypeId \rightarrow \{mi\}$ */
+    \FOR{(each $mi \in MeasSet$)}
+    \STATE $ObsTypeId$ = GetObsTypeId ($A.MeasType, mi.MeasTypeId$);
+	\STATE Update $ObsType2MeasIdx$ by changing the item $ObsTypeId \rightarrow \{mi\}$;
+    \ENDFOR
+    \RETURN $ObsType2MeasIdx$;
+\end{algorithmic}
+}
+\end{algorithm}
+
+\begin{algorithm} [htb]
+\caption{{\bf MaterializeEntity}($ObsTypeId, ObsType2MeasIdx, EntIdx, A.*, OBOE.*)$}
+{\small
+\begin{algorithmic}
+\STATE $KeyVal$ = GetObsTypeKeys ($ObsTypeId, ObsType2MeasIdx$);
+\STATE $HasKey$ = false;
+\STATE {\bf if }{($ObsTypeId$ has key measurements OR is specified with distinct yes)}
+     $HasKey$ = true;
 \STATE $EntType$ = GetObsEntityType ($A.ObservationType, ObsTypeId$)
@@
 	\STATE $EntId$ = GetEntId($ObsTypeId, KeyVal, EntIdx$); %/*From the current entity key index get the id of the entity instance*/
-	\IF{($EntId$ is not Null)}
-		\STATE $CrtNewEntInst$ = false;
-	\ENDIF
+	\STATE {\bf  if} {($EntId!=NULL$)} $CrtNewEntInst$ = false;
 \ENDIF
@@
 	\STATE $EntId$ = CrtEntId($EntType$);
-	\STATE Create an entity instance $ei = \langle EntId, EntType\rangle $;
-	\STATE Put $ei$ to $OBOE.Entity$;
-	\IF{$HasKey$==true)}
-        \STATE /*Only when this is the key measurement, we need to maintain the index*/
-        \STATE $EntIdx = EntIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow EntId$\};
+	\STATE Create an entity instance $ei = \langle EntId, EntType\rangle $ and put $ei$ to $OBOE.Entity$;
+	\STATE {\bf if} ($HasKey$==true) $EntIdx = EntIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow EntId$\};
 \ENDIF
-\ENDIF
-\STATE return $EntId$;
+\RETURN $EntId$;
 \end{algorithmic}
+}
 \end{algorithm}
@@
 \begin{algorithm} [htb]
-\caption{{\bf MaterializeObs}($ObsTypeId, HasKey,KeyVal,EntId, ObsIdx, OBOE$)}
+\caption{{\bf MaterializeObs}($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, OBOE.*$)}
+{\small 
 \begin{algorithmic}
+    \STATE $KeyVal$ = GetObsTypeKeys ($ObsTypeId, ObsType2MeasIdx$);
+    \STATE $IsObsDistinct$ = CheckIfObsDistinct($A.ObservationType, ObsTypeId$);
 		\STATE $CrtNewObsInst$ = true;
-		\IF{ ($HasKey$==true) }
-			\STATE $ObsId$ = GetObsId($ObsTypeId, KeyVal, ObsIdx$); %/*From the current observation key index get the id of the observation instance*/
-			\IF{($ObsId$ is not Null)}
-				\STATE $CrtNewObsInst$ = false;
-			\ENDIF
+	\IF{ ($IsObsDistinct$==true) }
+        %/*From the current observation key index get the id of the observation instance*/
+		\STATE $ObsId$ = GetObsId($ObsTypeId, KeyVal, ObsIdx$);
+		\STATE {\bf if}{($ObsId!=NULL$)}  $CrtNewObsInst$ = false;
 	\ENDIF
 	\IF{($CrtNewObsInst==true)$}
-    		\STATE Create an observation instance $oi = \langle ObsId,  EntId\rangle $
-    		\STATE Put $oi$ to $OBOE.Observation$;
-    		\IF{($HasKey$==true)}
-                \STATE /*Only when it has key measurement and it is identified as unique, we need to maintain the index*/
-                \STATE $ObsIdx = ObsIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow ObsId$\};
+		\STATE Create an observation instance $oi = \langle ObsId,  EntId\rangle$ and put $oi$ to $OBOE.Observation$;
+		\STATE {\bf  if} ($IsObsDistinct$==true) $ObsIdx = ObsIdx \cup \{\langle ObsTypeId, KeyVal \rangle \rightarrow ObsId$\};
            \ENDIF
-		\ENDIF
-		\STATE Return $ObsId$;
+    \STATE
+    \STATE /*Maintain the measurement instances for this observation*/
+    \STATE $miSet$ = GetMeasInst($ObsType2MeasIdx, ObsTypeId$);
+    \STATE {\bf if} {($ObsId$ is a new one)}
+         Set the $obsId$ to each $mi \in miSet$ so that $mi$-s are not orphans;
+    \STATE {\bf else} 
+         Discard all the $mi \in miSet$;
+
+    \STATE Put all the $mi \in miSet$ to $OBOE.Measurement$;
+    \STATE
+
+	\STATE $ContextIdx = ContextIdx \cup \{ObsTypeId \rightarrow ObsId\}$; /* $ContextIdx$ is also output*/
+
 \end{algorithmic}
+}
 \end{algorithm}
@@
 \begin{algorithm} [htb]
-\caption{{\bf MaterializeContext}($ContextIdx, A, OBOE$)}
+\caption{{\bf MaterializeContext}($ContextIdx, A.*, OBOE.*$)}
+{\small 
 \begin{algorithmic}
@@
 \STATE $ContextObsTypeId, Rel$ = GetContextObsTypeRel($A.ContextType, ObsTypeId$);
-\IF{($ContextObsTypeId$ is not Null)}
+\IF{($ContextObsTypeId!= NULL$)}
 	\STATE $ContextObsId$ = GetContextObsId($ContextIdx, ContextObsTypeId$);
-	\STATE Create a context $c=\langle ObsId, ContextObsId, Rel\rangle$;
-	\STATE Put $c$ to $OBOE.Context$;
+	\STATE Create a context instance $ci=\langle ObsId, ContextObsId, Rel\rangle$;
+	\STATE Put $ci$ to $OBOE.Context$;
 \ENDIF
@@
 \end{algorithmic}
+}
 \end{algorithm}
 
-We can run the example in page 6 in the powerpoint file to explain the algorithm.
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{|l|l|l|l|}
+\hline
+yr & spec & spp & dbh\\\hline
+2007 & 1 & piru & 35.8 \\\hline
+2008 & 1 & piru & 36.2 \\\hline
+2008 & 2 & abba & 33.2 \\\hline
+\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset 1}
+\label{tb:dataset1}
+\end{table}
 
-For Row(2007, 1, piru, 35.8)
+\begin{example} [Example with ``key yes'' and ``distinct yes'', without ``identifying yes'']
+Take the data in Table \ref{tb:dataset1} \footnote{This is in page 6 in the powerpoint file.}
+as an example to explain the algorithm. \\
+For {\bf Row(2007, 1, piru, 35.8)}
 \begin{itemize}
-    \item Create a measurement $mi_1=\langle meas1, null, Year, 2007\rangle$
-    \item Create a measurement $mi_2=\langle meas2, null, Year, 35.8\rangle$
-    \item Create a measurement $mi_3=\langle meas3, null, TaxonomicTypeName, Picea~rubens\rangle$
-    \item Create a measurement $mi_4=\langle meas4, null, EntityName, 1\rangle$
-    \item Get $MeasSet = \{mi_1, mi_2, mi_3, mi_4\}$;
-    \item Step 2: Get $KeyIdx = \{o_1\rightarrow \{mi_1\}, o_2 \rightarrow \{mi_2, mi_3, mi_4\}\}$
-    \item Step 3-4: for each observation types $o_1$ and $o_2$
+\item Step 1 creates four measurement instances: 
+$\langle mi_1, null, Year, 2007\rangle$, $\langle mi_2, null, DBH, 35.8\rangle$, \\
+$\langle mi_3, null, TaxonomicTypeName, Picea~rubens\rangle$, $\langle mi_4, null, EntityName, 1\rangle$,\\
+and returns $MeasSet = \{mi_1, mi_2, mi_3, mi_4\}$;
+\footnote{For all the instances, the measurement characteristic is set to represent Measurement Type}.
+
+\item Step 2 returns $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_1\}, o_2 \rightarrow \{mi_2, mi_3, mi_4\}\}$. 
+ \item Step 3-4: for each observation types $o_1$ and $o_2$,  materialize entity and observation instance
+    \begin{itemize}
+    \item for $o_1$ (with associated instance $mi_1$ of type $m_1$
+    \begin{itemize}
+        \item Since $m_1$ is specified as key, get the $KeyVal=2007$;
+        \item No entity with this key exists in $EntIdx$, create an entity $\langle ei_1, TemporalRange\rangle$; Now, $EntIdx=\{\langle o_1, 2007\rangle\rightarrow ei_1\}$.
+        \item Since $o_1$ is specified as {\em distinct}, need to make sure we do not create redundant observation instances.
+No entry with the key $\langle o_1, 2007\rangle$ exists in $ObsIdx$, so, create an observation instance $oi_1$, which is of entity $ei_1$ and represented as $\langle oi_1, ei_1\rangle$.
+\\
+	Now, $ObsIdx=\{\langle o_1, 2007 \rangle \rightarrow oi_1\}$
+        \item Connect $mi_1$ to $oi_1$;
+    \end{itemize}
+
+    \item When deal with $o_2$,
+        \begin{itemize}
+        \item $KeyVal=1$.
+        \item Create an entity instance $\langle ei_2, Tree\rangle$; $EntIdx=\{\langle o_1, 2007\rangle\rightarrow ei_1, \langle o_2, 1\rangle\rightarrow ei_2\}$.
+        \item Create an observation $\langle oi_2, ei_2 \rangle$.
+            No need to update $ObsIdx$ because $o_2$ is not identified as {\em distinct}.
+        \item Connect $mi_2$, $mi_3$ and $mi_4$ to $oi_2$;
+        \end{itemize}
+    \end{itemize}
+    \item Step 5 assigns the context relationship between $oi_1$ and $oi_2$;
+    \end{itemize}
+
+For {\bf Row (2008, 1, piru, 36.2)}
+    \begin{itemize}
+    \item Step 1 creates measurement instances $\langle mi_5, null, Year, 2007\rangle$,
+    $\langle mi_6, null, DBH, 35.8\rangle$,\\
+    $\langle mi_7, null, TaxonomicTypeName, Picea~rubens\rangle$,
+    $\langle mi_8, null, EntityName, 1\rangle$\\
+    and returns $MeasSet = \{mi_5, mi_6, mi_7, mi_8\}$;
+    \item Step 2 gets $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_5\}, o_2 \rightarrow \{mi_6, mi_7, mi_8\}\}$
+    \item Step 3-4: for each observation types $o_1$ and $o_2$ materialize entity and observation instance
+        \begin{itemize}
             \item for $o_1$
             \begin{itemize}
-        \item Since $m_1$ is specified as key, get the $KeyVal=2007$;
-        \item No entity with this key exists in $EntIdx$, create an entity $ei=\langle ent1, TemporalRange\rangle$; Now, $EntIdx=\{\langle o1, 2007\rightarrow ent1\rangle\}$.
-        \item Since $o_1$ is specified as {\em distinct}, need to make sure we do not create redundant observations. No entry with this key exists in $ObsIdx$, so, create an observation
-        $oi = \langle obs1, ent1\rangle$. Now, $ObsIdx=\{\langle o1, 2007 \rangle \rightarrow obs1\}$
+            \item  $KeyVal=2008$;
+            \item Create an entity instance $\langle ei_3, TemporalRange\rangle$; \\
+                $EntIdx=\{\langle o_1, 2007 \rangle \rightarrow ei_1, \langle o_2, 1 \rangle \rightarrow ei_2, \langle o_1, 2008 \rangle \rightarrow ei_3\}$.
+             \item Create an observation instance $\langle oi_3, ei_3 \rangle$;\\
+              $ObsIdx=\{\langle o_1, 2007\rangle\rightarrow oi_1, \langle o_2, 1\rangle\rightarrow oi_2, \langle o_1, 2008\rangle\rightarrow oi_3\}$
+		\item Connect $mi_5$ to $oi_3$; 
             \end{itemize}
@@
              \item $KeyVal=1$.
-        \item Create an entity $ei=\langle ent2, Tree\rangle$; $EntIdx=\{\langle o1, 2007\rightarrow ent1\rangle, \langle o2, 1\rightarrow ent2\rangle\}$.
-        \item Create an observation $\langle obs2, ent2 \rangle$. No need to update $ObsIdx$ because $o2$ is not identified as {\em unique}.
+            \item {\bf item $\langle o_2, 1 \rangle \rightarrow ei_2$ is already in $EntIdx$, so get the entity id $ei_2$. No need to create an entity.}
+            \item Since $o_2$ is not specified with {\em distinct yet}, we NEED to create an observation $\langle oi_4, ei_2 \rangle$. No need to update $ObsIdx$.
+		\item Connect $mi_6, mi_7, mi_8$ to $oi_4$; 
             \end{itemize}
-    \item Connect $mi_1$ to $obs1$;
-    \item Connect $mi_2$, $mi_3$ and $mi_4$ to $obs2$;
-    \item Set the context relationship between $obs1$ and $obs2$;
+        \end{itemize}
+    \end{itemize}
+For {\bf ROW (2008, 2, abba, 33.2)}
+    \begin{itemize}
+        \item For $o_1$'s measurement $2008$, 
+	\begin{itemize}
+	\item Since $\langle o_1, 2008 \rangle \rightarrow ei_3$ already exists in $EntIdx$, {\bf no need to create a new entity}.
+        \item Since $o_1$ is specified with {\em distinct yes}, and $\langle o_1, 2008 \rangle \rightarrow oi_3 $ already exists in $ObsIdx$, 
+	{\bf no need to create a new OBSERVATION} and no need to put the measurement instance for $2008$ into OBOE model.
+	\end{itemize}
     \end{itemize}
 
-For Row (2008, 1, piru, 36.2)
+\end{example}
+
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{|l|l|l|}
+\hline
+plt & spp & dbh\\\hline
+A & piru & 35.8 \\\hline
+A & piru & 36.2 \\\hline
+B & piru &33.2 \\\hline
+\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset 2}
+\label{tb2}
+\end{table}
+
+\begin{example} [Example with identifying]\label{eg2}
+Let us use the data in Table \ref{tb2} as an example.
+Before we go through the algorithm step by step,  we first note that  $o_1$ and $o_2$ have key measurements $m1$ and $m_2$ respectively. 
+So, we need to maintain the distinct entity instances for both of these two observation types. 
+In addition, $o_1$ is specified with {\em distinct yes} while $o_2$ is not. So, we need to maintain the distinct observation instances for $o_1$ but not for $o_2$.
+
+For the {\bf first row}, 
 \begin{itemize}
-    \item Create a measurement $mi_5=\langle meas5, null, Year, 2007\rangle$
-    \item Create a measurement $mi_6=\langle meas6, null, Year, 35.8\rangle$
-    \item Create a measurement $mi_7=\langle meas7, null, TaxonomicTypeName, Picea~rubens\rangle$
-    \item Create a measurement $mi_8=\langle meas8, null, EntityName, 1\rangle$
-    \item Get $MeasSet = \{mi_5, mi_6, mi_7, mi_8\}$;
-    \item Step 2: Get $KeyIdx = \{o_1\rightarrow \{mi_5\}, o_2 \rightarrow \{mi_6, mi_7, mi_8\}\}$
-    \item for type $o_1$
+\item The first step generates three measurement instances $MeasSet=\{\langle mi_1, null, EntityName, A\rangle$, \\
+$\langle mi_2, null, TaxonomicTypeName, Picea~rubens\rangle$, $\langle mi_3, null, DBH, 35.8\rangle\}$.
+\item The second step gets $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_1\}, o_2 \rightarrow \{mi_2, mi_3\}\}$.
+\item For each observation type, create entity and observation instances.
 \begin{itemize}
-        \item  $KeyVal=2008$;
-        \item Create an entity $\langle ent3, TemporalRange\rangle$; $EntIdx=\{\langle o1, 2007 \rangle \rightarrow ent1, \langle o2, 1 \rangle \rightarrow ent2, \langle o1, 2008 \rangle \rightarrow ent3\}$.
-         \item Create an observation $\langle obs3, ent3 \rangle$, and $ObsIdx=\{\langle o1, 2007\rightarrow obs1\rangle, \langle o2, 1\rightarrow obs2\rangle, \langle o1, 2008\rightarrow obs3\rangle\}$
+\item For $o_1$, the key value is $A$. Since there is no such a key in $EntIdx$, we create an entity $ei_1$ of type $Plot$.
+	$EntIdx = \{\langle o_1, A\rangle \rightarrow ei_1\}$. \\
+	We create an observation instance $oi_1$ whose entity is $ei_1$.\\
+	$ObsIdx = \{\langle o_1, A\rangle \rightarrow oi_1\}$.
+	Connect the measurement instances to observation instances  $mi_1 \rightarrow oi_1$,
+\item For $o_2$, 
+	the key value is $(A, Picea~rubens)$ since it has context $o_1$ with ``{\em identifying yes}'',  we create an entity instance $ei_2$ of type $Tree$. \\
+	$EntIdx = \{\langle o_1, A\rangle \rightarrow ei_1, \langle o_2, (A, Picea~rubens) \rangle \rightarrow ei_2\}$.\\
+	We  create an observation instance $oi_2$ whose entity is $ei_2$.\\
+	Connect the measurement instances $mi_2$ and $mi_3$  to observation instance $oi_2$.
 \end{itemize}
-    \item When deal with $o_2$,
+\item The last step for this row is to connect the observations using context relationship. For this instance, we connect $oi_1$ to $oi_2$ with context ``{\em Within}''.
+\end{itemize}
+
+For the {\bf second row}, 
 \begin{itemize} 
-         \item $KeyVal=1$.
-        \item {\bf Some item $\langle o2, 1 \rangle \rightarrow ent2$ is already in $EntIdx$, so get the entity id $ent2$. No need to create an entity.}
-        \item Since $o2$ is not specified with {\em unique yet}, we NEED to create an observation $\langle obs4, ent2 \rangle$. No need to update $ObsIdx$.
+\item The first step defines three measurement instances $MeasSet=\{\langle mi_4, null, EntityName, A\rangle$, \\
+$\langle mi_5, null,TaxonomicTypeName, Picea~rubens\rangle, \langle mi_6, null, DBH, 36.2\rangle\}$.
+\item The second step gets $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_4\}, o_2 \rightarrow \{mi_5, mi_6\}\}$.
+\item For each observation type, create entity and observation instances.
+\begin{itemize}
+\item For $o_1$, the key value is $\langle o_1, A\rangle$, $EntIdx$ already has an item for it with entity instance $ei_1$. No need to create a new instance for it. \\
+To create observation instance, since $o_1$ is specified with ``{\em distinct yes}'' and the key value is $\langle o_1,A\rangle$, 
+which corresponds to observatin instance $oi_1$. So, we do not need to create a new observation for it.\\
+When we try to connect the measurement instance $mi_4$ to observation instance,  we realize that we did not create a new observation instance for type $o_1$. 
+So its related measurement instance $mi_4$ can be discarded. 
+\item For $o_2$, the new key value is $\langle o_2, (A, Picea~rubens)\rangle$ , which corresponds to $ei_2$ in $EntIdx$, so no need to create a new instance for it either. \\
+To create observation instances, since no ``{\em distinct yes}'' is specified, we create a new observation instance $oi_3$ for it.
+Then, we connct the measurement instances $\{m_5, m_6\}$ to observation instance $oi_3$.
 \end{itemize}
 \end{itemize}
-For ROW (2008, 2, abba, 33.2)
-    \begin{itemize}
-        \item For $o1$'s measurement $2008$, since $\langle o1, 2008 \rangle \rightarrow ent3$ already exists in $EntIdx$, {\bf No need to create a new ENTITY}.
-        \item Since $o1$ is specified with {\em unique yes}, and since $\langle o1, 2008 \rangle \rightarrow obs3 $ already exists in $ObsIdx$, {\bf No need to create a new OBSERVATION} and no need to put the measurement for $2008$ into OBOE model.
-    \end{itemize}
+
+When we process the {\bf third row}, we have a new key value $\langle o_1, B\rangle$ for $o_1$, thus we create a new entity instance for it.
+For $o_2$, we have new key value $\langle o_2, (B,~Picea~rubens)\rangle$ and create a new entity instance for it.
+
+\end{example}
 

### 2010-01-28 docs/algdraft/UseCaseKey.tex
update the use cases for key yes, distinct yes and identifying.
--

@@
 
+Mark and Huiping had a discussion about use cases to use ``key yes'', ``distinct yes'' and ``identifying yes''.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+Plot & sub-plot & Tmnt & Sp & Ind &wt\\\hline
+1 & A & X &Aus&1&10\\\hline
+1 & A & C &Bus&1&20\\\hline
+1 & B & X &Aus&3&10\\\hline
+1 & B & C &Bus&4&10\\\hline
+2 & A & X &Aus&1&20\\\hline
+2 & A & C &Bus&4&10\\\hline
+2 & B & X &Aus&5&20\\\hline
+2 & B & C &Bus&4&10\\\hline
+\end{tabular}
+\end{center}
+\caption{A dataset with more complex information}
+\label{tb:complexdb}
+\end{table}
+
+\noindent Give the data in Table \ref{tb:complexdb}. Assume we have the following observation types (Correct me if it's not very appropriate.)
+\begin{itemize}
+\item Plot (with measurement type: PlotLabel)
+\item SubPlot (with measurement type: SubPlotLabel)
+\item Tmnt (represent {\em Treatment} (With measurement type TmntType))
+\item Sp (represent {\em Species}, with measurement type: SpName)
+\item Ind (represent {\em Individual}, with measurement type: IndLabel and Weight)
+\end{itemize}
+
+\noindent  Given the dataset in Table \ref{tb:complexdb}, users have different situations to catch. 
+\begin{itemize}
+\item Case 1: {\em Plot} with label ``1'' should refer to the same one physical plot (i.e., the Plot in the first 4 row means the same thing); similarly, 
+  {\em plot} with label ``2'' should refer to the second physical plot (i.e., the Plot in the last 4 row means the same thing). 
+\begin{itemize} 
+\item This can be captured in annotation by putting ``Distinct yes'' for observation type {\em Plot} and ``key yes'' for its measurement type {\em PlotLabel}. 
+\end{itemize}
+\item Case 2: {\em sub-plot}s with the same lable should refer to the same physical sub-plot if they are within the same plot; 
+but the sub-plot with the same label with different {\em Plot} label are different sub-plots.
+E.g., Row 1 \{Plot=1, sub-plot=A\} refers the same sub-plot as that in Row 2, but is different from the one in Row 5 \{Plot=2, sub-plot=A\}. 
+ \begin{itemize} 
+\item This can be captured by putting ``Distinct yes'' for observation type {\em SubPlot} , ``key yes'' for its measurement type {\em SubPlotLabel}.
+We need to denote {\em Plot} is its context and with {\em identifying yes} specified on this context.
+\end{itemize}
+\item Case 3: {\em Tmnt} with the same lable should refer to the same treatment process (So that we can aggregate on different treatment process, e.g., on ``X'' or on ``C''.)  But the treatment at different sub-plot should refer to different treatment.
+ \begin{itemize} 
+\item The first requirement can be captured by treating all the Treatment with value ``X'' as the same entity. 
+The second requirement can be captured by treating the treatment in different sub-plots as different observations߆. 
+I.e., treatments in row 1 and row 3 are of the same entity, but are different observations. 
+\item At the first glance, to represent this, {\em TmntType} for {\em Tmnt} should be specified with ``key yes''. 
+It should have context {\em sub-plot} which is specified with ``identifying yes''. 
+\item After further analysis, {\bf  one question arises: 
+The key measurements for the treatment observation is different from the key measurement  of the entity treatment}. 
+After considering the context, the key measurements for the treatment observation are \{Plotlabel, SubPlotLabel, TmntType\}.  
+When two rows have the same value on these measurements, they represent the same observation instance.
+However, the key measurement for the treatment entity is just \{TmntType\}. When two rows have the same value on it, they represent the same entity instance. 
+The {\em identifying} constraint can only capture the observation context. 
+{\bf This problem is more obvious when we analyze Case 5}. 
+\item Another different annotation may be applied to catch this semantic. E.g., treat the {\em treatments} in different rows as different entity instances. 
+This way, the observation type and the entity type have the same key measurement types \{Plotlabel, SubPlotLabel, TmntType\}. 
+However, this problem still exists for Case 4 and Case 5. 
+\end{itemize}
+
+\item Case 4:  {\em Sp} with the same name should refer to the same species (e.g., a bird named {\em Aus} flies from sub-plot (1,A) to (1,B).)  
+But the {\em Sp} with the same name at different sub-plot should refer to different observations of a specie.
+\begin{itemize}
+\item At the first glance, to represent this, {\em SpName} for {\em Sp} should be specified with ``key yes''. 
+It should have context {\em sub-plot} which is specified with ``identifying yes''. 
+\item {\bf The same problem as Case 3: The key measurements for the {\em Sp} observation is different from the key measurement  of the entity {\em Sp}}.
+the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName\}. 
+However, the key measurements for the species entity is just \{SpName\}. 
+\end{itemize}
+
+\item Case 5:  {\em Ind} with the same label and and the same species name should refer to the same species.
+But the individual (with the same lable and the same species name) at different sub-plot should refer to different species observations.
+\begin{itemize}
+\item {\bf The same problem as Case 4 and Case 5: The key measurements for the {\em Ind} observation is different from the key measurement  of the entity {\em Ind}.}
+the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName, Ind\}. 
+However, the key measurement for the species entity is just \{SpName, Ind\}.  When two rows have the same value on these two columns, they represent the same entity instance. 
+In this case, the observation context of Ind is {\em Sp} and {\em Sub-plot}. But the entity context of Ind is just {\em Sp}. 
+\end{itemize}
+
+\end{itemize}
+
+In summary, we can get a better idea about the problem described in the above use cases can when we answer the following two simple questions: \\
+Q1:  Will an entity type and an observation type (which is of the given entity type) always have the same key measurement type(s)?
+     The above use cases give situations that the answer is no.  \\
+Q2: is {\em identifying} itself enough to distinguish the key measurement for observation types and for entity types? 
+    My temporary answer to this question is no.
+
+A general thinking: 
+the counterpart in RDB (Relational DataBase) is a relational scheme with key attributes. 
+Here, we have two levels of objects: entity level and instance level.
+Then, for different levels of objects, we need to have different ways to specify their key measurements.
+
+\newpage
+
 {\bf Use cases to show that it's needed to have {\em key yes}, {\em identifying yes} and {distinct yes}}
@@
 
-Assume the following tabe is the measurement for some plant tree at different spots.
+Assume the following table is the measurement for some plant tree at different spots.
 
@@
 
-\begin{table}[htb]
-\begin{center}
-
-\end{center}
-\vspace{-0.2in}
-\caption{Dataset}
-\end{table}
-
 Based on this note, it seems like it is not useful to denote {\em distinct yes}.

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPage.java
playing with font size for the help. still some clipping issues.

@@ -98,6 +98,9 @@ public class AnnotationPage extends AbstractUIPage {
 	private Characteristic currentCharacteristic;
 	private Standard currentStandard;
 	private Protocol currentProtocol;
+	
+	private String MEASUREMENT_HELP_TEXT = "The Measurement 'Is Key' when it identifies the uniqueness of an Observation instance. Multiple measurements may be combined as a compound key.";
+	private String OBSERVATION_HELP_TEXT = "The Observation 'Is Distinct' when the observation Entity identifies the uniqueness of the observation Instance.";
 
 	// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
 	// *
@@ -198,7 +201,7 @@ public class AnnotationPage extends AbstractUIPage {
 				8 * WizardSettings.PADDING));
 		
 		measurementPanel.add(measurementLabelPanel);
-		measurementPanel.add(WidgetFactory.makeHTMLLabel("The Measurement 'Is Key' when it identifies the uniqueness of an Observation instance. Multiple measurements may be combined as a compound key.", 2));
+		measurementPanel.add(WidgetFactory.makeHTMLLabel(MEASUREMENT_HELP_TEXT , 2));
 		this.add(measurementPanel);
 		this.add(WidgetFactory.makeDefaultSpacer());
 		
@@ -221,7 +224,8 @@ public class AnnotationPage extends AbstractUIPage {
 				8 * WizardSettings.PADDING));
 		
 		observationPanel.add(labelPanel);
-		observationPanel.add(WidgetFactory.makeHTMLLabel("The Observation 'Is Distinct' when the observation Entity identifies the uniqueness of the observation Instance.", 2));
+		observationPanel.add(WidgetFactory.makeHTMLLabel(OBSERVATION_HELP_TEXT , 2));
+
 		this.add(observationPanel);
 		this.add(WidgetFactory.makeDefaultSpacer());
 		

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/table/AnnotationTable.java
reorder the annotation table after the underlying annotation has been modified and the table "resets". do this only if it has previously been reordered http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4739

@@ -45,6 +45,8 @@ import org.ecoinformatics.sms.annotation.Observation;
  */
 public class AnnotationTable extends MultiSpanCellTable {
 
+	private boolean reordered = false;
+	
 	public AnnotationTable(AnnotationTableModel annotationTableModel) {
 		super(annotationTableModel);
 //		this.setIntercellSpacing(new Dimension(0,0));
@@ -57,7 +59,18 @@ public class AnnotationTable extends MultiSpanCellTable {
 		return new ObservationCellRenderer();
 	}
 	
+	public boolean isReordered() {
+		return reordered;
+	}
+
+	public void setReordered(boolean reordered) {
+		this.reordered = reordered;
+	}
+
 	public void reorder(boolean includeContext) {
+		
+		this.reordered = true;
+		
 		List<Observation> sortedObservations = new ArrayList<Observation>();
 		
 		for (int i = 0; i < this.getColumnCount(); i++) {

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/table/AnnotationTablePanel.java

@@ -140,6 +140,10 @@ public class AnnotationTablePanel extends JPanel implements StateChangeListener 
 		annotationScrollPane.setRowHeaderView(rowheaders);
 		
 		annotationTable.reset();
+		
+		if (annotationTable.isReordered()) {
+			annotationTable.reorder(true);
+		}
 	
 	}
 

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
refactor plugins into pages/commands packages

@@ -36,6 +36,11 @@ import javax.swing.table.TableColumnModel;
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.ontology.OntologyClass;
+import org.ecoinformatics.sms.plugins.commands.AnnotationCommand;
+import org.ecoinformatics.sms.plugins.commands.AnnotationSearchCommand;
+import org.ecoinformatics.sms.plugins.commands.ContextCommand;
+import org.ecoinformatics.sms.plugins.commands.ObservationCommand;
+import org.ecoinformatics.sms.plugins.commands.SaveAnnotationCommand;
 import org.ecoinformatics.sms.plugins.table.AnnotationTableModel;
 import org.ecoinformatics.sms.plugins.table.AnnotationTablePanel;
 import org.ecoinformatics.sms.plugins.table.DataTableModelListener;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/AnnotationCommand.java

@@ -24,7 +24,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
 import java.util.List;
@@ -33,6 +33,8 @@ import javax.swing.JTable;
 
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
+import org.ecoinformatics.sms.plugins.pages.AnnotationPage;
 
 import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
 import edu.ucsb.nceas.morpho.datapackage.DataViewContainerPanel;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/AnnotationSearchCommand.java

@@ -24,7 +24,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import edu.ucsb.nceas.morpho.Morpho;
 import edu.ucsb.nceas.morpho.framework.ModalDialog;
@@ -40,6 +40,8 @@ import edu.ucsb.nceas.morpho.util.UISettings;
 
 import java.util.Vector;
 import java.awt.event.ActionEvent;
+
+import org.ecoinformatics.sms.plugins.pages.AnnotationQueryPage;
 
 /**
  * Class to handle annotation search command

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/ContextCommand.java

@@ -24,7 +24,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
 import java.util.ArrayList;
@@ -40,6 +40,8 @@ import org.ecoinformatics.sms.annotation.Mapping;
 import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.annotation.Relationship;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
+import org.ecoinformatics.sms.plugins.pages.ContextPage;
 
 import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
 import edu.ucsb.nceas.morpho.datapackage.DataViewContainerPanel;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/DirectAnnotationCommand.java

@@ -24,7 +24,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
 import java.util.List;
@@ -39,6 +39,8 @@ import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
 import org.ecoinformatics.sms.ontology.OntologyClass;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
+import org.ecoinformatics.sms.plugins.pages.OntologyClassSelectionPage;
 import org.ecoinformatics.sms.plugins.table.AnnotationTable;
 import org.ecoinformatics.sms.plugins.table.AnnotationTableModel;
 

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/ObservationCommand.java

@@ -24,7 +24,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
 import java.util.ArrayList;
@@ -37,6 +37,7 @@ import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Mapping;
 import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
 import edu.ucsb.nceas.morpho.datapackage.DataViewContainerPanel;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/SaveAnnotationCommand.java

@@ -24,9 +24,11 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
+
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
 import edu.ucsb.nceas.morpho.framework.UIController;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java

@@ -26,7 +26,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.pages;
 
 import java.awt.Dimension;
 import java.awt.GridLayout;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationQueryPage.java

@@ -26,7 +26,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.pages;
 
 import java.util.List;
 
@@ -39,6 +39,7 @@ import org.ecoinformatics.sms.annotation.Characteristic;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.ui.SimpleAnnotationPanel;
 
 import edu.ucsb.nceas.morpho.Morpho;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ContextPage.java

@@ -26,7 +26,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.pages;
 
 import java.awt.event.MouseAdapter;
 import java.awt.event.MouseEvent;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/pages/OntologyClassSelectionPage.java

@@ -26,7 +26,7 @@
  * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
  */
 
-package org.ecoinformatics.sms.plugins;
+package org.ecoinformatics.sms.plugins.pages;
 
 import java.awt.BorderLayout;
 import java.awt.event.MouseAdapter;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/table/AnnotationTablePanel.java

@@ -45,7 +45,7 @@ import javax.swing.ListSelectionModel;
 import javax.swing.UIManager;
 import javax.swing.table.TableCellRenderer;
 
-import org.ecoinformatics.sms.plugins.DirectAnnotationCommand;
+import org.ecoinformatics.sms.plugins.commands.DirectAnnotationCommand;
 
 import edu.ucsb.nceas.morpho.util.Command;
 import edu.ucsb.nceas.morpho.util.GUIAction;

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/ui/SimpleAnnotationPanel.java

@@ -41,7 +41,7 @@ import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.plugins.AnnotationPlugin;
-import org.ecoinformatics.sms.plugins.OntologyClassSelectionPage;
+import org.ecoinformatics.sms.plugins.pages.OntologyClassSelectionPage;
 
 import edu.ucsb.nceas.morpho.framework.ModalDialog;
 import edu.ucsb.nceas.morpho.framework.UIController;

### 2010-02-05 docs/algdraft/model.tex
add the files for model comparison

\documentclass{article}
\begin{document}

\section{How to evaluate the quality of a conceptual model}

This question is posted based on the need on evaluating which conceptual model 
(e.g., OBOE, O\&M, etc) is a good one, or is better than the other. 
Unfortunately, the literature \cite{DBLP:journals/dke/Moody05, ***} shows that
there is no well-formed standard way to evaluae the quality of a conceptual model. 
On the contrary, most of the conceptual models are evaluated in an {\em ad hoc} manner. 
However, this article highlights a major principle in evaluating a conceptual model: 
a conceptual model is valuable only it is used in practice.

Following this principle, we consider several factors in building our final model 
(either OBOE or an extended OBOE) in SONET project. The several factors include:\\  
1. Metadata model how complex it is?\\
2. How easy users can use the CM?\\
3. How easy it is to perform operations on the model? \\

\section{Term correspondences}
\begin{table}[htb]
\begin{tabular}{|l|l|l|l|}
\hline 
ER model & OBOE model term & O\&M term &others\\\hliine
Entity & Observation & Feature of interest & \\\hline
Characteristic & Measurement & ... & \\\hline
               & Characteristic & Property&\\\hline
               & Standard & ? & \\\hline
\end{tabular}
\end{table}

In O\&M, an {\em Observation} is defined as an action. 
It is modeled as a {\em Feature}. 
{\em Feature of Interest} is a feature that a user is interested in. 
Each feature can have properties. The {\em Property} can be {\em observed property} and other non-observed property. 
The value of {\em observed property} is obtained by an observer (e.g., a person)'s observation action.

Procedure, Result,

It seems to me: \\
Measurement  -- Feature type\\
Characteristic -- Property type\\
Standard \& Value -- Result\\
Procedure/protocal -- Procedure\\

How to represent the observation and measurement relationship in O\&M? \\
how to represent the observation - observation relationship in O\&M?\\
Any limitation for

Several questions in understanding O\&M, 
what is {\em feature} (type and instance), {\em property} (type and instance).
According to the definition  4.19 
property-type is a characteristic of one or more feature types. 
It can have value. Or It must have value? 
It can have one/multiple values?

Several questions to ask:
\begin{itemize}
\item How O\&M represent the relationship between Observation and Measurement in OBOE (1:m)
\end{itemize}

\section{Use cases}
We have several use cases that we can test the different data models. \\
{\bf to come. }

Some notes for myself: \\
sms: semantic mediation system

\bibliographystyle{abbrv}
\bibliography{model.bib}

\end{document}

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
added the remove option http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4737

@@ -35,11 +35,14 @@ import javax.swing.table.TableColumnModel;
 
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.annotation.Measurement;
+import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.plugins.commands.AnnotationCommand;
 import org.ecoinformatics.sms.plugins.commands.AnnotationSearchCommand;
 import org.ecoinformatics.sms.plugins.commands.ContextCommand;
 import org.ecoinformatics.sms.plugins.commands.ObservationCommand;
+import org.ecoinformatics.sms.plugins.commands.RemoveCommand;
 import org.ecoinformatics.sms.plugins.commands.SaveAnnotationCommand;
 import org.ecoinformatics.sms.plugins.table.AnnotationTableModel;
 import org.ecoinformatics.sms.plugins.table.AnnotationTablePanel;
@@ -93,6 +96,10 @@ public class AnnotationPlugin
 	private GUIAction addContextAction;
 
 	private GUIAction removeContextAction;
+
+	private GUIAction removeObservationAction;
+
+	private GUIAction removeMeasurementAction;
 	
 
 	/**
@@ -112,6 +119,8 @@ public class AnnotationPlugin
 			Log.debug(6, see.toString());
 		}
 
+		int menuPosition = 0;
+		
 		// initialize the actions
 		annotateAction =
 			new GUIAction(
@@ -121,7 +130,7 @@ public class AnnotationPlugin
 		annotateAction.setToolTipText(
 			"Add/edit annotation or this data table attribute");
 		annotateAction.setSeparatorPosition(Morpho.SEPARATOR_PRECEDING);
-		annotateAction.setMenuItemPosition(0);
+		annotateAction.setMenuItemPosition(menuPosition++);
 		annotateAction.setMenu(ANNOTATION_MENU_LABEL, ANNOTATIONMENUPOSITION);
 		annotateAction.setEnabled(false);
 
@@ -136,7 +145,7 @@ public class AnnotationPlugin
 				new ObservationCommand(true));
 		mergeObservationAction.setToolTipText(
 			"Selected columns are for the same Observation");
-		mergeObservationAction.setMenuItemPosition(5);
+		mergeObservationAction.setMenuItemPosition(menuPosition++);
[…]

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/RemoveCommand.java

@@ -1,0 +1,192 @@
+/**
+ *  '$RCSfile: EditColumnMetaDataCommand.java,v $'
+ *  Copyright: 2000 Regents of the University of California and the
+ *              National Center for Ecological Analysis and Synthesis
+ *    Authors: @tao@
+ *    Release: @release@
+ *
+ *   '$Author$'
+ *     '$Date$'
+ * '$Revision$'
+ *
+ * This program is free software; you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation; either version 2 of the License, or
+ * (at your option) any later version.
+ *
+ * This program is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with this program; if not, write to the Free Software
+ * Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+ */
+
+package org.ecoinformatics.sms.plugins.commands;
+
+import java.awt.event.ActionEvent;
+import java.util.ArrayList;
+import java.util.List;
+
+import javax.swing.JOptionPane;
+import javax.swing.JTable;
+
+import org.ecoinformatics.sms.SMS;
+import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.annotation.Context;
+import org.ecoinformatics.sms.annotation.Mapping;
+import org.ecoinformatics.sms.annotation.Measurement;
+import org.ecoinformatics.sms.annotation.Observation;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
+
+import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
+import edu.ucsb.nceas.morpho.datapackage.DataViewContainerPanel;
+import edu.ucsb.nceas.morpho.datapackage.DataViewer;
+import edu.ucsb.nceas.morpho.framework.MorphoFrame;
+import edu.ucsb.nceas.morpho.framework.UIController;
+import edu.ucsb.nceas.morpho.util.Command;
+import edu.ucsb.nceas.morpho.util.Log;
+import edu.ucsb.nceas.morpho.util.StateChangeEvent;
[…]

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
organize the Annotation menu and popup entries with separators

@@ -129,7 +129,7 @@ public class AnnotationPlugin
 				new AnnotationCommand());
 		annotateAction.setToolTipText(
 			"Add/edit annotation or this data table attribute");
-		annotateAction.setSeparatorPosition(Morpho.SEPARATOR_PRECEDING);
+		annotateAction.setSeparatorPosition(Morpho.SEPARATOR_FOLLOWING);
 		annotateAction.setMenuItemPosition(menuPosition++);
 		annotateAction.setMenu(ANNOTATION_MENU_LABEL, ANNOTATIONMENUPOSITION);
 		annotateAction.setEnabled(false);
@@ -207,6 +207,7 @@ public class AnnotationPlugin
 			"Define a Context relationship between this Observation and another");
 		addContextAction.setMenuItemPosition(menuPosition++);
 		addContextAction.setMenu(ANNOTATION_MENU_LABEL, ANNOTATIONMENUPOSITION);
+		addContextAction.setSeparatorPosition(Morpho.SEPARATOR_PRECEDING);
 		addContextAction.setEnabled(false);
 
 		addContextAction.setEnabledOnStateChange(
@@ -237,6 +238,7 @@ public class AnnotationPlugin
 	    saveAction.setMenuItemPosition(menuPosition++);
 	    saveAction.setToolTipText("Save Annotations...");
 		saveAction.setMenu(ANNOTATION_MENU_LABEL, ANNOTATIONMENUPOSITION);
+		saveAction.setSeparatorPosition(Morpho.SEPARATOR_PRECEDING);
 	    saveAction.setEnabled(false);
 	    saveAction.setEnabledOnStateChange(
                 StateChangeEvent.CREATE_ENTITY_DATAPACKAGE_FRAME,
@@ -539,11 +541,13 @@ public class AnnotationPlugin
 				DataViewer dataView = resultPane.getCurrentDataViewer();
 				if (dataView != null) {
 					dataView.addPopupMenuItem(UIController.getGUIActionCloneUsedByMorphoFrame(annotateAction, morphoFrame), true);
-					dataView.addPopupMenuItem(UIController.getGUIActionCloneUsedByMorphoFrame(mergeObservationAction, morphoFrame), false);
+					// separator
+					dataView.addPopupMenuItem(UIController.getGUIActionCloneUsedByMorphoFrame(mergeObservationAction, morphoFrame), true);
[…]

### 2010-02-05 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
when the DP is closed and unsaved annotation changes exist, we now throw them away in favor of the annotation as it exists on disk. http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4738

@@ -17,6 +17,8 @@ import java.awt.BorderLayout;
 import java.awt.Component;
 import java.awt.event.AdjustmentListener;
 import java.awt.event.MouseListener;
+import java.awt.event.WindowAdapter;
+import java.awt.event.WindowEvent;
 import java.io.ByteArrayOutputStream;
 import java.io.File;
 import java.io.FileInputStream;
@@ -279,11 +281,24 @@ public class AnnotationPlugin
 		initializeOntologies();
 		
 		// initialize the annotations
-		initializeAnnotations();
+		initializeAnnotations(null);
+		
+	}
+	
+	private void initializeWindowAdapter() {
+		morphoFrame = UIController.getInstance().getCurrentActiveWindow();
+		WindowAdapter windowListener = new WindowAdapter() {
+			public void windowClosed(WindowEvent we) {
+				String docid = morphoFrame.getAbstractDataPackage().getAccessionNumber();
+				initializeAnnotations(docid);
+				morphoFrame.removeWindowListener(this);
+			}
+		};
+		morphoFrame.addWindowListener(windowListener);
 	}
 	
 	// TODO: search Metacat for annotations
-	private void initializeAnnotations() {
+	private void initializeAnnotations(String forDocid) {
 		FileSystemDataStore fds = new FileSystemDataStore(Morpho.thisStaticInstance);
 		String querySpec = getAnnotationQuery();
 		Query query = new Query(querySpec, Morpho.thisStaticInstance);
@@ -291,16 +306,22 @@ public class AnnotationPlugin
 		query.setSearchMetacat(false);
 		ResultSet rs = query.execute();
 		Vector<Vector> resultVector = rs.getResultsVector();
-		Vector<String> docids = new Vector<String>();
 		for (Vector row: resultVector) {
 			String docid = (String) row.get(ResultSet.DOCIDINDEX);
 			try {
 				InputStream is = new FileInputStream(fds.openFile(docid));
-				SMS.getInstance().getAnnotationManager().importAnnotation(is, docid);
+				Annotation annotation = Annotation.read(is);
+				// if we are filtering, skip any docs that don't match
+				if (forDocid != null) {
+					String docidFromAnnotation = annotation.getEMLPackage();
+					if (!forDocid.equals(docidFromAnnotation)) {
+						continue;
+					}
+				}
+				SMS.getInstance().getAnnotationManager().importAnnotation(annotation, docid);
[…]

### 2010-03-05 dev/sms/src/org/ecoinformatics/sms/plugins/commands/ViewAnnotationCommand.java
add jgraphx visualization of the Annotation

@@ -68,9 +68,6 @@ public class ViewAnnotationCommand implements Command {
 					UISettings.POPUPDIALOG_HEIGHT,
 					true);
 		
-//		JOptionPane.showMessageDialog(
-//				UIController.getInstance().getCurrentActiveWindow(), 
-//				annotation.toString());
 	}
 
 }

### 2010-03-05 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ViewAnnotationPage.java

@@ -28,13 +28,17 @@
 
 package org.ecoinformatics.sms.plugins.pages;
 
+import java.awt.Component;
+
 import javax.swing.BoxLayout;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 import javax.swing.JScrollPane;
+import javax.swing.JTabbedPane;
 import javax.swing.JTextArea;
 
 import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.renderer.AnnotationGraph;
 
 import edu.ucsb.nceas.morpho.framework.AbstractUIPage;
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WidgetFactory;
@@ -51,6 +55,8 @@ public class ViewAnnotationPage extends AbstractUIPage {
 	private final String title = "View Annotation";
 	private final String subtitle = "";
 	private JTextArea annotationXML;
+	
+	private JTabbedPane tabPane = null;
 	
 	private Annotation annotation;
 
@@ -75,16 +81,13 @@ public class ViewAnnotationPage extends AbstractUIPage {
 
 		this.setLayout(new BoxLayout(this, BoxLayout.Y_AXIS));
 		
-		JLabel desc = WidgetFactory
-				.makeHTMLLabel(
-						"<b>Annotation XML</b>",
-						2);
+		JLabel desc = WidgetFactory.makeHTMLLabel("<b>View Annotation</b>", 2);
 		this.add(desc);
-		this.add(WidgetFactory.makeDefaultSpacer());
-		
-		this.add(WidgetFactory.makeDefaultSpacer());
-		
-		// annotation
+
+		tabPane = new JTabbedPane();
+		this.add(tabPane);
+				
+		// annotation XML
 		JPanel annotationPanel = WidgetFactory.makePanel();
 		annotationPanel.add(WidgetFactory.makeLabel("Annotation:", false));
 		annotationXML = WidgetFactory.makeTextArea("", 20, false);
@@ -92,9 +95,20 @@ public class ViewAnnotationPage extends AbstractUIPage {
 		annotationPanel.add(annotationScrollPane);
 		annotationPanel.setBorder(new javax.swing.border.EmptyBorder(0, 0, 0,
 				8 * WizardSettings.PADDING));
-		this.add(annotationPanel);
+		
+		tabPane.insertTab("Annotation XML", null, annotationPanel, "View XML source", 0);
+		
+		// annotation graph
[…]

### 2010-03-05 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java

@@ -1,0 +1,87 @@
+package org.ecoinformatics.sms.renderer;
+
+import java.awt.Component;
+
+import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.annotation.Characteristic;
+import org.ecoinformatics.sms.annotation.Measurement;
+import org.ecoinformatics.sms.annotation.Observation;
+import org.ecoinformatics.sms.annotation.Standard;
+
+import com.mxgraph.model.mxCell;
+import com.mxgraph.swing.mxGraphComponent;
+import com.mxgraph.view.mxGraph;
+
+public class AnnotationGraph {
+	
+	public static Component createAnnotationGraph(Annotation annotation) {
+		
+		// the graph
+		mxGraph graph = new mxGraph();
+		Object parent = graph.getDefaultParent();
+		int x = 20;
+		int y = 20;
+		int width = 80;
+		int height = 30;
+		
+		for (Observation observation: annotation.getObservations()) {
+			// start at the left again
+			x = 20;
+			graph.getModel().beginUpdate();
+			try {
+				
+				Object observationCell = graph.addCell(new mxCell(observation), parent);
+				Object observationNode = 
+					graph.insertVertex(observationCell, null, observation, x, y, width, height);
+				
+				for (Measurement measurement: observation.getMeasurements()) {
+					
+					// shift over to right
+					x = (width*3);
+					
+					// add measurement
+					Object measurementNode = 
+						graph.insertVertex(observationCell, null, measurement.getLabel(), x, y, width, height);
+					
+					graph.insertEdge(observationCell, null, "hasMeasurement", observationNode, measurementNode);
+					
+					// shift over to right
+					x = (width*6);
+
+					// add characteristic, if available
+					Characteristic characteristic = null;
+					try {
+						characteristic = measurement.getCharacteristics().get(0);
+					} catch (Exception e) {}
+					Object characteristicNode = 
+						graph.insertVertex(observationCell, null, characteristic, x, y, width, height);
+					graph.insertEdge(observationCell, null, "of", measurementNode, characteristicNode);
+					
+					// shift down
+					y += (height*2);
+					
+					// add standard
+					Standard standard = measurement.getStandard();
+					Object standardNode = 
+						graph.insertVertex(observationCell, null, standard, x, y, width, height);
+					graph.insertEdge(observationCell, null, "usesStandard", measurementNode, standardNode);
[…]

### 2010-03-05 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java
include Protocol node include ofEntity edge and Entity node

@@ -6,6 +6,7 @@ import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
 import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
+import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
 
 import com.mxgraph.model.mxCell;
@@ -31,9 +32,19 @@ public class AnnotationGraph {
 			try {
 				
 				Object observationCell = graph.addCell(new mxCell(observation), parent);
-				Object observationNode = 
-					graph.insertVertex(observationCell, null, observation, x, y, width, height);
 				
+				// add observation
+				Object observationNode = 
+					graph.insertVertex(observationCell, null, observation.getLabel(), x, y, width, height);
+				
+				// shift down
+				//y += (height*2);
+				
+				// of entity
+				Object entityNode = 
+					graph.insertVertex(observationCell, null, observation.getEntity(), x, y + (height*2), width, height);
+				graph.insertEdge(observationCell, null, "ofEntity", observationNode, entityNode);
+
 				for (Measurement measurement: observation.getMeasurements()) {
 					
 					// shift over to right
@@ -55,7 +66,7 @@ public class AnnotationGraph {
 					} catch (Exception e) {}
 					Object characteristicNode = 
 						graph.insertVertex(observationCell, null, characteristic, x, y, width, height);
-					graph.insertEdge(observationCell, null, "of", measurementNode, characteristicNode);
+					graph.insertEdge(observationCell, null, "ofCharacteristic", measurementNode, characteristicNode);
 					
 					// shift down
 					y += (height*2);
@@ -68,6 +79,16 @@ public class AnnotationGraph {
 					
 					// shift down
 					y += (height*2);
+					
+					// add protocol
+					Protocol protocol = measurement.getProtocol();
+					Object protocolNode = 
+						graph.insertVertex(observationCell, null, protocol, x, y, width, height);
+					graph.insertEdge(observationCell, null, "usesProtocol", measurementNode, protocolNode);
+					
+					// shift down
+					y += (height*2);
+					
 				}
 				// shift down
 				y += height;

### 2010-03-05 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java
add context/relationship edges between observation nodes

@@ -1,12 +1,16 @@
 package org.ecoinformatics.sms.renderer;
 
 import java.awt.Component;
+import java.util.HashMap;
+import java.util.Map;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
+import org.ecoinformatics.sms.annotation.Context;
 import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.annotation.Protocol;
+import org.ecoinformatics.sms.annotation.Relationship;
 import org.ecoinformatics.sms.annotation.Standard;
 
 import com.mxgraph.model.mxCell;
@@ -25,6 +29,8 @@ public class AnnotationGraph {
 		int width = 80;
 		int height = 30;
 		
+		Map<Observation, mxCell> observationMap = new HashMap<Observation, mxCell>();
+		
 		for (Observation observation: annotation.getObservations()) {
 			// start at the left again
 			x = 20;
@@ -33,6 +39,8 @@ public class AnnotationGraph {
 				
 				Object observationCell = graph.addCell(new mxCell(observation), parent);
 				
+				observationMap.put(observation, (mxCell) observationCell);
+
 				// add observation
 				Object observationNode = 
 					graph.insertVertex(observationCell, null, observation.getLabel(), x, y, width, height);
@@ -95,6 +103,24 @@ public class AnnotationGraph {
 			}
 			finally {
 				graph.getModel().endUpdate();
+			}
+		}
+		
+		// process context
+		Object[] observationCells = graph.getChildCells(parent);
+		for (int i = 0; i < observationCells.length; i++) {
+			mxCell observationCell = (mxCell) observationCells[i];
+			Observation observation = (Observation) observationCell.getValue();
+			if (observation.getContexts() != null) {
+				for (Context context: observation.getContexts()) {
+					Observation targetObservation = context.getObservation();
+					Relationship relationship = context.getRelationship();
+					mxCell targetObservationCell = observationMap.get(targetObservation);
+					// add the context from one observation node to the other observation node
+					// here we know the Observation is the first child in the observationCell
+					// TODO: figure out how to effectively do sub-graphs so that i can connect individual observation cells
+					graph.insertEdge(parent, null, relationship, observationCell.getChildAt(0), targetObservationCell.getChildAt(0));
+				}	
 			}
 		}
 		

### 2010-03-05 docs/algdraft/MaterializeDB.tex
Update the documents.

--- 2010-01-22 docs/algdraft/Materialize_DB.tex
@@
 \newtheorem{example}{Example}[section]
+\newtheorem{definition}{Definition}[section]
+\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
 \begin{document}
 
-\section{Valid annotation constraints}
-Among the {\em key}, {\em distinct}, and {\em identifying} constraints,
-there are correlations between ``key yes'' and ``distinct yes''.
+\section{Key constraints}
+
+%motivation to have key constraints.
+In real world, the dataset collected by an ecologist to describe
+measurents and observations is not normalized. I.e., two different
+rows in the table may have a big amount of ``redundant'' 'information
+to describe an object (e.g., entity).
+
+As discussed in \cite{***} %ER paper, 
+we can use the OBOE cenceptual model to capture the semantic
+relationships from the columns to their real meaning. 
+However, the mapping from the columns to the OBOE model still cannot
+directly solve the problem of getting concise informatin (e.g., unique
+entity instances, unique observation instances, etc.)
+
+Given Table \ref{tb:pltarea}(a), it records the fact that four times
+of dbh of some trees are taken. 
+Very often, an ecologist may ask questions like this:  Give me the
+average dbh for every different trees. 
+First, we have four rows here for trees. 
+But how many tree entity instances are here is a question.
+
+There are several cases to consider:
 \begin{itemize}
-\item ``Distinct yes'' on an observation type implies that its measurements collectively form a key of this observation type.
-\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
-In this case, when two rows (for the observation type) have the same values for these key measurements, they represent the different observation instances.
-It means that there may be other ``variables'' that can distinguish one observation instance from the other, but this dataset does not catch this variable ({\bf HP: is ``variable'' the right word to use?
-}).
-So, in real application, we do not encourage such situation to appear.
+\item Case 1: The naive extreme way to interpret the data is that each
+  observation is from different tree entity. Then, we have {\bf four}
+  tree entities. This may be too strict. People may say, well, I have
+  some observations for the same entity. 
+\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
+That's obvious that {\em piru} is different from {\em abba}. 
+With this constraint, we get {\bf two} tree entities. However, an
+obvious fact is also true that one {\em piru} tree cannot at the
+different plots. (It is another story if this tree was migrated from one plot to another.) 
+\item Case 3: Following Case 2, people want to further limit that the
+  same {\em spp} in the same {\em plt} should represent the same tree
+  entity set.  
+%To achieve this, we use {\em identifying yes}.
+%   Now, it should return
+%	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
 \end{itemize}
 
-An example with invalid annotation: given the data in Table \ref{tb:pltarea}. The following annotation is {\em invalid}.
-According to the annotation, if two plots have the same value on EntityName, they represent the same plot observation. 
-Obviously, it will have probelem to interprete the data in Table \ref{tb:pltarea} with this annotatioin. 
-E.g., the first and second row catching plot with EntityName $A$ should be the same plot obervation.
-However, the data shows that this one plot has two different areas $1.0$ and $1.1$, so, there is confusion here.
-\\
+%Q2: Why we need to distinguish the same entities using {\em key yes}
+%and {\em distinct yes}) to identify the same observation? 
+%For the example, we have one observation for splot {\em A}.
+%What's the semantic purpose of this? What kind of query may need this? 
+%E.g., how many spots in this dataset?
+
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{cc}
+\begin{tabular}{|l|l|l|}
+\hline
+plt & spp & dbh\\\hline
+A & piru & 35.8 \\\hline
+A & piru & 36.2 \\\hline
+B & piru &33.2 \\\hline
+B&abba&34\\\hline
+\end{tabular}
+&
+\begin{tabular}{|l|l|l|l|}
+\hline
+plt & area & spp & dbh\\\hline
+A & 1.0 & piru & 35.8 \\\hline
+A & 1.1 & piru & 36.2 \\\hline
+B & 2.0 &piru &33.2 \\\hline
+B& 2.0 & abba&34\\\hline
+\end{tabular}\\
+(a) & (b)
+\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset}
+\label{tb:pltarea}
+\end{table}
+
+Obviously, to answer this question, we not only need to figure out
+what are the observations and measurements, we also need proper
+constraints to help us understand whether different observations are
+from the same entity or not.
+
+To tackle this problem, we proposed to use {\em key yes}, {\em
+  distinct yes} and {\em identifying yes} constraints.
+
+\from{HP} {Describe or define {\em key yes},  {\em distinct yes},  {\em identifying yes}}
+
+\begin{definition}[Key measurement types]{
+The key mensurement tytpes of an observation are the measurement types
+that are used to distinguish one observation/entity instance from the
+other. }
+\end{definition}
+In most simple case, when an observation type does not have context, 
+the key measurement types are specified with ``key yes''.
+
+{\bf Key constraints}: in this report (paper),  we use the term {\em
+  key constraints} to refer to the general constraints of the above
+three types {\em key yes}, {\em distinct yes} and 
+{\em identifying yes}.
+
+\section{Valid annotation constraints}
+
+Based on the semantic meaning of these key constraints, we can see
+that they have some correlations. 
+Before we give any further examples to explain such correlations, we
+list the correlation rules among them.
+
+\begin{itemize}
+\item {\bf Rule 1}: If an observation type is specified with {\em distinct yes}, all
+  its measurement types are automatically marked with {\em key yes}. 
+\item {\bf Rule 2}: If the context of an observation type is specified with {\em identifying
+    yes}. This observation type MUST have some key measurement
+  types. And the context observation type also MUST have key
+  measurement types too. 
+\end{itemize}
+
+The following example shows an example why Rule 1 is needed. 
+Given the data in Table \ref{tb:pltarea}(b) and the following
+annotation.
 
@@
 \verb|        | {\bf standard} ``sqft''\\
-\verb|       |$\cdots$ Not finished....
+\verb|        |$\cdots$
 
-\begin{table}[htb]
-\begin{center}
-%\begin{tabular}{cc}
-%\begin{tabular}{|l|l|l|}
-%\hline
-%plt & spp & dbh\\\hline
-%A & piru & 35.8 \\\hline
-%A & piru & 36.2 \\\hline
-%B & piru &33.2 \\\hline
-%B&abba&34\\\hline
-%\end{tabular}
-%&
-\begin{tabular}{|l|l|l|l|}
-\hline
-plt & area & spp & dbh\\\hline
-A & 1.0 & piru & 35.8 \\\hline
-A & 1.1 & piru & 36.2 \\\hline
-B & 2.0 &piru &33.2 \\\hline
-B& 2.0 & abba&34\\\hline
-\end{tabular}\\
-%(a) & (b)
-%\end{tabular}
-\end{center}
-\vspace{-0.2in}
-\caption{Dataset}
-\label{tb:pltarea}
-\end{table}
+According to the annotation, if two plots have the same value for
+``EntityName'', they represent the same plot observation. 
+Obviously, it has problem to interprete the data in Table
+\ref{tb:pltarea} with this annotatioin. 
+E.g., the first and second rows catch information about plot with
+EntityName $A$. Accoring to the annotation, they should be the same
+plot obervation. 
+However, the data shows that this one plot has two different areas
+$1.0$ and $1.1$, so, there is confusion here.
 
-In our algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the correlation rules among these constraints.
+For this case, we can have several ways to make the annoation
+consistent. 
+The first alternative to tackle this confusion is to change the
+annotation.  We can let $m_2$ automatically becomes key measurement type of
+$o_1$. Then, every different ($m_1$, $m_2$) value pair denote a
+different observation instance which come from the different entity
+instance.
+The second alternative is to get rid of {\em distinct yes} for $o_1$,
+which means that $(A,1.0)$ and $(A,1.1)$ are different observation instances of the same entity instance
+$A$.
 
-{\bf From HP:} More to come on this part...
+%Here are some very basic rules to restrict the correlations among these constraints. 
+%\begin{itemize}
+%\item ``Distinct yes'' on an observation type implies that all its measurements collectively form a key of this observation type. 
+%In the above example, we can specify ``distinct yes'' on ``$o_1$''
+%without {\em explicitly} denote ``key yes'' on the measurements $m_1$
+%and $m_2$. 
+%The semantic meaning is that ($m_1$, $m_2$) collectively form key measurement for $o_1$.  
+%\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
+%In this case, when two rows (for the observation type) have the same
+%values for these key measurements, they represent the different
+%observation instances. Such a case means that there may exist other
+%``variables'' that can distinguish one observation instance from the
+%other, but this dataset does not catch this variable ({\bf HP: is
+%``variable'' the right word to use? }). So, in real application, we
+%do not encourage such situation to appear. 
+%\end{itemize}
+
+In our materialization algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the rules among these constraints.
 
@@
 	Note that, we hide the ``AnnotId'' in this schema, which denotes the resource the annotation is on.
-    We do not include this in the algorithm because the algorithm focuses on dealing with annotations on one resource.
+    We do not include this because the algorithm focuses on dealing with annotations on one resource.
     This way, we can simplify the description.
@@
 \begin{itemize}
-\item $OBOE.Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances materialized from $Dataset$.
-\item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$
-\item $OBOE.Entity=\{\langle \underline{EntId}, EntType\rangle\}$
-\item $OBOE.Context =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$
+\item $OBOE.Observation=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
+\item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
+\item $OBOE.Entity=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
+\item $OBOE.Context =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
 \end{itemize}
@@
 Algorithm \ref{alg:matdb} shows the framework of our algorithm. 
-In this algorithm, $ObsIdx$ is the index structure maintained for the observation instances 
+In this algorithm, we maintain two intermediate index structures ($EntIdx$ and $ObsIdx$) to keep track of the distinct entity and observation instances. 
+$ObsIdx$ is the index structure maintained for the observation instances 
 whose types are specified with {\em distinct yes}. 
 This index maintains the mapping from observation type and key values to its corresponding observation instance id. 
-Obviously, only when observation type is specified with {\em distinct yes}, i.e., we want to keep track of the same observation instances, we need to maintain their instances in this index. 
+Obviously, only when an observation type is specified with {\em distinct yes}, (i.e., we want to keep track of the same observation instances) we need to maintain their instances in this index.
+
 The key values can be calculated in three different cases.
-In the first case where just one measurement of an observation type is specified with ``key yes'' the key value is the value of this ``key'' measurement.
-In the second case, several measurement types are marked with ``key yes'', the key value is the combined value of these several measurement.
-In the third case, this object type is marked with ``identifying yes'', the key value is the combined value of its context observation's key measurement values.
+In the first case where only one measurement of an observation type is specified with ``key yes'', the key value is the value of this ``key'' measurement.
+ In the second case, several measurement types are marked with ``key yes'', the key value is the combined instance value of these several measurement types. 
+In the third case, some context of this object type is marked with ``identifying yes'', the key value is the combination of the instance values of its own key measurement types and the instance values of its context observation's key measurements.
 
-Similarly, in case that some measurement type(s) of an observation type is/are specified with ``key yes'', 
+$Entidx$ is the index structure for tracking the distinct entities. In case that some measurement type(s) of an observation type is/are specified with ``key yes'', 
 if different observation instances have the same value on these key measurements, semantically, we interpret that they are of the same entity. 
-$Entidx$ is the index structure for tracking the distinct entities.
-The key value is computed in the same way as that for an observation instance.
+The key value for each entity instance is computed in the same way as that for an observation instance.
 
@@
 In dealing with each row, five steps are involved. 
-The first step generates orphan measurement instances which are not connected any observation instances.
-The second step group these measurement instances according to their observation types.
-Then, for different observation types, we materialize entity instances and observation instances by either creating a new one or return an existing one.
+The first step generates orphan measurement instances which are not connected to any observation instances.
+The second step groups these measurement instances according to their observation types.
+Then, for different observation types, the third and fourth steps materialize entity instances and observation instances respectively by either creating new oned or return existing ones. 
 The last step assigns the context relationship among the different observation instances.
+
+\noindent {\bf Analysis of MaterializeDB}. \\
+{\bf Time}: As can be seen that  this algorihm scans the original data file in a row-by-row manner without revisiting the already seen rows. So, it is linear in the size of the dataset. 
+In addition, we use $EntIdx$ and $ObsIdx$ to facilitate the checking of unique entity and observation instances. Let $m$ be the number of distinct keys, each checking could take O(log(m)) time. 
+So, in total, the algorithm runs in O(n log(m)) where $m\ll n$ generally. \\
+{\bf Space}: $EntIdx$ and $ObsIdx$ are the intermdedite structures that we use in the algorithm. Since they keep the distinct entity and observation key values. The space complexity is of $O(m)$.
 
@@
 \STATE
-\FOR { (each $Row\langle A1, A2, \cdots, An\rangle \in Dataset$)}
+\FOR { (each $Row\langle A_1, A_2, \cdots, A_n\rangle \in Dataset$)}
     \STATE /* Step 1: Define measurement instances */\\
@@
     \STATE /* Step 2: Partition the measurement instances according to observation types*/\\
-        $ObsType2MeasIdx$ = PartMeas($MeasSet, A.*$);
+        $ObsType2MeasIdx$ = PartMeas($MeasSet, A.*$);  /*$ObsType2MeasIdx=\{ObsTypeId \rightarrow \{mi\}\}$*/
     \STATE
     \STATE $ContextIdx = \emptyset$;  /* Keep index $ObsTypeId \rightarrow ObsId$ to materialize context*/
-    \FOR{(each $ObsTypeId \in ObsType2MeasIdx.keys$)}
+    \FOR{(each $ObsTypeId$ in $ObsType2MeasIdx$)}
       \STATE /* Step 3: Find or create the entity instance for each observation type partition */
@@
         \STATE
-        \STATE /* Step 4 Find or create the observation instance for each observation type partition */
-        \STATE MaterializeObs($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, OBOE.*$);
+        \STATE /* Step 4: Find or create the observation instance for each observation type partition of entity $EntId$ */
+        \STATE MaterializeObs($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, A.*, OBOE.*$);
     \ENDFOR
@@
 
-{\bf Analysis of MaterializeDB}. \\
-{\bf Time}: As can be seen that  this algorihm scans the original data file in a row-by-row manner without revisiting the already seen rows. So, it is linear in the size of the dataset. 
- We use $EntIdx$ and $ObsIdx$ to facilitate the checking of unique entity and observation instances. Let $m$ be the number of unique keys, each checking could take O(log(m)) time. 
-So, in total, the algorithm runs in O(n log(m)) where $m\ll n$ generally. \\
-{\bf Space}: $EntIdx$ and $ObsIdx$ are the intermdedite structures that we use in the algorithm. Since they keep the unique entity and observation key values. The space complexity is of $O(m)$.
-
 \begin{algorithm} [htb]
@@
 \begin{algorithmic}
-    \STATE/* Create new orphan measurement instances*/
-    \STATE $MeasSet = \emptyset$; /* Keep the set of new measurement instances*/
+    \STATE/* Create new orphan measurement instances */
+    \STATE $MeasSet = \emptyset$; /* Keep the set of new measurement instances */
 	\FOR {(each $m = \langle MeasTypeId, ResAttribute, Cond, Val\rangle \in A.Map$)}
 
-		\STATE {\bf if}{(($m.ResAttribute != Row.Ai.Attrname$) OR ($Row.Ai$ does not satisfy $m.Cond$))}
+		\STATE {\bf if}{(($m.ResAttribute != Row.A_i.Attrname$) OR ($Row.A_i$ does not satisfy $m.Cond$))}
 		  {\bf continue};
-		%\STATE $ObsTypeId$ = GetObsTypeId($A.MeasType, m.MeasTypeId$);
-		%\STATE
 		\STATE $mi_{id}$ = GetNewMeasId($OBOE.Measurement$);
 		\STATE {\bf if} ($m.Val!=NULL$) $MeasVal$ = $m.Val$;
-		\STATE {\bf else} $MeasVal$ = $Row.Ai.Val$;
+		\STATE {\bf else} $MeasVal$ = $Row.A_i.Val$;
 		\STATE Create a measurement instance $\langle mi_{id}, null, MeasType, MeasVal\rangle$ and add it to $MeasSet$;
 	\ENDFOR
-    \RETURN $MeasSet$
+    \RETURN $MeasSet$;
 \end{algorithmic}
@@
      $HasKey$ = true;
-\STATE $EntType$ = GetObsEntityType ($A.ObservationType, ObsTypeId$)
+\STATE $EntType$ = GetObsEntityType ($A.ObservationType, ObsTypeId$);
 \STATE $CrtNewEntInst$ = true;
@@
 
-\begin{algorithm} [htb]
-\caption{{\bf MaterializeObs}($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, OBOE.*$)}
+\begin{algorithm} [ht]
+\caption{{\bf MaterializeObs}($ObsTypeId, EntId, ObsType2MeasIdx, ObsIdx, ContextIdx, A.*, OBOE.*$)}
 {\small 
@@
 	\IF{ ($IsObsDistinct$==true) }
-        %/*From the current observation key index get the id of the observation instance*/
 		\STATE $ObsId$ = GetObsId($ObsTypeId, KeyVal, ObsIdx$);
@@
     \STATE
-    \STATE /*Maintain the measurement instances for this observation*/
+    \STATE /* Maintain the measurement instances for this observation instance */
     \STATE $miSet$ = GetMeasInst($ObsType2MeasIdx, ObsTypeId$);
-    \STATE {\bf if} {($ObsId$ is a new one)}
-         Set the $obsId$ to each $mi \in miSet$ so that $mi$-s are not orphans;
-    \STATE {\bf else} 
-         Discard all the $mi \in miSet$;
-
+    \IF{($ObsId$ is a new one)} 
+         \STATE Set the $obsId$ to each $mi \in miSet$ so that $mi$-s are not orphans;
 	\STATE Put all the $mi \in miSet$ to $OBOE.Measurement$;
-    \STATE
+    \ELSE
+         \STATE Discard all the $mi \in miSet$;
+     \ENDIF
 
@@
 
+\newpage
 \begin{table}[htb]
@@
 \begin{example} [Example with ``key yes'' and ``distinct yes'', without ``identifying yes'']
-Take the data in Table \ref{tb:dataset1} \footnote{This is in page 6 in the powerpoint file.}
+Take the data in Table \ref{tb:dataset1} \footnote{The detailed annotation is in page 6 of Shawn's powerpoint file.}
 as an example to explain the algorithm. \\
@@
 \item Step 2 returns $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_1\}, o_2 \rightarrow \{mi_2, mi_3, mi_4\}\}$. 
- \item Step 3-4: for each observation types $o_1$ and $o_2$,  materialize entity and observation instance
+ \item Step 3-4: for  observation types $o_1$ and $o_2$,  materialize entity and observation instance
     \begin{itemize}
-    \item for $o_1$ (with associated instance $mi_1$ of type $m_1$
+    \item for $o_1$ (with associated instance $mi_1$ of type $m_1$)
     \begin{itemize}
@@
         \item Create an entity instance $\langle ei_2, Tree\rangle$; $EntIdx=\{\langle o_1, 2007\rangle\rightarrow ei_1, \langle o_2, 1\rangle\rightarrow ei_2\}$.
-        \item Create an observation $\langle oi_2, ei_2 \rangle$.
+        \item Create an observation instance $\langle oi_2, ei_2 \rangle$.
             No need to update $ObsIdx$ because $o_2$ is not identified as {\em distinct}.
@@
     \begin{itemize}
-    \item Step 1 creates measurement instances $\langle mi_5, null, Year, 2007\rangle$,
-    $\langle mi_6, null, DBH, 35.8\rangle$,\\
+    \item Step 1 creates measurement instances $\langle mi_5, null, Year, 2008\rangle$,
+    $\langle mi_6, null, DBH, 36.2\rangle$,\\
     $\langle mi_7, null, TaxonomicTypeName, Picea~rubens\rangle$,
@@
     \item Step 2 gets $ObsType2MeasIdx = \{\{o_1\rightarrow \{mi_5\}, o_2 \rightarrow \{mi_6, mi_7, mi_8\}\}$
-    \item Step 3-4: for each observation types $o_1$ and $o_2$ materialize entity and observation instance
+    \item Step 3-4: for observation types $o_1$ and $o_2$ materialize entity and observation instance
         \begin{itemize}
@@
 \begin{example} [Example with identifying]\label{eg2}
-Let us use the data in Table \ref{tb2} as an example.
+Let us use the data in Table \ref{tb2} as an example. \footnote{The detailed annotation information is at page 8 in Shawn's powerpoint file.}
 Before we go through the algorithm step by step,  we first note that  $o_1$ and $o_2$ have key measurements $m1$ and $m_2$ respectively. 
@@
 \begin{itemize}
-\item For $o_1$, the key value is $A$. Since there is no such a key in $EntIdx$, we create an entity $ei_1$ of type $Plot$.
+\item For $o_1$, the key value is $A$. Since there is no such a key in $EntIdx$, we create an entity $ei_1$ of type $Plot$.\\
 	$EntIdx = \{\langle o_1, A\rangle \rightarrow ei_1\}$. \\
 	We create an observation instance $oi_1$ whose entity is $ei_1$.\\
-	$ObsIdx = \{\langle o_1, A\rangle \rightarrow oi_1\}$.
-	Connect the measurement instances to observation instances  $mi_1 \rightarrow oi_1$,
+	$ObsIdx = \{\langle o_1, A\rangle \rightarrow oi_1\}$.\\
+	Connect the measurement instance $mi_1$ to observation instances  $oi_1$,
 \item For $o_2$, 
-	the key value is $(A, Picea~rubens)$ since it has context $o_1$ with ``{\em identifying yes}'',  we create an entity instance $ei_2$ of type $Tree$. \\
+	the key value is $(A, Picea~rubens)$ since it has context $o_1$ with ``{\em identifying yes}''.\\
+  	We create an entity instance $ei_2$ of type $Tree$. \\
 	$EntIdx = \{\langle o_1, A\rangle \rightarrow ei_1, \langle o_2, (A, Picea~rubens) \rangle \rightarrow ei_2\}$.\\
@@
 To create observation instance, since $o_1$ is specified with ``{\em distinct yes}'' and the key value is $\langle o_1,A\rangle$, 
-which corresponds to observatin instance $oi_1$. So, we do not need to create a new observation for it.\\
+which corresponds to an existing observatin instance $oi_1$. So, we do not need to create a new observation for it.\\
 When we try to connect the measurement instance $mi_4$ to observation instance,  we realize that we did not create a new observation instance for type $o_1$. 
@@
 When we process the {\bf third row}, we have a new key value $\langle o_1, B\rangle$ for $o_1$, thus we create a new entity instance for it.
-For $o_2$, we have new key value $\langle o_2, (B,~Picea~rubens)\rangle$ and create a new entity instance for it.
+For $o_2$, we have new key value $\langle o_2, (B,~Picea~rubens)\rangle$ and create a new entity instance for it. Similarly, we need to create new observation instances for both type.
 

### 2010-03-05 docs/algdraft/MaterializeDBV1.tex
Update the documents.

--- 2010-03-05 docs/algdraft/MaterializeDB.tex
@@
 \newtheorem{example}{Example}[section]
-\newtheorem{definition}{Definition}[section]
-\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
 \begin{document}
 
-\section{Key constraints}
+\section{Valid annotation constraints}
+Among the {\em key}, {\em distinct}, and {\em identifying} constraints,
+there are correlations between ``key yes'' and ``distinct yes''.
 
-%motivation to have key constraints.
-In real world, the dataset collected by an ecologist to describe
-measurents and observations is not normalized. I.e., two different
-rows in the table may have a big amount of ``redundant'' 'information
-to describe an object (e.g., entity).
-
-As discussed in \cite{***} %ER paper, 
-we can use the OBOE cenceptual model to capture the semantic
-relationships from the columns to their real meaning. 
-However, the mapping from the columns to the OBOE model still cannot
-directly solve the problem of getting concise informatin (e.g., unique
-entity instances, unique observation instances, etc.)
-
-Given Table \ref{tb:pltarea}(a), it records the fact that four times
-of dbh of some trees are taken. 
-Very often, an ecologist may ask questions like this:  Give me the
-average dbh for every different trees. 
-First, we have four rows here for trees. 
-But how many tree entity instances are here is a question.
-
-There are several cases to consider:
-\begin{itemize}
-\item Case 1: The naive extreme way to interpret the data is that each
-  observation is from different tree entity. Then, we have {\bf four}
-  tree entities. This may be too strict. People may say, well, I have
-  some observations for the same entity. 
-\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
-That's obvious that {\em piru} is different from {\em abba}. 
-With this constraint, we get {\bf two} tree entities. However, an
-obvious fact is also true that one {\em piru} tree cannot at the
-different plots. (It is another story if this tree was migrated from one plot to another.) 
-\item Case 3: Following Case 2, people want to further limit that the
-  same {\em spp} in the same {\em plt} should represent the same tree
-  entity set.  
-%To achieve this, we use {\em identifying yes}.
-%   Now, it should return
-%	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
-\end{itemize}
-
-%Q2: Why we need to distinguish the same entities using {\em key yes}
-%and {\em distinct yes}) to identify the same observation? 
-%For the example, we have one observation for splot {\em A}.
-%What's the semantic purpose of this? What kind of query may need this? 
-%E.g., how many spots in this dataset?
-
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{cc}
-\begin{tabular}{|l|l|l|}
-\hline
-plt & spp & dbh\\\hline
-A & piru & 35.8 \\\hline
-A & piru & 36.2 \\\hline
-B & piru &33.2 \\\hline
-B&abba&34\\\hline
-\end{tabular}
-&
-\begin{tabular}{|l|l|l|l|}
-\hline
-plt & area & spp & dbh\\\hline
-A & 1.0 & piru & 35.8 \\\hline
-A & 1.1 & piru & 36.2 \\\hline
-B & 2.0 &piru &33.2 \\\hline
-B& 2.0 & abba&34\\\hline
-\end{tabular}\\
-(a) & (b)
-\end{tabular}
-\end{center}
-\vspace{-0.2in}
-\caption{Dataset}
-\label{tb:pltarea}
-\end{table}
-
-Obviously, to answer this question, we not only need to figure out
-what are the observations and measurements, we also need proper
-constraints to help us understand whether different observations are
-from the same entity or not.
-
-To tackle this problem, we proposed to use {\em key yes}, {\em
-  distinct yes} and {\em identifying yes} constraints.
-
-\from{HP} {Describe or define {\em key yes},  {\em distinct yes},  {\em identifying yes}}
-
-\begin{definition}[Key measurement types]{
-The key mensurement tytpes of an observation are the measurement types
-that are used to distinguish one observation/entity instance from the
-other. }
-\end{definition}
-In most simple case, when an observation type does not have context, 
-the key measurement types are specified with ``key yes''.
-
-{\bf Key constraints}: in this report (paper),  we use the term {\em
-  key constraints} to refer to the general constraints of the above
-three types {\em key yes}, {\em distinct yes} and 
-{\em identifying yes}.
-
-\section{Valid annotation constraints}
-
-Based on the semantic meaning of these key constraints, we can see
-that they have some correlations. 
-Before we give any further examples to explain such correlations, we
-list the correlation rules among them.
-
-\begin{itemize}
-\item {\bf Rule 1}: If an observation type is specified with {\em distinct yes}, all
-  its measurement types are automatically marked with {\em key yes}. 
-\item {\bf Rule 2}: If the context of an observation type is specified with {\em identifying
-    yes}. This observation type MUST have some key measurement
-  types. And the context observation type also MUST have key
-  measurement types too. 
-\end{itemize}
-
-The following example shows an example why Rule 1 is needed. 
-Given the data in Table \ref{tb:pltarea}(b) and the following
-annotation.
+The following example gives us a hint of invalid annotations. Given the data in Table \ref{tb:pltarea}. The following annotation is {\em invalid}.
 
@@
 \verb|        | {\bf standard} ``sqft''\\
-\verb|        |$\cdots$
+\verb|       |$\cdots$ Not finished....
+\\
+According to the annotation, if two plots have the same value on EntityName, they represent the same plot observation. 
+Obviously, it has probelem to interprete the data in Table \ref{tb:pltarea} with this annotatioin. 
+E.g., the first and second row catching information about plot with EntityName $A$. Accoring to the annotation, they should be the same plot obervation.
+However, the data shows that this one plot has two different areas $1.0$ and $1.1$, so, there is confusion here.
+\\
+\begin{table}[htb]
+\begin{center}
+%\begin{tabular}{cc}
+%\begin{tabular}{|l|l|l|}
+%\hline
+%plt & spp & dbh\\\hline
+%A & piru & 35.8 \\\hline
+%A & piru & 36.2 \\\hline
+%B & piru &33.2 \\\hline
+%B&abba&34\\\hline
+%\end{tabular}
+%&
+\begin{tabular}{|l|l|l|l|}
+\hline
+plt & area & spp & dbh\\\hline
+A & 1.0 & piru & 35.8 \\\hline
+A & 1.1 & piru & 36.2 \\\hline
+B & 2.0 &piru &33.2 \\\hline
+B& 2.0 & abba&34\\\hline
+\end{tabular}\\
+%(a) & (b)
+%\end{tabular}
+\end{center}
+\vspace{-0.2in}
+\caption{Dataset}
+\label{tb:pltarea}
+\end{table}
 
-According to the annotation, if two plots have the same value for
-``EntityName'', they represent the same plot observation. 
-Obviously, it has problem to interprete the data in Table
-\ref{tb:pltarea} with this annotatioin. 
-E.g., the first and second rows catch information about plot with
-EntityName $A$. Accoring to the annotation, they should be the same
-plot obervation. 
-However, the data shows that this one plot has two different areas
-$1.0$ and $1.1$, so, there is confusion here.
+{\bf From HP:}  There are several ways to let the data comply with the annotaion without confusion. 
+Maybe too detailed, I omit this here.
 
-For this case, we can have several ways to make the annoation
-consistent. 
-The first alternative to tackle this confusion is to change the
-annotation.  We can let $m_2$ automatically becomes key measurement type of
-$o_1$. Then, every different ($m_1$, $m_2$) value pair denote a
-different observation instance which come from the different entity
-instance.
-The second alternative is to get rid of {\em distinct yes} for $o_1$,
-which means that $(A,1.0)$ and $(A,1.1)$ are different observation instances of the same entity instance
-$A$.
+Here are some very basic rules to restrict the correlations among these constraints. 
+\begin{itemize}
+\item ``Distinct yes'' on an observation type implies that all its measurements collectively form a key of this observation type. 
+In the above example, we can specify ``distinct yes'' on ``$o_1$'' without {\em explicitly} denote ``key yes'' on the measurements $m_1$ and $m_2$. 
+The semantic meaning is that ($m_1$, $m_2$) collectively form key measurement for $o_1$.  
+\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
+In this case, when two rows (for the observation type) have the same values for these key measurements, they represent the different observation instances.
+Such a case means that there may exist other ``variables'' that can distinguish one observation instance from the other, but this dataset does not catch this variable ({\bf HP: is ``variable'' the right word to use?
+}).
+So, in real application, we do not encourage such situation to appear.
+\end{itemize}
 
-%Here are some very basic rules to restrict the correlations among these constraints. 
-%\begin{itemize}
-%\item ``Distinct yes'' on an observation type implies that all its measurements collectively form a key of this observation type. 
-%In the above example, we can specify ``distinct yes'' on ``$o_1$''
-%without {\em explicitly} denote ``key yes'' on the measurements $m_1$
-%and $m_2$. 
-%The semantic meaning is that ($m_1$, $m_2$) collectively form key measurement for $o_1$.  
-%\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
-%In this case, when two rows (for the observation type) have the same
-%values for these key measurements, they represent the different
-%observation instances. Such a case means that there may exist other
-%``variables'' that can distinguish one observation instance from the
-%other, but this dataset does not catch this variable ({\bf HP: is
-%``variable'' the right word to use? }). So, in real application, we
-%do not encourage such situation to appear. 
-%\end{itemize}
+In our algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the rules among these constraints.
 
-In our materialization algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the rules among these constraints.
+{\bf From HP:} How detailed to discuss this?
 
@@
 \begin{itemize}
-\item $OBOE.Observation=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
+\item $OBOE.Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances;
 \item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;

### 2010-03-05 docs/algdraft/UseCaseKey.tex
Update the documents.

--- 2010-01-15 docs/algdraft/UseCaseKey.tex
@@
 \newtheorem{example}{Example}[section]
+\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
 \begin{document}
 
-{\bf Use cases to show that it's needed to have {\em key yes}, {\em identifying yes} and {distinct yes}}
+How to annotate the above dataset using ``key yes'', ``distinct yes''
+and ``identifying yes''.
 
-Q1: Why we need to distinguish the same entities using {\em key yes} and {\em identifying yes})?
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+Plot & sub-plot & Tmnt & Sp & Ind &wt\\\hline
+1 & A & X &Aus&1&10\\\hline
+1 & A & C &Bus&1&20\\\hline
+1 & B & X &Aus&3&10\\\hline
+1 & B & C &Bus&4&10\\\hline
+2 & A & X &Aus&1&20\\\hline
+2 & A & C &Bus&4&10\\\hline
+2 & B & X &Aus&5&20\\\hline
+2 & B & C &Bus&4&10\\\hline
+\end{tabular}
+\end{center}
+\caption{A dataset with more complex information}
+\label{tb:complexdb}
+\end{table}
 
-Assume the following tabe is the measurement for some plant tree at different spots.
+\noindent  Given the dataset in Table \ref{tb:complexdb}, users have different situations to catch. 
+\begin{itemize}
+\item Requirement 1: {\em Plot} with label ``1'' should refer to the same one physical plot (i.e., the Plot in the first 4 row means the same thing); similarly, 
+  {\em plot} with label ``2'' should refer to the second physical plot (i.e., the Plot in the last 4 row means the same thing). 
+\begin{itemize} 
+\item This can be captured in annotation by putting ``Distinct yes'' for observation type {\em Plot} and ``key yes'' for its measurement type {\em PlotLabel}. 
+\end{itemize}
+\item Requirement 2: {\em sub-plot}s with the same lable should refer to the same physical sub-plot if they are within the same plot; 
+but the sub-plot with the same label with different {\em Plot} label are different sub-plots.
+E.g., Row 1 \{Plot=1, sub-plot=A\} refers the same sub-plot as that in Row 2, but is different from the one in Row 5 \{Plot=2, sub-plot=A\}. 
+\begin{itemize} 
+\item This can be captured by putting ``Distinct yes'' for observation type {\em SubPlot} , ``key yes'' for its measurement type {\em SubPlotLabel}.
+We need to denote {\em Plot} is its context and with {\em identifying yes} specified on this context.
+\end{itemize}
+\item Requirement 3: {\em Tmnt} with the same lable should refer to the same treatment process (So that we can aggregate on different treatment process, e.g., on ``X'' or on ``C''.)  But the treatment at different sub-plot should refer to different treatment.
+ \begin{itemize} 
+\item The first requirement can  be captured by treating all the
+  Treatment with value ``X'' as the different entity instances with
+  the same type (TmntType). The second requirement can be captured by
+  treating the treatments in different sub-plots as observations of
+  the different entity instances.  I.e., treatments in row 1 and row 3
+  are of but are different observation instances which are of
+  different entity instances. 
+\item {\em Tmnt} has the context {\em sub-plot}. 
+We don't need to specify ``key yes'', ``distinct yes'', or ``identifying yes''. 
+%At the first glance, to represent this, {\em TmntType} for {\em Tmnt} should be specified with ``key yes''. 
+%It should have context {\em sub-plot} which is specified with ``identifying yes''. 
+%\item After further analysis, {\bf  one question arises: 
+%The key measurements for the treatment observation is different from the key measurement  of the entity treatment}. 
+%After considering the context, the key measurements for the treatment observation are \{Plotlabel, SubPlotLabel, TmntType\}.  
+%When two rows have the same value on these measurements, they represent the same observation instance.
+%However, the key measurement for the treatment entity is just \{TmntType\}. When two rows have the same value on it, they represent %the same entity instance. 
+%The {\em identifying} constraint can only capture the observation context. 
+%{\bf This problem is more obvious when we analyze Case 5}. 
+%\item Another different annotation may be applied to catch this semantic. E.g., treat the {\em treatments} in different rows as different entity instances. 
+%This way, the observation type and the entity type have the same key measurement types \{Plotlabel, SubPlotLabel, TmntType\}. 
+%However, this problem still exists for Case 4 and Case 5.
 
-Consider this question that a user may ask. Give me the average dbh for every {\em piru} tree (i.e., tree entity). 
-First, we have three observations here.
-But how many tree entities here is a question.
-
-There are several cases to consider:
-\begin{itemize}
-\item Case 1: The naive extreme way to interpret the data is that each observation is from different tree entity. Then, we have {\bf four} tree entities. 
-	This may be too strict. People may say, well, I have some observations for the same entity. 
-\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
-That's obvious that {\em piru} is different from {\em abba}. 
-With this constraint, we get {\bf two} tree entities. 
-\item Case 3: the assumption of case two has some obvious problem. 
-People want to further limit that the same {\em spp} in the same {\em plt} should represent the same tree entity set. 
-To achieve this, we use {\em identifying yes}.
-   Now, it should return
-	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
+Summarization questions: \\
+(1) Give me the average weight of the individuals with treatment ``X''. 
+How can this question be answered after the annotation and materialization? 
+This need to be answered after we annotate {\em Sp}, {\em Ind}, and {\em wt}. 
 \end{itemize}
 
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{cc}
-\begin{tabular}{|l|l|l|}
-\hline
-plt & spp & dbh\\\hline
-A & piru & 35.8 \\\hline
-A & piru & 36.2 \\\hline
-B & piru &33.2 \\\hline
-B&abba&34\\\hline
-\end{tabular}
-&
-\begin{tabular}{|l|l|l|l|}
-\hline
-plt & area & spp & dbh\\\hline
-A & 1.0 & piru & 35.8 \\\hline
-A & 1.1 & piru & 36.2 \\\hline
-B & &piru &33.2 \\\hline
-B& & abba&34\\\hline
-\end{tabular}\\
-(a) & (b)
-\end{tabular}
-\end{center}
-\vspace{-0.2in}
-\caption{Dataset}
-\end{table}
+\item Requirement 4:  {\em Sp} with the same name should refer to the same species (e.g., a bird named {\em Aus} flies from sub-plot (1,A) to (1,B).)  
+But the {\em Sp} with the same name at different sub-plot should refer to different observations of a specie.
+%\begin{itemize}
+%\item At the first glance, to represent this, {\em SpName} for {\em Sp} should be specified with ``key yes''. 
+%It should have context {\em sub-plot} which is specified with ``identifying yes''. 
+%\item {\bf The same problem as Case 3: The key measurements for the {\em Sp} observation is different from the key measurement  of the entity {\em Sp}}.
+%the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName\}. 
+%However, the key measurements for the species entity is just \{SpName\}. 
+%\end{itemize}
 
-Q2: Why we need to distinguish the same entities using {\em key yes} and {\em distinct yes}) to identify the same observation? 
-For the example, we have one observation for splot {\em A}.
-What's the semantic purpose of this? What kind of query may need this? 
-E.g., how many spots in this dataset?
+\item Requirement 5:  {\em Ind} with the same label and and the same species name should refer to the same species.
+But the individual (with the same label and the same species name) at different sub-plot should refer to different species observations.
+%\begin{itemize}
+%\item {\bf The same problem as Case 4 and Case 5: The key measurements for the {\em Ind} observation is different from the key measurement  of the entity {\em Ind}.}
+%the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName, Ind\}. 
+%However, the key measurement for the species entity is just \{SpName, Ind\}.  When two rows have the same value on these two columns, they represent the same entity instance. 
+%In this case, the observation context of Ind is {\em Sp} and {\em Sub-plot}. But the entity context of Ind is just {\em Sp}. 
+%\end{itemize} 
+\end{itemize}
 
-{\bf Note 1}: if one observation type is marked with {\em distinct yes}, all its measurements should be marked with {\em key yes}. 
-Otherwise, we may have the same observation with different measurement values. 
-E.g., what will happen for the following:
+The annotation can be done as follows. 
+\begin{itemize}
+\item Observation type: Plot {\em distinct yes}
+  \begin{itemize}
+  \item Measurement type: PlotLabel {\em key yes}
+  \end{itemize}
+\item Observation type: SubPlot {\em distinct yes}
+  \begin{itemize}
+    \item Measurement type: SubPlotLabel {\em key yes}
+      \item Context: Plot {\em identifying yes}. This means that the
+        key measurement for SubPlot (SubPlotLabel) and the key
+        measurement for Plot (PlotLabel) together form the key of
+        SubPlot observation (and its related entity too.)
+      \end{itemize}
+\item Observation type: Treatment {\em distinct yes}
+  \begin{itemize}
+    \item EntityType: TreatmentProcedure 
+    \item Measurement type: TmntType {\em key yes}
+   \item Context: SubPlot {\em Identifying yes}
+  \end{itemize}
+\item Observation type: SpeciesIndividual
+  \begin{itemize}
+      \item Measurement type: Species, {\em key yes}
+      \item Measurement type: IndLabel (represent {\em Individual}), {\em key yes} 
+      \item Measurement type: Weight
+      \item Context: SubPlot. Note: (1) if the users want to treat the
+        individual (with the same lavel and the same species) at
+        different sub-plots as different observation instances, here,
+        {\em identifying yes} should be specified. Otherwise, (2) if the users want to treat the
+        individual (with the same lavel and the same species) at
+        different sub-plots as SAME observation instances, here, we
+        don't need to specify {\em  identifying yes}.
+   \end{itemize} 
+\end{itemize}
 
-\noindent
-{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
-\verb|    |{\bf  entity} ``Plot''\\
-\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
-\verb|        | {\bf characteristic} ``EntityName'' \\
-\verb|        | {\bf standard} ``Nominal''\\
-\verb|    | {\bf measurement} ``m2''\\
-\verb|        | {\bf characteristic} ``area'' \\
-\verb|        | {\bf standard} ``sqft''\\
+%In summary, we can get a better idea about the problem described in
+%the above use cases can when we answer the following two simple
+%questions: \\
+%Q1:  Will an entity type and an observation type (which is of the
+%given entity type) always have the same key measurement type(s)?
+%     The above use cases give situations that the answer is no.  \\
+%Q2: is {\em identifying} itself enough to distinguish the key
+%measurement for observation types and for entity types? 
+%    My temporary answer to this question is no.
 
-Will $(A, 1.0)$ and $(A,1,1)$ be treated as the same observation? 
-According to the semantic meaning, they are the same observation because they have the same value on the key measurement and this observation type is marked witht {\em distinct yes}. 
-However, there is something wrong here.
+%A general thinking: 
+%the counterpart in RDB (Relational DataBase) is a relational scheme with key attributes. 
+%Here, we have two levels of objects: entity level and instance level.
+%Then, for different levels of objects, we need to have different ways to specify their key measurements.
 
-\begin{table}[htb]
-\begin{center}
-
-\end{center}
-\vspace{-0.2in}
-\caption{Dataset}
-\end{table}
-
-Based on this note, it seems like it is not useful to denote {\em distinct yes}.
-Basically, once all the measurements are marked with {\em key yes}, it automatically infers that it is distinct yes.
+%\newpage
 

### 2010-03-05 dev/sms/src/org/ecoinformatics/oboe/model/EntityInstance.java
add two generated data files for testing purpose

@@ -6,6 +6,7 @@ import java.io.PrintStream;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Entity;
+
 public class EntityInstance {
 	private static long gEntId=0;
 	

### 2010-03-06 dev/sms/src/org/ecoinformatics/sms/plugins/commands/ViewAnnotationCommand.java
include "ROUNDED" style for the nodes. can do much more with this - different parts of the annotation with different colors/shapes/etc.

@@ -28,8 +28,6 @@ package org.ecoinformatics.sms.plugins.commands;
 
 import java.awt.event.ActionEvent;
 
-import javax.swing.JOptionPane;
-
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.pages.ViewAnnotationPage;

### 2010-03-06 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java

@@ -15,7 +15,9 @@ import org.ecoinformatics.sms.annotation.Standard;
 
 import com.mxgraph.model.mxCell;
 import com.mxgraph.swing.mxGraphComponent;
+import com.mxgraph.util.mxConstants;
 import com.mxgraph.view.mxGraph;
+import com.mxgraph.view.mxStylesheet;
 
 public class AnnotationGraph {
 	
@@ -24,11 +26,24 @@ public class AnnotationGraph {
 		// the graph
 		mxGraph graph = new mxGraph();
 		Object parent = graph.getDefaultParent();
+		
+		//the style
+		mxStylesheet stylesheet = graph.getStylesheet();
+		Map<String, Object> styleMap = new HashMap<String, Object>();
+		styleMap.put(mxConstants.STYLE_SHAPE, mxConstants.SHAPE_ELLIPSE);
+		//styleMap.put(mxConstants.STYLE_OPACITY, 50);
+		//styleMap.put(mxConstants.STYLE_FONTCOLOR, "#774400");
+		stylesheet.putCellStyle("ROUNDED", styleMap);
+		graph.setStylesheet(stylesheet);
+		
+		// for positioning
 		int x = 20;
 		int y = 20;
 		int width = 80;
 		int height = 30;
+		String style = "defaultVertex;ROUNDED";
 		
+		// keep track of the observaiton cells
 		Map<Observation, mxCell> observationMap = new HashMap<Observation, mxCell>();
 		
 		for (Observation observation: annotation.getObservations()) {
@@ -43,14 +58,14 @@ public class AnnotationGraph {
 
 				// add observation
 				Object observationNode = 
-					graph.insertVertex(observationCell, null, observation.getLabel(), x, y, width, height);
-				
+					graph.insertVertex(observationCell, null, observation.getLabel(), x, y, width, height, style);
+
 				// shift down
 				//y += (height*2);
 				
 				// of entity
 				Object entityNode = 
-					graph.insertVertex(observationCell, null, observation.getEntity(), x, y + (height*2), width, height);
+					graph.insertVertex(observationCell, null, observation.getEntity(), x, y + (height*2), width, height, style);
 				graph.insertEdge(observationCell, null, "ofEntity", observationNode, entityNode);
 
 				for (Measurement measurement: observation.getMeasurements()) {
@@ -60,7 +75,7 @@ public class AnnotationGraph {
 					
 					// add measurement
 					Object measurementNode = 
[…]

### 2010-03-31 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ComplexQueryPage.java
consolidate rendering/editing UI widgets into single class - now the editor and renderer use different instances of the same CriteriaPanel class. add any/all checkbox for conditions (like subcriteria). http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4905

@@ -34,6 +34,7 @@ import java.util.List;
 
 import javax.swing.AbstractAction;
 import javax.swing.BoxLayout;
+import javax.swing.JCheckBox;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 
@@ -72,13 +73,13 @@ public class ComplexQueryPage extends AbstractUIPage {
 		
 	// query list
 	private CustomList queryList;
-	private JLabel queryListLabel;
 	
 	// the final query
 	private Query query;
 	
 	// docids from the annotation query
 	private List<String> docids = new ArrayList<String>();
+	private JCheckBox anyAll;
 	
 
 	// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
@@ -107,12 +108,9 @@ public class ComplexQueryPage extends AbstractUIPage {
 		
 		this.add(WidgetFactory.makeDefaultSpacer());
 		
-		// Query list
-		JPanel queryListPanel = WidgetFactory.makePanel();
-		queryListLabel = WidgetFactory.makeLabel("Conditions:", false);
-		queryListPanel.add(queryListLabel);
+		// Query list		
 		String[] colNames = new String[] {"Conditions"};
-		Object[] editors = new Object[] {new CriteriaRenderer(1) };
+		Object[] editors = new Object[] {new CriteriaRenderer(true) };
 		queryList = WidgetFactory.makeList(
 				colNames, 
 				editors, 
@@ -138,34 +136,15 @@ public class ComplexQueryPage extends AbstractUIPage {
 				
 			}
 		});
-		/**
-		// edit
-		queryList.setCustomEditAction(new AbstractAction() {
-			public void actionPerformed(ActionEvent e) {
-				List rowList = queryList.getSelectedRowList();
-				AnnotationQueryPage aqp = (AnnotationQueryPage) rowList.get(0);
-				
-				// show the dialog
-				ModalDialog dialog = 
-					new ModalDialog(
-							aqp, 
-							UIController.getInstance().getCurrentActiveWindow(), 
-							UISettings.POPUPDIALOG_WIDTH,
-							UISettings.POPUPDIALOG_HEIGHT);
 
-				// get the response back
-				if (dialog.USER_RESPONSE == ModalDialog.OK_OPTION) {
-					// make sure the UI reflects the changes
-					queryList.revalidate();
-					queryList.repaint();
+		// the any/all checkbox
[…]

### 2010-03-31 dev/sms/src/org/ecoinformatics/sms/plugins/search/CriteriaPanel.java

@@ -29,19 +29,18 @@ public class CriteriaPanel extends JPanel {
 	private Criteria criteria;
 	
 	private JPanel criteriaPanel;
-	private JPanel subcriteriaPanel;
 	private JComboBox subject;
 	private JComboBox condition;
 	private OntologyClassField value;
 	
+	private JPanel subcriteriaPanel;
 	private JCheckBox anyAll;
 	private CustomList subCriteria;
 	
-	public CriteriaPanel(Criteria criteria) {
+	
+	public CriteriaPanel(boolean isGroup) {
 		super();
-		
-		this.criteria = criteria;
-		
+				
 		this.setLayout(new BoxLayout(this, BoxLayout.Y_AXIS));
 		
 		// change the filter class based on what is selected 
@@ -75,12 +74,13 @@ public class CriteriaPanel extends JPanel {
 		this.add(criteriaPanel);
 		
 		subcriteriaPanel = WidgetFactory.makePanel(5);
+		anyAll = WidgetFactory.makeCheckBox("Match All", false);
 
-		if (criteria.isGroup()) {
-			
-			anyAll = WidgetFactory.makeCheckBox("Match All", false);
+		// only create the subcriteria list if this is a grouping criteria (infinite loop possible otherwise)
+		if (isGroup) {
 			String[] colNames = new String[] {"Subcriteria"};
-			Object[] editors = new Object[] {new CriteriaRenderer(0) };
+			// non-group criteria are rendered by this list
+			Object[] editors = new Object[] {new CriteriaRenderer(false) };
 			subCriteria = WidgetFactory.makeList(
 					colNames, 
 					editors, 
@@ -92,10 +92,11 @@ public class CriteriaPanel extends JPanel {
 					false, //showMoveUpButton, 
 					false //showMoveDownButton
 					);		
-			// add
+			// add action
 			subCriteria.setCustomAddAction(new AbstractAction() {
 				public void actionPerformed(ActionEvent e) {
-		
+					
+					// we add non-group criteria at this point
 					Criteria criteria = new Criteria();
 					criteria.setGroup(false);
 					
@@ -105,12 +106,60 @@ public class CriteriaPanel extends JPanel {
 					
 				}
 			});
+			
+			// add the subcriteria widgets to the panel
 			subcriteriaPanel.setLayout(new BoxLayout(subcriteriaPanel, BoxLayout.X_AXIS));
-
 			subcriteriaPanel.add(anyAll);
[…]

### 2010-03-31 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ComplexQueryPage.java
gather Criteria for generating query. now it actually produces a real - albeit simple - query http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4905

@@ -176,58 +176,44 @@ public class ComplexQueryPage extends AbstractUIPage {
 		
 		for (int i = 0; i < queryList.getListOfRowLists().size(); i++) {
 			
-			// get the query page
+			// get the Criteria object from the list
 			List rowList = (List) queryList.getListOfRowLists().get(i);
-			AnnotationQueryPage aqp = (AnnotationQueryPage) rowList.get(0);
+			Criteria criteria = (Criteria) rowList.get(0);
 			
-			Characteristic currentCharacteristic;
-			Standard currentStandard;
-			Protocol currentProtocol;
-			Entity currentEntity;
-			Triple context;
-			
-			// get the values form the page
-			try {
-				currentCharacteristic = new Characteristic(
-						aqp.getSimpleAnnotationPanel().getObservationCharacteristic().getURI());
-				characteristics.add(currentCharacteristic);
-			} catch (Exception e) {
-				currentCharacteristic = null;
+			if (criteria.isGroup()) {
+				for (Criteria subcriteria: criteria.getSubCriteria()) {
+					// what criteria were given?
+					OntologyClass subject = subcriteria.getSubject();
+					OntologyClass value = subcriteria.getValue();
+					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Entity.class))) {
+						entities.add(value);
+					}
+					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class))) {
+						characteristics.add(value);
+					}
+					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Standard.class))) {
+						standards.add(value);
+					}
+					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Protocol.class))) {
+						protocols.add(value);
+					}
+					
+					// TODO: get context from the Criteria
+					Triple context = null;
+					if (context != null) {
+						contexts.add(context);
+					}
+				}
 			}
-	
-			try {
-				currentStandard = new Standard(aqp.getSimpleAnnotationPanel().getObservationStandard().getURI());
-				standards.add(currentStandard);
-			} catch (Exception e) {
-				currentStandard = null;
-			}
-			
-			try {
-				currentProtocol = new Protocol(aqp.getSimpleAnnotationPanel().getObservationProtocol().getURI());
-				protocols.add(currentProtocol);
[…]

### 2010-03-31 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ComplexQueryPage.java
gather Context Triples from the Criteria for the query. there's still no way to specify them in this UI version, but if they were there, they'd be used in crafting the query behind the scenes. http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4905

@@ -123,7 +123,7 @@ public class ComplexQueryPage extends AbstractUIPage {
 				false //showMoveDownButton
 				);		
 		// add
-		queryList.setCustomAddAction(new AbstractAction() {
+		queryList.setCustomAddAction(new AbstractAction("Add Group") {
 			public void actionPerformed(ActionEvent e) {
 	
 				Criteria criteria = new Criteria();
@@ -182,26 +182,32 @@ public class ComplexQueryPage extends AbstractUIPage {
 			
 			if (criteria.isGroup()) {
 				for (Criteria subcriteria: criteria.getSubCriteria()) {
-					// what criteria were given?
-					OntologyClass subject = subcriteria.getSubject();
-					OntologyClass value = subcriteria.getValue();
-					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Entity.class))) {
-						entities.add(value);
-					}
-					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class))) {
-						characteristics.add(value);
-					}
-					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Standard.class))) {
-						standards.add(value);
-					}
-					if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Protocol.class))) {
-						protocols.add(value);
-					}
 					
-					// TODO: get context from the Criteria
-					Triple context = null;
-					if (context != null) {
-						contexts.add(context);
+					// handle context
+					if (criteria.isContext()) {
+						Triple context = criteria.getContextTriple();
+						if (context != null) {
+							contexts.add(context);
+						}
+					} else {
+						// what criteria were given?
+						OntologyClass subject = subcriteria.getSubject();
+						OntologyClass value = subcriteria.getValue();
+						if (value == null) {
+							continue;
+						}
+						if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Entity.class))) {
+							entities.add(value);
+						}
+						if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class))) {
+							characteristics.add(value);
+						}
+						if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Standard.class))) {
+							standards.add(value);
+						}
[…]

### 2010-03-31 dev/sms/src/org/ecoinformatics/sms/plugins/search/Criteria.java

@@ -2,6 +2,7 @@ package org.ecoinformatics.sms.plugins.search;
 
 import java.util.List;
 
+import org.ecoinformatics.sms.annotation.Triple;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 
 public class Criteria  {
@@ -10,6 +11,9 @@ public class Criteria  {
 	private String condition;
 	private OntologyClass value;
 	
+	private boolean context = false;
+	private Triple contextTriple;
+
 	private boolean group = false;
 	private boolean all = true;
 	private List<Criteria> subCriteria;
@@ -50,6 +54,18 @@ public class Criteria  {
 	}
 	public void setAll(boolean all) {
 		this.all = all;
+	}
+	public boolean isContext() {
+		return context;
+	}
+	public void setContext(boolean context) {
+		this.context = context;
+	}
+	public Triple getContextTriple() {
+		return contextTriple;
+	}
+	public void setContextTriple(Triple contextTriple) {
+		this.contextTriple = contextTriple;
 	}
 	public String toString() {
 		StringBuffer sb = new StringBuffer();

### 2010-04-01 docs/algdraft/model.tex
minor change to the model file.

@@
 
-\section{How to evaluate the quality of a conceptual model}
+\section{How to evaluate the quality of a conceptual model (CM)}
 
@@
 (e.g., OBOE, O\&M, etc) is a good one, or is better than the other. 
-Unfortunately, the literature \cite{DBLP:journals/dke/Moody05, ***} shows that
+Unfortunately, the literature \cite{DBLP:journals/dke/Moody05} shows that
 there is no well-formed standard way to evaluae the quality of a conceptual model. 
@@
 \begin{tabular}{|l|l|l|l|}
-\hline 
-ER model & OBOE model term & O\&M term &others\\\hliine
+\hline ER model & OBOE model term & O\&M term &others\\\hline
+\\\hline
+                      & Entity & &\\\hline
 Entity             & Observation & Feature of interest & \\\hline
@@
                        & Standard & ? & \\\hline
+Relationship   & Context & ? & \\\hline
+Value              & Standard Value & ? & \\\hline
 \end{tabular}
@@
 
-In O\&M, an {\em Observation} is defined as an action. 
+In OBOE, an {\em observation} represents any {\em measurement} of some {\em characteristic} 
+(attribute) of some real-world entity or phenomenon.
+A {\em measurement} consists of a realized value of some characteristic of an {\em entity}, 
+expressed in some well-specified units (drawn from a measurement standard) 
+Observations can provide {\em context} for other observations  
+(e.g. observations of spatial or temporal information would often provide context for some other observation).
+
+In O\&M, 
+An {\em observation} is an {\em action} with a {\em result} which has a value describing some phenomenon.  
+The observation is modelled as a {\em Feature} within the context of the General Feature Model.  
+An observation feature binds a {\em result} to a {\em feature of interest}, upon which the observation was made.  
+The observed {\em property} is a property of the feature of interest. 
+An observation uses a {\em procedure} to determine the value of the result, 
+which may involve a sensor or observer, analytical procedure, simulation or other numerical process.  
+The observation pattern and feature is primarily useful for capturing metadata associated 
+with the estimation of feature properties, which is important  particularly when error in this estimate is of interest.
+
+An {\em Observation} is defined as an action. 
 It is modeled as a {\em Feature}. 
@@
 \bibliographystyle{abbrv}
-\bibliography{model.bib}
+\bibliography{model}
 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/search/Criteria.java
refactor Criteria class to non-plugin package

@@ -1,4 +1,4 @@
-package org.ecoinformatics.sms.plugins.search;
+package org.ecoinformatics.sms.annotation.search;
 
 import java.util.List;
 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ComplexQueryPage.java

@@ -45,9 +45,9 @@ import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
 import org.ecoinformatics.sms.annotation.Triple;
+import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.plugins.AnnotationPlugin;
-import org.ecoinformatics.sms.plugins.search.Criteria;
 import org.ecoinformatics.sms.plugins.search.CriteriaRenderer;
 
 import edu.ucsb.nceas.morpho.Morpho;

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/search/CriteriaPanel.java

@@ -17,6 +17,7 @@ import org.ecoinformatics.sms.annotation.Characteristic;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
+import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.ui.OntologyClassField;

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/search/CriteriaRenderer.java

@@ -8,6 +8,8 @@ import javax.swing.JTable;
 import javax.swing.table.TableCellEditor;
 import javax.swing.table.TableCellRenderer;
 
+import org.ecoinformatics.sms.annotation.search.Criteria;
+
 public class CriteriaRenderer 
 	extends AbstractCellEditor 
 		implements 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/AnnotationManager.java
include getMatchingAnnotations(Criteria) method

@@ -34,7 +34,9 @@ package org.ecoinformatics.sms;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Triple;
+import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
+
 import java.util.List;
 import java.io.InputStream;
 import java.io.OutputStream;
@@ -130,6 +132,13 @@ public interface AnnotationManager {
    public List<OntologyClass> getOntologyClasses(Annotation a,
            boolean addSubclasses, boolean addSuperclasses);
 
+   /**
+    * Get Annotations that match the given Criteria
+    * @param criteria Criteria to match
+    * @return list of matching annotations
+    */
+   public List<Annotation> getMatchingAnnotations(Criteria criteria);
+   
    /**
     * Get annotations that contain an entity in the given list and a 
     * measurement with a characteristic and standard in the given lists

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/DefaultAnnotationManager.java

@@ -35,6 +35,7 @@ package org.ecoinformatics.sms.annotation;
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.OntologyManager;
 import org.ecoinformatics.sms.AnnotationManager;
+import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 
 import java.util.HashMap;
@@ -234,6 +235,11 @@ public class DefaultAnnotationManager implements AnnotationManager {
             result.add(c);
       return result;
    }
+   
+   public List<Annotation> getMatchingAnnotations(Criteria criteria) {
+	   return getAnnotations();
+   }
+
 
    /**
     * Get annotations that contain an entity in the given list and a 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/commands/CompoundAnnotationSearchCommand.java
use the newer "Complex" query page

@@ -42,6 +42,7 @@ import java.util.List;
 import java.util.Vector;
 import java.awt.event.ActionEvent;
 
+import org.ecoinformatics.sms.plugins.pages.ComplexQueryPage;
 import org.ecoinformatics.sms.plugins.pages.CompoundQueryPage;
 import org.ecoinformatics.sms.plugins.search.ResultSetComparator;
 
@@ -51,7 +52,7 @@ import org.ecoinformatics.sms.plugins.search.ResultSetComparator;
 public class CompoundAnnotationSearchCommand implements Command {
 
 
-	private CompoundQueryPage cqp;
+	private ComplexQueryPage cqp;
 
 	/**
 	 * Constructor of AnnotationSearchCommand
@@ -75,7 +76,7 @@ public class CompoundAnnotationSearchCommand implements Command {
 		if (morphoFrame != null) {
 			// easier to perform searches over and over
 			if (cqp == null) {
-				cqp = new CompoundQueryPage();
+				cqp = new ComplexQueryPage();
 			}
 			// show the dialog
 			ModalDialog dialog = 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/pages/CompoundQueryPage.java

@@ -58,6 +58,12 @@ import edu.ucsb.nceas.morpho.query.Query;
 import edu.ucsb.nceas.morpho.util.UISettings;
 import edu.ucsb.nceas.utilities.OrderedMap;
 
+/**
+ * @deprecated in favor of the ComplexQueryPage
+ * @see ComplexQueryPage		
+ * @author leinfelder
+ *
+ */
 public class CompoundQueryPage extends AbstractUIPage {
 
 	// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/DefaultAnnotationManager.java
implement the getMatchingAnnotations(Criteria) method

@@ -37,6 +37,7 @@ import org.ecoinformatics.sms.OntologyManager;
 import org.ecoinformatics.sms.AnnotationManager;
 import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import java.util.HashMap;
 import java.util.List;
@@ -237,7 +238,16 @@ public class DefaultAnnotationManager implements AnnotationManager {
    }
    
    public List<Annotation> getMatchingAnnotations(Criteria criteria) {
-	   return getAnnotations();
+	   //let the fun begin!
+	   List<Annotation> results = new ArrayList<Annotation>();
+	   for(Annotation annotation : getAnnotations()) {
+		   int match = matchesCriteria(annotation, criteria);
+		   if (match < 0) {
+			   continue;
+		   }
+		   results.add(annotation);
+	   }
+	   return results;
    }
 
 
@@ -842,6 +852,76 @@ public class DefaultAnnotationManager implements AnnotationManager {
          return result;
       addSuperclasses(result);
       return result;
+   }
+   
+   /** 
+    * This method is called recursively on the subcriteria, building up a weighted return
+    * @param a the Annotation to be evaluated for matching
+    * @param criteria potentially complex 
+    * @return
+    */
+   private int matchesCriteria(Annotation a, Criteria criteria) {
+	   
+	   //simple first case
+	   if (!criteria.isGroup()) {
+		   // everything in a list
+			List<OntologyClass> characteristics = new ArrayList<OntologyClass>();
+			List<OntologyClass> standards = new ArrayList<OntologyClass>();
+			List<OntologyClass> protocols = new ArrayList<OntologyClass>();
+			List<OntologyClass> entities = new ArrayList<OntologyClass>();
+			List<Triple> contexts = new ArrayList<Triple>();
+			
+			// try context
+			if (criteria.isContext()) {
+				Triple context = criteria.getContextTriple();
+				if (context != null) {
+					contexts.add(context);
+					return hasMatchingContext(a, contexts);
+				}
+			} else {
+				// what criteria was given?
+				OntologyClass subject = criteria.getSubject();
+				OntologyClass value = criteria.getValue();
+				if (value == null) {
+					return 0;
+				}
[…]

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ComplexQueryPage.java

@@ -40,13 +40,7 @@ import javax.swing.JPanel;
 
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.annotation.Annotation;
-import org.ecoinformatics.sms.annotation.Characteristic;
-import org.ecoinformatics.sms.annotation.Entity;
-import org.ecoinformatics.sms.annotation.Protocol;
-import org.ecoinformatics.sms.annotation.Standard;
-import org.ecoinformatics.sms.annotation.Triple;
 import org.ecoinformatics.sms.annotation.search.Criteria;
-import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.search.CriteriaRenderer;
 
@@ -167,69 +161,23 @@ public class ComplexQueryPage extends AbstractUIPage {
 
 	private void generateQuery() {
 
-		List<OntologyClass> characteristics = new ArrayList<OntologyClass>();
-		List<OntologyClass> standards = new ArrayList<OntologyClass>();
-		List<OntologyClass> protocols = new ArrayList<OntologyClass>();
-		List<OntologyClass> entities = new ArrayList<OntologyClass>();
-		List<Triple> contexts = new ArrayList<Triple>();
-
+		// holds the subcriteria in the customlist
+		Criteria criteria = new Criteria();
+		criteria.setGroup(true);
+		criteria.setAll(anyAll.isSelected());
+		List<Criteria> subCriteria= new ArrayList<Criteria>();
 		
 		for (int i = 0; i < queryList.getListOfRowLists().size(); i++) {
-			
 			// get the Criteria object from the list
 			List rowList = (List) queryList.getListOfRowLists().get(i);
-			Criteria criteria = (Criteria) rowList.get(0);
-			
-			if (criteria.isGroup()) {
-				for (Criteria subcriteria: criteria.getSubCriteria()) {
-					
-					// handle context
-					if (criteria.isContext()) {
-						Triple context = criteria.getContextTriple();
-						if (context != null) {
-							contexts.add(context);
-						}
-					} else {
-						// what criteria were given?
-						OntologyClass subject = subcriteria.getSubject();
-						OntologyClass value = subcriteria.getValue();
-						if (value == null) {
-							continue;
-						}
[…]

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/Annotation.java
refactor the mapping from annotation classes (Java Class) to OBOE class (OntologyClass) since the DefaultAnnotationManager now references this and I don't want a dependency on the AnnotationPlugin

@@ -32,8 +32,10 @@
  */
 package org.ecoinformatics.sms.annotation;
 
+import java.util.HashMap;
 import java.util.List;
 import java.util.ArrayList;
+import java.util.Map;
 import java.io.OutputStream;
 import java.io.ByteArrayOutputStream;
 import java.io.InputStream;
@@ -359,6 +361,12 @@ public class Annotation {
    private List<Ontology> _ontologies = new ArrayList();
    private List<Observation> _observations = new ArrayList();
    private List<Mapping> _mappings = new ArrayList();
+   
+   /**
+    * Maps an Annotation Java Class to a specific OBOE OntologyClass
+    */
+   public static Map<Class,OntologyClass> OBOE_CLASSES = new HashMap<Class, OntologyClass>();
+
    public static String ANNOTATION_NS =
       "http://ecoinformatics.org/sms/annotation.0.9";
 

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/DefaultAnnotationManager.java

@@ -37,7 +37,6 @@ import org.ecoinformatics.sms.OntologyManager;
 import org.ecoinformatics.sms.AnnotationManager;
 import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
-import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import java.util.HashMap;
 import java.util.List;
@@ -885,19 +884,19 @@ public class DefaultAnnotationManager implements AnnotationManager {
 				if (value == null) {
 					return 0;
 				}
-				if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Entity.class))) {
+				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Entity.class))) {
 					entities.add(value);
 					return hasMatchingEntity(a, entities);
 				}
-				if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class))) {
+				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Characteristic.class))) {
 					characteristics.add(value);
 					return hasMatchingCharacteristic(a, characteristics);
 				}
-				if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Standard.class))) {
+				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Standard.class))) {
 					standards.add(value);
 					return hasMatchingStandard(a, standards);
 				}
-				if (subject != null && subject.equals(AnnotationPlugin.OBOE_CLASSES.get(Protocol.class))) {
+				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Protocol.class))) {
 					protocols.add(value);
 					return hasMatchingProtocol(a, protocols);
 				}

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java

@@ -26,10 +26,8 @@ import java.io.File;
 import java.io.FileInputStream;
 import java.io.InputStream;
 import java.io.StringReader;
-import java.util.HashMap;
 import java.util.Hashtable;
 import java.util.List;
-import java.util.Map;
 import java.util.Vector;
 import java.util.Map.Entry;
 
@@ -101,10 +99,7 @@ public class AnnotationPlugin
 
     public static final String PHYSICAL_URI_TAG_NAME = "physicalURI";
     
-    // for filtering
-    public static Map<Class,OntologyClass> OBOE_CLASSES = new HashMap<Class, OntologyClass>();
-
-	private MorphoFrame morphoFrame = null;
+    private MorphoFrame morphoFrame = null;
 	
 	private GUIAction annotateAction = null;
 
@@ -383,7 +378,7 @@ public class AnnotationPlugin
 		}
 		// initialize the static classes
 		Hashtable<String, String> oboeClasses = Morpho.getConfiguration().getHashtable("oboeClasses", "className", "classURI");
-		OBOE_CLASSES.clear();
+		Annotation.OBOE_CLASSES.clear();
 		for (Entry<String, String> entry: oboeClasses.entrySet()) {
 			String name = entry.getKey();
 			String uri = entry.getValue();
@@ -396,7 +391,7 @@ public class AnnotationPlugin
 					oboeClass = new OntologyClass();
 				}
 				oboeClass.setURI(uri);
-				OBOE_CLASSES.put(objectClass.getClass(), oboeClass);
+				Annotation.OBOE_CLASSES.put(objectClass.getClass(), oboeClass);
 			} catch (Exception e) {
 				// TODO Auto-generated catch block
 				e.printStackTrace();

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/commands/DirectAnnotationCommand.java

@@ -243,13 +243,13 @@ public class DirectAnnotationCommand implements Command {
 				// we don't edit this directly
 				return false;
 			case AnnotationTableModel.OBSERVATION_ROW:
-				ontologyPage.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Entity.class));
+				ontologyPage.setFilterClass(Annotation.OBOE_CLASSES.get(Entity.class));
 				if (currentObservation != null) {
 					currentClass = currentObservation.getEntity();
 				}
 				break;
 			case AnnotationTableModel.CHARACTERISTIC_ROW:
-				ontologyPage.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class));
+				ontologyPage.setFilterClass(Annotation.OBOE_CLASSES.get(Characteristic.class));
 				if (currentMeasurement != null) {
 					if (!currentMeasurement.getCharacteristics().isEmpty()) {
 						currentCharacteristic = currentMeasurement.getCharacteristics().get(0);
@@ -258,14 +258,14 @@ public class DirectAnnotationCommand implements Command {
 				}
 				break;
 			case AnnotationTableModel.STANDARD_ROW:
-				ontologyPage.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Standard.class));
+				ontologyPage.setFilterClass(Annotation.OBOE_CLASSES.get(Standard.class));
 				if (currentMeasurement != null) {
 					currentStandard = currentMeasurement.getStandard();
 					currentClass = currentStandard;
 				}
 				break;
 			case AnnotationTableModel.PROTOCOL_ROW:
-				ontologyPage.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Protocol.class));
+				ontologyPage.setFilterClass(Annotation.OBOE_CLASSES.get(Protocol.class));
 				if (currentMeasurement != null) {
 					currentProtocol = currentMeasurement.getProtocol();
 					currentClass = currentProtocol;

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AddContextPage.java

@@ -36,10 +36,10 @@ import javax.swing.JComboBox;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 
+import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Context;
 import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.annotation.Relationship;
-import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.ui.OntologyClassField;
 
 import edu.ucsb.nceas.morpho.framework.AbstractUIPage;
@@ -175,7 +175,7 @@ public class AddContextPage extends AbstractUIPage {
 
 		// context
 		contextRelationship = OntologyClassField.makeLabel("<relationship>", false, null);
-		contextRelationship.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Relationship.class));
+		contextRelationship.setFilterClass(Annotation.OBOE_CLASSES.get(Relationship.class));
 		contextPanel.add(contextRelationship);
 		
 		contextPanel.add(WidgetFactory.makeLabel(" the ", false, null));

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/search/CriteriaPanel.java

@@ -13,13 +13,13 @@ import javax.swing.JCheckBox;
 import javax.swing.JComboBox;
 import javax.swing.JPanel;
 
+import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
 import org.ecoinformatics.sms.annotation.search.Criteria;
 import org.ecoinformatics.sms.ontology.OntologyClass;
-import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.ui.OntologyClassField;
 
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.CustomList;
@@ -52,10 +52,10 @@ public class CriteriaPanel extends JPanel {
 			}
 		};
 		Object[] subjectValues = new OntologyClass[] {
-				AnnotationPlugin.OBOE_CLASSES.get(Entity.class), 
-				AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class), 
-				AnnotationPlugin.OBOE_CLASSES.get(Standard.class), 
-				AnnotationPlugin.OBOE_CLASSES.get(Protocol.class)};
+				Annotation.OBOE_CLASSES.get(Entity.class), 
+				Annotation.OBOE_CLASSES.get(Characteristic.class), 
+				Annotation.OBOE_CLASSES.get(Standard.class), 
+				Annotation.OBOE_CLASSES.get(Protocol.class)};
 		subject = WidgetFactory.makePickList(subjectValues, false, 0, subjectListener);
 		
 		Object[] conditionValues = new String[] {"is", "is not"};

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/ui/ContextTriplePanel.java

@@ -32,10 +32,10 @@ import javax.swing.BoxLayout;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 
+import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Relationship;
 import org.ecoinformatics.sms.annotation.Triple;
-import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WidgetFactory;
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WizardSettings;
@@ -74,21 +74,21 @@ public class ContextTriplePanel extends JPanel {
 		
 		// entity A
 		entityA = OntologyClassField.makeLabel("", false, null);
-		entityA.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Entity.class));
+		entityA.setFilterClass(Annotation.OBOE_CLASSES.get(Entity.class));
 		contextPanel.add(entityA);
 		
 		contextPanel.add(WidgetFactory.makeLabel(" was ", false, null));
 
 		// context
 		contextRelationship = OntologyClassField.makeLabel("", false, null);
-		contextRelationship.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Relationship.class));
+		contextRelationship.setFilterClass(Annotation.OBOE_CLASSES.get(Relationship.class));
 		contextPanel.add(contextRelationship);
 		
 		contextPanel.add(WidgetFactory.makeLabel(" the ", false, null));
 		
 		// entity B
 		entityB = OntologyClassField.makeLabel("", false, null);
-		entityB.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Entity.class));
+		entityB.setFilterClass(Annotation.OBOE_CLASSES.get(Entity.class));
 		contextPanel.add(entityB);
 		
 		contextPanel.setBorder(new javax.swing.border.EmptyBorder(0, 0, 0,

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/ui/SimpleAnnotationPanel.java

@@ -38,6 +38,7 @@ import javax.swing.BoxLayout;
 import javax.swing.JPanel;
 
 import org.ecoinformatics.sms.SMS;
+import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Measurement;
@@ -47,7 +48,6 @@ import org.ecoinformatics.sms.ontology.Ontology;
 import org.ecoinformatics.sms.ontology.OntologyClass;
 import org.ecoinformatics.sms.ontology.OntologyObjectProperty;
 import org.ecoinformatics.sms.ontology.OntologyProperty;
-import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WidgetFactory;
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WizardSettings;
@@ -91,7 +91,7 @@ public class SimpleAnnotationPanel extends JPanel {
 		JPanel entityLabelPanel = WidgetFactory.makePanel(2);
 		entityLabelPanel.add(WidgetFactory.makeLabel("Entity:", false));
 		observationEntity = OntologyClassField.makeLabel("", false, null);
-		observationEntity.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Entity.class));
+		observationEntity.setFilterClass(Annotation.OBOE_CLASSES.get(Entity.class));
 		entityLabelPanel.add(observationEntity);
 		
 		entityPanel.add(entityLabelPanel);
@@ -107,7 +107,7 @@ public class SimpleAnnotationPanel extends JPanel {
 		characteristicLabelPanel.add(WidgetFactory.makeLabel("Characteristic:",
 				false));
 		observationCharacteristic = OntologyClassField.makeLabel("", false, null);
-		observationCharacteristic.setFilterClass(AnnotationPlugin.OBOE_CLASSES.get(Characteristic.class));
+		observationCharacteristic.setFilterClass(Annotation.OBOE_CLASSES.get(Characteristic.class));
 		characteristicLabelPanel.add(observationCharacteristic);
 		
 		characteristicPanel.add(characteristicLabelPanel);
@@ -122,7 +122,7 @@ public class SimpleAnnotationPanel extends JPanel {
 		JPanel standardLabelPanel = WidgetFactory.makePanel(2);
[…]

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/ontology/OntologyClass.java
prevent NPEs when testing for equality

@@ -131,7 +131,7 @@ public class OntologyClass {
         if(!(obj instanceof OntologyClass))
             return false;
         OntologyClass c = (OntologyClass) obj;
-        if(c.getURI().equals(getURI()))
+        if(c != null && c.getURI().equals(getURI()))
             return true;
         return false;
     }

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/DefaultAnnotationManager.java
refactored to remove the dependency that Criteria and the AnnotationManager had on the OBOE classes map. While it's still a useful map (for UI), the DefaultAnnotationManager can still function (search) without it. http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4905

@@ -879,24 +879,24 @@ public class DefaultAnnotationManager implements AnnotationManager {
 				}
 			} else {
 				// what criteria was given?
-				OntologyClass subject = criteria.getSubject();
+				Class subject = criteria.getSubject();
 				OntologyClass value = criteria.getValue();
 				if (value == null) {
 					return 0;
 				}
-				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Entity.class))) {
+				if (subject != null && subject.equals(Entity.class)) {
 					entities.add(value);
 					return hasMatchingEntity(a, entities);
 				}
-				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Characteristic.class))) {
+				if (subject != null && subject.equals(Characteristic.class)) {
 					characteristics.add(value);
 					return hasMatchingCharacteristic(a, characteristics);
 				}
-				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Standard.class))) {
+				if (subject != null && subject.equals(Standard.class)) {
 					standards.add(value);
 					return hasMatchingStandard(a, standards);
 				}
-				if (subject != null && subject.equals(Annotation.OBOE_CLASSES.get(Protocol.class))) {
+				if (subject != null && subject.equals(Protocol.class)) {
 					protocols.add(value);
 					return hasMatchingProtocol(a, protocols);
 				}

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/annotation/search/Criteria.java

@@ -7,7 +7,7 @@ import org.ecoinformatics.sms.ontology.OntologyClass;
 
 public class Criteria  {
 
-	private OntologyClass subject;
+	private Class subject;
 	private String condition;
 	private OntologyClass value;
 	
@@ -18,10 +18,10 @@ public class Criteria  {
 	private boolean all = true;
 	private List<Criteria> subCriteria;
 	
-	public OntologyClass getSubject() {
+	public Class getSubject() {
 		return subject;
 	}
-	public void setSubject(OntologyClass subject) {
+	public void setSubject(Class subject) {
 		this.subject = subject;
 	}
 	public String getCondition() {

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/search/CriteriaPanel.java

@@ -5,7 +5,9 @@ import java.awt.event.ActionEvent;
 import java.awt.event.ItemEvent;
 import java.awt.event.ItemListener;
 import java.util.ArrayList;
+import java.util.Iterator;
 import java.util.List;
+import java.util.Map.Entry;
 
 import javax.swing.AbstractAction;
 import javax.swing.BoxLayout;
@@ -51,6 +53,7 @@ public class CriteriaPanel extends JPanel {
 				value.setFilterClass((OntologyClass) source.getSelectedItem());
 			}
 		};
+		// use the mapped OntologyClasses for the drop down
 		Object[] subjectValues = new OntologyClass[] {
 				Annotation.OBOE_CLASSES.get(Entity.class), 
 				Annotation.OBOE_CLASSES.get(Characteristic.class), 
@@ -126,8 +129,17 @@ public class CriteriaPanel extends JPanel {
 	 * @return criteria as it exists in this panel
 	 */
 	public Criteria getCriteria() {
-		// main criteria options
-		criteria.setSubject((OntologyClass) subject.getSelectedItem());
+		// find the Java class we want for the selected OntologyClass - it is the key in the map
+		OntologyClass selectedSubject = (OntologyClass) subject.getSelectedItem();
+		Iterator<Entry<Class, OntologyClass>> subjectIter = Annotation.OBOE_CLASSES.entrySet().iterator();
+		while (subjectIter.hasNext()) {
+			Entry<Class, OntologyClass> entry = subjectIter.next();
+			if (entry.getValue().equals(selectedSubject)) {
+				criteria.setSubject(entry.getKey());
+				break;
+			}
+		}
+		
 		criteria.setCondition((String) condition.getSelectedItem());
 		criteria.setValue(value.getOntologyClass());
 		criteria.setAll(anyAll.isSelected());
@@ -155,7 +167,7 @@ public class CriteriaPanel extends JPanel {
 		subcriteriaPanel.setVisible(criteria.isGroup());
 		
 		// set the values in the UI
-		subject.setSelectedItem(criteria.getSubject());
+		subject.setSelectedItem(Annotation.OBOE_CLASSES.get(criteria.getSubject()));
 		condition.setSelectedItem(criteria.getCondition());
 		this.value.setOntologyClass(criteria.getValue());
 		

### 2010-04-01 dev/sms/src/org/ecoinformatics/sms/plugins/ui/OntologyClassField.java
set the tooltip to the selected class name, otherwise use the filter class name (as before) http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4844

@@ -67,9 +67,13 @@ public class OntologyClassField extends JTextField {
 		this.ontologyClass = ontologyClass;
 		if (ontologyClass != null) {
 			this.setText(ontologyClass.getName());
+			this.setToolTipText(ontologyClass.getName());
 		}
 		else {
 			this.setText("");
+			if (filterClass != null) {
+				this.setToolTipText(filterClass.getName());
+			}
 		}
 		this.setCaretPosition(0);
 		this.revalidate();

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ContextPage.java
added toggle button that enables/disables the editing of the screen. it does not drive the save action at this point - just the visualization (i.e. exiting edit mode does not do the save, switching tabs or column selection does the save). still a todo. http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4950

@@ -28,11 +28,14 @@
 
 package org.ecoinformatics.sms.plugins.pages;
 
-import java.awt.GridLayout;
+import java.awt.event.ActionEvent;
 
+import javax.swing.AbstractAction;
+import javax.swing.Box;
 import javax.swing.BoxLayout;
 import javax.swing.JButton;
 import javax.swing.JPanel;
+import javax.swing.JToggleButton;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Mapping;
@@ -65,6 +68,9 @@ public class ContextPage extends AbstractUIPage {
 	// context options
 	private ContextPanelList contextList;
 	
+	// edit toggle
+	JToggleButton editButton;
+	
 
 	// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
 	// *
@@ -85,14 +91,35 @@ public class ContextPage extends AbstractUIPage {
 
 		this.setLayout(new BoxLayout(this, BoxLayout.Y_AXIS));
 		
-		JPanel descPanel = WidgetFactory.makePanel(2); 
-		descPanel.setLayout(new GridLayout(2,1));
-		descPanel.add(WidgetFactory.makeHTMLLabel("The Observation was made where the...", 1));
 		final ContextPage pageRef = this;
-		JButton helpButton = Help.createHelpButton(pageRef, "What's this?");
-		descPanel.add(helpButton);
+
+		// edit button
+		AbstractAction toggleAction = new AbstractAction("Edit") {
+			public void actionPerformed(ActionEvent e) {
+				if (editButton.isSelected()) {
+					// active only if we have an observation
+					pageRef.setEnabled(observation != null);
+				}
+				else {
+					pageRef.setEnabled(false);
+				}
+			}
+		};
+		editButton = new JToggleButton(toggleAction);
 		
-		this.add(descPanel);
+		// help button
+		JButton helpButton = Help.createHelpButton(pageRef, "What's this?");
+		
+		JPanel lineOnePanel = WidgetFactory.makePanel(); 
+		lineOnePanel.add(WidgetFactory.makeLabel("The Observation was made where the...", false, null));
+		lineOnePanel.add(Box.createHorizontalGlue());
+		lineOnePanel.add(editButton);
+		
+		JPanel topPanel = WidgetFactory.makePanel(1);
+		topPanel.add(helpButton);
+			
+		this.add(lineOnePanel);
+		this.add(topPanel);
 				
[…]

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
Column Annotation tab now has the edit toggle button. again, it's not yet driving the save action, but will eventually. http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4950

@@ -29,13 +29,17 @@
 package org.ecoinformatics.sms.plugins.pages;
 
 import java.awt.GridLayout;
+import java.awt.event.ActionEvent;
 
+import javax.swing.AbstractAction;
+import javax.swing.Box;
 import javax.swing.BoxLayout;
 import javax.swing.JButton;
 import javax.swing.JCheckBox;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 import javax.swing.JTextField;
+import javax.swing.JToggleButton;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
@@ -73,6 +77,9 @@ public class AnnotationPage extends AbstractUIPage {
 	
 	private boolean showAll = true;
 	
+	// edit toggle
+	JToggleButton editButton;
+	
 	private JLabel attributeLabel;
 
 	// observation
@@ -99,11 +106,14 @@ public class AnnotationPage extends AbstractUIPage {
 	public AnnotationPage(boolean showAll) {
 		this.showAll = showAll;
 		init();
-		setFieldsEnabled(false);
+		setEnabled(false);
 	}
 
-	private void setFieldsEnabled(boolean enabled) {		
-		this.simpleAnnotationPanel.setEnabled(enabled);
+	public void setEnabled(boolean enabled) {
+		super.setEnabled(enabled);
+		simpleAnnotationPanel.setEnabled(enabled);
+		editButton.setSelected(enabled);
+
 	}
 	
 	/**
@@ -129,17 +139,33 @@ public class AnnotationPage extends AbstractUIPage {
 		final AnnotationPage pageRef = this;
 		JButton helpButton = Help.createHelpButton(pageRef, "What's this?");
 		
+		// edit button
+		AbstractAction toggleAction = new AbstractAction("Edit") {
+			public void actionPerformed(ActionEvent e) {
+				if (editButton.isSelected()) {
+					// active only if we have an observation
+					pageRef.setEnabled(currentAttributeName != null);
+				}
+				else {
+					pageRef.setEnabled(false);
+				}
+			}
+		};
+		editButton = new JToggleButton(toggleAction);
+		
 		// actually show the help
 		if (showAll) {
 			this.add(descPanel);
 		}
 		
 		// Attribute Label
-		JPanel attributeLabelPanel = WidgetFactory.makePanel(1);
+		JPanel attributeLabelPanel = WidgetFactory.makePanel();
[…]

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/context/ContextPanel.java
disable/enable remove button on the context panel when editing stops/starts http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4950

@@ -145,6 +145,7 @@ public class ContextPanel extends JPanel {
 		contextRelationship.setEnabled(enabled);
 		observationIsIdentifying.setEnabled(enabled);
 		observationList.setEnabled(enabled);
+		removeButton.setEnabled(enabled);
 	}
 
 	public Observation getSelectedObservation() {

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
the edit toggles are now driving the save action http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4950

@@ -49,6 +49,7 @@ import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Standard;
+import org.ecoinformatics.sms.plugins.AnnotationPlugin;
 import org.ecoinformatics.sms.plugins.ui.SimpleAnnotationPanel;
 
 import edu.ucsb.nceas.morpho.framework.AbstractUIPage;
@@ -148,6 +149,13 @@ public class AnnotationPage extends AbstractUIPage {
 				}
 				else {
 					pageRef.setEnabled(false);
+					// save the current state
+					annotation = pageRef.getAnnotation();
+					if (annotation != null) {
+						Log.debug(40, "Persisting Annotation: " + annotation.getURI() );
+						// save
+						AnnotationPlugin.saveAnnotation(annotation);
+					}
 				}
 			}
 		};

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ContextPage.java

@@ -46,6 +46,7 @@ import org.ecoinformatics.sms.plugins.context.ContextPanelList;
 
 import edu.ucsb.nceas.morpho.framework.AbstractUIPage;
 import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WidgetFactory;
+import edu.ucsb.nceas.morpho.util.Log;
 import edu.ucsb.nceas.utilities.OrderedMap;
 
 public class ContextPage extends AbstractUIPage {
@@ -102,6 +103,13 @@ public class ContextPage extends AbstractUIPage {
 				}
 				else {
 					pageRef.setEnabled(false);
+					// save annotation
+					annotation = pageRef.getAnnotation();
+					if (annotation != null) {
+						Log.debug(40, "Persisting Annotation from Context page: " + annotation.getURI() );
+						// save
+						AnnotationPlugin.saveAnnotation(annotation);
+					}
 				}
 			}
 		};

### 2010-04-22 dev/sms/src/org/ecoinformatics/sms/plugins/table/AnnotationTabPane.java

@@ -30,8 +30,6 @@ public class AnnotationTabPane extends JTabbedPane implements StateChangeListene
 		TAB_NAMES.add(CONTEXT_ANNOTATION);
 	}
 	
-	private int previousTab = -1;
-
 	private Annotation annotation;
 	
 	private AnnotationPage annotationPage;
@@ -60,7 +58,7 @@ public class AnnotationTabPane extends JTabbedPane implements StateChangeListene
 			handleSelectColumn();
 		}
 		if (event.getChangedState().equals(AnnotationPlugin.ANNOTATION_CHANGE_EVENT)) {
-			showColumnAnnotation();
+			showAnnotation();
 		}
 	}
 	
@@ -73,71 +71,17 @@ public class AnnotationTabPane extends JTabbedPane implements StateChangeListene
 	}
 	
 	private void handleSelectColumn() {
-
-		// save the state of the annotation if we are on the annotation tab
-		int currentTab = getSelectedIndex();
-		Log.debug(40, "Column selected, Tab = " + currentTab );
-
-		if (currentTab == TAB_NAMES.indexOf(COLUMN_ANNOTATION)) {
-			persistColumnAnnotation();
-		}
-		
-		if (currentTab == TAB_NAMES.indexOf(CONTEXT_ANNOTATION)) {
-			persistContextAnnotation();
-		}
-		
-		// always set the annotation
-		showColumnAnnotation();
-		
+		// show the annotation
+		showAnnotation();
 	}
 	
 	private void handleSelectTab() {
-		
-		// save when moving away from the annotation tab
-		int currentTab = getSelectedIndex();
-		Log.debug(40, "Tab selected, Tab = " + currentTab );
-
-		if (currentTab != TAB_NAMES.indexOf(COLUMN_ANNOTATION)) {
-			if (previousTab == TAB_NAMES.indexOf(COLUMN_ANNOTATION)) {
-				persistColumnAnnotation();
-			}
-		}
-		
-		if (currentTab != TAB_NAMES.indexOf(CONTEXT_ANNOTATION)) {
-			if (previousTab == TAB_NAMES.indexOf(CONTEXT_ANNOTATION)) {
-				persistContextAnnotation();
-			}
-		}
-		
-		// always set the latest state in the annotation tab
-		showColumnAnnotation();
-		
-		previousTab = currentTab;
+		// show the latest state in the tabs
+		showAnnotation();
 	}
 	
-	private void persistColumnAnnotation() {
-		annotation = annotationPage.getAnnotation();
[…]

### 2010-04-23 dev/sms/src/org/ecoinformatics/oboe/MaterializeDB.java
Add the classes to export the data to RDB for querying purpose.

@@ -27,7 +27,6 @@ import org.ecoinformatics.datamanager.parser.Attribute;
 import org.ecoinformatics.datamanager.parser.AttributeList;
 import org.ecoinformatics.datamanager.parser.DataPackage;
 import org.ecoinformatics.datamanager.parser.Entity;
-//import org.ecoinformatics.owlifier.OwlifierSpreadsheet;
 import org.ecoinformatics.oboe.model.ConditionInstance;
 import org.ecoinformatics.oboe.model.ContextInstance;
 import org.ecoinformatics.oboe.model.EntityInstance;
@@ -303,13 +302,15 @@ public class MaterializeDB {
 	 */
 	private static EntityInstance MaterializeEntity(
 			Observation obsType, 
+			String uniqueRecordIdPrefix,
 			Map<Observation, Set<MeasurementInstance>> obsType2MeasIdx, 
 			Map<ObsTypeKey, EntityInstance> entIdx, 
 			Annotation A, OboeModel OBOE)
 	{
 		String keyValue = GetObsTypeKeys(obsType, obsType2MeasIdx); 
 		List<Measurement> keyMeasTypes = obsType.getKeyMeasurements();
-		
+		String uniqueRecordId = uniqueRecordIdPrefix + obsType.getLabel();
+		//System.out.println(Debugger.getCallerPosition()+",keyValue="+keyValue+",keyMeasTypes="+keyMeasTypes);
 		boolean hasKey = false;
 		EntityInstance entityInstance = null;
 		
@@ -337,7 +338,9 @@ public class MaterializeDB {
 		if(crtNewEntityInstance){
 			org.ecoinformatics.sms.annotation.Entity entType = obsType.getEntity();
 			entityInstance = new EntityInstance(entType);
-			//System.out.println("New ei: " + entityInstance);
+			entityInstance.setUniqueRecordId(uniqueRecordId);
+			
+			//System.out.println(Debugger.getCallerPosition()+"New ei: " + entityInstance);
 			OBOE.AddEntityInstance(entityInstance);
 			
 			if(hasKey){
@@ -671,11 +674,23 @@ public class MaterializeDB {
 		Map<ObsTypeKey, ObservationInstance> obsIdx = 
 			new TreeMap<ObsTypeKey, ObservationInstance>(); //<ObsTypeId, KeyVal> --> observation instance
 		
[…]

### 2010-04-23 dev/sms/src/org/ecoinformatics/oboe/datastorage/PostgresDB.java

@@ -1,0 +1,363 @@
+package org.ecoinformatics.oboe.datastorage;
+
+import java.util.List;
+import java.util.Set;
+import java.util.TreeSet;
+
+import org.ecoinformatics.oboe.Debugger;
+import org.ecoinformatics.oboe.model.ContextInstance;
+import org.ecoinformatics.oboe.model.EntityInstance;
+import org.ecoinformatics.oboe.model.MeasurementInstance;
+import org.ecoinformatics.oboe.model.ObservationInstance;
+
+import org.ecoinformatics.sms.annotation.*;
+import org.ecoinformatics.oboe.model.*;
+
+import java.sql.Connection;
+import java.sql.DriverManager;
+import java.sql.ResultSet;
+import java.sql.SQLException;
+import java.sql.Statement;
+import java.sql.PreparedStatement;
+
+public class PostgresDB {
+
+	private String m_entityTypeTable = "entity_type";
+	private String m_entityInstanceTable = "entity_instance";
+	
+	private String m_obsTypeTable = "observation_type";
+	private String m_obsInstanceTable = "observation_instance";
+	
+	
+	private String m_measTypeTable = "measurement_type";
+	private String m_measInstanceTable = "measurement_instance";
+	
+	private String m_contextTypeTable = "context_relationship";
+	private String m_contextInstanceTable = "context_instance";
+	
+	private String m_url = "jdbc:postgresql://localhost:5432/oboe";
+	private String m_user = "oboe";
+	private String m_password = "nceas";
+	
+	Connection m_conn = null;
+	
+	String m_insertEntityInstance ="INSERT INTO " +m_entityInstanceTable + " VALUES(?,?,?);";
+	
+	String m_insertObservationType = "INSERT INTO " + m_obsTypeTable + " VALUES(?,?,?,?);";
+	String m_insertMeasurementType = "INSERT INTO " + m_measTypeTable + " VALUES(?,?,?,?,?,?)";
+	
+	/**
+	 * Open the postgres connection 
+	 * 
+	 * @throws Exception
+	 */
+	public void open()
+		throws Exception
+	{
+		try {
+			Class.forName("org.postgresql.Driver");
+		} catch (ClassNotFoundException cnfe) {
+			System.err.println(Debugger.getCallerPosition()+"Couldn't find driver class:");
+			System.out.println(Debugger.getCallerPosition()+"Couldn't find driver class:");
+			throw cnfe;
+		}
+		
+		try {
[…]

### 2010-04-23 dev/sms/src/org/ecoinformatics/oboe/model/EntityInstance.java

@@ -10,6 +10,8 @@ import org.ecoinformatics.sms.annotation.Entity;
 public class EntityInstance {
 	private static long gEntId=0;
 	
+	private String m_uniqueRecordId;
+	
 	private long entId;
 	private Entity entityType;
 	
@@ -34,6 +36,20 @@ public class EntityInstance {
 	public void setEntityType(Entity entityType) {
 		this.entityType = entityType;
 	}
+	
+	public String getUniqueRecordId() {
+		return m_uniqueRecordId;
+	}
+	
+	public String getUniqueEntityId()
+	{
+		return (m_uniqueRecordId+"_eid_"+entId);
+	}
+
+	public void setUniqueRecordId(String mUniqueRecordId) {
+		m_uniqueRecordId = mUniqueRecordId;
+	}
+	
 	public OboeModel getOboe() {
 		return oboe;
 	}
@@ -51,6 +67,7 @@ public class EntityInstance {
 	{
 		String str = "[ei";
 		str += entId;
+		str += ",rid="+m_uniqueRecordId+",";
 		str +="("+ entityType.getName()+")";
 		str +="]";
 		return str;

### 2010-04-23 dev/sms/src/org/ecoinformatics/oboe/model/OboeModel.java

@@ -18,13 +18,17 @@ import com.hp.hpl.jena.rdf.model.Resource;
 import com.hp.hpl.jena.rdf.model.Statement;
 import com.hp.hpl.jena.rdf.model.RDFNode;
 
+import org.ecoinformatics.oboe.Debugger;
+import org.ecoinformatics.oboe.datastorage.PostgresDB;
+import org.ecoinformatics.sms.annotation.*;
+
 public class OboeModel {
 
-	 private List<EntityInstance> m_entityInstances;
-	 private List<ObservationInstance> m_observationInstances;
-	 private List<MeasurementInstance> m_measurementInstances;
+	 public List<EntityInstance> m_entityInstances;
+	 public List<ObservationInstance> m_observationInstances;
+	 public List<MeasurementInstance> m_measurementInstances;
 	 
-	 private List<ContextInstance> m_contextInstances;
+	 public List<ContextInstance> m_contextInstances;
 	 
 	 //index from oi --> ci list, used in materialize DB
 	 private Map<ObservationInstance, List<ContextInstance>> m_oi2ciList;
@@ -277,6 +281,24 @@ public class OboeModel {
 	 
 	 
 	 /**
+	  * Import this the materialized dataset and the annotation (with type information) to databases
+	  * 
+	  * @param A
+	 * @throws Exception 
+	  */
+	 public void toRDB(String annotationFileName, Annotation A) throws Exception
+	 {
+		 System.out.println(Debugger.getCallerPosition()+"Begin...");
+		 PostgresDB db = new PostgresDB();
+		 
+		 db.open();
+		 db.importAnnotation(A, annotationFileName); //export type information
+		 db.importInstance(this); //export data instance information.
+		 db.close();
+		 System.out.println(Debugger.getCallerPosition()+"End...");
+	 }
+	 
+	 /**
 	  * output the materialized data to CSV file
 	  * @param csvFileName: the absolute file name to store the CSV file 
 	  */
@@ -316,5 +338,20 @@ public class OboeModel {
 			e.printStackTrace();
 			throw e;
 		}
+	 }
+	 
+
+	 
+	 /**
+	  * Show the space used for this materialized OBOE view
+	  */
+	 public void calSpace()
+	 {
+		 System.out.println("Number of entities:"+m_entityInstances.size());
+		 System.out.println("Number of observations:"+m_observationInstances.size());
+		 System.out.println("Number of measurements:" + m_measurementInstances.size());
[…]

### 2010-04-23 dev/sms/src/org/ecoinformatics/sms/annotation/Context.java

@@ -34,7 +34,7 @@ package org.ecoinformatics.sms.annotation;
 /**
  * Objects of this class represent observation contexts
  */
-public class Context{
+public class Context implements Comparable<Context>{
 
     /**
      * Default constructor
@@ -152,5 +152,25 @@ public class Context{
 		
 		boolean cmp3 = (_isIdentifying==o.isIdentifying());
 		return cmp3;
+	}
+
+	/**
+	 * For comparing two contexts
+	 * @author cao
+	 */
+	public int compareTo(Context c)	
+	{
+		
+		int cmp1 = _observation.getLabel().compareTo(c.getObservation().getLabel());
+		
+		if(cmp1!=0) return cmp1;
+		
+		int cmp2=0;
+		if(_relationship!=null||c.getRelationship()!=null){
+			cmp2 = _relationship.getURI().compareTo(c.getRelationship().getURI());
+		}
+		if(cmp2!=0) return cmp2;
+		
+		return 0;
 	}
 } 

### 2010-04-23 dev/sms/src/org/ecoinformatics/oboe/datastorage/PostgresDB.java
Finish the loading of materialized data. Todo: set proper index on the tables + set benchmark queries

@@ -42,6 +42,10 @@ public class PostgresDB {
 	Connection m_conn = null;
 	
 	String m_insertEntityInstance ="INSERT INTO " +m_entityInstanceTable + " VALUES(?,?,?);";
+	String m_insertObservationInstance ="INSERT INTO " +m_obsInstanceTable + " VALUES(?,?,?,?);";
+	String m_insertMeasurementInstance ="INSERT INTO " +m_measInstanceTable + " VALUES(?,?,?,?,?);";
+	String m_insertContextInstance ="INSERT INTO " +m_contextInstanceTable + " VALUES(?,?,?,?,?);";
+	
 	
 	String m_insertObservationType = "INSERT INTO " + m_obsTypeTable + " VALUES(?,?,?,?);";
 	String m_insertMeasurementType = "INSERT INTO " + m_measTypeTable + " VALUES(?,?,?,?,?,?)";
@@ -84,13 +88,16 @@ public class PostgresDB {
 			}
 		}
 	}
+	
+	
 	/**
 	 * Form the SQL to insert entity instances
 	 * @param ei
 	 * @return
+	 * @deprecated
 	 */
 	private String formEntityInstanceSQL(EntityInstance ei){
-		String sql = "INSERT INTO " +m_entityInstanceTable + " values("+
+		String sql = "INSERT INTO " +m_entityInstanceTable + " VALUES("+
 					ei.getUniqueRecordId()+","+ //used to join with the observation instance table
 					ei.getEntId()+"," +			//used to join with the observation instance table
 					ei.getEntityType().getName()+");";
@@ -99,25 +106,28 @@ public class PostgresDB {
 		return sql;
 	}
 	
-//	private String formEntityTypeSQL(Entity et){
-//		String sql = "INSERT INTO " +m_entityTypeTable + " values("+
-//					et.getOntology().getPrefix()+","+ //used to join with the observation instance table
-//					et.getURI()+"," +			//used to join with the observation instance table
-//					et.getName()+");";
-//		
-//		//System.out.println(Debugger.getCallerPosition()+"sql="+sql);
-//		return sql;
-//	}
+	/**
+	 * Set the parameters for inserting entity instances
+	 * 
+	 * @param pstmt
+	 * @param ei
+	 * @throws SQLException
+	 */
+	private void setEntityInstanceParam(PreparedStatement pstmt, EntityInstance ei) 
+		throws SQLException
+	{
+		pstmt.setString(1, ei.getUniqueRecordId());
[…]

### 2010-04-23 docs/algdraft/DataIntegration.tex
add the starting draft for the data integration paper.

\documentclass[10pt]{article}
%\title{}
%\author{}
%\date{}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{a4}
\usepackage{color}
\renewcommand\floatpagefraction{0.99}
\renewcommand\topfraction{0.99}
\renewcommand\bottomfraction{0.99}
\renewcommand\textfraction{.05}
\setcounter{totalnumber}{5}

%\setlength{\parindent}{0pt}
\setlength\topmargin{0in}
\setlength\headheight{0in}
\setlength\headsep{0in}
\setlength\textheight{8.5in}
\setlength\textwidth{6.5in}
\setlength\oddsidemargin{0in}
\setlength\evensidemargin{0in}

\newtheorem{example}{Example}[section]
\newtheorem{definition}{Definition}[section]
\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}

\title{Data integration}
\author{}
\begin{document}
\maketitle

\section{Materialize Database}

{\bf HP:}{Copy the materials from materiazeDB to here. Skip now.}

\section{Query rewriting}

What kinds of queries?

The information that users can have: metadata about the table.

In the scientific data, it is unrealistic for the users to know the
table structure to perform the query.

IR-style query: 
\begin{itemize}
\item KQ1: Give me the datasets that contain species ``Picea rubens'' observations.
\item KQ2: give me the datasets that have measurements on ``area''
  characteristics. 
\item KQ3: Give me the datasets that contain species ``Picea rubens''
  observations in ``California''. 
\item KQ: Give me the records....
\end{itemize}

IR-style summarization query: 
\begin{itemize}
SQ1: Give me the dataset that contains at least five ``Picea rubens'' observations.
SQ2: Give me the datasets that have measurements with average ``area''
bigger than 5.0 square feet. 
SQ3: 
\end{itemize}

Use the materialized database, we would have four tables for all the datasets: 
entity, observation, measurement, context. 
and 4 tables for entity types, observation types, measurement types,
and context relationships.

To answer KQ1: it's simple join between

\section{Experiments}

1. Materialize DB (time + space) \\
2. Query over materialized DB 
3.

\end{document}

### 2010-04-27 dev/sms/src/org/ecoinformatics/sms/annotation/persistent/manager/DbAnnotationManager.java
only include dataTable param if it is given

@@ -395,7 +395,9 @@ public class DbAnnotationManager extends DefaultAnnotationManager {
 					"emlPackage = $emlPackage and dataTable = $dataTable");
 		Map<String, String> params = new HashMap<String, String>();
 		params.put("emlPackage", emlPackage);
-		params.put("dataTable", dataTable);
+		if (dataTable != null) {
+			params.put("dataTable", dataTable);
+		}
 		SelectQuery query = new SelectQuery(DbAnnotation.class, expression.expWithParameters(params));
 		List<DbAnnotation> values = context.performQuery(query);
 		if (values != null) {

### 2010-04-27 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
copy the annotation when the EML package is duplicated http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4853

@@ -21,6 +21,7 @@ import java.awt.event.AdjustmentListener;
 import java.awt.event.MouseListener;
 import java.awt.event.WindowAdapter;
 import java.awt.event.WindowEvent;
+import java.io.ByteArrayInputStream;
 import java.io.ByteArrayOutputStream;
 import java.io.File;
 import java.io.FileInputStream;
@@ -40,6 +41,7 @@ import javax.swing.table.TableColumnModel;
 
 import org.ecoinformatics.sms.SMS;
 import org.ecoinformatics.sms.annotation.Annotation;
+import org.ecoinformatics.sms.annotation.AnnotationException;
 import org.ecoinformatics.sms.annotation.Measurement;
 import org.ecoinformatics.sms.annotation.Observation;
 import org.ecoinformatics.sms.ontology.OntologyClass;
@@ -724,17 +726,42 @@ public class AnnotationPlugin
 		String oldId = saveEvent.getInitialId();
 		String newId = saveEvent.getFinalId();
 		String location = saveEvent.getLocation();
-			
+		boolean duplicate = saveEvent.isDuplicate();
+		
 		// get all the annotations for the original docid
 		List<Annotation> annotations = SMS.getInstance().getAnnotationManager().getAnnotations(oldId, null);
 		for (Annotation annotation: annotations) {
-			// set the updated packageId
-			annotation.setEMLPackage(newId);
-			// save the annotation
-			saveAnnotation(annotation);
-			// serialize to disk
-			serializeAnnotation(newId, location);
+			if (duplicate) {
+				try {
+					// copy the annotation
+					ByteArrayOutputStream baos = new ByteArrayOutputStream();
+					annotation.write(baos);
+					Annotation annotationCopy = Annotation.read(new ByteArrayInputStream(baos.toByteArray()));
+					// make an id for the annotation
+					AccessionNumber an = new AccessionNumber(Morpho.thisStaticInstance);
+					String annotationId = an.getNextId();
+					// set the data package association and the id
+					annotationCopy.setEMLPackage(newId);
+					annotationCopy.setURI(annotationId);
+					// save the copy
+					saveAnnotation(annotationCopy);
+					// serialize the copy
+					serializeAnnotation(newId, location);
+				} catch (AnnotationException e) {
+					Log.debug(5, "Error saving Annotation duplicate");
+					e.printStackTrace();
+				}
+			}
+			else {
+				// set the updated packageId
[…]

### 2010-04-27 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
prompt for duplicating annotations with EML package http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4853

@@ -33,6 +33,7 @@ import java.util.Vector;
 import java.util.Map.Entry;
 
 import javax.swing.JLabel;
+import javax.swing.JOptionPane;
 import javax.swing.JTabbedPane;
 import javax.swing.JTable;
 import javax.swing.UIManager;
@@ -732,6 +733,19 @@ public class AnnotationPlugin
 		List<Annotation> annotations = SMS.getInstance().getAnnotationManager().getAnnotations(oldId, null);
 		for (Annotation annotation: annotations) {
 			if (duplicate) {
+				// prompt?
+				boolean saveDuplicate = false;
+				int response = 
+					JOptionPane.showConfirmDialog(
+							null, 
+							"Would you like to copy the existing Annotation information to the new package?\n" +
+							"id = " + annotation.getURI(),
+							"Duplicate Annotation?", JOptionPane.YES_NO_OPTION);
+				saveDuplicate = (response == JOptionPane.YES_OPTION);
+				if (!saveDuplicate) {
+					continue;
+				}
+				
 				try {
 					// copy the annotation
 					ByteArrayOutputStream baos = new ByteArrayOutputStream();

### 2010-04-27 docs/algdraft/DataIntegration.tex

@@
   observations in ``California''. 
-\item KQ: Give me the records....
 \end{itemize}
@@
 \begin{itemize}
-SQ1: Give me the dataset that contains at least five ``Picea rubens'' observations.
-SQ2: Give me the datasets that have measurements with average ``area''
+\item SQ1: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
+\item SQ2: Give me the datasets that have measurements with average ``area''
 bigger than 5.0 square feet. 
-SQ3: 
 \end{itemize}
@@
 
+How to perform query rewriting? 
+m1:
+
+data1 --> table 1
+data2 --> table 2 
+When we put the data as data tables, we add a unique record id for
+each row in the data. This is used to get the unique record id.
+
+Metadata
+Table definition
+
+Table column definition
+
+Map annotation information
+
+\begin{example}
+KQ1: <concept>: <cond>
+Concept can be observation, measurement, characteristic, standard,
+unit, etc.
+
+Use query rewriting method: 
+For KQ1, give me the datasets that contain species ``Picea rubens''
+observations.
+concept = observation, cond is species = ``Picea rubens'';
+
+The first step is to find the dataset that contain all the metadata. 
+From the annotation, find the {\em measurement type} with {\em
+  Characteristic} species. 
+Here, it is $m_3$. From the mapping, we find that the attribute in the
+table is {\em spp}.
+
+The second step is to find the data file which really have this
+value. 
+For this one, we find the data table, and do a select on the table
+content.
+
+Use materialized database: 
+The first step is the same, after we know the data file, then we
+search the table {\em measurement\_instance} with the {\em mvalue=``Picea rubens''}.
+
+In this example, the computation cost is to search the different data
+tables. To reduce this computation cost, we can materialize just the
+measurement values. (How to do this? ) need more details.
+
+Half materialized: how to perform query over it. 
+\end{example}
+
+\begin{example}
+SQ1: Give me the dataset that contains at least five distinct ``Picea
+rubens'' observations.
+
+The first step is to find the datasets that contain the metadata. 
+The second step is to perform a query over the materialized database. 
+%select mlabel, mvalue, count(*) as cnt from measurement_instance
+%group by mvalue,mlabel having count(*)>=2;
+
+Using the query rewriting, 
+First, we find the data tables that contain the needed concepts. 
+In each table, we need to figure out the distinct observations. 
+From $m_i$ $obstype_i$, all the other measurements that are of
+$obstype_i$. 
+Group by the key measurements of $obstype_i$ if distinct yes.
+If this observation type does not have any context, then the key
+measurements are all the measurements directly defined under this
+observatioin type. Otherwise, i.e., this observation type has context,
+then the key measurements are all the measurements in the context
+chain. 
+\end{example}
+
+\begin{example}
+KQ3: Give me the datasets that contain species ``Picea rubens''
+  observations in ``California''.
+
+Species = ``Picea rubens'' and observation has context {\em in}
+another observation which has measurement of value ``California''. 
+\end{example}
+
+Using materializedDB, we can answer the query this way
+First, Species --> mlabel, observation type
+
+select mtype, olabel from measurement\_type  \\
+where chracteristic = 'Species';  \\
+
+For the olabel found, \\
+select context\_olabel\\
+from context\_relationship\\
+where olabel = 'olabel';
+
+context\_olabel\_set is a set containing all the observation type
+labels in the context chain.
+
+SELECT record\_id\\
+FROM observation\_instance oi, observation\_instance coi,
+context\_instance ci,
+measurement\_instance mi, measurement\_instance cmi \\
+WHERE mi.oid=ci.oid AND ci.oid = oi.oid and  cmi.oid=coi.oid AND ci.context\_oid =
+coi.oid and oi.otype = 'olabel' and coi.otype $\in$
+context\_olabel\_set
+AND mi.mvalue=`Picea rubens' and cmi.mvalue='California';\\
+
+Use query rewriting, what I can do? 
+It's the same till to getting context\_olabel\_set.
+
+Get the asked condition columns and the context columns,
+
+select record\_id from data\q_tables
+where condition column 1='value1' and (context column1 = 'context
+value' or context column2 = `context value', etc.);
+
+\begin{example}
+SQ2: Give me the datasets that have measurements with average ``area''
+bigger than 5.0 square feet. 
+\end{example}
+
 \section{Experiments}
 
-1. Materialize DB (time + space) \\
-2. Query over materialized DB 
-3.
+Synthetic data generator.
+
+Algorithm to materialize DB. 
+Report the time and space.
+
+Synthetic query generator.
+
+Test algorithm to perform query over materiazed DB.
+Test algorithm to perform query-rewriting over materiazed DB.
+
+Test algorithm for half-materialized data.
 

### 2010-04-27 docs/DataIntegration/DataIntegration.tex
This is for a specific target conference. SO, put it here separately.Later on, we can move it to other part.

--- 2010-04-27 docs/algdraft/DataIntegration.tex
@@
 
-select record\_id from data\q_tables
+select record\_id from data\_tables
 where condition column 1='value1' and (context column1 = 'context

### 2010-04-27 dev/sms/src/org/ecoinformatics/sms/plugins/ui/OntologyClassField.java
highlight the current field that is responding to the ontology browser http://bugzilla.ecoinformatics.org/show_bug.cgi?id=4856

@@ -1,5 +1,6 @@
 package org.ecoinformatics.sms.plugins.ui;
 
+import java.awt.Color;
 import java.awt.Component;
 import java.awt.Container;
 import java.awt.Dimension;
@@ -64,10 +65,13 @@ public class OntologyClassField extends JTextField {
 	
 	private static Point lastPosition = null;
 	
+	private static JDialog popup = null;
+	
+	private static OntologyClassField currentField;
+	
 	private OntologyClass ontologyClass;
 	private OntologyClass filterClass;
 	private OntologyClassSelectionPanel selectionPanel;
-	private static JDialog popup = null;
 	private boolean editing = false;
 
 	public OntologyClassField(String text) {
@@ -309,6 +313,8 @@ public class OntologyClassField extends JTextField {
 		if (close) {
 			popup.setVisible(false);
 			selectionPanel = null;
+			// make sure we are not highlighted
+			currentField.setBorder(new JTextField().getBorder());
 		}
 	}
 	
@@ -388,7 +394,7 @@ public class OntologyClassField extends JTextField {
 			popup.setDefaultCloseOperation(JDialog.DISPOSE_ON_CLOSE);
 			final OntologyClassField sourceRef = source;
 			popup.addWindowListener(new WindowAdapter() {
-				public void windowClosed(WindowEvent e) {
+				public void windowClosing(WindowEvent e) {
 					sourceRef.stopEditing(false, true);
 				}
 			});
@@ -413,9 +419,19 @@ public class OntologyClassField extends JTextField {
 			//ignore
 		}
 		
-		//begin editing on this field
+		// begin editing on this field
 		source.editing = true;
 		source.selectionPanel = selectionPanel;
+		
+		// highlight the field
+		if (currentField != null) {
+			currentField.setBorder(new JTextField().getBorder());
+		}
+		source.setBorder(
+				BorderFactory.createCompoundBorder(
+						BorderFactory.createLineBorder(Color.BLUE, 2), 
+						new JTextField().getBorder()));
+		currentField = source;
 		
 		// the select button
 		boolean showSelect = true;

### 2010-04-28 docs/algdraft/DataIntegration/DataIntegration.tex
update the data integration draft with more clarified query definition and example.

--- 2010-04-27 docs/DataIntegration/DataIntegration.tex
@@
-\documentclass[10pt]{article}
-%\title{}
-%\author{}
-%\date{}
+\documentclass[conference]{IEEEtran}
+
 \usepackage{algorithm}
 \usepackage{algorithmic}
-\usepackage{a4}
-\usepackage{color}
 \renewcommand\floatpagefraction{0.99}
@@
 
-%\setlength{\parindent}{0pt}
-\setlength\topmargin{0in}
-\setlength\headheight{0in}
-\setlength\headsep{0in}
-\setlength\textheight{8.5in}
-\setlength\textwidth{6.5in}
-\setlength\oddsidemargin{0in}
-\setlength\evensidemargin{0in}
-
 \newtheorem{example}{Example}[section]
@@
 
-\title{Data integration}
-\author{}
+\hyphenation{op-tical net-works semi-conduc-tor}
+
 \begin{document}
+
+\title{Querying Integrated Ecological Data}
+
+% author names and affiliations
+% use a multiple column layout for up to three different
+% affiliations
+\author{\IEEEauthorblockN{Michael Shell}
+\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
+Georgia Institute of Technology\\
+Atlanta, Georgia 30332--0250\\
+Email: http://www.michaelshell.org/contact.html}
+\and
+\IEEEauthorblockN{Homer Simpson}
+\IEEEauthorblockA{Twentieth Century Fox\\
+Springfield, USA\\
+Email: homer@thesimpsons.com}
+\and
+\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
+\IEEEauthorblockA{Starfleet Academy\\
+San Francisco, California 96678-2391\\
+Telephone: (800) 555--1212\\
+Fax: (888) 555--1212}}
+
 \maketitle
 
-\section{Materialize Database}
+\begin{abstract}
+%\boldmath
+The abstract goes here.
+\end{abstract}
+% IEEEtran.cls defaults to using nonbold math in the Abstract.
+% This preserves the distinction between vectors and scalars. However,
+% if the conference you are submitting to favors bold math in the abstract,
+% then you can use LaTeX's standard command \boldmath at the very start
+% of the abstract to achieve this. Many IEEE journals/conferences frown on
+% math in the abstract anyway.
+
+% no keywords
+
+\section{Introduction}
+In this work, we study the query processing over integrated ecological
+data through OBOE model\cite{DBLP:conf/er/BowersMS0}, a conceptual model used to interpret
+observation and measurement data.
+
+Contributions of this work:
+\begin{itemize}
+\item Formalize the queries over observation and measurement query.
+\item Propose three methods to evaluate queries. 
+\end{itemize}
+
+\section{Related work}
+
+This is related
+
+\section{Data model and query}
+
+In the ecological domain, observation and measurement data are the
+core. As a consequence, people are more interested in finding datasets
+related to specific observation and measurements.
+
+\subsection{Data model}
+
+We a scientist contributes dataset into the data repository, generally
+each dataset is converted to a data table. 
+So, the database contains metadata about each dataset, the data table
+definition (e.g., attribute/column name, attribute type, etc).
+
+To formalize the senario, we use $D$ to denote the set of data tables. Let $d$ be one data table, 
+$d[attr]$ or $d[idx]$ represents the set of values for the attribute
+$attr$ or for the $idx$-th attribut.
+
+Generally, the dataset is connected to the OBOE model using
+annotations.
+
+\from{HP}{Put the screendump of annotation.}
+
+Map annotation information 
+Through the annotation, 
+We use $A$ to denote the annotation of one dataset.
+
+\from{HP}{Another way to organize the data is to store them in one big
+data table with three columns data\_file, attribute\_name,
+attribute\_value. The problem with this organizationo is that the
+attribute\_value are different types. }
+
+%Implementation consideration: 
+%When we put the data to data tables, we add a unique record id for
+%each row in the data. This is used to get the unique record id.
+
+\subsection{Observation and measurement query}
+
+Given dataset in Figure \ref{}.
+
+\begin{itemize}
+\item $Q_1$: Give me the datasets that contain species ``Picea rubens'' observations.
+%$Q_2$: Give me the datasets that have measurements on ``area''  characteristics. 
+\item $Q_2$: Give me the datasets that contain species ``Picea rubens''
+  observations in ``California''. 
+\item $Q_3$: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
+\item $Q_4$: Give me the datasets that have measurements with average ``area''
+bigger than 5.0 square feet. 
+\end{itemize}
+
+We formalize the queries as follows.
+
+\begin{definition}
+KQ1: $<concept>:<cond>$
+Here, <concept> can be observation, measurement, characteristic, standard,
+unit, etc.. 
+<cond> is in the form of $func([distinct]~attr)~op~value$, here $op$ can be some
+algorithmmic operators $=$, $\neq$, $>$, or $<$. 
+$func$ are aggregation functions including sum, average, min and max.
+
+The result is a set of $d$, $\{d|d\in D \wedge (value~op~dv = true |dv\in d[attr])\}$.
+\end{definition}
+
+For each of such result, we denote that $d~s.t.~KQ$ meaning that
+dataset $d$ satisfies the query.
+
+This is the baseline keyword style query. 
+The following two queries are the basic queries.
+
+\begin{definition}
+KQ2: $Q_1~context(Q_1.concept,Q_2)~Q_2 $, here $Q_i$ is a basic query
+in the form of $<concept>:<cond>$. 
+\end{definition}
+
+\begin{example}
+The given example queries can be represented in the formal query
+format. 
+\begin{itemize}
+\item $Q_1 = Observation: Species='Picea rubens'$
+\item $Q_2 = <Observation_1: Species='Picea rubens'>
+  IN(Observation_1,Observation_2) <Observation_2:
+  measurement_{unknown}=`California'>$
+\item $Q_3 = Observation: Species='Picea rubens' \wedge
+  count(distinct~Species)\geq 5$ \from{HP}{TODO}
+\item $Q_4$ 
+\end{itemize}
+
+\end{example}
+
+\section{Querying annotated and integrated ecological data}
+
+All the datasets can be imported to a RDBMS for
+
+To answer such queries, a naive method is to
+
+\subsection{Materialized database}
 
@@
 
-\section{Query rewriting}
-
-What kinds of queries?
+\subsection{Query rewriting}
 
@@
 table structure to perform the query.
-
-IR-style query: 
-\begin{itemize}
-\item KQ1: Give me the datasets that contain species ``Picea rubens'' observations.
-\item KQ2: give me the datasets that have measurements on ``area''
-  characteristics. 
-\item KQ3: Give me the datasets that contain species ``Picea rubens''
-  observations in ``California''. 
-\end{itemize}
-
-IR-style summarization query: 
-\begin{itemize}
-\item SQ1: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
-\item SQ2: Give me the datasets that have measurements with average ``area''
-bigger than 5.0 square feet. 
-\end{itemize}
 
@@
 
-How to perform query rewriting? 
-m1:
-
-data1 --> table 1
-data2 --> table 2 
-When we put the data as data tables, we add a unique record id for
-each row in the data. This is used to get the unique record id.
-
-Metadata
-Table definition
-
-Table column definition
-
-Map annotation information
-
 \begin{example}
-KQ1: <concept>: <cond>
-Concept can be observation, measurement, characteristic, standard,
-unit, etc.
-
-Use query rewriting method: 
-For KQ1, give me the datasets that contain species ``Picea rubens''
-observations.
-concept = observation, cond is species = ``Picea rubens'';
+Take KQ1 as example. 
+concept = observation, cond is {\em species = ``Picea rubens''};
 
@@
 \begin{example}
-SQ1: Give me the dataset that contains at least five distinct ``Picea
-rubens'' observations.
+We use $SQ_1$ as an example.
 
@@
 
+Use materialized database: 
+Analysis: ``Area'' is the context of some observation.
+Area --> mlabel, observation type
+
+From the context observation lable, find the observation labels, \\
+select olabel\\
+from context\_relationship\\
+where context\_olabel = 'olabel';
+
+Let olabel\_set  be the set containing all the observation type
+labels that use ``Area'' as context.
+
+SELECT oi.record\_id \\
+FROM observation\_instance oi, measurement\_instance mi, 
+observation\_instance coi, measurement\_instance cmi,context\_instance ci\\
+WHERE (mi.oid=ci.oid AND ci.oid = oi.oid) and (cmi.oid=coi.oid AND ci.context\_oid =
+coi.oid) AND coi.otype $\in$ olabel\_set 
+AND cmi.mlabel=`Area'
+GROUP BY cmi.oid
+HAVING avg(cmi.mvalue)>5.0in;
+
 \section{Experiments}
@@
 
+\section{Conclusion}
+
+%\bibliographystyle{abbrv}
+%\bibliography{model}
+%\bibliographystyle{IEEEtran}
+%\bibliography{IEEEabrv,DataIntegration}
+
+%\begin{thebibliography}{1}
+%\bibliographystyle{abbrv}
+%\bibliography{model}
+%\end{thebibliography}
+
 \end{document}

### 2010-04-28 dev/sms/src/org/ecoinformatics/oboe/datastorage/PostgresDB.java
Update some functions.

@@ -471,7 +471,7 @@ public class PostgresDB {
 			//Add the context types related to this observation type
 			for(Context context: obsType.getContexts()){
 				setContextTypeParam(pstmtContext,annotationId,obsType,context);
-				System.out.println(Debugger.getCallerPosition()+pstmtContext.toString());
+				//System.out.println(Debugger.getCallerPosition()+pstmtContext.toString());
 				pstmtContext.execute();
 			}
 		}

### 2010-04-28 dev/sms/src/org/ecoinformatics/sms/plugins/AnnotationPlugin.java
handle DeleteEvents to clean up the annotations when their EML packages are removed.

@@ -62,6 +62,8 @@ import org.ecoinformatics.sms.plugins.table.DataTableModelListener;
 import org.ecoinformatics.sms.plugins.table.ScrollBarAdjustmentListener;
 import org.ecoinformatics.sms.renderer.OntologyClassSelectionPanel;
 
+import com.hp.hpl.jena.rdf.arp.impl.Location;
+
 import edu.ucsb.nceas.morpho.Morpho;
 import edu.ucsb.nceas.morpho.datapackage.AbstractDataPackage;
 import edu.ucsb.nceas.morpho.datapackage.AccessionNumber;
@@ -80,6 +82,7 @@ import edu.ucsb.nceas.morpho.plugins.ServiceExistsException;
 import edu.ucsb.nceas.morpho.plugins.ServiceProvider;
 import edu.ucsb.nceas.morpho.query.Query;
 import edu.ucsb.nceas.morpho.query.ResultSet;
+import edu.ucsb.nceas.morpho.util.DeleteEvent;
 import edu.ucsb.nceas.morpho.util.GUIAction;
 import edu.ucsb.nceas.morpho.util.Log;
 import edu.ucsb.nceas.morpho.util.SaveEvent;
@@ -289,6 +292,7 @@ public class AnnotationPlugin
 		StateChangeMonitor.getInstance().addStateChangeListener(StateChangeEvent.CREATE_ENTITY_DATAPACKAGE_FRAME, this);
 		StateChangeMonitor.getInstance().addStateChangeListener(StateChangeEvent.SELECT_DATA_VIEWER, this);
 		StateChangeMonitor.getInstance().addStateChangeListener(StateChangeEvent.SAVE_DATAPACKAGE, this);
+		StateChangeMonitor.getInstance().addStateChangeListener(StateChangeEvent.DELETE_DATAPACKAGE, this);
 		
 		//initialize the ontologies
 		initializeOntologies();
@@ -522,6 +526,31 @@ public class AnnotationPlugin
 		}
 	}
 	
+	public static void deleteAnnotations(String packageId, String location) {
+		FileSystemDataStore fds = new FileSystemDataStore(Morpho.thisStaticInstance);
+		MetacatDataStore mds = new MetacatDataStore(Morpho.thisStaticInstance);
+		
+		List<Annotation> annotations = SMS.getInstance().getAnnotationManager().getAnnotations(packageId, null);
+		for (Annotation annotation: annotations) {
+			try {
+				String annotationId = annotation.getURI();
+				// remove from manager
[…]

### 2010-04-29 docs/algdraft/DataIntegration/DataIntegration.tex
update more writing.

@@
 \documentclass[conference]{IEEEtran}
+
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
 
@@
 \end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the conference you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals/conferences frown on
-% math in the abstract anyway.
 
@@
 In this work, we study the query processing over integrated ecological
-data through OBOE model\cite{DBLP:conf/er/BowersMS0}, a conceptual model used to interpret
+data through OBOE model\cite{DBLP:conf/er/BowersMS08}, a conceptual model used to interpret
 observation and measurement data.
 
+\subsection{Background}
+Introduce OBOE, annotation
+Give an example of them.
+
+\subsection{Contribution and paper organization}
 Contributions of this work:
@@
 
-\section{Related work}
+This paper is organized as follows. 
+Section \ref{sec:relatedwork} reviews works that are related to this
+research. Section \ref{sec:dataquery} formalize the data model and the
+queries that people are interested to ask.
 
-This is related
+\section{Related work}\label{sec:relatedwork}
 
-\section{Data model and query}
+\section{Data model, annotation and query}\label{sec:dataquery}
 
-In the ecological domain, observation and measurement data are the
-core. As a consequence, people are more interested in finding datasets
-related to specific observation and measurements.
+In the ecology domain, most data people collected are observation and
+measurement data. {\bf Add one sentence to explain this.}
+Naturally, when searching ecological data, people are more interested in finding datasets
+related to specific observation and measurement. 
+In this section, we first illustrate the data model. Then, we
+formalize the queries that scientists in this domain tend to ask.
 
@@
 
-We a scientist contributes dataset into the data repository, generally
-each dataset is converted to a data table. 
+When a scientist contributes dataset into the data repository, generally
+each dataset is converted to a data table\cite{tdar} or treated as a separate
+object entity\cite{knb}. 
 So, the database contains metadata about each dataset, the data table
 definition (e.g., attribute/column name, attribute type, etc). 
-
-To formalize the senario, we use $D$ to denote the set of data tables. Let $d$ be one data table, 
+%\from{HP}{Another way to organize the data is to store them in one big
+%data table with three columns data\_file, attribute\_name,
+%attribute\_value. The problem with this organization is that the
+%attribute\_value are different types. } 
+To formalize the scenario, we use $D$ to denote the set of data
+tables. Each data table contains metadata about the attribute
+definition.  Sometimes, we also use {\em column} to denote the attribute. 
+Let $d$ be one data table, 
 $d[attr]$ or $d[idx]$ represents the set of values for the attribute
-$attr$ or for the $idx$-th attribut.
-
-Generally, the dataset is connected to the OBOE model using
-annotations.
-
-\from{HP}{Put the screendump of annotation.}
-
-Map annotation information 
-Through the annotation, 
-We use $A$ to denote the annotation of one dataset.
-
-\from{HP}{Another way to organize the data is to store them in one big
-data table with three columns data\_file, attribute\_name,
-attribute\_value. The problem with this organizationo is that the
-attribute\_value are different types. }
-
-%Implementation consideration: 
+$attr$ or for the $idx$-th attribute. 
+Besides the dataset, in the data store, we also keep the annotation
+information. 
+%Implementation detail (no need to put it in the paper): 
 %When we put the data to data tables, we add a unique record id for
@@
 
+\subsection{Annotation}
+In the Ecology domain, people are using annotations to connect
+datasets to the OBOE model. 
+{\bf TODO: Put a screen-dump of annotation.}
+
+Map annotation information 
+We use $A$ to denote the annotation of one dataset. 
+More specifically,
+To describe an observation, we need to denote on which entity (object in the
+real world) the observation is made.
+Very often, more than one observation can be made on one entity.   
+\begin{itemize}
+\item A.ObsType = $\{\langle \underline{AnnotId,ObsTypeId}, EntTypeId,
+  isDistinct \rangle\}$ 
+\item A.MeasType = $\{\langle \underline{AnnotId, MeasTypeId}, ObsTypeId,
+  CharType, StdType, ProtType, Precision, isKey \rangle\}$ 
+\item A.ContextType = $\{\langle \underline{AnnotId, ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$
+\item A.Map = $\{\langle \underline{AnnotId,MeasTypeId, ResAttribute, Cond}, Val\rangle\}$
+\end{itemize}
+
+``AnnotId'' denotes the resource (e.g., data table) the annotation is on.
+
+{\bf TODO:} Describe the constraints in annotation
+
 \subsection{Observation and measurement query}
 
-Given dataset in Figure \ref{}.
+{\bf TODO: Have a running example from.}
+Observation and measurement (OM) query
+
+\begin{definition}[Basic OM query]\label{def:basic_omq}
+A basic observation and measurement query is defined as $Q$::={\tt
+  concept:cond}. 
+Here, {\tt concept} is the main term in an observation model. 
+In using the OBOE model, it can be observation, measurement,
+characteristic or standard. 
+{\tt cond} is in the form of $f([distinct]~attr)~op~value$ where $op$ is
+basic operator from $\{=, \neq, >, <\}$ and $f$ is an
+aggregation function from $\{sum, avg, count, min, max\}$. 
+\end{definition}
+
+\begin{definition}\label{def:qresult}
+The result of a query $Q$ is a set of data objects (e.g., data tables) 
+$\{d|d\in D \wedge d~satisfies~{\tt cond}\}$. 
+%$\{d|d\in D \wedge (value~op~dv = true, dv\in d[attr])\}$. 
+For each of such result $d$, we use $d~s.t.~Q$ to denote that
+data object $d$ satisfies the query. 
+\end{definition}
+
+Definition \ref{def:basic_omq} can formulate queries on the
+one specific observation and measurement. 
+However, it fails to catch the observation context relationship. 
+E.g., {\bf TODO: }
+So, we generalize the basic OM query to contextualized OM query as
+follows.
+
+\begin{definition}[Contextualized OM query]\label{def:context_omq}
+$CQ$::=$Q_1~${\tt context}($Q_1.concept, Q_2.concept$)$~Q_2$. Here $Q_i$ is a basic query
+in the form of {\tt concept:cond}. 
+\end{definition}
+
+Given the dataset in Figure \ref{}. 
+People may ask the following queries.
 
@@
 
-We formalize the queries as follows.
-
-\begin{definition}
-KQ1: $<concept>:<cond>$
-Here, <concept> can be observation, measurement, characteristic, standard,
-unit, etc.. 
-<cond> is in the form of $func([distinct]~attr)~op~value$, here $op$ can be some
-algorithmmic operators $=$, $\neq$, $>$, or $<$. 
-$func$ are aggregation functions including sum, average, min and max.
-
-The result is a set of $d$, $\{d|d\in D \wedge (value~op~dv = true |dv\in d[attr])\}$.
-\end{definition}
-
-For each of such result, we denote that $d~s.t.~KQ$ meaning that
-dataset $d$ satisfies the query.
-
-This is the baseline keyword style query. 
-The following two queries are the basic queries.
-
-\begin{definition}
-KQ2: $Q_1~context(Q_1.concept,Q_2)~Q_2 $, here $Q_i$ is a basic query
-in the form of $<concept>:<cond>$. 
-\end{definition}
+They can be formalized as the following queries.
 
@@
 \begin{itemize}
-\item $Q_1 = Observation: Species='Picea rubens'$
-\item $Q_2 = <Observation_1: Species='Picea rubens'>
-  IN(Observation_1,Observation_2) <Observation_2:
-  measurement_{unknown}=`California'>$
-\item $Q_3 = Observation: Species='Picea rubens' \wedge
-  count(distinct~Species)\geq 5$ \from{HP}{TODO}
-\item $Q_4$ 
+\item $Observation: \langle Species='Picea~rubens'\rangle$
+\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
+  $IN(Observation_1, Observation_2$) \\
+$Observation_2: \langle Meas_{unknown}=`California'\rangle$
+\item $Observation: \langle Species='Picea~rubens' $\\
+$\wedge
+  count(distinct~Species)\geq 5\rangle$
+\item $??$
 \end{itemize}
@@
 
+\section{Naive method}\label{sec:naive}
+Index the data with all the values, then search over values. 
+This may get wrong results.
+
+{\bf Experiement: show the error rate (e.g., precision, recall, $F_1$ of this method.}
+
 \section{Querying annotated and integrated ecological data}
 
-All the datasets can be imported to a RDBMS for
+As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
+in querying integrated data. 
+One is to rewrite the original query to a series of new queries over
+the data; the other is to materialize the data to a consistent data
+model and then answer the query over the materialized database. 
+In Section \ref{sec:matdb} and \ref{sec:queryrewrite}, we introduce
+our strategies in answering the queries.
 
-To answer such queries, a naive method is to
+\subsection{Framework and operators}
 
-\subsection{Materialized database}
+\subsection{Querying over Materialized database}\label{sec:matdb}
 
-{\bf HP:}{Copy the materials from materiazeDB to here. Skip now.}
+\subsubsection{Materialize data}
 
-\subsection{Query rewriting}
+\subsubsection{Querying materialized database}
+
+\subsection{Query rewriting}\label{sec:queryrewrite}
+
+For query rewriting, the first step is to find the candidate data
+tables and the attribute name lists that can be used to answer our queries. 
+To find the candidate data table, the metadata tables and annotation
+tables are searched. 
+When the candidate data tables and the needed attributes are known, 
+the second step translate the needed queries over the candidate data
+tables.
+
+When a query {\tt concept:cond} is given, the {\tt concept} can only be interpreted using
+annotations. When seeing a basic concept (observation, measurement,
+characteristic, protocal), their related concepts information can be
+retrieved through $A.MeasType$, $A.ObsType$.
+
+From the {\em measuremen type} and $A.Map$, we can get the attribute
+names; Then, using the attribute names and the metadata table (with
+table and column definition), we can get the candidate data tables
+with their corresponding attribute lists.
+
+If there is context, we need to use $A.ContextType$.
+
+If there is distinct constraint, need to find the key measurements. 
+then, perform ``group by'' operation.
+
+From the annotation, find the {\em measurement type} with {\em
+  Characteristic} species. 
+Here, it is $m_3$. From the mapping, we find that the attribute in the
+table is {\em spp}.
 
@@
 concept = observation, cond is {\em species = ``Picea rubens''};
-
-The first step is to find the dataset that contain all the metadata. 
-From the annotation, find the {\em measurement type} with {\em
-  Characteristic} species. 
-Here, it is $m_3$. From the mapping, we find that the attribute in the
-table is {\em spp}.
 
@@
 
-%\bibliographystyle{abbrv}
-%\bibliography{model}
-%\bibliographystyle{IEEEtran}
-%\bibliography{IEEEabrv,DataIntegration}
-
-%\begin{thebibliography}{1}
-%\bibliographystyle{abbrv}
-%\bibliography{model}
-%\end{thebibliography}
+\bibliographystyle{IEEEtranS}
+\bibliography{IEEEabrv,DataIntegration}
 

### 2010-04-29 docs/algdraft/DataIntegration/DataIntegration.tex
Refine the model and query rewriting.

@@
 
-\title{Querying Integrated Ecological Data}
+\title{Querying Integrated Scientific Observation and Measurement Data}
 
@@
 
-In the ecology domain, most data people collected are observation and
-measurement data. {\bf Add one sentence to explain this.}
-Naturally, when searching ecological data, people are more interested in finding datasets
-related to specific observation and measurement. 
+In many scientific domains (e.g., ecology, hydrology, earth science, geology), people collect observation and
+measurement data. Such data record the observed value of some real
+world entity at some specific place. E.g., ecologist studying the
+growth pattern and treatments often need to record the tree height,
+The collected data reflect the fact the tree height of a specific tree is 1.0in on
+May 1, 2009 and 1.5in on May 1, 2010. ({\bf check whether this is
+  realistic or not.})
+
+In scientific domains where people collect observation and measurement data, there are
+several commonly used and widely recognized canonical
+concepts(\cite{oboe, om}). 
+In this paper, we refer these concepts as OM concetps. 
+The canonical concepts include {\em observation}, {\em measurement},
+{\em characteristic}, {\em standard}, {\em protoca} or ({\em procedure}). 
+For example, {\bf add an example to illustrate these concepts. }
+
+When a user poses a query, it is unrealistic for a him/her to know the 
+underlying data structure of the data.  
+But they all know the well recognized OM conceps as discussed above. 
+So, naturally, when searching such scientific data, people are more interested in finding datasets
+related to such concepts, e.g., specific observation and measurement.
+
 In this section, we first illustrate the data model. Then, we
@@
 
-\subsection{Data model}
+\subsection{Data model}\label{sec:datamodel}
 
@@
 
-\subsection{Annotation}
+\subsection{Annotation}\label{sec:annotation}.
 In the Ecology domain, people are using annotations to connect
@@
 {\bf TODO: Put a screen-dump of annotation.}
+When some annotations are done over data tables, the system keeps the
+information $\langle AnnotId, DTableId\rangle$.
 
@@
 
-{\bf TODO:} Describe the constraints in annotation
+{\bf TODO: Describe the constraints in annotation}
 
-\subsection{Observation and measurement query}
+\subsection{Observation and measurement (OM) query}
 
-{\bf TODO: Have a running example from.}
+{\bf TODO: Have a running example from here.}
 Observation and measurement (OM) query
@@
 
-For query rewriting, the first step is to find the candidate data
-tables and the attribute name lists that can be used to answer our queries. 
-To find the candidate data table, the metadata tables and annotation
-tables are searched. 
-When the candidate data tables and the needed attributes are known, 
-the second step translate the needed queries over the candidate data
+As described in Section \ref{sec:dataquery}, 
+a user's query generally does not contain the underlying data
+structure because the user cannot get such information. 
+On the contrary, they formulate their query with the well recognized
+OM concept terminologies such as observation
+measurement, characteristic, etc. 
+To answer such query from the original data model (i.e., dataset
+metadata and data tables), we need to {\em rewrite} the quey over the
+real data table. 
+In what follows, we first sketch the re-writing process. Then, we
+detailed the procedure in using the data model and
+structures. Finally, we include the process for the more complicated
+cases (e.g., with distinct, aggregate).
+
+Roughly speaking, the query rewriting consists of two steps:
+\begin{itemize}
+\item From the given query, find the relevant data tables and
+  attributes that need to be used answer the given query. 
+\item Translate the given query to queries over the relevant data
   tables. 
+\end{itemize}
 
-When a query {\tt concept:cond} is given, the {\tt concept} can only be interpreted using
-annotations. When seeing a basic concept (observation, measurement,
-characteristic, protocal), their related concepts information can be
-retrieved through $A.MeasType$, $A.ObsType$.
+{\bf TODO: put a figure here to illustrate the steps}.
 
-From the {\em measuremen type} and $A.Map$, we can get the attribute
-names; Then, using the attribute names and the metadata table (with
-table and column definition), we can get the candidate data tables
-with their corresponding attribute lists.
+The first step is to map the OM query to the real data structure. 
+We can utilize the annotation structure $A.Map$ which keeps the correspondences between
+measurement type and table attributes. 
+So, when the given concept is {\em measurement}, we can directly get
+its related attribute names and annotation id $AnnotId$. 
+In case that the given OM concept is a non-measurement concept, we can use
+other Annotation structures $A.*$ to figure out the measurementt types. 
+E.g., if the OM concept is Characteristic ``Species'', we can use
+$A.MeasType$ to get the measurement type id.
 
+As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
+information $\langle AnnotId, DTableId\rangle$. 
+With the $AnnotId$, we can ge the data table information from the
+metadata  $\langle AnnotId, DTableId\rangle$. 
+Then, we get the relevant data table information and the needed attribute
+information.
+
+Let $RD$ be the set of relevant data tables and $d_{a_1} \cdots d_{a_m}$
+are the related attributes for table $d\in RD$. We can translate the original OM query to SQL queries
+over the data tables.
+
+{\tt 
+SELECT DISTINCT d.record\_id\\
+FROM d\\
+WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
+$[$GROUP BY$]$\\
+$[$HAVING$]$
+}
+
+The above describes the process for the most basic OM query. 
+However, in the real application, people ask queries with distinct
+observation or entity constraints. Also, people ask queries about
+different observatitons using context.
+
+{\bf TODO: Discuss how to deal with the distinct constraints. }
+If there is distinct constraint, need to find the key measurements. 
+then, perform ``group by'' operation.
+
+{\bf TODO: Discuss how to deal with context}. 
 If there is context, we need to use $A.ContextType$.
 
-If there is distinct constraint, need to find the key measurements. 
-then, perform ``group by'' operation.
+% Use the materialized database, we would have four tables for all the datasets: 
+% entity, observation, measurement, context. 
+% and 4 tables for entity types, observation types, measurement types,
+% and context relationships.
+
+\begin{example}
+{\bf TODO: refine this example}
+Take KQ1 as example. 
+concept = observation, cond is {\em species = ``Picea rubens''};
 
@@
 table is {\em spp}.
-
-The information that users can have: metadata about the table.
-
-In the scientific data, it is unrealistic for the users to know the
-table structure to perform the query.
-
-Use the materialized database, we would have four tables for all the datasets: 
-entity, observation, measurement, context. 
-and 4 tables for entity types, observation types, measurement types,
-and context relationships.
-
-To answer KQ1: it's simple join between
-
-\begin{example}
-Take KQ1 as example. 
-concept = observation, cond is {\em species = ``Picea rubens''};
 

### 2010-04-30 docs/algdraft/DataIntegration/DataIntegration.tex
Rerich query rewriting processing.

@@
 
-\section{Querying annotated and integrated ecological data}
+\section{Querying annotated and integrated OM data}
 
@@
 \subsection{Framework and operators}
-
-\subsection{Querying over Materialized database}\label{sec:matdb}
-
-\subsubsection{Materialize data}
-
-\subsubsection{Querying materialized database}
 
@@
 
-\begin{example}
-{\bf TODO: refine this example}
-Take KQ1 as example. 
-concept = observation, cond is {\em species = ``Picea rubens''};
+\begin{example}\label{eg:query_reqerite_q1}
+Take $Q_1$ as example. The {\tt concept} is ``observation'' and {\tt cond} is {\em species = ``Picea rubens''};
 
-From the annotation, find the {\em measurement type} with {\em
-  Characteristic} species. 
-Here, it is $m_3$. From the mapping, we find that the attribute in the
-table is {\em spp}.
+From the annotation $A.MeasType$, we can find the measurement type
+$m_3$ with the characteristic {\em species}. 
+From the mapping $A.Map$, we further find that the attribute {\em spp} that the
+measurement type $m_3$ is mapped to.
 
-The second step is to find the data file which really have this
-value. 
-For this one, we find the data table, and do a select on the table
-content.
+The second step is to find the data file which really have this value. 
+For this one, we find the data table, and do a selection on the table content. 
+\end{example}
 
-Use materialized database: 
-The first step is the same, after we know the data file, then we
-search the table {\em measurement\_instance} with the {\em mvalue=``Picea rubens''}.
+From the above description, we can tell that the computation cost is to search the different data
+tables. This way, for each candidate data table, we need to send an
+SQL query to the server for evaluation purpose.
+If the number of candidate data tables are small. This method should
+work well. However, when we have a lot of candidate data tables, the
+efficiency may be affected. 
+In what follows, we introduce another strategy, which logically merge
+the data content in different tables and perform queries ober the
+materialized database.
 
-In this example, the computation cost is to search the different data
-tables. To reduce this computation cost, we can materialize just the
-measurement values. (How to do this? ) need more details.
+\subsection{Querying over materialized database}\label{sec:matdb}
 
-Half materialized: how to perform query over it. 
+The previous section shows the method to rewrite a given query. 
+In this section, we would discribe another strategy.
+
+In this method, we merge the data in different tables into several
+core OBOE tables, denoted by $OBOE.*$
+In detail,
+\begin{itemize}
+\item $OBOE.EntityIns=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
+\item $OBOE.ObsIns=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
+\item $OBOE.MeasIns=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
+\item $OBOE.ContextIns =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
+\end{itemize}
+
+\subsubsection{Materialize data}
+{\bf: TODO: refine the writing in MaterializeDB and move them here.}
+
+\subsubsection{Querying materialized database}
+
+\begin{example} \label{eg:materialize_db_q1}
+Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
+the query rewriting method. 
+Using the  materialized database, the first step is the same.
+After we know the data file, then we search the table {\em measurement\_instance} with the {\em mvalue=``Picea rubens''}. 
 \end{example}

### 2010-04-30 docs/algdraft/DataIntegration/DataIntegration.tex
Finish the rough writeup for query-rewriting.

@@
 \item $Q_3$: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
-\item $Q_4$: Give me the datasets that have measurements with average ``area''
-bigger than 5.0 square feet. 
+\item $Q_4$: Give me the datasets that have trees with average ``height''
+  than 20in in year 2009. 
+\item $Q_5$: an example on summarize one measure with respect to another.
+This could take two forms: one where the context forms the groups
+(like by year), and another where the two measurements are within the
+same context (like finding correlations).
 \end{itemize}
@@
 
-{\bf Experiement: show the error rate (e.g., precision, recall, $F_1$ of this method.}
+{\bf TODO: An example to show that the naive method give wrong answers.}
+{\bf Experiement: show the error rate (e.g., precision, recall, (or
+  directly $F_1$) of this method.}
 
@@
 
+[
 {\tt 
@@
 }
+]
 
@@
 
-{\bf TODO: Discuss how to deal with the distinct constraints. }
-If there is distinct constraint, need to find the key measurements. 
-then, perform ``group by'' operation.
+%Discuss how to deal with the distinct constraints.
+Many times, a user may want to find {\em distinct} observations or
+entities. In this case, the above simple solution cannot work
+correctly because it treats observations in different rows as the same
+observation. So, we need to use the annotation to figure out what
+distinguishes one observation from another. 
+When an observation type is denoted with {\em distinct yes} and some
+of its measurement types are denoted with {\em key yes}, 
+the values of the measurements on the key measurement types can
+uniquely distinguish one observation from the other. 
+To implement this, in SQL, we can perform ``group by'' operations on
+the key measurement types of the observations.
 
-{\bf TODO: Discuss how to deal with context}. 
-If there is context, we need to use $A.ContextType$.
+%Discuss how to deal with context
+The above description discusses cases without context. 
+However, in real application, the observation has context. 
+When the context is not {\em identifying}, i.e., the context value
+does not affect the uniqueness of one observation, this can be
+processed as the basic way. 
+When the context is denoted with {\em identifying yes}, it means the
+identity of one observation also depend on the contextual
+observation. 
+In this case, we need to figure out the all the key measurement types
+of one observation, then the ``GROUP BY'' operation should be on all
+these key measurement types.
 
-% Use the materialized database, we would have four tables for all the datasets: 
-% entity, observation, measurement, context. 
-% and 4 tables for entity types, observation types, measurement types,
-% and context relationships.
+{\bf TODO: put an example here.}
+
+In this case, there is no need to change the SQL scripts.
+the only change are the relevant attribute set $d_{a_1} \cdots
+d_{a_m}$, which corresponds to not only the measurement types that
+directly related to the observation, but also the measurement types
+that are in the identifying context chain of the observation.
 
 \begin{example}\label{eg:query_reqerite_q1}
+{\bf TODO: refine this example} 
 Take $Q_1$ as example. The {\tt concept} is ``observation'' and {\tt cond} is {\em species = ``Picea rubens''};
@@
 The second step is to find the data file which really have this value. 
-For this one, we find the data table, and do a selection on the table content. 
+For this one, we find the data table, and do a selection on the table
+content.
+
+{\bf TODO: SQL here.} 
 \end{example}

### 2010-04-30 docs/algdraft/DataIntegration/DataIntegration.tex
add more on the materialized db part.

@@
   than 20in in year 2009. 
-\item $Q_5$: an example on summarize one measure with respect to another.
-This could take two forms: one where the context forms the groups
-(like by year), and another where the two measurements are within the
-same context (like finding correlations).
+% Shawn: an example on summarize one measure with respect to another.
+% This could take two forms: one where the context forms the groups
+% (like by year), and another where the two measurements are within the
+% same context (like finding correlations).
 \end{itemize}
@@
 \section{Naive method}\label{sec:naive}
-Index the data with all the values, then search over values. 
+A very direct way to help searching would be to index the data with
+the data with all the values, then search over these values. 
 This may get wrong results.
@@
 
-\subsection{Framework and operators}
+%\subsection{Framework and operators}
 
@@
 The previous section shows the method to rewrite a given query. 
-In this section, we would discribe another strategy.
+However, as we analyzed, the computation cost may be higher when the
+candidate data tables are many. In this section, we propose another
+strategy to make use of the existing optimization strategies of
+current DBMS.
 
-In this method, we merge the data in different tables into several
-core OBOE tables, denoted by $OBOE.*$
-In detail,
+The basic idea is to materialize the data into some centralized
+concept tables with instances of entities, observations, measurements
+and the contexts between observations. 
+We call such concept tables core OBOE tables, and denote them as
+$OBOE.*$. 
+In partibular, we have the following concept instance tables. 
 \begin{itemize}
@@
 
-\subsubsection{Materialize data}
+In what follows, we describe how to materialize the data into such
+concept instance tables in Section \ref{sec:materialilzedb} when
+considering all the distinct and context constratins. 
+Then, in Section \ref{sec:materialilzedb}, we describe our strategy to
+perform queries over materialized database.
+
+\subsubsection{Materialize data}\label{sec:materialilzedb}
 {\bf: TODO: refine the writing in MaterializeDB and move them here.}
 
-\subsubsection{Querying materialized database}
+\subsubsection{Querying materialized database}\label{sec:materialilzedb}
 

### 2010-05-07 docs/algdraft/DataIntegration/DataIntegration.tex
Refine the definition and add introduction. Is refining the materialization part.

@@
 \section{Introduction}
-In this work, we study the query processing over integrated ecological
-data through OBOE model\cite{DBLP:conf/er/BowersMS08}, a conceptual model used to interpret
-observation and measurement data.
+In this work, we study the query processing over scientific
+observation and measurement data using OBOE
+model\cite{DBLP:conf/er/BowersMS08}. OBOE model is a conceptual model
+used to interpret observation and measurement data.
 
 \subsection{Background}
-Introduce OBOE, annotation
-Give an example of them.
+
+%Application background
+In many scientific domains (e.g., ecology, hydrology, earth science,
+geology), people collect observational data. Such data
+record the observed value of some real world entity at some specific
+place and time. E.g., ecologists studying relationship between the 
+growth pattern and the treatments often need to record the tree
+heights. The collected data reflect the fact the tree height of a
+specific tree is 30.1in on May 1, 2009 and 30.3in on May 1, 2010.
+
+%Characteristics of scientific observational data: non-normalized
+Almost all such scientific data do not follow database higher normal
+forms. Generally, scientists have their way in interpreting their
+data, but they are not ready for any normalization process. 
+For example, Table \ref{tb:dataset} is a simplified but typical dataset collected by
+a scientist who study the growth pattern of trees. 
+Obviously, the ``plt'' of the first two rows is the same place, and
+the ``plt'' of the last two rows is the same. 
+Here, their area information is redundant. 
+In the real application, many columns have redundant information.
+
+%HP: this one has the coordinates, it's troublesome to explain. 
+%% \begin{table}[htb]
+%% \begin{center}
+%% \begin{tabular}{|l|l|l|l|l|}
+%% \hline
+%% code & no & ht & plt & area\\\hline
+%% piru & 1 & 35.8 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
+%% piru & 2 & 36.2 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
+%% piru & 1 & 25.7 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0 \\\hline
+%% %abba & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
+%% capo & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
+%% \end{tabular}
+%% \end{center}
+%% \caption{Dataset}
+%% \label{tb:dataset}
+%% \end{table}
+
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+code & no & ht & plt & area\\\hline
+piru & 1 & 35.8 & California & 4.0\\\hline
+piru & 2 & 36.2 & California & 4.0\\\hline
+piru & 1 & 25.7 & Oregon & 3.0 \\\hline
+%capo & 1 & 7.8 & Oregon & 3.0\\\hline
+Oriental poppy& 1 & 7.8 & Oregon & 3.0\\\hline 
+\end{tabular}
+\end{center}
+\caption{Dataset}
+\label{tb:dataset}
+\end{table}
+%HP: California poppy info
+%Oregon iris: http://en.wikipedia.org/wiki/List_of_native_Oregon_plants
+%http://en.wikipedia.org/wiki/California_poppy
+
+%Introduce major concepts
+In scientific domains where people collect observation and measurement
+data, there are several commonly used and widely recognized canonical
+concepts(\cite{oboe, om}). 
+In this paper, we refer these concepts as OM concepts. 
+The canonical concepts include {\em observation}, {\em measurement},
+{\em characteristic}, {\em standard}, {\em protocol} or ({\em procedure}). 
+For example, {\bf add an example to illustrate these concepts. }
+
+%Introduce annotation, Motivation to use annotation
+Generally, every scientific dataset goes into the data repository with
+some metadata, e.g., Dublin core metadata\cite{***}, Darwin core metadata\cite{***}. 
+However, all these metadata are at the dataset level, they did not
+provide enough information for the data {\em content} inside the
+dataset. 
+To better make use of the data content in the repository, more systems
+are embracing the ideas of having metadata on the data content. 
+E.g., Some systems \cite{tdar} provides a mechanism to collect the
+column/attribute level metadata, some system \cite{semtools} uses
+{\em annotation} to add more semantic information to the data
+content. In this work, we follow the terminology in \cite{semtools} and
+use the term {\em annotation} to distinguish the 
+metadata on the data content from those on data objects at a coarser
+granularity. 
+%How to represent the annotation and the symbols used
+We propose tools for scientists to provide annotations to scientific
+observational data. 
+\from{HP}{Put a screen-dump of annotation.}
+
+%Challenge to query such data
+\from{HP}{Queries to such data and challenges to answer such queries.}
+When a user formulates a query, it is unrealistic for a him/her to know the 
+underlying data structure of the data.  
+But they all know the well recognized OM concepts as discussed above. 
+So, naturally, when searching such scientific data, people are more
+interested in finding datasets related to such OM terminologies as observation
+measurement, characteristic, etc. 
+Given the dataset in Table \ref{tb:dataset}. 
+People may ask the following queries.
+
+\begin{example}\label{eg:query}
+Simple and summarization queries: 
+\begin{itemize}
+\item $Q_1$: Give me the data sets that contain species ``Picea rubens'' observations.
+%$Q_2$: Give me the data sets that have measurements on ``area''  characteristics. 
+\item $Q_2$: Give me the data sets that contain species ``Picea rubens''
+  observations in ``California''. 
+\item $Q_3$: Give me the data sets that contain at least five distinct
+  ``Picea rubens'' observations.
+\item $Q_4$: Give me the data sets that have trees with average ``height''
+  than 20.0 in ``California''. 
+% Shawn: an example on summarize one measure with respect to another.
+% This could take two forms: one where the context forms the groups
+% (like by year), and another where the two measurements are within the
+% same context (like finding correlations).
+\end{itemize}
+\end{example}
+
+\from{HP}{Current data integration effort}
+
+\from{HP}{Our effort in querying observational data: introduce OBOE, query}
+
+\from{HP}{Need re-organization. The description in the
+  following several sections may be moved to here later.}
 
@@
 
+\from{HP}{This will come after the real problem definition and the
+  solution. The description in the following several sections may be
+  moved to here later.}
+
 \section{Data model, annotation and query}\label{sec:dataquery}
-
-In many scientific domains (e.g., ecology, hydrology, earth science, geology), people collect observation and
-measurement data. Such data record the observed value of some real
-world entity at some specific place. E.g., ecologist studying the
-growth pattern and treatments often need to record the tree height,
-The collected data reflect the fact the tree height of a specific tree is 1.0in on
-May 1, 2009 and 1.5in on May 1, 2010. ({\bf check whether this is
-  realistic or not.})
-
-In scientific domains where people collect observation and measurement data, there are
-several commonly used and widely recognized canonical
-concepts(\cite{oboe, om}). 
-In this paper, we refer these concepts as OM concetps. 
-The canonical concepts include {\em observation}, {\em measurement},
-{\em characteristic}, {\em standard}, {\em protoca} or ({\em procedure}). 
-For example, {\bf add an example to illustrate these concepts. }
-
-When a user poses a query, it is unrealistic for a him/her to know the 
-underlying data structure of the data.  
-But they all know the well recognized OM conceps as discussed above. 
-So, naturally, when searching such scientific data, people are more interested in finding datasets
-related to such concepts, e.g., specific observation and measurement.
 
@@
 
-When a scientist contributes dataset into the data repository, generally
-each dataset is converted to a data table\cite{tdar} or treated as a separate
-object entity\cite{knb}. 
-So, the database contains metadata about each dataset, the data table
-definition (e.g., attribute/column name, attribute type, etc). 
-%\from{HP}{Another way to organize the data is to store them in one big
+%Data model: Data, attribute, data value set
+When a scientist contributes data into an integrated data repository,
+a widely accepted way is to convert each dataset to a data
+table\cite{tdar} or treated as a separate object entity\cite{knb}. 
+{\bf (HP: is it really widely used? any other system uses this
+  way?. Add more citations here.)}
+In using this method, the database contains metadata of each
+dataset and the definition of the data table (e.g., attribute/column
+name, attribute type, etc).
+
+%HP: 
+%An alternative way to organize the data is to store them in one big
 %data table with three columns data\_file, attribute\_name,
-%attribute\_value. The problem with this organization is that the
-%attribute\_value are different types. } 
-To formalize the scenario, we use $D$ to denote the set of data
-tables. Each data table contains metadata about the attribute
-definition.  Sometimes, we also use {\em column} to denote the attribute. 
-Let $d$ be one data table, 
-$d[attr]$ or $d[idx]$ represents the set of values for the attribute
-$attr$ or for the $idx$-th attribute. 
-Besides the dataset, in the data store, we also keep the annotation
-information. 
-%Implementation detail (no need to put it in the paper): 
+%attribute\_value. The problem with this structure is that the
+%attribute\_value are different types.
+In our work, we focus on querying this kind of data repository (or
+databases). To formalize the scenario, we use $D$ to denote the set of data
+tables in the data repository and $d$ to refer to a specific data
+table. Each data table $d$ contains metadata about the attribute
+definition $Attr_{d}$. Sometimes, we also use {\em column} to refer to an
+attribute. 
+Given one data table $d$, an attribute $attr_i \in Attr_{d}$ or column index $i$, 
+$d[attr_i]$ or $d[i]$ represents the set of values for the attribute
+$attr_i$ or for the $i$-th column.
+%HP: Implementation detail (no need to put it in the paper): 
 %When we put the data to data tables, we add a unique record id for
@@
 
-\subsection{Annotation}\label{sec:annotation}.
-In the Ecology domain, people are using annotations to connect
-datasets to the OBOE model. 
-{\bf TODO: Put a screen-dump of annotation.}
-When some annotations are done over data tables, the system keeps the
-information $\langle AnnotId, DTableId\rangle$.
+\subsection{Annotation}\label{sec:annotation}
 
-Map annotation information 
+%What kinds of annotation information is needed.
+
 We use $A$ to denote the annotation of one dataset. 
-More specifically,
-To describe an observation, we need to denote on which entity (object in the
-real world) the observation is made.
-Very often, more than one observation can be made on one entity.   
+Internally, we keep the following information for the
+annotation.
+
+For data table $d$ with annotation $A$, the system keeps the
+information $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$.
+As we described before, we have four main objects to describe: 
+observation type (OT), measurement type (MT), and
+context type (CT), and the mapping (Map) from the measurement type to
+resource attributes.  
+So, the annotation contains the following information.
+
 \begin{itemize}
-\item A.ObsType = $\{\langle \underline{AnnotId,ObsTypeId}, EntTypeId,
-  isDistinct \rangle\}$ 
-\item A.MeasType = $\{\langle \underline{AnnotId, MeasTypeId}, ObsTypeId,
-  CharType, StdType, ProtType, Precision, isKey \rangle\}$ 
-\item A.ContextType = $\{\langle \underline{AnnotId, ObsTypeId, ContextObsTypeId, RelType}, isIdentify\rangle\}$
-\item A.Map = $\{\langle \underline{AnnotId,MeasTypeId, ResAttribute, Cond}, Val\rangle\}$
+\item $A.OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
+  describe an observation type, which denotes on which
+  entity type (object in the real world) the observation is made. 
+Very often, more than one observation can be made on one entity. Thus,{\em
+  isDistinct} is used to denote whether the same value of key
+meaasurements of an obbservation types can uniquely identify one
+observation or not. 
+\item $A.MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
+  Cha, \cdots)\}$. Generally, $MT$ contains information about
+  characteristic ($Cha$), Standard, Protocol and Precisio. We do not include them
+  here to make the description clearer. Here, {\em isKey} is used to denote
+  whether one measurement type is the key measurement for the
+  observation type $OT_{id}$ or not. 
+\item $A.CT$ = $\{(\underline{A_{id}, ot_{id}, cot_{id},
+  Rel}, isIdentify)\}$ where $cot$ represents the context observation
+  type. 
+\item $A.Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, Cond},
+  Val\}$
 \end{itemize}
 
-``AnnotId'' denotes the resource (e.g., data table) the annotation is on.
+We use the following example to illustrate the concepts in the
+annotation.
 
-{\bf TODO: Describe the constraints in annotation}
+\begin{example}
+$A.OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
+  denotes that annotation $A_1$ are for two observation types. One
+  type $OT_1$ is on real world entity ``tree'' and the values of its key measurements
+  do not uniquely identify a distinct observation. The other type
+  $OT_2$ is for real world entity ``GeoSpot'' (i.e., geospatial
+  location) and its key measurement types can identify it's unique
+  observations. \from{HP}{Introduce key measurements before this.}
+
+$A.MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
+($A_1, MT_2, OT_1$, SpecNo, true), 
+($A_1, MT_3, OT_1$, Height, false),
+($A_1, MT_4, OT_2$, Plot\_State, true), 
+($A_1, MT_5, OT_2$, Plot\_Area, false)\} represents that $OT_1$
+(for ``tree'') has three measurement types ($MT_1, MT_2$, $MT_3$) for characteristics
+``Species'', ``SpecNo'', and ``Height''. The fist two measurement
+  types together form the key measurement type for $OT_1$. 
+$OT_2$ (for ``GeoSpot'') has one key measurement type $MT_4$ for
+  characteristic ``Plot\_State'' and another measurement type $MT_5$
+  for characteristic ``Plot\_Area''.
+
+$A.CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
+observation of a {\em tree} is within the context of a geo-spatial
+location. And this context is used to identify the uniqueness of the
+observation. E.g., a tree with species name ``piru'' and species no
+$1$ in California
+%$(34^{\circ}8'3N,118^{\circ}14'37W)$ (Los Angeles in California) 
+is different from a tree with 
+the same species name ``piru'' at Oregon. 
+%location $36^{\circ}10'30N,115^{\circ}8'11W$ (Las Vegas in
+%California).
+
+%HP coordinates material: 
+%http://www.traveljournals.net/explore/united_states/map/u1662328/los_angeles.html
+%http://en.wikipedia.org/wiki/Las_Vegas,_Nevada
+
+$A.Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
+%($A_1$, $MT_1$, ``code'', ``eq `abba''', ``Abies balsamea''),
+%($A_1$, $MT_1$, ``code'', ``eq `capo''', ``California poppy''),
+($A_1$, $MT_2$,``no'', null,), 
+($A_1$, $MT_3$, ``plt'', null,), 
+($A_1$, $MT_4$,``area'', null,\}. 
+The first mapping rules maps ``code''
+attribute to the measurement type $MT_1$ (for ``Species''
+characteristic) and change the value to ``Picea rubens'' if the code
+is ``piru''. The meaning of the other mapping rules are very
+obvious here, so we skip the details. 
+\end{example}
 
 \subsection{Observation and measurement (OM) query}
+Above we defined the data model and the annotation, in this
+sub-section, we define the queries that scientists are interested in
+We denote the queries on the observational data as {\em Observation
+and measurement (OM) query}.
 
-{\bf TODO: Have a running example from here.}
-Observation and measurement (OM) query
+In using the observational data, people are generally interested in
+the following concepts: {\em observation, measurement,characteristic or standard}.
 
@@
 Here, {\tt concept} is the main term in an observation model. 
-In using the OBOE model, it can be observation, measurement,
-characteristic or standard. 
 {\tt cond} is in the form of $f([distinct]~attr)~op~value$ where $op$ is
@@
 $\{d|d\in D \wedge d~satisfies~{\tt cond}\}$. 
-%$\{d|d\in D \wedge (value~op~dv = true, dv\in d[attr])\}$. 
 For each of such result $d$, we use $d~s.t.~Q$ to denote that
@@
 
+Based on the different {\tt cond} definition, ``$d~satisfies~{\tt cond}$''
+is translated into different formulas. 
+E.g., if {\tt cond} is defined that the value of attribute ``area'' need to be
+smaller than $3.0$, then, ``$d~satisfies~{\tt cond}$'' is translated to
+``$dv<3.0|dv\in d[area]$''.
+
 Definition \ref{def:basic_omq} can formulate queries on the
 one specific observation and measurement. 
-However, it fails to catch the observation context relationship. 
-E.g., {\bf TODO: }
+In real application, we need to consider the context relationship. 
 So, we generalize the basic OM query to contextualized OM query as
@@
 
-Given the dataset in Figure \ref{}. 
-People may ask the following queries.
+For example, the queries in Example \ref{eg:query} can be formalized
+as the following formal OM queries.
 
-\begin{itemize}
-\item $Q_1$: Give me the datasets that contain species ``Picea rubens'' observations.
-%$Q_2$: Give me the datasets that have measurements on ``area''  characteristics. 
-\item $Q_2$: Give me the datasets that contain species ``Picea rubens''
-  observations in ``California''. 
-\item $Q_3$: Give me the dataset that contains at least five distinct ``Picea rubens'' observations.
-\item $Q_4$: Give me the datasets that have trees with average ``height''
-  than 20in in year 2009. 
-% Shawn: an example on summarize one measure with respect to another.
-% This could take two forms: one where the context forms the groups
-% (like by year), and another where the two measurements are within the
-% same context (like finding correlations).
-\end{itemize}
-
-They can be formalized as the following queries.
-
-\begin{example}
-The given example queries can be represented in the formal query
-format. 
+%\begin{example}
 \begin{itemize}
@@
 
-\end{example}
+%\end{example}
 
-\section{Naive method}\label{sec:naive}
-A very direct way to help searching would be to index the data with
-the data with all the values, then search over these values. 
-This may get wrong results.
-
-{\bf TODO: An example to show that the naive method give wrong answers.}
-{\bf Experiement: show the error rate (e.g., precision, recall, (or
-  directly $F_1$) of this method.}
+%\from{HP}{Add experiement: show the error rate (e.g., precision, recall, (or
+%  directly $F_1$) of this method.? not needed?}
 
 \section{Querying annotated and integrated OM data}
+
+In this section, we discuss several methods that we can use to
+answer OM queries over scientific observational data with
+annotations.
+
+%\subsection{Naive method}\label{sec:naive}
+
+A very direct way search such data is to extract all the content in
+the data cells and index them, this way, the system may help answer
+very simple queries with the help of some thesaurus. 
+For example, for $Q_1$ in Example \ref{eg:query}, as long as we can
+translate the abbreviation ``piru'' in the dataset to ``Picea rubens''
+based on some predefined thesaurus. We can find the results. 
+Even for this kind of simple query, however, if we want to find
+observation of ``California poppy'', an approximate match may return
+the dataset in Table \ref{tb:dataset} as a result because it contains
+both {\em California} and {\em Poppy}.
+If such compound word search can be alleviated using some existing
+techniques \cite{***}, %citation on compound word querying. 
+it is almost impossible to directly use such IR-style system to answer a query
+in the form of $Q_3$ and $Q_4$. 
+I.e., answer queries with context.
+
+This naive method cannot answer the desired query. It is because it
+fails to catch the semantics in the data. 
+In what follows, we would use the annotation during our search to get
+more semantic support.
 
@@
 model and then answer the query over the materialized database. 
-In Section \ref{sec:matdb} and \ref{sec:queryrewrite}, we introduce
-our strategies in answering the queries.
-
-%\subsection{Framework and operators}
+We what follows we work from these two directions to leverage the
+semantic information in answering such queries.
 
@@
 As described in Section \ref{sec:dataquery}, 
-a user's query generally does not contain the underlying data
-structure because the user cannot get such information. 
-On the contrary, they formulate their query with the well recognized
+an OM query is represented using 
 OM concept terminologies such as observation
@@
 So, when the given concept is {\em measurement}, we can directly get
-its related attribute names and annotation id $AnnotId$. 
+its related attribute names and annotation id $A_{id}$. 
 In case that the given OM concept is a non-measurement concept, we can use
@@
 As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
-information $\langle AnnotId, DTableId\rangle$. 
-With the $AnnotId$, we can ge the data table information from the
-metadata  $\langle AnnotId, DTableId\rangle$. 
+information $\langle A_{id}, DTableId\rangle$. 
+With the $A_{id}$, we can ge the data table information from the
+metadata  $\langle A_{id}, DTableId\rangle$. 
 Then, we get the relevant data table information and the needed attribute
@@
 \begin{itemize}
-\item $OBOE.EntityIns=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
-\item $OBOE.ObsIns=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
-\item $OBOE.MeasIns=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
-\item $OBOE.ContextIns =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
+\item $OBOE.EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
+\item $OBOE.OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
+  all the observation instances;
+\item $OBOE.MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
+  mVal)\}$ for all the measurement instances;
+\item $OBOE.CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
+  all the context instances;
 \end{itemize}
@@
 Using the  materialized database, the first step is the same.
-After we know the data file, then we search the table {\em measurement\_instance} with the {\em mvalue=``Picea rubens''}. 
+After we know the data file, then we search the OBOE.MI with the
+condition that {\em mVal=``Picea rubens''}.
 \end{example}
 
-\begin{example}
-We use $SQ_1$ as an example.
+{\tt 
+SELECT distinct $AD.d_{id}$\\
+FROM MI, MT, AD\\ 
+WHERE $MT.Cha='Species'$ \\
+AND $MI.mVal = `Picea rubens'$\\
+AND $MI.mt_{id}=MT.mt_{id}$ AND $MT.A_{id}=AD.A_{id}$; %\#Join the MI and MT tables
+}
 
-The first step is to find the datasets that contain the metadata. 
-The second step is to perform a query over the materialized database. 
-%select mlabel, mvalue, count(*) as cnt from measurement_instance
-%group by mvalue,mlabel having count(*)>=2;
+In this example, the computation cost is the selection cost of the
+measurement instance table. 
+It should be faster than the query rewriting.
+
+If the query has some restriction on the aggregate result, we can
+easily add the GROUP BY and HAVING clause to deal with it. 
+We take $Q_3$ as an example, which asks for data sets that contain at
+least five distinct ``Picea rubens'' observations.
+The SQL query can be written as: \\
+{\tt 
+SELECT distinct $d_{id}$\\
+FROM MI, MT\\ 
+WHERE $MT.Cha='Species'$ \\
+AND $MI.mVal = `Picea rubens'$\\
+AND $MI.mt_{id}=MT.mt_{id}$\\
+GROUP BY $d_{id},oi_{id}$\\
+HAVING COUNT(*)$>$5; %\#Join the MI and MT tables
+}
 
@@
 chain. 
-\end{example}
+%\end{example}
 
-\begin{example}
-KQ3: Give me the datasets that contain species ``Picea rubens''
-  observations in ``California''.
+When we are given contextualized query, e.g., $Q_2$. 
+Give me the datasets that contain species ``Picea rubens'' observations in ``California''.
 
@@
 another observation which has measurement of value ``California''.
-\end{example}
 
-Using materializedDB, we can answer the query this way
-First, Species --> mlabel, observation type
+%Using materializedDB, we can answer the query this way
+%First, from the Characteristic ``Species'', we can get  --> mlabel, observation type
 
-select mtype, olabel from measurement\_type  \\
-where chracteristic = 'Species';  \\
+%select mtype, olabel from measurement\_type  \\
+%where chracteristic = 'Species';  \\
 
-For the olabel found, \\
-select context\_olabel\\
-from context\_relationship\\
-where olabel = 'olabel';
-
-context\_olabel\_set is a set containing all the observation type
+{\tt 
+SELECT record\_id\\
+FROM CI as ci, MI as mi, MI as cmi,MT as mt, MT as cmt, \\
+WHERE $mi.oi_{id}=ci.oi_{id}$ AND $cmi.oi_{id}=ci.coi_{id}$ \\
+AND $ci.coi_{id} = coi.oi_{id}$ AND oi.otype = 'olabel' \\
+AND mi.mvalue=`Picea rubens' AND cmi.mvalue='California'\\
+AND MT.mt_{id}=mi.mt_{id} AND MT.mLabel = 'Species'\\
+AND CMT.mt_{id}=cmi.mt_{id} AND cmt.mLabel=`State'\\
+AND coi.otype IN (\\
+\verb|    |SELECT context\_olabel\\
+\verb|    |FROM context\_relationship\\
+\verb|    |WHERE olabel = 'olabel');
+}
+The sub-query finds the set containing all the observation type
 labels in the context chain.
-
-SELECT record\_id\\
-FROM observation\_instance oi, observation\_instance coi,
-context\_instance ci,
-measurement\_instance mi, measurement\_instance cmi \\
-WHERE mi.oid=ci.oid AND ci.oid = oi.oid and  cmi.oid=coi.oid AND ci.context\_oid =
-coi.oid and oi.otype = 'olabel' and coi.otype $\in$
-context\_olabel\_set
-AND mi.mvalue=`Picea rubens' and cmi.mvalue='California';\\
 

### 2010-05-07 docs/algdraft/DataIntegration/DataIntegration.tex
Refine the part for querying materialized database. The big structure of the paper is set. Need further refinement....

@@
 % affiliations
-\author{\IEEEauthorblockN{Michael Shell}
+\author{\IEEEauthorblockN{Author1}
 \IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
@@
 \and
-\IEEEauthorblockN{Homer Simpson}
+\IEEEauthorblockN{Author2}
 \IEEEauthorblockA{Twentieth Century Fox\\
@@
 \and
-\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
+\IEEEauthorblockN{Author3}
 \IEEEauthorblockA{Starfleet Academy\\
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 In this work, we study the query processing over scientific
@@
 
-%Challenge to query such data
-\from{HP}{Queries to such data and challenges to answer such queries.}
-When a user formulates a query, it is unrealistic for a him/her to know the 
+%Queries to such data and challenges to answer such queries.
+The internal data structure of such scientific data repository is
+generally unknown to a user who wants to pose a query to such scientific data
+repository and get some useful information. 
+So, it is unrealistic for a him/her to formulate a query based on the 
 underlying data structure of the data.  
-But they all know the well recognized OM concepts as discussed above. 
-So, naturally, when searching such scientific data, people are more
-interested in finding datasets related to such OM terminologies as observation
-measurement, characteristic, etc. 
+The easiest query they can pose to such data repository are keyword-query. 
+Such keyword style query works well for the dataset metadata
+(e.g., contributor information, dataset description, etc) using the
+current Information Retrieval (IR) techniques. However, it
+is far from enough to be used to search the data content because the
+data content need to be interpreted with semantics. 
+It is important in the scientific domain to have typical
+kinds of queries. One fact we can use is that the domain
+scientists know well the widely used OM concepts as discussed above. 
+So, naturally, when searching such scientific data, they can provide
+the concept information to restrain their queries and to find data sets
+related to such OM terminologies. 
 Given the dataset in Table \ref{tb:dataset}. 
-People may ask the following queries.
+People may ask the following observation and measurement (OM) queries.
 
@@
 \end{example}
+\from{HP}{Challenges to answer such queries: uniqueness}
 
@@
 
-\from{HP}{Our effort in querying observational data: introduce OBOE, query}
+\from{HP}{Our effort in querying observational data: introduce OBOE}
 
@@
 \begin{itemize}
-\item Formalize the queries over observation and measurement query.
-\item Propose three methods to evaluate queries. 
+\item We formulate a canonical set of queries over observation and
+  measurement scientific data repository. Such formal queries can
+  formalize most OM concepts related searches. 
+\item We propose three methods to evaluate such OM
+  queries. \from{HP}{Elaborate the three methods.}
 \end{itemize}
@@
 Section \ref{sec:relatedwork} reviews works that are related to this
-research. Section \ref{sec:dataquery} formalize the data model and the
+research. Section \ref{sec:dataquery} formalizes the data model and the
 queries that people are interested to ask.
@@
 %What kinds of annotation information is needed. 
-
 We use $A$ to denote the annotation of one dataset. 
@@
 \begin{itemize}
-\item $A.OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
+\item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
   describe an observation type, which denotes on which
@@
 observation or not. 
-\item $A.MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
+\item $MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
   Cha, \cdots)\}$. Generally, $MT$ contains information about
-  characteristic ($Cha$), Standard, Protocol and Precisio. We do not include them
+  characteristic ($Cha$), Standard, Protocol and Precision. We do not include them
   here to make the description clearer. Here, {\em isKey} is used to denote
@@
   observation type $OT_{id}$ or not. 
-\item $A.CT$ = $\{(\underline{A_{id}, ot_{id}, cot_{id},
-  Rel}, isIdentify)\}$ where $cot$ represents the context observation
+\item $CT$ = $\{(\underline{A_{id}, ct_{id}}, ot_{id}, cot_{id}, Rel,
+  isIdentify)\}$ where $cot$ represents the context observation
   type. 
-\item $A.Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, Cond},
+\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, MapCond},
   Val\}$
@@
 \begin{example}
-$A.OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
+$OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
   denotes that annotation $A_1$ are for two observation types. One
@@
 
-$A.MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
+$MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
 ($A_1, MT_2, OT_1$, SpecNo, true), 
@@
 
-$A.CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
+$CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
 observation of a {\em tree} is within the context of a geo-spatial
@@
 
-$A.Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
+$Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
 %($A_1$, $MT_1$, ``code'', ``eq `abba''', ``Abies balsamea''),
@@
 \subsection{Observation and measurement (OM) query}
-Above we defined the data model and the annotation, in this
-sub-section, we define the queries that scientists are interested in
-We denote the queries on the observational data as {\em Observation
+Above we defined the data model and the annotation. In this
+sub-section, we formalize the queries that scientists are interested
+in. We denote the queries on the observational data as {\em Observation
 and measurement (OM) query}.
 
-In using the observational data, people are generally interested in
-the following concepts: {\em observation, measurement,characteristic or standard}.
+As discussed in Section \ref{sec:intro}, it is generally impossible for a
+scientist to formulate an OM query using the internal data structure
+of a dataset because such data structures differ from contributors to
+contributors. More realistically, they can pose queries
+(Example \ref{eg:query}) related to key OM concepts, e.g., {\em
+  observation, measurement,characteristic, and standard}, and these
+OM concepts satisfy given keyword and context conditions.
 
 \begin{definition}[Basic OM query]\label{def:basic_omq}
-A basic observation and measurement query is defined as $Q$::={\tt
-  concept:cond}. 
-Here, {\tt concept} is the main term in an observation model. 
-{\tt cond} is in the form of $f([distinct]~attr)~op~value$ where $op$ is
-basic operator from $\{=, \neq, >, <\}$ and $f$ is an
+A basic observation and measurement query is defined as 
+\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]. 
+Here, 
+\begin{itemize}
+\item {\tt OMconcept} refers to one of the major OM terminologies. 
+\item 
+$\langle${\tt cond}$\rangle$ is in the form of 
+{\tt f([distinct entity.]measurement) op value} 
+where {\tt op} is a basic operator from $\{=, \neq, >, <\}$ 
+and {\tt f} is an
 aggregation function from $\{sum, avg, count, min, max\}$. 
+{\tt distinct} and {\tt entity} are optional. 
+\end{itemize}
+\from{HP}{The symbol fonts are a little bit messy...need to
+  distinguish between the keywords and variables.}
 \end{definition}
 
-\begin{definition}\label{def:qresult}
+\begin{definition}[Result of OM query]\label{def:qresult}
 The result of a query $Q$ is a set of data objects (e.g., data tables) 
-$\{d|d\in D \wedge d~satisfies~{\tt cond}\}$. 
+\{$d|d\in D \wedge d~satisfies~\langle ${\tt cond} $\rangle$\}. 
 For each of such result $d$, we use $d~s.t.~Q$ to denote that
-data object $d$ satisfies the query. 
+an data object $d$ satisfies the query. 
 \end{definition}
 
-Based on the different {\tt cond} definition, ``$d~satisfies~{\tt cond}$''
-is translated into different formulas. 
-E.g., if {\tt cond} is defined that the value of attribute ``area'' need to be
+Based on the different $\langle${\tt cond}$\rangle$ definitions,
+``$d~satisfies~\langle${\tt cond}$\rangle$''
+is translated differently. 
+E.g., if $\langle${\tt cond}$\rangle$ is defined that the value of ``{\em area}'' need to be
 smaller than $3.0$, then, ``$d~satisfies~{\tt cond}$'' is translated to
@@
 
-Definition \ref{def:basic_omq} can formulate queries on the
+Definition \ref{def:basic_omq} can formulate queries on
 one specific observation and measurement. 
 In real application, we need to consider the context relationship. 
+\from{HP}{E.g., etc.}
 So, we generalize the basic OM query to contextualized OM query as
@@
 
-\begin{definition}[Contextualized OM query]\label{def:context_omq}
-$CQ$::=$Q_1~${\tt context}($Q_1.concept, Q_2.concept$)$~Q_2$. Here $Q_i$ is a basic query
-in the form of {\tt concept:cond}. 
+\begin{definition}[Contextualized OM query]\label{def:context_omq} A
+  contextualized OM query is defined as:
+\[CQ ::=Q_i \wedge {\tt context}(Q_i.{\tt OMconcept}, Q_j.{\tt
+  OMconcept})\wedge Q_j\]
+Here, $Q_i, Q_j$ are basic queries.
 \end{definition}
@@
 \item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
-  $IN(Observation_1, Observation_2$) \\
-$Observation_2: \langle Meas_{unknown}=`California'\rangle$
-\item $Observation: \langle Species='Picea~rubens' $\\
-$\wedge
-  count(distinct~Species)\geq 5\rangle$
-\item $??$
+  $\wedge IN(Observation_1, Observation_2$) \\
+$\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
+\item $Observation: \langle Species='Picea~rubens'$\\
+$\wedge count(distinct~Species)\geq 5\rangle$
+\item $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
+$\wedge IN(Observation_1, Observation_2)$\\
+$\wedge Observation_2: \langle State='California'\rangle$\\
+%Give me the data sets that have trees with average ``height''
+%  than 20.0 in ``California''
 \end{itemize}
@@
 
-In this section, we discuss several methods that we can use to
+In this section, we propose several methods that can be used to
 answer OM queries over scientific observational data with
@@
 %\subsection{Naive method}\label{sec:naive}
-
-A very direct way search such data is to extract all the content in
-the data cells and index them, this way, the system may help answer
+A very direct way to search such data is to extract all the content in
+the data cells and index them. This way the system may help answer
 very simple queries with the help of some thesaurus. 
@@
 translate the abbreviation ``piru'' in the dataset to ``Picea rubens''
-based on some predefined thesaurus. We can find the results. 
-Even for this kind of simple query, however, if we want to find
+based on some predefined thesaurus, we can find the results. 
+However, even for this kind of simple query, if we want to find
 observation of ``California poppy'', an approximate match may return
@@
 both {\em California} and {\em Poppy}.
-If such compound word search can be alleviated using some existing
+If the search problem caused by the compound words can be alleviated using some existing
 techniques \cite{***}, %citation on compound word querying. 
-it is almost impossible to directly use such IR-style system to answer a query
+it is still almost impossible to directly use such IR-style system to answer a query
 in the form of $Q_3$ and $Q_4$. 
-I.e., answer queries with context.
-
-This naive method cannot answer the desired query. It is because it
-fails to catch the semantics in the data. 
-In what follows, we would use the annotation during our search to get
-more semantic support.
+This naive method cannot be directly used answer the desired queries. It is because it
+fails to catch the semantics of the context and uniqueness of observations in the data. 
+In what follows, we utilize the annotation information to get
+more semantic support, and provide new strategies to answer the
+queries.
 
@@
 model and then answer the query over the materialized database. 
-We what follows we work from these two directions to leverage the
+In what follows we work from these two directions to leverage the
 semantic information in answering such queries.
@@
 
-[
+\vspace{0.1in}
 {\tt 
-SELECT DISTINCT d.record\_id\\
-FROM d\\
+\begin{tabular}{l} 
+SELECT DISTINCT $d_{id}$\\
+FROM $d$\\
 WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
-$[$GROUP BY$]$\\
-$[$HAVING$]$
+$[$GROUP BY {\em distinct requirement}$]$\\
+$[$HAVING {\em aggregation condition}$]$;
+\end{tabular}
 }
-]
+\vspace{0.1in}
 
@@
 
+\from{HP}{Refine the process of $Q_2$.\\
+For the first step, we find the data tables that contain the needed concepts. 
+In each table, we need to figure out the distinct observations. 
+From $m_i$ $obstype_i$, all the other measurements that are of
+$obstype_i$. 
+Group by the key measurements of $obstype_i$ if distinct yes.
+If this observation type does not have any context, then the key
+measurements are all the measurements directly defined under this
+observatioin type. Otherwise, i.e., this observation type has context,
+then the key measurements are all the measurements in the context
+chain. }
+
+%Analysis of this method.
+
+{\bf Execution cost analysis:} 
 From the above description, we can tell that the computation cost is to search the different data
@@
 
-\subsection{Querying over materialized database}\label{sec:matdb}
+\subsection{Querying materialized database}\label{sec:matdb}
 
@@
 \begin{itemize}
-\item $OBOE.EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
-\item $OBOE.OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
-  all the observation instances;
-\item $OBOE.MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
-  mVal)\}$ for all the measurement instances;
-\item $OBOE.CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
-  all the context instances;
+%\item $EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
+\item $OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
+  all the observation instances where $ot_{id}$ is used to link to the
+  annotation information in $OT$; 
+\item $MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
+  mVal)\}$ for all the measurement instances where $mt_{id}$ is used to link to the
+  annotation information in $MT$; 
+\item $CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
+  all the context instances where $ct_{id}$ is used to link to the
+  annotation information in $CT$. 
 \end{itemize}
@@
 considering all the distinct and context constratins. 
-Then, in Section \ref{sec:materialilzedb}, we describe our strategy to
+Then, in Section \ref{sec:query_materialilzedb}, we describe our strategy to
 perform queries over materialized database.
@@
 
-\subsubsection{Querying materialized database}\label{sec:materialilzedb}
+\subsubsection{Querying materialized database}\label{sec:query_materialilzedb}
+
+Once the data are materialized using the annotation constraints, we
+can perform the query over them. In what follows, we describe how to
+use the materialized data to answer the query.
+
+When the given query is a basic OM query, i.e., there is not context
+in the query. 
+Then, we can form a template query as the following;
+
+\begin{figure}[htb]
+{\tt
+\begin{tabular}{l}
+SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
+FROM $AD, MI~as~mi, MT~as~mt[$, {\em other tables}$]$\\
+$[$WHERE    {\em table join condition, \\
+$\qquad\qquad$ AND selection condition}$]$,\\
+$[$GROUP BY {\em distinct requirement} $]$\\
+$[$HAVING   {\em aggregation condition} $]$;
+\end{tabular}
+}
+\caption{SQL template to answer OM using materialized DB}
+\label{fig:sql_matdb_basic_omq}
+\end{figure}
+
+In particular,
+
+\vspace{0.1in}
+\begin{tabular}{l}
+$table~join~condition=$\\
+%$\qquad (mi.oi_{id}=ci.oi_{id})~AND~(cmi.oi_{id}=ci.coi_{id})~AND$\\
+$\qquad (mi.mt_{id}=mt.mt_{id})~AND~(mt.A_{id}=AD.A_{id})$
+\end{tabular}
+\vspace{0.1in}
+
+For the very basic query (e.g., $Q_1$ in Example \ref{eg:query}), the
+SQL generally need only the first three clauses. 
+If the query has some restriction on the aggregated result, the {\tt
+  GROUP BY} and {\tt HAVING} clauses to deal with it.
+
+We use $Q_1$ and $Q_3$ to illustrate the query on the materialized
+database. 
+can be answered using the above
+template. 
+Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
+the query rewriting method.
 
 \begin{example} \label{eg:materialize_db_q1}
-Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
-the query rewriting method. 
-Using the  materialized database, the first step is the same.
-After we know the data file, then we search the OBOE.MI with the
-condition that {\em mVal=``Picea rubens''}.
+Using the  materialized database,  we just need to search the MI table
+with the condition that {\em mVal=``Picea rubens''}.
+
+\vspace{0.1in}
+{\tt 
+\begin{tabular}{l}
+SELECT DISTINCT $AD.d_{id}$\\
+FROM $AD, MI~as~mi, MT~as~mt$\\ 
+WHERE $mt.Cha='Species'$ \\
+$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
+$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
+$\qquad\quad$ AND $mi.A_{id}=AD.A_{id}$; 
+%\#Join the MI and MT tables
+\end{tabular}
+}
+\vspace{0.1in}
+
+For $Q_3$, which asks for data sets that contain at
+least five distinct ``Picea rubens'' observations.
+The SQL query can be written as:
+
+\vspace{0.1in}
+{\tt 
+\begin{tabular}{l}
+SELECT distinct $AD.d_{id}$\\
+FROM $AD, MI~as~mi, MT~as~mt$\\
+WHERE $mt.Cha='Species'$ \\
+$\qquad\quad$ AND $mi.mVal = `Picea rubens'$\\
+$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
+$\qquad\quad$ AND $mt.A_{id}=AD.A_{id}$\\
+GROUP BY $d_{id},oi_{id}$\\
+HAVING COUNT(*)$>$5; 
+}
+\end{tabular}
+}
+\vspace{0.1in}
 \end{example}
 
+For this kind of query, the computation cost is the selection cost of the
+measurement instance (MI) table. 
+Compared with the query rewriting strategy, which need to pose the
+query over multiple data tables, this query should answer the queries
+more efficiently.
+
+Figure \ref{fig:sql_matdb_basic_omq} shows the SQL template to process
+the basic OM query without context. In case the OM query has context,
+we need to add the context into consideration.
+
+\begin{enumerate}
+\item The first step is similar to that in the basic OM query
+  processing. Besides this, we need to figure out what are the concept
+  obserbationi types of the needed observations.  
+\item \label{step:singlesql} Form an SQL for the observation and every of its context
+  observations. 
+\item Intersect all the SQLs from Step \ref{step:singlesql}). 
+\end{enumerate}
+
+\begin{figure}[htb]
 {\tt
-SELECT distinct $AD.d_{id}$\\
-FROM MI, MT, AD\\ 
-WHERE $MT.Cha='Species'$ \\
-AND $MI.mVal = `Picea rubens'$\\
-AND $MI.mt_{id}=MT.mt_{id}$ AND $MT.A_{id}=AD.A_{id}$; %\#Join the MI and MT tables
+\begin{tabular}{l}
+SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
+FROM $AD, CI~as~ci,MI~as~mi, MT~as~mt$\\
+$\qquad MI~as~cmi, MT~as~cmt[,$ {\em other tables}$]$\\
+$[$WHERE    {\em table join condition, \\
+$\qquad\qquad$ AND selection condition}$]$,\\
+$[$GROUP BY {\em distinct requirement} $]$\\
+$[$HAVING   {\em aggregation condition} $]$;
+\end{tabular}
 }
+\caption{SQL template to answer contextualized OM using materialized DB}
+\label{fig:sql_matdb_context_omq}
+\end{figure}
 
-In this example, the computation cost is the selection cost of the
-measurement instance table. 
-It should be faster than the query rewriting.
+In this case, the table join condition is much more complex since we
+need to use the context instance table $ci$ to connect the
+observation and context observation table. 
+In particular,
 
-If the query has some restriction on the aggregate result, we can
-easily add the GROUP BY and HAVING clause to deal with it. 
-We take $Q_3$ as an example, which asks for data sets that contain at
-least five distinct ``Picea rubens'' observations.
-The SQL query can be written as: \\
+\vspace{0.1in}
 {\tt 
-SELECT distinct $d_{id}$\\
-FROM MI, MT\\ 
-WHERE $MT.Cha='Species'$ \\
-AND $MI.mVal = `Picea rubens'$\\
-AND $MI.mt_{id}=MT.mt_{id}$\\
-GROUP BY $d_{id},oi_{id}$\\
-HAVING COUNT(*)$>$5; %\#Join the MI and MT tables
+\begin{tabular}{l}
+$table~join~condition=$\\
+$\qquad (mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
+$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
+$\qquad (mt.A_{id}=AD.A_{id})$
+\end{tabular}
 }
+\vspace{0.1in}
 
-Using the query rewriting, 
-First, we find the data tables that contain the needed concepts. 
-In each table, we need to figure out the distinct observations. 
-From $m_i$ $obstype_i$, all the other measurements that are of
-$obstype_i$. 
-Group by the key measurements of $obstype_i$ if distinct yes.
-If this observation type does not have any context, then the key
-measurements are all the measurements directly defined under this
-observatioin type. Otherwise, i.e., this observation type has context,
-then the key measurements are all the measurements in the context
-chain. 
-%\end{example}
+{\bf Execution cost analysis:} From the join condition, we can see that the expensive cost comes from
+the join operation between the measurement instance, context
+measurement instance, and context instance table. 
+As we analyzed in Section \ref{sec:materialilzedb}, the two tables with
+growing space are measurement instance table and context instance
+table.
 
-When we are given contextualized query, e.g., $Q_2$. 
-Give me the datasets that contain species ``Picea rubens'' observations in ``California''.
+{\bf Example:} When we are given contextualized query, e.g., $Q_2$, which asks for
+the datasets that contain species ``Picea rubens'' observations in
+``California''.
 
-Species = ``Picea rubens'' and observation has context {\em in}
-another observation which has measurement of value ``California''.
+In the first step, we get the requirement is on Characteristic
+``Species'' with value ($mVal$) ``Picea rubens'' and observation has context {\em in}
+another observation for ``State'' of value ``California''.
 
@@
 %where chracteristic = 'Species';  \\
+In the first step, we need to find the context type id $tmp\_ct\_id$
+that satisfy the context relationship.
 
+\vspace{0.1in}
 {\tt 
-SELECT record\_id\\
-FROM CI as ci, MI as mi, MI as cmi,MT as mt, MT as cmt, \\
-WHERE $mi.oi_{id}=ci.oi_{id}$ AND $cmi.oi_{id}=ci.coi_{id}$ \\
-AND $ci.coi_{id} = coi.oi_{id}$ AND oi.otype = 'olabel' \\
-AND mi.mvalue=`Picea rubens' AND cmi.mvalue='California'\\
-AND MT.mt_{id}=mi.mt_{id} AND MT.mLabel = 'Species'\\
-AND CMT.mt_{id}=cmi.mt_{id} AND cmt.mLabel=`State'\\
-AND coi.otype IN (\\
-\verb|    |SELECT context\_olabel\\
-\verb|    |FROM context\_relationship\\
-\verb|    |WHERE olabel = 'olabel');
+\begin{tabular}{l}
+SELECT DISTINCT AD.d_{id}\\
+FROM $AD, CI~as~ci, MI~as~mi, MT~as~mt,$ \\
+$\quad\qquad MI~as~cmi, MT~as~cmt,$ \\
+WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
+$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
+$\qquad (mt.A_{id}=AD.A_{id})$\\
+$\qquad$ AND $mi.mVal=`Picea rubens'$ \\
+$\qquad$ AND $cmi.mVal='California'$\\
+$\qquad$ AND $mt.Cha = 'Species'$\\
+$\qquad$ AND $cmt.Cha=`State'$\\
+$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
+\end{tabular}
 }
-The sub-query finds the set containing all the observation type
-labels in the context chain.
+\vspace{0.1in}
 
-Use query rewriting, what I can do? 
-It's the same till to getting context\_olabel\_set.
+%Use query rewriting, what I can do? 
+%It's the same till to getting context\_olabel\_set.
 
-Get the asked condition columns and the context columns,
+%Get the asked condition columns and the context columns,
 
-select record\_id from data\_tables
-where condition column 1='value1' and (context column1 = 'context
-value' or context column2 = `context value', etc.);
+%select record\_id from data\_tables
+%where condition column 1='value1' and (context column1 = 'context
+%value' or context column2 = `context value', etc.);
 
-\begin{example}
-SQ2: Give me the datasets that have measurements with average ``area''
-bigger than 5.0 square feet. 
-\end{example}
+\from{HP}{Need to refinement the following. }
 
+=============
+
+Now let's think about the query $Q_4$ with aggregation and context
+requirement.
+
+Also, in the first step, we discover the context type id $tmp\_ct\_id$
+that satisfy the context relationship. 
 Use materialized database: 
@@
 
-SELECT oi.record\_id \\
-FROM observation\_instance oi, measurement\_instance mi, 
-observation\_instance coi, measurement\_instance cmi,context\_instance ci\\
-WHERE (mi.oid=ci.oid AND ci.oid = oi.oid) and (cmi.oid=coi.oid AND ci.context\_oid =
-coi.oid) AND coi.otype $\in$ olabel\_set 
-AND cmi.mlabel=`Area'
-GROUP BY cmi.oid
-HAVING avg(cmi.mvalue)>5.0in;
+\vspace{0.1in}
+{\tt 
+\begin{tabular}{l}
+SELECT DISTINCT AD.d_{id} \\
+FROM $AD, CI~as~ci, MI~as~mi, MI~as~cmi,$ \\
+$\quad\qquad MT~as~mt, MT~as~cmt,$ \\
+WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
+$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
+$\qquad (mt.A_{id}=AD.A_{id})$\\
+$\qquad$ AND $cmi.Cha=`State'$\\
+$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
+GROUP BY $cmi.oid$\\
+HAVING $avg(cmi.mVal)>5.0$;
+\end{tabular}
+}
+\vspace{0.1in}
+
+\subsection{Query partially materialized database}
+As we analyzed in the first two sub-sections, 
+the query-rewriting suffers the multiple queries posed on different
+candidate dataset. The querying of materialized database suffers from
+the multiple self-join operation of the measurement instance (MI)
+table, which has the most amount of information.
+
+In this section, we propose to materialize only the measurement
+instance and also put the measurement instance and context measurement
+instances together in the same table to avoid the expensive
+computation cost.
+
+\from{HP}{Let's see how this can work. }
 

### 2010-05-07 docs/algdraft/DataIntegration/DataIntegration.tex
Refine texts (little change). No adjustment on framework structure.

@@
 
-\newtheorem{example}{Example}[section]
-\newtheorem{definition}{Definition}[section]
+\newtheorem{example}{\bf Example}[section]
+\newtheorem{definition}{\bf Definition}[section]
 \newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
@@
 A basic observation and measurement query is defined as 
-\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]. 
+\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]
 Here, 
@@
 
-{\bf TODO: put a figure here to illustrate the steps}.
+\from{HP}{todo:put a figure here to illustrate the steps}.
 
@@
 
-{\bf Execution cost analysis:} 
+{\bf Cost analysis:} 
 From the above description, we can tell that the computation cost is to search the different data
@@
 $OBOE.*$. 
-In partibular, we have the following concept instance tables. 
+In particular, we have the following concept instance tables. 
 \begin{itemize}
@@
 \subsubsection{Materialize data}\label{sec:materialilzedb}
-{\bf: TODO: refine the writing in MaterializeDB and move them here.}
+\from{HP}{TODO: refine the writing in MaterializeDB and move them here.}
 
@@
 use the materialized data to answer the query.
+
+\from{HP}{TODO:Formalize the translation of operators in the OM
+  query}.
 
@@
 
+{\tt 
+\begin{tabular}{l}
+SELECT $ct.ct_{id} as tmp\_ct_\id$\\
+FROM OT as ot, OT as cot, CT as ct\\
+WHERE $ct.ot_{id} = ot.ot_{id}$ AND \\
+      $\qquad ct.cot_{id}= cot.ot_{id}$ AND \\
+      $\qquad ct.Rel='IN'$ AND $ot.et='tree'$;
+\end{tabular}
+}
+
+Then, the next step is to find the distinct data sets.
+
 \vspace{0.1in}
@@
 
-\from{HP}{Need to refinement the following. }
-
-=============
-
 Now let's think about the query $Q_4$ with aggregation and context
 requirement. 
+% Give me the data sets that have trees with average ``height''
+%  than 20.0 in ``California''.
+$Q_4 = Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
+$\wedge IN(Observation_1, Observation_2)$\\
+$\wedge Observation_2: \langle State='California'\rangle$
 
@@
 that satisfy the context relationship.
-Use materialized database: 
-Analysis: ``Area'' is the context of some observation.
-Area --> mlabel, observation type
-
-From the context observation lable, find the observation labels, \\
-select olabel\\
-from context\_relationship\\
-where context\_olabel = 'olabel';
-
-Let olabel\_set  be the set containing all the observation type
-labels that use ``Area'' as context.
 
@@
 $\qquad (mt.A_{id}=AD.A_{id})$\\
-$\qquad$ AND $cmi.Cha=`State'$\\
+$\qquad$ AND $mt.Cha=`height'$\\
+$\qquad$ AND $cmt.Cha=`State'$\\
+$\qquad$ AND $cmi.mVal=`California'$\\
 $\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
-GROUP BY $cmi.oid$\\
-HAVING $avg(cmi.mVal)>5.0$;
+GROUP BY $AD.d_{id}, oi_{id}$\\
+HAVING $avg(mi.mVal)>5.0$;
 \end{tabular}
@@
 \vspace{0.1in}
+
+Note that, in the above example, there is only one context.
+In the real application, there may be multiple context observations,
+or even context chains. In this case, we need to get {\em all} the
+context and get the intersection of the results.
 

### 2010-05-10 docs/algdraft/DataIntegration/DataIntegration.tex
(1) Refine description of query rewriting (Sec IV.A) (except the examples) (2) Change some inconsistent symboles in Section III (model) and Sec IV.B (materialization) (3) Refine the first 4 paragraphs of Section IV.B.

@@
 
-For data table $d$ with annotation $A$, the system keeps the
-information $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$.
-As we described before, we have four main objects to describe: 
+For a data table $d$ with annotation $A$, the system keeps the
+corresponding information between them in $AD$ tables. 
+Besides it, we also have four main concepts to describe: 
 observation type (OT), measurement type (MT), and
@@
 \begin{itemize}
+\item $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$ to keep the information
+  of data sets $d_{id}$-s and their related annotations
+  $A_{id}$. Here, $d_{meta}$ is an abstract attribute for all the
+  other metadata of $d_{id}$. 
 \item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
-  describe an observation type, which denotes on which
+  describe an observation type. It denotes on which
   entity type (object in the real world) the observation is made. 
@@
   isDistinct} is used to denote whether the same value of key
-meaasurements of an obbservation types can uniquely identify one
+measurements of an obbservation types can uniquely identify one
 observation or not. 
@@
   type. 
-\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, MapCond},
-  Val\}$
+\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, mapCond, mapVal},
+  Val\}$. This structure denotes the correspondences between an
+  measurement type ($mt_{id}$ and an attribute $attr_i$ in an dataset
+  $d_{id}$ for a given condition $mapCond$. 
 \end{itemize}
@@
 \begin{itemize}
-\item $Observation: \langle Species='Picea~rubens'\rangle$
-\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
+\item $Q_1$ = $Observation: \langle Species='Picea~rubens'\rangle$
+\item $Q_2$ = $Observation_1: \langle Species='Picea~rubens'\rangle$\\
   $\wedge IN(Observation_1, Observation_2$) \\
 $\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
-\item $Observation: \langle Species='Picea~rubens'$\\
+\item $Q_3$ = $Observation: \langle Species='Picea~rubens'$\\
 $\wedge count(distinct~Species)\geq 5\rangle$
-\item $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
+\item $Q_4$ = $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
 $\wedge IN(Observation_1, Observation_2)$\\
@@
 an OM query is represented using 
-OM concept terminologies such as observation
+OM concept terminologies such as observation, 
 measurement, characteristic, etc. 
-To answer such query from the original data model (i.e., dataset
-metadata and data tables), we need to {\em rewrite} the quey over the
-real data table. 
-In what follows, we first sketch the re-writing process. Then, we
-detailed the procedure in using the data model and
-structures. Finally, we include the process for the more complicated
+To answer such queries from the original data model (i.e., dataset 
+metadata and translated data tables), 
+we need to {\em rewrite} the OM query over the real data tables. 
+In what follows, we first sketch the re-writing process. 
+Then, we detail the procedure in using the data model and structures. 
+Finally, we include the process for the more complicated
 cases (e.g., with distinct, aggregate).
 
+%Framework of query rewriting
 Roughly speaking, the query rewriting consists of two steps:
 \begin{itemize}
-\item From the given query, find the relevant data tables and
-  attributes that need to be used answer the given query. 
-\item Translate the given query to queries over the relevant data
+\item From the given query, search the database metadata and
+  annotation information to find the relevant data tables and
+  attributes that need to be used to answer the given query. 
+\item Translate the given OM query to queries over the relevant data
   tables. 
@@
 
-The first step is to map the OM query to the real data structure. 
-We can utilize the annotation structure $A.Map$ which keeps the correspondences between
-measurement type and table attributes. 
-So, when the given concept is {\em measurement}, we can directly get
+%Elaborate the first step
+The first step is to map the OM concepts in the query to the real data
+structure. 
+With the annotation structures $OT$, $CT$, and $MT$, any given OM
+concepts can be linked to a measurement type $mt_{id}$. 
+Then, the annotation structure $Map$ can be used to bridge OM concepts and data sets. 
+Utilizing $Map$, we can find the attributes $attr_i$ and the
+annotation $A_{id}$).
+E.g., when the given concept is {\em measurement}, we can directly get
 its related attribute names and annotation id $A_{id}$. 
 In case that the given OM concept is a non-measurement concept, we can use
-other Annotation structures $A.*$ to figure out the measurementt types. 
-E.g., if the OM concept is Characteristic ``Species'', we can use
-$A.MeasType$ to get the measurement type id.
+other annotation structures to figure out the measurementt types. 
+E.g., if the OM concept is {\em characteristic} ``Species'', we can use
+$MT$ to get the measurement type id $mt_{id}$. 
+After we get $A_{id}$, we can quickly find the dataset id $d_{id}$
+using $AD$ metadata table (Section \ref{sec:annotation}). 
+Then, we get the relevant data tables and the needed attributes.
 
-As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
-information $\langle A_{id}, DTableId\rangle$. 
-With the $A_{id}$, we can ge the data table information from the
-metadata  $\langle A_{id}, DTableId\rangle$. 
-Then, we get the relevant data table information and the needed attribute
-information.
+%Elaborate the second step
+Let $D_{cand}$ be the set of relevant (or candidate) data tables and $d_{a_1} \cdots d_{a_m}$
+are the related attributes for table $d\in D_{cand}$. 
+The second step translates the original OM query to SQL queries over the data tables.
 
-Let $RD$ be the set of relevant data tables and $d_{a_1} \cdots d_{a_m}$
-are the related attributes for table $d\in RD$. We can translate the original OM query to SQL queries
-over the data tables.
+In particular, for each candidate data table $d$,  
+the translation process works in the following rules.
+
+\begin{itemize}
+\item The SELECT clause always get the distinct $d_{id}$ and/or
+  record id $r_{id}$ (when it is needed);
+\item The FROM clause has table $d \in D_{cand}$;
+\item The WHERE clause contains all the function $f(\cdot)$ in the OM
+  query;
+\item When there is distinct constraint on entity or observation, the
+  GROUP BY clause contains of all the attributes corresponding to the
+  {\em key measurement types} of the entity or observation. we detail
+  all process of getting the key measurement types of an observation
+  later. 
+\item The aggregation requirement in $f(\cdot)$ is translated to
+  HAVING clause. 
+\end{itemize}
+
+Based on these rules, an OM query can be translated to the following
+SQL scripts.
 
@@
 
-The above describes the process for the most basic OM query. 
-However, in the real application, people ask queries with distinct
+The above describes the process for the OM query. 
+In real application, people ask queries with distinct
 observation or entity constraints. Also, people ask queries about
 different observatitons using context. 
+The following elaborate the details in processing these two cases.
 
-%Discuss how to deal with the distinct constraints.
-Many times, a user may want to find {\em distinct} observations or
-entities. In this case, the above simple solution cannot work
-correctly because it treats observations in different rows as the same
-observation. So, we need to use the annotation to figure out what
-distinguishes one observation from another. 
+%In detail,  how to deal with the distinct constraints.
+When a query with distinct constraints, e.g., find data sets with {\em distinct} observations or
+entities. To deal with this case, 
+we need to be able to figure out what distinguishes one observation
+from another. Here, one observation type's key measurement types can
+play this role well. 
 When an observation type is denoted with {\em distinct yes} and some
@@
 
-%Discuss how to deal with context
-The above description discusses cases without context. 
-However, in real application, the observation has context. 
-When the context is not {\em identifying}, i.e., the context value
+%In detail, how to deal with context
+When the given OM query has context, we need to consider two cases. 
+In the first case the context is not {\em identifying} to its
+observation types, i.e., the context value
 does not affect the uniqueness of one observation, this can be
@@
 these key measurement types. 
-
-{\bf TODO: put an example here.}
-
+\from{HP}{Put an example here.}
 In this case, there is no need to change the SQL scripts.
@@
 
+\from{HP}{Put a specific routine to get key measurements of
+  observation and/or entity.}
+
 \begin{example}\label{eg:query_reqerite_q1}
 {\bf TODO: refine this example} 
-Take $Q_1$ as example. The {\tt concept} is ``observation'' and {\tt cond} is {\em species = ``Picea rubens''};
+Take $Q_1$ ($Observation: \langle Species='Picea~rubens'\rangle$) as
+example. 
+The {\tt OMconcept} is ``observation'' and {\tt cond} is ``{\em species = `Picea rubens'}'';
 
-From the annotation $A.MeasType$, we can find the measurement type
+From the annotation $MT$, we can find the measurement type
 $m_3$ with the characteristic {\em species}. 
-From the mapping $A.Map$, we further find that the attribute {\em spp} that the
+From the mapping $Map$, we further find that the attribute {\em spp} that the
 measurement type $m_3$ is mapped to.
 
-The second step is to find the data file which really have this value. 
+The second step is to find the data sets which really have this value. 
 For this one, we find the data table, and do a selection on the table
@@
 
-{\bf TODO: SQL here.} 
+\vspace{0.1in}
+{\tt 
+\begin{tabular}{l} 
+SELECT DISTINCT $d_{id}$\\
+FROM $d$\\
+WHERE $spp = 'piru'$
+\end{tabular}
+}
+\vspace{0.1in}
+
 \end{example}
 
-\from{HP}{Refine the process of $Q_2$.\\
+\begin{example}
+Take $Q_3$ ($Observation:$\\
+$\langle Species='Picea~rubens'$$\wedge count(distinct~Species)\geq
+5\rangle$) as an example.
+
+\from{HP}{Refine\\
 For the first step, we find the data tables that contain the needed concepts. 
@@
 
+\vspace{0.1in}
+{\tt 
+\begin{tabular}{l} 
+SELECT DISTINCT $d_{id}$\\
+FROM $d$\\
+WHERE $spp = 'piru'$\\
+GROUP BY $d_{id}$, $spp$\\
+HAVING count($d_{id}$,$spp$) $\geq 5$\\
+\end{tabular}
+}
+\vspace{0.1in}
+
+\end{example}
+
 %Analysis of this method. 
-
 {\bf Cost analysis:} 
@@
 The previous section shows the method to rewrite a given query. 
-However, as we analyzed, the computation cost may be higher when the
-candidate data tables are many. In this section, we propose another
-strategy to make use of the existing optimization strategies of
+However, as we analyzed, the computation cost may be very high when
+there are many candidate data tables. 
+In this section, we propose another strategy to answer the given OM
+query by making use of the existing optimization strategies of
 current DBMS.
 
-The basic idea is to materialize the data into some centralized
-concept tables with instances of entities, observations, measurements
-and the contexts between observations. 
-We call such concept tables core OBOE tables, and denote them as
+%rough idea of this strategy
+In this strategy, we do not directly work on the original data table
+$D$. On the other hand, we {\em materialize} the original data-sets to
+centralized concept tables with instances of entities, observations, measurements
+and contexts. Such concept tables are called core OBOE tables, and denoted as
 $OBOE.*$. 
+Then, the query process works on the materialized databases (i.e., the
+conceptual data tables).
+
+%structure of this sub-section
+In what follows, we first describe the structure of the core
+conceptual tables and present our algorithm to materialize the data into such
+concept instance tables in Section \ref{sec:materialilzedb}. Our
+algorithm considers all the distinct and context constraints. 
+Then, Section \ref{sec:query_materialilzedb} provides the detailed
+process of performing queries over materialized database.
+
+\subsubsection{Materialize data}\label{sec:materialilzedb}
+
+%Structure of conceptual instances
 In particular, we have the following concept instance tables. 
@@
 
-In what follows, we describe how to materialize the data into such
-concept instance tables in Section \ref{sec:materialilzedb} when
-considering all the distinct and context constratins. 
-Then, in Section \ref{sec:query_materialilzedb}, we describe our strategy to
-perform queries over materialized database.
-
-\subsubsection{Materialize data}\label{sec:materialilzedb}
 \from{HP}{TODO: refine the writing in MaterializeDB and move them here.}
@@
 Once the data are materialized using the annotation constraints, we
-can perform the query over them. In what follows, we describe how to
-use the materialized data to answer the query.
+can perform the OM query over them. 
+%In what follows, we describe how to use the materialized data to answer the query.
 
@@
 WHERE $mt.Cha='Species'$ \\
-$\qquad\quad$ AND $mi.mVal = `Picea rubens'$\\
+$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
 $\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
change Edit/Save toggle button text depending on what action will be performed

@@ -146,6 +146,7 @@ public class AnnotationPage extends AbstractUIPage {
 				if (editButton.isSelected()) {
 					// active only if we have an observation
 					pageRef.setEnabled(currentAttributeName != null);
+					editButton.setText("Save");
 				}
 				else {
 					pageRef.setEnabled(false);
@@ -156,11 +157,12 @@ public class AnnotationPage extends AbstractUIPage {
 						// save
 						AnnotationPlugin.saveAnnotation(annotation);
 					}
+					editButton.setText("Edit");
 				}
 			}
 		};
-		//editButton = new JToggleButton(toggleAction);
-		editButton = WidgetFactory.makeCheckBox("Edit", false);
+		editButton = new JToggleButton(toggleAction);
+		//editButton = WidgetFactory.makeCheckBox("Edit", false);
 		editButton.addActionListener(toggleAction);
 		
 		// actually show the help

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
change Edit/Save toggle button text depending on state of the panel

@@ -114,6 +114,11 @@ public class AnnotationPage extends AbstractUIPage {
 		super.setEnabled(enabled);
 		simpleAnnotationPanel.setEnabled(enabled);
 		editButton.setSelected(enabled);
+		if (enabled) {
+			editButton.setText("Save");
+		} else {
+			editButton.setText("Edit");
+		}
 
 	}
 	

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ContextPage.java
change Edit/Save toggle button text depending on what action will be performed

@@ -100,6 +100,7 @@ public class ContextPage extends AbstractUIPage {
 				if (editButton.isSelected()) {
 					// active only if we have an observation
 					pageRef.setEnabled(observation != null);
+					editButton.setText("Save");
 				}
 				else {
 					pageRef.setEnabled(false);
@@ -110,11 +111,12 @@ public class ContextPage extends AbstractUIPage {
 						// save
 						AnnotationPlugin.saveAnnotation(annotation);
 					}
+					editButton.setText("Edit");
 				}
 			}
 		};
-		//editButton = new JToggleButton(toggleAction);
-		editButton = WidgetFactory.makeCheckBox("Edit", false);
+		editButton = new JToggleButton(toggleAction);
+		//editButton = WidgetFactory.makeCheckBox("Edit", false);
 		editButton.addActionListener(toggleAction);
 		
 		// help button
@@ -145,6 +147,11 @@ public class ContextPage extends AbstractUIPage {
 		super.setEnabled(enabled);
 		contextList.setEnabled(enabled);
 		editButton.setSelected(enabled);
+		if (enabled) {
+			editButton.setText("Save");
+		} else {
+			editButton.setText("Edit");
+		}
 	}
 	
 	public void setObservation(Observation observation) {

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/renderer/OntologyClassSelectionPanel.java
"Search" button at the bottom of the panel so that the "Select" button is near it

@@ -212,10 +212,6 @@ public class OntologyClassSelectionPanel extends JPanel {
 		treePanel.add(panel1);
 		treePanel.add(Box.createRigidArea(new Dimension(0, 2)));
 		treePanel.add(treeView);
-		if (showSearch) {
-			treePanel.add(Box.createRigidArea(new Dimension(0, 5)));
-			treePanel.add(searchPanel);
-		}
 
 		// top portion
 		JPanel panel7 = new JPanel();
@@ -236,6 +232,10 @@ public class OntologyClassSelectionPanel extends JPanel {
 		panel.add(panel8);
 		panel.add(Box.createRigidArea(new Dimension(0, 2)));
 		panel.add(commentView);
+		if (showSearch) {
+			panel.add(Box.createRigidArea(new Dimension(0, 5)));
+			panel.add(searchPanel);
+		}
 
 		//panel.setMaximumSize(new Dimension(width, height));
 		//panel.setPreferredSize(new Dimension(width, height));

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/plugins/ui/OntologyClassField.java
make the madlib appear as one line of fields with the "sentence" above the fields

@@ -118,7 +118,7 @@ public class OntologyClassField extends JTextField {
 			// set the label border if we have it wrapped
 			Container parent = this.getParent();
 			if (parent instanceof OntologyClassFieldPanel) {
-				setPanelBorder((OntologyClassFieldPanel) parent, filterClass.getName());
+				setPanelBorder((OntologyClassFieldPanel) parent, filterClass.getName(), TitledBorder.ABOVE_BOTTOM);
 			}
 		}
 	}
@@ -135,14 +135,22 @@ public class OntologyClassField extends JTextField {
 		panel.setLayout(new BoxLayout(panel, BoxLayout.Y_AXIS));
 		panel.add(field);
 		if (field.getFilterClass() != null) {
-			setPanelBorder(panel, field.getFilterClass().getName());
+			setPanelBorder(panel, field.getFilterClass().getName(), TitledBorder.ABOVE_BOTTOM);
 		}
-		
 		return panel;
-		
 	}
 	
-	public static void setPanelBorder(OntologyClassFieldPanel panel, String title) {
+	public static JPanel wrapField(OntologyClassField field, String label) {
+		OntologyClassFieldPanel panel = new OntologyClassFieldPanel();
+		panel.setLayout(new BoxLayout(panel, BoxLayout.Y_AXIS));
+		panel.add(field);
+		if (field.getFilterClass() != null) {
+			setPanelBorder(panel, label, TitledBorder.ABOVE_TOP);
+		}
+		return panel;	
+	}
+	
+	public static void setPanelBorder(OntologyClassFieldPanel panel, String title, int location) {
 		Font titleFont = new Font("Sans-Serif", Font.PLAIN, 8);
 		Border existingBorder = BorderFactory.createEmptyBorder();
 		panel.setBorder(
@@ -150,7 +158,7 @@ public class OntologyClassField extends JTextField {
 						existingBorder, 
 						title, 
 						TitledBorder.CENTER, 
-						TitledBorder.ABOVE_BOTTOM, 
+						location, 
 						titleFont, 
 						WizardSettings.WIZARD_CONTENT_TEXT_COLOR));
 	}

### 2010-05-10 dev/sms/src/org/ecoinformatics/sms/plugins/ui/SimpleAnnotationPanel.java

@@ -218,46 +218,23 @@ public class SimpleAnnotationPanel extends JPanel {
 		
 		// Characteristic and Entity
 		JPanel characteristicPanel = WidgetFactory.makePanel(2);
-		
-		characteristicPanel.add(WidgetFactory.makeLabel("The ", false, null));
-		
+				
 		observationCharacteristic = OntologyClassField.makeLabel("", true, null);
 		observationCharacteristic.setFilterClass(Annotation.OBOE_CLASSES.get(Characteristic.class));
-		characteristicPanel.add(OntologyClassField.wrapField(observationCharacteristic));
-		
-		characteristicPanel.add(WidgetFactory.makeLabel(" of the ", false, null));
-		
+		characteristicPanel.add(OntologyClassField.wrapField(observationCharacteristic, "The Characteristic"));
+				
 		observationEntity = OntologyClassField.makeLabel("", true, null);
 		observationEntity.setFilterClass(Annotation.OBOE_CLASSES.get(Entity.class));
-		characteristicPanel.add(OntologyClassField.wrapField(observationEntity));
-		characteristicPanel.add(WidgetFactory.makeLabel(" was recorded ", false, null));
-		
-//		characteristicPanel.setBorder(new javax.swing.border.EmptyBorder(0, 0,
-//				0, 8 * WizardSettings.PADDING));
-		
-	
-		// Standard and Protocol
-		JPanel standardPanel = WidgetFactory.makePanel(2);
-				
-		standardPanel.add(WidgetFactory.makeLabel(" using the ", false, null));
-
+		characteristicPanel.add(OntologyClassField.wrapField(observationEntity, "of the Entity was recorded"));
+			
+		// Standard and Protocol	
 		observationStandard = OntologyClassField.makeLabel("", true, null);
 		observationStandard.setFilterClass(Annotation.OBOE_CLASSES.get(Standard.class));
-
-		standardPanel.add(OntologyClassField.wrapField(observationStandard));
-		standardPanel.add(WidgetFactory.makeLabel(" standard, ", false, null));
-
-		standardPanel.add(WidgetFactory.makeLabel("and the ", false, null));
-
+		characteristicPanel.add(OntologyClassField.wrapField(observationStandard, "using the MeasurementStandard"));
+		
 		observationProtocol = OntologyClassField.makeLabel("", true, null);
[…]

### 2010-05-11 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
add "same as observation x" drop down. need to hook it up so that it is used when saving. then try to do it as a drop down for attributes rather than existing observations.

@@ -30,12 +30,14 @@ package org.ecoinformatics.sms.plugins.pages;
 
 import java.awt.GridLayout;
 import java.awt.event.ActionEvent;
+import java.util.List;
 
 import javax.swing.AbstractAction;
 import javax.swing.Box;
 import javax.swing.BoxLayout;
 import javax.swing.JButton;
 import javax.swing.JCheckBox;
+import javax.swing.JComboBox;
 import javax.swing.JLabel;
 import javax.swing.JPanel;
 import javax.swing.JTextField;
@@ -87,6 +89,7 @@ public class AnnotationPage extends AbstractUIPage {
 	private JTextField observationLabel;
 	private JLabel observationLabelLabel;
 	private JCheckBox observationIsDistinct;
+	private JComboBox existingObservationList;
 	
 	// measurement
 	private JTextField measurementLabel;
@@ -113,6 +116,7 @@ public class AnnotationPage extends AbstractUIPage {
 	public void setEnabled(boolean enabled) {
 		super.setEnabled(enabled);
 		simpleAnnotationPanel.setEnabled(enabled);
+		existingObservationList.setEnabled(enabled);
 		editButton.setSelected(enabled);
 		if (enabled) {
 			editButton.setText("Save");
@@ -188,6 +192,13 @@ public class AnnotationPage extends AbstractUIPage {
 		//add the main panel here
 		simpleAnnotationPanel = new SimpleAnnotationPanel(true, showAll);
 		this.add(simpleAnnotationPanel);
+		
+		// same as other observation?
+		JPanel sameObservationPanel = WidgetFactory.makePanel(2);
+		sameObservationPanel.add(WidgetFactory.makeLabel("Same entity sample or individual as:", false, null));
+		existingObservationList = WidgetFactory.makePickList(null, false, 0, null);
+		sameObservationPanel.add(existingObservationList);
+		this.add(sameObservationPanel);
 		
 		// Measurement Label
 		JPanel measurementPanel = WidgetFactory.makePanel(2);
@@ -301,6 +312,16 @@ public class AnnotationPage extends AbstractUIPage {
 			try {
 				currentProtocol = currentMeasurement.getProtocol();
 				this.simpleAnnotationPanel.setObservationProtocol(currentProtocol);
+			} catch (Exception e) {
+			}
+			try {
+				List<Observation> existingObservations = annotation.getObservations();
+				existingObservationList.removeAllItems();
+				existingObservationList.addItem(null);
[…]

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/MaterializeDB.java
Add the first batch of codes querying materialized database.

@@ -127,9 +127,10 @@ public class MaterializeDB {
 	 * @param annotFileName
 	 * @return
 	 */
-	private static Annotation readAnnotation(String annotFileName)
+	private static Annotation readRemoteAnnotation(String annotFileName)
 	{
 	
+		annotFileName = "https://code.ecoinformatics.org/code/semtools/trunk/dev/sms/examples/er-2008-ex2-annot.xml";
 		String annot1 = annotFileName;
 		URL url = null;
 	    URLConnection connection = null;
@@ -167,7 +168,6 @@ public class MaterializeDB {
 	    
 	    for(Annotation a: annotations){
 	    	List<Mapping> mappingList = a.getMappings();
-	    	//System.out.println("a : EML package = " +a.getEMLPackage()+", Data Table = " + a.getDataTable());
 	    	System.out.println("a : mappingList size = " +mappingList.size());
 	    	for(Mapping m: mappingList){
 	    		System.out.println(m.getAttribute()+", " + m.getMeasurement().getLabel() +", "+m.getValue() +", "+m.getConditions());
@@ -656,7 +656,10 @@ public class MaterializeDB {
 		System.out.println(Debugger.getCallerPosition() +"2. Read annotation ...");
 		Annotation A = null;
 		if(annotFileName.endsWith(".xml")){ //this is an annotation file
-			A = readAnnotation(annotFileName);
+			//A = readRemoteAnnotation(annotFileName);
+			InputStream is = new FileInputStream(annotFileName);
+			A = Annotation.read(is);
+			//A = readAnnotation(annotFileName);
 		}else{
 			//this is an annotation specification file
 			AnnotationSpecifier a = new AnnotationSpecifier();
@@ -731,49 +734,60 @@ public class MaterializeDB {
 	 * @author cao
 	 * @param args
 	 * 
-	 * 
-	E.g., to set parameters in the Arguments field
-
-	er-2008-ex3-eml.xml //now this file is not used
-	er-2008-ex3-data.txt
-	er-2008-ex3-annot.xml
-	er-2008-ex3-oboe.csv
-
 	 */
 	public static void main(String[] args) {
 		
-		if(args.length!=2&&args.length!=3){
-			System.out.println("Usage: ./MaterializeDB <1. file prefix name> <2. row num> [<3. bool: materialize context chain>]");
[…]

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/QueryProcessor.java

@@ -1,0 +1,53 @@
+package org.ecoinformatics.oboe;
+
+import java.io.IOException;
+import java.util.ArrayList;
+
+import org.ecoinformatics.oboe.query.Query;
+import org.ecoinformatics.oboe.query.QueryList;
+import org.ecoinformatics.oboe.query.OboeQueryResult;
+import org.ecoinformatics.oboe.query.OboeQueryResultContainer;
+
+public class QueryProcessor {
+
+	public static void main(String[] args) throws Exception {
+		if(args.length<2||args.length>3){
+			System.out.println("Usage: ./QueryProcessor <1. query file> <2. query strategy> [3. query result file name]");
+			System.out.println("\t <2. Query stragety> 1-query rewriting; 2-query materialized db; 3-others; ");
+			return;
+		}
+		
+		//1. process parameters
+		String queryFile =args[0];
+		int queryStrategy = Integer.parseInt(args[1]);
+		if(queryStrategy<=0||queryStrategy>3){
+			System.out.println("<2. Query stragety> 1-query rewriting; 2-query materialized db; 3-others; ");
+			return;
+		}
+		String queryResultFile = "";
+		if(args.length==3){
+			queryResultFile = args[2];
+		}
+		
+		System.out.println("1. queryFile="+queryFile);
+		System.out.println("2. queryStrategy="+queryStrategy);
+		System.out.println("3. queryResultFile="+queryResultFile);
+		
+		//2. read queries
+		QueryList queryList= new QueryList();
+		//queryList.read(queryFile);
+		queryList.setTest();
+		System.out.println(Debugger.getCallerPosition()+"queryList="+queryList);
+		OboeQueryResultContainer queryResultList = new OboeQueryResultContainer(); 
+		
+		//3. perform queries
+		for(int i=0;i<queryList.size();i++){
+			Query queryI = queryList.getQuery(i);
+			OboeQueryResult queryResult = queryI.execute(queryStrategy);
+			queryResultList.add(queryResult);
+		}
+		
+		//4. write the query result to the result file
+		queryResultList.write(queryResultFile);		
+	}
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/SyntheticDataGenerator.java

@@ -77,7 +77,7 @@ public class SyntheticDataGenerator {
 	 */
 	public static void main(String[] args) {
 		if(args.length!=2){
-			System.out.println("Usage: ./MaterializeDB <0. file name prefix> <1. number of rows> ");
+			System.out.println("Usage: ./SyntheticDataGenerator <0. file name prefix> <1. number of rows> ");
 			return;
 		}
 		

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/SyntheticDataValidator.java

@@ -52,7 +52,7 @@ public class SyntheticDataValidator {
 	public static void main(String[] args)
 	{
 		if(args.length!=2){
-			System.out.println("Usage: ./MaterializeDB <0. file name prefix> <1. number of rows> ");
+			System.out.println("Usage: ./SyntheticDataValidator <0. file name prefix> <1. number of rows> ");
 			return;
 		}
 		

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/SyntheticQueryGenerator.java

@@ -1,0 +1,13 @@
+package org.ecoinformatics.oboe;
+
+public class SyntheticQueryGenerator {
+	
+	public static void main(String[] args) {
+		//if(args.length!=2){
+		//	System.out.println("Usage: ./SyntheticQueryGenerator <0. file name prefix> <1. number of rows> ");
+		//	return;
+		//}
+		
+		System.out.println("This has not been implemented yet....");		
+	}
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/datastorage/PostgresDB.java

@@ -29,8 +29,8 @@ public class PostgresDB {
 	private String m_obsInstanceTable = "observation_instance";
 	
 	
-	private String m_measTypeTable = "measurement_type";
-	private String m_measInstanceTable = "measurement_instance";
+	protected String m_measTypeTable = "measurement_type";
+	protected String m_measInstanceTable = "measurement_instance";
 	
 	private String m_contextTypeTable = "context_relationship";
 	private String m_contextInstanceTable = "context_instance";
@@ -75,6 +75,11 @@ public class PostgresDB {
             System.out.println(Debugger.getCallerPosition()+"Couldn't connect: print out a stack trace and exit.");
             throw se;
         }
+	}
+	
+	public Connection getConnection()
+	{
+		return m_conn;
 	}
 	
 	/**
@@ -286,9 +291,13 @@ public class PostgresDB {
 	 * @throws SQLException 
 	 */
 	private void setMeasTypeParam(PreparedStatement stmt, Long annotId,Observation ot, Measurement mt) throws SQLException{
+		String characteristicName = "";
 		String standardName = "";
 		String protocalName= "";
 		
+		if(mt.getCharacteristics()!=null){
+			characteristicName = mt.getCharacteristics().get(0).getName();
+		}
 		if(mt.getStandard()!=null){
 			standardName = mt.getStandard().getName();
 		}
@@ -301,7 +310,7 @@ public class PostgresDB {
 		stmt.setString(2,mt.getLabel());
 		stmt.setString(3,ot.getLabel());
 		stmt.setBoolean(4,mt.isKey());
-		stmt.setString(5,null);
+		stmt.setString(5,characteristicName);
 		stmt.setString(6,standardName);
 		stmt.setString(7, protocalName);
 	}
@@ -452,7 +461,7 @@ public class PostgresDB {
 		}
 		
 		long annotationId = insAnnotationFile(annotationUri);
-		System.out.println(Debugger.getCallerPosition()+",annotationUri="+annotationUri+",annotationId="+annotationId);
+		System.out.println(Debugger.getCallerPosition()+"annotationUri="+annotationUri+", annotationId="+annotationId);
 		
 		PreparedStatement pstmtObs = m_conn.prepareStatement(m_insertObservationType);
 		PreparedStatement pstmtMeas = m_conn.prepareStatement(m_insertMeasurementType);

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/model/MeasurementInstance.java

@@ -16,6 +16,12 @@ public class MeasurementInstance<T> implements Comparable<MeasurementInstance> {
 	private OboeModel oboe;
 	private Annotation a;
 	
+	public MeasurementInstance(Measurement _measType, String _measValue)
+	{
+		this.setMeasValue(_measValue);
+		this.setMeasurementType(_measType);		
+	}
+	
 	public MeasurementInstance(Measurement _measType, ObservationInstance _obsInstance, String _measValue)
 	{
 		setMeasId(gMeasId++);

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/Constant.java

@@ -1,0 +1,11 @@
+package org.ecoinformatics.oboe.query;
+
+public class Constant {
+	public static String QUERY_COMMENT_PREFIX = "#";
+	
+	//Query strategy, this need to be consistent with the values used to call QueryProcessor.main function
+	public static int QUERY_REWRITE = 1;
+	public static int QUERY_MATERIALIZED_DB = 2;
+	public static int QUERY_PARTIAL_MATERIALIZED_DB = 3;
+	
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/MDB.java

@@ -1,0 +1,62 @@
+package org.ecoinformatics.oboe.query;
+
+import java.sql.Connection;
+import java.sql.ResultSet;
+import java.sql.Statement;
+
+import org.ecoinformatics.oboe.Debugger;
+import org.ecoinformatics.oboe.datastorage.PostgresDB;
+
+/**
+ * Class for materialized database 
+ * 
+ * @author cao
+ *
+ */
+public class MDB extends PostgresDB{
+
+	/**
+	 * Translate the given query to sql on matererialized database 
+	 * @param query
+	 * @return
+	 */
+	private String translateQuery(Query query)
+	{
+		String sql = "";
+		sql +="SELECT DISTINCT record_id ";
+		sql +="FROM "+super.m_measInstanceTable+" AS mi," + super.m_measTypeTable + " AS mt ";
+		String whereSql = query.getQueryMeasurementString();
+		if(whereSql.length()>0){
+			sql +="WHERE " + whereSql;
+		}
+		//TODO: add GROUP BY and HAVING clause
+		sql+=";";
+		System.out.println(Debugger.getCallerPosition()+"\n"+sql);
+		return sql;
+	}
+	
+	public OboeQueryResult query(Query query) throws Exception
+	{
+		OboeQueryResult queryResult = new OboeQueryResult();
+		
+		//open database connection
+		super.open();
+		
+		String sql = translateQuery(query);
+		
+		Connection conn = super.getConnection();
+		Statement stmt = conn.createStatement();
+		ResultSet rs = stmt.executeQuery(sql);
+		while(rs.next()){
+			String id = rs.getString(1);
+			System.out.println("id= "+ id);
+		}
+		rs.close();
+		stmt.close();
+			
+		//close database connectioni
+		super.close();
+		
+		return queryResult;		
+	}
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/OboeQueryResult.java

@@ -1,0 +1,5 @@
+package org.ecoinformatics.oboe.query;
+
+public class OboeQueryResult {
+
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/OboeQueryResultContainer.java

@@ -1,0 +1,43 @@
+package org.ecoinformatics.oboe.query;
+
+import java.io.FileNotFoundException;
+import java.io.PrintStream;
+import java.util.ArrayList;
+
+import org.ecoinformatics.oboe.Debugger;
+
+public class OboeQueryResultContainer {
+	ArrayList<OboeQueryResult> m_queryResult = null;
+	
+	public OboeQueryResultContainer(){
+		m_queryResult = new ArrayList<OboeQueryResult>(); 		
+	}
+	
+	public void add(OboeQueryResult queryResult){
+		m_queryResult.add(queryResult);
+	}
+	
+	private void write(PrintStream o)
+	{
+		System.out.println(Debugger.getCallerPosition()+"Not finished...");
+		System.exit(0);
+	}
+
+	/**
+	 * Write the query results to a file
+	 * @param queryResultFile
+	 * @throws FileNotFoundException 
+	 */
+	public void write(String queryResultFile) throws FileNotFoundException
+	{
+		if(queryResultFile==null||(queryResultFile.trim().length()==0))
+			return;
+		
+		PrintStream resultPrintStream = new PrintStream(queryResultFile);
+		write(resultPrintStream);
+		resultPrintStream.close();
+		
+		System.out.println("Result is written to "+queryResultFile);		
+		
+	}
+}

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/Query.java

@@ -1,0 +1,127 @@
+package org.ecoinformatics.oboe.query;
+
+import java.util.List;
+import java.util.ArrayList;
+
+import org.ecoinformatics.oboe.Debugger;
+import org.ecoinformatics.sms.annotation.Context;
+import org.ecoinformatics.sms.annotation.Measurement;
+import org.ecoinformatics.sms.annotation.Observation;
+import org.ecoinformatics.sms.annotation.Characteristic;
+
+import org.ecoinformatics.oboe.model.*;
+
+public class Query {
+	
+	private Observation m_qObsType = null;
+	//private Measurement m_qMeasurementType = null;
+	private MeasurementInstance m_queryMi = null;
+	private Context m_qContextType = null;
+	
+	/**
+	 * Parse the query text to a query structure
+	 * @param queryText
+	 */
+	public void parse(String queryText){
+		System.out.println(queryText);
+		System.out.println(Debugger.getCallerPosition()+"Notfinished...");
+		System.exit(0);
+	}
+	
+	/**
+	 * Set a test query: with TaxonomicTypeName characteristic = Picea rubens
+	 */
+	public void setTest()
+	{
+		Characteristic characteristic = new Characteristic();
+		characteristic.setName("TaxonomicTypeName");
+		Measurement measurementType = new Measurement();
+		measurementType.addCharacteristic(characteristic);
+		m_queryMi = new MeasurementInstance(measurementType,"Picea rubens");
+	}
+	
+	/**
+	 * Convert a query to a string, for output purpose
+	 */
+	public String toString()
+	{
+		String str="";
+		if(m_queryMi!=null){
+			Measurement mt = m_queryMi.getMeasurementType();
+			str +="Measurment ";
+			if(mt.getCharacteristics()!=null){
+				Characteristic cha = mt.getCharacteristics().get(0);
+				str += ("Characteristic "+cha.getName()+" is ");
+			}
+			String mVal = m_queryMi.getMeasValue();
+			if(mVal!=null){
+				str +=" Value: " + mVal;
+			}
+		}
+		
+		return str;
+	}
+	
+	/**
+	 * Get the list of query measurement instances
+	 * 	
+	 * @return
+	 */
+	public List<MeasurementInstance> getQueryMeasements(){
+		List<MeasurementInstance> queryMeasList = new ArrayList<MeasurementInstance>();
[…]

### 2010-05-11 dev/sms/src/org/ecoinformatics/oboe/query/QueryList.java

@@ -1,0 +1,100 @@
+package org.ecoinformatics.oboe.query;
+
+import java.io.BufferedReader;
+import java.io.FileNotFoundException;
+import java.io.FileReader;
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.List;
+import java.util.Map;
+
+public class QueryList {
+	ArrayList<Query> m_queryList = null;
+	
+	public QueryList(){
+		m_queryList = new ArrayList<Query>(); 		
+	}
+	
+	/**
+	 * Return the query list as a string
+	 */
+	public String toString()
+	{
+		String str="";
+		for(int i=0;i<m_queryList.size();i++){
+			str += m_queryList.get(i).toString() + "\n";			
+		}
+		return str;
+	}
+
+	/**
+	 * Return the size of the query list
+	 * @return
+	 */
+	public int size()
+	{
+		return m_queryList.size();
+	}
+	
+	/**
+	 * Get the i-th query. i is in [0, size()-1]
+	 * 
+	 * @param i
+	 * @return
+	 */
+	public Query getQuery(int i)
+	{
+		return m_queryList.get(i);
+	}
+	
+	
+	/**
+	 * Read from a file r all the queries
+	 * and put them to the query lsit
+	 * @param r
+	 * @throws IOException 
+	 */
+	private void read(BufferedReader r) throws IOException
+	{
+		String oneQuery = "";
+		
+		try {
+			while((oneQuery = r.readLine())!=null){
+				if(oneQuery.trim().length()>0&&oneQuery.startsWith(Constant.QUERY_COMMENT_PREFIX))
+					continue;
+				Query query = new Query();
+				
+				query.parse(oneQuery);
+				m_queryList.add(query);
+			}
+		} catch (IOException e) {
+			e.printStackTrace();
+			throw e;
+		}
+	}
+	
+	/**
+	 * Read the given query file to the query list
+	 * @param queryFile
+	 * @throws IOException
+	 */
+	public void read(String queryFile) throws IOException
+	{
+		BufferedReader bufferedReader = null; 
+		bufferedReader = new BufferedReader(new FileReader(queryFile));
+		
+		read(bufferedReader);
+		
+		if (bufferedReader != null) bufferedReader.close();		
+	}
+	
+	/**
+	 * Set a test query listt
+	 */
+	public void setTest()
+	{
+		Query query = new Query();
+		query.setTest();
+		m_queryList.add(query);
+	}
+}

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
rearrange/align annotation fields, text and buttons

@@ -28,11 +28,13 @@
 
 package org.ecoinformatics.sms.plugins.pages;
 
+import java.awt.Color;
 import java.awt.GridLayout;
 import java.awt.event.ActionEvent;
 import java.util.List;
 
 import javax.swing.AbstractAction;
+import javax.swing.BorderFactory;
 import javax.swing.Box;
 import javax.swing.BoxLayout;
 import javax.swing.JButton;
@@ -42,6 +44,7 @@ import javax.swing.JLabel;
 import javax.swing.JPanel;
 import javax.swing.JTextField;
 import javax.swing.JToggleButton;
+import javax.swing.border.TitledBorder;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
@@ -180,11 +183,12 @@ public class AnnotationPage extends AbstractUIPage {
 		}
 		
 		// Attribute Label
-		JPanel attributeLabelPanel = WidgetFactory.makePanel();
-		attributeLabelPanel.add(WidgetFactory.makeLabel("Current Attribute:", false, null));
-		attributeLabel = WidgetFactory.makeLabel("?", true);
+		JPanel attributeLabelPanel = WidgetFactory.makePanel(2);
+		attributeLabel = WidgetFactory.makeLabel("<none selected>", true);
+		
+		attributeLabelPanel.add(WidgetFactory.makeLabel("Selected Attribute:", false, null));
 		attributeLabelPanel.add(attributeLabel);
-		attributeLabelPanel.add(WidgetFactory.makeLabel("(select a column to begin)", false, null));
+		//attributeLabelPanel.add(WidgetFactory.makeLabel("(select a column to begin)", false, null));
 		attributeLabelPanel.add(Box.createHorizontalGlue());
 		attributeLabelPanel.add(helpButton);
 		attributeLabelPanel.add(editButton);
@@ -194,12 +198,21 @@ public class AnnotationPage extends AbstractUIPage {
 		//add the main panel here
 		simpleAnnotationPanel = new SimpleAnnotationPanel(true, showAll);
 		this.add(simpleAnnotationPanel);
+		this.add(WidgetFactory.makeDefaultSpacer());
 		
 		// same as other observation?
 		JPanel sameObservationPanel = WidgetFactory.makePanel(2);
-		sameObservationPanel.add(WidgetFactory.makeLabel("Same entity sample or individual as:", false, null));
[…]

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/plugins/ui/SimpleAnnotationPanel.java

@@ -194,22 +194,22 @@ public class SimpleAnnotationPanel extends JPanel {
 				
 		// Measurement template
 		JPanel measurementPanel = WidgetFactory.makePanel(2);
-		//measurementPanel.add(Box.createHorizontalGlue());
-		measurementPanel.add(WidgetFactory.makeLabel("Select the Measurement template used: ", false, null));
+		measurementPanel.add(Box.createHorizontalGlue());
+		measurementPanel.add(WidgetFactory.makeLabel("Use pre-configured measurement template: ", false, null));
 		observationMeasurement = OntologyClassField.makeLabel("", true, null);
 		observationMeasurement.setFilterClass(Annotation.OBOE_CLASSES.get(Measurement.class));
 		//measurementPanel.add(OntologyClassField.wrapField(observationMeasurement, "The Measurement template used"));
 		measurementPanel.add(observationMeasurement);
 		//measurementPanel.add(WidgetFactory.makeLabel(" or select classes below", false, null));
-		measurementPanel.add(Box.createHorizontalGlue());
-//		measurementPanel.setBorder(
-//				BorderFactory.createTitledBorder(
-//						BorderFactory.createMatteBorder(1, 0, 0, 0, Color.gray), 
-//						null, //"Measurement Template", 
-//						TitledBorder.RIGHT, 
-//						TitledBorder.TOP,
-//						WizardSettings.WIZARD_CONTENT_BOLD_FONT,
-//						WizardSettings.WIZARD_CONTENT_TEXT_COLOR));
+		//measurementPanel.add(Box.createHorizontalGlue());
+		measurementPanel.setBorder(
+				BorderFactory.createTitledBorder(
+						BorderFactory.createMatteBorder(1, 0, 0, 0, Color.gray), 
+						null, //"Measurement Template", 
+						TitledBorder.RIGHT, 
+						TitledBorder.TOP,
+						WizardSettings.WIZARD_CONTENT_BOLD_FONT,
+						WizardSettings.WIZARD_CONTENT_TEXT_COLOR));
 		
 		// listen for the measurement to be set
 		observationMeasurement.addActionListener(
@@ -221,7 +221,8 @@ public class SimpleAnnotationPanel extends JPanel {
 		
 		// Characteristic and Entity
 		JPanel characteristicPanel = WidgetFactory.makePanel(2);
-				
[…]

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java
add experimental layouts, but they remain commented out

@@ -18,9 +18,15 @@ import org.ecoinformatics.sms.annotation.Protocol;
 import org.ecoinformatics.sms.annotation.Relationship;
 import org.ecoinformatics.sms.annotation.Standard;
 
+import com.mxgraph.layout.mxCircleLayout;
+import com.mxgraph.layout.mxCompactTreeLayout;
+import com.mxgraph.layout.mxEdgeLabelLayout;
 import com.mxgraph.layout.mxFastOrganicLayout;
 import com.mxgraph.layout.mxGraphLayout;
+import com.mxgraph.layout.mxOrganicLayout;
+import com.mxgraph.layout.mxStackLayout;
 import com.mxgraph.layout.hierarchical.mxHierarchicalLayout;
+import com.mxgraph.layout.orthogonal.mxOrthogonalLayout;
 import com.mxgraph.model.mxCell;
 import com.mxgraph.swing.mxGraphComponent;
 import com.mxgraph.util.mxConstants;
@@ -56,7 +62,9 @@ public class AnnotationGraph {
 		// the edge style
 		Map<String, Object> edgeStyleMap = new HashMap<String, Object>();
 		edgeStyleMap.put(mxConstants.STYLE_ELBOW, mxConstants.ELBOW_VERTICAL);
+		//edgeStyleMap.put(mxConstants.STYLE_ELBOW, mxConstants.ELBOW_HORIZONTAL);
 		edgeStyleMap.put(mxConstants.STYLE_EDGE, mxConstants.EDGESTYLE_ELBOW);
+		//edgeStyleMap.put(mxConstants.STYLE_EDGE, mxConstants.EDGESTYLE_LOOP);
 		edgeStyleMap.put(mxConstants.STYLE_VERTICAL_ALIGN, mxConstants.ALIGN_BOTTOM);
 		stylesheet.putCellStyle("Edge", edgeStyleMap);
 
@@ -205,6 +213,10 @@ public class AnnotationGraph {
 			Object parent = graph.getDefaultParent();
 			
 			//mxGraphLayout layout = new mxFastOrganicLayout(graph);
+			//mxGraphLayout layout = new mxCircleLayout(graph);
+			//mxGraphLayout layout = new mxOrganicLayout(graph);
+			//mxGraphLayout layout = new mxCompactTreeLayout(graph);
+
 			mxHierarchicalLayout layout = new mxHierarchicalLayout(graph, SwingConstants.WEST);
[…]

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/plugins/pages/AnnotationPage.java
The "Same individual" dropdown now informs the save process and merges the observation as needed. If an observation no longer has any measurements, it is completely removed from the annotation (including any context relationships it may be providing).

@@ -31,6 +31,10 @@ package org.ecoinformatics.sms.plugins.pages;
 import java.awt.Color;
 import java.awt.GridLayout;
 import java.awt.event.ActionEvent;
+import java.awt.event.ItemEvent;
+import java.awt.event.ItemListener;
+import java.util.ArrayList;
+import java.util.Iterator;
 import java.util.List;
 
 import javax.swing.AbstractAction;
@@ -48,6 +52,7 @@ import javax.swing.border.TitledBorder;
 
 import org.ecoinformatics.sms.annotation.Annotation;
 import org.ecoinformatics.sms.annotation.Characteristic;
+import org.ecoinformatics.sms.annotation.Context;
 import org.ecoinformatics.sms.annotation.Entity;
 import org.ecoinformatics.sms.annotation.Mapping;
 import org.ecoinformatics.sms.annotation.Measurement;
@@ -203,7 +208,15 @@ public class AnnotationPage extends AbstractUIPage {
 		// same as other observation?
 		JPanel sameObservationPanel = WidgetFactory.makePanel(2);
 		sameObservationPanel.add(WidgetFactory.makeLabel("Measurements were taken on the same sample or individual as:", false, null));
-		existingObservationList = WidgetFactory.makePickList(null, false, 0, null);
+		ItemListener observationListener = new ItemListener() {
+			public void itemStateChanged(ItemEvent e) {
+				Observation obs = (Observation) existingObservationList.getSelectedItem();
+				if (obs != null) {
+					simpleAnnotationPanel.setObservationEntity(obs.getEntity());
+				}
+			}
+		};
+		existingObservationList = WidgetFactory.makePickList(null, false, 0, observationListener);
 		sameObservationPanel.add(existingObservationList);
 		sameObservationPanel.setBorder(
 				BorderFactory.createTitledBorder(
@@ -276,6 +289,17 @@ public class AnnotationPage extends AbstractUIPage {
 		try {
 			// what are we editing:
 			attributeLabel.setText(currentAttributeName);
+			
+			try {
+				List<Observation> existingObservations = annotation.getObservations();
+				existingObservationList.removeAllItems();
+				existingObservationList.addItem(null);
+				for (Observation o: existingObservations) {
+					existingObservationList.addItem(o);
+				}
+				
+			} catch (Exception e) {
+			}
 
 			// is there a measurement mapping for the attribute?
[…]

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/plugins/pages/ViewAnnotationPage.java
invert the boolean switch for showing context only/all parts of the annotation. default is to show all using a manual graph layout.

@@ -93,7 +93,7 @@ public class ViewAnnotationPage extends AbstractUIPage {
 		// annotation graph
 		JPanel annotationGraphPanel = WidgetFactory.makePanel();
 		// create the graph
-		final Component graphComponent = AnnotationGraph.createAnnotationGraph(annotation, false);
+		final Component graphComponent = AnnotationGraph.createAnnotationGraph(annotation, true);
 		JButton refreshButton = WidgetFactory.makeJButton("Re-Layout", new ActionListener() {
 			public void actionPerformed(ActionEvent e) {
 				AnnotationGraph.layoutGraph(graphComponent);

### 2010-05-12 dev/sms/src/org/ecoinformatics/sms/renderer/AnnotationGraph.java

@@ -37,7 +37,7 @@ import edu.ucsb.nceas.morpho.plugins.datapackagewizard.WidgetFactory;
 
 public class AnnotationGraph {
 	
-	public static Component createAnnotationGraph(Annotation annotation, boolean contextOnly) {
+	public static Component createAnnotationGraph(Annotation annotation, boolean showAll) {
 		
 		// the graph
 		mxGraph graph = new mxGraph();
@@ -106,7 +106,7 @@ public class AnnotationGraph {
 
 				observationMap.put(observation, (mxCell)observationNode);
 				
-				if (contextOnly) {
+				if (!showAll) {
 					continue;
 				}
 				
@@ -194,7 +194,7 @@ public class AnnotationGraph {
 		//add to the page
 		mxGraphComponent graphComponent = new mxGraphComponent(graph);
 		
-		if (contextOnly) {
+		if (!showAll) {
 			layoutGraph(graphComponent);
 		}
 		

### 2010-05-13 docs/algdraft/DataIntegration/intro.tex
Decompose the draft to several parts so that we can work on it collaboratively.
Until further discussion, 
from now on, 
Huiping will work on method.tex and Shawn probably will work on modeldef.tex

--- 2010-05-07 docs/algdraft/DataIntegration/DataIntegration.tex
@@
-\documentclass[conference]{IEEEtran}
-
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-
-\usepackage{algorithm}
-\usepackage{algorithmic}
- \renewcommand\floatpagefraction{0.99}
- \renewcommand\topfraction{0.99}
- \renewcommand\bottomfraction{0.99}
- \renewcommand\textfraction{.05}
- \setcounter{totalnumber}{5}
-
-\newtheorem{example}{Example}[section]
-\newtheorem{definition}{Definition}[section]
-\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
-
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-\begin{document}
-
-\title{Querying Integrated Scientific Observation and Measurement Data}
-
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-\author{\IEEEauthorblockN{Author1}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
-\and
-\IEEEauthorblockN{Author2}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{Author3}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678-2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
-
-\maketitle
-
-\begin{abstract}
-%\boldmath
-The abstract goes here.
-\end{abstract}
-
-% no keywords
-
 \section{Introduction}\label{sec:intro}
@@
 queries that people are interested to ask.
-
-\section{Related work}\label{sec:relatedwork}
-
-\from{HP}{This will come after the real problem definition and the
-  solution. The description in the following several sections may be
-  moved to here later.}
-
-\section{Data model, annotation and query}\label{sec:dataquery}
-
-In this section, we first illustrate the data model. Then, we
-formalize the queries that scientists in this domain tend to ask.
-
-\subsection{Data model}\label{sec:datamodel}
-
-%Data model: Data, attribute, data value set
-When a scientist contributes data into an integrated data repository,
-a widely accepted way is to convert each dataset to a data
-table\cite{tdar} or treated as a separate object entity\cite{knb}. 
-{\bf (HP: is it really widely used? any other system uses this
-  way?. Add more citations here.)}
-In using this method, the database contains metadata of each
-dataset and the definition of the data table (e.g., attribute/column
-name, attribute type, etc).
-
-%HP: 
-%An alternative way to organize the data is to store them in one big
-%data table with three columns data\_file, attribute\_name,
-%attribute\_value. The problem with this structure is that the
-%attribute\_value are different types.
-In our work, we focus on querying this kind of data repository (or
-databases). To formalize the scenario, we use $D$ to denote the set of data
-tables in the data repository and $d$ to refer to a specific data
-table. Each data table $d$ contains metadata about the attribute
-definition $Attr_{d}$. Sometimes, we also use {\em column} to refer to an
-attribute. 
-Given one data table $d$, an attribute $attr_i \in Attr_{d}$ or column index $i$, 
-$d[attr_i]$ or $d[i]$ represents the set of values for the attribute
-$attr_i$ or for the $i$-th column.
-%HP: Implementation detail (no need to put it in the paper): 
-%When we put the data to data tables, we add a unique record id for
-%each row in the data. This is used to get the unique record id.
-
-\subsection{Annotation}\label{sec:annotation}
-
-%What kinds of annotation information is needed. 
-We use $A$ to denote the annotation of one dataset. 
-Internally, we keep the following information for the
-annotation.
-
-For data table $d$ with annotation $A$, the system keeps the
-information $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$.
-As we described before, we have four main objects to describe: 
-observation type (OT), measurement type (MT), and
-context type (CT), and the mapping (Map) from the measurement type to
-resource attributes.  
-So, the annotation contains the following information.
-
-\begin{itemize}
-\item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
-  describe an observation type, which denotes on which
-  entity type (object in the real world) the observation is made. 
-Very often, more than one observation can be made on one entity. Thus,{\em
-  isDistinct} is used to denote whether the same value of key
-meaasurements of an obbservation types can uniquely identify one
-observation or not. 
-\item $MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
-  Cha, \cdots)\}$. Generally, $MT$ contains information about
-  characteristic ($Cha$), Standard, Protocol and Precision. We do not include them
-  here to make the description clearer. Here, {\em isKey} is used to denote
-  whether one measurement type is the key measurement for the
-  observation type $OT_{id}$ or not. 
-\item $CT$ = $\{(\underline{A_{id}, ct_{id}}, ot_{id}, cot_{id}, Rel,
-  isIdentify)\}$ where $cot$ represents the context observation
-  type. 
-\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, MapCond},
-  Val\}$
-\end{itemize}
-
-We use the following example to illustrate the concepts in the
-annotation.
-
-\begin{example}
-$OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
-  denotes that annotation $A_1$ are for two observation types. One
-  type $OT_1$ is on real world entity ``tree'' and the values of its key measurements
-  do not uniquely identify a distinct observation. The other type
-  $OT_2$ is for real world entity ``GeoSpot'' (i.e., geospatial
-  location) and its key measurement types can identify it's unique
-  observations. \from{HP}{Introduce key measurements before this.}
-
-$MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
-($A_1, MT_2, OT_1$, SpecNo, true), 
-($A_1, MT_3, OT_1$, Height, false),
-($A_1, MT_4, OT_2$, Plot\_State, true), 
-($A_1, MT_5, OT_2$, Plot\_Area, false)\} represents that $OT_1$
-(for ``tree'') has three measurement types ($MT_1, MT_2$, $MT_3$) for characteristics
-``Species'', ``SpecNo'', and ``Height''. The fist two measurement
-  types together form the key measurement type for $OT_1$. 
-$OT_2$ (for ``GeoSpot'') has one key measurement type $MT_4$ for
-  characteristic ``Plot\_State'' and another measurement type $MT_5$
-  for characteristic ``Plot\_Area''.
-
-$CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
-observation of a {\em tree} is within the context of a geo-spatial
-location. And this context is used to identify the uniqueness of the
-observation. E.g., a tree with species name ``piru'' and species no
-$1$ in California
-%$(34^{\circ}8'3N,118^{\circ}14'37W)$ (Los Angeles in California) 
-is different from a tree with 
-the same species name ``piru'' at Oregon. 
-%location $36^{\circ}10'30N,115^{\circ}8'11W$ (Las Vegas in
-%California).
-
-%HP coordinates material: 
-%http://www.traveljournals.net/explore/united_states/map/u1662328/los_angeles.html
-%http://en.wikipedia.org/wiki/Las_Vegas,_Nevada
-
-$Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
-%($A_1$, $MT_1$, ``code'', ``eq `abba''', ``Abies balsamea''),
-%($A_1$, $MT_1$, ``code'', ``eq `capo''', ``California poppy''),
-($A_1$, $MT_2$,``no'', null,), 
-($A_1$, $MT_3$, ``plt'', null,), 
-($A_1$, $MT_4$,``area'', null,\}. 
-The first mapping rules maps ``code''
-attribute to the measurement type $MT_1$ (for ``Species''
-characteristic) and change the value to ``Picea rubens'' if the code
-is ``piru''. The meaning of the other mapping rules are very
-obvious here, so we skip the details. 
-\end{example}
-
-\subsection{Observation and measurement (OM) query}
-Above we defined the data model and the annotation. In this
-sub-section, we formalize the queries that scientists are interested
-in. We denote the queries on the observational data as {\em Observation
-and measurement (OM) query}.
-
-As discussed in Section \ref{sec:intro}, it is generally impossible for a
-scientist to formulate an OM query using the internal data structure
-of a dataset because such data structures differ from contributors to
-contributors. More realistically, they can pose queries
-(Example \ref{eg:query}) related to key OM concepts, e.g., {\em
-  observation, measurement,characteristic, and standard}, and these
-OM concepts satisfy given keyword and context conditions.
-
-\begin{definition}[Basic OM query]\label{def:basic_omq}
-A basic observation and measurement query is defined as 
-\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]. 
-Here, 
-\begin{itemize}
-\item {\tt OMconcept} refers to one of the major OM terminologies. 
-\item 
-$\langle${\tt cond}$\rangle$ is in the form of 
-{\tt f([distinct entity.]measurement) op value} 
-where {\tt op} is a basic operator from $\{=, \neq, >, <\}$ 
-and {\tt f} is an
-aggregation function from $\{sum, avg, count, min, max\}$. 
-{\tt distinct} and {\tt entity} are optional. 
-\end{itemize}
-\from{HP}{The symbol fonts are a little bit messy...need to
-  distinguish between the keywords and variables.}
-\end{definition}
-
-\begin{definition}[Result of OM query]\label{def:qresult}
-The result of a query $Q$ is a set of data objects (e.g., data tables) 
-\{$d|d\in D \wedge d~satisfies~\langle ${\tt cond} $\rangle$\}. 
-For each of such result $d$, we use $d~s.t.~Q$ to denote that
-an data object $d$ satisfies the query. 
-\end{definition}
-
-Based on the different $\langle${\tt cond}$\rangle$ definitions,
-``$d~satisfies~\langle${\tt cond}$\rangle$''
-is translated differently. 
-E.g., if $\langle${\tt cond}$\rangle$ is defined that the value of ``{\em area}'' need to be
-smaller than $3.0$, then, ``$d~satisfies~{\tt cond}$'' is translated to
-``$dv<3.0|dv\in d[area]$''.
-
-Definition \ref{def:basic_omq} can formulate queries on
-one specific observation and measurement. 
-In real application, we need to consider the context relationship. 
-\from{HP}{E.g., etc.}
-So, we generalize the basic OM query to contextualized OM query as
-follows.
-
-\begin{definition}[Contextualized OM query]\label{def:context_omq} A
-  contextualized OM query is defined as:
-\[CQ ::=Q_i \wedge {\tt context}(Q_i.{\tt OMconcept}, Q_j.{\tt
-  OMconcept})\wedge Q_j\]
-Here, $Q_i, Q_j$ are basic queries.
-\end{definition}
-
-For example, the queries in Example \ref{eg:query} can be formalized
-as the following formal OM queries.
-
-%\begin{example}
-\begin{itemize}
-\item $Observation: \langle Species='Picea~rubens'\rangle$
-\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
-  $\wedge IN(Observation_1, Observation_2$) \\
-$\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
-\item $Observation: \langle Species='Picea~rubens'$\\
-$\wedge count(distinct~Species)\geq 5\rangle$
-\item $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
-$\wedge IN(Observation_1, Observation_2)$\\
-$\wedge Observation_2: \langle State='California'\rangle$\\
-%Give me the data sets that have trees with average ``height''
-%  than 20.0 in ``California''
-\end{itemize}
-
-%\end{example}
-
-%\from{HP}{Add experiement: show the error rate (e.g., precision, recall, (or
-%  directly $F_1$) of this method.? not needed?}
-
-\section{Querying annotated and integrated OM data}
-
-In this section, we propose several methods that can be used to
-answer OM queries over scientific observational data with
-annotations.
-
-%\subsection{Naive method}\label{sec:naive}
-A very direct way to search such data is to extract all the content in
-the data cells and index them. This way the system may help answer
-very simple queries with the help of some thesaurus. 
-For example, for $Q_1$ in Example \ref{eg:query}, as long as we can
-translate the abbreviation ``piru'' in the dataset to ``Picea rubens''
-based on some predefined thesaurus, we can find the results. 
-However, even for this kind of simple query, if we want to find
-observation of ``California poppy'', an approximate match may return
-the dataset in Table \ref{tb:dataset} as a result because it contains
-both {\em California} and {\em Poppy}.
-If the search problem caused by the compound words can be alleviated using some existing
-techniques \cite{***}, %citation on compound word querying. 
-it is still almost impossible to directly use such IR-style system to answer a query
-in the form of $Q_3$ and $Q_4$. 
-This naive method cannot be directly used answer the desired queries. It is because it
-fails to catch the semantics of the context and uniqueness of observations in the data. 
-In what follows, we utilize the annotation information to get
-more semantic support, and provide new strategies to answer the
-queries.
-
-As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
-in querying integrated data. 
-One is to rewrite the original query to a series of new queries over
-the data; the other is to materialize the data to a consistent data
-model and then answer the query over the materialized database. 
-In what follows we work from these two directions to leverage the
-semantic information in answering such queries.
-
-\subsection{Query rewriting}\label{sec:queryrewrite}
-
-As described in Section \ref{sec:dataquery}, 
-an OM query is represented using 
-OM concept terminologies such as observation
-measurement, characteristic, etc. 
-To answer such query from the original data model (i.e., dataset
-metadata and data tables), we need to {\em rewrite} the quey over the
-real data table. 
-In what follows, we first sketch the re-writing process. Then, we
-detailed the procedure in using the data model and
-structures. Finally, we include the process for the more complicated
-cases (e.g., with distinct, aggregate).
-
-Roughly speaking, the query rewriting consists of two steps:
-\begin{itemize}
-\item From the given query, find the relevant data tables and
-  attributes that need to be used answer the given query. 
-\item Translate the given query to queries over the relevant data
-  tables. 
-\end{itemize}
-
-{\bf TODO: put a figure here to illustrate the steps}.
-
-The first step is to map the OM query to the real data structure. 
-We can utilize the annotation structure $A.Map$ which keeps the correspondences between
-measurement type and table attributes. 
-So, when the given concept is {\em measurement}, we can directly get
-its related attribute names and annotation id $A_{id}$. 
-In case that the given OM concept is a non-measurement concept, we can use
-other Annotation structures $A.*$ to figure out the measurementt types. 
-E.g., if the OM concept is Characteristic ``Species'', we can use
-$A.MeasType$ to get the measurement type id.
-
-As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
-information $\langle A_{id}, DTableId\rangle$. 
-With the $A_{id}$, we can ge the data table information from the
-metadata  $\langle A_{id}, DTableId\rangle$. 
-Then, we get the relevant data table information and the needed attribute
-information.
-
-Let $RD$ be the set of relevant data tables and $d_{a_1} \cdots d_{a_m}$
-are the related attributes for table $d\in RD$. We can translate the original OM query to SQL queries
-over the data tables.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l} 
-SELECT DISTINCT $d_{id}$\\
-FROM $d$\\
-WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
-$[$GROUP BY {\em distinct requirement}$]$\\
-$[$HAVING {\em aggregation condition}$]$;
-\end{tabular}
-}
-\vspace{0.1in}
-
-The above describes the process for the most basic OM query. 
-However, in the real application, people ask queries with distinct
-observation or entity constraints. Also, people ask queries about
-different observatitons using context.
-
-%Discuss how to deal with the distinct constraints.
-Many times, a user may want to find {\em distinct} observations or
-entities. In this case, the above simple solution cannot work
-correctly because it treats observations in different rows as the same
-observation. So, we need to use the annotation to figure out what
-distinguishes one observation from another. 
-When an observation type is denoted with {\em distinct yes} and some
-of its measurement types are denoted with {\em key yes}, 
-the values of the measurements on the key measurement types can
-uniquely distinguish one observation from the other. 
-To implement this, in SQL, we can perform ``group by'' operations on
-the key measurement types of the observations.
-
-%Discuss how to deal with context
-The above description discusses cases without context. 
-However, in real application, the observation has context. 
-When the context is not {\em identifying}, i.e., the context value
-does not affect the uniqueness of one observation, this can be
-processed as the basic way. 
-When the context is denoted with {\em identifying yes}, it means the
-identity of one observation also depend on the contextual
-observation. 
-In this case, we need to figure out the all the key measurement types
-of one observation, then the ``GROUP BY'' operation should be on all
-these key measurement types.
-
-{\bf TODO: put an example here.}
-
-In this case, there is no need to change the SQL scripts.
-the only change are the relevant attribute set $d_{a_1} \cdots
-d_{a_m}$, which corresponds to not only the measurement types that
-directly related to the observation, but also the measurement types
-that are in the identifying context chain of the observation.
-
-\begin{example}\label{eg:query_reqerite_q1}
-{\bf TODO: refine this example} 
-Take $Q_1$ as example. The {\tt concept} is ``observation'' and {\tt cond} is {\em species = ``Picea rubens''};
-
-From the annotation $A.MeasType$, we can find the measurement type
-$m_3$ with the characteristic {\em species}. 
-From the mapping $A.Map$, we further find that the attribute {\em spp} that the
-measurement type $m_3$ is mapped to.
-
-The second step is to find the data file which really have this value. 
-For this one, we find the data table, and do a selection on the table
-content.
-
-{\bf TODO: SQL here.} 
-\end{example}
-
-\from{HP}{Refine the process of $Q_2$.\\
-For the first step, we find the data tables that contain the needed concepts. 
-In each table, we need to figure out the distinct observations. 
-From $m_i$ $obstype_i$, all the other measurements that are of
-$obstype_i$. 
-Group by the key measurements of $obstype_i$ if distinct yes.
-If this observation type does not have any context, then the key
-measurements are all the measurements directly defined under this
-observatioin type. Otherwise, i.e., this observation type has context,
-then the key measurements are all the measurements in the context
-chain. }
-
-%Analysis of this method.
-
-{\bf Execution cost analysis:} 
-From the above description, we can tell that the computation cost is to search the different data
-tables. This way, for each candidate data table, we need to send an
-SQL query to the server for evaluation purpose.
-If the number of candidate data tables are small. This method should
-work well. However, when we have a lot of candidate data tables, the
-efficiency may be affected. 
-In what follows, we introduce another strategy, which logically merge
-the data content in different tables and perform queries ober the
-materialized database.
-
-\subsection{Querying materialized database}\label{sec:matdb}
-
-The previous section shows the method to rewrite a given query. 
-However, as we analyzed, the computation cost may be higher when the
-candidate data tables are many. In this section, we propose another
-strategy to make use of the existing optimization strategies of
-current DBMS.
-
-The basic idea is to materialize the data into some centralized
-concept tables with instances of entities, observations, measurements
-and the contexts between observations. 
-We call such concept tables core OBOE tables, and denote them as
-$OBOE.*$. 
-In partibular, we have the following concept instance tables. 
-\begin{itemize}
-%\item $EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
-\item $OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
-  all the observation instances where $ot_{id}$ is used to link to the
-  annotation information in $OT$; 
-\item $MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
-  mVal)\}$ for all the measurement instances where $mt_{id}$ is used to link to the
-  annotation information in $MT$; 
-\item $CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
-  all the context instances where $ct_{id}$ is used to link to the
-  annotation information in $CT$. 
-\end{itemize}
-
-In what follows, we describe how to materialize the data into such
-concept instance tables in Section \ref{sec:materialilzedb} when
-considering all the distinct and context constratins. 
-Then, in Section \ref{sec:query_materialilzedb}, we describe our strategy to
-perform queries over materialized database.
-
-\subsubsection{Materialize data}\label{sec:materialilzedb}
-{\bf: TODO: refine the writing in MaterializeDB and move them here.}
-
-\subsubsection{Querying materialized database}\label{sec:query_materialilzedb}
-
-Once the data are materialized using the annotation constraints, we
-can perform the query over them. In what follows, we describe how to
-use the materialized data to answer the query.
-
-When the given query is a basic OM query, i.e., there is not context
-in the query. 
-Then, we can form a template query as the following;
-
-\begin{figure}[htb]
-{\tt
-\begin{tabular}{l}
-SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt[$, {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
-$[$GROUP BY {\em distinct requirement} $]$\\
-$[$HAVING   {\em aggregation condition} $]$;
-\end{tabular}
-}
-\caption{SQL template to answer OM using materialized DB}
-\label{fig:sql_matdb_basic_omq}
-\end{figure}
-
-In particular,
-
-\vspace{0.1in}
-\begin{tabular}{l}
-$table~join~condition=$\\
-%$\qquad (mi.oi_{id}=ci.oi_{id})~AND~(cmi.oi_{id}=ci.coi_{id})~AND$\\
-$\qquad (mi.mt_{id}=mt.mt_{id})~AND~(mt.A_{id}=AD.A_{id})$
-\end{tabular}
-\vspace{0.1in}
-
-For the very basic query (e.g., $Q_1$ in Example \ref{eg:query}), the
-SQL generally need only the first three clauses. 
-If the query has some restriction on the aggregated result, the {\tt
-  GROUP BY} and {\tt HAVING} clauses to deal with it.
-
-We use $Q_1$ and $Q_3$ to illustrate the query on the materialized
-database. 
-can be answered using the above
-template. 
-Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
-the query rewriting method.
-
-\begin{example} \label{eg:materialize_db_q1}
-Using the  materialized database,  we just need to search the MI table
-with the condition that {\em mVal=``Picea rubens''}.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt$\\ 
-WHERE $mt.Cha='Species'$ \\
-$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
-$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
-$\qquad\quad$ AND $mi.A_{id}=AD.A_{id}$; 
-%\#Join the MI and MT tables
-\end{tabular}
-}
-\vspace{0.1in}
-
-For $Q_3$, which asks for data sets that contain at
-least five distinct ``Picea rubens'' observations.
-The SQL query can be written as:
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT distinct $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt$\\
-WHERE $mt.Cha='Species'$ \\
-$\qquad\quad$ AND $mi.mVal = `Picea rubens'$\\
-$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
-$\qquad\quad$ AND $mt.A_{id}=AD.A_{id}$\\
-GROUP BY $d_{id},oi_{id}$\\
-HAVING COUNT(*)$>$5; 
-}
-\end{tabular}
-}
-\vspace{0.1in}
-\end{example}
-
-For this kind of query, the computation cost is the selection cost of the
-measurement instance (MI) table. 
-Compared with the query rewriting strategy, which need to pose the
-query over multiple data tables, this query should answer the queries
-more efficiently.
-
-Figure \ref{fig:sql_matdb_basic_omq} shows the SQL template to process
-the basic OM query without context. In case the OM query has context,
-we need to add the context into consideration.
-
-\begin{enumerate}
-\item The first step is similar to that in the basic OM query
-  processing. Besides this, we need to figure out what are the concept
-  obserbationi types of the needed observations.  
-\item \label{step:singlesql} Form an SQL for the observation and every of its context
-  observations. 
-\item Intersect all the SQLs from Step \ref{step:singlesql}). 
-\end{enumerate}
-
-\begin{figure}[htb]
-{\tt
-\begin{tabular}{l}
-SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
-FROM $AD, CI~as~ci,MI~as~mi, MT~as~mt$\\
-$\qquad MI~as~cmi, MT~as~cmt[,$ {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
-$[$GROUP BY {\em distinct requirement} $]$\\
-$[$HAVING   {\em aggregation condition} $]$;
-\end{tabular}
-}
-\caption{SQL template to answer contextualized OM using materialized DB}
-\label{fig:sql_matdb_context_omq}
-\end{figure}
-
-In this case, the table join condition is much more complex since we
-need to use the context instance table $ci$ to connect the
-observation and context observation table. 
-In particular,
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-$table~join~condition=$\\
-$\qquad (mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$
-\end{tabular}
-}
-\vspace{0.1in}
-
-{\bf Execution cost analysis:} From the join condition, we can see that the expensive cost comes from
-the join operation between the measurement instance, context
-measurement instance, and context instance table. 
-As we analyzed in Section \ref{sec:materialilzedb}, the two tables with
-growing space are measurement instance table and context instance
-table.
-
-{\bf Example:} When we are given contextualized query, e.g., $Q_2$, which asks for
-the datasets that contain species ``Picea rubens'' observations in
-``California''.
-
-In the first step, we get the requirement is on Characteristic
-``Species'' with value ($mVal$) ``Picea rubens'' and observation has context {\em in}
-another observation for ``State'' of value ``California''.
-
-%Using materializedDB, we can answer the query this way
-%First, from the Characteristic ``Species'', we can get  --> mlabel, observation type
-
-%select mtype, olabel from measurement\_type  \\
-%where chracteristic = 'Species';  \\
-In the first step, we need to find the context type id $tmp\_ct\_id$
-that satisfy the context relationship.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT AD.d_{id}\\
-FROM $AD, CI~as~ci, MI~as~mi, MT~as~mt,$ \\
-$\quad\qquad MI~as~cmi, MT~as~cmt,$ \\
-WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$\\
-$\qquad$ AND $mi.mVal=`Picea rubens'$ \\
-$\qquad$ AND $cmi.mVal='California'$\\
-$\qquad$ AND $mt.Cha = 'Species'$\\
-$\qquad$ AND $cmt.Cha=`State'$\\
-$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
-\end{tabular}
-}
-\vspace{0.1in}
-
-%Use query rewriting, what I can do? 
-%It's the same till to getting context\_olabel\_set.
-
-%Get the asked condition columns and the context columns,
-
-%select record\_id from data\_tables
-%where condition column 1='value1' and (context column1 = 'context
-%value' or context column2 = `context value', etc.);
-
-\from{HP}{Need to refinement the following. }
-
-=============
-
-Now let's think about the query $Q_4$ with aggregation and context
-requirement.
-
-Also, in the first step, we discover the context type id $tmp\_ct\_id$
-that satisfy the context relationship. 
-Use materialized database: 
-Analysis: ``Area'' is the context of some observation.
-Area --> mlabel, observation type
-
-From the context observation lable, find the observation labels, \\
-select olabel\\
-from context\_relationship\\
-where context\_olabel = 'olabel';
-
-Let olabel\_set  be the set containing all the observation type
-labels that use ``Area'' as context.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT AD.d_{id} \\
-FROM $AD, CI~as~ci, MI~as~mi, MI~as~cmi,$ \\
-$\quad\qquad MT~as~mt, MT~as~cmt,$ \\
-WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$\\
-$\qquad$ AND $cmi.Cha=`State'$\\
-$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
-GROUP BY $cmi.oid$\\
-HAVING $avg(cmi.mVal)>5.0$;
-\end{tabular}
-}
-\vspace{0.1in}
-
-\subsection{Query partially materialized database}
-As we analyzed in the first two sub-sections, 
-the query-rewriting suffers the multiple queries posed on different
-candidate dataset. The querying of materialized database suffers from
-the multiple self-join operation of the measurement instance (MI)
-table, which has the most amount of information.
-
-In this section, we propose to materialize only the measurement
-instance and also put the measurement instance and context measurement
-instances together in the same table to avoid the expensive
-computation cost.
-
-\from{HP}{Let's see how this can work. }
-
-\section{Experiments}
-
-Synthetic data generator.
-
-Algorithm to materialize DB. 
-Report the time and space.
-
-Synthetic query generator.
-
-Test algorithm to perform query over materiazed DB.
-Test algorithm to perform query-rewriting over materiazed DB.
-
-Test algorithm for half-materialized data.
-
-\section{Conclusion}
-
-\bibliographystyle{IEEEtranS}
-\bibliography{IEEEabrv,DataIntegration}
-
-\end{document}

### 2010-05-13 docs/algdraft/DataIntegration/method.tex
Decompose the draft to several parts so that we can work on it collaboratively.
Until further discussion, 
from now on, 
Huiping will work on method.tex and Shawn probably will work on modeldef.tex

--- 2010-05-10 docs/algdraft/DataIntegration/DataIntegration.tex
@@
-\documentclass[conference]{IEEEtran}
-
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-
-\usepackage{algorithm}
-\usepackage{algorithmic}
- \renewcommand\floatpagefraction{0.99}
- \renewcommand\topfraction{0.99}
- \renewcommand\bottomfraction{0.99}
- \renewcommand\textfraction{.05}
- \setcounter{totalnumber}{5}
-
-\newtheorem{example}{\bf Example}[section]
-\newtheorem{definition}{\bf Definition}[section]
-\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
-
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-\begin{document}
-
-\title{Querying Integrated Scientific Observation and Measurement Data}
-
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-\author{\IEEEauthorblockN{Author1}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
-\and
-\IEEEauthorblockN{Author2}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{Author3}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678-2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
-
-\maketitle
-
-\begin{abstract}
-%\boldmath
-The abstract goes here.
-\end{abstract}
-
-% no keywords
-
-\section{Introduction}\label{sec:intro}
-In this work, we study the query processing over scientific
-observation and measurement data using OBOE
-model\cite{DBLP:conf/er/BowersMS08}. OBOE model is a conceptual model
-used to interpret observation and measurement data.
-
-\subsection{Background}
-
-%Application background
-In many scientific domains (e.g., ecology, hydrology, earth science,
-geology), people collect observational data. Such data
-record the observed value of some real world entity at some specific
-place and time. E.g., ecologists studying relationship between the 
-growth pattern and the treatments often need to record the tree
-heights. The collected data reflect the fact the tree height of a
-specific tree is 30.1in on May 1, 2009 and 30.3in on May 1, 2010.
-
-%Characteristics of scientific observational data: non-normalized
-Almost all such scientific data do not follow database higher normal
-forms. Generally, scientists have their way in interpreting their
-data, but they are not ready for any normalization process. 
-For example, Table \ref{tb:dataset} is a simplified but typical dataset collected by
-a scientist who study the growth pattern of trees. 
-Obviously, the ``plt'' of the first two rows is the same place, and
-the ``plt'' of the last two rows is the same. 
-Here, their area information is redundant. 
-In the real application, many columns have redundant information.
-
-%HP: this one has the coordinates, it's troublesome to explain. 
-%% \begin{table}[htb]
-%% \begin{center}
-%% \begin{tabular}{|l|l|l|l|l|}
-%% \hline
-%% code & no & ht & plt & area\\\hline
-%% piru & 1 & 35.8 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
-%% piru & 2 & 36.2 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
-%% piru & 1 & 25.7 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0 \\\hline
-%% %abba & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
-%% capo & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
-%% \end{tabular}
-%% \end{center}
-%% \caption{Dataset}
-%% \label{tb:dataset}
-%% \end{table}
-
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-code & no & ht & plt & area\\\hline
-piru & 1 & 35.8 & California & 4.0\\\hline
-piru & 2 & 36.2 & California & 4.0\\\hline
-piru & 1 & 25.7 & Oregon & 3.0 \\\hline
-%capo & 1 & 7.8 & Oregon & 3.0\\\hline
-Oriental poppy& 1 & 7.8 & Oregon & 3.0\\\hline 
-\end{tabular}
-\end{center}
-\caption{Dataset}
-\label{tb:dataset}
-\end{table}
-%HP: California poppy info
-%Oregon iris: http://en.wikipedia.org/wiki/List_of_native_Oregon_plants
-%http://en.wikipedia.org/wiki/California_poppy
-
-%Introduce major concepts
-In scientific domains where people collect observation and measurement
-data, there are several commonly used and widely recognized canonical
-concepts(\cite{oboe, om}). 
-In this paper, we refer these concepts as OM concepts. 
-The canonical concepts include {\em observation}, {\em measurement},
-{\em characteristic}, {\em standard}, {\em protocol} or ({\em procedure}). 
-For example, {\bf add an example to illustrate these concepts. }
-
-%Introduce annotation, Motivation to use annotation
-Generally, every scientific dataset goes into the data repository with
-some metadata, e.g., Dublin core metadata\cite{***}, Darwin core metadata\cite{***}. 
-However, all these metadata are at the dataset level, they did not
-provide enough information for the data {\em content} inside the
-dataset. 
-To better make use of the data content in the repository, more systems
-are embracing the ideas of having metadata on the data content. 
-E.g., Some systems \cite{tdar} provides a mechanism to collect the
-column/attribute level metadata, some system \cite{semtools} uses
-{\em annotation} to add more semantic information to the data
-content. In this work, we follow the terminology in \cite{semtools} and
-use the term {\em annotation} to distinguish the 
-metadata on the data content from those on data objects at a coarser
-granularity. 
-%How to represent the annotation and the symbols used
-We propose tools for scientists to provide annotations to scientific
-observational data. 
-\from{HP}{Put a screen-dump of annotation.}
-
-%Queries to such data and challenges to answer such queries.
-The internal data structure of such scientific data repository is
-generally unknown to a user who wants to pose a query to such scientific data
-repository and get some useful information. 
-So, it is unrealistic for a him/her to formulate a query based on the 
-underlying data structure of the data.  
-The easiest query they can pose to such data repository are keyword-query. 
-Such keyword style query works well for the dataset metadata
-(e.g., contributor information, dataset description, etc) using the
-current Information Retrieval (IR) techniques. However, it
-is far from enough to be used to search the data content because the
-data content need to be interpreted with semantics. 
-It is important in the scientific domain to have typical
-kinds of queries. One fact we can use is that the domain
-scientists know well the widely used OM concepts as discussed above. 
-So, naturally, when searching such scientific data, they can provide
-the concept information to restrain their queries and to find data sets
-related to such OM terminologies. 
-Given the dataset in Table \ref{tb:dataset}. 
-People may ask the following observation and measurement (OM) queries.
-
-\begin{example}\label{eg:query}
-Simple and summarization queries: 
-\begin{itemize}
-\item $Q_1$: Give me the data sets that contain species ``Picea rubens'' observations.
-%$Q_2$: Give me the data sets that have measurements on ``area''  characteristics. 
-\item $Q_2$: Give me the data sets that contain species ``Picea rubens''
-  observations in ``California''. 
-\item $Q_3$: Give me the data sets that contain at least five distinct
-  ``Picea rubens'' observations.
-\item $Q_4$: Give me the data sets that have trees with average ``height''
-  than 20.0 in ``California''. 
-% Shawn: an example on summarize one measure with respect to another.
-% This could take two forms: one where the context forms the groups
-% (like by year), and another where the two measurements are within the
-% same context (like finding correlations).
-\end{itemize}
-\end{example}
-\from{HP}{Challenges to answer such queries: uniqueness}
-
-\from{HP}{Current data integration effort}
-
-\from{HP}{Our effort in querying observational data: introduce OBOE}
-
-\from{HP}{Need re-organization. The description in the
-  following several sections may be moved to here later.}
-
-\subsection{Contribution and paper organization}
-Contributions of this work:
-\begin{itemize}
-\item We formulate a canonical set of queries over observation and
-  measurement scientific data repository. Such formal queries can
-  formalize most OM concepts related searches. 
-\item We propose three methods to evaluate such OM
-  queries. \from{HP}{Elaborate the three methods.}
-\end{itemize}
-
-This paper is organized as follows. 
-Section \ref{sec:relatedwork} reviews works that are related to this
-research. Section \ref{sec:dataquery} formalizes the data model and the
-queries that people are interested to ask.
-
-\section{Related work}\label{sec:relatedwork}
-
-\from{HP}{This will come after the real problem definition and the
-  solution. The description in the following several sections may be
-  moved to here later.}
-
-\section{Data model, annotation and query}\label{sec:dataquery}
-
-In this section, we first illustrate the data model. Then, we
-formalize the queries that scientists in this domain tend to ask.
-
-\subsection{Data model}\label{sec:datamodel}
-
-%Data model: Data, attribute, data value set
-When a scientist contributes data into an integrated data repository,
-a widely accepted way is to convert each dataset to a data
-table\cite{tdar} or treated as a separate object entity\cite{knb}. 
-{\bf (HP: is it really widely used? any other system uses this
-  way?. Add more citations here.)}
-In using this method, the database contains metadata of each
-dataset and the definition of the data table (e.g., attribute/column
-name, attribute type, etc).
-
-%HP: 
-%An alternative way to organize the data is to store them in one big
-%data table with three columns data\_file, attribute\_name,
-%attribute\_value. The problem with this structure is that the
-%attribute\_value are different types.
-In our work, we focus on querying this kind of data repository (or
-databases). To formalize the scenario, we use $D$ to denote the set of data
-tables in the data repository and $d$ to refer to a specific data
-table. Each data table $d$ contains metadata about the attribute
-definition $Attr_{d}$. Sometimes, we also use {\em column} to refer to an
-attribute. 
-Given one data table $d$, an attribute $attr_i \in Attr_{d}$ or column index $i$, 
-$d[attr_i]$ or $d[i]$ represents the set of values for the attribute
-$attr_i$ or for the $i$-th column.
-%HP: Implementation detail (no need to put it in the paper): 
-%When we put the data to data tables, we add a unique record id for
-%each row in the data. This is used to get the unique record id.
-
-\subsection{Annotation}\label{sec:annotation}
-
-%What kinds of annotation information is needed. 
-We use $A$ to denote the annotation of one dataset. 
-Internally, we keep the following information for the
-annotation.
-
-For a data table $d$ with annotation $A$, the system keeps the
-corresponding information between them in $AD$ tables. 
-Besides it, we also have four main concepts to describe: 
-observation type (OT), measurement type (MT), and
-context type (CT), and the mapping (Map) from the measurement type to
-resource attributes.  
-So, the annotation contains the following information.
-
-\begin{itemize}
-\item $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$ to keep the information
-  of data sets $d_{id}$-s and their related annotations
-  $A_{id}$. Here, $d_{meta}$ is an abstract attribute for all the
-  other metadata of $d_{id}$. 
-\item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
-  describe an observation type. It denotes on which
-  entity type (object in the real world) the observation is made. 
-Very often, more than one observation can be made on one entity. Thus,{\em
-  isDistinct} is used to denote whether the same value of key
-measurements of an obbservation types can uniquely identify one
-observation or not. 
-\item $MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
-  Cha, \cdots)\}$. Generally, $MT$ contains information about
-  characteristic ($Cha$), Standard, Protocol and Precision. We do not include them
-  here to make the description clearer. Here, {\em isKey} is used to denote
-  whether one measurement type is the key measurement for the
-  observation type $OT_{id}$ or not. 
-\item $CT$ = $\{(\underline{A_{id}, ct_{id}}, ot_{id}, cot_{id}, Rel,
-  isIdentify)\}$ where $cot$ represents the context observation
-  type. 
-\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, mapCond, mapVal},
-  Val\}$. This structure denotes the correspondences between an
-  measurement type ($mt_{id}$ and an attribute $attr_i$ in an dataset
-  $d_{id}$ for a given condition $mapCond$. 
-\end{itemize}
-
-We use the following example to illustrate the concepts in the
-annotation.
-
-\begin{example}
-$OT$ = \{($A_1, OT_1$, tree, false),($A_1, OT_2$, GeoSpot, true) \}
-  denotes that annotation $A_1$ are for two observation types. One
-  type $OT_1$ is on real world entity ``tree'' and the values of its key measurements
-  do not uniquely identify a distinct observation. The other type
-  $OT_2$ is for real world entity ``GeoSpot'' (i.e., geospatial
-  location) and its key measurement types can identify it's unique
-  observations. \from{HP}{Introduce key measurements before this.}
-
-$MT$ =\{($A_1, MT_1, OT_1$, Species, true), 
-($A_1, MT_2, OT_1$, SpecNo, true), 
-($A_1, MT_3, OT_1$, Height, false),
-($A_1, MT_4, OT_2$, Plot\_State, true), 
-($A_1, MT_5, OT_2$, Plot\_Area, false)\} represents that $OT_1$
-(for ``tree'') has three measurement types ($MT_1, MT_2$, $MT_3$) for characteristics
-``Species'', ``SpecNo'', and ``Height''. The fist two measurement
-  types together form the key measurement type for $OT_1$. 
-$OT_2$ (for ``GeoSpot'') has one key measurement type $MT_4$ for
-  characteristic ``Plot\_State'' and another measurement type $MT_5$
-  for characteristic ``Plot\_Area''.
-
-$CT$ = \{($A_1$,$OT_1$, $OT_2$, within, true)\}. It shows that
-observation of a {\em tree} is within the context of a geo-spatial
-location. And this context is used to identify the uniqueness of the
-observation. E.g., a tree with species name ``piru'' and species no
-$1$ in California
-%$(34^{\circ}8'3N,118^{\circ}14'37W)$ (Los Angeles in California) 
-is different from a tree with 
-the same species name ``piru'' at Oregon. 
-%location $36^{\circ}10'30N,115^{\circ}8'11W$ (Las Vegas in
-%California).
-
-%HP coordinates material: 
-%http://www.traveljournals.net/explore/united_states/map/u1662328/los_angeles.html
-%http://en.wikipedia.org/wiki/Las_Vegas,_Nevada
-
-$Map$ = \{($A_1$, $MT_1$, ``code'', ``eq `piru''', ``Picea rubens''),
-%($A_1$, $MT_1$, ``code'', ``eq `abba''', ``Abies balsamea''),
-%($A_1$, $MT_1$, ``code'', ``eq `capo''', ``California poppy''),
-($A_1$, $MT_2$,``no'', null,), 
-($A_1$, $MT_3$, ``plt'', null,), 
-($A_1$, $MT_4$,``area'', null,\}. 
-The first mapping rules maps ``code''
-attribute to the measurement type $MT_1$ (for ``Species''
-characteristic) and change the value to ``Picea rubens'' if the code
-is ``piru''. The meaning of the other mapping rules are very
-obvious here, so we skip the details. 
-\end{example}
-
-\subsection{Observation and measurement (OM) query}
-Above we defined the data model and the annotation. In this
-sub-section, we formalize the queries that scientists are interested
-in. We denote the queries on the observational data as {\em Observation
-and measurement (OM) query}.
-
-As discussed in Section \ref{sec:intro}, it is generally impossible for a
-scientist to formulate an OM query using the internal data structure
-of a dataset because such data structures differ from contributors to
-contributors. More realistically, they can pose queries
-(Example \ref{eg:query}) related to key OM concepts, e.g., {\em
-  observation, measurement,characteristic, and standard}, and these
-OM concepts satisfy given keyword and context conditions.
-
-\begin{definition}[Basic OM query]\label{def:basic_omq}
-A basic observation and measurement query is defined as 
-\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]
-Here, 
-\begin{itemize}
-\item {\tt OMconcept} refers to one of the major OM terminologies. 
-\item 
-$\langle${\tt cond}$\rangle$ is in the form of 
-{\tt f([distinct entity.]measurement) op value} 
-where {\tt op} is a basic operator from $\{=, \neq, >, <\}$ 
-and {\tt f} is an
-aggregation function from $\{sum, avg, count, min, max\}$. 
-{\tt distinct} and {\tt entity} are optional. 
-\end{itemize}
-\from{HP}{The symbol fonts are a little bit messy...need to
-  distinguish between the keywords and variables.}
-\end{definition}
-
-\begin{definition}[Result of OM query]\label{def:qresult}
-The result of a query $Q$ is a set of data objects (e.g., data tables) 
-\{$d|d\in D \wedge d~satisfies~\langle ${\tt cond} $\rangle$\}. 
-For each of such result $d$, we use $d~s.t.~Q$ to denote that
-an data object $d$ satisfies the query. 
-\end{definition}
-
-Based on the different $\langle${\tt cond}$\rangle$ definitions,
-``$d~satisfies~\langle${\tt cond}$\rangle$''
-is translated differently. 
-E.g., if $\langle${\tt cond}$\rangle$ is defined that the value of ``{\em area}'' need to be
-smaller than $3.0$, then, ``$d~satisfies~{\tt cond}$'' is translated to
-``$dv<3.0|dv\in d[area]$''.
-
-Definition \ref{def:basic_omq} can formulate queries on
-one specific observation and measurement. 
-In real application, we need to consider the context relationship. 
-\from{HP}{E.g., etc.}
-So, we generalize the basic OM query to contextualized OM query as
-follows.
-
-\begin{definition}[Contextualized OM query]\label{def:context_omq} A
-  contextualized OM query is defined as:
-\[CQ ::=Q_i \wedge {\tt context}(Q_i.{\tt OMconcept}, Q_j.{\tt
-  OMconcept})\wedge Q_j\]
-Here, $Q_i, Q_j$ are basic queries.
-\end{definition}
-
-For example, the queries in Example \ref{eg:query} can be formalized
-as the following formal OM queries.
-
-%\begin{example}
-\begin{itemize}
-\item $Q_1$ = $Observation: \langle Species='Picea~rubens'\rangle$
-\item $Q_2$ = $Observation_1: \langle Species='Picea~rubens'\rangle$\\
-  $\wedge IN(Observation_1, Observation_2$) \\
-$\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
-\item $Q_3$ = $Observation: \langle Species='Picea~rubens'$\\
-$\wedge count(distinct~Species)\geq 5\rangle$
-\item $Q_4$ = $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
-$\wedge IN(Observation_1, Observation_2)$\\
-$\wedge Observation_2: \langle State='California'\rangle$\\
-%Give me the data sets that have trees with average ``height''
-%  than 20.0 in ``California''
-\end{itemize}
-
-%\end{example}
-
-%\from{HP}{Add experiement: show the error rate (e.g., precision, recall, (or
-%  directly $F_1$) of this method.? not needed?}
-
 \section{Querying annotated and integrated OM data}
@@
 FROM $AD, MI~as~mi, MT~as~mt[$, {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
+$[$WHERE    {\em table join condition}, \\
+$\qquad\qquad$ AND {\em selection condition}$]$,\\
 $[$GROUP BY {\em distinct requirement} $]$\\
@@
 HAVING COUNT(*)$>$5; 
-}
 \end{tabular}
@@
 $\qquad MI~as~cmi, MT~as~cmt[,$ {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
+$[$WHERE    {\em table join condition}, \\
+$\qquad\qquad$ AND {\em selection condition}$]$,\\
 $[$GROUP BY {\em distinct requirement} $]$\\
@@
 \begin{tabular}{l}
-SELECT $ct.ct_{id} as tmp\_ct_\id$\\
+SELECT $ct.ct_{id} as tmp\_ct\_id$\\
 FROM OT as ot, OT as cot, CT as ct\\
@@
 \begin{tabular}{l}
-SELECT DISTINCT AD.d_{id}\\
+SELECT DISTINCT $AD.d_{id}$\\
 FROM $AD, CI~as~ci, MI~as~mi, MT~as~mt,$ \\
@@
 \begin{tabular}{l}
-SELECT DISTINCT AD.d_{id} \\
+SELECT DISTINCT $AD.d_{id}$ \\
 FROM $AD, CI~as~ci, MI~as~mi, MI~as~cmi,$ \\
@@
 \from{HP}{Let's see how this can work. }
-
-\section{Experiments}
-
-Synthetic data generator.
-
-Algorithm to materialize DB. 
-Report the time and space.
-
-Synthetic query generator.
-
-Test algorithm to perform query over materiazed DB.
-Test algorithm to perform query-rewriting over materiazed DB.
-
-Test algorithm for half-materialized data.
-
-\section{Conclusion}
-
-\bibliographystyle{IEEEtranS}
-\bibliography{IEEEabrv,DataIntegration}
-
-\end{document}

### 2010-05-13 docs/algdraft/DataIntegration/modeldef.tex
Decompose the draft to several parts so that we can work on it collaboratively.
Until further discussion, 
from now on, 
Huiping will work on method.tex and Shawn probably will work on modeldef.tex

--- 2010-05-10 docs/algdraft/DataIntegration/DataIntegration.tex
@@
-\documentclass[conference]{IEEEtran}
-
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-
-\usepackage{algorithm}
-\usepackage{algorithmic}
- \renewcommand\floatpagefraction{0.99}
- \renewcommand\topfraction{0.99}
- \renewcommand\bottomfraction{0.99}
- \renewcommand\textfraction{.05}
- \setcounter{totalnumber}{5}
-
-\newtheorem{example}{\bf Example}[section]
-\newtheorem{definition}{\bf Definition}[section]
-\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
-
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-\begin{document}
-
-\title{Querying Integrated Scientific Observation and Measurement Data}
-
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-\author{\IEEEauthorblockN{Author1}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
-\and
-\IEEEauthorblockN{Author2}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{Author3}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678-2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
-
-\maketitle
-
-\begin{abstract}
-%\boldmath
-The abstract goes here.
-\end{abstract}
-
-% no keywords
-
-\section{Introduction}\label{sec:intro}
-In this work, we study the query processing over scientific
-observation and measurement data using OBOE
-model\cite{DBLP:conf/er/BowersMS08}. OBOE model is a conceptual model
-used to interpret observation and measurement data.
-
-\subsection{Background}
-
-%Application background
-In many scientific domains (e.g., ecology, hydrology, earth science,
-geology), people collect observational data. Such data
-record the observed value of some real world entity at some specific
-place and time. E.g., ecologists studying relationship between the 
-growth pattern and the treatments often need to record the tree
-heights. The collected data reflect the fact the tree height of a
-specific tree is 30.1in on May 1, 2009 and 30.3in on May 1, 2010.
-
-%Characteristics of scientific observational data: non-normalized
-Almost all such scientific data do not follow database higher normal
-forms. Generally, scientists have their way in interpreting their
-data, but they are not ready for any normalization process. 
-For example, Table \ref{tb:dataset} is a simplified but typical dataset collected by
-a scientist who study the growth pattern of trees. 
-Obviously, the ``plt'' of the first two rows is the same place, and
-the ``plt'' of the last two rows is the same. 
-Here, their area information is redundant. 
-In the real application, many columns have redundant information.
-
-%HP: this one has the coordinates, it's troublesome to explain. 
-%% \begin{table}[htb]
-%% \begin{center}
-%% \begin{tabular}{|l|l|l|l|l|}
-%% \hline
-%% code & no & ht & plt & area\\\hline
-%% piru & 1 & 35.8 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
-%% piru & 2 & 36.2 & $(34^{\circ}8'3N,118^{\circ}14'37W)$ & 4.0\\\hline
-%% piru & 1 & 25.7 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0 \\\hline
-%% %abba & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
-%% capo & 1 & 15.6 & $(36^{\circ}10'30N,115^{\circ}8'11W)$ & 3.0\\\hline
-%% \end{tabular}
-%% \end{center}
-%% \caption{Dataset}
-%% \label{tb:dataset}
-%% \end{table}
-
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-code & no & ht & plt & area\\\hline
-piru & 1 & 35.8 & California & 4.0\\\hline
-piru & 2 & 36.2 & California & 4.0\\\hline
-piru & 1 & 25.7 & Oregon & 3.0 \\\hline
-%capo & 1 & 7.8 & Oregon & 3.0\\\hline
-Oriental poppy& 1 & 7.8 & Oregon & 3.0\\\hline 
-\end{tabular}
-\end{center}
-\caption{Dataset}
-\label{tb:dataset}
-\end{table}
-%HP: California poppy info
-%Oregon iris: http://en.wikipedia.org/wiki/List_of_native_Oregon_plants
-%http://en.wikipedia.org/wiki/California_poppy
-
-%Introduce major concepts
-In scientific domains where people collect observation and measurement
-data, there are several commonly used and widely recognized canonical
-concepts(\cite{oboe, om}). 
-In this paper, we refer these concepts as OM concepts. 
-The canonical concepts include {\em observation}, {\em measurement},
-{\em characteristic}, {\em standard}, {\em protocol} or ({\em procedure}). 
-For example, {\bf add an example to illustrate these concepts. }
-
-%Introduce annotation, Motivation to use annotation
-Generally, every scientific dataset goes into the data repository with
-some metadata, e.g., Dublin core metadata\cite{***}, Darwin core metadata\cite{***}. 
-However, all these metadata are at the dataset level, they did not
-provide enough information for the data {\em content} inside the
-dataset. 
-To better make use of the data content in the repository, more systems
-are embracing the ideas of having metadata on the data content. 
-E.g., Some systems \cite{tdar} provides a mechanism to collect the
-column/attribute level metadata, some system \cite{semtools} uses
-{\em annotation} to add more semantic information to the data
-content. In this work, we follow the terminology in \cite{semtools} and
-use the term {\em annotation} to distinguish the 
-metadata on the data content from those on data objects at a coarser
-granularity. 
-%How to represent the annotation and the symbols used
-We propose tools for scientists to provide annotations to scientific
-observational data. 
-\from{HP}{Put a screen-dump of annotation.}
-
-%Queries to such data and challenges to answer such queries.
-The internal data structure of such scientific data repository is
-generally unknown to a user who wants to pose a query to such scientific data
-repository and get some useful information. 
-So, it is unrealistic for a him/her to formulate a query based on the 
-underlying data structure of the data.  
-The easiest query they can pose to such data repository are keyword-query. 
-Such keyword style query works well for the dataset metadata
-(e.g., contributor information, dataset description, etc) using the
-current Information Retrieval (IR) techniques. However, it
-is far from enough to be used to search the data content because the
-data content need to be interpreted with semantics. 
-It is important in the scientific domain to have typical
-kinds of queries. One fact we can use is that the domain
-scientists know well the widely used OM concepts as discussed above. 
-So, naturally, when searching such scientific data, they can provide
-the concept information to restrain their queries and to find data sets
-related to such OM terminologies. 
-Given the dataset in Table \ref{tb:dataset}. 
-People may ask the following observation and measurement (OM) queries.
-
-\begin{example}\label{eg:query}
-Simple and summarization queries: 
-\begin{itemize}
-\item $Q_1$: Give me the data sets that contain species ``Picea rubens'' observations.
-%$Q_2$: Give me the data sets that have measurements on ``area''  characteristics. 
-\item $Q_2$: Give me the data sets that contain species ``Picea rubens''
-  observations in ``California''. 
-\item $Q_3$: Give me the data sets that contain at least five distinct
-  ``Picea rubens'' observations.
-\item $Q_4$: Give me the data sets that have trees with average ``height''
-  than 20.0 in ``California''. 
-% Shawn: an example on summarize one measure with respect to another.
-% This could take two forms: one where the context forms the groups
-% (like by year), and another where the two measurements are within the
-% same context (like finding correlations).
-\end{itemize}
-\end{example}
-\from{HP}{Challenges to answer such queries: uniqueness}
-
-\from{HP}{Current data integration effort}
-
-\from{HP}{Our effort in querying observational data: introduce OBOE}
-
-\from{HP}{Need re-organization. The description in the
-  following several sections may be moved to here later.}
-
-\subsection{Contribution and paper organization}
-Contributions of this work:
-\begin{itemize}
-\item We formulate a canonical set of queries over observation and
-  measurement scientific data repository. Such formal queries can
-  formalize most OM concepts related searches. 
-\item We propose three methods to evaluate such OM
-  queries. \from{HP}{Elaborate the three methods.}
-\end{itemize}
-
-This paper is organized as follows. 
-Section \ref{sec:relatedwork} reviews works that are related to this
-research. Section \ref{sec:dataquery} formalizes the data model and the
-queries that people are interested to ask.
-
-\section{Related work}\label{sec:relatedwork}
-
-\from{HP}{This will come after the real problem definition and the
-  solution. The description in the following several sections may be
-  moved to here later.}
-
 \section{Data model, annotation and query}\label{sec:dataquery}
@@
 A basic observation and measurement query is defined as 
-\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle}\]
+\[Q::={\tt OMconcept:}\langle{\tt cond}\rangle\}\]
 Here, 
@@
 %  directly $F_1$) of this method.? not needed?}
-
-\section{Querying annotated and integrated OM data}
-
-In this section, we propose several methods that can be used to
-answer OM queries over scientific observational data with
-annotations.
-
-%\subsection{Naive method}\label{sec:naive}
-A very direct way to search such data is to extract all the content in
-the data cells and index them. This way the system may help answer
-very simple queries with the help of some thesaurus. 
-For example, for $Q_1$ in Example \ref{eg:query}, as long as we can
-translate the abbreviation ``piru'' in the dataset to ``Picea rubens''
-based on some predefined thesaurus, we can find the results. 
-However, even for this kind of simple query, if we want to find
-observation of ``California poppy'', an approximate match may return
-the dataset in Table \ref{tb:dataset} as a result because it contains
-both {\em California} and {\em Poppy}.
-If the search problem caused by the compound words can be alleviated using some existing
-techniques \cite{***}, %citation on compound word querying. 
-it is still almost impossible to directly use such IR-style system to answer a query
-in the form of $Q_3$ and $Q_4$. 
-This naive method cannot be directly used answer the desired queries. It is because it
-fails to catch the semantics of the context and uniqueness of observations in the data. 
-In what follows, we utilize the annotation information to get
-more semantic support, and provide new strategies to answer the
-queries.
-
-As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
-in querying integrated data. 
-One is to rewrite the original query to a series of new queries over
-the data; the other is to materialize the data to a consistent data
-model and then answer the query over the materialized database. 
-In what follows we work from these two directions to leverage the
-semantic information in answering such queries.
-
-\subsection{Query rewriting}\label{sec:queryrewrite}
-
-As described in Section \ref{sec:dataquery}, 
-an OM query is represented using 
-OM concept terminologies such as observation, 
-measurement, characteristic, etc. 
-To answer such queries from the original data model (i.e., dataset 
-metadata and translated data tables), 
-we need to {\em rewrite} the OM query over the real data tables. 
-In what follows, we first sketch the re-writing process. 
-Then, we detail the procedure in using the data model and structures. 
-Finally, we include the process for the more complicated
-cases (e.g., with distinct, aggregate).
-
-%Framework of query rewriting
-Roughly speaking, the query rewriting consists of two steps:
-\begin{itemize}
-\item From the given query, search the database metadata and
-  annotation information to find the relevant data tables and
-  attributes that need to be used to answer the given query. 
-\item Translate the given OM query to queries over the relevant data
-  tables. 
-\end{itemize}
-
-\from{HP}{todo:put a figure here to illustrate the steps}.
-
-%Elaborate the first step
-The first step is to map the OM concepts in the query to the real data
-structure. 
-With the annotation structures $OT$, $CT$, and $MT$, any given OM
-concepts can be linked to a measurement type $mt_{id}$. 
-Then, the annotation structure $Map$ can be used to bridge OM concepts and data sets. 
-Utilizing $Map$, we can find the attributes $attr_i$ and the
-annotation $A_{id}$).
-E.g., when the given concept is {\em measurement}, we can directly get
-its related attribute names and annotation id $A_{id}$. 
-In case that the given OM concept is a non-measurement concept, we can use
-other annotation structures to figure out the measurementt types. 
-E.g., if the OM concept is {\em characteristic} ``Species'', we can use
-$MT$ to get the measurement type id $mt_{id}$. 
-After we get $A_{id}$, we can quickly find the dataset id $d_{id}$
-using $AD$ metadata table (Section \ref{sec:annotation}). 
-Then, we get the relevant data tables and the needed attributes.
-
-%Elaborate the second step
-Let $D_{cand}$ be the set of relevant (or candidate) data tables and $d_{a_1} \cdots d_{a_m}$
-are the related attributes for table $d\in D_{cand}$. 
-The second step translates the original OM query to SQL queries over the data tables.
-
-In particular, for each candidate data table $d$,  
-the translation process works in the following rules.
-
-\begin{itemize}
-\item The SELECT clause always get the distinct $d_{id}$ and/or
-  record id $r_{id}$ (when it is needed);
-\item The FROM clause has table $d \in D_{cand}$;
-\item The WHERE clause contains all the function $f(\cdot)$ in the OM
-  query;
-\item When there is distinct constraint on entity or observation, the
-  GROUP BY clause contains of all the attributes corresponding to the
-  {\em key measurement types} of the entity or observation. we detail
-  all process of getting the key measurement types of an observation
-  later. 
-\item The aggregation requirement in $f(\cdot)$ is translated to
-  HAVING clause. 
-\end{itemize}
-
-Based on these rules, an OM query can be translated to the following
-SQL scripts.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l} 
-SELECT DISTINCT $d_{id}$\\
-FROM $d$\\
-WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
-$[$GROUP BY {\em distinct requirement}$]$\\
-$[$HAVING {\em aggregation condition}$]$;
-\end{tabular}
-}
-\vspace{0.1in}
-
-The above describes the process for the OM query. 
-In real application, people ask queries with distinct
-observation or entity constraints. Also, people ask queries about
-different observatitons using context. 
-The following elaborate the details in processing these two cases.
-
-%In detail,  how to deal with the distinct constraints.
-When a query with distinct constraints, e.g., find data sets with {\em distinct} observations or
-entities. To deal with this case, 
-we need to be able to figure out what distinguishes one observation
-from another. Here, one observation type's key measurement types can
-play this role well. 
-When an observation type is denoted with {\em distinct yes} and some
-of its measurement types are denoted with {\em key yes}, 
-the values of the measurements on the key measurement types can
-uniquely distinguish one observation from the other. 
-To implement this, in SQL, we can perform ``group by'' operations on
-the key measurement types of the observations.
-
-%In detail, how to deal with context
-When the given OM query has context, we need to consider two cases. 
-In the first case the context is not {\em identifying} to its
-observation types, i.e., the context value
-does not affect the uniqueness of one observation, this can be
-processed as the basic way. 
-When the context is denoted with {\em identifying yes}, it means the
-identity of one observation also depend on the contextual
-observation. 
-In this case, we need to figure out the all the key measurement types
-of one observation, then the ``GROUP BY'' operation should be on all
-these key measurement types. 
-\from{HP}{Put an example here.}
-In this case, there is no need to change the SQL scripts.
-the only change are the relevant attribute set $d_{a_1} \cdots
-d_{a_m}$, which corresponds to not only the measurement types that
-directly related to the observation, but also the measurement types
-that are in the identifying context chain of the observation.
-
-\from{HP}{Put a specific routine to get key measurements of
-  observation and/or entity.}
-
-\begin{example}\label{eg:query_reqerite_q1}
-{\bf TODO: refine this example} 
-Take $Q_1$ ($Observation: \langle Species='Picea~rubens'\rangle$) as
-example. 
-The {\tt OMconcept} is ``observation'' and {\tt cond} is ``{\em species = `Picea rubens'}'';
-
-From the annotation $MT$, we can find the measurement type
-$m_3$ with the characteristic {\em species}. 
-From the mapping $Map$, we further find that the attribute {\em spp} that the
-measurement type $m_3$ is mapped to.
-
-The second step is to find the data sets which really have this value. 
-For this one, we find the data table, and do a selection on the table
-content.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l} 
-SELECT DISTINCT $d_{id}$\\
-FROM $d$\\
-WHERE $spp = 'piru'$
-\end{tabular}
-}
-\vspace{0.1in}
-
-\end{example}
-
-\begin{example}
-Take $Q_3$ ($Observation:$\\
-$\langle Species='Picea~rubens'$$\wedge count(distinct~Species)\geq
-5\rangle$) as an example.
-
-\from{HP}{Refine\\
-For the first step, we find the data tables that contain the needed concepts. 
-In each table, we need to figure out the distinct observations. 
-From $m_i$ $obstype_i$, all the other measurements that are of
-$obstype_i$. 
-Group by the key measurements of $obstype_i$ if distinct yes.
-If this observation type does not have any context, then the key
-measurements are all the measurements directly defined under this
-observatioin type. Otherwise, i.e., this observation type has context,
-then the key measurements are all the measurements in the context
-chain. }
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l} 
-SELECT DISTINCT $d_{id}$\\
-FROM $d$\\
-WHERE $spp = 'piru'$\\
-GROUP BY $d_{id}$, $spp$\\
-HAVING count($d_{id}$,$spp$) $\geq 5$\\
-\end{tabular}
-}
-\vspace{0.1in}
-
-\end{example}
-
-%Analysis of this method. 
-{\bf Cost analysis:} 
-From the above description, we can tell that the computation cost is to search the different data
-tables. This way, for each candidate data table, we need to send an
-SQL query to the server for evaluation purpose.
-If the number of candidate data tables are small. This method should
-work well. However, when we have a lot of candidate data tables, the
-efficiency may be affected. 
-In what follows, we introduce another strategy, which logically merge
-the data content in different tables and perform queries ober the
-materialized database.
-
-\subsection{Querying materialized database}\label{sec:matdb}
-
-The previous section shows the method to rewrite a given query. 
-However, as we analyzed, the computation cost may be very high when
-there are many candidate data tables. 
-In this section, we propose another strategy to answer the given OM
-query by making use of the existing optimization strategies of
-current DBMS.
-
-%rough idea of this strategy
-In this strategy, we do not directly work on the original data table
-$D$. On the other hand, we {\em materialize} the original data-sets to
-centralized concept tables with instances of entities, observations, measurements
-and contexts. Such concept tables are called core OBOE tables, and denoted as
-$OBOE.*$. 
-Then, the query process works on the materialized databases (i.e., the
-conceptual data tables).
-
-%structure of this sub-section
-In what follows, we first describe the structure of the core
-conceptual tables and present our algorithm to materialize the data into such
-concept instance tables in Section \ref{sec:materialilzedb}. Our
-algorithm considers all the distinct and context constraints. 
-Then, Section \ref{sec:query_materialilzedb} provides the detailed
-process of performing queries over materialized database.
-
-\subsubsection{Materialize data}\label{sec:materialilzedb}
-
-%Structure of conceptual instances
-In particular, we have the following concept instance tables. 
-\begin{itemize}
-%\item $EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
-\item $OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
-  all the observation instances where $ot_{id}$ is used to link to the
-  annotation information in $OT$; 
-\item $MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
-  mVal)\}$ for all the measurement instances where $mt_{id}$ is used to link to the
-  annotation information in $MT$; 
-\item $CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
-  all the context instances where $ct_{id}$ is used to link to the
-  annotation information in $CT$. 
-\end{itemize}
-
-\from{HP}{TODO: refine the writing in MaterializeDB and move them here.}
-
-\subsubsection{Querying materialized database}\label{sec:query_materialilzedb}
-
-Once the data are materialized using the annotation constraints, we
-can perform the OM query over them. 
-%In what follows, we describe how to use the materialized data to answer the query.
-
-\from{HP}{TODO:Formalize the translation of operators in the OM
-  query}.
-
-When the given query is a basic OM query, i.e., there is not context
-in the query. 
-Then, we can form a template query as the following;
-
-\begin{figure}[htb]
-{\tt
-\begin{tabular}{l}
-SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt[$, {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
-$[$GROUP BY {\em distinct requirement} $]$\\
-$[$HAVING   {\em aggregation condition} $]$;
-\end{tabular}
-}
-\caption{SQL template to answer OM using materialized DB}
-\label{fig:sql_matdb_basic_omq}
-\end{figure}
-
-In particular,
-
-\vspace{0.1in}
-\begin{tabular}{l}
-$table~join~condition=$\\
-%$\qquad (mi.oi_{id}=ci.oi_{id})~AND~(cmi.oi_{id}=ci.coi_{id})~AND$\\
-$\qquad (mi.mt_{id}=mt.mt_{id})~AND~(mt.A_{id}=AD.A_{id})$
-\end{tabular}
-\vspace{0.1in}
-
-For the very basic query (e.g., $Q_1$ in Example \ref{eg:query}), the
-SQL generally need only the first three clauses. 
-If the query has some restriction on the aggregated result, the {\tt
-  GROUP BY} and {\tt HAVING} clauses to deal with it.
-
-We use $Q_1$ and $Q_3$ to illustrate the query on the materialized
-database. 
-can be answered using the above
-template. 
-Example \ref{eg:query_reqerite_q1} shows how to deal with $Q_1$ using
-the query rewriting method.
-
-\begin{example} \label{eg:materialize_db_q1}
-Using the  materialized database,  we just need to search the MI table
-with the condition that {\em mVal=``Picea rubens''}.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt$\\ 
-WHERE $mt.Cha='Species'$ \\
-$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
-$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
-$\qquad\quad$ AND $mi.A_{id}=AD.A_{id}$; 
-%\#Join the MI and MT tables
-\end{tabular}
-}
-\vspace{0.1in}
-
-For $Q_3$, which asks for data sets that contain at
-least five distinct ``Picea rubens'' observations.
-The SQL query can be written as:
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT distinct $AD.d_{id}$\\
-FROM $AD, MI~as~mi, MT~as~mt$\\
-WHERE $mt.Cha='Species'$ \\
-$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
-$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
-$\qquad\quad$ AND $mt.A_{id}=AD.A_{id}$\\
-GROUP BY $d_{id},oi_{id}$\\
-HAVING COUNT(*)$>$5; 
-}
-\end{tabular}
-}
-\vspace{0.1in}
-\end{example}
-
-For this kind of query, the computation cost is the selection cost of the
-measurement instance (MI) table. 
-Compared with the query rewriting strategy, which need to pose the
-query over multiple data tables, this query should answer the queries
-more efficiently.
-
-Figure \ref{fig:sql_matdb_basic_omq} shows the SQL template to process
-the basic OM query without context. In case the OM query has context,
-we need to add the context into consideration.
-
-\begin{enumerate}
-\item The first step is similar to that in the basic OM query
-  processing. Besides this, we need to figure out what are the concept
-  obserbationi types of the needed observations.  
-\item \label{step:singlesql} Form an SQL for the observation and every of its context
-  observations. 
-\item Intersect all the SQLs from Step \ref{step:singlesql}). 
-\end{enumerate}
-
-\begin{figure}[htb]
-{\tt
-\begin{tabular}{l}
-SELECT $[$DISTINCT$]$ $AD.d_{id}$\\
-FROM $AD, CI~as~ci,MI~as~mi, MT~as~mt$\\
-$\qquad MI~as~cmi, MT~as~cmt[,$ {\em other tables}$]$\\
-$[$WHERE    {\em table join condition, \\
-$\qquad\qquad$ AND selection condition}$]$,\\
-$[$GROUP BY {\em distinct requirement} $]$\\
-$[$HAVING   {\em aggregation condition} $]$;
-\end{tabular}
-}
-\caption{SQL template to answer contextualized OM using materialized DB}
-\label{fig:sql_matdb_context_omq}
-\end{figure}
-
-In this case, the table join condition is much more complex since we
-need to use the context instance table $ci$ to connect the
-observation and context observation table. 
-In particular,
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-$table~join~condition=$\\
-$\qquad (mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$
-\end{tabular}
-}
-\vspace{0.1in}
-
-{\bf Execution cost analysis:} From the join condition, we can see that the expensive cost comes from
-the join operation between the measurement instance, context
-measurement instance, and context instance table. 
-As we analyzed in Section \ref{sec:materialilzedb}, the two tables with
-growing space are measurement instance table and context instance
-table.
-
-{\bf Example:} When we are given contextualized query, e.g., $Q_2$, which asks for
-the datasets that contain species ``Picea rubens'' observations in
-``California''.
-
-In the first step, we get the requirement is on Characteristic
-``Species'' with value ($mVal$) ``Picea rubens'' and observation has context {\em in}
-another observation for ``State'' of value ``California''.
-
-%Using materializedDB, we can answer the query this way
-%First, from the Characteristic ``Species'', we can get  --> mlabel, observation type
-
-%select mtype, olabel from measurement\_type  \\
-%where chracteristic = 'Species';  \\
-In the first step, we need to find the context type id $tmp\_ct\_id$
-that satisfy the context relationship.
-
-{\tt 
-\begin{tabular}{l}
-SELECT $ct.ct_{id} as tmp\_ct_\id$\\
-FROM OT as ot, OT as cot, CT as ct\\
-WHERE $ct.ot_{id} = ot.ot_{id}$ AND \\
-      $\qquad ct.cot_{id}= cot.ot_{id}$ AND \\
-      $\qquad ct.Rel='IN'$ AND $ot.et='tree'$;
-\end{tabular}
-}
-
-Then, the next step is to find the distinct data sets.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT AD.d_{id}\\
-FROM $AD, CI~as~ci, MI~as~mi, MT~as~mt,$ \\
-$\quad\qquad MI~as~cmi, MT~as~cmt,$ \\
-WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$\\
-$\qquad$ AND $mi.mVal=`Picea rubens'$ \\
-$\qquad$ AND $cmi.mVal='California'$\\
-$\qquad$ AND $mt.Cha = 'Species'$\\
-$\qquad$ AND $cmt.Cha=`State'$\\
-$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
-\end{tabular}
-}
-\vspace{0.1in}
-
-%Use query rewriting, what I can do? 
-%It's the same till to getting context\_olabel\_set.
-
-%Get the asked condition columns and the context columns,
-
-%select record\_id from data\_tables
-%where condition column 1='value1' and (context column1 = 'context
-%value' or context column2 = `context value', etc.);
-
-Now let's think about the query $Q_4$ with aggregation and context
-requirement. 
-% Give me the data sets that have trees with average ``height''
-%  than 20.0 in ``California''.
-$Q_4 = Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
-$\wedge IN(Observation_1, Observation_2)$\\
-$\wedge Observation_2: \langle State='California'\rangle$
-
-Also, in the first step, we discover the context type id $tmp\_ct\_id$
-that satisfy the context relationship.
-
-\vspace{0.1in}
-{\tt 
-\begin{tabular}{l}
-SELECT DISTINCT AD.d_{id} \\
-FROM $AD, CI~as~ci, MI~as~mi, MI~as~cmi,$ \\
-$\quad\qquad MT~as~mt, MT~as~cmt,$ \\
-WHERE $(mi.oi_{id}=ci.oi_{id})$ AND $(cmi.oi_{id}=ci.coi_{id})$ AND\\
-$\qquad (mi.mt_{id}=mt.mt_{id})$ AND $(cmi.mt_{id}=cmt.mt_{id})$ AND\\
-$\qquad (mt.A_{id}=AD.A_{id})$\\
-$\qquad$ AND $mt.Cha=`height'$\\
-$\qquad$ AND $cmt.Cha=`State'$\\
-$\qquad$ AND $cmi.mVal=`California'$\\
-$\qquad$ AND $ci.ct_{id} = tmp\_ct\_id)$\\
-GROUP BY $AD.d_{id}, oi_{id}$\\
-HAVING $avg(mi.mVal)>5.0$;
-\end{tabular}
-}
-\vspace{0.1in}
-
-Note that, in the above example, there is only one context.
-In the real application, there may be multiple context observations,
-or even context chains. In this case, we need to get {\em all} the
-context and get the intersection of the results.
-
-\subsection{Query partially materialized database}
-As we analyzed in the first two sub-sections, 
-the query-rewriting suffers the multiple queries posed on different
-candidate dataset. The querying of materialized database suffers from
-the multiple self-join operation of the measurement instance (MI)
-table, which has the most amount of information.
-
-In this section, we propose to materialize only the measurement
-instance and also put the measurement instance and context measurement
-instances together in the same table to avoid the expensive
-computation cost.
-
-\from{HP}{Let's see how this can work. }
-
-\section{Experiments}
-
-Synthetic data generator.
-
-Algorithm to materialize DB. 
-Report the time and space.
-
-Synthetic query generator.
-
-Test algorithm to perform query over materiazed DB.
-Test algorithm to perform query-rewriting over materiazed DB.
-
-Test algorithm for half-materialized data.
-
-\section{Conclusion}
-
-\bibliographystyle{IEEEtranS}
-\bibliography{IEEEabrv,DataIntegration}
-
-\end{document}

### 2010-06-07 docs/algdraft/DataIntegration/method.tex
update the algorithm with a framework (to finish)

@@
 
-As we analyzed in Section \ref{sec:relatedwork}, there are two extrems
+As we analyzed in Section \ref{sec:relatedwork}, there are two extremes
 in querying integrated data.
@@
 \begin{itemize}
-\item From the given query, search the database metadata and
+\item From the given query with entity, characteristic information, search the database metadata and
   annotation information to find the relevant data tables and
@@
 \item Translate the given OM query to queries over the relevant data
-  tables. 
+  tables. To do this translation, we decompose an OM query to have several basic component operators. 
+  \begin{itemize}
+  \item Basic OM query with one entity type. While the characteristic condition may be conjunctive or disjunctive, it may also has aggregation. E.g., $A(Cha1>25.0 && Count(Cha)>3, Height>2.0)$
+  \item Context OM chain refer to a set of queries having ``HasContext'' relationship. Obviously, these queries have different entity condition. E.g., $Ent1(...) --> Ent2(), Ent1()-->Ent3()$
+  \item Disjunctive OM query with multiple entity types. $Ent1(...), Ent2(), Ent3(...)$.
+  \end{itemize}
+
 \end{itemize}
@@
 In case that the given OM concept is a non-measurement concept, we can use
-other annotation structures to figure out the measurementt types. 
+other annotation structures to figure out the measurement types.
 E.g., if the OM concept is {\em characteristic} ``Species'', we can use
@@
 are the related attributes for table $d\in D_{cand}$.
+
 The second step translates the original OM query to SQL queries over the data tables.
+From now on, we assume that we know the attribute already for all the needed characteristics.
+
+To translate the given query:
+ Group the basic queries (with only one entity type) to list of context chains.
+
+\begin{algorithm}[htb]
+\begin{algorithmic}[1]
+%\begin{footnotesize}
+\STATE $Result$ := $\emptyset$; 
+\COMMENT{//From the given query, get the DNF of context queries.}
+\STATE $contextChainDNF$ := ${\bf GroupBasicQuery}(OMQ)$;
+\STATE
+\COMMENT{//Execute every DNF int the context chain set and union their results}
+\FOR{each context chain $oneContextQuery$ in $contextChainDNF$}
+	\STATE $result_{CNF}$ := $\emptyset$;
+    \COMMENT{//Execute every basic query in this CNF query and intersect their results}
+    \FOR{each basic query $OMBasic$ in $ContextQuery$}
+    \STATE $result_{basic}$ := $OMBasic.execute()$;
+    \STATE $result_{CNF}$ := $result_{CNF} \cap result_{basic}$;
+    \ENDFOR
+	\STATE $Result$ := $Result\cup result_{CNF}\$
+\ENDFOR
+\RETURN $RESULT$;
+%\end{footnotesize}
+\end{algorithmic}
+\caption{Execute(OMQ)}
+\label{alg:omq}
+\end{algorithm}
+
+\begin{figure}[htb]
+%\normalsize
+%\begin{minipage}{5in}
+%\sf
+\begin{scriptsize}
+\begin{tabbing}
+1166\= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= \kill
+{\bf Algorithm GroupBasicQuery}($OMQ$);\\
+1).\>$contextChainSet$ := $\emptyset$;\\
+\\
+\>//Put the basic queries in context into context chains
+2).\>{\bf for} (each basic context $OMBasic-->ContextOMBasic$ in $OMQ$)\\
+3).\>\>$added$ := false;
+4).\>\>{\bf for} (each set $oneContextChain \in contextChainDNF)\\
+5).\>\>\>{\bf if} (($OMBasic \in oneContextChain$) or $ContextOMBasic \in oneContextChain$))\\
+6).\>\>\>\> $oneContextChain$ = $oneContextChain \cup OMBasic \cup ContextOMBasic$\\
+7).\>\>\>\> $added$ :=true;\\
+8).\>\>\>\> {\bf break;}\\
+9).\>\>{\bf if}($added$==false)\\
+10).\>\>\>Create a set $newContextChain$ with two elements $OMBasic$ and $ContextOMBasic$;\\
+11).\>\>\>Add $newContextChain$ to $contextChainSet$;\\
+\\
+\>//Put the basic queries in context into context chains
+12).\>{\bf for} (each $OMBasic$ not in context of $OMQ$)\\
+13).\>\>Create a set $newContextChain$ with one element $OMBasic$\\
+14).\>\>Add $newContextChain$ to $contextChainSet$;\\
+15).\>{\bf return} $contextChainSet$;\\
+\end{tabbing}
+\end{scriptsize}
+%\end{minipage}
+\caption{GroupBasicQuery(OMQ)}
+\label{alg:omq}
+\end{figure}
 
@@
 the translation process works in the following rules.
+
+First, for the basic OM query with one entity type.
+
+Second, for the context chain query.
 

### 2010-06-08 docs/algdraft/DataIntegration/method.tex
update the write-up for the algorithm method.

--- 2010-05-13 docs/algdraft/DataIntegration/method.tex
@@
 \item Translate the given OM query to queries over the relevant data
-  tables. 
+  tables. To do this translation, we decompose an OM query to have several basic component operators.
+  \begin{itemize}
+  \item Basic OM query with one entity type. While the characteristic condition may be conjunctive or disjunctive, it may also has aggregation. E.g., $A(Cha1>25.0 \&\& Count(Cha)>3, Height>2.0)$
+  \item Context OM chain refer to a set of queries having ``HasContext'' relationship.
+Obviously, these queries have different entity condition.
+E.g., $Ent1(...) \rightarrow Ent2(), Ent1() \rightarrow Ent3()$
+  \item Disjunctive OM query with multiple entity types. $Ent1(...), Ent2(), Ent3(...)$.
+  \end{itemize}
+
 \end{itemize}
@@
 are the related attributes for table $d\in D_{cand}$.
+
 The second step translates the original OM query to SQL queries over the data tables.
+From now on, we assume that we know the attribute already for all the needed characteristics.
+
+To translate the given query:
+Figure \ref{alg:omq} shows the algorithm framework to deal with one OM query.
+\from{HP}{More description comes later.}
+
+\begin{figure}[htb]
+%\normalsize
+%\begin{minipage}{5in}
+%\sf
+\begin{scriptsize}
+\begin{tabbing}
+1166\= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= 16 \= \kill
+{\bf Algorithm Execute}(OMQ)\\
+1).\> $Result$ := $\emptyset$;\\
+2).\>//From the given query, get the DNF of context queries.\\
+3).\> $contextChainDNF$ := ${\bf GroupBasicQuery}(OMQ)$;\\
+4).\> \\
+5).\> //Execute every DNF int the context chain set and union their results\\
+6).\> {\bf for} each context chain $oneContextQuery$ in $contextChainDNF$\\
+7).\>\> $result_{CNF}$ := $\emptyset$;\\
+8).\>\>//Execute every basic query in this CNF query and intersect their results\\
+9).\>\>{\bf for} each basic query $OM_{b}$ in $ContextQuery$\\
+10).\>\>\>$result_{b}$ := ${\bf ExecuteBasicOM}(M_{b})$;\\
+11).\>\>\>$result_{CNF}$ := $result_{CNF} \cap result_{b}$;\\
+12).\>\>$Result$ := $Result\cup result_{CNF}$\\
+13).\>Return $Result$;\\
+\\
+{\bf Algorithm GroupBasicQuery}($OMQ$);\\
+1).\>$contextChainSet$ := $\emptyset$;\\
+2).\\
+3).\>//Put the basic queries in context into context chains\\
+4).\>{\bf for} (each basic context $OM_{b}\rightarrow COM_{b}$ in $OMQ$)\\
+5).\>\>$added$ := false;\\
+6).\>\>{\bf for} (each set $oneContextChain \in contextChainDNF$)\\
+7).\>\>\>{\bf if} (($OM_{b} \in oneContextChain$) or $OM_{b} \in oneContextChain$))\\
+8).\>\>\>\> $oneContextChain$ := $oneContextChain \cup OM_{b} \cup COM_{b}$\\
+9).\>\>\>\> $added$ :=true;\\
+10).\>\>\>\> {\bf break;}\\
+11).\>\>{\bf if}($added$==false)\\
+12).\>\>\>Create a set $newContextChain$ with two elements $OM_{b}$ and $COM_{b}$;\\
+13).\>\>\>Add $newContextChain$ to $contextChainSet$;\\
+14).\\
+15).\>//Put the basic queries without context into context chains\\
+16).\>{\bf for} (each $OM_{b}$ not in context of $OMQ$)\\
+17).\>\>Create a set $newContextChain$ with one element $OM_{b}$\\
+18).\>\>Add $newContextChain$ to $contextChainSet$;\\
+19).\>{\bf return} $contextChainSet$;
+\end{tabbing}
+\end{scriptsize}
+%\end{minipage}
+\caption{Algorithm to execute an OM query}
+\label{alg:omq}
+\end{figure}
 
@@
 the translation process works in the following rules.
+For an OMBasic query with one entity type.
+We need to do the
 
 \begin{itemize}
-\item The SELECT clause always get the distinct $d_{id}$ and/or
-  record id $r_{id}$ (when it is needed);
+\item The SELECT clause always get the distinct $d_{id}$ and the key attributes for this entity type;
 \item The FROM clause has table $d \in D_{cand}$;
-\item The WHERE clause contains all the function $f(\cdot)$ in the OM
+\item The WHERE clause contains all the non-aggregational conditions $f(\cdot)$ in the OM
   query;
 \item When there is distinct constraint on entity or observation, the
-  GROUP BY clause contains of all the attributes corresponding to the
-  {\em key measurement types} of the entity or observation. we detail
+  GROUP BY clause contains of all the key attributes corresponding to the of the entity. we detail
   all process of getting the key measurement types of an observation
@@
 
+For non-aggregate conditions,
+
 \vspace{0.1in}
@@
 \begin{tabular}{l}
-SELECT DISTINCT $d_{id}$\\
+SELECT DISTINCT $d_{id}, key~attribute~list$\\
 FROM $d$\\
-WHERE $f(d_{a_1} \cdots d_{a_m}) = true$\\
+WHERE $non-aggregate~conditions$)
+\end{tabular}
+}
+\vspace{0.1in}
+
+For each aggregate condition,
+
+\vspace{0.1in}
+{\tt
+\begin{tabular}{l}
+SELECT DISTINCT $d_{id}, key~attribute~list$\\
+FROM $d$\\
+WHERE $key~attribute_values$ IN(\\
+\hspace{0.6in}$non-aggregate~SQL$)\\
 $[$GROUP BY {\em distinct requirement}$]$\\
 $[$HAVING {\em aggregation condition}$]$;
+\end{tabular}
+}
+\vspace{0.1in}
+
+When there are more than one aggregation condition,
+the above SQL is formed for each aggregation condition
+and their results are intersected.
+
+\from{HP}{Output: record id or key attribute values.}
+If the output is record id, we further need to
+
+\vspace{0.1in}
+{\tt
+\begin{tabular}{l}
+SELECT $record\_id$\\
+FROM $d$\\
+WHERE ($key~measurement~value$ IN(\\
+\hspace{0.5in}SELECT DISTINCT $key~attribute~list$\\
+\hspace{0.5in}FROM ($results~from~aggregation functions$)
 \end{tabular}
@@
 
+For a non-aggregate condition,
+
+{\tt
+\begin{tabular}{l}
+SELECT DISTINCT did, record\_id, eid \\
+\hspace{0.5in}characteristic, mvalue\\
+FROM $mi, mt, oi, ei$\\
+$[$WHERE    {\em table join condition}, \\
+$\qquad\qquad$ AND {\em selection condition}$]$\\
+INTERSECT $other~measurement~cond$\\
+\end{tabular}
+}
+
+For an aggregate condition,
+
+{\tt
+\begin{tabular}{l}
+SELECT DISTINCT did, record\_id, eid\\
+FROM $oi$\\
+WHERE (did, eid) IN (\\
+\hspace{0.5in}SELECT did, eid\\
+\hspace{0.5in}FROM $non-aggregate~sql$\\
+\hspace{0.5in}WHERE $characteristic~condition$\\
+\hspace{0.5in}GROUP BY did, eid\\
+\hspace{0.5in}HAVING {\em aggregation condition};
+\end{tabular}
+}
 When the given query is a basic OM query, i.e., there is not context

### 2011-05-26 docs/pubs/EIM-2011/abstract.tex

\begin{abstract}
  The Data Manager paper from the last EIM included semantic
  foreshadowing in the 'future directions' section and this is a
  logical continuation of that work. Previously, we saw that the DM
  library enabled data integration tasks provided we knew exactly
  which data attributes from exactly which datasets were "compatible"
  with each other. Here "compatible" means not simply structurally
  similar (e.g. decimal numbers) but also conceptually similar
  (e.g. ocean water temperature in C). By including rich semantic
  annotations we initially show simple integration capabilities where
  the subject domain of our data corpus is relatively constrained
  (i.e. SBC-LTER) and the observational models represented are
  minimally variable (i.e. context excluded). Semantically-enahnced
  query capabilities facilitate this 'smart integration' in multiple
  ways. First, the corpus is refined through compound concept-based
  queries that exercise OBOE-compatible ontological subsumption
  hierarchies to locate roughly equivalent attributes across
  datasets. Additional query constraints are defined in relation to
  the precise observational model formally expressed in OBOE. Finally,
  the actual data values are interrogated so that only those that fall
  in the desired range are returned. From this resultant subset of
  disparate data objects we create a single synthetic data product
  that represents a 'materialization' of the observations. Technical
  discussion will include high-level descriptions of the
  infrastructure and tools employed with focus on our development in
  the rapidly evolving semantic frontier where scalability and best
  practices are particularly germane and equally elusive.
\end{abstract}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2011-05-26 docs/pubs/EIM-2011/intro.tex

--- 2011-05-26 docs/pubs/EIM-2011/abstract.tex
@@
-\begin{abstract}
-  The Data Manager paper from the last EIM included semantic
-  foreshadowing in the 'future directions' section and this is a
-  logical continuation of that work. Previously, we saw that the DM
-  library enabled data integration tasks provided we knew exactly
-  which data attributes from exactly which datasets were "compatible"
-  with each other. Here "compatible" means not simply structurally
-  similar (e.g. decimal numbers) but also conceptually similar
-  (e.g. ocean water temperature in C). By including rich semantic
-  annotations we initially show simple integration capabilities where
-  the subject domain of our data corpus is relatively constrained
-  (i.e. SBC-LTER) and the observational models represented are
-  minimally variable (i.e. context excluded). Semantically-enahnced
-  query capabilities facilitate this 'smart integration' in multiple
-  ways. First, the corpus is refined through compound concept-based
-  queries that exercise OBOE-compatible ontological subsumption
-  hierarchies to locate roughly equivalent attributes across
-  datasets. Additional query constraints are defined in relation to
-  the precise observational model formally expressed in OBOE. Finally,
-  the actual data values are interrogated so that only those that fall
-  in the desired range are returned. From this resultant subset of
-  disparate data objects we create a single synthetic data product
-  that represents a 'materialization' of the observations. Technical
-  discussion will include high-level descriptions of the
-  infrastructure and tools employed with focus on our development in
-  the rapidly evolving semantic frontier where scalability and best
-  practices are particularly germane and equally elusive.
-\end{abstract}
+\section{Introduction}
+\label{sec-introduction}
 

### 2011-06-01 docs/pubs/EIM-2011/framework.tex
draft framework section

\section{Framework}

Semtools has focused development effort on three components:
a Java library for accessing and manipulating OBOE ontology extensions and Annotations,
an Annotation plugin for the Morpho metadata editor, and query extensions for the Metacat data catalog.

\subsection{Semantic mediation API}
The SMS API includes ontology management features, annotation manipulation capabilities, and simple concept navigation and visualization widgets. OWL API primarily underlies the ontology management interface providing the RDF/OWL parsing and serialization as well as simple class and property exploration. The Pellet reasoner exposes inferred axioms and relationships that are not explicitly defined in the source ontologies. Inference is particularly powerful when utilizing OBOE Measurement 'templates' that express exactly what (Entity and Characteristic) can be observed and how (Standard and Protocol). 
Annotations are ultimately managed and stored in an opaque local relational database for scalible persistence and query performance; the overhead of in-memory Annotation exploration having proven to be prohibitively limiting for large sets of metadata.

\subsection{Morpho editor plugin}
The semantic editor plugin for Morpho provides a front-end to the SMS API and allows metadata editors to build up semantic Annotations for existing data package descriptions. The editor focuses on a simplictic fill-in-the-blank style interface with reusable, searchable, hierachical concept selection widgets. The optional plugin seamlessly integrates with a standard Morpho installation augmenting the existing features to provide semantic query capabilities for locating data packages, marking up data packages with semantic Annotations, and saving Annotations locally or to a shared repository where they can be discovered and explored by a broader audience.

\subsection{Metacat query extensions}
The semantic plugin for Metacat augements existing metadata storage and search by allowing Annotations to be saved and queried alongside the metadata and data that they annotate. In addition to traditional keyword and spatial search criteria, the Metacat plugin allows semantic criteria to be included where they may either increase query recall using term-expansion (i.e. traversing the class subsumption hierarchy) or refine the resultset by limiting matches to datasets that contain the specified observational model (e.g. combinations of OBOE-compatible Entity, Characteristic, or Protocol concepts). The observational model can be leveraged further by materializing the Annotation and data artifact (via the Data Manager Library) into a full OBOE model and inspecting the observational values themselves.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2011-06-01 docs/pubs/EIM-2011/annotation.tex
add more sections (for me)
annotations draft

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Annotation}
+
+Semantic Annotations describe the observation model contained in a data object.
+The Annotation serves as a template for constructing a fully fleshed OBOE model of the [usually tabular] data.
+Less formal descriptions of the data attributes are often included in traditional metadata formats like EML.
+while there is some overlap between these two mechanisms - particularly with respect to measurement standards or units - semantic concepts applied to each attribute have the benefit of placing the descriptive burden on the ontology from which the concepts are drawn.
+Because the Annotation represents a potentially subjective perspective on the data, they are stored independently; the former referencing the latter.
+
+The Annotation structure largely mirrors the core classes in OBOE.
+Observations are composed of Measurements of a specific Entity and are represented by a collection of Characteristics (usually just one) collected using a defined Protocol and Standard (i.e.
+unit).
+Tablular data objects map each attribute to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation.
+These Observtions can provide context for other Observations such that the structure of the observational data model and collection paradigm are formally represented in the Annotation.
 

### 2011-06-01 docs/pubs/EIM-2011/annotation.tex
minor wordsmith; add XML note

@@
 while there is some overlap between these two mechanisms - particularly with respect to measurement standards or units - semantic concepts applied to each attribute have the benefit of placing the descriptive burden on the ontology from which the concepts are drawn.
-Because the Annotation represents a potentially subjective perspective on the data, they are stored independently; the former referencing the latter.
+Because the Annotation represents a potentially subjective perspective on the data, they are stored independently as XML; the former referencing the latter.
 
@@
 unit).
-Tablular data objects map each attribute to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation.
+Tablular data object attributes are mapped to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation.
 These Observtions can provide context for other Observations such that the structure of the observational data model and collection paradigm are formally represented in the Annotation.

### 2011-06-01 docs/pubs/EIM-2011/discovery.tex
add more sections (for me)
discovery draft

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Discovery}
+
+\subsection{Concept query}
+The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria.
+Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.
+By leveraging the relationships defined and/or inferred from the ontology we are able to increase recall beyond what is possible for simple keyword-based searches (Berkely et al, 20??).
+Broad queries return direct matches and also n-depth subclass matches.
+The queries can be quickly honed when using this browsing paradigm which allows rapid exploration of the data repository without the oneous of defining complete observational queries \emph{de novo}.
+Measurement templates as defined in OBOE compatible ontologies enable a single concept to proxy its constituent parts, namely the valid Characteristics of certain Entities that can be measured with certain Protocols and Standards.
+This short-hand query generation highlights one of the compelling applications of using OBOE compatible ontologies and can also be leveraged when authoring Annotations.
+Using compound semantic query criteria applies a fine-grained filter on the datasets that are returned.
+Results may include only those datasets that include measurements for a set of specific Characteristics of a particular Entity.
+Furthermore, a query can specify that those measurements come from precisely the \emph{same instance of that Entity} as belaboured by the authors of the OBOE specification (Schildhauer here somewhere??).
+
+\subsection{Data query}
+For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied.
+Different techniques are available for merging the Annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept-based query are collated and only that subset is materialized.
+The Data Manager Library is conveniently used to load raw data objects because our corpus is described using EML and the Annotation syntax - both of which inform the correct use of the Data Manager query features.
+For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query.
+Figure XX illustrates a semantic data query in which datasets recording organismal mass is greater than 0.5 grams are returned.
 

### 2011-06-01 docs/pubs/EIM-2011/framework.tex
a few more notes about SMS API

@@
 The SMS API includes ontology management features, annotation manipulation capabilities, and simple concept navigation and visualization widgets.
-OWL API primarily underlies the ontology management interface providing the RDF/OWL parsing and serialization as well as simple class and property exploration.
-The Pellet reasoner exposes inferred axioms and relationships that are not explicitly defined in the source ontologies.
+The API is intended to be a centralized toolkit for use in multiple contexts - be they client or server - that abstracts the specifics of any single semantic library.
+That being said, OWL API primarily underlies the ontology management interface providing RDF/OWL parsing and serialization as well as simple class and property exploration.
+A Pellet reasoner exposes inferred axioms and relationships that are not explicitly defined in the source ontologies.
 Inference is particularly powerful when utilizing OBOE Measurement 'templates' that express exactly what (Entity and Characteristic) can be observed and how (Standard and Protocol).
-Annotations are ultimately managed and stored in an opaque local relational database for scalible persistence and query performance; the overhead of in-memory Annotation exploration having proven to be prohibitively limiting for large sets of metadata.
+
+Annotations are ultimately managed and stored in an opaque local relational database chosen for scalible persistence and query performance; the overhead of in-memory Annotation queries having proven to be prohibitively limiting for large sets of data.
 

### 2011-06-02 docs/pubs/EIM-2011/intro.tex

@@
 \section{Introduction}
-\label{sec-introduction}
+\label{sec-intro}
+
+A major challenge in environmental information management concerns
+providing effective approaches for the discovery and integration of
+heteregenous data sets. For instance, locating and combining relevant
+observational data are often critical and time-consuming steps for
+researchers studying phenomena at broad spatial, temporal, and
+biological scales
+\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do,green05:_compl_in_ecolog_and_conser,sorokina09:_detec_inter_variab_inter_obser_ornit_data,jackson01:_histor_overf_and_recen_collap}.
+The
+underlying data sets used within such studies frequently differ in
+subtle and complex ways, due in part to the protocols used for data
+collection, the types of observations made, and the experimental and
+other contextual information associated with the data set. These
+differences in turn can lead to structural and semantic heterogeneity
+among data sets that make them hard to discover using current data
+management approaches and require considerable manual effort by
+researchers needing to combine data sets.
+
+A number of recent efforts within the earth and environmental
+informatics communities are adopting the notion of an observation as a
+key modeling concept for enabling improved discovery and integration
+of scientific data
+\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
+approaches provide higher-level observational data models for
+describing and representing observations and measurements found in
+underlying datasets by defining common ``core'' concepts such as the
+entities or features being observed, measurement units and protocols,
+and context relationships between observations \cite{om,bowers08}.  A
+major goal of these approaches is to enable interoperability and
+uniform access to data by abstracting away the underlying
+representation details that often exists across scientific data sets.
+
+In this paper we describe extensions to the Ecological Metadata
+Language (EML) \cite{} and suppporting tools for enabling improved
+discovery and integration of ecological data sets. Our work is based
+on the Extensible Observations Ontology (OBOE) \cite{}, which
+represents a generic observational model implemented in OWL-DL for
+describing domain-specific observation and measurement types. Our
+approach adds additional metadata in the form of semantic annotations
+that link attributes within datasets to OBOE terms for describing the
+implicit observation and measurement types found within data
+sets. Semantic annotations are executable in the sense that they can
+be used to convert a data set into a collection of observation and
+measurement instances, providing a more uniform representation for
+expressing queries and performing integration.  To support the
+creation of annotations, we have extended the Morpho metadata editor
+with a high-level user interface as well as the Metacat data catalog
+for storing and querying annotations through a new Semantic Mediation
+API. This API can also be used to perform basic data-level integration
+tasks using our prior work on the EML Data Manager Library \cite{}.
+
+The rest of this paper is organized as follows. \secref{sec-framework}
+briefly describes the various components used within our
+approach. \secref{sec-annotation} describes the extensions we have
+developed for Morpho and Metacat to support semantic
+annotation. \secref{sec-applications} describes the types of data
+discovery queries and integration services supported by our
+framework. \secref{sec-related} briefly describes related work, and we
+summarize our contributions in \secref{sec-summary}.
+
+%\section{Preliminaries}
+
+%Briefly describe EML, Metacat, and Morhpo. Briefly describe oboe and
+%semantic annotations.
 

### 2011-06-02 docs/pubs/EIM-2011/integration.tex
add more sections (for me)
integration draft

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Integration}
+
+The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
+In addition to inspecting the data for values within a range and returning the packages that contain a match, the "smart" data integration feature goes further by constructing a synthetic data product that represents the complete results of the query both in terms of the attributes and the values that are returned.
+Each original data package match may have wildly different syntactic structures (e.g.
+column number, naming, order) but could share a subset of attributes that are asserted as semantically compatible.
+These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly.
+While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g.
+gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g.
+1000 milligrams in a gram) will enable more thorough data integration.
 

### 2011-06-02 docs/pubs/EIM-2011/abstract.tex

@@
 \begin{abstract}
-  The Data Manager paper from the last EIM included semantic
-  foreshadowing in the 'future directions' section and this is a
-  logical continuation of that work. Previously, we saw that the DM
-  library enabled data integration tasks provided we knew exactly
-  which data attributes from exactly which datasets were "compatible"
-  with each other. Here "compatible" means not simply structurally
-  similar (e.g. decimal numbers) but also conceptually similar
-  (e.g. ocean water temperature in C). By including rich semantic
-  annotations we initially show simple integration capabilities where
-  the subject domain of our data corpus is relatively constrained
-  (i.e. SBC-LTER) and the observational models represented are
-  minimally variable (i.e. context excluded). Semantically-enahnced
-  query capabilities facilitate this 'smart integration' in multiple
-  ways. First, the corpus is refined through compound concept-based
-  queries that exercise OBOE-compatible ontological subsumption
-  hierarchies to locate roughly equivalent attributes across
-  datasets. Additional query constraints are defined in relation to
-  the precise observational model formally expressed in OBOE. Finally,
-  the actual data values are interrogated so that only those that fall
-  in the desired range are returned. From this resultant subset of
-  disparate data objects we create a single synthetic data product
-  that represents a 'materialization' of the observations. Technical
-  discussion will include high-level descriptions of the
-  infrastructure and tools employed with focus on our development in
-  the rapidly evolving semantic frontier where scalability and best
-  practices are particularly germane and equally elusive.
+  Effective discovery and integration of ecological data within data
+  management systems requires rich semantic information that can
+  describe and relate the types of information contained within
+  disparate data sets. Within the Semtools project, we have developed
+  approaches for expressing and representing semantic annotations of
+  data sets for supplementing attribute and data-level metadata with
+  terms drawn from domain-specific ontologies. Annotations provide a
+  formal mechansism that can be used together with reasoning systems
+  to enhance existing data discovery and integration approaches.  In
+  this paper we describe extensions to the Ecological Metadata
+  Language (EML) and associated tools for storing and using semantic
+  annotations.  Specifically, we describe new user interface
+  components implemented within the Morpho EML metadata editor for
+  capturing user-supplied semantic annotations, extensions to the
+  Metacata system for storing and accessing annotations and
+  corresponding OWL-DL ontologies, and a new API within Metacat that
+  uses annotation metadata to provide concept-based search and
+  integration of data sets.
+%   The Data Manager paper from the last EIM included semantic
+%   foreshadowing in the 'future directions' section and this is a
+%   logical continuation of that work. Previously, we saw that the DM
+%   library enabled data integration tasks provided we knew exactly
+%   which data attributes from exactly which datasets were "compatible"
+%   with each other. Here "compatible" means not simply structurally
+%   similar (e.g. decimal numbers) but also conceptually similar
+%   (e.g. ocean water temperature in C). By including rich semantic
+%   annotations we initially show simple integration capabilities where
+%   the subject domain of our data corpus is relatively constrained
+%   (i.e. SBC-LTER) and the observational models represented are
+%   minimally variable (i.e. context excluded). Semantically-enahnced
+%   query capabilities facilitate this 'smart integration' in multiple
+%   ways.
+%   First, the corpus is refined through compound concept-based
+%   queries that exercise OBOE-compatible ontological subsumption
+%   hierarchies to locate roughly equivalent attributes across
+%   datasets. Additional query constraints are defined in relation to
+%   the precise observational model formally expressed in OBOE. Finally,
+%   the actual data values are interrogated so that only those that fall
+%   in the desired range are returned. From this resultant subset of
+%   disparate data objects we create a single synthetic data product
+%   that represents a 'materialization' of the observations. Technical
+%   discussion will include high-level descriptions of the
+%   infrastructure and tools employed with focus on our development in
+%   the rapidly evolving semantic frontier where scalability and best
+%   practices are particularly germane and equally elusive.
 \end{abstract}

### 2011-06-02 docs/pubs/EIM-2011/intro.tex

@@
 In this paper we describe extensions to the Ecological Metadata
-Language (EML) \cite{} and suppporting tools for enabling improved
-discovery and integration of ecological data sets. Our work is based
-on the Extensible Observations Ontology (OBOE) \cite{}, which
-represents a generic observational model implemented in OWL-DL for
-describing domain-specific observation and measurement types. Our
-approach adds additional metadata in the form of semantic annotations
-that link attributes within datasets to OBOE terms for describing the
-implicit observation and measurement types found within data
-sets. Semantic annotations are executable in the sense that they can
-be used to convert a data set into a collection of observation and
-measurement instances, providing a more uniform representation for
-expressing queries and performing integration.  To support the
-creation of annotations, we have extended the Morpho metadata editor
-with a high-level user interface as well as the Metacat data catalog
-for storing and querying annotations through a new Semantic Mediation
-API. This API can also be used to perform basic data-level integration
-tasks using our prior work on the EML Data Manager Library \cite{}.
+Language (EML) \cite{Fegraus07} and suppporting tools for enabling
+improved discovery and integration of ecological data sets. Our work
+is based on the Extensible Observations Ontology (OBOE)
+\cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
+generic observational model implemented in OWL-DL for describing
+domain-specific observation and measurement types. Our approach adds
+additional metadata in the form of semantic annotations that link
+attributes within datasets to OBOE terms for describing the implicit
+observation and measurement types found within data sets. Semantic
+annotations are executable in the sense that they can be used to
+convert a data set into a collection of observation and measurement
+instances, providing a more uniform representation for expressing
+queries and performing integration.  To support the creation of
+annotations, we have extended the Morpho metadata editor
+\cite{metacat02:_manag_heter_ecolog_data_using_morph} with a
+high-level user interface as well as the Metacat data catalog
+\cite{berkley01:_metac} for storing and querying annotations through a
+new Semantic Mediation API. This API can also be used to perform basic
+data-level integration tasks using our prior work on the EML Data
+Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin}.
 

### 2011-06-02 docs/pubs/EIM-2011/application.tex
condense discovery/integration into one section (application.tex) reorder so that annotation/oboe section is before framework section.

--- 2011-06-01 docs/pubs/EIM-2011/discovery.tex
@@
-\section{Discovery}
+\section{Discovery and Integration}
 
@@
 
+\subsection{Data integration}
+The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
+In addition to inspecting the data for values within a range and returning the packages that contain a match, the "smart" data integration feature goes further by constructing a synthetic data product that represents the complete results of the query both in terms of the attributes and the values that are returned.
+Each original data package match may have wildly different syntactic structures (e.g.
+column number, naming, order) but could share a subset of attributes that are asserted as semantically compatible.
+These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly.
+While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g.
+gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g.
+1000 milligrams in a gram) will enable more thorough data integration.
+
 %%% Local Variables: 

### 2011-06-02 docs/pubs/EIM-2011/intro.tex
condense discovery/integration into one section (application.tex) reorder so that annotation/oboe section is before framework section.

@@
 uniform access to data by abstracting away the underlying
-representation details that often exists across scientific data sets.
+representation details that often exist across scientific data sets.
 

### 2011-06-02 docs/pubs/EIM-2011/application.tex
figures for: integration, metacat query, morpho annotation

@@
 The Data Manager Library is conveniently used to load raw data objects because our corpus is described using EML and the Annotation syntax - both of which inform the correct use of the Data Manager query features.
-For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query.
-Figure XX illustrates a semantic data query in which datasets recording organismal mass is greater than 0.5 grams are returned.
+For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query (\figref{fig:metacat-query}).
 
@@
 column number, naming, order) but could share a subset of attributes that are asserted as semantically compatible.
-These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly.
+These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly (\figref{fig:integration}).
 While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g.
@@
 1000 milligrams in a gram) will enable more thorough data integration.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{images/metacat-query.png}
+\caption{Semantic data query web interface.
+Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned.
+Additional search options and compund query criteria can be specified within the other tabs.
+Matches can be saved in the data cart for later exploration.}
+\label{fig:metacat-query}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{images/integration.png}
+\caption{Data integration query across multiple data packages (A, B).
+Annotations (C, D) are used to determine semantically equivalent data attributes contained in the data objects (E, F).
+Attributes 'plot' and 'site' are considered equivalent measurements of the characteristic 'Location'; attributes 'weight' and 'wt' both map to the same characteristic 'Mass'.
+The Semantic Mediation API utilizes the Data Manager Library to load and query the source data informed by semantic similarities between the structurally disparate data attributes.}
+\label{fig:integration}
+\end{figure}
 

### 2011-06-02 docs/pubs/EIM-2011/framework.tex
figures for: integration, metacat query, morpho annotation

@@
 The semantic editor plugin for Morpho provides a front-end to the SMS API and allows metadata editors to build up semantic Annotations for existing data package descriptions.
-The editor focuses on a simplictic fill-in-the-blank style interface with reusable, searchable, hierachical concept selection widgets.
+The editor focuses on a simplictic fill-in-the-blank style interface with reusable, searchable, hierachical concept selection widgets (\figref{fig:morpho-annotation}).
 The optional plugin seamlessly integrates with a standard Morpho installation augmenting the existing features to provide semantic query capabilities for locating data packages, marking up data packages with semantic Annotations, and saving Annotations locally or to a shared repository where they can be discovered and explored by a broader audience.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{images/morpho-annotation.png}
+\caption{Morpho metadata editor with Semantic plugin.
+Fill-in-the-blank style editing interface to encourage natural language descriptions.
+A searchable hierarchical browser is used to select concepts from an ontology.}
+\label{fig:morpho-annotation}
+\end{figure}
 

### 2011-06-02 docs/pubs/EIM-2011/intro.tex
resolve section references from intro.tex to other .tex files

@@
 
-The rest of this paper is organized as follows. \secref{sec-framework}
+The rest of this paper is organized as follows. \secref{sec:annotation}
 briefly describes the various components used within our
-approach. \secref{sec-annotation} describes the extensions we have
+approach. \secref{sec:framework} describes the extensions we have
 developed for Morpho and Metacat to support semantic
-annotation. \secref{sec-applications} describes the types of data
+annotation. \secref{sec:application} describes the types of data
 discovery queries and integration services supported by our
-framework. \secref{sec-related} briefly describes related work, and we
-summarize our contributions in \secref{sec-summary}.
+framework. \secref{sec:related} briefly describes related work, and we
+summarize our contributions in \secref{sec:summary}.
 

### 2011-06-03 docs/pubs/EIM-2011/annotation.tex
resolve section references from intro.tex to other .tex files
OBOE/Annotation figure for Kelp CHN dataset. needs better caption

@@
 \section{Annotation}
+\label{sec:annotation}
 
@@
 unit).
-Tablular data object attributes are mapped to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation.
+Tablular data object attributes are mapped to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation (\figref{fig:kelp-mass-model}).
 These Observtions can provide context for other Observations such that the structure of the observational data model and collection paradigm are formally represented in the Annotation.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
+\caption{Partial OBOE Annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.}
+\label{fig:kelp-mass-model}
+\end{figure}
 

### 2011-06-03 docs/pubs/EIM-2011/framework.tex
resolve section references from intro.tex to other .tex files
use two columns for screen shots. crop images rescale diagram in omnigraffle
merge widget and annotation UI into single image with widget floating atop

@@
-\section{Framework}
+\section{Discovery and Integration Framework}
+\label{sec:framework}
 
-Semtools has focused development effort on three components:
-a Java library for accessing and manipulating OBOE ontology extensions and Annotations,
-an Annotation plugin for the Morpho metadata editor, and query extensions for the Metacat data catalog.
+The Semtools project has focused development efforts on three main
+components: a Java library for accessing and manipulating OBOE
+ontology extensions and semantic annotations, an annotation plugin for
+the Morpho metadata editor, and query extensions for the Metacat data
+catalog. Below we briefly describe these components and give a brief
+overview of the OBOE model (for more in-depth presentation of OBOE see
+\cite{madin07:_ontol_for_descr_and_synth,bowers08}) and the semantic
+annotation approach used in Semtools.
 
-\subsection{Semantic mediation API}
-The SMS API includes ontology management features, annotation manipulation capabilities, and simple concept navigation and visualization widgets.
-The API is intended to be a centralized toolkit for use in multiple contexts - be they client or server - that abstracts the specifics of any single semantic library.
-That being said, OWL API primarily underlies the ontology management interface providing RDF/OWL parsing and serialization as well as simple class and property exploration.
-A Pellet reasoner exposes inferred axioms and relationships that are not explicitly defined in the source ontologies.
-Inference is particularly powerful when utilizing OBOE Measurement 'templates' that express exactly what (Entity and Characteristic) can be observed and how (Standard and Protocol).
-
-Annotations are ultimately managed and stored in an opaque local relational database chosen for scalible persistence and query performance; the overhead of in-memory Annotation queries having proven to be prohibitively limiting for large sets of data.
-
-\subsection{Morpho editor plugin}
-The semantic editor plugin for Morpho provides a front-end to the SMS API and allows metadata editors to build up semantic Annotations for existing data package descriptions.
-The editor focuses on a simplictic fill-in-the-blank style interface with reusable, searchable, hierachical concept selection widgets (\figref{fig:morpho-annotation}).
-The optional plugin seamlessly integrates with a standard Morpho installation augmenting the existing features to provide semantic query capabilities for locating data packages, marking up data packages with semantic Annotations, and saving Annotations locally or to a shared repository where they can be discovered and explored by a broader audience.
+\mypara{The OBOE Observational Model.}  \figref{fig:oboe} shows the
+main modeling constructs of
+OBOE\footnote{\url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}}. An
+{\em observation} is made of an {\em entity} (e.g., biological
+organisms, geographic locations, environmental features) and serves to
+group a set of measurements together to form a single
+``\emph{observation event}''. A \emph{measurement} assigns a value to
+a {\em characteristic} of the observed entity (e.g., the weight of a
+plant), and can also include \emph{standards} (e.g., units) and
+collection \emph{protocols}. An observation can occur within the
+surrounding \emph{context} of other observations (e.g., as part of a
+temporal or spatial context), and context may include a named
+relationship (e.g., ``partOf'', ``within'') that existed during the
+observation event. A key feature of OBOE is that it allows properties
+(characteristics and relationships) of entities to be asserted without
+being interpreted as \emph{inherently} (i.e., {always}) true of the
+entity.  Depending on the context in which the entity was observed or
+how the measurements were performed, an entity's properties may take
+on different values.  OBOE allows RDF-style assertions about entities
+to be contextualized, and thus different values can be assigned for
+the same entity under distinct contexts, which is a crucial feature
+for modeling ecological as well as many other types of scientific data
+\cite{bowers08,mungall07:_repres_phenot_in_owl}.
 
@@
   \centering
-\includegraphics[width=0.5\textwidth]{images/morpho-annotation.png}
+  \includegraphics[width=0.5\textwidth]{images/oboe}
+  \caption{The main classes and properties of the extensible
+    observation ontology (OBOE). While shown using UML, the model is
+    defined using OWL-DL.}
+  \label{fig:oboe}
+\end{figure}
+
+\mypara{Semantic Annotations.} Semantic Annotations describe the
+observation model contained in a data object. The Annotation serves as
+a template for constructing a fully fleshed OBOE model of the [usually
+tabular] data. Less formal descriptions of the data attributes are
+often included in traditional metadata formats like EML. while there
+is some overlap between these two mechanisms - particularly with
+respect to measurement standards or units - semantic concepts applied
+to each attribute have the benefit of placing the descriptive burden
+on the ontology from which the concepts are drawn. Because the
+Annotation represents a potentially subjective perspective on the
+data, they are stored independently as XML; the former referencing the
+latter.
+
+The Annotation structure largely mirrors the core classes in
+OBOE. Observations are composed of Measurements of a specific Entity
+and are represented by a collection of Characteristics (usually just
+one) collected using a defined Protocol and Standard (i.e. unit).
+Tablular data object attributes are mapped to a Measurement such that
+a collection of attributes usually pertain to the Entity of a shared
+Observation (\figref{fig:kelp-mass-model}). These Observtions can
+provide context for other Observations such that the structure of the
+observational data model and collection paradigm are formally
+represented in the Annotation.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
+\caption{Partial OBOE Annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.}
+\label{fig:kelp-mass-model}
+\end{figure}
+
+\mypara{The Semantic Mediation API.} The SMS API includes ontology
+management features, annotation manipulation capabilities, and simple
+concept navigation and visualization widgets. The API is intended to
+be a centralized toolkit for use in multiple contexts - be they client
+or server - that abstracts the specifics of any single semantic
+library. That being said, OWL API primarily underlies the ontology
+management interface providing RDF/OWL parsing and serialization as
+well as simple class and property exploration. A Pellet reasoner
+exposes inferred axioms and relationships that are not explicitly
+defined in the source ontologies. Inference is particularly powerful
+when utilizing OBOE Measurement 'templates' that express exactly what
+(Entity and Characteristic) can be observed and how (Standard and
+Protocol).
+
+Annotations are ultimately managed and stored in an opaque local
+relational database chosen for scalible persistence and query
+performance; the overhead of in-memory Annotation queries having
+proven to be prohibitively limiting for large sets of data.
+
+\mypara{The Morpho Editor Plugin.}  The semantic editor plugin for
+Morpho provides a front-end to the SMS API and allows metadata editors
+to build up semantic Annotations for existing data package
+descriptions. The editor focuses on a simplictic fill-in-the-blank
+style interface with reusable, searchable, hierachical concept
+selection widgets (\figref{fig:morpho-annotation}). The optional
+plugin seamlessly integrates with a standard Morpho installation
+augmenting the existing features to provide semantic query
+capabilities for locating data packages, marking up data packages with
+semantic Annotations, and saving Annotations locally or to a shared
+repository where they can be discovered and explored by a broader
+audience.
+
+\begin{figure*}
+\centering
+\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
 \caption{Morpho metadata editor with Semantic plugin.
@@
 \label{fig:morpho-annotation}
-\end{figure}
+\end{figure*}
 
-\subsection{Metacat query extensions}
-The semantic plugin for Metacat augements existing metadata storage and search by allowing Annotations to be saved and queried alongside the metadata and data that they annotate.
-In addition to traditional keyword and spatial search criteria, the Metacat plugin allows semantic criteria to be included where they may either increase query recall using term-expansion (i.e.
-traversing the class subsumption hierarchy) or refine the resultset by limiting matches to datasets that contain the specified observational model (e.g.
-combinations of OBOE-compatible Entity, Characteristic, or Protocol concepts).
-The observational model can be leveraged further by materializing the Annotation and data artifact (via the Data Manager Library) into a full OBOE model and inspecting the observational values themselves.
+\mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
+augements existing metadata storage and search by allowing Annotations
+to be saved and queried alongside the metadata and data that they
+annotate. In addition to traditional keyword and spatial search
+criteria, the Metacat plugin allows semantic criteria to be included
+where they may either increase query recall using term-expansion
+(i.e. traversing the class subsumption hierarchy) or refine the
+resultset by limiting matches to datasets that contain the specified
+observational model (e.g. combinations of OBOE-compatible Entity,
+Characteristic, or Protocol concepts). The observational model can be
+leveraged further by materializing the Annotation and data artifact
+(via the Data Manager Library) into a full OBOE model and inspecting
+the observational values themselves.
 

### 2011-06-03 docs/pubs/EIM-2011/framework.tex
combined annotation and framework sections, did a pass all the way through the new (combined) section II

@@
 \mypara{The OBOE Observational Model.}  \figref{fig:oboe} shows the
-main modeling constructs of
-OBOE\footnote{\url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}}. An
-{\em observation} is made of an {\em entity} (e.g., biological
-organisms, geographic locations, environmental features) and serves to
-group a set of measurements together to form a single
-``\emph{observation event}''. A \emph{measurement} assigns a value to
-a {\em characteristic} of the observed entity (e.g., the weight of a
-plant), and can also include \emph{standards} (e.g., units) and
-collection \emph{protocols}. An observation can occur within the
-surrounding \emph{context} of other observations (e.g., as part of a
-temporal or spatial context), and context may include a named
-relationship (e.g., ``partOf'', ``within'') that existed during the
-observation event. A key feature of OBOE is that it allows properties
-(characteristics and relationships) of entities to be asserted without
-being interpreted as \emph{inherently} (i.e., {always}) true of the
-entity.  Depending on the context in which the entity was observed or
-how the measurements were performed, an entity's properties may take
-on different values.  OBOE allows RDF-style assertions about entities
-to be contextualized, and thus different values can be assigned for
-the same entity under distinct contexts, which is a crucial feature
-for modeling ecological as well as many other types of scientific data
-\cite{bowers08,mungall07:_repres_phenot_in_owl}.
+main modeling constructs of OBOE (see:
+\url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}). An {\em
+  observation} is made of an {\em entity} (e.g., biological organisms,
+geographic locations, environmental features) and serves to group a
+set of measurements together to form a single ``\emph{observation
+  event}''. A \emph{measurement} assigns a value to a {\em
+  characteristic} of the observed entity (e.g., the weight of a
+plant), and can also include \emph{standards} (e.g., units as well as
+standards for coded values) and collection \emph{protocols}. An
+observation can occur within the surrounding \emph{context} of other
+observations (e.g., as part of a temporal or spatial context), and
+context may include a named relationship (e.g., ``partOf'',
+``within'') that existed during the observation event. A key feature
+of OBOE is that it allows properties (characteristics and
+relationships) of entities to be asserted without being interpreted as
+\emph{inherently} (i.e., {always}) true of the entity.  Depending on
+the context in which the entity was observed or how the measurements
+were performed, an entity's properties may take on different values.
+OBOE allows RDF-style assertions about entities to be contextualized,
+and thus different values can be assigned for the same entity under
+distinct contexts, which is a crucial feature for modeling ecological
+as well as many other types of scientific data
+\cite{bowers08,mungall07:_repres_phenot_in_owl}. In addition, OBOE is
+currently implemented as an OWL-DL ontology that can be easily used
+with (or extended by) other ontologies for specifying domain-specific
+types of entities, characteristics, measurement standards, protocols,
+and relationships. For instance, the Semtools project has defined
+specific OBOE extensions in collaboration with the Santa Barbara
+Coastal Long-Term Ecological Research Project as well as through
+ongoing collaborations with other projects, and general extensions
+exist for OBOE that define a number of common entities, measurement
+units, and corresponding physical characteristics.
 
-\begin{figure}
+\begin{figure}[!t]
   \centering
-  \includegraphics[width=0.5\textwidth]{images/oboe}
+  \includegraphics[width=0.49\textwidth]{images/oboe}
   \caption{The main classes and properties of the extensible
@@
 
-\mypara{Semantic Annotations.} Semantic Annotations describe the
-observation model contained in a data object. The Annotation serves as
-a template for constructing a fully fleshed OBOE model of the [usually
-tabular] data. Less formal descriptions of the data attributes are
-often included in traditional metadata formats like EML. while there
-is some overlap between these two mechanisms - particularly with
-respect to measurement standards or units - semantic concepts applied
-to each attribute have the benefit of placing the descriptive burden
-on the ontology from which the concepts are drawn. Because the
-Annotation represents a potentially subjective perspective on the
-data, they are stored independently as XML; the former referencing the
-latter.
+\mypara{Semantic Annotations.} A semantic annotation consists of two
+parts: (i) a ``configuration'' of the observation model containing the
+specific entities, characteristics, observations, measurements, and so
+on (drawn from one or more domain ontologies) that appropriately
+capture the semantics of the data set; and (ii) a mapping between the
+attributes in the data set to specific measurements defined in the
+model configuration. \figref{fig:kelp-mass-model} shows a high-level
+example of an annotation defined for a simple Kelp sampling data
+set. Here, the data set consists of five attributes (bottom of
+\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
+measurement type (where only the characteristic of each measurement
+type is shown), and measurement types are organized into observations
+of specific Kelp entities (shown of type ``Macrocystis''), temporal
+points (denoted by date-times), and spatial locations (given as site
+names). Each measurement associated with a Kelp observation is assumed
+to have occurred within the site and during the given time as
+specified by the context relationships.  Semantic annotations provide
+a formal description of attribute semantics, whereas in many
+commonly-used metadata formats for describing data sets, largely only
+informal text-based descriptions of data attributes are permitted.  In
+the case of EML, there is some overlap between these two
+mechanisms---particularly with respect to measurement
+standards---however, semantic annotations extend this approach by
+providing a general mechanism to formally associate concepts drawn
+from domain ontologies to attributes. In the Semtools framework, we
+employ an XML serialization syntax for semantic annotations that is
+compatible with EML but that is stored separately from the EML
+documentation of a data set (allowing, e.g., annotations to be used
+independently of EML if needed). In addition, semantic annotations can
+be used to ``materialize'' a given data set into a set of triples
+conforming to the model configuration given in the annotation.
+Materializing a data set to a set of OBOE triples in this way provides
+a more uniform structural representation that can make a number of
+integration tasks easier (described further below).
 
-The Annotation structure largely mirrors the core classes in
-OBOE. Observations are composed of Measurements of a specific Entity
-and are represented by a collection of Characteristics (usually just
-one) collected using a defined Protocol and Standard (i.e. unit).
-Tablular data object attributes are mapped to a Measurement such that
-a collection of attributes usually pertain to the Entity of a shared
-Observation (\figref{fig:kelp-mass-model}). These Observtions can
-provide context for other Observations such that the structure of the
-observational data model and collection paradigm are formally
-represented in the Annotation.
+% Because the Annotation represents a potentially subjective
+% perspective on the data, they are stored independently as XML; the
+% former referencing the latter.  The Annotation structure largely
+% mirrors the core classes in OBOE. Observations are composed of
+% Measurements of a specific Entity and are represented by a
+% collection of Characteristics (usually just one) collected using a
+% defined Protocol and Standard (i.e. unit).  Tablular data object
+% attributes are mapped to a Measurement such that a collection of
+% attributes usually pertain to the Entity of a shared Observation
+% (\figref{fig:kelp-mass-model}). These Observtions can provide
+% context for other Observations such that the structure of the
+% observational data model and collection paradigm are formally
+% represented in the Annotation.
 
-\begin{figure}
+\begin{figure}[!b]
 \centering
 \includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
-\caption{Partial OBOE Annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.}
+\caption{Partial OBOE semantic annotation for Kelp sampling
+  data. Shaded nodes represent ontological concepts; rectangular nodes
+  are data table attibutes mapped to OBOE measurement
+  characteristics.}
 \label{fig:kelp-mass-model}
@@
 
-\mypara{The Semantic Mediation API.} The SMS API includes ontology
-management features, annotation manipulation capabilities, and simple
-concept navigation and visualization widgets. The API is intended to
-be a centralized toolkit for use in multiple contexts - be they client
-or server - that abstracts the specifics of any single semantic
-library. That being said, OWL API primarily underlies the ontology
-management interface providing RDF/OWL parsing and serialization as
-well as simple class and property exploration. A Pellet reasoner
-exposes inferred axioms and relationships that are not explicitly
-defined in the source ontologies. Inference is particularly powerful
-when utilizing OBOE Measurement 'templates' that express exactly what
-(Entity and Characteristic) can be observed and how (Standard and
-Protocol).
+\mypara{The Semantic Mediation API.} The Semantic Mediation API
+includes basic ontology management features, annotation manipulation
+capabilities, and simple concept navigation and visualization
+components. The API is intended to be a centralized toolkit for use in
+multiple application contexts (on either the client or server side).
+The Semantic Mediation API uses both the OWL API \cite{} for ontology
+management services including RDF/OWL parsing, serialization, and
+simple class and property exploration as well as the Pellet
+description-logic reasoner \cite{} for classification and exposing
+inferred axioms in source ontologies. The inference services exposed
+through the Semantic Mediation API is used in both discovery and
+integration described below.
+% is particularly powerful when utilizing OBOE Measurement 'templates'
+% that express exactly what (Entity and Characteristic) can be
+% observed and how (Standard and Protocol).
+In our current Morpho and Metacat extensions, semantic annotations are
+managed and stored automatically in an underlying, local relational
+database. While it is also possible to use in-memory approaches for
+storing and querying annotations, we found the overhead to be
+prohibitive when large numbers of data sets are managed.
 
-Annotations are ultimately managed and stored in an opaque local
-relational database chosen for scalible persistence and query
-performance; the overhead of in-memory Annotation queries having
-proven to be prohibitively limiting for large sets of data.
+% store and query annotations inmemory
+% approachesfor storing for scalable persistence and improved query
+% performance, which we found to be essential when large numbers of data
+% sets are supported. , where the overhead of in-memory annotation
+% queries having proven to be prohibitively limiting for large sets of
+% data.
 
-\mypara{The Morpho Editor Plugin.}  The semantic editor plugin for
-Morpho provides a front-end to the SMS API and allows metadata editors
-to build up semantic Annotations for existing data package
-descriptions. The editor focuses on a simplictic fill-in-the-blank
-style interface with reusable, searchable, hierachical concept
-selection widgets (\figref{fig:morpho-annotation}). The optional
-plugin seamlessly integrates with a standard Morpho installation
-augmenting the existing features to provide semantic query
-capabilities for locating data packages, marking up data packages with
-semantic Annotations, and saving Annotations locally or to a shared
-repository where they can be discovered and explored by a broader
-audience.
+\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
+plugin for Morpho provides a front-end to the Semantic Mediation API
+and allows users to define annotations for existing data package
+descriptions. The editor provides a simple ``fill-in-the-blank'' style
+form-based interface with a searchable hierachical concept selection
+widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
+integrates with a standard Morpho installation and provides semantic
+query capabilities for locating data packages, marking up data
+packages with semantic annotations, and saving annotations locally or
+to a shared repository where they can be discovered and explored by
+other users. The annotation editor in Morpho allows a user to view the
+data set being annotated and fill in (by selecting an appropriate
+ontology term) the characteristic, measurement standard, protocol, and
+associated entity for each data-set attribute. Users can also specify
+whether an observation spans multiple columns, and can provide context
+relationships between attributes (i.e., obserations). The editor
+provides a number of additional features including allowing users to
+view the entire annotation (similar to \figref{fig:kelp-mass-model})
+and to specify additional mapping constraints for observations and
+measurements (not shown in \figref{fig:morpho-annotation}).
 
@@
 \includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
-\caption{Morpho metadata editor with Semantic plugin.
-Fill-in-the-blank style editing interface to encourage natural language descriptions.
-A searchable hierarchical browser is used to select concepts from an ontology.}
+\caption{Morpho metadata editor with Semantic
+  plugin. Fill-in-the-blank style editing interface to encourage
+  natural language descriptions. A searchable hierarchical browser is
+  used to select concepts from an ontology.}
 \label{fig:morpho-annotation}
@@
 \mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
-augements existing metadata storage and search by allowing Annotations
-to be saved and queried alongside the metadata and data that they
-annotate. In addition to traditional keyword and spatial search
-criteria, the Metacat plugin allows semantic criteria to be included
-where they may either increase query recall using term-expansion
-(i.e. traversing the class subsumption hierarchy) or refine the
-resultset by limiting matches to datasets that contain the specified
-observational model (e.g. combinations of OBOE-compatible Entity,
-Characteristic, or Protocol concepts). The observational model can be
-leveraged further by materializing the Annotation and data artifact
-(via the Data Manager Library) into a full OBOE model and inspecting
-the observational values themselves.
+augements Metacat's existing metadata storage and search by allowing
+annotations to be saved and queried alongside the metadata and data
+that they annotate. In addition to traditional keyword and spatial
+search criteria, the Metacat plugin allows semantic criteria to be
+included where they may either increase query recall using
+term-expansion (i.e., traversing the class subsumption hierarchy) or
+refine the resultset by limiting matches to datasets that contain the
+specified observational model (e.g., combinations of OBOE-compatible
+entity, characteristic, measurement standard, or protocol
+concepts). The observational model can be leveraged further by
+materializing the annotation and data artifact (via the Data Manager
+Library \cite{leinfelder10:_metad_driven_approac_to_loadin}) into a
+fully instantiated OBOE model and inspecting (and querying over) the
+observational values themselves.
 

### 2011-06-03 docs/pubs/EIM-2011/intro.tex
combined annotation and framework sections, did a pass all the way through the new (combined) section II

@@
 
-The rest of this paper is organized as follows. \secref{sec:annotation}
-briefly describes the various components used within our
-approach. \secref{sec:framework} describes the extensions we have
-developed for Morpho and Metacat to support semantic
-annotation. \secref{sec:application} describes the types of data
-discovery queries and integration services supported by our
-framework. \secref{sec:related} briefly describes related work, and we
-summarize our contributions in \secref{sec:summary}.
+The rest of this paper is organized as follows. \secref{sec:framework}
+briefly describes the various components used within our approach
+including the extensions we have developed for Morpho and Metacat to
+support semantic annotation. \secref{sec:application} describes the
+types of data discovery queries and integration services supported by
+our framework. \secref{sec:related} briefly describes related work,
+and we summarize our contributions in \secref{sec:summary}.
 

### 2011-06-04 docs/pubs/EIM-2011/application.tex
resolve section references from intro.tex to other .tex files
use two columns for screen shots. crop images rescale diagram in omnigraffle
scale to 1.0 for omni diagram

@@
 \section{Discovery and Integration}
+\label{sec:application}
 
@@
 Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.
-By leveraging the relationships defined and/or inferred from the ontology we are able to increase recall beyond what is possible for simple keyword-based searches (Berkely et al, 20??).
-Broad queries return direct matches and also n-depth subclass matches.
-The queries can be quickly honed when using this browsing paradigm which allows rapid exploration of the data repository without the oneous of defining complete observational queries \emph{de novo}.
-Measurement templates as defined in OBOE compatible ontologies enable a single concept to proxy its constituent parts, namely the valid Characteristics of certain Entities that can be measured with certain Protocols and Standards.
-This short-hand query generation highlights one of the compelling applications of using OBOE compatible ontologies and can also be leveraged when authoring Annotations.
-Using compound semantic query criteria applies a fine-grained filter on the datasets that are returned.
-Results may include only those datasets that include measurements for a set of specific Characteristics of a particular Entity.
-Furthermore, a query can specify that those measurements come from precisely the \emph{same instance of that Entity} as belaboured by the authors of the OBOE specification (Schildhauer here somewhere??).
+
+By leveraging the relationships defined and/or inferred from the
+ontology we are able to increase recall beyond what is possible for
+simple keyword-based searches
+\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
+direct matches and also n-depth subclass matches. The queries can be
+quickly honed when using this browsing paradigm which allows rapid
+exploration of the data repository without the oneous of defining
+complete observational queries \emph{de novo}. Measurement templates
+as defined in OBOE compatible ontologies enable a single concept to
+proxy its constituent parts, namely the valid Characteristics of
+certain Entities that can be measured with certain Protocols and
+Standards. This short-hand query generation highlights one of the
+compelling applications of using OBOE compatible ontologies and can
+also be leveraged when authoring Annotations.  Using compound semantic
+query criteria applies a fine-grained filter on the datasets that are
+returned. Results may include only those datasets that include
+measurements for a set of specific Characteristics of a particular
+Entity. Furthermore, a query can specify that those measurements come
+from precisely the \emph{same instance of that Entity} as belaboured
+by the authors of the OBOE specification (Schildhauer here
+somewhere??).
 
@@
 For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query (\figref{fig:metacat-query}).
+
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=1.0\textwidth]{images/metacat-query.png}
+\caption{Semantic data query web interface.
+Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned.
+Additional search options and compund query criteria can be specified within the other tabs.
+Matches can be saved in the data cart for later exploration.}
+\label{fig:metacat-query}
+\end{figure*}
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{images/metacat-query.png}
-\caption{Semantic data query web interface.
-Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned.
-Additional search options and compund query criteria can be specified within the other tabs.
-Matches can be saved in the data cart for later exploration.}
-\label{fig:metacat-query}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{images/integration.png}
+\includegraphics[scale=1.0]{images/integration.png}
 \caption{Data integration query across multiple data packages (A, B).

### 2011-06-04 docs/pubs/EIM-2011/related.tex
resolve section references from intro.tex to other .tex files

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Related Work}
+\label{sec:related}
+
+The need for uniform mechanisms to describe observational data has led
+to many proposals for observational data models (e.g., \cite{om,
+  tarboton07:_cuahs_commun_obser_data_model,netcdf}), and ontologies
+(e.g.,
+\cite{fox09:_ontol,sweet,sheth08:_seman_sensor_web,mungall110:_integ}).
+The work here is complementary to these efforts by providing a
+framework and extensions to popular metadata tools (namely, Metacat
+\cite{} and Morpho \cite{}) for supporting a generic observational
+model through relatively light-weight, formal annotations. Our
+approach also supports the use of domain-specific ontologies for
+describing data as well as high-level query support for discovering
+data sets. Our extensions also support ... integration ... tasks that
+are crucial for determining the relevance of data.
+
+Data management systems are increasingly employing various forms of
+annotations to help improve search (e.g.,
+\cite{DBLP:conf/icde/GeertsKM06,Reeve05SemanticAnnotation,DBLP:conf/vldb/Bhagwat
+  CTV04}). For example, MONDRIAN \cite{DBLP:conf/icde/GeertsKM06}
+employs an annotation model and query operators to manipulate both
+data and annotations.  However, users must be familiar with the
+underlying data structures (schemas) to take advantage of these
+operators, which is generally not feasible for observational data in
+which data sets exhibit a high degree of structural and semantic
+heterogeneity.  Some efforts have also been carried out for leveraging
+annotations, e.g., for the discovery of domain-specific data
+\cite{obsdatasearch09,DBLP:conf/icde/StoyanovichMR10}, however, these
+approaches are largely based on keyword queries, and do not consider
+structured searches. Our work differs in that we consider a highly
+structured and generic model for annotations with the aim of providing
+a uniform approach for issuing structured data-discovery searches. Our
+work is closely aligned to traditional data integration approaches
+(e.g., \cite{HalevyDataIntegration06,kolaitis05}), where a global
+mediated schema is used to (physically or logically) merge the
+structures of heterogeneous data sources using mapping constraints
+among the source and target schemas. As such, the observational model
+we employ in our framqework can be viewed as a (general-purpose)
+mediation schema for observational data sets.  This schema can be
+augmented with logic rules (as target constraints) and uses the
+semantic annotations as mapping constraints. However, instead of users
+specifying logic constraints directly, we provide a high-level
+annotation language that simplifies the specification of mappings and
+more naturally aligns with the observation model. In addition, our
+work focuses on implementing practical approaches for rewriting and
+optimizing queries (that can include aggregation and summarization
+operators) over our annotation approach.  In particular, our goal is
+to create a feasible, scalable, and deployable system for applying
+these approaches for data discovery and exploratory data analysis
+within existing scientific data repositories.
 

### 2011-06-04 docs/pubs/EIM-2011/related.tex

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Related Work}
+\label{sec:related}
+
+The need for more semantic mechanisms to describe observational data
+has led to many proposals for observational data models (e.g.,
+\cite{om,
+  tarboton07:_cuahs_commun_obser_data_model,netcdf,cushing07:_compon_based_end_user_datab})
+and ontologies (e.g.,
+\cite{fox09:_ontol,sweet,sheth08:_seman_sensor_web,mungall110:_integ}).
+The work presented here is complementary to these efforts by providing
+a concrete set of software components that have been integrated with
+popular metadata tools (namely, Metacat
+\cite{metacat02:_manag_heter_ecolog_data_using_morph} and Morpho
+\cite{berkley01:_metac}) to provide a more uniform, semantic view of
+heterogeneous observational data. By extending Morpho and Metacat to
+support semantic annotations, these tools can provide additional help
+to researchers interested in performing synthetic studies by providing
+semantically-enhanced discovery and integration services, which are
+largely lacking in many existing environmental information management
+frameworks \cite{jones_new_2006}.
+
+Our work on using semantic annotations for data integration is closely
+aligned to traditional information integration approaches (e.g.,
+\cite{HalevyDataIntegration06,kolaitis05}), where a global mediated
+schema is used to (physically or logically) merge the structures of
+heterogeneous data sources using mapping constraints among the source
+and target schemas. As such, the observational model we employ in our
+framework can be viewed as a (general-purpose) mediation schema for
+observational data sets.  This schema can be augmented with logic
+rules (as target constraints) where semantic annotations are used as
+mapping constraints. However, instead of users specifying logic
+constraints directly, we provide a high-level annotation language and
+user-interface components (through Morpho) that can simplify the
+specification of mappings and more naturally aligns with the
+observation model.
+
+Finally, the use of annotations are playing a more prominent role in
+database systems (e.g.,
+\cite{GeertsKM06,Reeve05SemanticAnnotation,BhagwatCTV04}), e.g., the
+MONDRIAN system \cite{GeertsKM06} employs an annotation model and a
+set of query operators to manipulate both data and annotations.
+However, users must be familiar with the underlying data structures
+(schemas) to take advantage of these operators, which is generally not
+feasible for observational data in which data sets exhibit a high
+degree of structural and semantic heterogeneity. Our annotation
+approach used to extend EML is also similar in spirit to a number of
+other high-level mapping languages used for data exchange (e.g.,
+\cite{fagin09:_clio,an06:_build_seman_mappin_datab_ontol}). Our
+approach differs by being specifically tailored to the OBOE
+observational model, which in turn simplifies the annotation language,
+making it in general easier for users to specify annotations for
+observational data. Our approach also provides well-defined and
+unambiguous mappings from data sets to the observation and measurement
+model, which is critical for providing automated, high-quality data
+integration services over heterogeneous observational data.
+
+% Some efforts have also been carried out for leveraging annotations,
+% e.g., for the discovery of domain-specific data
+% \cite{obsdatasearch09,StoyanovichMR10}, however, these approaches are
+% largely based on keyword queries, and do not consider structured
+% searches. Our work differs in that we consider a highly structured and
+% generic model for annotations with the aim of providing a uniform
+% approach for issuing structured data-discovery searches.
 

### 2011-06-04 docs/pubs/EIM-2011/abstract.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

@@
   capturing user-supplied semantic annotations, extensions to the
-  Metacata system for storing and accessing annotations and
+  Metacat system for storing and accessing annotations and
   corresponding OWL-DL ontologies, and a new API within Metacat that

### 2011-06-04 docs/pubs/EIM-2011/annotation.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

@@
 
-Semantic Annotations describe the observation model contained in a data object.
-The Annotation serves as a template for constructing a fully fleshed OBOE model of the [usually tabular] data.
-Less formal descriptions of the data attributes are often included in traditional metadata formats like EML.
-while there is some overlap between these two mechanisms - particularly with respect to measurement standards or units - semantic concepts applied to each attribute have the benefit of placing the descriptive burden on the ontology from which the concepts are drawn.
-Because the Annotation represents a potentially subjective perspective on the data, they are stored independently as XML; the former referencing the latter.
+Semantic Annotations describe the observation model contained in a
+data object. The Annotation serves as a template for constructing a
+fully fleshed OBOE model of the [usually tabular] data. Less formal
+descriptions of the data attributes are often included in traditional
+metadata formats like EML. while there is some overlap between these
+two mechanisms - particularly with respect to measurement standards or
+units - semantic concepts applied to each attribute have the benefit
+of placing the descriptive burden on the ontology from which the
+concepts are drawn. Because the Annotation represents a potentially
+subjective perspective on the data, they are stored independently as
+XML; the former referencing the latter.
 
-The Annotation structure largely mirrors the core classes in OBOE.
-Observations are composed of Measurements of a specific Entity and are represented by a collection of Characteristics (usually just one) collected using a defined Protocol and Standard (i.e.
-unit).
-Tablular data object attributes are mapped to a Measurement such that a collection of attributes usually pertain to the Entity of a shared Observation (\figref{fig:kelp-mass-model}).
-These Observtions can provide context for other Observations such that the structure of the observational data model and collection paradigm are formally represented in the Annotation.
+The Annotation structure largely mirrors the core classes in
+OBOE. Observations are composed of Measurements of a specific Entity
+and are represented by a collection of Characteristics (usually just
+one) collected using a defined Protocol and Standard (i.e. unit).
+Tablular data object attributes are mapped to a Measurement such that
+a collection of attributes usually pertain to the Entity of a shared
+Observation (\figref{fig:kelp-mass-model}). These Observtions can
+provide context for other Observations such that the structure of the
+observational data model and collection paradigm are formally
+represented in the Annotation.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{images/oboe}
+  \caption{The main classes and properties of the extensible
+    observation ontology (OBOE). While shown here using UML, the model is
+    defined using OWL-DL.}
+  \label{fig:oboe}
+\end{figure}
 
@@
 \includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
-\caption{Partial OBOE Annotation for Kelp sampling data. Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.}
+\caption{Partial OBOE annotation for Kelp sampling data.
+Shaded nodes represent ontological concepts; rectangular nodes are data table attibutes mapped to OBOE characteristics.
+Measurement standards (e.g., grams) and collection protocols are not illustrated.}
 \label{fig:kelp-mass-model}

### 2011-06-04 docs/pubs/EIM-2011/application.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

@@
 
-\subsection{Concept query}
-The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria.
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=1.0\textwidth]{images/metacat-query.png}
+\caption{Semantic data query web interface.
+Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned.
+Additional search options and compound query criteria can be specified within the other tabs.
+Matches can be saved in the data cart for later exploration.}
+\label{fig:metacat-query}
+\end{figure*}
+
+\mypara{Concept query.} The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria.
 Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.
@@
 exploration of the data repository without the oneous of defining
-complete observational queries \emph{de novo}. Measurement templates
+complete observational queries \emph{de novo}. Measurement ``templates''
 as defined in OBOE compatible ontologies enable a single concept to
-proxy its constituent parts, namely the valid Characteristics of
-certain Entities that can be measured with certain Protocols and
-Standards. This short-hand query generation highlights one of the
-compelling applications of using OBOE compatible ontologies and can
-also be leveraged when authoring Annotations.  Using compound semantic
-query criteria applies a fine-grained filter on the datasets that are
-returned. Results may include only those datasets that include
-measurements for a set of specific Characteristics of a particular
-Entity. Furthermore, a query can specify that those measurements come
-from precisely the \emph{same instance of that Entity} as belaboured
-by the authors of the OBOE specification (Schildhauer here
-somewhere??).
+proxy its constituent parts, namely the characteristics of
+particular entities that can be measured with a set of protocols and
+standards. This short-hand query generation highlights a
+compelling reason for using OBOE extension ontologies and can
+also be leveraged when creating or editing semantic annotations in the Morpho interface.
 
-\subsection{Data query}
-For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied.
-Different techniques are available for merging the Annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept-based query are collated and only that subset is materialized.
-The Data Manager Library is conveniently used to load raw data objects because our corpus is described using EML and the Annotation syntax - both of which inform the correct use of the Data Manager query features.
-For any data attributes that match the concept query criteria, we then verify that those attributes contain data values within the range specifed by the initial semantic data query (\figref{fig:metacat-query}).
+Using compound semantic query criteria applies a finer-grained filter on the datasets that are
+returned. Results can be restricted to only those datasets that include 
+measurements for a set of specific characteristics of a particular
+observational entity. Furthermore, a query can specify that those measurements come
+from precisely the \emph{same instance of that entity}; a feature that fully exercises the comprensive observational structure expressed in the annotation.
 
-\begin{figure*}[!t]
-\centering
-\includegraphics[width=1.0\textwidth]{images/metacat-query.png}
-\caption{Semantic data query web interface.
-Data packages containing observations of Kelp Wet Mass less than or equal to 5 [grams] are returned.
-Additional search options and compund query criteria can be specified within the other tabs.
-Matches can be saved in the data cart for later exploration.}
-\label{fig:metacat-query}
-\end{figure*}
+\mypara{Data query.} For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied.
+Different techniques are available for merging the annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept query are collated and only that subset is materialized.
+Because our corpus is described using EML in conjunction with the annotation syntax, the Data Manager Library can handily load the described raw data while the annotation informs the correct use of the Data Manager query and filtering features.
+For any measurements that match the concept query criteria, we verify that those measurements (e.g., attributes) contain data values within the range specifed in the initial semantic data query and return the data packages that contain them (\figref{fig:metacat-query}).
 
-\subsection{Data integration}
-The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
-In addition to inspecting the data for values within a range and returning the packages that contain a match, the "smart" data integration feature goes further by constructing a synthetic data product that represents the complete results of the query both in terms of the attributes and the values that are returned.
-Each original data package match may have wildly different syntactic structures (e.g.
-column number, naming, order) but could share a subset of attributes that are asserted as semantically compatible.
-These compatible, in-common attributes then become the facets and metrics in the synthetic dataset with their values filtered accordingly (\figref{fig:integration}).
-While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g.
-gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g.
-1000 milligrams in a gram) will enable more thorough data integration.
+\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
+In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned.
+Each original data object may have wildly different syntactic structures (e.g.
+column number, naming, order) but could share a subset of attributes that are semantically compatible as defined in accompanying annotations.
+These compatible, in-common attributes become the facets and metrics in a synthetic dataset; their values filtered accordingly (\figref{fig:integration}).
+While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g., gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g., 1000 milligrams in a gram) will enable increasingly thorough, automated data integration practices.
 

### 2011-06-04 docs/pubs/EIM-2011/framework.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

@@
-\section{Discovery and Integration Framework}
+\section{Semtools Framework}
 \label{sec:framework}
@@
 catalog. Below we briefly describe these components and give a brief
-overview of the OBOE model (for more in-depth presentation of OBOE see
-\cite{madin07:_ontol_for_descr_and_synth,bowers08}) and the semantic
-annotation approach used in Semtools.
+overview of the OBOE model as well as the semantic
+annotation approach used in Semtools. For a more in-depth presentation of OBOE see
+\cite{madin07:_ontol_for_descr_and_synth,bowers08}.
 
@@
 a formal description of attribute semantics, whereas in many
-commonly-used metadata formats for describing data sets, largely only
+commonly-used metadata formats for describing data sets,  only
 informal text-based descriptions of data attributes are permitted.  In
@@
 documentation of a data set (allowing, e.g., annotations to be used
-independently of EML if needed). In addition, semantic annotations can
+independently of EML or with other metadata standards if needed). In addition, semantic annotations can
 be used to ``materialize'' a given data set into a set of triples
@@
 
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
+\caption{Morpho metadata editor with Semantic
+  plugin. The fill-in-the-blank interface uses
+  natural language descriptions for intuitive editing. A searchable, hierarchical browser is
+  used to select concepts from domain-specific ontologies.}
+\label{fig:morpho-annotation}
+\end{figure*}
+
 \mypara{The Semantic Mediation API.} The Semantic Mediation API
@@
 components. The API is intended to be a centralized toolkit for use in
-multiple application contexts (on either the client or server side).
-The Semantic Mediation API uses both the OWL API \cite{} for ontology
+multiple application contexts (on either client or server deployments).
+The Semantic Mediation API uses both the OWL API \cite{owlapi} for ontology
 management services including RDF/OWL parsing, serialization, and
 simple class and property exploration as well as the Pellet
-description-logic reasoner \cite{} for classification and exposing
+description-logic reasoner \cite{pellet} for classification and exposing
 inferred axioms in source ontologies. The inference services exposed
-through the Semantic Mediation API is used in both discovery and
+through the Semantic Mediation API are used in both discovery and
 integration described below.
@@
 and to specify additional mapping constraints for observations and
-measurements (not shown in \figref{fig:morpho-annotation}).
-
-\begin{figure*}
-\centering
-\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
-\caption{Morpho metadata editor with Semantic
-  plugin. Fill-in-the-blank style editing interface to encourage
-  natural language descriptions. A searchable hierarchical browser is
-  used to select concepts from an ontology.}
-\label{fig:morpho-annotation}
-\end{figure*}
+measurements.
 

### 2011-06-04 docs/pubs/EIM-2011/intro.tex
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

@@
 A number of recent efforts within the earth and environmental
-informatics communities are adopting the notion of an observation as a
+informatics communities are adopting the notion of an \emph{observation} as a
 key modeling concept for enabling improved discovery and integration

### 2011-06-04 docs/pubs/EIM-2011/summary.tex
resolve section references from intro.tex to other .tex files
added summary section, citation entries for owl api, sonet, and pellet. edited full content of paper. would like to a) automatically spell check b) do a final editing on it before submission. very close!

--- 2011-05-26 docs/pubs/EIM-2011/intro.tex
@@
-\section{Introduction}
-\label{sec-introduction}
+\section{Summary}
+\label{sec:summary}
+
+The Semtools project has been successful in exploring and codifying technologies and techniques for applying semantic concepts to observational data.
+By providing mechanisms for linking datasets to ontological terms organized in a high-level observational model (e.g., OBOE), these new extensions to Metacat and Morpho help to overcome a number of limitations in
+existing metadata management systems that strive to provide effective data discovery and integration features.
+Our close involvement with the SONet Project (Scientific Observations Network) \cite{sonet} encourages continued use-case refinement that will inform future semantic tool development and place an emphasis on intuitive interfaces and incremental adoption.
+This varied community of stakeholders is firmly invested in the use of cutting edge semantic solutions that will ultimately benefit multiple science disciplines by reducing obstacles to broad data sharing and innovative reuse.
 

### 2011-06-05 docs/pubs/EIM-2011/abstract.tex
Trivial spelling and grammar fixes.

@@
   annotations.  Specifically, we describe new user interface
-  components implemented within the Morpho EML metadata editor for
+  components implemented within the Morpho metadata editor for
   capturing user-supplied semantic annotations, extensions to the

### 2011-06-05 docs/pubs/EIM-2011/application.tex
Trivial spelling and grammar fixes.

@@
 quickly honed when using this browsing paradigm which allows rapid
-exploration of the data repository without the oneous of defining
+exploration of the data repository without the onerous task of defining
 complete observational queries \emph{de novo}. Measurement ``templates''
@@
 
-\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
+\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration approach.
 In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned.

### 2011-06-05 docs/pubs/EIM-2011/intro.tex
Trivial spelling and grammar fixes.

@@
 uniform access to data by abstracting away the underlying
-representation details that often exist across scientific data sets.
+representation details that often impede integration across scientific data sets.
 

### 2011-06-05 docs/pubs/EIM-2011/framework.tex
Trivial spelling and grammar fixes.
added more to annotation and materialization based on matt's suggestions

@@
 catalog. Below we briefly describe these components and give a brief
-overview of the OBOE model as well as the semantic
-annotation approach used in Semtools. For a more in-depth presentation of OBOE see
+overview of the OBOE model as well as the semantic annotation approach
+used in Semtools. For a more in-depth presentation of OBOE see
 \cite{madin07:_ontol_for_descr_and_synth,bowers08}.
+
+\begin{figure}[!t]
+  \centering
+  \includegraphics[width=0.49\textwidth]{images/oboe}
+  \caption{The main classes and properties of the extensible
+    observation ontology (OBOE). While shown using UML, the model is
+    defined using OWL-DL.}
+  \label{fig:oboe}
+\end{figure}
 
@@
 
-\begin{figure}[!t]
+\begin{figure}[!b]
 \centering
-  \includegraphics[width=0.49\textwidth]{images/oboe}
-  \caption{The main classes and properties of the extensible
-    observation ontology (OBOE). While shown using UML, the model is
-    defined using OWL-DL.}
-  \label{fig:oboe}
+\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
+\caption{Partial OBOE semantic annotation for Kelp sampling
+  data. Shaded nodes represent ontological concepts; rectangular nodes
+  are data table attibutes mapped to OBOE measurement
+  characteristics.}
+\label{fig:kelp-mass-model}
 \end{figure}
@@
 to have occurred within the site and during the given time as
-specified by the context relationships.  Semantic annotations provide
-a formal description of attribute semantics, whereas in many
-commonly-used metadata formats for describing data sets,  only
-informal text-based descriptions of data attributes are permitted.  In
-the case of EML, there is some overlap between these two
-mechanisms---particularly with respect to measurement
-standards---however, semantic annotations extend this approach by
-providing a general mechanism to formally associate concepts drawn
-from domain ontologies to attributes. In the Semtools framework, we
-employ an XML serialization syntax for semantic annotations that is
-compatible with EML but that is stored separately from the EML
-documentation of a data set (allowing, e.g., annotations to be used
-independently of EML or with other metadata standards if needed). In addition, semantic annotations can
-be used to ``materialize'' a given data set into a set of triples
-conforming to the model configuration given in the annotation.
-Materializing a data set to a set of OBOE triples in this way provides
-a more uniform structural representation that can make a number of
-integration tasks easier (described further below).
+specified by the context relationships.
+
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
+\caption{Morpho metadata editor with Semantic
+  plugin. The fill-in-the-blank interface uses
+  natural language descriptions for intuitive editing. A searchable, hierarchical browser is
+  used to select concepts from domain-specific ontologies.}
+\label{fig:morpho-annotation}
+\end{figure*}
+
+Semantic annotations can be used to facilitate discovery and
+integration of heterogeneous data sets. For instance, combining
+semantic annotations with OBOE, it is possible to discover data sets
+based on searches expressed over types of observations and
+measurements of interest. As simple examples, users can pose queries
+such as ``\emph{find all data sets containing observations of Kelp
+  entities}'' and ``\emph{find all data sets containing Mass
+  measurements of Kelp}''.\footnote{Note that queries also can include
+  contextual constraints, e.g., Mass measurements of Kelp within a
+  specific spatial or temporal context.} Both of these queries would
+return the example data set in \figref{fig:kelp-mass-model} since the
+attribute {\tt DRY} is linked to a DryMass measurement of observations
+of Macrocystis (where DryMass is defined as a special kind, or
+\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
+Kelp). Using semantic annotations in this way can help to increase
+both query recall and precision over standard keyword-based approaches
+\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
+defining terms as subclasses of other terms (e.g., Macrocystis as a
+subclass of Kelp), term expansion can be used to increase the number
+(recall) of data sets returned (where subclasses of query terms are
+also searched), and the precision of the result can be improved since
+queries can specify the desired connections between terms (e.g.,
+measurements of Mass for Kelp observations) as opposed to returning
+all data sets that simply mention the terms but without any explicit
+connections (i.e., where Mass was measured, but not for Kelp samples).
+Annotations also help facilitate integration by allowing tools to
+align data-set attributes based on their declared measurement and
+observation types.
+
+In general, semantic annotations provide a formal description of
+attribute semantics, whereas in many commonly-used metadata formats
+for describing data sets, only informal text-based descriptions of
+data attributes are permitted.  In the case of EML, there is some
+overlap between these two mechanisms---particularly with respect to
+measurement standards---however, semantic annotations extend this
+approach by providing a general mechanism to formally associate
+concepts drawn from domain ontologies to attributes. In the Semtools
+framework, we employ an XML serialization syntax for semantic
+annotations that is compatible with EML but that is stored separately
+from the EML documentation of a data set (allowing, e.g., annotations
+to be used independently of EML or with other metadata standards if
+needed). In addition, semantic annotations can be used to
+``materialize'' a given data set into a set of triples conforming to
+the model configuration given in the annotation. In other words, a
+tabular data set such as the one shown in \figref{fig:kelp-mass-model}
+can automatically be converted into a corresponding collection of
+observation and measurement instances. This in turn enables a simple
+form of structural integration, where instead of having a large number
+of different tabular data structures, all data is represented using
+the standard set of structures defined by the OBOE model (see
+\figref{fig:oboe}). Thus, materializing a data set in this way
+provides a more uniform structural representation that can make a
+number of discovery and integration tasks easier. For instance,
+materialization can be used to increase query expressivity by allowing
+searches of the form ``\emph{find all data sets containing Mass
+  measurements of Kelp with values less than 1 Kilogram}'', which in
+our example can be answered by generating (i.e., materializing) the
+measurements associated with the {\tt DRY} attribute in the data set
+of \figref{fig:kelp-mass-model}.
 
@@
 
-\begin{figure}[!b]
-\centering
-\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
-\caption{Partial OBOE semantic annotation for Kelp sampling
-  data. Shaded nodes represent ontological concepts; rectangular nodes
-  are data table attibutes mapped to OBOE measurement
-  characteristics.}
-\label{fig:kelp-mass-model}
-\end{figure}
-
-\begin{figure*}[!t]
-\centering
-\includegraphics[width=1.0\textwidth]{images/morpho-annotation-widget.png}
-\caption{Morpho metadata editor with Semantic
-  plugin. The fill-in-the-blank interface uses
-  natural language descriptions for intuitive editing. A searchable, hierarchical browser is
-  used to select concepts from domain-specific ontologies.}
-\label{fig:morpho-annotation}
-\end{figure*}
-
 \mypara{The Semantic Mediation API.} The Semantic Mediation API
@@
 components. The API is intended to be a centralized toolkit for use in
-multiple application contexts (on either client or server deployments).
-The Semantic Mediation API uses both the OWL API \cite{owlapi} for ontology
-management services including RDF/OWL parsing, serialization, and
-simple class and property exploration as well as the Pellet
-description-logic reasoner \cite{pellet} for classification and exposing
-inferred axioms in source ontologies. The inference services exposed
-through the Semantic Mediation API are used in both discovery and
-integration described below.
+multiple application contexts (on either client or server
+deployments).  The Semantic Mediation API uses both the OWL API
+\cite{owlapi} for ontology management services including RDF/OWL
+parsing, serialization, and simple class and property exploration as
+well as the Pellet description-logic reasoner \cite{pellet} for
+classification and exposing inferred axioms in source ontologies. The
+inference services exposed through the Semantic Mediation API are used
+in both discovery and integration features described below.
 % is particularly powerful when utilizing OBOE Measurement 'templates'

### 2011-06-05 docs/pubs/EIM-2011/application.tex
added more to annotation and materialization based on matt's suggestions
more on section 3

--- 2011-06-04 docs/pubs/EIM-2011/application.tex
@@
-\section{Discovery and Integration}
-\label{sec:application}
-
 \begin{figure*}[!t]
@@
 
-\mypara{Concept query.} The semantic query interface primarily supports locating datasets by how well their observational models match the given criteria.
-Query criteria largely mirror the structure of an Annotation in that combinations of Entity, Characteristic and Protocol are specified and optionally compounded when increased precision is sought.
+\section{Discovery and Integration}
+\label{sec:application}
 
-By leveraging the relationships defined and/or inferred from the
-ontology we are able to increase recall beyond what is possible for
-simple keyword-based searches
+In this section we describe the new data discovery and integration
+applications we have built using the components described above as
+part of the Semtools project.
+
+\mypara{Concept query.} The semantic query interface (see
+\figref{fig:metacat-query}) is implemented as a Web application over
+Metacat that primarily supports locating datasets by how well their
+observational models match the given criteria. The interface provides
+\emph{structured} as opposed to \emph{unstructured}, i.e.,
+keyword-based queries. In particular, query criteria given by users
+largely mirror the structure of a semantic annotation in that
+combinations of Entity, Characteristic, and Protocol are specified and
+optionally compounded when increased precision is sought.
+
+As discussed above, by leveraging the relationships defined and
+inferred from the ontology we are able to increase recall beyond what
+is possible for simple keyword-based searches
 \cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
-direct matches and also n-depth subclass matches. The queries can be
-quickly honed when using this browsing paradigm which allows rapid
-exploration of the data repository without the oneous of defining
-complete observational queries \emph{de novo}. Measurement ``templates''
-as defined in OBOE compatible ontologies enable a single concept to
-proxy its constituent parts, namely the characteristics of
-particular entities that can be measured with a set of protocols and
-standards. This short-hand query generation highlights a
-compelling reason for using OBOE extension ontologies and can
-also be leveraged when creating or editing semantic annotations in the Morpho interface.
+direct matches as well as subclass matches. The queries can be quickly
+refined when using the Web application by allowing rapid exploration
+of the data repository without having to define complete observational
+queries \emph{de novo}. The interface allows users to specify
+individual classes of a measurement as well as pre-configured
+measurement types (representing standard data-set attribute types) as
+defined in OBOE compatible ontologies to enable a single concept to
+proxy its constituent parts, namely the characteristics of particular
+entities that can be measured with a set of protocols and
+standards. This short-hand query generation can save users time in
+specifying their queries, and highlights a compelling reason for using
+OBOE extension ontologies and can also be leveraged when creating or
+editing semantic annotations in the Morpho interface.
 
-Using compound semantic query criteria applies a finer-grained filter on the datasets that are
-returned. Results can be restricted to only those datasets that include 
-measurements for a set of specific characteristics of a particular
-observational entity. Furthermore, a query can specify that those measurements come
-from precisely the \emph{same instance of that entity}; a feature that fully exercises the comprensive observational structure expressed in the annotation.
+Using compound semantic query criteria applies a finer-grained filter
+on the datasets that are returned. Results can be restricted to only
+those datasets that include measurements for a set of specific
+characteristics of a particular observational entity. Furthermore, a
+query can specify that those measurements come from precisely the
+\emph{same instance of that entity}; a feature that fully exercises
+the comprensive observational structure expressed in the annotation
+and enables higher query precision as described above.
 
-\mypara{Data query.} For even more precise recall, the OBOE model can be partially \emph{materialized} during the query stage after which a data range filter can be applied.
-Different techniques are available for merging the annotation with the data that it describes, but for performance reasons a hybrid approach has been adopted in which premliminary search results from a concept query are collated and only that subset is materialized.
-Because our corpus is described using EML in conjunction with the annotation syntax, the Data Manager Library can handily load the described raw data while the annotation informs the correct use of the Data Manager query and filtering features.
-For any measurements that match the concept query criteria, we verify that those measurements (e.g., attributes) contain data values within the range specifed in the initial semantic data query and return the data packages that contain them (\figref{fig:metacat-query}).
+\mypara{Data-level query.} For even more precise recall, the OBOE
+model can be (partially) materialized (see above) during the query
+stage after which a data range filter can be applied. Different
+techniques are available for merging the annotation with the data that
+it describes, but for performance reasons a hybrid approach has been
+adopted in which premliminary search results from a structured query
+are collated and only that subset is materialized. Because our corpus
+is described using EML in conjunction with the annotation syntax, the
+Data Manager Library \cite{leinfelder10} is used to load the described
+raw data (into a relational database) while the annotation informs the
+correct use of the Data Manager query and filtering features. For any
+measurements that match the concept query criteria, we verify that
+those measurements (e.g., attributes) contain data values within the
+range specifed in the initial semantic data query and return the data
+packages that contain them (see \figref{fig:metacat-query}).
 
-\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits.
-In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned.
-Each original data object may have wildly different syntactic structures (e.g.
-column number, naming, order) but could share a subset of attributes that are semantically compatible as defined in accompanying annotations.
-These compatible, in-common attributes become the facets and metrics in a synthetic dataset; their values filtered accordingly (\figref{fig:integration}).
-While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g., gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g., 1000 milligrams in a gram) will enable increasingly thorough, automated data integration practices.
+\mypara{Data integration.} The materialization routine used for
+semantic data queries laid the groundwork for enabling data
+integration. In addition to inspecting the data for values within a
+range and returning the datasets that contain a match, the data
+integration feature of the Web application goes further by
+constructing a synthetic data product (table) that represents the
+complete results of the query in terms of both the attributes and the
+values that are returned. Each original data set may have very
+different syntactic structures (e.g. column number, naming, order) but
+could share a subset of attributes that are semantically compatible as
+defined in accompanying annotations. These compatible, in-common
+attributes become the facets and metrics in a synthetic dataset; their
+values filtered accordingly. \figref{fig:integration} further
+illustrates the data integration support provided in the current
+implementation of the Web application. Consider two data packages,
+denoted by A and B in \figref{fig:integration}.  Annotations (denoted
+C and D) are used to determine semantically equivalent data attributes
+contained in the data sets (denoted by E and F). The attributes {\tt
+  plot} and {\tt site} are considered equivalent measurements of the
+characteristic Location; attributes {\tt weight} and {\tt wt} both map
+to the same characteristic Mass. The Semantic Mediation API parses
+each annotation and then computes an equivalance mapping among
+attributes based on their corresponding annotations. The Data Manager
+Library is then used to load the data sets and then query each data
+set to produce and merge the final, synthetic result data set.
+
+While this approach provides a preliminary form of data-level
+integration, we are currently developing additional algorithms for
+determining compatibility of annotated measurements (e.g., to include
+unit information such as that gram and ounce are both mass units) and
+for converting measurement values using ontologically-defined unit
+conversions (e.g., 1000 milligrams in a gram), which will further
+support automated data integration through the Web application.
 
@@
 \caption{Data integration query across multiple data packages (A, B).
-Annotations (C, D) are used to determine semantically equivalent data attributes contained in the data objects (E, F).
-Attributes 'plot' and 'site' are considered equivalent measurements of the characteristic 'Location'; attributes 'weight' and 'wt' both map to the same characteristic 'Mass'.
-The Semantic Mediation API utilizes the Data Manager Library to load and query the source data informed by semantic similarities between the structurally disparate data attributes.}
+  Annotations (C, D) are used to determine semantically equivalent
+  data attributes contained in the data objects (E, F). Attributes
+  'plot' and 'site' are considered equivalent measurements of the
+  characteristic 'Location'; attributes 'weight' and 'wt' both map to
+  the same characteristic 'Mass'. The Semantic Mediation API utilizes
+  the Data Manager Library to load and query the source data informed
+  by semantic similarities between the structurally disparate data
+  attributes.}
 \label{fig:integration}

### 2011-06-05 docs/pubs/EIM-2011/abstract.tex

@@
   terms drawn from domain-specific ontologies. Annotations provide a
-  formal mechansism that can be used together with reasoning systems
+  formal mechanism that can be used together with reasoning systems
   to enhance existing data discovery and integration approaches.  In

### 2011-06-05 docs/pubs/EIM-2011/application.tex

@@
 \figref{fig:metacat-query}) is implemented as a Web application over
-Metacat that primarily supports locating datasets by how well their
+Metacat that primarily supports locating data sets by how well their
 observational models match the given criteria. The interface provides
@@
 Using compound semantic query criteria applies a finer-grained filter
-on the datasets that are returned. Results can be restricted to only
-those datasets that include measurements for a set of specific
+on the data sets that are returned. Results can be restricted to only
+those data sets that include measurements for a set of specific
 characteristics of a particular observational entity. Furthermore, a
@@
 \emph{same instance of that entity}; a feature that fully exercises
-the comprensive observational structure expressed in the annotation
+the comprehensive observational structure expressed in the annotation
 and enables higher query precision as described above.
@@
 it describes, but for performance reasons a hybrid approach has been
-adopted in which premliminary search results from a structured query
+adopted in which preliminary search results from a structured query
 are collated and only that subset is materialized. Because our corpus
@@
 those measurements (e.g., attributes) contain data values within the
-range specifed in the initial semantic data query and return the data
+range specified in the initial semantic data query and return the data
 packages that contain them (see \figref{fig:metacat-query}).
@@
 integration. In addition to inspecting the data for values within a
-range and returning the datasets that contain a match, the data
+range and returning the data sets that contain a match, the data
 integration feature of the Web application goes further by
@@
 defined in accompanying annotations. These compatible, in-common
-attributes become the facets and metrics in a synthetic dataset; their
+attributes become the facets and metrics in a synthetic data set; their
 values filtered accordingly. \figref{fig:integration} further
@@
 to the same characteristic Mass. The Semantic Mediation API parses
-each annotation and then computes an equivalance mapping among
+each annotation and then computes an equivalence mapping among
 attributes based on their corresponding annotations. The Data Manager

### 2011-06-05 docs/pubs/EIM-2011/framework.tex

@@
   data. Shaded nodes represent ontological concepts; rectangular nodes
-  are data table attibutes mapped to OBOE measurement
+  are data table attributes mapped to OBOE measurement
   characteristics.}
@@
 descriptions. The editor provides a simple ``fill-in-the-blank'' style
-form-based interface with a searchable hierachical concept selection
+form-based interface with a searchable hierarchical concept selection
 widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
@@
 whether an observation spans multiple columns, and can provide context
-relationships between attributes (i.e., obserations). The editor
+relationships between attributes (i.e., observations). The editor
 provides a number of additional features including allowing users to
@@
 \mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
-augements Metacat's existing metadata storage and search by allowing
+augments Metacat's existing metadata storage and search by allowing
 annotations to be saved and queried alongside the metadata and data
@@
 term-expansion (i.e., traversing the class subsumption hierarchy) or
-refine the resultset by limiting matches to datasets that contain the
-specified observational model (e.g., combinations of OBOE-compatible
-entity, characteristic, measurement standard, or protocol
-concepts). The observational model can be leveraged further by
-materializing the annotation and data artifact (via the Data Manager
-Library \cite{leinfelder10:_metad_driven_approac_to_loadin}) into a
-fully instantiated OBOE model and inspecting (and querying over) the
-observational values themselves.
+refine the result set by limiting matches to data sets that contain
+the specified observational model (e.g., combinations of
+OBOE-compatible entity, characteristic, measurement standard, or
+protocol concepts). The observational model can be leveraged further
+by materializing the annotation and data artifact (via the Data
+Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin})
+into a fully instantiated OBOE model and inspecting (and querying
+over) the observational values themselves.
 

### 2011-06-05 docs/pubs/EIM-2011/intro.tex

@@
 providing effective approaches for the discovery and integration of
-heteregenous data sets. For instance, locating and combining relevant
+heterogeneous data sets. For instance, locating and combining relevant
 observational data are often critical and time-consuming steps for
@@
 describing and representing observations and measurements found in
-underlying datasets by defining common ``core'' concepts such as the
+underlying data sets by defining common ``core'' concepts such as the
 entities or features being observed, measurement units and protocols,
@@
 In this paper we describe extensions to the Ecological Metadata
-Language (EML) \cite{Fegraus07} and suppporting tools for enabling
+Language (EML) \cite{Fegraus07} and supporting tools for enabling
 improved discovery and integration of ecological data sets. Our work
@@
 additional metadata in the form of semantic annotations that link
-attributes within datasets to OBOE terms for describing the implicit
+attributes within data sets to OBOE terms for describing the implicit
 observation and measurement types found within data sets. Semantic

### 2011-06-05 docs/pubs/EIM-2011/summary.tex

@@
 
-The Semtools project has been successful in exploring and codifying technologies and techniques for applying semantic concepts to observational data.
-By providing mechanisms for linking datasets to ontological terms organized in a high-level observational model (e.g., OBOE), these new extensions to Metacat and Morpho help to overcome a number of limitations in
-existing metadata management systems that strive to provide effective data discovery and integration features.
-Our close involvement with the SONet Project (Scientific Observations Network) \cite{sonet} encourages continued use-case refinement that will inform future semantic tool development and place an emphasis on intuitive interfaces and incremental adoption.
-This varied community of stakeholders is firmly invested in the use of cutting edge semantic solutions that will ultimately benefit multiple science disciplines by reducing obstacles to broad data sharing and innovative reuse.
+The Semtools project has been successful in exploring and codifying
+technologies and techniques for applying semantic concepts to
+observational data. By providing mechanisms for linking data sets to
+ontological terms organized in a high-level observational model (e.g.,
+OBOE), these new extensions to Metacat and Morpho help to overcome a
+number of limitations in existing metadata management systems that
+strive to provide effective data discovery and integration
+features. Our close involvement with the SONet Project (Scientific
+Observations Network) \cite{sonet} encourages continued use-case
+refinement that will inform future semantic tool development and place
+an emphasis on intuitive interfaces and incremental adoption. This
+varied community of stakeholders is firmly invested in the use of
+cutting edge semantic solutions that will ultimately benefit multiple
+science disciplines by reducing obstacles to broad data sharing and
+innovative reuse.
 

### 2011-06-06 docs/pubs/EIM-2011/application.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)

@@
 specifying their queries, and highlights a compelling reason for using
-OBOE extension ontologies and can also be leveraged when creating or
+OBOE extension ontologies. Measurement templates can also be leveraged when creating or
 editing semantic annotations in the Morpho interface.
@@
 is described using EML in conjunction with the annotation syntax, the
-Data Manager Library \cite{leinfelder10} is used to load the described
+Data Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to load the described
 raw data (into a relational database) while the annotation informs the
@@
   data attributes contained in the data objects (E, F). Attributes
-  'plot' and 'site' are considered equivalent measurements of the
-  characteristic 'Location'; attributes 'weight' and 'wt' both map to
-  the same characteristic 'Mass'. The Semantic Mediation API utilizes
+  {\tt plot} and {\tt site} are considered equivalent measurements of the
+  characteristic Location; attributes {\tt weight} and {\tt wt} both map to
+  the same characteristic Mass. The Semantic Mediation API utilizes
   the Data Manager Library to load and query the source data informed

### 2011-06-06 docs/pubs/EIM-2011/framework.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)

@@
 the Morpho metadata editor, and query extensions for the Metacat data
-catalog. Below we briefly describe these components and give a brief
+catalog. Below we describe these components and give a brief
 overview of the OBOE model as well as the semantic annotation approach
@@
 return the example data set in \figref{fig:kelp-mass-model} since the
-attribute {\tt DRY} is linked to a DryMass measurement of observations
-of Macrocystis (where DryMass is defined as a special kind, or
+attribute {\tt WET} is linked to a WetMass measurement for observations
+of Macrocystis (where WetMass is defined as a special kind, or
 \emph{subclass} of Mass, and Macrocystis is defined as a subclass of
@@
 (recall) of data sets returned (where subclasses of query terms are
-also searched), and the precision of the result can be improved since
-queries can specify the desired connections between terms (e.g.,
+also searched). The precision of the result can be improved since
+queries may specify the desired connections between terms (e.g.,
 measurements of Mass for Kelp observations) as opposed to returning
@@
 searches of the form ``\emph{find all data sets containing Mass
-  measurements of Kelp with values less than 1 Kilogram}'', which in
+  measurements of Kelp with values less than or equal to 5 grams}'', which in
 our example can be answered by generating (i.e., materializing) the
-measurements associated with the {\tt DRY} attribute in the data set
+measurements associated with the {\tt WET} and {\tt DRY} attributes in the data set
 of \figref{fig:kelp-mass-model}.
@@
 plugin for Morpho provides a front-end to the Semantic Mediation API
-and allows users to define annotations for existing data package
+and allows data owners and curators to define annotations for existing data package
 descriptions. The editor provides a simple ``fill-in-the-blank'' style
@@
 other users. The annotation editor in Morpho allows a user to view the
-data set being annotated and fill in (by selecting an appropriate
+data set being annotated as they fill in (by selecting an appropriate
 ontology term) the characteristic, measurement standard, protocol, and
@@
 relationships between attributes (i.e., observations). The editor
-provides a number of additional features including allowing users to
+provides a number of additional features including the ability to
 view the entire annotation (similar to \figref{fig:kelp-mass-model})

### 2011-07-29 docs/pubs/EIM-2011/application.tex
use pdf for figure 5 with larger font
incorporate comments from Reviewer #1

@@
 individual classes of a measurement as well as pre-configured
-measurement types (representing standard data-set attribute types) as
+measurement types (representing standard data set attribute types) as
 defined in OBOE compatible ontologies to enable a single concept to
@@
 range and returning the data sets that contain a match, the data
-integration feature of the Web application goes further by
+integration feature of the Semtools Web application goes further by
 constructing a synthetic data product (table) that represents the
@@
 \centering
-\includegraphics[scale=1.0]{images/integration.png}
+\includegraphics[width=0.45\textwidth]{images/integration.pdf}
 \caption{Data integration query across multiple data packages (A, B).

### 2011-07-29 docs/pubs/EIM-2011/framework.tex
use pdf (vector) graphic for figure 2 and let latex do the scaling. Font is increased to 12 (mostly)
incorporate comments from Reviewer #1

@@
   \caption{The main classes and properties of the extensible
-    observation ontology (OBOE). While shown using UML, the model is
+    observation ontology (OBOE). While shown using Unified Modeling Language (UML), the model is
     defined using OWL-DL.}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
+\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.pdf}
 \caption{Partial OBOE semantic annotation for Kelp sampling
@@
 Annotations also help facilitate integration by allowing tools to
-align data-set attributes based on their declared measurement and
+align data set attributes based on their declared measurement and
 observation types.
@@
 deployments).  The Semantic Mediation API uses both the OWL API
-\cite{owlapi} for ontology management services including RDF/OWL
+\cite{owlapi} for ontology management services including ontology 
 parsing, serialization, and simple class and property exploration as
@@
 plugin for Morpho provides a front-end to the Semantic Mediation API
-and allows data owners and curators to define annotations for existing data package
+and allows data owners and curators to define annotations for existing EML data 
 descriptions. The editor provides a simple ``fill-in-the-blank'' style
@@
 integrates with a standard Morpho installation and provides semantic
-query capabilities for locating data packages, marking up data
-packages with semantic annotations, and saving annotations locally or
+query capabilities for locating data packages, marking up data sets within a 
+package using semantic annotations, and saving annotations locally or
 to a shared repository where they can be discovered and explored by
@@
 ontology term) the characteristic, measurement standard, protocol, and
-associated entity for each data-set attribute. Users can also specify
+associated entity for each data set attribute. Users can also specify
 whether an observation spans multiple columns, and can provide context

### 2011-07-29 docs/pubs/EIM-2011/intro.tex
incorporate comments from Reviewer #1

@@
 \cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
-generic observational model implemented in OWL-DL for describing
+generic observational model implemented in OWL-DL \cite{owldl} for describing
 domain-specific observation and measurement types. Our approach adds

### 2011-07-30 docs/pubs/EIM-2011/abstract.tex
snapshot copy -- will finish up rest later

@@
+\noindent
 \begin{abstract}
@@
   terms drawn from domain-specific ontologies. Annotations provide a
-  formal mechanism that can be used together with reasoning systems
-  to enhance existing data discovery and integration approaches.  In
-  this paper we describe extensions to the Ecological Metadata
-  Language (EML) and associated tools for storing and using semantic
-  annotations.  Specifically, we describe new user interface
-  components implemented within the Morpho metadata editor for
-  capturing user-supplied semantic annotations, extensions to the
-  Metacat system for storing and accessing annotations and
-  corresponding OWL-DL ontologies, and a new API within Metacat that
-  uses annotation metadata to provide concept-based search and
-  integration of data sets.
+  formal mechanism that can be used together with reasoning systems to
+  enhance existing data discovery and integration approaches.  We
+  describe extensions to the Ecological Metadata Language (EML) and
+  associated tools for storing and using semantic annotations.
+  Specifically, we describe new user interface components implemented
+  within the Morpho metadata editor for capturing user-supplied
+  semantic annotations, extensions to the Metacat system for storing
+  and accessing annotations and corresponding OWL-DL ontologies, and a
+  new API within Metacat that uses annotation metadata to provide
+  concept-based search and integration of data sets.
 %   The Data Manager paper from the last EIM included semantic
@@
 
+\vspace{3pt}
+\noindent {\small{\bf{\em Keywords}---ontologies; annotation; data
+  discovery and integration}}
+
 %%% Local Variables: 

### 2011-07-30 docs/pubs/EIM-2011/application.tex
snapshot copy -- will finish up rest later

@@
 
-\mypara{Concept query.} The semantic query interface (see
-\figref{fig:metacat-query}) is implemented as a Web application over
-Metacat that primarily supports locating data sets by how well their
-observational models match the given criteria. The interface provides
-\emph{structured} as opposed to \emph{unstructured}, i.e.,
-keyword-based queries. In particular, query criteria given by users
-largely mirror the structure of a semantic annotation in that
-combinations of Entity, Characteristic, and Protocol are specified and
-optionally compounded when increased precision is sought.
+\subsection{Concept Query}
+
+The semantic query interface (see \figref{fig:metacat-query}) is
+implemented as a Web application over Metacat that primarily supports
+locating data sets by how well their observational models match the
+given criteria. The interface provides \emph{structured} as opposed to
+\emph{unstructured}, i.e., keyword-based queries. In particular, query
+criteria given by users largely mirror the structure of a semantic
+annotation in that combinations of Entity, Characteristic, and
+Protocol are specified and optionally compounded when increased
+precision is sought.
 
@@
 
-\mypara{Data-level query.} For even more precise recall, the OBOE
-model can be (partially) materialized (see above) during the query
-stage after which a data range filter can be applied. Different
-techniques are available for merging the annotation with the data that
-it describes, but for performance reasons a hybrid approach has been
-adopted in which preliminary search results from a structured query
-are collated and only that subset is materialized. Because our corpus
-is described using EML in conjunction with the annotation syntax, the
-Data Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to load the described
-raw data (into a relational database) while the annotation informs the
-correct use of the Data Manager query and filtering features. For any
-measurements that match the concept query criteria, we verify that
-those measurements (e.g., attributes) contain data values within the
-range specified in the initial semantic data query and return the data
-packages that contain them (see \figref{fig:metacat-query}).
+\subsection{Data-Level Query}
 
-\mypara{Data integration.} The materialization routine used for
-semantic data queries laid the groundwork for enabling data
-integration. In addition to inspecting the data for values within a
-range and returning the data sets that contain a match, the data
-integration feature of the Semtools Web application goes further by
-constructing a synthetic data product (table) that represents the
-complete results of the query in terms of both the attributes and the
-values that are returned. Each original data set may have very
-different syntactic structures (e.g. column number, naming, order) but
-could share a subset of attributes that are semantically compatible as
-defined in accompanying annotations. These compatible, in-common
-attributes become the facets and metrics in a synthetic data set; their
-values filtered accordingly. \figref{fig:integration} further
-illustrates the data integration support provided in the current
-implementation of the Web application. Consider two data packages,
-denoted by A and B in \figref{fig:integration}.  Annotations (denoted
-C and D) are used to determine semantically equivalent data attributes
-contained in the data sets (denoted by E and F). The attributes {\tt
-  plot} and {\tt site} are considered equivalent measurements of the
-characteristic Location; attributes {\tt weight} and {\tt wt} both map
-to the same characteristic Mass. The Semantic Mediation API parses
-each annotation and then computes an equivalence mapping among
-attributes based on their corresponding annotations. The Data Manager
-Library is then used to load the data sets and then query each data
-set to produce and merge the final, synthetic result data set.
+For even more precise recall, the OBOE model can be (partially)
+materialized (see above) during the query stage after which a data
+range filter can be applied. Different techniques are available for
+merging the annotation with the data that it describes, but for
+performance reasons a hybrid approach has been adopted in which
+preliminary search results from a structured query are collated and
+only that subset is materialized. Because our corpus is described
+using EML in conjunction with the annotation syntax, the Data Manager
+Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to
+load the described raw data (into a relational database) while the
+annotation informs the correct use of the Data Manager query and
+filtering features. For any measurements that match the concept query
+criteria, we verify that those measurements (e.g., attributes) contain
+data values within the range specified in the initial semantic data
+query and return the data packages that contain them (see
+\figref{fig:metacat-query}).
+
+\subsection{Data Integration}
+
+The materialization routine used for semantic data queries laid the
+groundwork for enabling data integration. In addition to inspecting
+the data for values within a range and returning the data sets that
+contain a match, the data integration feature of the Semtools Web
+application goes further by constructing a synthetic data product
+(table) that represents the complete results of the query in terms of
+both the attributes and the values that are returned. Each original
+data set may have very different syntactic structures (e.g. column
+number, naming, order) but could share a subset of attributes that are
+semantically compatible as defined in accompanying annotations. These
+compatible, in-common attributes become the facets and metrics in a
+synthetic data set; their values filtered
+accordingly. \figref{fig:integration} further illustrates the data
+integration support provided in the current implementation of the Web
+application. Consider two data packages, denoted by A and B in
+\figref{fig:integration}.  Annotations (denoted C and D) are used to
+determine semantically equivalent data attributes contained in the
+data sets (denoted by E and F). The attributes {\tt plot} and {\tt
+  site} are considered equivalent measurements of the characteristic
+Location; attributes {\tt weight} and {\tt wt} both map to the same
+characteristic Mass. The Semantic Mediation API parses each annotation
+and then computes an equivalence mapping among attributes based on
+their corresponding annotations. The Data Manager Library is then used
+to load the data sets and then query each data set to produce and
+merge the final, synthetic result data set.
 

### 2011-07-30 docs/pubs/EIM-2011/framework.tex
snapshot copy -- will finish up rest later

@@
   \centering
-  \includegraphics[width=0.49\textwidth]{images/oboe}
-  \caption{The main classes and properties of the extensible
-    observation ontology (OBOE). While shown using Unified Modeling Language (UML), the model is
-    defined using OWL-DL.}
+  \includegraphics[width=0.494\textwidth]{images/oboe}
+  \caption{Main classes and properties of the extensible observation
+    ontology (OBOE). While shown using the Unified Modeling Language (UML),
+    the model is defined as an OWL-DL ontology.}
   \label{fig:oboe}
+  \vspace{-12pt}
 \end{figure}
 
-\mypara{The OBOE Observational Model.}  \figref{fig:oboe} shows the
-main modeling constructs of OBOE (see:
+\subsection{The OBOE Observational Model}
+
+\figref{fig:oboe} shows the main modeling constructs of OBOE (see:
 \url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}). An {\em
@@
 
-\mypara{Semantic Annotations.} A semantic annotation consists of two
-parts: (i) a ``configuration'' of the observation model containing the
-specific entities, characteristics, observations, measurements, and so
-on (drawn from one or more domain ontologies) that appropriately
-capture the semantics of the data set; and (ii) a mapping between the
-attributes in the data set to specific measurements defined in the
-model configuration. \figref{fig:kelp-mass-model} shows a high-level
-example of an annotation defined for a simple Kelp sampling data
-set. Here, the data set consists of five attributes (bottom of
+\subsection{Semantic Annotations}
+
+A semantic annotation consists of two parts: (i) a ``configuration''
+of the observation model containing the specific entities,
+characteristics, observations, measurements, and so on (drawn from one
+or more domain ontologies) that appropriately capture the semantics of
+the data set; and (ii) a mapping between the attributes in the data
+set to specific measurements defined in the model
+configuration. \figref{fig:kelp-mass-model} shows a high-level example
+of an annotation defined for a simple Kelp sampling data set. Here,
+the data set consists of five attributes (bottom of
 \figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
@@
 measurements of interest. As simple examples, users can pose queries
-such as ``\emph{find all data sets containing observations of Kelp
-  entities}'' and ``\emph{find all data sets containing Mass
-  measurements of Kelp}''.\footnote{Note that queries also can include
-  contextual constraints, e.g., Mass measurements of Kelp within a
-  specific spatial or temporal context.} Both of these queries would
-return the example data set in \figref{fig:kelp-mass-model} since the
-attribute {\tt WET} is linked to a WetMass measurement for observations
-of Macrocystis (where WetMass is defined as a special kind, or
-\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
-Kelp). Using semantic annotations in this way can help to increase
-both query recall and precision over standard keyword-based approaches
+such as ``\emph{find all data sets containing observations of Kelp}''
+and ``\emph{find all data sets containing Mass measurements of
+  Kelp}''.  Both of these queries would return the example data set in
+\figref{fig:kelp-mass-model} since the attribute {\tt WET} is linked
+to a WetMass measurement for observations of Macrocystis (where
+WetMass is defined as a special kind, or \emph{subclass} of Mass, and
+Macrocystis is defined as a subclass of Kelp). Using semantic
+annotations in this way can help to increase both query recall and
+precision over standard keyword-based approaches
 \cite{berkley09:_improv_data_discov_for_metad}. In particular, by
@@
 connections (i.e., where Mass was measured, but not for Kelp samples).
-Annotations also help facilitate integration by allowing tools to
-align data set attributes based on their declared measurement and
-observation types.
+% Queries can additionally include context constraints, e.g., Mass
+% measurements of Kelp within a specific spatial or temporal context.
+Annotations also help facilitate integration by allowing
+tools to align data set attributes based on their declared measurement
+and observation types.
 
@@
 
-\mypara{The Semantic Mediation API.} The Semantic Mediation API
-includes basic ontology management features, annotation manipulation
-capabilities, and simple concept navigation and visualization
-components. The API is intended to be a centralized toolkit for use in
-multiple application contexts (on either client or server
-deployments).  The Semantic Mediation API uses both the OWL API
-\cite{owlapi} for ontology management services including ontology 
-parsing, serialization, and simple class and property exploration as
-well as the Pellet description-logic reasoner \cite{pellet} for
-classification and exposing inferred axioms in source ontologies. The
-inference services exposed through the Semantic Mediation API are used
-in both discovery and integration features described below.
+\subsection{The Semantic Mediation API}
+
+The Semantic Mediation API includes basic ontology management
+features, annotation manipulation capabilities, and simple concept
+navigation and visualization components. The API is intended to be a
+centralized toolkit for use in multiple application contexts (on
+either client or server deployments).  The Semantic Mediation API uses
+both the OWL API \cite{owlapi} for ontology management services
+including ontology parsing, serialization, and simple class and
+property exploration as well as the Pellet description-logic reasoner
+\cite{pellet} for classification and exposing inferred axioms in
+source ontologies. The inference services exposed through the Semantic
+Mediation API are used in both discovery and integration features
+described below.
 % is particularly powerful when utilizing OBOE Measurement 'templates'
@@
 
-\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
-plugin for Morpho provides a front-end to the Semantic Mediation API
-and allows data owners and curators to define annotations for existing EML data 
-descriptions. The editor provides a simple ``fill-in-the-blank'' style
-form-based interface with a searchable hierarchical concept selection
-widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
-integrates with a standard Morpho installation and provides semantic
-query capabilities for locating data packages, marking up data sets within a 
+\subsection{The Morpho Editor Plugin}
+
+The semantic-annotation editor plugin for Morpho provides a front-end
+to the Semantic Mediation API and allows data owners and curators to
+define annotations for existing EML data descriptions. The editor
+provides a simple ``fill-in-the-blank'' style form-based interface
+with a searchable hierarchical concept selection widget (see
+\figref{fig:morpho-annotation}). The plugin seamlessly integrates with
+a standard Morpho installation and provides semantic query
+capabilities for locating data packages, marking up data sets within a
 package using semantic annotations, and saving annotations locally or
@@
 relationships between attributes (i.e., observations). The editor
-provides a number of additional features including the ability to
-view the entire annotation (similar to \figref{fig:kelp-mass-model})
-and to specify additional mapping constraints for observations and
+provides a number of additional features including the ability to view
+the entire annotation (similar to \figref{fig:kelp-mass-model}) and to
+specify additional mapping constraints for observations and
 measurements.
 
-\mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
-augments Metacat's existing metadata storage and search by allowing
-annotations to be saved and queried alongside the metadata and data
-that they annotate. In addition to traditional keyword and spatial
-search criteria, the Metacat plugin allows semantic criteria to be
-included where they may either increase query recall using
-term-expansion (i.e., traversing the class subsumption hierarchy) or
-refine the result set by limiting matches to data sets that contain
-the specified observational model (e.g., combinations of
-OBOE-compatible entity, characteristic, measurement standard, or
-protocol concepts). The observational model can be leveraged further
-by materializing the annotation and data artifact (via the Data
-Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin})
-into a fully instantiated OBOE model and inspecting (and querying
-over) the observational values themselves.
+\subsection{Metacat Query Extensions}
+
+The semantic plugin for Metacat augments Metacat's existing metadata
+storage and search by allowing annotations to be saved and queried
+alongside the metadata and data that they annotate. In addition to
+traditional keyword and spatial search criteria, the Metacat plugin
+allows semantic criteria to be included where they may either increase
+query recall using term-expansion (i.e., traversing the class
+subsumption hierarchy) or refine the result set by limiting matches to
+data sets that contain the specified observational model (e.g.,
+combinations of OBOE-compatible entity, characteristic, measurement
+standard, or protocol concepts). The observational model can be
+leveraged further by materializing the annotation and data artifact
+(via the Data Manager Library
+\cite{leinfelder10:_metad_driven_approac_to_loadin}) into a fully
+instantiated OBOE model and inspecting (and querying over) the
+observational values themselves.
 

### 2011-07-30 docs/pubs/EIM-2011/intro.tex
snapshot copy -- will finish up rest later

@@
 biological scales
-\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do,green05:_compl_in_ecolog_and_conser,sorokina09:_detec_inter_variab_inter_obser_ornit_data,jackson01:_histor_overf_and_recen_collap}.
-The
+\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do}. The
 underlying data sets used within such studies frequently differ in
@@
 A number of recent efforts within the earth and environmental
-informatics communities are adopting the notion of an \emph{observation} as a
-key modeling concept for enabling improved discovery and integration
-of scientific data
-\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
+informatics communities are adopting the notion of an
+\emph{observation} as a key modeling concept for enabling improved
+discovery and integration of scientific data
+\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,mungall07:_repres_phenot_in_owl,bowers08}. These
 approaches provide higher-level observational data models for
@@
 uniform access to data by abstracting away the underlying
-representation details that often impede integration across scientific data sets.
+representation details that often impede integration across scientific
+data sets.
 

### 2011-07-30 docs/pubs/EIM-2011/related.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)
snapshot copy -- will finish up rest later

@@
 has led to many proposals for observational data models (e.g.,
-\cite{om,
-  tarboton07:_cuahs_commun_obser_data_model,netcdf,cushing07:_compon_based_end_user_datab})
-and ontologies (e.g.,
-\cite{fox09:_ontol,sweet,sheth08:_seman_sensor_web,mungall110:_integ}).
-The work presented here is complementary to these efforts by providing
-a concrete set of software components that have been integrated with
+\cite{om, tarboton07:_cuahs_commun_obser_data_model,netcdf}) and
+ontologies (e.g.,
+\cite{fox09:_ontol,sweet,mungall07:_repres_phenot_in_owl}).  The work
+presented here is complementary to these efforts by providing a
+concrete set of software components that have been integrated with
 popular metadata tools (namely, Metacat
@@
 aligned to traditional information integration approaches (e.g.,
-\cite{HalevyDataIntegration06,kolaitis05}), where a global mediated
+\cite{kolaitis05}), where a global mediated
 schema is used to (physically or logically) merge the structures of
@@
 
-Finally, the use of annotations are playing a more prominent role in
-database systems (e.g.,
-\cite{GeertsKM06,Reeve05SemanticAnnotation,BhagwatCTV04}), e.g., the
-MONDRIAN system \cite{GeertsKM06} employs an annotation model and a
-set of query operators to manipulate both data and annotations.
-However, users must be familiar with the underlying data structures
-(schemas) to take advantage of these operators, which is generally not
-feasible for observational data in which data sets exhibit a high
-degree of structural and semantic heterogeneity. Our annotation
-approach used to extend EML is also similar in spirit to a number of
-other high-level mapping languages used for data exchange (e.g.,
-\cite{fagin09:_clio,an06:_build_seman_mappin_datab_ontol}). Our
+Annotations are playing a more prominent role in database systems,
+e.g., the MONDRIAN system \cite{GeertsKM06} employs an annotation
+model and a set of query operators to manipulate both data and
+annotations.  However, users must be familiar with the underlying data
+structures (schemas) to take advantage of these operators, which is
+generally not feasible for observational data in which data sets
+exhibit a high degree of structural and semantic heterogeneity. Our
+annotation approach used to extend EML is also similar in spirit to a
+number of other high-level mapping languages used for data exchange
+(e.g., \cite{fagin09:_clio,an06:_build_seman_mappin_datab_ontol}). Our
 approach differs by being specifically tailored to the OBOE

### 2011-08-01 docs/pubs/EIM-2011/application.tex
finished edits for final version

@@
 
-The materialization routine used for semantic data queries laid the
-groundwork for enabling data integration. In addition to inspecting
-the data for values within a range and returning the data sets that
-contain a match, the data integration feature of the Semtools Web
-application goes further by constructing a synthetic data product
-(table) that represents the complete results of the query in terms of
-both the attributes and the values that are returned. Each original
-data set may have very different syntactic structures (e.g. column
-number, naming, order) but could share a subset of attributes that are
-semantically compatible as defined in accompanying annotations. These
-compatible, in-common attributes become the facets and metrics in a
-synthetic data set; their values filtered
-accordingly. \figref{fig:integration} further illustrates the data
-integration support provided in the current implementation of the Web
-application. Consider two data packages, denoted by A and B in
+The materialization routine for semantic data queries can help in
+enabling data integration. In addition to inspecting data for values
+within a range and returning the data sets that contain a match, the
+data integration feature of the Semtools Web application goes further
+by constructing a synthetic data product (table) that represents the
+complete results of the query in terms of both the attributes and the
+values that are returned. Each original data set may have very
+different syntactic structures (e.g., column number, naming, order)
+but could share a subset of attributes that are semantically
+compatible as defined in accompanying annotations. These compatible
+attributes can then become the basis for a synthetic data set.
+\figref{fig:integration} illustrates the data integration support
+provided in the current implementation of the Web
+application. Consider the two data packages (denoted A and B) in
 \figref{fig:integration}.  Annotations (denoted C and D) are used to
@@
 Location; attributes {\tt weight} and {\tt wt} both map to the same
-characteristic Mass. The Semantic Mediation API parses each annotation
-and then computes an equivalence mapping among attributes based on
-their corresponding annotations. The Data Manager Library is then used
-to load the data sets and then query each data set to produce and
-merge the final, synthetic result data set.
+characteristic Mass. The Semantic Mediation API computes an
+equivalence among attributes based on their corresponding
+annotations. The Data Manager Library is then used to load the data
+sets and then query each data set to produce and merge a synthetic
+result data set.
 
@@
 \includegraphics[width=0.45\textwidth]{images/integration.pdf}
-\caption{Data integration query across multiple data packages (A, B).
-  Annotations (C, D) are used to determine semantically equivalent
-  data attributes contained in the data objects (E, F). Attributes
-  {\tt plot} and {\tt site} are considered equivalent measurements of the
-  characteristic Location; attributes {\tt weight} and {\tt wt} both map to
-  the same characteristic Mass. The Semantic Mediation API utilizes
-  the Data Manager Library to load and query the source data informed
-  by semantic similarities between the structurally disparate data
+\caption{Integration query across multiple data packages (A, B).
+  Annotations (C, D) determine semantically equivalent data attributes
+  contained in the data objects (E, F). Attributes {\tt plot} and {\tt
+    site} are considered equivalent measurements of the characteristic
+  Location; attributes {\tt weight} and {\tt wt} both map to the same
+  characteristic Mass. The Semantic Mediation API utilizes the Data
+  Manager Library to load and query the source data informed by
+  semantic similarities between the structurally disparate data
   attributes.}

### 2011-08-01 docs/pubs/EIM-2011/framework.tex
finished edits for final version

@@
 The Semtools project has focused development efforts on three main
-components: a Java library for accessing and manipulating OBOE
-ontology extensions and semantic annotations, an annotation plugin for
-the Morpho metadata editor, and query extensions for the Metacat data
-catalog. Below we describe these components and give a brief
-overview of the OBOE model as well as the semantic annotation approach
-used in Semtools. For a more in-depth presentation of OBOE see
+components: a Java library to access and manipulate OBOE ontologies
+and semantic annotations, an annotation plugin for the Morpho metadata
+editor, and query extensions for the Metacat data catalog. Below we
+briefly describe the OBOE model, the semantic annotation approach used
+by Semtools, and the corresponding software components. For a more
+in-depth presentation of OBOE see
 \cite{madin07:_ontol_for_descr_and_synth,bowers08}.
